### 2014-01-06 doc/sfincsPaper/sfincsPaper-20131205-01.tex
Added documentation

%\documentclass{amsart}
%\documentclass[aip,graphicx]{revtex4-1}
\documentclass[12pt]{revtex4}
\usepackage{latexsym, color, graphicx, comment}
\usepackage{amssymb}
\usepackage{amsmath}

%\graphicspath{{../MATLAB/}}

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
\newcommand{\ExB}{\vect{E}\times\vect{B}}
\newcommand{\Cnl}{C_{\mathrm{nl}a}}
\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
\newcommand{\vE}{\vect{v}_{E}}
\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
\newcommand{\vEo}{\vect{v}_{E0}}
\newcommand{\vpar}{v_{||}}
\newcommand{\vperp}{v_{\bot}}
\newcommand{\rhop}{\rho_{\theta}}
\newcommand{\rT}{r_{T}}
\newcommand{\Ti}{T_{\mathrm{i}}}
\newcommand{\rTi}{r_{T\mathrm{i}}}
\newcommand{\rTe}{r_{T\mathrm{e}}}
\newcommand{\reta}{r_{\eta}}
\newcommand{\rn}{r_{n}}
\newcommand{\rperp}{r_\bot}
\newcommand{\fM}{f_{\mathrm{M}}}
\newcommand{\fMa}{f_{\mathrm{M}a}}
\newcommand{\fMi}{f_{\mathrm{M}i}}
\newcommand{\fMe}{f_{\mathrm{Me}}}
\newcommand{\fe}{f_{\mathrm{e}}}
\newcommand{\Te}{T_{\mathrm{e}}}
\newcommand{\vth}{v_{\mathrm{th}}}
\newcommand{\nuii}{\nu_{ii}}
\newcommand{\nuPrime}{\hat{\nu}}
\newcommand{\gradpar}{\nabla_{||}}
\newcommand{\kpar}{k_{||}}
\newcommand{\psia}{\psi_0}
\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\todo}[1]{\textcolor{red}{To do: #1}}
\newcommand{\changed}[1]{\textcolor{green}{#1}}
\newcommand{\sgn}{\mathrm{sgn}}

\begin{document}

\title{Comparison of particle trajectories and collision operators for collisional transport in nonaxisymmetric plasmas}

% repeat the \author .. \affiliation  etc. as needed
% \email, \thanks, \homepage, \altaffiliation all apply to the current author.
% Explanatory text should go in the []'s,
% actual e-mail address or url should go in the {}'s for \email and \homepage.
% Please use the appropriate macro for the type of information

% \affiliation command applies to all authors since the last \affiliation command.
% The \affiliation command should follow the other information.

\author{M Landreman}
\email[]{mattland@umd.edu}
%\homepage[]{Your web page}
%\thanks{}
\affiliation{Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA}
\author{H M Smith}
\author{P Helander}
\affiliation{Max-Planck-Institut f\"{u}r Plasmaphysik, 17491 Greifswald, Germany}

% Collaboration name, if desired (requires use of superscriptaddress option in \documentclass).
% \noaffiliation is required (may also be used with the \author command).
%\collaboration{}
%\noaffiliation

\date{\today}

\begin{abstract}

Neoclassical phenomena are computed numerically for the LHD and W7-X stellarators using several variants of the drift-kinetic
equation, differing in the terms involving the radial electric field. It is found that
for electric fields below roughly one third of the resonant value, the different formulations give
nearly identical results, demonstrating the incompressible ExB-drift approximation
is quite accurate in this regime. 
However, near the electric field resonance,
the models yield substantially different results.
We also compare results for various collision operators.
In many cases it is found to be important that collisions conserve momentum,
while differences between the full linearized Fokker-Planck operator
and a simpler momentum-conserving model operator are often relatively minor.

\end{abstract}

\pacs{}% insert suggested PACS numbers in braces on next line

\maketitle %\maketitle must follow title, authors, abstract and \pacs

\section{Introduction}

One important difference between axisymmetric and nonaxisymmetric plasmas is that
neoclassical effects in the latter are much more sensitive to small values of the radial
electric field $E_r$. In axisymmetric plasmas, in order for the radial electric to modify the collisional ion heat flux
and other neoclassical phenomena,
the poloidal ion Mach number $(B / B_{pol}) |\vect{v}_E|/v_i$ must approach $\sim 1$,
since an $E_r$ of corresponding magnitude is required to modify the trapped region of phase space\cite{GrishaNeo}.
Here, $B$ is the magnetic field magnitude, $B_{pol}$ is the poloidal magnetic field,
$\vect{v}_E$ is the $\vect{E}\times\vect{B}$ drift, and $v_i=\sqrt{2 T_i/m_i}$ is the ion thermal speed.
However, in nonaxisymmetric plasmas, 
a much smaller value of $E_r$ can modify the collisional fluxes \cite{Galeev, HoKulsrud, BeidlerBigBenchmarking}.
The reason is that helically trapped particles experience a secular radial magnetic drift, and
whichever process first interrupts this radial motion will thereby determine the step size for radial diffusion.
When $E_r=0$, the radial magnetic drift is interrupted by collisions, which cause
the particle to gain parallel momentum and de-trap.
But if $E_r$ is sufficient for the poloidal $\vect{E}\times\vect{B}$
precession frequency to exceed the effective collisional detrapping rate, $\vect{E}\times\vect{B}$ precession
begins to carry helically trapped particles onto untrapped trajectories, thereby limiting the radial step size and transport.
This transition from collisional ($1/\nu$-regime) to $E_r$-limited ($\sqrt{\nu}$-regime) transport
typically occurs at values of $E_r$ for which the poloidal Mach number is still $\ll 1$, due to the low collisionality
in typical experiments. For this reason, stellarator transport is sensitive to small values of $E_r$.

However, including the physics of $\vect{E}\times\vect{B}$ precession in a $\delta f$ drift-kinetic equation (or code to
solve such an equation)
is complicated by several issues.
First, if a rigorous expansion in $\rho_* \ll 1$ is employed, 
$\vect{E}\times\vect{B}$ is formally excluded
when the usual drift ordering $\vect{v}_E \sim \rho_* v_i$ is used,
but the high-flow ordering $\vect{v}_E \sim \rho_*^0 v_i$ is
not a useful ordering either, since it leads to contradictions in a general nonaxisymmetric field \cite{HelanderRotation, SugamaHighFlow}.
Here, $\rho_* = \rho/L$ where $\rho$ is the ion gyroradius and $L$ is a typical
macroscopic scale length.
Second, if the $\vect{v}_{E}$ poloidal precession term
is included in a radially local, time-independent kinetic equation
for $\delta f$ (the departure of the distribution function from a Maxwellian),
unphysical constraints are placed on the distribution function,
as we will prove in section \ref{sec:sources}.
These constraints only appear when $E_r \ne 0$, meaning a small but nonzero
$E_r$ is a singular perturbation of the $E_r=0$ case.
These unphysical behaviors have been elininated in previous codes\cite{BeidlerBigBenchmarking}
by making the ad-hoc replacement $1/B^2 \to 1/\left< B^2 \right>$ (where $\left< \ldots \right>$ denotes
a flux surface average) in the $\vect{E}\times\vect{B}$ drift.
At the same time, variation in the particles' energy and pitch angle associated with $E_r$ is neglected.
These replacements and omissions are chosen so as to restore the variational form of the kinetic equation  \cite{DKES1, DKES2}.
These changes to the kinetic equation may be called the ``incompressible-$\vect{E}\times\vect{B}$-drift''
approximation \cite{Beidler}.
Some investigations have indicated that the incompressible-$\vect{E}\times\vect{B}$-drift
approximation may be reasonably accurate for small $E_r$ but a poor approximation
for larger $E_r$ \cite{Beidler, meMonoenergetic}.
This issue of which collisionless terms to include in the kinetic equation
is effectively a choice between particle trajectories, since the collisionless guiding center trajectories
are equivalent to the characteristic curves of the drift-kinetic equation.

\begin{comment}
If the typical drift ordering $\vect{v}_E \sim \rho_* v_i$ is used,
with the usual ordering of the collision frequency $\nu$ as $\sim \rho_*^0 v_i/L$,
and a rigorous ordering of terms in $\rho_*$ is made, collisionless detrapping is formally
excluded. Here, $\rho_* = \rho/L$ where $\rho$ is the ion gyroradius and $L$ is a typical
macroscopic scale length.  Collisionless detrapping cannot be included by
the alternative ordering $\vect{v}_E \sim v_i$,
since this ordering leads to contradictions in a general nonaxisymmetric field \cite{HelanderRotation, SugamaHighFlow}.
In many stellarator neoclassical calculations and codes, then, the $\rho_*$ ordering is not rigidly followed,
with the $v_E$ poloidal precession term

To include the physics of collisionless detrapping, $\vect{E}\times\vect{B}$-drift terms
must be retained in the kinetic equation even though these terms are smaller
than other terms in the equation according to the $\rho_* \ll 1$ expansion.
At the same time, the ad-hoc replacement $1/B^2 \to 1/\left< B^2 \right>$ (where $\left< \ldots \right>$ denotes
a flux surface average) is sometimes made in the $\vect{E}\times\vect{B}$ drift \cite{DKES1, DKES2},
since this replacement happens to restore the variational form of the kinetic equation.
This replacement may be called the ``incompressible-$\vect{E}\times\vect{B}$-drift''
approximation \cite{Beidler}.
...
You can think of these various terms as related to effective particle trajectories.
\end{comment}

Another limitation of many past stellarator neoclassical calculations
is that they are often performed with simplified models for collisions \todo{Cite other stellarator codes here}.
The linearized Fokker-Planck collision operator -- the most accurate linear operator available -- 
has been implemented in a variety of 
tokamak neoclassical codes \cite{Sauter0, Sauter, WongChan, NEOFP, speedGrids}.
However, due to the numerical challenge of the extra toroidal dimension in stellarators, 
many stellarator neoclassical codes retain only pitch-angle scattering collisions,
so coupling in the energy dimension is eliminated.
The pitch-angle scattering operator
lacks the momentum conservation property of the Fokker-Planck operator,
which is known to be important in many situations \cite{Briesemeister}.
Several techniques have been devised and implemented \cite{Taguchi2, SugamaNishimura, Maassberg, PENTA}
to effectively restore momentum conservation by post-processing
the transport coefficients obtained with a pure pitch-angle scattering operator,
but these methods will not exactly reproduce calculations with the full linearized Fokker-Planck operator.
The NEO-2 code has implemented the full linearized Fokker-Planck operator for
stellarator geometry \cite{Kernbichler1, Kernbichler2}, but using a field-line-tracing method
which makes it difficult to add the important effect of poloidal $\vect{E}\times\vect{B}$ precession.

Here, we describe a new stellarator neoclassical code
SFINCS (the Stellarator Fokker-Planck Iterative Neoclassical Conservative Solver)
that can be used to explore the aforementioned issues,
comparing various models for effective particle trajectories and collisions.
The code solves the 4D drift-kinetic equation for the distribution function,
retaining coupling in 2 spatial independent variables (toroidal and poloidal angle)
and 2 velocity independent variables (speed and pitch angle),
but neglecting radial coupling.
(For comparison, DKES \cite{DKES1, DKES2} is 3D since energy coupling is neglected, while FORTEC-3D \cite{FORTEC} is 5D since
radial coupling is retained.)
General nonaxisymmetric nested flux surface geometry is allowed,
one or more species may be included, and several models for collisions are available,
including the full inter-species linearized Fokker-Planck operator.
The incompressible-$\vect{E}\times\vect{B}$-drift trajectories are implemented,
as are several other options for trajectories that include the true $\vect{E}\times\vect{B}$ drift.
As we shall demonstrate, retaining the true form of the $\vect{E}\times\vect{B}$
drift comes at a cost, requiring sources/sinks in the kinetic equation
in order for the solutions to be well behaved.
While all of the various options for the particle trajectories have disadvantages,
SFINCS allows the three options to be compared.
As we will show in several calculations for the LHD and W7-X stellarators,
in many experimentally relevant cases,
the transport matrix elements are nearly identical for the various choices of particle trajectories.
However, differences between the trajectory models emerge when the radial electric field
grows comparable to the ``resonant'' value.

In the following section, we motivate the form of the kinetic equation solved by SFINCS,
and detail the three models for particle trajectories that will be compared.
Section \ref{sec:collisionOperators} describes several available models for collisions.
For several of the particle trajectory models,
additional sources/sinks and constraints must be included in the system of equations
for the equations to be well posed and for the solutions to be well behaved.
These issues are explored in section \ref{sec:sources}.
Details of the numerical implementation are given in section \ref{sec:numerics}.
Some of the numerical results presented are given in terms of a transport
matrix, which is defined in section \ref{sec:transportMatrix}.
The numerical results are presented in sections \ref{sec:ErComparison} and \ref{sec:collisionComparison},
in which we discuss the transport matrix elements for the geometries of the LHD and W7-X
stellarators, comparing a variety of assumptions about the particle trajectories and collision operator.
In section \ref{sec:conclusions} we discuss the results and conclude.

\todo{cite EUTERPE and FORTEC-3D}

\section{Kinetic equations}
\label{sec:equations}

We begin with the drift-kinetic equation (19) of Ref. \cite{Hazeltine}.
The standard drift ordering is applied at first: $\rho_* \ll 1$ where $\rho_* = \rho/L$,
$\vE/v_a \sim \rho_*$, $\partial/\partial t \sim \rho_*^2 v_a/L$, and $\nu_a \sim v_a/L$,
where $L$ is a typical scale length.
We expand the distribution function as $f_a = f_{a0} + f_{a1} + \ldots$.
The leading order distribution function $f_{a0}$ is taken to be a Maxwellian
that is constant on flux surfaces when expressed in terms of total energy
$W_a = v^2/2+Z_a e \Phi/m_a$:
\begin{equation}
f_{a0} = \eta_a(\psi) \left[ \frac{m_a}{2\pi T_a(\psi)}\right]^{3/2} \exp\left( -\frac{m_a W_a}{T_a(\psi)}\right).
\label{eq:Maxwellian}
\end{equation}
The mean flow of this Maxwellian is taken to be zero since, as argued in Refs. \cite{HelanderRotation, SugamaHighFlow}, sonic
flows are not permitted in a general stellarator.
Taking $f_{a1}/f_{a0}\sim\rho_*$, the terms of order $\sim\rho_* (v/L)f_{a0}$ in (19) of Ref. \cite{Hazeltine} are then
\begin{equation}
\vpar \vect{b} \cdot\left(\nabla f_{a1}\right)_{W_a,\mu}
-C_{a}
=-\left(\vma + \vE\right)\cdot\nabla\psi \left(\frac{\partial f_{a0}}{\partial \psi}\right)_{W_a}
+\frac{Z_a e}{m_a c}\vpar\vect{b}\cdot\frac{\partial\vect{A}}{\partial t} \frac{\partial f_{a0}}{\partial W_a}
\label{eq:kinetic1}
\end{equation}
where
\begin{equation}
v_{ma}\cdot\nabla\psi = \frac{m_a c v_{||}^2}{Z_a e B} \vect{b}\times (\vect{b}\cdot\nabla\vect{b})\cdot\nabla\psi
+ \frac{m_a c v_{\perp}^2}{2 Z_a e B^2} \vect{b}\times \nabla B \cdot\nabla\psi
=\frac{m_a c}{2 Z_a e B^2}\left(v_{||}^2 + \frac{v_{\bot}^2}{2}\right) \vect{b}\times \nabla B \cdot\nabla\psi
\end{equation}
(exactly true for any $\beta$) and $\vE=(c/B^2)\vect{B}\times\nabla\Phi$.
Here, $C_a$ is the collision term for species $a$, linearized about the Maxwellians (\ref{eq:Maxwellian}).
Subscripts on gradients and partial derivatives indicate the quantities held fixed,
and $\mu=v^2/(2B)$ is the magnetic moment.

Unfortunately, (\ref{eq:kinetic1}) does not contain the physics of $\vect{E}\times\vect{B}$ precession,
since the characteristic curves of this equation correspond only to motion along the magnetic field lines.
Consequently, important transport regimes such as the $\sqrt{\nu}$ regime cannot be obtained
using (\ref{eq:kinetic1}).
To retain $\vect{E}\times\vect{B}$ precession, we also keep the term $(\vE+\vma)\cdot\nabla f_{a1}$
in (\ref{eq:kinetic1}), even though according to the formal ordering it should
appear at next order.  A similar step is made in other stellarator neoclassical
calculations \cite{DKES1, DKES2}

As shown in Appendix C of Ref. \cite{usPedestal}, we may choose the gauge
for the electromagnetic potentials such that $-c^{-1}\vect{b}\cdot\partial\vect{A}/\partial t = \left<E_{||} B\right> B/\left<B^2\right>$
on the right-hand side of (\ref{eq:kinetic1}).
Here, angle brackets denote a flux surface average:
\begin{equation}
\label{eq:FSA}
\left\langle \ldots \right\rangle
=\frac{1}{{V}'}
\int_0^{2\pi }d\theta
\int_0^{2\pi }d\zeta
{\frac{(\ldots)}{\vect{B}\cdot \nabla \zeta }}
\end{equation}
where
${V}'=\int_0^{2\pi } d\theta \int_0^{2\pi } d\zeta / \vect{B}\cdot \nabla \zeta $.
Thus, (\ref{eq:kinetic1}) becomes
\begin{equation}
\left(\vpar \vect{b} + \vE + \vma\right) \cdot\left(\nabla f_{a1}\right)_{W_a,\mu}
-C_{a}
=-\left(\vma + \vE\right)\cdot\nabla\psi \left(\frac{\partial f_{a0}}{\partial \psi}\right)_{W_a}
+\frac{Z_a e}{T_a}\vpar\frac{B\left<E_{||} B\right>}{\left< B^2\right>} f_{a0}.
\label{eq:kinetic2}
\end{equation}
Even if the radial electric field is considered an input, this form of the kinetic equation remains
nonlinear in the unknowns since the $\nabla f_{a1}$ term depends on the variation of $\Phi$
on a flux surface, and this variation is an unknown like $f_{a1}$.

To make the problem linear, we make use of the fact that the potential is nearly a flux function.
We define $\Phi_0 = \left<\Phi\right>$ and $\Phi_1 = \Phi-\Phi_0$.
We assume $\Phi_1 \ll \Phi_0$,
and we will show shortly that this assumption is self-consistent.
Since $e \Phi_0/T_a \sim 1$ in the drift ordering, then $e \Phi_1/T_a \ll 1$.
We do not expand in the ion charge $Z_a$.
Equation (\ref{eq:Maxwellian}) then gives
$f_{a0} \approx F_{a}\left[ 1 - Z_a e \Phi_1/T_a\right]$ where
\begin{equation}
F_{a}=n_a(\psi) \left[\frac{m_a}{2\pi T_a(\psi)}\right]^{3/2}\exp\left(-\frac{m_a v^2}{2 T_a(\psi)}\right)
\end{equation}
and $n_a = \eta_a \exp(-Z_a e \Phi_0/T_a)$ is the leading order density.
We define the leading-order total energy
$W_{a0} =v^2/2+ Z_a e\Phi_0 /m_a$,
and leading-order $\vect{E}\times\vect{B}$ drift
$\vEo=(c/B^2)(d\Phi_0/d\psi)\vect{B}\times\nabla\psi$.
As the relative differences between $f_{a0}$ and $F_a$, between $W_a$ and $W_{a0}$, and between $\vE$ and $\vEo$
are all small, we may replace the former quantities with the latter ones in (\ref{eq:kinetic2}).
At the same time, we note
\begin{equation}
\frac{\vE\cdot\nabla\psi}{\vma\cdot\nabla\psi} \sim \frac{1}{\epsilon} \frac{Z_a e \Phi_1}{T_a}
\label{eq:radialExB}
\end{equation}
where $\epsilon$ is the relative variation of $B$ on a flux surface, and taking this ratio to be small, the $\vE\cdot\nabla\psi$
term in (\ref{eq:kinetic2}) may be neglected. Thus, we obtain
\begin{equation}
\left(\vpar \vect{b} + \vEo + \vma\right) \cdot\left(\nabla f_{a1}\right)_{W_{a0},\mu}
-C_{a}
=-(\vma\cdot\nabla\psi) \left(\frac{\partial F_{a}}{\partial \psi}\right)_{W_{a0}}
+\frac{Z_a e}{T_a}\vpar\frac{B\left<E_{||} B\right>}{\left< B^2\right>} F_{a}.
\label{eq:kinetic3}
\end{equation}
where $C_a$ is now the collision operator linearized about $F_a$ rather than $f_{a0}$,
\begin{equation}
\left(\frac{\partial F_{a}}{\partial \psi}\right)_{W_{a0}}
=\left[ \frac{1}{p_a}\frac{dp_a}{d\psi} + \frac{Z_a e}{T_a} \frac{d\Phi_0}{d\psi} + \left( x_a^2-\frac{5}{2}\right) \frac{1}{T_a} \frac{dT_a}{d\psi}\right]F_a,
\end{equation}
and $x_a = v/v_a$.
If $F_a$ and $\Phi_0$ are considered known, then (\ref{eq:kinetic3}) is now linear in the unknowns $f_{a1}$,
and $\Phi_1$ has decoupled from the kinetic equations.

We note that in some circumstances the ratio (\ref{eq:radialExB}) may not be small \cite{HoKulsrud},
particularly for impurities\cite{Jose} with $Z_a \gg 1$.
However, treating the ratio (\ref{eq:radialExB}) as finite leads to a kinetic equation
that is nonlinear in the unknowns.  We neglect these nonlinear effects of $\Phi_1$ in the present
linear study, but such effects will be important to examine in future work.

For numerical computations, it is convenient to use coordinates for which the ranges of allowed values
are independent of the other coordinates.  As $W_{a0}$ and $\mu$ do not have this property, it is convenient
to switch to coordinates $x_a$ and $\xi = \vpar/v$.  Carrying out this change of variables on the first term
of (\ref{eq:kinetic3}), we find
\begin{equation}
%\left(\vpar \vect{b} + \vEo + \vma\right)
\dot{\vect{r}} \cdot\left(\nabla f_{a1}\right)_{W_{a0},\mu}
=
\dot{\vect{r}}\cdot \left(\nabla f_{a1}\right)_{x_a,\xi}
+\dot{x}_a \left(\frac{\partial f_{a1}}{\partial x_a}\right)_{\vect{r},\xi}
+\dot{\xi}_a \left(\frac{\partial f_{a1}}{\partial \xi}\right)_{\vect{r},x_a}
\end{equation}
where
\begin{equation}
\dot{\vect{r}} = \vpar \vect{b} + \vEo + \vma,
\label{eq:rdot1}
\end{equation}
\begin{equation}
\dot{x}_a = (\vma\cdot\nabla\psi) \left(-\frac{x_a}{2T_a}\frac{dT_a}{d\psi} - \frac{Z_a e}{2T_a x_a}\frac{d\Phi_0}{d\psi}\right),
\label{eq:xdot1}
\end{equation}
and
\begin{equation}
\dot{\xi}_a = -\frac{1-\xi^2}{2 B \xi} \vpar \vect{b}\cdot\nabla B
+\xi(1-\xi^2) \frac{c}{2 B^3} \frac{d\Phi_0}{d\psi}\vect{B}\times\nabla\psi\cdot\nabla B
-\frac{1-\xi^2}{2 B \xi} \vma\cdot\nabla B
.
\label{eq:xidot1}
\end{equation}
For the rest of this work, we will neglect the $\vma$ term in (\ref{eq:rdot1}), the $dT_a/d\psi$ term
in (\ref{eq:xdot1}), and the $\vma\cdot\nabla B$ term in (\ref{eq:xidot1}),
since these terms are small in the $\rho_*$ expansion and do not involve the radial electric field $d\Phi_0/d\psi$.
(The omitted terms may be important in other situations, but here our primary interest is the treatment of the $d\Phi_0/d\psi$ terms.)
Our kinetic equation then becomes
\begin{equation}
\dot{\vect{r}}\cdot \left(\nabla f_{a1}\right)_{x_a,\xi}
+\dot{x}_a \left(\frac{\partial f_{a1}}{\partial x_a}\right)_{\vect{r},\xi}
+\dot{\xi}_a \left(\frac{\partial f_{a1}}{\partial \xi}\right)_{\vect{r},x_a}
-C_{a}
=-(\vma\cdot\nabla\psi) \left(\frac{\partial F_{a}}{\partial \psi}\right)_{W_{a0}}
+\frac{Z_a e}{T_a}\vpar\frac{B\left<E_{||} B\right>}{\left< B^2\right>} F_{a}.
\label{eq:kinetic4}
\end{equation}
where the effective particle trajectory equations are
\begin{eqnarray}
\label{eq:fullTrajectories}
\dot{\vect{r}} &=& \vpar \vect{b} + \frac{c}{B^2}\frac{d\Phi_0}{d\psi} \vect{B}\times\nabla\psi,
\\
\dot{x}_a &=& -(\vma\cdot\nabla\psi) \frac{Z_a e}{2T_a x_a}\frac{d\Phi_0}{d\psi},
\nonumber \\
\dot{\xi}_a &=& -\frac{1-\xi^2}{2 B \xi} \vpar \vect{b}\cdot\nabla B
+\xi(1-\xi^2) \frac{c}{2 B^3} \frac{d\Phi_0}{d\psi}\vect{B}\times\nabla\psi\cdot\nabla B
.\nonumber
\end{eqnarray}
We will refer to (\ref{eq:fullTrajectories}) as the ``full trajectories.''

The $d\Phi_0/d\psi$ terms in $\dot{x}_a$ and $\dot{\xi}_a$ may be interpreted as a finite orbit width effect.
As a particle drifts radially, it experiences a varying electrostatic potential (even if the potential is a flux function.)
Thus the potential energy of the particle changes, so to maintain a constant total energy, the kinetic energy must change at
an equal and opposite rate, giving rise to the $d\Phi/d\psi$ term in $\dot{x}_a$.
Then to conserve $\mu$ while $v$ changes, $\xi$ must also change appropriately,
giving rise to the $d\Phi_0/d\psi$ term in $\dot{\xi}_a$.  Without these $d\Phi_0/d\psi$
terms in $\dot{x}_a$ and $\dot{\xi}$, $\mu$ will not be conserved,
whereas you can verify that $\mu$ is indeed conserved by (\ref{eq:fullTrajectories}).
Note that the $d\Phi_0/d\psi$ term in $\dot{\vect{r}}$
is the same order in the $\rho_*$ expansion as the
$d\Phi_0/d\psi$ terms in $\dot{x}_a$ and $\dot{\xi}_a$,
suggesting that if the former term is retained, the latter terms should be retained as well.

A large number of stellarator neoclassical codes \cite{DKES1, DKES2, BeidlerBigBenchmarking} effectively
solve (\ref{eq:kinetic4}) with the alternative trajectory equations
\begin{eqnarray}
\label{eq:DKESTrajectories}
\dot{\vect{r}} &=& \vpar \vect{b} + \frac{c}{\left<B^2\right>}\frac{d\Phi_0}{d\psi} \vect{B}\times\nabla\psi,
\\
\dot{x}_a &=& 0,
\nonumber \\
\dot{\xi}_a &=& -\frac{1-\xi^2}{2 B \xi} \vpar \vect{b}\cdot\nabla B
.\nonumber
\end{eqnarray}
We refer to these equations as the ``DKES trajectories,'' in light of their use in the widely applied code DKES \cite{DKES1, DKES2}.
These trajectories differ from (\ref{eq:fullTrajectories}) both in the neglect of the $d\Phi_0/d\psi$ terms in $\dot{x}_a$ and $\dot{\xi}_a$,
and in the replacement $B^2 \to \left< B^2 \right>$ in $\dot{\vect{r}}$.
The motivation for approximating the $\vect{E}\times\vect{B}$ drift in this matter will be clarified in section \ref{sec:sources}.
As shown in Refs. \cite{meMonoenergetic, finiteErOmnigenous}, in a symmetric magnetic field, the model (\ref{eq:DKESTrajectories}) possesses a conserved quantity
which is equal to $\mu$ when $d\Phi_0/d\psi=0$ but which differs from $\mu$ when $d\Phi_0/d\psi\neq 0$.

For comparison, we will also consider the following set of trajectory equations:
\begin{eqnarray}
\label{eq:partialTrajectories}
\dot{\vect{r}} &=& \vpar \vect{b} + \frac{c}{B^2}\frac{d\Phi_0}{d\psi} \vect{B}\times\nabla\psi,
\\
\dot{x}_a &=& 0,
\nonumber \\
\dot{\xi}_a &=& -\frac{1-\xi^2}{2 B \xi} \vpar \vect{b}\cdot\nabla B
\nonumber
\end{eqnarray}
which will be referred to as the ``partial trajectories.''  Equations (\ref{eq:partialTrajectories})
represent an intermediate step between (\ref{eq:DKESTrajectories}) and (\ref{eq:fullTrajectories}),
in that (\ref{eq:partialTrajectories}) includes the correct $\vect{E}\times\vect{B}$ drift, but not the
$d\Phi_0/d\psi$ terms in $\dot{x}_a$ and $\dot{\xi}_a$ required to conserve $\mu$.

For all three trajectory models, the quasineutrality equation is effectively decoupled from the kinetic equation (\ref{eq:kinetic4}).
%and it determines $\Phi_1$.
At leading order, quasineutrality implies $\sum_a Z_a n_a = 0$.
At next order, noting that both $f_{a0}$ and $f_{a1}$ contribute to density variation on a flux surface,
\begin{equation}
\sum_a \left( -\frac{Z_a^2 e \Phi_1}{T_a}n_a + Z_a \int d^3v\, f_{a1} \right) = 0.
\end{equation}
This equation may be solved for $\Phi_1$, giving the variation of the potential on a flux surface.
It follows that $e \Phi_1 / T_a \sim f_{a1} / f_{a0} \sim \rho_*$, so our earlier assumption that
$e \Phi_1 / T_a \ll 1$ is self-consistent.

\section{Collision operators}
\label{sec:collisionOperators}

\todo{write this section.}

\section{Conservation properties and sources}
\label{sec:sources}

If one attempts to solve the kinetic equation (\ref{eq:kinetic4}) numerically using either the full or partial trajectories
and $E_r \neq 0$, unphysical results will be obtained, with
the numerical solution not converging as resolution parameters are increased.
We now explore the reason for this behavior.
We will then describe a modified form of the kinetic equation which robustly produces more sensible results.

Consider the result of applying the operation
\begin{equation}
\left< \int d^3v(\ldots)\right>
\label{eq:massConservation}
\end{equation}
to the kinetic equation (\ref{eq:kinetic4}) for each of the trajectory models (\ref{eq:fullTrajectories})-(\ref{eq:partialTrajectories}).
This operation annihilates the streaming and mirror terms, the collision operator, and the
inhomogeneous drive terms.
The operation (\ref{eq:massConservation}) effectively produces a flux-surface-averaged
mass conservation equation for each model.  For the full trajectories and DKES trajectories,
the $d\Phi_0/d\psi$ terms are also annihilated by
(\ref{eq:massConservation}), so the
resulting mass conservation equation is just $0=0$. However, for the partial trajectories, the $d\Phi_0/d\psi$
term ($\vEo\cdot\nabla f_{a1}$) is not annihilated by (\ref{eq:massConservation}),
leaving
\begin{equation}
c \left< \frac{1}{B^2}
\vect{B}\times\nabla\psi\cdot\nabla
\int d^3v\, f_{a1}\right> \frac{d\Phi_0}{d\psi}=0.
\label{eq:unphysicalMassConservation}
\end{equation}
Thus, a nonzero $d\Phi_0/d\psi$ gives a singular perturbation to the $d\Phi_0/d\psi=0$ limit in this partial trajectory model:
the $d\Phi_0/d\psi=0$ solution for $f_{a1}$ need not satisfy 
$\left< (1/B^2) \vect{B}\times\nabla\psi\cdot\nabla\int d^3v\, f_{a1}\right>=0$,
so $f_{a1}$ must change dramatically as $E_r$ is raised from 0 to a small nonzero value,
a behavior which is unphysical.
When $d\Phi_0/d\psi \not= 0$, (\ref{eq:unphysicalMassConservation}) constrains $f_{a1}$ in an unphysical
manner, for there is no analogue to (\ref{eq:unphysicalMassConservation}) in the more accurate
averaged fluid mass conservation equation $0=\left< \partial N_a/\partial t + \nabla\cdot(N_a \vect{V}_a)\right>$
(i.e. the moment of the full Fokker-Planck equation with no expansion in $\rho_*$ or other parameters),
where $N_a$ and $\vect{V}_a$ are the full fluid density and velocity.
The unphysical nature of (\ref{eq:unphysicalMassConservation}) can also be seen from the fact that when the
$d\Phi_0/d\psi$ terms in $\dot{x}_a$ and $\dot{\xi}_a$ are retained in the more accurate trajectories (\ref{eq:fullTrajectories}),
these terms precisely cancel (\ref{eq:unphysicalMassConservation}).

Similarly, we can obtain an averaged energy conservation equation for each trajectory model by applying the operation
\begin{equation}
\sum_a \left< \int d^3v\frac{m_a v^2}{2}(\ldots)\right>
\end{equation}
to (\ref{eq:kinetic4}). Again, the result is $0=0$ for the DKES trajectories. However,
this time both the full and partial trajectory models give nonzero results: the partial trajectories give
\begin{equation}
c \sum_a\left< \frac{1}{B^2}
\vect{B}\times\nabla\psi\cdot\nabla
\int d^3v \frac{m_a v^2}{2} f_{a1}\right> \frac{d\Phi_0}{d\psi}=0.
\label{eq:unphysicalEnergyConservation1}
\end{equation}
and the full trajectories give
\begin{equation}
-c \sum_a\left< \frac{1}{B^2}
\vect{B}\times\nabla\psi\cdot\nabla
\int d^3v \frac{m_a v^2}{2}\frac{(1+\xi^2)}{2} f_{a1}\right> \frac{d\Phi_0}{d\psi}=0.
\label{eq:unphysicalEnergyConservation2}
\end{equation}
The quantity in (\ref{eq:unphysicalEnergyConservation2})
without the $d\Phi_0/d\psi$ factor is proportional to the radial current
$\sum_a Z_a \left< \int d^3v\, f_{a1} v_{ma}\cdot\nabla\psi\right>$,
so it vanishes naturally when $E_r$ is at the ambipolar value.  However, as
the radial current would usually not be zero when $E_r=0$, (\ref{eq:unphysicalEnergyConservation2})
again implies a small nonzero $E_r$ would be a singular perturbation of the
$E_r=0$ limit.

One motivation for use of the DKES trajectory model is now apparent:
it is the only model that avoids the imposition of one or more unphysical
constraints on the distribution function when $d\Phi_0/d\psi \ne 0$,
constraints which cause an $E_r \ne 0$ calculation to be a singular perturbation
of an $E_r = 0$ calculation.
\begin{comment}
if one is forced to drop energy coupling (for reasons of numerical efficiency) by 
neglecting $\dot{x}_a$, unphysical constraints 
would be placed on the distribution function unless the $1/B^2  \to 1/\left< B^2 \right>$
replacement was made.
\end{comment}

The aforementioned problems with the partial and full trajectory models may be eliminated 
in the following manner.
The kinetic equation becomes well behaved if we introduce small particle and heat sources
\begin{equation}
S_a(\psi, v) = S_{ap}(\psi) F_{a}(\psi,v) \left[ x_a^2 - \frac{5}{2}\right] + S_{ah}(\psi) F_a(\psi, v) \left[x_a^2 - \frac{3}{2}\right]
\label{eq:sources}
\end{equation}
where $S_{ap}$ and $S_{ah}$ are considered to be unknowns.
(The $v$ dependence in (\ref{eq:sources}) is chosen so $S_{ap}$ provides a particle source but no heat source, while $S_{ah}$ provides
a heat source but no particle source.)
As these two new unknowns are now included in the system of equations on each flux surface, we must supply an equal number of additional constraints.
The constraints we supply are $\left< \int d^3v\, f_{a1}\right>=0$ and $\left< \int d^3v\, v^2 f_{a1}\right>=0$,
the sensible requirements that all the flux-surface-averaged density and pressure reside in $F_a$ rather than $f_{a1}$.
This method of modifying the system of equations has the added benefit of eliminating the null solutions $f_{a1} = F_a$ and $f_{a1} = F_a v^2$ 
even in the $d\Phi_0/d\psi=0$ limit.
When $S_a$ is included in the kinetic equation, new terms proportional to $S_{ap}$ and/or $S_{ah}$
now appear in the mass and energy conservation equations such as (\ref{eq:unphysicalMassConservation})-(\ref{eq:unphysicalEnergyConservation2}).
These conservation equations imply that when $d\Phi_0/d\psi=0$, $S_{ap}$ and $S_{ah}$ must vanish.
However, now when $d\Phi_0/d\psi$ is increased from 0 to a small finite number, the sources can turn on
to satisfy  (\ref{eq:unphysicalMassConservation})-(\ref{eq:unphysicalEnergyConservation2}),
eliminating the singular perturbation in $f_{a1}$.  We find that numerical results are then well behaved,
converging appropriately as numerical resolution parameters are increased, and smoothly going to the
$E_r=0$ results as $E_r$ is decreased.

We do not claim that the method proposed here is an ideal solution: the sources (\ref{eq:sources})
are ad-hoc and are not derived rigorously. However, by the techniques proposed here, we can at least
compare the three different trajectory models, and for most experimentally relevant values of $E_r$, we will show
that the three models give nearly identical results.

The aforemention mass and energy sources and constraints work when either the Fokker-Planck or momentum-conserving model collision operators
are used, 
but not when the pure pitch-angle scattering collision operator is used, since in this case the kinetic equation has a larger null space
(any function of $v$ is a solution.)
For this collision operator, we instead allow a general source $S_a(\psi,v)$,
introducing $N_x$ unknowns on each flux surface, where $N_x$ is the number of grid points in $v$.
To supply an equal number of constraints, we impose the equations $\left< \int_{-1}^1 d\xi\, f_{a1}\right>=0$
at each grid point in $v$.

\section{Numerical implementation}
\label{sec:numerics}

The SFINCS code solves the drift-kinetic equation (\ref{eq:kinetic4})
with (\ref{eq:sources})
for the three trajectory models (\ref{eq:fullTrajectories})-(\ref{eq:partialTrajectories}),
for general nonaxisymmetric nested flux surface geometry, and for an arbitrary number of species.
SFINCS is based on the Fokker-Planck code described in Ref. \cite{speedGrids},
generalized to allow nonaxisymmetry.
Briefly, the kinetic equation is discretized using finite differences with a 5-point stencil in $\theta$
and $\zeta$, using a truncated Legendre modal expansion in $\xi$, and using
a spectral collocation method in $x_a$.
The time-independent kinetic equation is solved directly (by solving a single sparse linear system),
so the rate of convergence is not limited by the timescale of physical relaxation.
A preconditioned iterative Krylov solver is employed.
The modifications compared to the code of Ref. \cite{speedGrids} are the following.
(1) $f_{a1}$, $B$, and other geometric operators are allowed to depend on the toroidal angle $\zeta$,
and the numerical grid is expanded to include this new coordinate.
(2) The additional $d\Phi_0/d\psi$ terms in $\vect{\dot{r}}_a$, $\dot{x}_a$, and $\dot{\xi}_a$
are included.
(3) The additional collision operators discussed above are included.
(4) The extra constraint equations and sources are implemented as in (19) of Ref. \cite{PERFECT}.

Note that poloidal and toroidal magnetic drifts could be included in the kinetic equation
without increasing the density of the matrix, i.e. without increasing the computational expense
of the method here.  However, to include radial drifts acting on $f_{a1}$,
the number of independent variables would increase from 4 to 5 since different flux surfaces
would couple. This increase in dimensionality would be numerically challenging.

The magnetic geometry is specified in Boozer coordinates $\theta$ and $\zeta$, in which $\vect{B}$
is represented as
\begin{eqnarray}
\vect{B} &=&
\nabla \psi \times \nabla \theta + \iota \nabla\zeta\times\nabla\psi \\
&=& \beta(\psi,\theta,\zeta)\nabla\psi + I(\psi)\nabla\theta + G(\psi) \nabla\zeta.
\end{eqnarray}
Here, $\iota = 1/q$ is the rotational transform with $q$ the safety factor,
$cI/2$ is the toroidal current inside the flux surface, and $cG/2$ is the poloidal
current outside the flux surface.
The geometric operators needed in the kinetic equation are then
\begin{equation}
\vect{B}\cdot\nabla X = \left( \iota \frac{\partial X}{\partial\theta} + \frac{\partial X}{\partial\zeta}\right)\vect{B}\cdot\nabla\zeta
\end{equation}
and
\begin{equation}
\vect{B}\times\nabla\psi\cdot\nabla X = \left( G \frac{\partial X}{\partial\theta} -I \frac{\partial X}{\partial\zeta}\right)\vect{B}\cdot\nabla\zeta
\end{equation}
where $X$ can be any scalar quantity,
and the coordinate Jacobian is
$\vect{B}\cdot\nabla\zeta = B^2/(G+\iota I)$. Thus, the magnetic geometry
enters the kinetic equation only through the quantities $I$, $G$, $\iota$,
and $B(\theta,\zeta)$.

As a test of the code, we have verified the quasisymmetry isomorphism,
discussed in Refs. \cite{Pytte, Boozer83, meQS}.
In this test, the magnetic field is specified as $B(\theta,\zeta) = y(M\theta-N\zeta)$
for some given periodic function $y$ and integers $M$ and $N$.
The kinetic equation is solved for several values of $M$ and $N$,
varying the collision frequency with each run so that
$\nu_{ii} / \left|M - N/\iota \right|$ remains fixed.
As $M$ and $N$ are varied, $G$, $I$, $\iota$, $dn/d\psi$, $dT/d\psi$, and other input quantities are also kept fixed.
Under these circumstances, it is known analytically \cite{Pytte, Boozer83, meQS} that the flow
and heat flux should obey the transformations
\begin{equation}
\left< V_{||} B \right> \frac{M-N/\iota}{NI+MG} = C_V
\end{equation}
and
\begin{equation}
\left< \int d^3v\, f_{i1} \frac{v^2}{2}\vect{v}_{mi}\cdot\nabla\psi \right> \frac{\left| M-N/\iota\right| }{(NI+MG)^2} = C_q
\end{equation}
respectively, where the quantities $C_V$ and $C_q$ should be independent of $M$ and $N$.
It was verified that the SFINCS code obeyed these isomorphism transformations
for a range of $y$, collisionality regimes, and collision operators.

\section{Ion transport matrix}
\label{sec:transportMatrix}

For the case of ions in a pure plasma, it is instructive to examine the transport matrix $L_{jk}$, defined as follows:
\begin{equation}
\label{eq:matrix}
\left(\begin{array}{c}
\frac{Ze(G+\iota I)}{ncTG} \left< \int d^3v\, f \vm\cdot\nabla\psi\right> \\
\frac{Ze(G+\iota I)}{ncTG} \left< \int d^3v\, f \frac{mv^2}{2T}\vm\cdot\nabla\psi\right> \\
\frac{1}{v_i B_0} \left< B V_{||} \right>
\end{array}\right)
=
\left(\begin{array}{ccc}
L_{11} & L_{12} & L_{13} \\
L_{21} & L_{22} & L_{23} \\
L_{31} & L_{32} & L_{33}
\end{array}\right)
\left(\begin{array}{c}
\frac{GTc}{ZeB_0 v_{i}} \left[ \frac{1}{n}\frac{dn}{d\psi} + \frac{Ze}{T}\frac{d\Phi}{d\psi} -\frac{3}{2T}\frac{dT}{d\psi}\right] \\
\frac{GTc}{Ze B_0 v_{i} T}\frac{dT}{d\psi} \\
\frac{Ze}{T}(G+\iota I)\frac{\left<E_{||} B\right>}{\left<B^2\right>}
\end{array}\right)
\end{equation}
Here, $B_0$ is the $(0,0)$ Fourier mode amplitude of $B(\theta,\zeta)$, and we have dropped $i$ subscripts.
When the DKES trajectories (\ref{eq:DKESTrajectories}) are used, it can be shown that $L_{jk}$ is symmetric for any value of $E_r$.
When the trajectories (\ref{eq:fullTrajectories}) or (\ref{eq:partialTrajectories})
are used and $E_r=0$, $L_{jk}$ is symmetric as well. However, when the trajectories (\ref{eq:fullTrajectories}) or (\ref{eq:partialTrajectories}) are used and $E_r \neq 0$,
the transport matrix defined in this manner is generally not symmetric.

Different definitions of the transport matrix have been given elsewhere in the literature \cite{Beidler},
the definition here has several nice properties. First, the matrix is dimensionless. Second, $L_{jk}$ is symmetric (in the cases described above).
Third, $L_{jk}$ depends on the magnetic geometry and physical parameters only through
$B/B_0$, $I/G$, $\iota$, a normalized collisionality
\begin{equation}
\nu' = \frac{(G+\iota I)\nu_{ii}}{v_i B_0},
\end{equation}
and a normalized electric field
\begin{equation}
E_* = \frac{c G}{\iota v_i B_0}\frac{d\Phi_0}{d\psi},
\end{equation}
and not on any other individual parameters such as density, temperature, $I$, etc.
Typically, $I \ll G$ and $G \approx B_0 R$ where $R$ is the major radius of the device,
so $\nu' \approx \nu_{ii} R / v_i$.
In axisymmetry, $E_*$ corresponds to the poloidal Mach number: $E_* \approx (B / B_{pol}) | \vEo | / v_i$
where $B_{pol}$ is the poloidal magnetic field.
Therefore, $E_*$ corresponds to the electric field normalized by the so-called resonant electric field \cite{Beidler}.

Several properties of the matrix $L_{jk}$ are noteworthy. Using the property of the linearized
ion-ion collision operator $\int d^3v (g/F_i) C_{ii}\{g\} \le 0$
for any $g$, then $\sgn(L_{11}) = \sgn(L_{22}) = -\sgn(L_{33}) = -\sgn((G+\iota I)/B_0)$.
\todo{I think this property is true for the FP and PAS operators but not for the model operator.
Confirm this. Also, for the FP and PAS operators, check this for $E_r \ne 0$ for the 3 trajectory models}.
Second, for all three trajectory models, the elements $L_{jk}$
are independent of the sign of the elecric field: $L_{jk}(E_*) = L_{jk}(-E_*)$,
assuming the stellarator symmetry property $B(\theta,\zeta) = B(-\theta,-\zeta)$
for some choice of the origin of $\theta$ and $\zeta$.
This symmetry of $L_{jk}$ follows from a symmetry in the kinetic equation:
if the signs of 
$\theta$, $\zeta$, $v_{||}$, and $d\Phi_0/d\psi$ are all reversed
in (\ref{eq:kinetic4}), the sign of $f_{i1}$ will reverse,
leaving the left-hand side of (\ref{eq:matrix}) unchanged.

\section{Comparison of $E_r$ terms}
\label{sec:ErComparison}

Figures (\ref{fig:ErComparison_LHD}) and (\ref{fig:ErComparison_W7X}) show a SFINCS computation of
the ion transport matrix elements for the 3 trajectory models
in two different stellarator geometries.
The calculations in figure (\ref{fig:ErComparison_LHD}) are performed for the $r/a=0.5$ surface of
the LHD stellarator \cite{LHD} in its standard configuration.
The calculations in figure (\ref{fig:ErComparison_W7X}) are performed for the $r/a=0.5$ surface of
the W7-X stellarator \cite{W7X1,W7X2} in its standard configuration.
In the LHD calculation, only the Boozer harmonics of $B(\theta,\zeta)/B_0$ with amplitude larger than $10^{-2}$
are retained, as listed in Table 1 of Ref. \cite{BeidlerBigBenchmarking}.
For both figures, the Fokker-Planck collision operator is used, and the collisionality
is set to $\nu'=0.01$.
As both figures illustrate, the electric field has negligible
effect on the transport matrix elements when $E_* < 0.01$.
For these small values of the electric field, the radial step size for diffusion is 
limited by collisions rather than by $\vect{E}\times\vect{B}$ precession.
As $E_r \to 0$, all the matrix elements converge smoothly to their $E_r=0$ limits.
For $E_*$ in the range $[0.01, 0.3]$,
the $\vect{E}\times\vect{B}$ precession suppresses radial transport,
as can be seen by the reduction in $|L_{11}|$ and $|L_{22}|$. In this regime of $E_*$, the three trajectory models give nearly
identical results for all the transport matrix elements.
However, once $E_*$ exceeds about 0.3, the results from the three trajectory models
begin to separate. 
%Roughly, the poloidal $\vect{E}\times\vect{B}$ will limit the radial step size for diffusion
%when $E_* > \nu'$, and this prediction is confirmed
\todo{Can we say anything about why the three trajectory models agree below the resonance? 
How large is $E_*$ likely to be in W7X and LHD?}

To ensure convergence with respect to the numerical resolution parameters $N_{\theta}$, $N_{\zeta}$, $N_{\xi}$ and $N_x$,
it was verified that the plotted curves did not change visibly on the scales shown
when each of these resolution parameters was doubled.

\todo{Add an example (with figure) involving more than 1 species. Perhaps compute $j_{bs}$ vs $E_r$ for different $E_r$ models at 1 or 2 values of nu*, perhaps assuming $L_n = L_{Te} = L_{Ti}, E||=0.$}

\begin{figure}[h!]
% Figure generated using m20131202_01_plotSFINCSErComparisonForPaper.m
% Data files:
% m20130324_01_SFINCS_ErScan_2013-03-27_03-59_ErScan_awesomer_mergeder.mat
% Code run on Matt's laptop
% *********************************
\includegraphics{m20131202_01_plotSFINCSErComparisonForPaper_LHD.eps}
%\includegraphics{m20131202_01_plotSFINCSErComparisonForPaper_LHD.pdf}
\caption{(Color online) Comparison of trajectory models for LHD standard geometry at $\nu' = 0.01$,
using linearized Fokker-Planck collisions.
\label{fig:ErComparison_LHD}}
\end{figure}

\begin{figure}[h!]
% Figure generated using m20131202_01_plotSFINCSErComparisonForPaper.m
% Data files:
% sfincs_2013-12-01_15-49_ErScan_noConvergence
% Code run on edison in 
% *********************************
\includegraphics{m20131202_01_plotSFINCSErComparisonForPaper_W7X.eps}
%\includegraphics{m20131202_01_plotSFINCSErComparisonForPaper_W7X.pdf}
\caption{(Color online) Comparison of trajectory models for W7X standard geometry at $\nu' = 0.01$,
using linearized Fokker-Planck collisions.
\todo{Fill in points for $E_*$ in the range 0.1 to 1.}
\label{fig:ErComparison_W7X}}
\end{figure}

\section{Comparison of collision operators}
\label{sec:collisionComparison}

Figures (\ref{fig:collisionComparison_LHD})-(\ref{fig:collisionComparison_W7X}) show the transport
matrix elements for the LHD and W7X geometries described earlier, this time comparing
the different collision operators as a function of collisionality.
The comparison is done for $d\Phi_0/d\psi=0$, so the three trajectory models
become identical, and the sources $S_a$ vanish.
It can be seen in the figures that at high collisionality,
momentum conservation is important for all the transport matrix elements (with the possible exception of $L_{22}$.)
At low collisionality, momentum conservation 
is unimportant for $L_{11}$, $L_{12}$, $L_{21}$, and $L_{22}$.
These matrix elements are associated with $1/\nu$-regime radial transport
(when $\nu' \ll 1$), which is associated with pitch-angle scattering of helically trapped particles.
Thus, the pitch-angle scattering approximation for collisions accurately
captures the dominant physics in these cases.
The other matrix elements are more sensitive to momentum conservation at low collisionality.
For all the matrix elements at all collisionalities, the momentum-conserving model operator
reproduces all the trends of the more accurate linearized Fokker-Planck operator,
though with some $O(1)$ differences.

Again, to ensure convergence with respect to the numerical resolution parameters $N_{\theta}$, $N_{\zeta}$, $N_{\xi}$ and $N_x$,
it was verified that the plotted curves did not change visibly on the scales shown
when each of these resolution parameters was doubled.
The required resolution in the $\zeta$ and $\xi$ coordinates increases substantially
as $\nu'$ decreases, due to the boundary layers that develop in the distribution function.

\begin{figure}[h!]
% Figure generated using m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper.m
% Data files:
% m20131117_04_sfincs_w7x_2013-11-26_02-27_nuPrimeScan_nuPrime0_01to100
% m20130226_04_SFINCS_2013-02-26_21-21_good
% W7X run was on edison in /global/scratch2/sd/landrema/20131123-01-sfincs/20131125-02-nuPrimeScan0_01to100
% *********************************
\includegraphics{m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper_LHD.eps}
%\includegraphics{m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper_LHD.pdf}
\caption{(Color online) Comparison of collision operators for LHD standard geometry at $E_r=0$.
\todo{Extend plots to $\nu' = 10^{-3}$.}
\label{fig:collisionComparison_LHD}}
\end{figure}

\begin{figure}[h!]
% Figure generated using m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper.m
% Data files:
% m20131117_04_sfincs_w7x_2013-11-26_02-27_nuPrimeScan_nuPrime0_01to100
% m20130226_04_SFINCS_2013-02-26_21-21_good
% W7X run was on edison in /global/scratch2/sd/landrema/20131123-01-sfincs/20131125-02-nuPrimeScan0_01to100
% *********************************
\includegraphics{m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper_W7X.eps}
%\includegraphics{m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper_W7X.pdf}
\caption{(Color online) Comparison of collision operators for W7X standard geometry at $E_r=0$.
\todo{Extend plots to $\nu' = 10^{-3}$.}
\label{fig:collisionComparison_W7X}}
\end{figure}

%\todo{Add an example (with figure) involving more than 1 species. Perhaps compute $j_{bs}$ vs nu* for different collision operators at $E_r=0$, perhaps assuming $L_n = L_{Te} = L_{Ti}, E||=0.$}

\begin{comment}
\begin{table}
\centering
\begin{tabular}{c c c c c}
$\nu_n$ & $N_\theta$ & $N_\zeta$ & $N_\xi$ & $N_x$ \\
\hline
100 & 11 & 37 & 13 & 8 \\
10 & 13 & 35 & 12 & 7 \\
1 & 13 & 31 & 24 & 6 \\
0.1 & 11 & 37 & 37 & 5 \\
0.01 & 11 & 64 & 68 & 5 \\
0.001 & 11 & 83 & 122 & 5 \\
%\begin{tabular}{c c c c c c c}
%$\nu_n$ & 0.001 & 0.01 & 0.1 & 1 & 10 & 100 \\
\end{tabular}
\caption{Resolution parameters required for convergence in the W7-X examples.
\label{tab:resolution}}
\end{table}
\end{comment}

\section{Discussion and conclusions}
\label{sec:conclusions}

\todo{Write this section}

\begin{acknowledgments}
Some of the computer simulations presented here
used resources of the National Energy Research Scientific Computing Center (NERSC),
which is supported by the Office of Science of the U.S. Department of Energy under Contract No. DE-AC02-05CH11231.
M. L. was supported by the
Fusion Energy Postdoctoral Research Program
administered by the Oak Ridge Institute for Science and Education.
We are grateful to J. Geiger for providing the W7X equilibrium data.
\end{acknowledgments}

%\appendix
%\section{Any appendix would go here}

\bibliography{sfincsPaper}

\end{document}

% To do:
% * Add references to other stellarator codes in intro.
% * Write conclusions
% * Finish E* scan for W7X
% * Figure out multi-species model operator
% * Write section on collision operators
% * Should I include any results showing ripple?
% * Think of a good example involving >1 species.
% * Paragraph that L_jk is invariant under EStar -> - EStar, L_11 and L_22 are negative, and L12=L21 for full trajectories even when E \ne 0.

### 2014-01-06 doc/sfincsPaper/sfincsPaper.tex
Added documentation

--- 2014-01-06 doc/sfincsPaper/sfincsPaper-20131205-01.tex
@@
 \newcommand{\rT}{r_{T}}
+\newcommand{\vi}{v_{\mathrm{i}}}
+\newcommand{\me}{m_{\mathrm{e}}}
+\newcommand{\mi}{m_{\mathrm{i}}}
 \newcommand{\Ti}{T_{\mathrm{i}}}
@@
 \newcommand{\vth}{v_{\mathrm{th}}}
-\newcommand{\nuii}{\nu_{ii}}
+\newcommand{\nuii}{\nu_{\mathrm{ii}}}
 \newcommand{\nuPrime}{\hat{\nu}}
@@
 \newcommand{\sgn}{\mathrm{sgn}}
+\newcommand{\Erres}{E_r^{\mathrm{res}}}
+\newcommand{\Bpol}{B_{\mathrm{pol}}}
+\newcommand{\Sap}{S_{a\mathrm{p}}}
+\newcommand{\Sah}{S_{a\mathrm{h}}}
 
@@
 \affiliation{Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA}
+\affiliation{Plasma Science and Fusion Center, Massachusetts Institute of Technology, Cambridge, MA, 02139, USA}
 \author{H M Smith}
@@
 
-Neoclassical phenomena are computed numerically for the LHD and W7-X stellarators using several variants of the drift-kinetic
-equation, differing in the terms involving the radial electric field. It is found that
+In this work, we test the validity of several common simplifying assumptions used in numerical neoclassical calculations
+for nonaxisymmetric plasmas.
+First, neoclassical phenomena are computed for the LHD and W7-X stellarators using several versions of the drift-kinetic
+equation, 
+including the commonly used incompressible-$\vect{E}\times\vect{B}$-drift approximation
+and two other variants,
+corresponding to different effective particle trajectories.
+It is found that
 for electric fields below roughly one third of the resonant value, the different formulations give
-nearly identical results, demonstrating the incompressible ExB-drift approximation
+nearly identical results, demonstrating the incompressible $\vect{E}\times\vect{B}$-drift approximation
 is quite accurate in this regime. 
@@
 the models yield substantially different results.
-We also compare results for various collision operators.
-In many cases it is found to be important that collisions conserve momentum,
-while differences between the full linearized Fokker-Planck operator
-and a simpler momentum-conserving model operator are often relatively minor.
+We also compare results for various collision operators, including the full linearized Fokker-Planck operator.
+At low collisionality, the radial transport driven by radial
+gradients is nearly identical for the different operators,
+while in other cases it is found to be important that collisions conserve momentum.
+%At high collisionality, it seems necessary to use the full Fokker-Planck operator to obtain all transport coefficients correctly.
 
@@
 
-One important difference between axisymmetric and nonaxisymmetric plasmas is that
-neoclassical effects in the latter are much more sensitive to small values of the radial
+One important difference between axisymmetric and nonaxisymmetric plasmas is
+that neoclassical effects in the latter are more sensitive to small values of the radial
 electric field $E_r$. In axisymmetric plasmas, in order for the radial electric to modify the collisional ion heat flux
 and other neoclassical phenomena,
-the poloidal ion Mach number $(B / B_{pol}) |\vect{v}_E|/v_i$ must approach $\sim 1$,
+the poloidal ion Mach number $(B / \Bpol) |\vect{v}_E|/\vi$ must approach $\sim 1$,
 since an $E_r$ of corresponding magnitude is required to modify the trapped region of phase space\cite{GrishaNeo}.
-Here, $B$ is the magnetic field magnitude, $B_{pol}$ is the poloidal magnetic field,
-$\vect{v}_E$ is the $\vect{E}\times\vect{B}$ drift, and $v_i=\sqrt{2 T_i/m_i}$ is the ion thermal speed.
+Here, $B$ is the magnetic field magnitude, $\Bpol$ is the poloidal magnetic field,
+$\vect{v}_E$ is the $\vect{E}\times\vect{B}$ drift, and $\vi=\sqrt{2 \Ti / \mi}$ is the ion thermal speed.
 However, in nonaxisymmetric plasmas, 
@@
 precession frequency to exceed the effective collisional detrapping rate, $\vect{E}\times\vect{B}$ precession
-begins to carry helically trapped particles onto untrapped trajectories, thereby limiting the radial step size and transport.
+begins to carry helically trapped particles onto untrapped trajectories, 
+and also confines the trapped orbits by convecting them (usually poloidally) around the torus,
+thereby limiting the radial step size and transport.
 This transition from collisional ($1/\nu$-regime) to $E_r$-limited ($\sqrt{\nu}$-regime) transport
 typically occurs at values of $E_r$ for which the poloidal Mach number is still $\ll 1$, due to the low collisionality
-in typical experiments. For this reason, stellarator transport is sensitive to small values of $E_r$.
-
-However, including the physics of $\vect{E}\times\vect{B}$ precession in a $\delta f$ drift-kinetic equation (or code to
+in typical experiments. (Here, $\nu$ denotes a collision frequency.)
+For this reason, stellarator transport at low collisionality is sensitive to small values of $E_r$.
+A variety of codes have been developed to compute 
+these neoclassical effects in stellarators \cite{DKES1, DKES2, GSRAKE, Tribaldos2001, PENTA, Kernbichler1, FORTEC_PFR, FORTEC_CPC, JOSE}.
+
+However, including the physics of $\vect{E}\times\vect{B}$ precession in a $\delta\! f$ drift-kinetic equation (or code to
 solve such an equation)
@@
 First, if a rigorous expansion in $\rho_* \ll 1$ is employed, 
-$\vect{E}\times\vect{B}$ is formally excluded
-when the usual drift ordering $\vect{v}_E \sim \rho_* v_i$ is used,
-but the high-flow ordering $\vect{v}_E \sim \rho_*^0 v_i$ is
+$\vect{E}\times\vect{B}$ precession is formally excluded
+when the usual drift ordering $\vect{v}_E \sim \rho_* \vi$ is used,
+but the high-flow ordering $\vect{v}_E \sim \vi$ is
 not a useful ordering either, since it leads to contradictions in a general nonaxisymmetric field \cite{HelanderRotation, SugamaHighFlow}.
@@
 is included in a radially local, time-independent kinetic equation
-for $\delta f$ (the departure of the distribution function from a Maxwellian),
+for $\delta\! f$ (the departure of the distribution function from a Maxwellian),
 unphysical constraints are placed on the distribution function,
-as we will prove in section \ref{sec:sources}.
+as we will prove in section \ref{sec:sources}
+by considering appropriate moments of the kinetic equation.
 These constraints only appear when $E_r \ne 0$, meaning a small but nonzero
 $E_r$ is a singular perturbation of the $E_r=0$ case.
-These unphysical behaviors have been elininated in previous codes\cite{BeidlerBigBenchmarking}
+These unphysical behaviors have been eliminated in previous codes\cite{BeidlerBigBenchmarking}
 by making the ad-hoc replacement $1/B^2 \to 1/\left< B^2 \right>$ (where $\left< \ldots \right>$ denotes
@@
 
-\begin{comment}
-If the typical drift ordering $\vect{v}_E \sim \rho_* v_i$ is used,
-with the usual ordering of the collision frequency $\nu$ as $\sim \rho_*^0 v_i/L$,
-and a rigorous ordering of terms in $\rho_*$ is made, collisionless detrapping is formally
-excluded. Here, $\rho_* = \rho/L$ where $\rho$ is the ion gyroradius and $L$ is a typical
-macroscopic scale length.  Collisionless detrapping cannot be included by
-the alternative ordering $\vect{v}_E \sim v_i$,
-since this ordering leads to contradictions in a general nonaxisymmetric field \cite{HelanderRotation, SugamaHighFlow}.
-In many stellarator neoclassical calculations and codes, then, the $\rho_*$ ordering is not rigidly followed,
-with the $v_E$ poloidal precession term
-
-To include the physics of collisionless detrapping, $\vect{E}\times\vect{B}$-drift terms
-must be retained in the kinetic equation even though these terms are smaller
-than other terms in the equation according to the $\rho_* \ll 1$ expansion.
-At the same time, the ad-hoc replacement $1/B^2 \to 1/\left< B^2 \right>$ (where $\left< \ldots \right>$ denotes
-a flux surface average) is sometimes made in the $\vect{E}\times\vect{B}$ drift \cite{DKES1, DKES2},
-since this replacement happens to restore the variational form of the kinetic equation.
-This replacement may be called the ``incompressible-$\vect{E}\times\vect{B}$-drift''
-approximation \cite{Beidler}.
-...
-You can think of these various terms as related to effective particle trajectories.
-\end{comment}
-
 Another limitation of many past stellarator neoclassical calculations
-is that they are often performed with simplified models for collisions \todo{Cite other stellarator codes here}.
+is that they are often performed with simplified models for collisions.
 The linearized Fokker-Planck collision operator -- the most accurate linear operator available -- 
@@
 The NEO-2 code has implemented the full linearized Fokker-Planck operator for
-stellarator geometry \cite{Kernbichler1, Kernbichler2}, but using a field-line-tracing method
+stellarator geometry \cite{Kernbichler1}, but using a field-line-tracing method
 which makes it difficult to add the important effect of poloidal $\vect{E}\times\vect{B}$ precession.
@@
 comparing various models for effective particle trajectories and collisions.
+Although we use the terminology of ``effective trajectories,'' the code
+uses continuum rather than Monte Carlo algorithms.
 The code solves the 4D drift-kinetic equation for the distribution function,
@@
 but neglecting radial coupling.
-(For comparison, DKES \cite{DKES1, DKES2} is 3D since energy coupling is neglected, while FORTEC-3D \cite{FORTEC} is 5D since
+(For comparison, DKES \cite{DKES1, DKES2} is 3D since energy coupling is neglected, 
+while FORTEC-3D \cite{FORTEC_PFR, FORTEC_CPC} is 5D since
 radial coupling is retained.)
@@
 and detail the three models for particle trajectories that will be compared.
-Section \ref{sec:collisionOperators} describes several available models for collisions.
 For several of the particle trajectory models,
@@
 
-\todo{cite EUTERPE and FORTEC-3D}
-
 \section{Kinetic equations}
@@
 We begin with the drift-kinetic equation (19) of Ref. \cite{Hazeltine}.
-The standard drift ordering is applied at first: $\rho_* \ll 1$ where $\rho_* = \rho/L$,
-$\vE/v_a \sim \rho_*$, $\partial/\partial t \sim \rho_*^2 v_a/L$, and $\nu_a \sim v_a/L$,
-where $L$ is a typical scale length.
+The standard drift ordering is applied at first: $\rho_{*a} \ll 1$ where $\rho_{*a} = \rho_a/L$,
+$\vE/v_a \sim \rho_{*a}$, $\partial/\partial t \sim \rho_{*a}^2 v_a/L$, and $\nu_a \sim v_a/L$.
+Here, $v_a = \sqrt{2 T_a / m_a}$ is the thermal speed of species $a$, $T_a$ is the temperature,
+$m_a$ is the mass, $\rho_a = v_a m_a c/(Z_a e B)$ is the gyroradius, $Z_a$ is the species charge in units
+of the proton charge $e$, $c$ is the speed of light, $L$ is a typical scale length,
+and $\nu_a$ is a collision frequency.
+No expansion in mass ratios or charges is made.
 We expand the distribution function as $f_a = f_{a0} + f_{a1} + \ldots$.
@@
 \end{equation}
+Here, $\Phi$ is the electrostatic potential and $v$ is the speed.
 The mean flow of this Maxwellian is taken to be zero since, as argued in Refs. \cite{HelanderRotation, SugamaHighFlow}, sonic
 flows are not permitted in a general stellarator.
-Taking $f_{a1}/f_{a0}\sim\rho_*$, the terms of order $\sim\rho_* (v/L)f_{a0}$ in (19) of Ref. \cite{Hazeltine} are then
+Taking $f_{a1}/f_{a0}\sim\rho_{*a}$, the terms of order $\sim\rho_{*a} (v/L)f_{a0}$ in (19) of Ref. \cite{Hazeltine} are then
 \begin{equation}
@@
 \end{equation}
-where
-\begin{equation}
-v_{ma}\cdot\nabla\psi = \frac{m_a c v_{||}^2}{Z_a e B} \vect{b}\times (\vect{b}\cdot\nabla\vect{b})\cdot\nabla\psi
+where the radial magnetic drift is
+\begin{equation}
+\vma\cdot\nabla\psi = \frac{m_a c v_{||}^2}{Z_a e B} \vect{b}\times (\vect{b}\cdot\nabla\vect{b})\cdot\nabla\psi
 + \frac{m_a c v_{\perp}^2}{2 Z_a e B^2} \vect{b}\times \nabla B \cdot\nabla\psi
@@
 \end{equation}
-(exactly true for any $\beta$) and $\vE=(c/B^2)\vect{B}\times\nabla\Phi$.
-Here, $C_a$ is the collision term for species $a$, linearized about the Maxwellians (\ref{eq:Maxwellian}).
+(exactly true for any $\beta$ in a magnetic equilibrium with isotropic pressure) and the $\vect{E}\times\vect{B}$ drift
+is $\vE=(c/B^2)\vect{B}\times\nabla\Phi$.
+Here, $\vect{b} = \vect{B}/B$ is the unit vector along the magnetic field, 
+$v_{||}$ and $v_{\bot}$ denote the components of velocity parallel and perpendicular to $\vect{B}$,
+$2\pi\psi$ is the toroidal flux, $\vect{A}$ is the magnetic vector potential, 
+and $C_a$ is the collision term for species $a$, linearized about the Maxwellians (\ref{eq:Maxwellian}).
 Subscripts on gradients and partial derivatives indicate the quantities held fixed,
-and $\mu=v^2/(2B)$ is the magnetic moment.
+and $\mu=v_{\bot}^2/(2B)$ is the magnetic moment.
 
@@
 appear at next order.  A similar step is made in other stellarator neoclassical
-calculations \cite{DKES1, DKES2}
+calculations \cite{DKES1, DKES2}.
+The mathematical reason why this term is important at low collisionality is that it has different symmetry properties than other, possible larger, terms in (\ref{eq:kinetic1}). For instance, it 
+survives if a bounce average is used to annihilate the first term.
+(We will not bounce average the kinetic equation here,
+but when the collisionality is low, the solution of the full equation becomes
+asymptotically close to the solution of the bounce-averaged equation.)
 
@@
 where
-${V}'=\int_0^{2\pi } d\theta \int_0^{2\pi } d\zeta / \vect{B}\cdot \nabla \zeta $.
+${V}'=\int_0^{2\pi } d\theta \int_0^{2\pi } d\zeta / \vect{B}\cdot \nabla \zeta $,
+$\theta$ and $\zeta$ are poloidal and toroidal magnetic angles satisfying
+\begin{equation}
+\vect{B} = \nabla \psi \times \nabla \theta + \iota \nabla\zeta\times\nabla\psi,
+\end{equation}
+$\iota = 1/q$ is the rotational transform, and $q$ is the safety factor.
 Thus, (\ref{eq:kinetic1}) becomes
@@
 
-To make the problem linear, we make use of the fact that the potential is nearly a flux function.
+To make the problem linear, we make use of the fact that the electrostatic potential is nearly a flux function.
 We define $\Phi_0 = \left<\Phi\right>$ and $\Phi_1 = \Phi-\Phi_0$.
@@
 \end{equation}
-where $\epsilon$ is the relative variation of $B$ on a flux surface, and taking this ratio to be small, the $\vE\cdot\nabla\psi$
+where $\epsilon$ is the relative variation of $B$ on a flux surface, and taking the ratio (\ref{eq:radialExB}) to be small, the $\vE\cdot\nabla\psi$
 term in (\ref{eq:kinetic2}) may be neglected. Thus, we obtain
@@
 We note that in some circumstances the ratio (\ref{eq:radialExB}) may not be small \cite{HoKulsrud},
-particularly for impurities\cite{Jose} with $Z_a \gg 1$.
+particularly for impurities \cite{JOSE} with $Z_a \gg 1$.
 However, treating the ratio (\ref{eq:radialExB}) as finite leads to a kinetic equation
@@
 \begin{equation}
-%\left(\vpar \vect{b} + \vEo + \vma\right)
 \dot{\vect{r}} \cdot\left(\nabla f_{a1}\right)_{W_{a0},\mu}
@@
 +\dot{x}_a \left(\frac{\partial f_{a1}}{\partial x_a}\right)_{\vect{r},\xi}
-+\dot{\xi}_a \left(\frac{\partial f_{a1}}{\partial \xi}\right)_{\vect{r},x_a}
-\end{equation}
-where
++\dot{\xi}_a \left(\frac{\partial f_{a1}}{\partial \xi}\right)_{\vect{r},x_a},
+\end{equation}
+where $\vect{r}$ denotes the position vector,
 \begin{equation}
@@
 in (\ref{eq:xdot1}), and the $\vma\cdot\nabla B$ term in (\ref{eq:xidot1}),
-since these terms are small in the $\rho_*$ expansion and do not involve the radial electric field $d\Phi_0/d\psi$.
-(The omitted terms may be important in other situations, but here our primary interest is the treatment of the $d\Phi_0/d\psi$ terms.)
+for several reasons.
+First, if the  $\vma$ term in (\ref{eq:rdot1}) were retained, we would need to solve a 5D rather
+than 4D problem due to the radial coupling (i.e. $\psi$ appearing
+as a derivative rather than merely as a parameter). Second,
+once radial coupling is dropped, we must also drop the $dT_a/d\psi$ term
+in (\ref{eq:xdot1}) and the $\vma\cdot\nabla B$ term in (\ref{eq:xidot1})
+in order for $\mu$ to be conserved.
+Third, dropping these terms conveniently eliminates all dependence of the transport matrix (defined in section \ref{sec:transportMatrix})
+on $d T_a/d\psi$ and $dB/d\psi$.
+Fourth, we wish to focus on the effects of the radial electric field.
+The omitted terms may be important in other situations, but here our primary interest is the treatment of the $d\Phi_0/d\psi$ terms.
 Our kinetic equation then becomes
@@
 giving rise to the $d\Phi_0/d\psi$ term in $\dot{\xi}_a$.  Without these $d\Phi_0/d\psi$
-terms in $\dot{x}_a$ and $\dot{\xi}$, $\mu$ will not be conserved,
+terms in $\dot{x}_a$ and $\dot{\xi}_a$, $\mu$ will not be conserved,
 whereas you can verify that $\mu$ is indeed conserved by (\ref{eq:fullTrajectories}).
@@
 For all three trajectory models, the quasineutrality equation is effectively decoupled from the kinetic equation (\ref{eq:kinetic4}).
-%and it determines $\Phi_1$.
 At leading order, quasineutrality implies $\sum_a Z_a n_a = 0$.
@@
 \sum_a \left( -\frac{Z_a^2 e \Phi_1}{T_a}n_a + Z_a \int d^3v\, f_{a1} \right) = 0.
+\label{eq:quasineutrality}
 \end{equation}
 This equation may be solved for $\Phi_1$, giving the variation of the potential on a flux surface.
-It follows that $e \Phi_1 / T_a \sim f_{a1} / f_{a0} \sim \rho_*$, so our earlier assumption that
+It follows that $e \Phi_1 / T_a \sim f_{a1} / f_{a0} \sim \rho_{*a}$, so our earlier assumption that
 $e \Phi_1 / T_a \ll 1$ is self-consistent.
 
-\section{Collision operators}
-\label{sec:collisionOperators}
-
-\todo{write this section.}
+Several choices can be made for the collision operator.
+The most accurate linear option is the Fokker-Planck operator \cite{RMJ,PerBook}
+linearized about the Maxwellians: $C_a = \sum_b C_{ab}^{\ell}$,
+where $C_{ab}^{\ell} = C_{ab}\{f_{a1},F_{b}\} + C_{ab} \{ F_{a}, f_{b1}\}$
+and $C_{ab}$ is the full bilinear Fokker-Planck operator between species $a$ and $b$.
+This linearized operator may be written in many forms, and for numerical implementation,
+we find it convenient to use the form detailed in equations (14)-(16) of Ref. \cite{speedGrids}.
+
+A simpler option used in many codes is the pitch-angle scattering operator \cite{BeidlerBigBenchmarking}.
+This operator lacks several properties of the linearized Fokker-Planck operator,
+such as the momentum conservation property $\int d^3v \, v_{||} C_{aa}^{\ell}=0$.
+Several more accurate approximate operators have been used in the literature.
+One such operator we will consider later consists of the pitch-angle scattering operator
+plus an ad-hoc momentum-restoring term, given for the case of self-collisions
+by eq (3.69) in Ref. \cite{PerBook}.
 
@@
 \end{equation}
-The quantity in (\ref{eq:unphysicalEnergyConservation2})
-without the $d\Phi_0/d\psi$ factor is proportional to the radial current
-$\sum_a Z_a \left< \int d^3v\, f_{a1} v_{ma}\cdot\nabla\psi\right>$,
+The quantity multiplying $d\Phi_0/d\psi$ in (\ref{eq:unphysicalEnergyConservation2})
+is proportional to the radial current
+$\sum_a Z_a \left< \int d^3v\, f_{a1} \vma\cdot\nabla\psi\right>$,
 so it vanishes naturally when $E_r$ is at the ambipolar value.  However, as
@@
 of an $E_r = 0$ calculation.
-\begin{comment}
-if one is forced to drop energy coupling (for reasons of numerical efficiency) by 
-neglecting $\dot{x}_a$, unphysical constraints 
-would be placed on the distribution function unless the $1/B^2  \to 1/\left< B^2 \right>$
-replacement was made.
-\end{comment}
 
@@
 \begin{equation}
-S_a(\psi, v) = S_{ap}(\psi) F_{a}(\psi,v) \left[ x_a^2 - \frac{5}{2}\right] + S_{ah}(\psi) F_a(\psi, v) \left[x_a^2 - \frac{3}{2}\right]
+S_a(\psi, v) = \Sap(\psi) F_{a}(\psi,v) \left[ x_a^2 - \frac{5}{2}\right] + \Sah(\psi) F_a(\psi, v) \left[x_a^2 - \frac{3}{2}\right]
 \label{eq:sources}
 \end{equation}
-where $S_{ap}$ and $S_{ah}$ are considered to be unknowns.
-(The $v$ dependence in (\ref{eq:sources}) is chosen so $S_{ap}$ provides a particle source but no heat source, while $S_{ah}$ provides
+where $\Sap$ and $\Sah$ are considered to be unknowns.
+(The factors involving $x_a^2$ in (\ref{eq:sources}) are chosen so $\Sap$ provides a particle source but no heat source, while $\Sah$ provides
 a heat source but no particle source.)
@@
 the sensible requirements that all the flux-surface-averaged density and pressure reside in $F_a$ rather than $f_{a1}$.
-This method of modifying the system of equations has the added benefit of eliminating the null solutions $f_{a1} = F_a$ and $f_{a1} = F_a v^2$ 
-even in the $d\Phi_0/d\psi=0$ limit.
-When $S_a$ is included in the kinetic equation, new terms proportional to $S_{ap}$ and/or $S_{ah}$
+This method of modifying the system of equations has the added benefit 
+(even in the $d\Phi_0/d\psi=0$ limit)
+of eliminating the null solutions $f_{a1} = F_a$ and $f_{a1} = F_a v^2$.
+When $S_a$ is included in the kinetic equation, new terms proportional to $\Sap$ and/or $\Sah$
 now appear in the mass and energy conservation equations such as (\ref{eq:unphysicalMassConservation})-(\ref{eq:unphysicalEnergyConservation2}).
-These conservation equations imply that when $d\Phi_0/d\psi=0$, $S_{ap}$ and $S_{ah}$ must vanish.
+These conservation equations imply that when $d\Phi_0/d\psi=0$, $\Sap$ and $\Sah$ must vanish.
 However, now when $d\Phi_0/d\psi$ is increased from 0 to a small finite number, the sources can turn on
@@
 that the three models give nearly identical results.
-
-The aforemention mass and energy sources and constraints work when either the Fokker-Planck or momentum-conserving model collision operators
-are used, 
-but not when the pure pitch-angle scattering collision operator is used, since in this case the kinetic equation has a larger null space
+As already mentioned, the need for source terms disappears when the radial electric field equals the value required for ambipolarity, so the source terms do not interfere with the aim of eventually calculating the actual plasma transport.
+
+The aforemention mass and energy sources and constraints work when the Fokker-Planck collision operator
+is used, 
+but not when the pure pitch-angle scattering or momentum-conserving model collision operator is used, since in these cases the kinetic equation has a larger null space
 (any function of $v$ is a solution.)
-For this collision operator, we instead allow a general source $S_a(\psi,v)$,
+For these collision operators, we instead allow a general source $S_a(\psi,v)$,
 introducing $N_x$ unknowns on each flux surface, where $N_x$ is the number of grid points in $v$.
@@
 generalized to allow nonaxisymmetry.
+SFINCS is also closely related to the radially global Fokker-Planck code for tokamaks described
+in Ref. \cite{PERFECT}.
 Briefly, the kinetic equation is discretized using finite differences with a 5-point stencil in $\theta$
@@
 so the rate of convergence is not limited by the timescale of physical relaxation.
-A preconditioned iterative Krylov solver is employed.
 The modifications compared to the code of Ref. \cite{speedGrids} are the following.
@@
 (4) The extra constraint equations and sources are implemented as in (19) of Ref. \cite{PERFECT}.
+Specifically, considering first the case of a single species for simplicity, the linear system has the block structure
+\begin{equation}
+\begin{array}{r}
+\mbox{Kinetic equation}\;\;\{ \\
+\left<\int d^3v f_{a1}\right>=0 \;\;\{ \\
+\left<\int d^3v\; f_{a1} v^2\right>=0\;\; \{
+\end{array}
+\left(\begin{array}{ccc}
+M_{11} & M_{12} & M_{13} \\
+M_{21} & 0 & 0 \\
+M_{31} & 0 & 0 \end{array}
+\right)
+\underbrace{
+\left(
+\begin{array}{c}
+f_{a1} \\
+\Sap \\
+\Sah \end{array}
+\right)}_{\mbox{Vector of unknowns}}
+ = \left(
+\begin{array}{c}
+R
+ \\
+0 \\
+0 \end{array}
+\right),
+\label{eq:blocks}
+\end{equation}
+where $R$ is the inhomogeneous term (i.e. the right-hand side) from (\ref{eq:kinetic4}),
+and the $M$ operators are as follows: $M_{11}$ represents the operator on the left-hand side of (\ref{eq:kinetic4}),
+$M_{12}$ and $M_{13}$ represent the $\Sap$ and $\Sah$ terms
+in (\ref{eq:sources}) respectively,
+and $M_{21}$ and $M_{31}$ represent the aforementioned extra constraint equations introduced.
+For the case of multiple particle species, the linear system consists of blocks of the form (\ref{eq:blocks}) for each
+species, with coupling between species only through the collision operators in the $M_{11}$ blocks.
+
+The resulting large sparse linear system is solved using the
+PETSc\cite{petsc-web-page, petsc-user-ref} library.
+A preconditioned iterative Krylov solver is employed,
+either GMRES\cite{GMRES} or BICGStab(l)\cite{BICGstabl}.
+An effective preconditioner is typically obtained by dropping all coupling in the $x_a$ coordinate,
+either for all Legendre modes in $\xi$, or for all but the first one or two Legendre modes.
+The preconditioner is $LU$-factorized directly using the SuperLU-dist\cite{superlu1, superlu2} package.
 
@@
 
-The magnetic geometry is specified in Boozer coordinates $\theta$ and $\zeta$, in which $\vect{B}$
-is represented as
-\begin{eqnarray}
-\vect{B} &=&
-\nabla \psi \times \nabla \theta + \iota \nabla\zeta\times\nabla\psi \\
-&=& \beta(\psi,\theta,\zeta)\nabla\psi + I(\psi)\nabla\theta + G(\psi) \nabla\zeta.
-\end{eqnarray}
-Here, $\iota = 1/q$ is the rotational transform with $q$ the safety factor,
+The magnetic geometry is specified in Boozer coordinates $\theta$ and $\zeta$, in which
+\begin{equation}
+\vect{B} =  \beta(\psi,\theta,\zeta)\nabla\psi + I(\psi)\nabla\theta + G(\psi) \nabla\zeta.
+\end{equation}
+Here, 
 $cI/2$ is the toroidal current inside the flux surface, and $cG/2$ is the poloidal
@@
 where $X$ can be any scalar quantity,
-and the coordinate Jacobian is
+and the inverse coordinate Jacobian is
 $\vect{B}\cdot\nabla\zeta = B^2/(G+\iota I)$. Thus, the magnetic geometry
@@
 
-As a test of the code, we have verified the quasisymmetry isomorphism,
-discussed in Refs. \cite{Pytte, Boozer83, meQS}.
-In this test, the magnetic field is specified as $B(\theta,\zeta) = y(M\theta-N\zeta)$
-for some given periodic function $y$ and integers $M$ and $N$.
-The kinetic equation is solved for several values of $M$ and $N$,
-varying the collision frequency with each run so that
-$\nu_{ii} / \left|M - N/\iota \right|$ remains fixed.
-As $M$ and $N$ are varied, $G$, $I$, $\iota$, $dn/d\psi$, $dT/d\psi$, and other input quantities are also kept fixed.
-Under these circumstances, it is known analytically \cite{Pytte, Boozer83, meQS} that the flow
-and heat flux should obey the transformations
-\begin{equation}
-\left< V_{||} B \right> \frac{M-N/\iota}{NI+MG} = C_V
-\end{equation}
-and
-\begin{equation}
-\left< \int d^3v\, f_{i1} \frac{v^2}{2}\vect{v}_{mi}\cdot\nabla\psi \right> \frac{\left| M-N/\iota\right| }{(NI+MG)^2} = C_q
-\end{equation}
-respectively, where the quantities $C_V$ and $C_q$ should be independent of $M$ and $N$.
-It was verified that the SFINCS code obeyed these isomorphism transformations
-for a range of $y$, collisionality regimes, and collision operators.
-
 \section{Ion transport matrix}
@@
 
-For the case of ions in a pure plasma, it is instructive to examine the transport matrix $L_{jk}$, defined as follows:
+We will present results of the numerical calculations in terms of the transport matrix $L_{jk}$, defined as follows:
 \begin{equation}
@@
 \frac{Ze(G+\iota I)}{ncTG} \left< \int d^3v\, f \frac{mv^2}{2T}\vm\cdot\nabla\psi\right> \\
-\frac{1}{v_i B_0} \left< B V_{||} \right>
+\frac{1}{\vi B_0} \left< B V_{||} \right>
 \end{array}\right)
@@
 \end{equation}
-Here, $B_0$ is the $(0,0)$ Fourier mode amplitude of $B(\theta,\zeta)$, and we have dropped $i$ subscripts.
+Here, $B_0$ is the $(0,0)$ Fourier mode amplitude of $B(\theta,\zeta)$, and we have dropped $i$ subscripts
+where possible to simplify the notation.
 When the DKES trajectories (\ref{eq:DKESTrajectories}) are used, it can be shown that $L_{jk}$ is symmetric for any value of $E_r$.
@@
 Different definitions of the transport matrix have been given elsewhere in the literature \cite{Beidler},
-the definition here has several nice properties. First, the matrix is dimensionless. Second, $L_{jk}$ is symmetric (in the cases described above).
+but the definition here has several nice properties. First, the matrix is dimensionless. Second, $L_{jk}$ is symmetric (in the cases described above).
 Third, $L_{jk}$ depends on the magnetic geometry and physical parameters only through
@@
 \begin{equation}
-\nu' = \frac{(G+\iota I)\nu_{ii}}{v_i B_0},
+\nu' = \frac{(G+\iota I)\nuii}{\vi B_0},
+\label{eq:nuPrime}
 \end{equation}
@@
 \begin{equation}
-E_* = \frac{c G}{\iota v_i B_0}\frac{d\Phi_0}{d\psi},
-\end{equation}
-and not on any other individual parameters such as density, temperature, $I$, etc.
+E_* = \frac{c G}{\iota \vi B_0}\frac{d\Phi_0}{d\psi},
+\label{eq:EStar}
+\end{equation}
+and not on any other individual parameters such as density, temperature, $G$, etc.
+In (\ref{eq:nuPrime}), $\nuii = 4\sqrt{2\pi} n Z^4 e^4 \ln\Lambda/(3 m^{1/2} T^{3/2})$
+is the ion-ion collision frequency.
 Typically, $I \ll G$ and $G \approx B_0 R$ where $R$ is the major radius of the device,
-so $\nu' \approx \nu_{ii} R / v_i$.
-In axisymmetry, $E_*$ corresponds to the poloidal Mach number: $E_* \approx (B / B_{pol}) | \vEo | / v_i$
-where $B_{pol}$ is the poloidal magnetic field.
-Therefore, $E_*$ corresponds to the electric field normalized by the so-called resonant electric field \cite{Beidler}.
-
-Several properties of the matrix $L_{jk}$ are noteworthy. Using the property of the linearized
-ion-ion collision operator $\int d^3v (g/F_i) C_{ii}\{g\} \le 0$
-for any $g$, then $\sgn(L_{11}) = \sgn(L_{22}) = -\sgn(L_{33}) = -\sgn((G+\iota I)/B_0)$.
-\todo{I think this property is true for the FP and PAS operators but not for the model operator.
-Confirm this. Also, for the FP and PAS operators, check this for $E_r \ne 0$ for the 3 trajectory models}.
+so $\nu' \approx \nuii R / \vi$.
+In axisymmetry, $E_*$ corresponds to the poloidal Mach number: $E_* \approx (B / \Bpol) | \vEo | / \vi$
+where $\Bpol$ is the poloidal magnetic field.
+Therefore, $E_*$ corresponds to the electric field normalized by the so-called resonant electric field \cite{Beidler}
+$\Erres = r\iota \vi B/(R c)$, with $r/R$ the inverse aspect ratio.
+
+Several properties of the matrix $L_{jk}$ are noteworthy. Using the property $\int d^3v (g/F_{\mathrm{i}}) C_{\mathrm{ii}}\{g\} \le 0$
+for any $g$, which holds for all three ion-ion collision operators considered here,
+then $\sgn(L_{11}) = \sgn(L_{22}) = -\sgn(L_{33}) = -\sgn((G+\iota I)/B_0)$.
+This property holds when $E_r=0$, and it holds when $E_r \ne 0$ for the DKES trajectories, but it
+may not hold when $E_r \ne 0$ for the partial or full trajectories.
 Second, for all three trajectory models, the elements $L_{jk}$
-are independent of the sign of the elecric field: $L_{jk}(E_*) = L_{jk}(-E_*)$,
+are independent of the sign of the electric field: $L_{jk}(E_*) = L_{jk}(-E_*)$,
 assuming the stellarator symmetry property $B(\theta,\zeta) = B(-\theta,-\zeta)$
@@
 
+When the pure pitch-angle scattering collision operator 
+and DKES trajectories are chosen, the kinetic equation
+solved in SFINCS becomes identical to the one solved in the DKES code \cite{DKES1, DKES2}.
+In this case, it was verified that the two codes agreed for all elements of the transport
+matrix.
+
 \section{Comparison of $E_r$ terms}
@@
 the LHD stellarator \cite{LHD} in its standard configuration.
+(Here the flux function $r$ is defined to be proportional to the square root of the toroidal flux enclosed by the flux surface in question.)
 The calculations in figure (\ref{fig:ErComparison_W7X}) are performed for the $r/a=0.5$ surface of
@@
 begin to separate.
-%Roughly, the poloidal $\vect{E}\times\vect{B}$ will limit the radial step size for diffusion
-%when $E_* > \nu'$, and this prediction is confirmed
-\todo{Can we say anything about why the three trajectory models agree below the resonance? 
-How large is $E_*$ likely to be in W7X and LHD?}
-
-To ensure convergence with respect to the numerical resolution parameters $N_{\theta}$, $N_{\zeta}$, $N_{\xi}$ and $N_x$,
-it was verified that the plotted curves did not change visibly on the scales shown
-when each of these resolution parameters was doubled.
-
-\todo{Add an example (with figure) involving more than 1 species.
-Perhaps compute $j_{bs}$ vs $E_r$ for different $E_r$ models at 1 or 2 values of nu*, perhaps assuming $L_n = L_{Te} = L_{Ti}, E||=0.$}
+
+In all probability, the reason why the three trajectory models agree
+so well with each other below the resonance is that they all capture
+the principal mechanism of transport in the $\sqrt{\nu}$-regime. The
+$\vect{E}\times\vect{B}$ drift convects most locally trapped particles
+poloidally around the torus, thus preventing them from drifting to the
+wall, and the transport is instead dominated by shallowly trapped and
+barely passing particles that are scattered back and forth across the
+trapped-passing boundary on a time scale equal to the poloidal
+convection time \cite{HoKulsrud}. This behavior is not likely to be
+affected by the approximations 
+made in the DKES and partial trajectory models.
+
+In most plasma scenarios, $E_*$ can be estimated by noting that the
+ambipolar electric field arises to bring the ion particle transport
+down to the electron level, and is therefore approximately determined so 
+as to reduce the magnitude of the ``thermodynamic
+force'' appearing as the first component of the vector on the
+right-hand side of (\ref{eq:matrix}). The electric field is thus of
+order $E_r \sim T/(eL_\perp)$, where $L_\perp$ denotes the length
+scale corresponding to the pressure gradient. It is thus expected that $E_*$
+is of order $E_* \sim \rho_\theta / L_\perp$, where $\rho_{\theta} =
+\rho / (\iota \epsilon)$, with $\epsilon$ the inverse aspect ratio. In
+W7-X, $E_r$ is predicted to be particularly large in regions with
+steep density gradients and will in normal plasma scenarios stay below
+a few tens of kV/m, corresponding to $E_*$ up to a few tenths\cite{Turkin}.
+In most of the plasma, $E_*$ is however expected to be smaller than a few percent.
 
@@
 using linearized Fokker-Planck collisions.
+The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
+of the normalized radial electric field (\ref{eq:EStar}).
 \label{fig:ErComparison_LHD}}
@@
 \begin{figure}[h!]
+% Figure generated using m20131205_05_plotSFINCSFortranErComparisonForPaper.m
+% SFINCS run on hopper in 
+% /global/scratch2/sd/landrema/20131204-02-sfincs-fortran-hopper-intel
+%******************************
+% Old version of figure from Dec 2 version:
 % Figure generated using m20131202_01_plotSFINCSErComparisonForPaper.m
@@
 % *********************************
-\includegraphics{m20131202_01_plotSFINCSErComparisonForPaper_W7X.eps}
-%\includegraphics{m20131202_01_plotSFINCSErComparisonForPaper_W7X.pdf}
-\caption{(Color online) Comparison of trajectory models for W7X standard geometry at $\nu' = 0.01$,
+\includegraphics{m20131205_05_plotSFINCSFortranErComparisonForPaper_W7X.eps}
+\caption{(Color online) Comparison of trajectory models for W7-X standard geometry at $\nu' = 0.01$,
 using linearized Fokker-Planck collisions.
-\todo{Fill in points for $E_*$ in the range 0.1 to 1.}
+The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
+of the normalized radial electric field (\ref{eq:EStar}).
 \label{fig:ErComparison_W7X}}
@@
 
+Further analysis of whether the choice of trajectory model is significant in W7-X
+is shown in figure  \ref{fig:2species}.
+This calculation is based on the scenario considered in figure 5 of Ref. \cite{Turkin}. 
+We focus on the radial location $r=0.45$ m ($r/a=0.88$)
+in which the pressure gradient is strong.  This gradient should result in a large $E_r$, as predicted 
+both by the argument in the preceding paragraph, and by
+the modeling in Ref. \cite{Turkin} based on incompressible-$\vect{E}\times\vect{B}$ computations.
+(Here, the flux label $r$ is defined by $\pi r^2 B_0 = 2 \pi \psi$.)
+Matching the parameters in that work, we consider a pure hydrogen plasma with $n=6.6\times 10^{19}\; \mathrm{m}^{-3}$,
+$\Te=\Ti=1$ keV, $dn/dr = -1.2\times 10^{21} \;\mathrm{m}^{-4}$, and $d\Te/dr=d\Ti/dr = -16$ keV/m.
+These values correspond to $\nu'=0.03$ and $\Erres = 100 $ kV/m.
+For this scenario, kinetic electrons are included in SFINCS along with the ions.
+Inter-species linearized Fokker-Planck collisions are included with no expansion in mass ratio.
+
+The radial fluxes of ions and electrons as functions of $E_r$ are shown in figure \ref{fig:2species}.A.
+The electron fluxes (dashed curves) are very small $(\sim \sqrt{\me/\mi})$ compared to the ion fluxes
+and are identical between the three trajectory models. No difference between the models is expected for
+the electrons, since $E_*$ defined using the electron rather than ion thermal speed is always $\ll 1$.
+The vertical magenta dotted line indicates the ambipolar value of $E_r \approx $ 33 kV/m, which is effectively
+identical for the three trajectory models, and comparable to the value predicted in \cite{Turkin}.
+This electric field is roughly one third of the resonant value, and therefore the ion transport coefficients are
+just beginning to separate for the three models. Heat fluxes are shown in figure \ref{fig:2species}.B, showing similar
+behavior to the particle fluxes.
+
+Figure \ref{fig:2species}.C shows the surface-averaged ion parallel flow. At the ambipolar value of $E_r$,
+the three trajectory models yield similar values for the predicted flow.
+At lower magnitudes of $E_r$, the flows predicted by the three models are nearly indistinguishable.
+However, at larger electric fields, the three models begin to give quite different predictions.
+This change in behavior around $E_* \sim 0.3$ is consistent with the patterns in figures \ref{fig:ErComparison_LHD}- \ref{fig:ErComparison_W7X}.
+A similar pattern is visible in the bootstrap current density, shown in figure \ref{fig:2species}.D.
+At the ambipolar value of $E_r$, the partial trajectory model predicts
+$27\%$ more bootstrap current than the full trajectory model,
+and the DKES model predicts $8\%$ more bootstrap current than the full trajectory model.
+Interestingly, if the electric field exceeds 60 kv/m in the inward (ion root) direction, the bootstrap current
+in the full trajectory model changes sign, whereas there is no sign change in the DKES model.
+
+\begin{figure}[h!]
+% Figure generated using m20131212_02_PlotSFINCS2SpeciesEScanForPaper.m
+% SFINCS run on hopper and edison in 
+% /global/scratch2/sd/landrema/20131204-02-sfincs-fortran-hopper-intel
+\includegraphics{m20131212_02_PlotSFINCS2SpeciesEScanForPaper.eps}
+\caption{(Color online)
+Fluxes and flows computed for a scenario of steep pressure gradient near the edge of W7-X.
+Magenta dotted line is the ambipolar $E_r$, effectively identical for the three
+trajectory models.
+\label{fig:2species}}
+\end{figure}
+
 \section{Comparison of collision operators}
@@
 Figures (\ref{fig:collisionComparison_LHD})-(\ref{fig:collisionComparison_W7X}) show the transport
-matrix elements for the LHD and W7X geometries described earlier, this time comparing
+matrix elements for the LHD and W7-X geometries described earlier, this time comparing
 the different collision operators as a function of collisionality.
@@
 is unimportant for $L_{11}$, $L_{12}$, $L_{21}$, and $L_{22}$.
-These matrix elements are associated with $1/\nu$-regime radial transport
+These matrix elements represent $1/\nu$-regime radial transport
 (when $\nu' \ll 1$), which is associated with pitch-angle scattering of helically trapped particles.
@@
 captures the dominant physics in these cases. 
-The other matrix elements are more sensitive to momentum conservation at low collisionality.
+When $d\Phi_0/d\psi \ne 0$, the same is true for the
+$\sqrt{\nu}$-regime, where the main effect of the collisions is to
+scatter particles across a thin collisional boundary layer in velocity
+space around the trapped-passing boundary.
+
+The other matrix elements ($L_{13}$, $L_{23}$, $L_{31}$, $L_{32}$, and $L_{33}$) 
+are more sensitive to momentum conservation at low collisionality.
 For all the matrix elements at all collisionalities, the momentum-conserving model operator
@@
 
-Again, to ensure convergence with respect to the numerical resolution parameters $N_{\theta}$, $N_{\zeta}$, $N_{\xi}$ and $N_x$,
-it was verified that the plotted curves did not change visibly on the scales shown
-when each of these resolution parameters was doubled.
-The required resolution in the $\zeta$ and $\xi$ coordinates increases substantially
-as $\nu'$ decreases, due to the boundary layers that develop in the distribution function.
+Note in Figs.~(\ref{fig:collisionComparison_LHD})-(\ref{fig:collisionComparison_W7X}) that the scaling of the $L_{11}$ and $L_{12}$ coefficients at high collisionality depends crucially on whether momentum is conserved in the collision operator.
+In the momentum-conserving calculations, these transport coefficients are inversely proportional to $\nu$, whereas they are proportional to $\nu$ if the collisions are approximated by pure pitch-angle scattering.
+To understand why, it is useful to recall that the Pfirsch-Schl{\"u}ter
+particle flux consists of two terms: one related to the parallel
+friction force and one related to parallel viscosity \cite{AndreiPer2009, Braun}.
+This is most easily seen by taking the scalar product of the lowest-order plasma current, which satisfies $\vect{J} \times \vect{B} = c p'(\psi) \nabla \psi$, with the momentum equation,
+\begin{equation}
+m_a n_a \vect{V}_a \cdot \nabla \vect{V}_a = n_a e_a \left( - \nabla \phi + c^{-1}\vect{V}_a \times \vect{B} \right)
+- \nabla p_a - \nabla \cdot \vect{\pi}_a + \vect{F}_a, 
+\end{equation}
+neglecting the left-hand side. Since $\nabla \cdot \vect{J} = 0$ and $n_a$ is a flux function in lowest order, we obtain
+	\begin{equation}
+	 \left< n_a \vect{V}_a \cdot \nabla \psi \right> 
+	= \frac{1}{e_a p'(\psi)} \left< \vect{J} \cdot \left( \vect{F}_a - \nabla \cdot \vect{\pi}_a \right) \right>,
+	\end{equation}
+where the term corresponding to the perpendicular component of the friction force $\vect{F}_a$ represents the classical particle flux and the other terms the neoclassical flux,
+		\begin{equation}
+	 \left< n_a \vect{V}_a \cdot \nabla \psi \right>_{nc} = \frac{1}{e_ap'} \left< J_\| F_{a\|} + \vect{\pi}:\nabla \vect{J} \right>
+	 \label{PS flux}
+	 \end{equation}
+where the viscosity tensor is $\vect{\pi}_a =(p_{\|} - p_{\perp})(\vect{b b} - {\bf I}/3)$.
+The first term in (\ref{PS flux}) is proportional to $\nu$ and
+therefore dominates at high collisionality, but vanishes when there is only a 
+single ion species because of momentum conservation in like-particle
+collisions. All that remains is therefore the particle flux caused
+by parallel viscosity, which is inversely proportional to $\nu$ at high collisionality \cite{PerBook}. In
+the pure pitch-angle-scattering model however, parallel momentum
+conservation is violated, leading to spurious friction-driven
+transport proportional to $\nu$. This is why the green curves have a
+slope of +1 for large $\nu$ in the logarithmic plots of $L_{11}$ and
+$L_{12}$ in Figures
+(\ref{fig:collisionComparison_LHD})-(\ref{fig:collisionComparison_W7X}),
+while the blue and red curves have the slope -1.
+
+A similar difference between the momentum-conserving and
+pitch-angle-scattering operators is evident in the parallel
+conductivity coefficient $L_{33}$. The flow that arises in response to
+a parallel electric field is determined by the parallel momentum equation
+\begin{equation} 
+n_a Z_a e E_{||} = \vect{b}\cdot \nabla p_a + \vect{b}\cdot(\nabla \cdot \vect{\pi}_a )+ F_{a||}, 
+\end{equation} 
+where the parallel friction force $F_{\|a}$ again vanishes when only a
+single ion species is considered. Hence 
+\begin{equation} 
+n_a Z_a e \left< B  E_{||} \right> = \left< \vect{B}\cdot(\nabla \cdot \vect{\pi}_a) \right> 
+= \left< (p_\perp - p_\| ) \nabla_\| B \right>.
+\end{equation}
+In
+the absence of radial gradients, the pressure anisotropy in the \PS
+regime is proportional to the parallel flow velocity and inversely
+proportional to the collision frequency \cite{AndreiPer2009}. The
+flow $\left< V_\| B \right>$ is therefore proportional to $\nu$ in
+the \PS regime unless momentum conservation is violated. In the
+latter case, the spurious friction force causes $\left< V_\| B
+\right>$ to be inversely proportional to $\nu$, as can be seen in
+Figs.~(\ref{fig:collisionComparison_LHD})-(\ref{fig:collisionComparison_W7X}).
+
+When $\nu' < 1$, the resolution required in the $\theta$, $\zeta$, and $\xi$ coordinates
+increases as $\nu'$ decreases,
+due to the boundary layers that develop in phase space.  The highest resolution used
+for results presented here, corresponding to the W7-X calculations at  $\nu'=10^{-3}$,
+was $N_\theta=29$, $N_\zeta = 83$, $N_\xi=180$, and $N_x=5$,
+giving a $2,166,302 \times 2,166,302$ linear system.
+Here, $N_j$ is the number of grid points or modes in coordinate $j$.
+Each calculation at this resolution with the Fokker-Planck collision operator
+required $\sim 30-50$ minutes to run on 4 nodes of the Edison computer at NERSC.
+At higher collisionality, lower resolution is sufficient, so memory and time requirements are reduced;
+computations with $\nu' > 10^{-2}$ can typically be run on a laptop.
 
 \begin{figure}[h!]
+% Figure generated using m20131208_01_plotSFINCSFortranAndMatlabNuScansForPaper.m
+% High collisionality runs done in matlab, with data in
+% m20130226_04_SFINCS_2013-02-26_21-21_good
+% Low collisionality runs done in fortran on edison.
+% *********************************
+% Old version of figure:
 % Figure generated using m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper.m
@@
 % *********************************
-\includegraphics{m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper_LHD.eps}
-%\includegraphics{m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper_LHD.pdf}
+\includegraphics{m20131208_01_plotSFINCSFortranAndMatlabNuScansForPaper_LHD.eps}
+%\includegraphics{m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper_LHD.eps}
 \caption{(Color online) Comparison of collision operators for LHD standard geometry at $E_r=0$.
-\todo{Extend plots to $\nu' = 10^{-3}$.}
+The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
+of the collisionality (\ref{eq:nuPrime}).
 \label{fig:collisionComparison_LHD}}
@@
 \begin{figure}[h!]
+% Figure generated using m20131208_01_plotSFINCSFortranAndMatlabNuScansForPaper.m
+% High collisionality runs done in matlab, with data in
+% m20131117_04_sfincs_w7x_2013-11-26_02-27_nuPrimeScan_nuPrime0_01to100
+% Low collisionality runs done in fortran on edison.
+% *********************************
+% Old version of figure:
 % Figure generated using m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper.m
@@
 % *********************************
-\includegraphics{m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper_W7X.eps}
-%\includegraphics{m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper_W7X.pdf}
-\caption{(Color online) Comparison of collision operators for W7X standard geometry at $E_r=0$.
-\todo{Extend plots to $\nu' = 10^{-3}$.}
+\includegraphics{m20131208_01_plotSFINCSFortranAndMatlabNuScansForPaper_W7X.eps}
+%\includegraphics{m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper_W7X.eps}
+\caption{(Color online) Comparison of collision operators for W7-X standard geometry at $E_r=0$.
+The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
+of the collisionality (\ref{eq:nuPrime}).
 \label{fig:collisionComparison_W7X}}
@@
 
-%\todo{Add an example (with figure) involving more than 1 species. Perhaps compute $j_{bs}$ vs nu* for different collision operators at $E_r=0$, perhaps assuming $L_n = L_{Te} = L_{Ti}, E||=0.$}
-
-\begin{comment}
-\begin{table}
-\centering
-\begin{tabular}{c c c c c}
-$\nu_n$ & $N_\theta$ & $N_\zeta$ & $N_\xi$ & $N_x$ \\
-\hline
-100 & 11 & 37 & 13 & 8 \\
-10 & 13 & 35 & 12 & 7 \\
-1 & 13 & 31 & 24 & 6 \\
-0.1 & 11 & 37 & 37 & 5 \\
-0.01 & 11 & 64 & 68 & 5 \\
-0.001 & 11 & 83 & 122 & 5 \\
-%\begin{tabular}{c c c c c c c}
-%$\nu_n$ & 0.001 & 0.01 & 0.1 & 1 & 10 & 100 \\
-\end{tabular}
-\caption{Resolution parameters required for convergence in the W7-X examples.
-\label{tab:resolution}}
-\end{table}
-\end{comment}
-
 \section{Discussion and conclusions}
@@
 
-\todo{Write this section}
+Different stellarator transport codes make different assumptions about the
+details of the drift-kinetic equation, and our goal in this work has been
+to examine the consequences of these details.
+In this work, we have compared three versions of the drift-kinetic equation
+for a stellarator, consisting of (\ref{eq:kinetic4}) with
+the coefficients (\ref{eq:fullTrajectories}), (\ref{eq:DKESTrajectories}),
+or (\ref{eq:partialTrajectories}).
+These three sets of expressions for $\vect{\dot{r}}_a$,
+$\dot{x}_a$, and $\dot{\xi}_a$ may be interpreted as effective
+particle trajectories (although we solve each form of the kinetic equation
+using continuum numerical methods).
+At first glance,  (\ref{eq:fullTrajectories}) and  (\ref{eq:partialTrajectories})
+appear more accurate than  (\ref{eq:DKESTrajectories}).
+However, as we have shown in section \ref{sec:sources},
+the kinetic equation (\ref{eq:kinetic4})
+with  (\ref{eq:fullTrajectories}) or  (\ref{eq:partialTrajectories})
+is not well behaved when $E_r \ne 0$,
+with an unphysical constraint forced upon the distribution function.
+This analytic property of the kinetic equation must be dealt with
+before attempting to solve the equation numerically.
+
+To eliminate this problem of unphysical constraints,
+we propose formulating the kinetic problem as in  (\ref{eq:blocks}) with (\ref{eq:sources}).
+A particle and heat source are introduced,
+along with the additional constraints that all the flux-surface-averaged
+density and pressure reside in the leading-order Maxwellian.
+For the model (\ref{eq:DKESTrajectories}), the sources always vanish.
+For the model  (\ref{eq:fullTrajectories}), the particle source
+vanishes for any $E_r$, and the energy source vanishes when $E_r$
+takes on its ambipolar value.
+The equations (\ref{eq:blocks}) have been implemented in a new time-independent continuum code SFINCS,
+and the resulting ion transport matrices have been compared for the geometries of the LHD and W7-X stellarators.
+When $E_r$ is below roughly one-third of the resonant value,
+the three models give nearly indistinguishable results. 
+This finding confirms that the incompressible-$\vect{E}\times\vect{B}$ trajectory
+model used in some codes \cite{DKES1,DKES2} is quite accurate in this
+small-$E_r$ regime, which is typically satisfied in experiments.
+Physically, the effect of $E_r$ in this regime is to generate a $\sqrt{\nu}$ regime
+of transport due to poloidal precession of helically trapped particles,
+and this process is retained (at least approximately) in all three trajectory models.
+Once $E_r$ approaches the resonance, however, the three trajectory models
+yield substantially different results.
+This $E_r \sim \Erres$ regime can be relevant to experiments with high ratios $\Te / \Ti$
+\cite{Lore, Briesemeister} and strong gradients \cite{Baldzuhn}.
+In figure \ref{fig:2species}, we find that in the large-$E_r$ region anticipated for the edge of W7-X,
+the bootstrap current density in the full trajectory model is modestly reduced (by $8\%$)
+compared to an incompressible-$\vect{E}\times\vect{B}$ calculation,
+but should larger values of $E_*$ arise, we expect the deviation could grow more significant.
+
+Since full coupling in the speed coordinate $x_{a}$ is retained in our numerical
+implementation, it is possible to directly compare results from the full linearized
+Fokker-Planck collision operator to results from simpler collision models.
+At low collisionality, the ion transport matrix elements $L_{11}$,
+$L_{12}$, $L_{21}$, and $L_{22}$ are nearly identical for the three collision models
+considered. This result makes sense physically since these matrix elements at low collisionality
+are associated with a piece of the distribution function that is localized to a narrow range of pitch angles,
+so pitch angle diffusion is the dominant collisional process.
+However, these same matrix elements at higher collisionality, or the other matrix elements
+at any collisionality, are sensitive to momentum conservation.
+The momentum-conserving model operator results in the correct scaling with collisionality
+when compared to the full linearized Fokker-Planck operator.
+However, there are still $O(1)$ differences in the transport coefficients
+computed with these two collision operators.
 
 \begin{acknowledgments}
+This work was supported by the US Department of Energy through grants DE-FG02-91ER-54109 and DE-FG02-93ER-54197.
+M. L. is grateful to the Plasma Science and Fusion Center at the Massachusetts Institute of Technology,
+where he carried out much of the code development, and for travel support from the
+Max-Planck-Institut f\"{u}r Plasmaphysik in Greifswald, Germany.
 Some of the computer simulations presented here
@@
 administered by the Oak Ridge Institute for Science and Education.
-We are grateful to J. Geiger for providing the W7X equilibrium data.
+We are thankful to J. Geiger for providing the W7-X equilibrium data,
+and to Craig Beidler, Albert Moll\'{e}n, and Peter Catto for other input on this work.
 \end{acknowledgments}
 
-%\appendix
-%\section{Any appendix would go here}
+\appendix
+\section{Quasisymmetry Isomorphism}
+
+A useful test of a stellarator neoclassical code such as the one
+described here is the quasisymmetry isomorphism,
+discussed analytically in Refs. \cite{Pytte, Boozer83, meQS}.
+In this test, the magnetic field is specified as $B(\theta,\zeta) = y(M\theta-N\zeta)$
+for some given periodic function $y$ and integers $M$ and $N$.
+The kinetic equation is solved for several values of $M$ and $N$,
+varying the collision frequency with each run so that
+$\nuii / \left|M - N/\iota \right|$ remains fixed.
+As $M$ and $N$ are varied, $G$, $I$, $\iota$, $dn/d\psi$, $dT/d\psi$, and other input quantities are also kept fixed.
+Under these circumstances, it is known analytically \cite{Pytte, Boozer83, meQS} that the flow
+and heat flux should obey the transformations
+\begin{equation}
+\left< V_{||} B \right> \frac{M-N/\iota}{NI+MG} = C_V
+\end{equation}
+and
+\begin{equation}
+\left< \int d^3v\, f_{\mathrm{i}1} \frac{v^2}{2}\vect{v}_{\mathrm{m}i}\cdot\nabla\psi \right> \frac{\left| M-N/\iota\right| }{(NI+MG)^2} = C_q
+\end{equation}
+respectively, where the quantities $C_V$ and $C_q$ should be independent of $M$ and $N$.
+It was verified that the SFINCS code obeyed these isomorphism transformations
+for a range of $y$, collisionality regimes, and collision operators.
 
@@
 \end{document}
-
-% To do:
-% * Add references to other stellarator codes in intro.
-% * Write conclusions
-% * Finish E* scan for W7X
-% * Figure out multi-species model operator
-% * Write section on collision operators
-% * Should I include any results showing ripple?
-% * Think of a good example involving >1 species.
-% * Paragraph that L_jk is invariant under EStar -> - EStar, L_11 and L_22 are negative, and L12=L21 for full trajectories even when E \ne 0.

### 2014-01-06 Albert/Matlab/tags/m20130318/m20121127_02_makeHighOrderInterpolationMatrix.m
Initial commit

@@ -1,0 +1,88 @@
+function m=m20121127_02_makeHighOrderInterpolationMatrix(x,y,L,GOrH)
+% This function produces a matrix for interpolating a function from one
+% grid (x) to another (y).  The interpolation is performed using cubic 
+% functions.  This routine could be applied in many ways, but the 
+% particular application it was designed for is interpolating Rosenbluth 
+% potentials.  A variety of options are therefore available for 
+% extrapolating the potentials off the right end of the x grid, depending 
+% on which of the two potentials we are considering.  This function is not
+% designed forextrapolating off the left end of the x grid, so use with 
+% caution.
+%
+% Inputs:
+% x = grid on which we know a function.
+% y = new grid onto which we want to interpolate the function.
+% L = Legendre mode number of the potential we are considering.  This
+%     parameter is ignored when 'GOrH' = 'f'.
+% GOrH = method for extrapolating to the right of the x grid.
+%   Possible options for GOrH:
+%   'f' = return 0.
+%   'H' = extrapolate as 1/(y ^ (L+1))
+%         (suitable for the 1st Rosenbluth potential.)
+%   'G' = extrapolate using equation (27) of 
+%         http://arxiv.org/pdf/1210.5289v1.pdf
+%         (suitable for d^2/dv^2 of the 2nd Rosenbluth potential.)
+%   Note: for the pure plasma code, it is not necessary to extrapolate the
+%   potentials off the grid, so the 'L' and 'GOrH' parameters do not
+%   matter.  They do matter in the impure-plasma code however.
+%
+% Outputs:
+% m = interpolation matrix
+
+m = zeros(numel(y), numel(x));
+for i=1:numel(y)
+    index = find(abs(x-y(i))<1e-12,1);
+    if ~isempty(index)
+        m(i,index)=1;
+        
+    else
+        
+        index = find(x>=y(i),1);
+        if numel(index)>1
+            error('numel(index)>1')
+        end
+        if numel(index)>0
+            if (index>(numel(x)-2))
+                indicesToUse = ((-3):0) + numel(x);
+            elseif index<3
+                indicesToUse = 1:4;
+            else
+                indicesToUse = ((-2):1) + index;
[…]

### 2014-01-06 Albert/Matlab/tags/m20130318/m20130318_02_SFINCS.m

@@ -1,0 +1,2075 @@
+function m20130318_02_SFINCS()
+
+% SFINCS:
+% The Stellarator Fokker-Planck Iterative Neoclassical Conservative Solver.
+% Original version written in 2013 by Matt Landreman
+% Massachusetts Institute of Technology
+% Plasma Science & Fusion Center
+
+% --------------------------------------------------
+% Program flow control parameters:
+% --------------------------------------------------
+
+programMode = 1;
+% 1 = single run.
+% 2 = Do a convergence scan and save the results.
+% 3 = Load a previous convergence scan and plot the results. (Don't do any new solves.)
+% 4 = Do a nuPrime scan and save the results.
+% 5 = Load a previous nuPrime scan and plot the results. (Don't do any new solves.)
+
+% The setting below matters for programMode=3 or programMode=5 only:
+dataFileToPlot = 'm20130318_02_SFINCS_2013-03-18_14-47_convergenceScan_convergenceScan.mat';
+
+RHSMode = 2;
+% 1 = Use a single right-hand side.
+% 2 = Use multiple right-hand sides to compute the transport matrix.
+
+%testQuasisymmetryIsomorphism = true;
+testQuasisymmetryIsomorphism = false;
+
+saveStuff = true;
+%saveStuff = false;
+
+filenameNote = 'myFirstScan';
+
+% --------------------------------------------------
+% Geometry parameters:
+% --------------------------------------------------
+
+geometryScheme = 2;
+% 1 = Two-helicity model
+% 2 = Three-helicity approximation of the LHD standard configuration
+% 3 = Four-helicity approximation of the LHD inward-shifted configuration
+
+B0OverBBar = 0.7;
+GHat = 1.0;
+IHat = 0.8;
+iota = 1.31;
+epsilon_t = 0.13;
+epsilon_h = 0.1;
+helicity_l = 2;
+helicity_n = 5;
+
+% --------------------------------------------------
+% Physics parameters:
+% --------------------------------------------------
+
+Delta = 0.0011;
+omega = 0.0014;
+psiAHat = 0.03;
+THat = 1.0;
+nHat = 1.0;
+dTHatdpsi = -0.7;
+dnHatdpsi = -0.5;
+EHat = 0;
+dPhiHatdpsi = 0;
+
+
+% If RHSMode == 1, nuN is used and nuPrime is ignored.
+% If RHSMode == 2, nuPrime is used and nuN is ignored.
+
+nuN = 1.0;
[…]

### 2014-01-06 Albert/Matlab/tags/m20130318/poldif.m

@@ -1,0 +1,73 @@
+function DM = poldif(x, malpha, B)
+
+%  The function DM =  poldif(x, maplha, B) computes the
+%  differentiation matrices D1, D2, ..., DM on arbitrary nodes.
+%
+%  The function is called with either two or three input arguments.
+%  If two input arguments are supplied, the weight function is assumed 
+%  to be constant.   If three arguments are supplied, the weights should 
+%  be defined as the second and third arguments.
+%
+%  Input (constant weight):
+%
+%  x:        Vector of N distinct nodes.
+%  malpha:   M, the number of derivatives required (integer).
+%  B:        Omitted.
+%
+%  Note:     0 < M < N-1.
+%
+%  Input (non-constant weight):
+%
+%  x:        Vector of N distinct nodes.
+%  malpha:   Vector of weight values alpha(x), evaluated at x = x(k).
+%  B:        Matrix of size M x N,  where M is the highest 
+%            derivative required.  It should contain the quantities 
+%            B(ell,j) = beta(ell,j) = (ell-th derivative
+%            of alpha(x))/alpha(x),   evaluated at x = x(j).
+%
+%  Output:
+%  DM:       DM(1:N,1:N,ell) contains ell-th derivative matrix, ell=1..M.
+
+%  J.A.C. Weideman, S.C. Reddy 1998
+
+       N = length(x);                      
+       x = x(:);                     % Make sure x is a column vector 
+
+if nargin == 2                       % Check if constant weight function
+       M = malpha;                   % is to be assumed.
+   alpha = ones(N,1);              
+       B = zeros(M,N);
+elseif nargin == 3
+   alpha = malpha(:);                % Make sure alpha is a column vector
+       M = length(B(:,1));           % First dimension of B is the number 
+end                                  % of derivative matrices to be computed
+ 
+        I = eye(N);                  % Identity matrix.
+        L = logical(I);              % Logical identity matrix.
+
+       XX = x(:,ones(1,N));
+       DX = XX-XX';                  % DX contains entries x(k)-x(j).
+
+    DX(L) = ones(N,1);               % Put 1's one the main diagonal.
+
+        c = alpha.*prod(DX,2);       % Quantities c(j).
+
+        C = c(:,ones(1,N)); 
+        C = C./C';                   % Matrix with entries c(k)/c(j).
+   
+        Z = 1./DX;                   % Z contains entries 1/(x(k)-x(j))
[…]

### 2014-01-06 Albert/Matlab/trunk/m20120703_03_polynomialInterpolationMatrix.m

@@ -1,0 +1,65 @@
+function interpMatrix = m20120703_03_polynomialInterpolationMatrix(xk, x, alpxk, alpx)
+% Returns the matrix for polynomial interpolation from the xk grid to the x
+% grid.
+%
+% This script is a modification of the polint.m routine from DMSuite,
+% by J.A.C. Weideman and S.C. Reddy, available here:
+% http://www.mathworks.com/matlabcentral/fileexchange/29
+% or here:
+% http://dip.sun.ac.za/~weideman/research/differ.html
+%
+% The original polint.m routine performed the interpolation explicitly.
+% This modified function returns the interpolation matrix, for use in
+% implicit codes.
+% 
+% Modification by Matt Landreman, MIT Plasma Science and Fusion Center,
+% July 3, 2012.
+%
+%  Inputs
+%  xk:    Vector of x-coordinates of data (assumed distinct).
+%  x:     Vector of x-values where polynomial interpolant is to be evaluated.
+%  alpxk: Vector of weight values sampled at the points xk.
+%  alpx:  Vector of weight values sampled at the points x.
+%
+%  Output:
+%  interpolation matrix
+%
+%  The code implements the barycentric formula; see page 252 in
+%  P. Henrici, Essentials of Numerical Analysis, Wiley, 1982.
+%  (Note that if some fk > 1/eps, with eps the machine epsilon,
+%  the value of eps in the code may have to be reduced.)
+%  Note added May 2003:  Except for certain nice node distributions
+%  polynomial interpolation of high-degree is an ill-conditioned
+%  problem.  This code does not test for conditioning so use with
+%  care. 
+
+%  J.A.C. Weideman, S.C. Reddy 1998.
+
+N = length(xk);
+     M = length(x);
+
+if nargin == 2
+    alpx = 1; alpxk = 1;
+    deWeightify = eye(N);
+elseif nargin > 2
+    deWeightify = diag(1./alpxk);
+else
+    error('Unexpected number of inputs.')
+end
+
+     x = x(:);                          % Make sure the data are column vectors
+    xk = xk(:);  
+ alpxk = alpxk(:); alpx = alpx(:);
+
+     L = logical(eye(N));
+
+     D = xk(:,ones(1,N))-xk(:,ones(1,N))';  % Compute the weights w(k)
+  D(L) = ones(N,1);
+     w = 1./prod(D)';                       
+ 
+     
[…]

### 2014-01-06 Albert/Matlab/trunk/m20130312_02_SpectralNodesWeightsAndDifferentiationMatricesForV.m

@@ -1,0 +1,45 @@
+function [x, ddx, d2dx2, weights] = m20130312_02_SpectralNodesWeightsAndDifferentiationMatricesForV(N, k, scale, pointAtZero)
+% Computes the abscissae (i.e. grid points), Gaussian integration weights,
+% and spectral colocation differentiation matrices for functions on the
+% domain [0, Inf] which behave roughly as exp(-x^2) for large x.
+%
+% Written by Matt Landreman - July 6, 2012
+%
+% Inputs:
+% N = desired number of abscissae
+% k = extra power of x in the orthogonality relation and weight. 
+%        (Usually, k=0 works well.)
+% scale = factor by which to scale all the abscissae, (typically in the range 0.7 to 1)
+%               so as to not waste points at very large x.
+% pointAtZero = true or false, depending on whether or not you require a
+%               grid point at x=0.
+%
+% Outputs:
+% x = column vector of abscissae
+% ddx = first derivative matrix
+% d2dx2 = second derivative matrix
+% weights = Gaussian integration weights
+%
+% This function depends on
+% m20121125_05_GaussWeightsAndAbscissae.m
+% as well as
+% poldif.m
+% The latter is part of the DMSuite package by S.C. Reddy and J.A.C. Weideman, available at
+% http://www.mathworks.com/matlabcentral/fileexchange/29
+% or here:
+% http://dip.sun.ac.za/~weideman/research/differ.html
+
+if pointAtZero
+    [x, weights]=m20121007_01_generateGaussRadauPointsAndWeights(N, @(x) (x.^k) .* exp(-x.*x), 0, Inf);
+else
+    [x, weights]=m20121125_05_GaussWeightsAndAbscissae(N, @(x) (x.^k) .* exp(-x.*x), 0, Inf);
+end
+weights=weights./((x.^k) .* exp(-x.*x));
+x = x*scale;
+weights = weights*scale;
+weightFactors=[k./x-2*x, k*(k-1)./(x.*x)-2*(2*k+1)+4*x.*x]';
+differentiationMatrices=poldif(x,(x.^k) .* exp(-x.*x),weightFactors);
+ddx=differentiationMatrices(:,:,1);
+d2dx2=differentiationMatrices(:,:,2);
+
+end

### 2014-01-06 Albert/Matlab/trunk/sfincsAlbert.m

@@ -1,0 +1,2534 @@
+function sfincsAlbert()
+
+% SFINCS:
+% The Stellarator Fokker-Planck Iterative Neoclassical Conservative Solver.
+% Single species version.
+% Original version written in 2013 by Matt Landreman
+% Massachusetts Institute of Technology
+% Plasma Science & Fusion Center
+    
+% Dimensional quantities in this program are normalized to "reference" values:
+% \bar{B} = reference magnetic field, typically 1 Tesla.
+% \bar{R} = reference length, typically 1 meter.
+% \bar{n} = reference density, typically 10^19 m^{-3}, 10^20 m^{-3}, or something similar.
+% \bar{m} = reference mass, typically either the mass of hydrogen or deuterium.
+% \bar{T} = reference temperature in energy units, typically 1 eV or 1 keV.
+% \bar{v} = \sqrt{2 * \bar{T} / \bar{m}} = reference speed
+% \bar{Phi} = reference electrostatic potential, typically 1 V or 1 kV.
+
+% You can choose any reference parameters you like, not just the values
+% suggested here. The code "knows" about the reference values only through
+% the 3 combinations Delta, omega, and nuN or nuPrime, input below.
+
+% Radial gradients of density, temperature, and electrostatic potential are
+% specified as derivatives with respect to psi_N, where psi_N is the 
+% toroidal flux normalized to the value at the last closed flux surface. 
+% (psi_N=0 is the magnetic axis, and psi_N=1 is the last closed flux 
+% surface.)
+    
+% --------------------------------------------------
+% Program flow control parameters:
+% --------------------------------------------------
+
+programMode = 1;
+% 1 = single run.
+% 2 = Do a convergence scan and save the results.
+% 3 = Load a previous convergence scan and plot the results. (Doesn't do any new solves.)
+% 4 = Do a nuPrime scan and save the results.
+% 5 = Load a previous nuPrime scan and plot the results. (Doesn't do any new solves.)
+
+% The setting below matters for programMode=3 or programMode=5 only:
+dataFileToPlot = 'm20130318_02_SFINCS_2013-03-18_14-47_convergenceScan_convergenceScan.mat';
+
+RHSMode = 1;
+% 1 = Use a single right-hand side.
[…]

### 2014-01-06 fortran/multiSpecies/globalVariables.F90

@@ -1,0 +1,293 @@
+module globalVariables
+
+  implicit none
+
+#include <finclude/petscsysdef.h>
+
+  integer, parameter :: integerToRepresentTrue  =  1
+  integer, parameter :: integerToRepresentFalse = -1
+
+  PetscScalar :: one = 1., oneHalf = 0.5d+0
+  PetscScalar :: zero = 0., two = 2., three = 3., four = 4., five = 5.
+
+  PetscScalar, parameter :: pi = 3.14159265358979d+0
+  PetscScalar, parameter :: sqrtpi = 1.77245385090552d+0
+
+  ! ********************************************************
+  ! ********************************************************
+  !
+  ! Options for program flow control:
+  !
+  ! ********************************************************
+  ! ********************************************************
+
+  integer :: programMode, RHSMode, numRHSs
+  logical :: saveMatlabOutput
+  integer :: outputScheme
+  character(len=200) :: MatlabOutputFilename
+  character(len=200) :: outputFilename
+  logical :: solveSystem
+
+  ! ********************************************************
+  ! ********************************************************
+  !
+  ! Geometry input parameters:
+  !
+  ! ********************************************************
+  ! ********************************************************
+
+  integer :: geometryScheme
+  PetscScalar ::  GHat, IHat, iota, B0OverBBar
+  PetscScalar :: epsilon_t, epsilon_h, epsilon_antisymm
+  integer :: NPeriods, helicity_l, helicity_n, helicity_antisymm_l, helicity_antisymm_n
+  character(len=200) :: fort996boozer_file, JGboozer_file
+  PetscScalar :: normradius_wish, min_Bmn_to_load
+
+  ! ********************************************************
+  ! ********************************************************
+  !
+  ! Species quantities
+  !
+  ! ********************************************************
+  ! ********************************************************
+
+  integer, parameter :: maxNumSpecies = 100
+  integer, parameter :: speciesNotInitialized = -9999
+  integer :: NSpecies
+  PetscScalar, dimension(maxNumSpecies) :: Zs, mHats, NHats, THats, dNHatdpsiNs, dTHatdpsiNs
+
+  ! ********************************************************
[…]

### 2014-01-06 fortran/multiSpecies/polynomialDiffMatrices.F90

@@ -1,0 +1,173 @@
+! This module computes the first and second spectral differentiation matrices.
+! Any weight and grid points are allowed.  However, the method is poorly
+! conditioned unless the grid points correspond to Gaussian integration
+! abscissae for the given weight.  No warning is given in the case of
+! poor conditioning.  Presently, the weight function is
+! set to exp(-x^2), but you are free to change it.  When changing the
+! weight, you must also change the functions dweightdxOverWeight
+! and d2weightdx2OverWeight appropriately.
+!
+! This algorithm closely follows poldif.m for MATLAB
+! Part of DMSuite
+! written by J.A.C. Weideman, S.C. Reddy 1998
+! Available here
+! http://www.mathworks.com/matlabcentral/fileexchange/29
+! or here
+! http://dip.sun.ac.za/~weideman/research/differ.html
+!
+! The type PetscScalar is used, so this module can be used in a PETSc
+! application. However, no other PETSc functionality is used, so you can
+! replace the type with e.g. real if you want to build a non-PETSc
+! application.
+!
+! Matt Landreman
+! Massachusetts Institute of Technology
+! Plasma Science & Fusion Center
+! November, 2012
+!
+module polynomialDiffMatrices
+
+  implicit none
+
+#include <finclude/petscsysdef.h>
+
+  public :: makeXPolynomialDiffMatrices
+  private :: weight, dweightdxOverWeight, d2weightdx2OverWeight
+
+contains
+
+  ! ------------------------------------------------------------------
+
+  function weight(x)
+
+    implicit none
+
+    PetscScalar, intent(in) :: x
+    PetscScalar :: weight
+
+    weight = exp(-x*x)
+
+  end function weight
+
+  ! ------------------------------------------------------------------
+
+  function dweightdxOverWeight(x)
+
+    implicit none
+
+    PetscScalar, intent(in) :: x
+    PetscScalar :: dweightdxOverWeight
+
+    dweightdxOverWeight = -2*x
+
+  end function dweightdxOverWeight
+
[…]

### 2014-01-06 fortran/multiSpecies/xGrid.F90

@@ -1,0 +1,221 @@
+  ! This module computes the nodes and weights for Gaussian quadrature for any
+  ! weight function on any interval.  The algorithm used is the Stieltjes
+  ! procedure described in section 4.6.3 of Numerical Recipes, 3rd edition,
+  ! together with the Jacobi matrix procedure of section 4.6.2.  See also
+  ! www.nr.com/webnotes?3
+  ! The module is presently configured to use the weight exp(-x^2) and to
+  ! use the interval [0, Inf), which yields the grid and integration weights
+  ! dicussed in the following paper:
+  ! http://arxiv.org/abs/1210.5289
+  ! This grid is useful for velocity space.
+  ! You can change the "weight" function in the module and the integration
+  ! domain in the makeXGrid subroutine if you wish.
+  !
+  ! The type PetscScalar is used, so this module can be used in a PETSc
+  ! application. However, no other PETSc functionality is used, so you can
+  ! replace the type with e.g. real if you want to build a non-PETSc
+  ! application.
+  !
+  ! The makeXGrid subroutine calls LAPACK, which is
+  ! automatically included in any program that is linked to the PETSc 
+  ! libraries.
+  !
+  ! The makeXGrid subroutine also calls QUADPACK for numerical
+  ! integration.  You can download QUADPACK here:
+  ! http://www.netlib.org/quadpack/
+  !
+  ! Matt Landreman
+  ! Massachusetts Institute of Technology
+  ! Plasma Science & Fusion Center
+  ! November, 2012
+  !
+#define workspaceSize 500  
+  
+  module xGrid
+
+    implicit none
+
+#include <finclude/petscsysdef.h>
+
+    PetscScalar, private, allocatable :: a(:), b(:)
+    integer, private :: j
+    private weight
+
+  contains
+
+    subroutine makeXGrid(N, abscissae, weights)
+      ! This is the main function of the module.
+      !
+      ! Inputs:
+      !  N = number of grid points to generate.
+      !
+      ! Outputs:
+      !   abscissae = grid points
+      !   weights = Gaussian integration weights
+      !
+      ! Both abscissae and weights should be allocated to have size N
+      ! before makeXGrid is called.
+
[…]

### 2014-01-06 fortran/singleSpecies/m20131008_02_plotSFINCSFortran1SpeciesConvergenceScan.m

@@ -1,0 +1,400 @@
+function m20131008_02_plotSFINCSFortran1SpeciesConvergenceScan()
+
+% Name of .h5 HDF5 file from SFINCS:
+filename='sfincsOutput.h5';
+%filename='C:\Users\landreman\Documents\Ubuntu\sfincsOutput_20131008-04_convergenceScan.h5';
+
+%excludeRunsThatDidntConverge = true;
+excludeRunsThatDidntConverge = false;
+
+figureOffset = 0;
+
+colors = [1,0,0;
+    0.8,0.6,0;
+    0,0.7,0;
+    0,0.8,0.9;
+    0,0,1;
+    1,0,1;
+    0.6,0.6,0.6;
+    0,0,0];
+
+
+figureOffset = 50;
+
+info = h5info(filename);
+fprintf('Fields saved in this HDF5 file:\n')
+for i=1:numel(info.Groups(1).Datasets)
+    fprintf('  %s\n',info.Groups(1).Datasets(i).Name)
+end
+
+numRuns = numel(info.Groups);
+
+programMode = h5read(filename,'/programMode');
+
+switch programMode
+    case 1
+        error('This HDF5 file corresponds to a single run (programMode=1), not a convergence scan (programMode=2).')
+    case {2}
+    otherwise
+        error('This HDF5 file does not correspond to a convergence scan (programMode=2).')
+end
+
+location  = getLocationString(1);
+integerToRepresentTrue = h5read(filename,[location,'integerToRepresentTrue']);
+    
+RHSMode1s = 0;
+RHSMode2s = 0;
+didItConverges = false(numRuns,1);
+for i=1:numRuns
+    location  = getLocationString(i);
+    didItConverge = h5read(filename,[location,'didItConverge']);
+    didItConverges(i) = (didItConverge == integerToRepresentTrue);
+    RHSMode = h5read(filename,[location,'RHSMode']);
+    switch RHSMode
+        case 1
+            RHSMode1s = RHSMode1s + 1;
+        case 2
+            RHSMode2s = RHSMode2s + 1;            
+        otherwise
+            error('Unrecognized RHSMode')
+    end
+    if didItConverge ~= integerToRepresentTrue
+        beep
+        fprintf('Warning: run %d did not converge.\n',i)
+    end
+end
+
+if RHSMode1s > 0 && RHSMode2s > 0
+    error('Either all runs should have RHSMode1 or all runs should have RHSMode2')
+end
+
+if RHSMode1s > 0
+    % Scan used RHSMode = 1
[…]

### 2014-01-06 fortran/singleSpecies/polynomialInterpolationMatrix.F90

@@ -1,0 +1,105 @@
+#include <finclude/petscsysdef.h>
+  
+subroutine polynomialInterpolationMatrix(N, M, xk, x, alpxk, alpx, matrix)
+  ! This module returns the matrix for spectral interpolation from any grid to
+  ! any other grid, using any weights. However, the method is poorly
+  ! conditioned unless the points of the initial grid (on which the function
+  ! is already known) correspond to Gaussian integration
+  ! abscissae for the given weight.  No warning is given in the case of
+  ! poor conditioning.  The grid points onto which you interpolate need not
+  ! be nicely distributed.  The weight function does not appear in this module;
+  ! rather, the weight evaluated on both grids is specified as an input.
+  !
+  ! This algorithm closely follows polint.m for MATLAB
+  ! Part of DMSuite
+  ! written by J.A.C. Weideman, S.C. Reddy 1998
+  ! Available here
+  ! http://www.mathworks.com/matlabcentral/fileexchange/29
+  ! or here
+  ! http://dip.sun.ac.za/~weideman/research/differ.html
+  !
+  ! Note from DMSuite's polint function:
+  !     The code implements the barycentric formula; see page 252 in
+  !     P. Henrici, Essentials of Numerical Analysis, Wiley, 1982.
+  !
+  ! The type PetscScalar is used, so this module can be used in a PETSc
+  ! application. However, no other PETSc functionality is used, so you can
+  ! replace the type with e.g. real if you want to build a non-PETSc
+  ! application.
+  !
+  ! Matt Landreman
+  ! Massachusetts Institute of Technology
+  ! Plasma Science & Fusion Center
+  ! November, 2012
+  !
+
+  ! Inputs:
+  !   N = number of grid points on which the function is known.
+  !   M = number of grid points onto which we would like to interpolate.
+  !   xk(N) = grid points on which the function is known.
+  !   alpxk(N) = weight function evaluated on the xk grid.
+  !   x(M) = grid onto which we would like to interpolate.
+  !   alpx(M) = weight function evaluated on the x grid.
+  !
+  ! Outputs:
+  !   matrix(M,N) = interpolation matrix.
+  !
[…]

### 2014-01-06 fortran/singleSpecies/quadpack/d1mach.f

@@ -1,0 +1,360 @@
+      DOUBLE PRECISION FUNCTION D1MACH(I)
+C
+C  DOUBLE-PRECISION MACHINE CONSTANTS
+C
+C  D1MACH( 1) = B**(EMIN-1), THE SMALLEST POSITIVE MAGNITUDE.
+C
+C  D1MACH( 2) = B**EMAX*(1 - B**(-T)), THE LARGEST MAGNITUDE.
+C
+C  D1MACH( 3) = B**(-T), THE SMALLEST RELATIVE SPACING.
+C
+C  D1MACH( 4) = B**(1-T), THE LARGEST RELATIVE SPACING.
+C
+C  D1MACH( 5) = LOG10(B)
+C
+C  TO ALTER THIS FUNCTION FOR A PARTICULAR ENVIRONMENT,
+C  THE DESIRED SET OF DATA STATEMENTS SHOULD BE ACTIVATED BY
+C  REMOVING THE C FROM COLUMN 1.
+C
+C  FOR IEEE-ARITHMETIC MACHINES (BINARY STANDARD), ONE OF THE FIRST
+C  TWO SETS OF CONSTANTS BELOW SHOULD BE APPROPRIATE.
+C
+C  WHERE POSSIBLE, DECIMAL, OCTAL OR HEXADECIMAL CONSTANTS ARE USED
+C  TO SPECIFY THE CONSTANTS EXACTLY.  SOMETIMES THIS REQUIRES USING
+C  EQUIVALENT INTEGER ARRAYS.  IF YOUR COMPILER USES HALF-WORD
+C  INTEGERS BY DEFAULT (SOMETIMES CALLED INTEGER*2), YOU MAY NEED TO
+C  CHANGE INTEGER TO INTEGER*4 OR OTHERWISE INSTRUCT YOUR COMPILER
+C  TO USE FULL-WORD INTEGERS IN THE NEXT 5 DECLARATIONS.
+C
+      INTEGER SMALL(4)
+      INTEGER LARGE(4)
+      INTEGER RIGHT(4)
+      INTEGER DIVER(4)
+      INTEGER LOG10(4)
+      INTEGER SC
+C
+      DOUBLE PRECISION DMACH(5)
+C
+      EQUIVALENCE (DMACH(1),SMALL(1))
+      EQUIVALENCE (DMACH(2),LARGE(1))
+      EQUIVALENCE (DMACH(3),RIGHT(1))
[…]

### 2014-01-06 fortran/singleSpecies/readInput.F90

@@ -1,0 +1,116 @@
+module readInput
+
+  use globalVariables
+
+  implicit none
+
+contains
+
+  subroutine readNamelistInput(filename)
+
+    implicit none
+
+    character(len=*), intent(in) :: filename
+    integer :: fileUnit, didFileAccessWork
+
+    namelist / flowControl / programMode, saveMatlabOutput, outputScheme, MatlabOutputFilename, &
+         outputFilename, parallelizeOverScan, solveSystem, RHSMode
+
+    namelist / geometryParameters / GHat, IHat, iota, epsilon_t, epsilon_h, &
+         helicity_l, helicity_n, B0OverBBar, geometryScheme, &
+         epsilon_antisymm, helicity_antisymm_l, helicity_antisymm_n, &
+         fort996boozer_file, JGboozer_file, normradius_wish, min_Bmn_to_load
+
+    namelist / physicsParameters / speciesMode, Delta, omega, psiAHat, nuN, nuPrime, THat, nHat, EHat, &
+         dnHatdpsi, dTHatdpsi, dPhiHatdpsi, EStar, collisionOperator, constraintScheme, includeXDotTerm, &
+         includeElectricFieldTermInXiDot, useDKESExBDrift, include_fDivVE_term, EStarMin, EStarMax, NEStar
+
+    namelist / resolutionParameters / forceOddNtheta, &
+         Ntheta, NthetaMaxFactor, NthetaMinFactor, NthetaNumRuns, &
+         Nzeta, NzetaMaxFactor, NzetaMinFactor, NzetaNumRuns, &
+         Nxi, NxiMaxFactor, NxiMinFactor, NxiNumRuns, &
+         NL, NLMaxFactor, NLMinFactor, NLNumRuns, &
+         Nx, NxMaxFactor, NxMinFactor, NxNumRuns, &
+         xMax, xMaxMaxFactor, xMaxMinFactor, xMaxNumRuns, &
+         solverTolerance, solverToleranceMinFactor, solverToleranceMaxFactor, solverToleranceNumRuns, &
+         NxPotentialsPerVth, NxPotentialsPerVthMaxFactor, NxPotentialsPerVthMinFactor, NxPotentialsPerVthNumRuns
+
+    namelist / otherNumericalParameters / &
+         useIterativeSolver, thetaDerivativeScheme, whichParallelSolverToFactorPreconditioner, &
+         PETSCPreallocationStrategy
+
[…]

### 2014-01-06 fortran/singleSpecies/sfincs.F90

@@ -1,0 +1,156 @@
+! For PETSc versions 3.4 and later, the line below should be un-commented:
+!#define PetscGetTime PetscTime
+
+! Main program
+
+program sfincs
+  use globalVariables
+  use printToStdout
+  use writeHDF5Output
+  use readInput
+  use scan
+  use petscsysdef
+
+  implicit none
+
+#include <finclude/petscsysdef.h>
+
+  PetscErrorCode ierr
+  PetscLogDouble :: time1, time2
+  integer :: runNum
+
+
+  call PetscInitialize(PETSC_NULL_CHARACTER, ierr)
+  CHKERRQ(ierr)
+
+  call MPI_COMM_SIZE(PETSC_COMM_WORLD, numProcs, ierr)
+  call MPI_COMM_RANK(PETSC_COMM_WORLD, myRank, ierr)
+  ! This program uses 1-based indices to number the MPI processes, consistent with the Fortran
+  ! convention for array indices, not the 0-based indexing used by C and MPI.
+  myRank = myRank + 1
+  masterProc = (myRank==1)
+
+  call printGreeting()
+
+  ! Default values for all input parameters are set in parameters.F90.
+  ! Defaults are replaced by any values given in the input namelist file:
+  call readNamelistInput("input.namelist")
+
+  select case (programMode)
+  case (-1)
+     print *,"Switching the sign of programMode to +1 would generate 1 run."
+     stop
+  case (-2)
+     call processConvergenceScanParameters()
+     print *,"Switching the sign of programMode would yield a scan with",numRunsInScan," runs."
+     stop
+  case (-6)
+     print *,"Switching the sign of programMode would yield a scan with",NEStar," runs."
+     stop
+  end select
+
+  call chooseParallelDirectSolver()
+
+  select case (programMode)
+  case (1)
+     ! Single run
+
+     ! There is no parallelization over runs, so set the sub-commmunicator 
+     ! to be equal to the global communicator:
+     MPIComm = PETSC_COMM_WORLD
+     myRankInSubComm = myRank
+     numProcsInSubComm = numProcs
+     masterProcInSubComm = masterProc
+     myCommunicatorIndex = 1
+
+     call openOutputFile()
+     call allocateArraysForSingleRun()
+     call createHDF5Structures()
+     call solveDKE()
+     call writeRunToOutputFile(1)
+
+  case (2)
+     ! Convergence scan
+
+     call processConvergenceScanParameters()
+
[…]

### 2014-01-06 fortran/singleSpecies/writeHDF5Output.F90

@@ -1,0 +1,883 @@
+#define HAVE_PARALLEL_HDF5
+
+module writeHDF5Output
+
+  use globalVariables
+  use scan
+  use petscsysdef
+  use HDF5
+
+  implicit none
+
+#include <finclude/petscsysdef.h>
+
+  integer, private :: HDF5Error
+  integer(HID_T), private :: HDF5FileID, parallelID, dspaceIDForScalar
+  integer(HID_T), dimension(:), allocatable, private :: dspaceIDForZeta
+  integer(HID_T), dimension(:), allocatable, private :: dspaceIDForTheta
+  integer(HID_T), dimension(:), allocatable, private :: dspaceIDForThetaZeta
+  integer(HID_T), dimension(:), allocatable, private :: dspaceIDForSources
+  integer(HID_T), private :: dspaceIDForTransportMatrix
+  integer(HID_T), dimension(:), allocatable, private :: groupIDs
+
+  integer(HID_T), private :: dsetID_programMode
+
+  integer(HID_T), dimension(:), allocatable, private :: dsetIDs_Ntheta
+  integer(HID_T), dimension(:), allocatable, private :: dsetIDs_Nzeta
+  integer(HID_T), dimension(:), allocatable, private :: dsetIDs_Nxi
+  integer(HID_T), dimension(:), allocatable, private :: dsetIDs_NL
+  integer(HID_T), dimension(:), allocatable, private :: dsetIDs_Nx
+  integer(HID_T), dimension(:), allocatable, private :: dsetIDs_NxPotentialsPerVth
+  integer(HID_T), dimension(:), allocatable, private :: dsetIDs_xMax
+  integer(HID_T), dimension(:), allocatable, private :: dsetIDs_solverTolerance
+  integer(HID_T), dimension(:), allocatable, private :: dsetIDs_theta
+  integer(HID_T), dimension(:), allocatable, private :: dsetIDs_zeta
+  integer(HID_T), dimension(:), allocatable, private :: dsetIDs_thetaDerivativeScheme
+  integer(HID_T), dimension(:), allocatable, private :: dsetIDs_preconditioner_x
+  integer(HID_T), dimension(:), allocatable, private :: dsetIDs_preconditioner_x_min_L
+  integer(HID_T), dimension(:), allocatable, private :: dsetIDs_preconditioner_xi
[…]

### 2014-01-06 matlab/multiSpecies/SFINCS.m

@@ -1,0 +1,2925 @@
+function SFINCS()
+
+% SFINCS:
+% The Stellarator Fokker-Planck Iterative Neoclassical Conservative Solver.
+% Multiple-species version.
+% Written in 2013 by Matt Landreman
+% Massachusetts Institute of Technology
+% Plasma Science & Fusion Center
+
+% Dimensional quantities in this program are normalized to "reference" values:
+% \bar{B} = reference magnetic field, typically 1 Tesla.
+% \bar{R} = reference length, typically 1 meter.
+% \bar{n} = reference density, typically 10^19 m^{-3}, 10^20 m^{-3}, or something similar.
+% \bar{m} = reference mass, typically either the mass of hydrogen or deuterium.
+% \bar{T} = reference temperature in energy units, typically 1 eV or 1 keV.
+% \bar{v} = \sqrt{2 * \bar{T} / \bar{m}} = reference speed
+% \bar{Phi} = reference electrostatic potential, typically 1 V or 1 kV.
+
+% You can choose any reference parameters you like, not just the values
+% suggested here. The code "knows" about the reference values only through
+% the 3 combinations Delta, alpha, and nu_n, input below.
+
+% Radial gradients of density, temperature, and electrostatic potential are
+% specified as derivatives with respect to psi_N, where psi_N is the 
+% toroidal flux normalized to the value at the last closed flux surface. 
+% (psi_N=0 is the magnetic axis, and psi_N=1 is the last closed flux 
+% surface.)
+
+% --------------------------------------------------
+% Program control parameters:
+% --------------------------------------------------
+
+programMode = 1;
+% 1 = single run.
+% 2 = Do a convergence scan and save the results.
+% 3 = Load a previous convergence scan and plot the results. (Doesn't do any new solves.)
+% 4 = Do a nuPrime scan and save the results.
+% 5 = Load a previous nuPrime scan and plot the results. (Doesn't do any new solves.)
+
+% The setting below matters for programMode=3 or programMode=5 only:
+dataFileToPlot = 'm20130318_02_SFINCS_2013-03-18_14-47_convergenceScan_convergenceScan.mat';
+
+RHSMode = 1;
+% 1 = Use a single right-hand side.
+% 2 = Use multiple right-hand sides to compute the transport matrix.
+% At present, RHSMode==2 is only allowed when a single species is used.
+
[…]

### 2014-01-06 matlab/multiSpecies/m20130312_02_SpectralNodesWeightsAndDifferentiationMatricesForV.m

@@ -1,0 +1,48 @@
+function [x, ddx, d2dx2, weights] = m20130312_02_SpectralNodesWeightsAndDifferentiationMatricesForV(N, k, scale, pointAtZero)
+% Computes the abscissae (i.e. grid points), Gaussian integration weights,
+% and spectral colocation differentiation matrices for functions on the
+% domain [0, Inf] which behave roughly as exp(-x^2) for large x.
+%
+% Written by Matt Landreman - July 6, 2012
+%
+% Inputs:
+% N = desired number of abscissae
+% k = extra power of x in the orthogonality relation and weight. 
+%        (Usually, k=0 works well.)
+% scale = factor by which to scale all the abscissae, (typically in the range 0.7 to 1)
+%               so as to not waste points at very large x.
+% pointAtZero = true or false, depending on whether or not you require a
+%               grid point at x=0.
+%
+% Outputs:
+% x = column vector of abscissae
+% ddx = first derivative matrix
+% d2dx2 = second derivative matrix
+% weights = Gaussian integration weights
+%
+% This function depends on
+% m20121125_05_GaussWeightsAndAbscissae.m
+% as well as
+% poldif.m
+% The latter is part of the DMSuite package by S.C. Reddy and J.A.C. Weideman, available at
+% http://www.mathworks.com/matlabcentral/fileexchange/29
+% or here:
+% http://dip.sun.ac.za/~weideman/research/differ.html
+
+if pointAtZero
+    if k ~= 0
+        error('When k ~= 0, this subroutine only works with no point at 0.')
+    end
+    [x, weights]=m20121007_01_generateGaussRadauPointsAndWeights(N, @(x) (x.^k) .* exp(-x.*x), 0, Inf);
+else
+    [x, weights]=m20121125_05_GaussWeightsAndAbscissae(N, @(x) (x.^k) .* exp(-x.*x), 0, Inf);
+end
+weights=weights./((x.^k) .* exp(-x.*x));
+x = x*scale;
+weights = weights*scale;
+weightFactors=[k./x-2*x, k*(k-1)./(x.*x)-2*(2*k+1)+4*x.*x]';
+differentiationMatrices=poldif(x,(x.^k) .* exp(-x.*x),weightFactors);
+ddx=differentiationMatrices(:,:,1);
+d2dx2=differentiationMatrices(:,:,2);
+
+end

### 2014-01-06 matlab/multiSpecies/m20130728_01_solveLinearSystem.m

@@ -1,0 +1,142 @@
+function [soln, didItConverge, residual] ...
+    = m20130728_01_solveLinearSystem(matrix, rhs, preconditionerHandle, ...
+    tryIterativeSolvers, orderOfSolversToTry, requestedAbsoluteTolerance, maxIterations, restart, ...
+    figureNum, tryDirectSolverIfIterativeSolversFail)
+% Allowed values for orderOfSolversToTry:
+% 1 = GMRES
+% 2 = BiCGStab
+% 3 = BiCGStab(l)
+% 4 = TFQMR
+% 5 = CGS
+%
+% rhs may have >1 column. soln will have the same size as rhs.
+% If figureNum <= 0, no figure will be generated.
+
+timeWhenSolveStarted = tic;
+
+numSolversToTry = numel(orderOfSolversToTry);
+
+if ~tryIterativeSolvers
+    
+    fprintf('Direct solution\n')
+    soln = matrix \ rhs;
+    didItConverge = true;
+    residual = 0;
+    
+else
+    % Use an iterative Krylov-space solver.
+    
+    soln = zeros(size(rhs));
+    numCols = size(rhs,2);
+    
+    for col=1:numCols
+        attempt=0;
+        keepTrying = true;
+        x0 = zeros(size(rhs));
+        relativeTolerance = requestedAbsoluteTolerance;
+        while keepTrying
+            attempt = attempt+1;
+            tic
+            switch orderOfSolversToTry(attempt)
+                case 1
+                    solverName = 'GMRES';
+                    fprintf('GMRES solution ***********\n')
+                    [soln0,fl0,rr0,it0,rv0]=gmres(matrix,rhs(:,col),restart,relativeTolerance,maxIterations/restart,preconditionerHandle, [], x0);
+                case 2
+                    solverName = 'BiCGStab';
+                    fprintf('BiCGStab solution ***********\n')
+                    [soln0,fl0,rr0,it0,rv0]=bicgstab(matrix,rhs(:,col),relativeTolerance,maxIterations,preconditionerHandle, [], x0);
+                case 3
+                    solverName = 'BiCGStab(l)';
+                    fprintf('BiCGStab(l) solution ***********\n')
+                    [soln0,fl0,rr0,it0,rv0]=bicgstabl(matrix,rhs(:,col),relativeTolerance,maxIterations,preconditionerHandle, [], x0);
+                case 4
+                    solverName = 'TFQMR';
[…]

### 2014-01-06 matlab/singleSpecies/sfincs.m

@@ -1,0 +1,2790 @@
+function sfincs()
+
+% SFINCS:
+% The Stellarator Fokker-Planck Iterative Neoclassical Conservative Solver.
+% Single species version.
+% Original version written in 2013 by Matt Landreman
+% Massachusetts Institute of Technology
+% Plasma Science & Fusion Center
+    
+% Dimensional quantities in this program are normalized to "reference" values:
+% \bar{B} = reference magnetic field, typically 1 Tesla.
+% \bar{R} = reference length, typically 1 meter.
+% \bar{n} = reference density, typically 10^19 m^{-3}, 10^20 m^{-3}, or something similar.
+% \bar{m} = reference mass, typically either the mass of hydrogen or deuterium.
+% \bar{T} = reference temperature in energy units, typically 1 eV or 1 keV.
+% \bar{v} = \sqrt{2 * \bar{T} / \bar{m}} = reference speed
+% \bar{Phi} = reference electrostatic potential, typically 1 V or 1 kV.
+
+% You can choose any reference parameters you like, not just the values
+% suggested here. The code "knows" about the reference values only through
+% the 3 combinations Delta, omega, and nuN or nuPrime, input below.
+
+% Radial gradients of density, temperature, and electrostatic potential are
+% specified as derivatives with respect to psi_N, where psi_N is the 
+% toroidal flux normalized to the value at the last closed flux surface. 
+% (psi_N=0 is the magnetic axis, and psi_N=1 is the last closed flux 
+% surface.)
+    
+% --------------------------------------------------
+% Program flow control parameters:
+% --------------------------------------------------
+
+programMode = 1;
+% 1 = single run.
+% 2 = Do a convergence scan and save the results.
+% 3 = Load a previous convergence scan and plot the results. (Doesn't do any new solves.)
+% 4 = Do a nuPrime scan and save the results.
+% 5 = Load a previous nuPrime scan and plot the results. (Doesn't do any new solves.)
+% 6 = Do a scan of E_r for several ways of treating E_r. Keep nuPrime fixed. Plot and save the results.
+% 7 = Load a previous programMode=5 scan and plot the results.
+
+% The setting below matters for programMode = 3, 5, or 7 only:
[…]

### 2014-02-04 doc/sfincsPaper/sfincsPaper.tex
Added appendix on Simakov benchmark

@@
 = \left< (p_\perp - p_\| ) \nabla_\| B \right>.
+\label{ParallelMomentum}
 \end{equation}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%Appendix on Simakov benchmark
+
+\section{Ion transport matrix at high collisionality}
+From the model presented in Ref.~\cite{AndreiPer2009} we can derive expressions for $L_{jk}$ of Eq.~(\ref{eq:matrix}) in the Pfirsch-Schl{\"u}ter regime. 
+The coefficients then depend on the function $u$ given by the solution to Eq.~(8) in Ref.~\cite{AndreiPer2009}, which is proportional to the parallel current divided by $B$. 
+All coefficients but $L_{33}$ are straightforwardly obtained from Eqs.~(14), (18) and (26) in Ref.~\cite{AndreiPer2009} for the radial ion heat flux, the parallel ion flow and the radial current respectively, by %properly choosing
+suitable choices of 
+the thermodynamic forces in the right-hand-side of Eq.~(\ref{eq:matrix}) for different solves and using the symmetry of the transport matrix.
+To find the parallel conductivity coefficient $L_{33}$ we substitute the pressure anisotropy, given by Eq.~(20) in Ref.~\cite{AndreiPer2009}, into Eq.~(\ref{ParallelMomentum}) (of this paper) 
+in the absence of radial gradients (i.e. when $E_{||}$ is the only thermodynamic force present). $L_{33}$ is then found from the flow $\left< V_\| B \right>$.
+
+Expressions for the matrix coefficients in the Pfirsch-Schl{\"u}ter regime are summarized in Eqs.~(\ref{eq:LPSregime})-(\ref{eq:PSregimeFunctions}). 
+Note that the constants in the function ${K}_2 \left(\psi\right)$ in $L_{23}$ have been modified from their values in ${K}_{2}^{\mathrm{Simakov}} \left(\psi\right)$ obtained from Ref.~\cite{AndreiPer2009}, by a more precise solution to the Spitzer problem.
+
+%\begin{equation}
+%\begin{array}{c}
+\begin{align}
+\begin{split}
+L_{11} & = 
+ 0.96 \cdot 2^{1/2} \cdot  \frac{3}{4}  \frac{\left(G + \iota I\right)^2}{ \iota^2 G^2} 
+ {G}_1 \left(\psi\right) \frac{1}{\nu'},\\
+ %%
+L_{12} & = L_{21} =  0.96 \cdot 2^{1/2} \frac{\left({G} + \iota {I}\right)^2}{ \iota^2 {G}^2} \left[ 3.245 \, {G}_1 \left(\psi\right) + 0.085 \, {G}_2 \left(\psi\right)  \right] \frac{1}{\nu'}, \\
+%% 
+L_{13} & = L_{31}  = \frac{\left\langle {u} {B}^2  \right\rangle}{{G} \iota} - \frac{\left\langle {B}^2\right\rangle}{{G} \iota} {K}_1\left(\psi\right), \\
+%%
+L_{22} & = 2^{1/2} \cdot  \frac{8}{5} \frac{1}{\iota^2 {G}^2} B_0^2 {H} \left(\psi\right) \, \nu', \\
+	%\textcolor{red}{+ \left( \,0.96 \cdot 2^{1/2} \cdot \frac{5}{2} \frac{\left({G} + \iota {I}\right)^2}{ \iota^2 {G}^2} \left[ 3.245 {G}_1 \left(\psi\right) + 0.085 {G}_2 \left(\psi\right)  \right] \frac{1}{\nu'}\right)}, \\
+	%%
+L_{23} & = L_{32} = \frac{1}{\iota {G}} \left[\frac{5}{2} \left\langle {u} {B}^2  \right\rangle - \frac{5}{2} {K}_1\left(\psi\right) \left\langle {B}^2\right\rangle + {K}_2\left(\psi\right) \left\langle {B}^2\right\rangle\right], \\
+%%
+L_{33} & = \frac{1}{3 \cdot 0.96 \cdot 2^{1/2}} \frac{1}{\left({G} + \iota {I}\right)^2} \frac{\left\langle {B}^2\right\rangle^2}{\left\langle \left( {\nabla}_\| {B}\right)^2\right\rangle} \, \nu',	
+%\end{array}
+\label{eq:LPSregime}
+\end{split}
+\end{align}
+%\end{equation}
+%where
+\begin{align}
+\begin{split}
+& {G}_1 \left(\psi\right) %& 
+= \frac{\left\langle \left( {\nabla}_\| \ln {B}\right)  {\nabla}_\| \left({u} {B}^2\right)\right\rangle^2}{\left\langle \left( {\nabla}_\| {B}\right)^2\right\rangle} -
+ \left\langle \left[\frac{ {\nabla}_\| \left({u} {B}^2\right)}{{B}}\right]^2\right\rangle,  \\
+ %%
+& {G}_2 \left(\psi\right) %& 
+ = \left\langle {u} \left({\nabla}_\| \ln {B}\right) {\nabla}_\| \left({u} {B}^2\right) \right\rangle
+ - \frac{\left\langle \left( {\nabla}_\| \ln {B}\right)  {\nabla}_\| \left({u} {B}^2\right)\right\rangle
+ \left\langle {u} \left( {\nabla}_\| {B}\right)^2\right\rangle}{\left\langle \left( {\nabla}_\| {B}\right)^2\right\rangle},  \\
+ %%
+& {K}_1 \left(\psi\right) %& 
+ = \frac{\left\langle \left( {\nabla}_\| \ln {B}\right)  {\nabla}_\| \left({u} {B}^2\right)\right\rangle}{2 \left\langle \left( {\nabla}_\| {B}\right)^2\right\rangle}, \\
+ %%
+& {K}_2 \left(\psi\right) %& 
+ = 1.97213 \frac{\left\langle {u} {B}^2  \right\rangle}{\left\langle {B}^2  \right\rangle} - 1.03287 \cdot 2 {K}_1 \left(\psi\right)
+ + 0.09361 \frac{
+ \left\langle {u} \left( {\nabla}_\| {B}\right)^2\right\rangle}{\left\langle \left( {\nabla}_\| {B}\right)^2\right\rangle}, \\
+ %%
+&  {K}_{2}^{\mathrm{Simakov}} \left(\psi\right) %& 
+  = 1.77 \frac{\left\langle {u} {B}^2  \right\rangle}{\left\langle {B}^2  \right\rangle} - 0.91 \cdot 2 {K}_1 \left(\psi\right)
+ + 0.05 \frac{
+ \left\langle {u} \left( {\nabla}_\| {B}\right)^2\right\rangle}{\left\langle \left( {\nabla}_\| {B}\right)^2\right\rangle}, \\
+ %%
+& {H} \left(\psi\right) %& 
+ = \frac{\left\langle {u} {B}^2  \right\rangle^2}{\left\langle {B}^2  \right\rangle} - \left\langle {u}^2 {B}^2  \right\rangle.
+\label{eq:PSregimeFunctions}
+\end{split}
+\end{align}
+
 \bibliography{sfincsPaper}

### 2014-02-04 doc/sfincsPaper/sfincsPaper.tex
Fixed lowercase kv -> kV in sfincsPaper.tex

@@
 and the DKES model predicts $8\%$ more bootstrap current than the full trajectory model.
-Interestingly, if the electric field exceeds 60 kv/m in the inward (ion root) direction, the bootstrap current
+Interestingly, if the electric field exceeds 60 kV/m in the inward (ion root) direction, the bootstrap current
 in the full trajectory model changes sign, whereas there is no sign change in the DKES model.

### 2014-02-05 doc/sfincsPaper/20131220_Submission_v1_to_arXiv/sfincsPaper.tex
Added all files from arXiv v1 submission of sfincs paper

--- 2014-01-06 doc/sfincsPaper/sfincsPaper.tex
@@
 \affiliation{Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA}
-\affiliation{Plasma Science and Fusion Center, Massachusetts Institute of Technology, Cambridge, MA, 02139, USA}
 \author{H M Smith}
@@
 E_* = \frac{c G}{\iota \vi B_0}\frac{d\Phi_0}{d\psi},
-\label{eq:EStar}
-\end{equation}
-and not on any other individual parameters such as density, temperature, $G$, etc.
+\end{equation}
+and not on any other individual parameters such as density, temperature, $I$, etc.
 In (\ref{eq:nuPrime}), $\nuii = 4\sqrt{2\pi} n Z^4 e^4 \ln\Lambda/(3 m^{1/2} T^{3/2})$
@@
 using linearized Fokker-Planck collisions.
-The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
-of the normalized radial electric field (\ref{eq:EStar}).
 \label{fig:ErComparison_LHD}}
@@
 using linearized Fokker-Planck collisions.
-The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
-of the normalized radial electric field (\ref{eq:EStar}).
 \label{fig:ErComparison_W7X}}
@@
 \begin{figure}[h!]
-% Figure generated using m20131212_02_PlotSFINCS2SpeciesEScanForPaper.m
+% Figure generated using m20131205_02_plotSFINCSFortranErComparisonForPaper.m
 % SFINCS run on hopper and edison in 
@@
 \caption{(Color online) Comparison of collision operators for LHD standard geometry at $E_r=0$.
-The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
-of the collisionality (\ref{eq:nuPrime}).
 \label{fig:collisionComparison_LHD}}
@@
 \caption{(Color online) Comparison of collision operators for W7-X standard geometry at $E_r=0$.
-The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
-of the collisionality (\ref{eq:nuPrime}).
 \label{fig:collisionComparison_W7X}}

### 2014-02-05 doc/sfincsPaper/sfincsPaper.tex
Added DKES and Simakov-Helander benchmark to figures 4-5, along with associated descriptive text, to sfincs paper.

@@
 %\documentclass[aip,graphicx]{revtex4-1}
-\documentclass[12pt]{revtex4}
+\documentclass[12pt,superscriptaddress]{revtex4}
 \usepackage{latexsym, color, graphicx, comment}
@@
 \newcommand{\todo}[1]{\textcolor{red}{To do: #1}}
-\newcommand{\changed}[1]{\textcolor{green}{#1}}
 \newcommand{\sgn}{\mathrm{sgn}}
@@
 \newcommand{\Sah}{S_{a\mathrm{h}}}
+%\newcommand{\changed}[1]{#1}
+\newcommand{\changed}[1]{\textcolor{red}{#1}}
 
@@
 \affiliation{Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA}
-\affiliation{Plasma Science and Fusion Center, Massachusetts Institute of Technology, Cambridge, MA, 02139, USA}
 \author{H M Smith}
+\affiliation{Max-Planck-Institut f\"{u}r Plasmaphysik, 17491 Greifswald, Germany}
+\author{A Moll\'{e}n}
+\affiliation{Department of Applied Physics, Nuclear Engineering,
+  Chalmers University of Technology and Euratom-VR Association,
+  G\"oteborg,
+  Sweden}
 \author{P Helander}
@@
 
-When the pure pitch-angle scattering collision operator 
-and DKES trajectories are chosen, the kinetic equation
-solved in SFINCS becomes identical to the one solved in the DKES code \cite{DKES1, DKES2}.
-In this case, it was verified that the two codes agreed for all elements of the transport
-matrix.
-
 \section{Comparison of $E_r$ terms}
@@
 
+\changed{
+In the short-mean-free-path limit $\nu' \gg 1$, the ion transport and flow can be computed
+analytically in terms of the parallel current \cite{AndreiPer2009}.
+The transport matrix elements associated with the Fokker-Planck collision operator
+may therefore be extracted from Ref. \cite{AndreiPer2009} and are summarized in Appendix \ref{apdx:SH}.
+Plotted in figures \ref{fig:collisionComparison_LHD} and \ref{fig:collisionComparison_W7X} (red dashed lines),
+these analytic high-collisionality limits agree quite well with the Fokker-Planck SFINCS computations
+in the appropriate $\nu' \gg 1$ limit.
+}
+
+\changed{
+At any collisionality, when the pure pitch-angle scattering collision operator 
+and DKES trajectories are chosen, the kinetic equation
+solved in SFINCS becomes identical to the one solved in the DKES code \cite{DKES1, DKES2}.
+In this case, it was verified that the two codes agreed for all elements of the transport
+matrix, as demonstrated in figure \ref{fig:collisionComparison_W7X}.
+For this figure, the monoenergetic transport coefficients
+computed by DKES have been integrated over velocity with the appropriate
+weights and normalized in the same way as (\ref{eq:matrix}).
+}
+
 \begin{figure}[h!]
-% Figure generated using m20131208_01_plotSFINCSFortranAndMatlabNuScansForPaper.m
+% Figure generated using m20140127_01_plotSFINCSFortranAndMatlabNuScansForPaper.m
+% *********************************
+% Previous version of figure generated using m20131208_01_plotSFINCSFortranAndMatlabNuScansForPaper.m
 % High collisionality runs done in matlab, with data in
@@
 % *********************************
-% Old version of figure:
+% Previous version of figure:
 % Figure generated using m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper.m
@@
 % *********************************
-\includegraphics{m20131208_01_plotSFINCSFortranAndMatlabNuScansForPaper_LHD.eps}
+\includegraphics{m20140127_01_plotSFINCSFortranAndMatlabNuScansForPaper_LHD.eps}
+%\includegraphics{m20131208_01_plotSFINCSFortranAndMatlabNuScansForPaper_LHD.eps}
 %\includegraphics{m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper_LHD.eps}
-\caption{(Color online) Comparison of collision operators for LHD standard geometry at $E_r=0$.
+\caption{(Color online) 
 The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
-of the collisionality (\ref{eq:nuPrime}).
+of the collisionality (\ref{eq:nuPrime}) for LHD geometry at $E_r=0$.
+SFINCS computations for three different collision operators are compared.
+\changed{
+Dashed red lines indicate the analytic high-collisionality limits
+for Fokker-Planck collisions, discussed in appendix \ref{apdx:SH} and in Ref. \cite{AndreiPer2009},
+which agree quite well with the SFINCS Fokker-Planck computations at high collisionality.
+}
 \label{fig:collisionComparison_LHD}}
@@
 \begin{figure}[h!]
-% Figure generated using m20131208_01_plotSFINCSFortranAndMatlabNuScansForPaper.m
+% Figure generated using m20140127_01_plotSFINCSFortranAndMatlabNuScansForPaper.m
+% *********************************
+% Previous version of figure generated using m20131208_01_plotSFINCSFortranAndMatlabNuScansForPaper.m
 % High collisionality runs done in matlab, with data in
@@
 % *********************************
-\includegraphics{m20131208_01_plotSFINCSFortranAndMatlabNuScansForPaper_W7X.eps}
+\includegraphics{m20140127_01_plotSFINCSFortranAndMatlabNuScansForPaper_W7X.eps}
+%\includegraphics{m20131208_01_plotSFINCSFortranAndMatlabNuScansForPaper_W7X.eps}
 %\includegraphics{m20131130_01_plotSFINCSCollisionOperatorComparisonForPaper_W7X.eps}
-\caption{(Color online) Comparison of collision operators for W7-X standard geometry at $E_r=0$.
+\caption{(Color online)
 The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
-of the collisionality (\ref{eq:nuPrime}).
+of the collisionality (\ref{eq:nuPrime}) for LHD geometry at $E_r=0$.
+SFINCS computations for three different collision operators are compared.
+\changed{
+Dashed red lines indicate the analytic high-collisionality limits
+for Fokker-Planck collisions, discussed in appendix \ref{apdx:SH} and in Ref. \cite{AndreiPer2009},
+which agree quite well with the SFINCS Fokker-Planck computations at high collisionality.
+Also shown (black crosses) are the transport matrix elements
+computed using the DKES code \cite{DKES1,DKES2}, which uses a pitch-angle scattering collision operator,
+demonstrating excellent agreement with SFINCS when the latter is run with the same collision model. 
+}
 \label{fig:collisionComparison_W7X}}
@@
 \section{Ion transport matrix at high collisionality}
-From the model presented in Ref.~\cite{AndreiPer2009} we can derive expressions for $L_{jk}$ of Eq.~(\ref{eq:matrix}) in the Pfirsch-Schl{\"u}ter regime. 
-The coefficients then depend on the function $u$ given by the solution to Eq.~(8) in Ref.~\cite{AndreiPer2009}, which is proportional to the parallel current divided by $B$. 
-All coefficients but $L_{33}$ are straightforwardly obtained from Eqs.~(14), (18) and (26) in Ref.~\cite{AndreiPer2009} for the radial ion heat flux, the parallel ion flow and the radial current respectively, by %properly choosing
+\label{apdx:SH}
+
+\changed{
+From the analytic calculations presented in Ref.~\cite{AndreiPer2009} we can derive expressions
+for $L_{jk}$ of Eq.~(\ref{eq:matrix}) in the Pfirsch-Schl{\"u}ter regime. 
+The transport matrix elements then depend on the function $u$ given by the solution to Eq.~(8) in Ref.~\cite{AndreiPer2009};
+$u$ is proportional to the parallel current divided by $B$. 
+All coefficients but $L_{33}$ are straightforwardly obtained from Eqs.~(14), (18) and (26) in Ref.~\cite{AndreiPer2009} for the radial ion heat flux, the parallel ion flow, and the radial current respectively, by %properly choosing
 suitable choices of 
-the thermodynamic forces in the right-hand-side of Eq.~(\ref{eq:matrix}) for different solves and using the symmetry of the transport matrix.
+the thermodynamic forces in the right-hand-side of Eq.~(\ref{eq:matrix}) and using the symmetry of the transport matrix.
 To find the parallel conductivity coefficient $L_{33}$ we substitute the pressure anisotropy, given by Eq.~(20) in Ref.~\cite{AndreiPer2009}, into Eq.~(\ref{ParallelMomentum}) (of this paper) 
-in the absence of radial gradients (i.e. when $E_{||}$ is the only thermodynamic force present). $L_{33}$ is then found from the flow $\left< V_\| B \right>$.
-
-Expressions for the matrix coefficients in the Pfirsch-Schl{\"u}ter regime are summarized in Eqs.~(\ref{eq:LPSregime})-(\ref{eq:PSregimeFunctions}). 
-Note that the constants in the function ${K}_2 \left(\psi\right)$ in $L_{23}$ have been modified from their values in ${K}_{2}^{\mathrm{Simakov}} \left(\psi\right)$ obtained from Ref.~\cite{AndreiPer2009}, by a more precise solution to the Spitzer problem.
-
+in the absence of radial gradients (i.e. when $E_{||}$ is the only thermodynamic force present). Then $L_{33}$ can be found from the flow $\left< V_\| B \right>$.
+}
+
+\changed{
+Expressions for the matrix coefficients in the Pfirsch-Schl{\"u}ter regime are summarized in 
+Eqs.~(\ref{eq:LPSregime})-(\ref{eq:PSregimeFunctions}). 
+Three numerical coefficients in the function ${K}_2^{\mathrm{Simakov}} \left(\psi\right)$
+arise from generalized Spitzer problems, 
+which were solved in Ref.~\cite{AndreiPer2009} by keeping a small number of Laguerre polynomials in kinetic energy.
+When these generalized Spitzer problems are solved numerically keeping many more energy polynomials,
+we obtain the more accurate coefficients given in ${K}_{2}$ below.
+}
+
+\changed{
 %\begin{equation}
@@
 \end{align}
+}
 

### 2014-02-05 doc/sfincsPaper/sfincsPaper.tex
Added Simakov-Helander benchmark as a new figure 6 in sfincs paper

@@
 \changed{
-In the short-mean-free-path limit $\nu' \gg 1$, the ion transport and flow can be computed
-analytically in terms of the parallel current \cite{AndreiPer2009}.
-The transport matrix elements associated with the Fokker-Planck collision operator
-may therefore be extracted from Ref. \cite{AndreiPer2009} and are summarized in Appendix \ref{apdx:SH}.
-Plotted in figures \ref{fig:collisionComparison_LHD} and \ref{fig:collisionComparison_W7X} (red dashed lines),
-these analytic high-collisionality limits agree quite well with the Fokker-Planck SFINCS computations
-in the appropriate $\nu' \gg 1$ limit.
-}
-
-\changed{
-At any collisionality, when the pure pitch-angle scattering collision operator 
+When the pure pitch-angle scattering collision operator 
 and DKES trajectories are chosen, the kinetic equation
@@
 
+\changed{
+In the short-mean-free-path limit $\nu' \gg 1$, the ion transport and flow can be computed
+analytically in terms of the parallel current \cite{AndreiPer2009}.
+The transport matrix elements associated with the Fokker-Planck collision operator
+may therefore be extracted from Ref. \cite{AndreiPer2009} and are summarized in Appendix \ref{apdx:SH}.
+Plotted in figure \ref{fig:SimakovHelanderComparison} (dashed and dot-dashed lines),
+these analytic high-collisionality limits agree quite well with the Fokker-Planck SFINCS computations
+in the appropriate $\nu' \gg 1$ limit.
+}
+
 \begin{figure}[h!]
-% Figure generated using m20140127_01_plotSFINCSFortranAndMatlabNuScansForPaper.m
+% Figure generated using m20140205_01_plotSFINCSFortranAndMatlabNuScansForPaper.m
 % *********************************
@@
 % *********************************
-\includegraphics{m20140127_01_plotSFINCSFortranAndMatlabNuScansForPaper_LHD.eps}
+\includegraphics{m20140205_01_plotSFINCSFortranAndMatlabNuScansForPaper_LHD.eps}
 %\includegraphics{m20131208_01_plotSFINCSFortranAndMatlabNuScansForPaper_LHD.eps}
@@
 SFINCS computations for three different collision operators are compared.
-\changed{
-Dashed red lines indicate the analytic high-collisionality limits
-for Fokker-Planck collisions, discussed in appendix \ref{apdx:SH} and in Ref. \cite{AndreiPer2009},
-which agree quite well with the SFINCS Fokker-Planck computations at high collisionality.
-}
 \label{fig:collisionComparison_LHD}}
@@
 \begin{figure}[h!]
-% Figure generated using m20140127_01_plotSFINCSFortranAndMatlabNuScansForPaper.m
+% Figure generated using m20140205_01_plotSFINCSFortranAndMatlabNuScansForPaper.m
 % *********************************
@@
 % *********************************
-\includegraphics{m20140127_01_plotSFINCSFortranAndMatlabNuScansForPaper_W7X.eps}
+\includegraphics{m20140205_01_plotSFINCSFortranAndMatlabNuScansForPaper_W7X.eps}
 %\includegraphics{m20131208_01_plotSFINCSFortranAndMatlabNuScansForPaper_W7X.eps}
@@
 The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
-of the collisionality (\ref{eq:nuPrime}) for LHD geometry at $E_r=0$.
+of the collisionality (\ref{eq:nuPrime}) for W7-X geometry at $E_r=0$.
 SFINCS computations for three different collision operators are compared.
 \changed{
-Dashed red lines indicate the analytic high-collisionality limits
-for Fokker-Planck collisions, discussed in appendix \ref{apdx:SH} and in Ref. \cite{AndreiPer2009},
-which agree quite well with the SFINCS Fokker-Planck computations at high collisionality.
 Also shown (black crosses) are the transport matrix elements
@@
 \label{fig:collisionComparison_W7X}}
+\end{figure}
+
+\begin{figure}[h!]
+% Figure generated using m20140205_02_plotSFINCSVsSimakovHelanderLimitForPaper.m
+\includegraphics{m20140205_02_plotSFINCSVsSimakovHelanderLimitForPaper.eps}
+\caption{(Color online)
+\changed{
+The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
+of the collisionality (\ref{eq:nuPrime}) for LHD and W7-X geometry at $E_r=0$.
+SFINCS results shown were computed using the linearized Fokker-Planck collision operator
+(so the solid curves here are
+identical to the red curves in figures \ref{fig:collisionComparison_LHD}-\ref{fig:collisionComparison_W7X}.)
+Dashed and dot-dashed lines indicate the analytic high-collisionality limits
+for Fokker-Planck collisions, discussed in Ref. \cite{AndreiPer2009} and in appendix \ref{apdx:SH},
+which agree quite well with the SFINCS computations at high collisionality.
+}
+\label{fig:SimakovHelanderComparison}}
 \end{figure}

### 2014-02-05 doc/sfincsPaper/refereeResponse/sfincsPaper_refereeResponse.tex
Started referee response

\documentclass[11pt]{article}
\usepackage[dvips,usenames,dvipsnames]{color}
\usepackage[dvips]{graphicx}
\usepackage{delarray}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\setlength {\parindent} { 10mm} 
\setlength{\textheight}{240mm} 
\setlength{\textwidth}{160mm} 
\setlength{\oddsidemargin}{0mm}
\setlength{\topmargin}{-10mm}

% newcommands
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
\newenvironment{referee}{\begin{quote}\it\color{Blue}}{\end{quote}}

\begin{document}  
\section*{Response to the referee}

Re: ``Comparison of particle trajectories and collision operators for
collisional transport in nonaxisymmetric plasmas'', by
M Landreman et al., manuscript POP42544.\newline

Thank you for your careful reading of the manuscript and constructive suggestions.
Below we address each of your points in detail.

\begin{referee}
If it is not difficult numerically to retain the poloidal and toroidal magnetic drifts in the
equations of motion, as it is written in the middle of page 13, why do authors keep these term
and remove only the component from eqs.(12) and (14)? I understand that it is to make
the ``DKES trajectories'', but it seems that approximating in (12) and (14) still
holds the conservation of $\mu$. Is it just because ``The omitted terms may be important in other
situations, but here our primary interest is the treatment of the $d\Phi/d\psi$ terms'' that the magnetic
drift term is completely dropped from (12) and (14), or is there any other unexplained problems
in solving the transport equation with the poloidal and toroidal magnetic drift terms kept?
\end{referee}

Response goes here.

\begin{referee}
Here the conservation properties of mass and energy in three trajectory models are discussed,
but how about the parallel momentum balance? Relating to this question, in page 20 - 21
authors have discussed about the relation between the momentum balance and friction force
from different collision operator models. I agree with the authors that the momentum
conservation property in collision operator is important to evaluate correct transport coefficients.
However, by seeing the figures 3(c)(d), parallel momentum balance seems to be broken in the
partial trajectory model, even if the full linearized Fokker-Planck operator is used.
The momentum balance equation (36) is derived by taking the $\vect{b}\cdot \int d^3v\, m\vect{v}$ moment of the drift
kinetic equation (15). Then, for each three trajectory model, taking the $\vect{b}\cdot \int d^3v\, m\vect{v}$ moment of
eq. (15) with combining one of eq. (17),(18), or (19), does it yield the proper parallel momentum
balance equation as eq.(36)? Or more simply, do these three trajectory models evaluate the
parallel viscosity $\left<\vect{b}\cdot\nabla\cdot\vect{\pi}\right>$ correctly?
\end{referee}

Response goes here.

\begin{referee}
I do not understand why the source term (25) does not work well with the pitch-angle scattering
operator or P.-A. scattering + momentum conservation operator. It is not the difference in the
collision operator but difference in the trajectory model that is related to the conservation
property of energy, (22). If the original motivation of introducing the term is to maintain the
mass and energy conservation property for finite-Er, then why you need to change the form of
source/sink term if you choose P.-A. scattering operator, which actually conserves the energy,
while you do not need according to the trajectory model? Does it mean diffusion process in
v-coordinate is essential for the source/sink term to work properly?
Please explain the reason of changing the source term according to the collision operator more
detail.
\end{referee}

Response goes here.

\begin{referee}
Concerning the source/sink term, I think that the usage of such an artificial term is relevant if
the effect of the term is small. Can you show the effect of the source/sink term is small or not in
some cases shown in Section VI, by comparing $\left< \int d^3v\, S(\psi,v) \right> \Delta t$ and $n(\psi)$, for example, where
$\Delta t$ is some characteristic time scale such as collision time or transport time scale.
\end{referee}

Response goes here.

\begin{referee}
Explanation for Fig 1 and 2: Though authors write ``As $E_r \to 0$, all the matrix elements converge
smoothly to their $E_r = 0$ limits'', since the horizontal axis of these figures are in log scale, we
cannot judge if their claim is true or not. Please show the value of $L_{jk}(E_r=0)$ in these figures,
too.
\end{referee}

Response goes here.

\begin{referee}
It is estimated that $E_* \sim 0.1 - 0.01$ in most of the plasma. However, if the core $T_e$ is higher than
$T_i$ and ambipolar $E_r$ is positive (electron root) in ECH-heated plasma, $E_*$ for ions easily becomes
nearly, or even larger than, unity near the magnetic axis, since $\rho_\theta \propto 1/B_p$ is large and usually
$|E_r|$(ion root) $\ll E_r$(electron root). See Yokoyama et al., Nuclear Fusion (2007), 1213 for example.
I think therefore the statement ``In most of the plasma, $E_*$ is however expected to be smaller
than a few percent.'' is too strong and may be misleading.
\end{referee}

Response goes here.

\begin{referee}
Appendix: The authors claims that the parallel flow and heat flux solved by SFINCS code obeys the
isomorphism, but is radial electric field ($d\Phi/dr$) contained in ``other input quantities''? Please
clarity that. Also, if the isomorphism is satisfied for finite $d\Phi/dr$ and finite source term, I
suppose that the usage of source term in the full trajectory model is justified at least in
quasisymmetric systems. Can author comment on this point in Appendix?
\end{referee}

Response goes here.

\vspace{0.3in}

In addition to the aforementioned changes, the following modifications have been made to the manuscript:

\begin{itemize}
\item Albert Moll\'{e}n has recently demonstrated precise agreement between SFINCS and an anlytical calculation
of Simakov and Helander for the limit of high collisionality [43]. His results are displayed in a new figure 6, with details given in Appendix B,
 and he has been included as a co-author.
\item Output from the DKES code is now plotted in figure 5, to demonstrate the good agreement with SFINCS.
\end{itemize}

Sincerely yours,\newline

M. Landreman

H. M. Smith

A. Moll\'{e}n

P. Helander

\end{document}

### 2014-02-06 doc/sfincsPaper/sfincsPaper.tex
Added Er=0 values to figures 1-2
Removed Albert from acknowledgements since he is now a co-author

@@
 \begin{figure}[h!]
-% Figure generated using m20131202_01_plotSFINCSErComparisonForPaper.m
+% Figure generated using m20140205_03_plotSFINCSFortranErComparisonForPaper.m
+%******************************
+% Previous version of figure generated using m20131202_01_plotSFINCSErComparisonForPaper.m
 % Data files:
@@
 % *********************************
-\includegraphics{m20131202_01_plotSFINCSErComparisonForPaper_LHD.eps}
+\includegraphics{m20140205_03_plotSFINCSFortranErComparisonForPaper_LHD.eps}
+%\includegraphics{m20131202_01_plotSFINCSErComparisonForPaper_LHD.eps}
 %\includegraphics{m20131202_01_plotSFINCSErComparisonForPaper_LHD.pdf}
@@
 of the normalized radial electric field (\ref{eq:EStar}).
+\changed{Results for $E_r=0$ are inicated by the $\blacktriangleright$ symbol to the left of each plot.}
 \label{fig:ErComparison_LHD}}
@@
 \begin{figure}[h!]
-% Figure generated using m20131205_05_plotSFINCSFortranErComparisonForPaper.m
+% Figure generated using m20140205_03_plotSFINCSFortranErComparisonForPaper.m
+%******************************
+% Previous version of figure generated using m20131205_05_plotSFINCSFortranErComparisonForPaper.m
 % SFINCS run on hopper in 
@@
 % *********************************
-\includegraphics{m20131205_05_plotSFINCSFortranErComparisonForPaper_W7X.eps}
+\includegraphics{m20140205_03_plotSFINCSFortranErComparisonForPaper_W7X.eps}
+%\includegraphics{m20131205_05_plotSFINCSFortranErComparisonForPaper_W7X.eps}
 \caption{(Color online) Comparison of trajectory models for W7-X standard geometry at $\nu' = 0.01$,
@@
 of the normalized radial electric field (\ref{eq:EStar}).
+\changed{Results for $E_r=0$ are inicated by the $\blacktriangleright$ symbol to the left of each plot.}
 \label{fig:ErComparison_W7X}}
@@
 We are thankful to J. Geiger for providing the W7-X equilibrium data,
-and to Craig Beidler, Albert Moll\'{e}n, and Peter Catto for other input on this work.
+and to Craig Beidler and Peter Catto for other input on this work.
 \end{acknowledgments}

### 2014-02-10 doc/sfincsPaper/refereeResponse/sfincsPaper_refereeResponse.tex
Added Er=0 values to figures 1-2
Added features to matlab 1-species version for testing quasisymmetry isomorphism and for non-stellarator-symmetric geometryScheme=1. Also expanded Appendix A of paper.

@@
 
-Response goes here.
+The values for $E_r=0$ are now plotted on figures (1)-(2), as $\blacktriangleright$ symbols on the left axis
+of each sub-figure.
 
@@
 
-Response goes here.
-
+We have changed and expanded the text in this appendix for clarification. 
+For all of the trajectory models described, an isomorphism does hold for the transport matrix elements
+if the electric field is varied so
+$(GM+IN)(\iota M-N)^{-1} d\Phi_0/d\psi$ remains constant as $M$ and $N$ are varied.
+%This required transformation of $d\Phi_0/d\psi$ can be found in the last line of table 1 in Ref. [49].
 \vspace{0.3in}

### 2014-02-10 doc/sfincsPaper/sfincsPaper.tex
Added features to matlab 1-species version for testing quasisymmetry isomorphism and for non-stellarator-symmetric geometryScheme=1. Also expanded Appendix A of paper.

@@
 of the normalized radial electric field (\ref{eq:EStar}).
-\changed{Results for $E_r=0$ are inicated by the $\blacktriangleright$ symbol to the left of each plot.}
+\changed{Results for $E_r=0$ are indicated by the $\blacktriangleright$ symbol to the left of each plot.}
 \label{fig:ErComparison_LHD}}
@@
 of the normalized radial electric field (\ref{eq:EStar}).
-\changed{Results for $E_r=0$ are inicated by the $\blacktriangleright$ symbol to the left of each plot.}
+\changed{Results for $E_r=0$ are indicated by the $\blacktriangleright$ symbol to the left of each plot.}
 \label{fig:ErComparison_W7X}}
@@
 discussed analytically in Refs. \cite{Pytte, Boozer83, meQS}.
-In this test, the magnetic field is specified as $B(\theta,\zeta) = y(M\theta-N\zeta)$
-for some given periodic function $y$ and integers $M$ and $N$.
-The kinetic equation is solved for several values of $M$ and $N$,
-varying the collision frequency with each run so that
-$\nuii / \left|M - N/\iota \right|$ remains fixed.
-As $M$ and $N$ are varied, $G$, $I$, $\iota$, $dn/d\psi$, $dT/d\psi$, and other input quantities are also kept fixed.
-Under these circumstances, it is known analytically \cite{Pytte, Boozer83, meQS} that the flow
+\changed{
+A quasisymmetric magnetic field is one satisfying $B(\theta,\zeta) = y(M\theta-N\zeta)$
+for some periodic function $y$ and integers $M$ and $N$.
+Magnetic fields with the same $y$ but different $M$ and $N$ are said to be isomorphic in that
+the associated transport matrices must be related in the following manner.
+Suppose the transport matrices are computed for several quasisymmetric magnetic fields with different values of $M$ and $N$,
+varying the collision frequency in each calculation so
+$\nuii / (\iota M - N)$ remains fixed,
+and varying the radial electric field so
+$(d\Phi_0/d\psi)(GM+IN)/(\iota M-N)$
+remains fixed.
+%Other input parameters such as $dn/d\psi$ and $dT/d\psi$ are kept fixed
+%as $M$ and $N$ are varied.
+%Under these circumstances, 
+In such a scan of $M$ and $N$,
+it can be shown analytically \cite{Pytte, Boozer83, meQS} that the 
+transport matrix elements should vary as follows:
+$L_{11}$, $L_{12}$, $L_{21}$, and $L_{22}$ vary
+$\propto (NI+MG)^2/(\iota M-N)$;
+$L_{13}$, $L_{23}$, $L_{31}$, and $L_{32}$ vary
+$\propto (NI+MG)/(\iota M-N)$;
+and $L_{33}$ varies
+$\propto 1/(\iota M-N)$.
+This isomorphism holds for all the trajectory models considered in this paper.
+}
+\begin{comment}
+flow
 and heat flux should obey the transformations
@@
 respectively, where the quantities $C_V$ and $C_q$ should be independent of $M$ and $N$.
-It was verified that the SFINCS code obeyed these isomorphism transformations
-for a range of $y$, collisionality regimes, and collision operators.
+\end{comment}
+
+\changed{
+As $\nuii / (\iota M - N)$ is to be held fixed in this test, while $\iota M - N$ can change sign as $M$ and $N$ are varied, 
+the collision frequency to use can be negative.
+While $\nuii <0$ does not make sense physically, it poses no mathematical or numerical problem.
+If the signs of the collision frequency, $\xi$, and $d\Phi_0/d\psi$ are simultaneously
+reversed in the kinetic equation, the part of $f_{i1}$ driven by $\left< E_{||}B\right>$ remains unchanged,
+while the part driven by radial gradients changes sign. Thus, $L_{11}$, $L_{12}$, $L_{21}$, $L_{22}$, and $L_{33}$ change sign, while
+$L_{13}$, $L_{23}$, $L_{31}$, and $L_{32}$ remain unchanged.
+%As discussed previously, reversing the sign of $d\Phi_0/d\psi$ leaves the transport matrix unchanged in a stellarator-symmetric field.
+Therefore, another way to express the quasisymmetry isomorphism 
+(even for non-stellarator-symmetric $y$)
+that preserves $\nuii > 0$ is the following:
+%in a stellarator-symmetric $y$
+if $M$ and $N$ are varied holding 
+$\nuii / \left|\iota M - N\right|$ and $(d\Phi_0/d\psi)(GM+IN)/|\iota M-N|$ fixed,
+$L_{11}$, $L_{12}$, $L_{21}$, and $L_{22}$ should vary
+$\propto (NI+MG)^2/|\iota M-N|$;
+$L_{13}$, $L_{23}$, $L_{31}$, and $L_{32}$ should vary
+$\propto (NI+MG)/(\iota M-N)$;
+and $L_{33}$ should vary
+$\propto 1/|\iota M-N|$.
+}
+
+It was verified that the SFINCS code obeyed both versions of these isomorphism transformations
+for various $y$, collisionality regimes, \changed{radial electric fields, trajectory models}, and collision operators.
 

### 2014-02-10 matlab/singleSpecies/sfincs.m

@@ -51,6 +51,10 @@ RHSMode = 1;
 %testQuasisymmetryIsomorphism = true;
 testQuasisymmetryIsomorphism = false;
 
+% The value of the variable below only matters in rare testing circumstances.
+preservePositiveNuInQuasisymmetryIsomorphism = true;
+%preservePositiveNuInQuasisymmetryIsomorphism = false;
+
 %saveStuff = true;
 saveStuff = false;
 
@@ -62,8 +66,8 @@ filenameNote = 'myFirstScan';
 % Geometry parameters:
 % --------------------------------------------------
 
-geometryScheme = 12;
-% 1 = Two-helicity model
+geometryScheme = 1;
+% 1 = Three-helicity model
 % 2 = Three-helicity approximation of the LHD standard configuration
 % 3 = Four-helicity approximation of the LHD inward-shifted configuration
 % 4 = Three-helicity approximation of the W7-X Standard configuration
@@ -73,28 +77,45 @@ geometryScheme = 12;
 
 % Additional parameters used only when geometryScheme=1:
 % B = BBar * B0OverBBar * [1 + epsilon_t * cos(theta) + epsilon_h * cos(helicity_l * theta - helicity_n * zeta)]
+%                            + epsilon_antisymm * sin(helicity_antisymm_l * theta - helicity_antisymm_n * zeta)]
 B0OverBBar = 1;
-epsilon_t = 0.13;
-epsilon_h = 0.0;
-helicity_l = 2;
+
+epsilon_t = 0.1;
+
+epsilon_h = 0.1;
+helicity_l = 1;
 helicity_n = 5;
+
+epsilon_antisymm = 0;
+% Note: when testQuasisymmetryIsomorphism=true, helicity_antisymm_l and helicity_antisymm_n will be over-written.
+helicity_antisymm_l = 0;
+helicity_antisymm_n = 0;
+
 % iota is the rotational transform = 1 / (safety factor q)
 iota = 0.4542;
+
 % G is c/2 * the poloidal current outside the flux
 % surface. Equivalently, G is the coefficient of grad zeta in the
 % covariant representation of vector B. GHat is G normalized by \bar{B}\bar{R}.
 GHat =  3.7481;
+
 % I is c/2 * the toroidal current inside the flux
 % surface. Equivalently, I is the coefficient of grad theta in the
 % covariant representation of vector B. IHat is I normalized by \bar{B}\bar{R}.
-IHat = 0.0;
+IHat = 0;
 
-% geometryScheme=10 parameters
[…]

### 2014-02-12 doc/sfincsPaper/refereeResponse/sfincsPaper_refereeResponse.tex
Progress on response to referee

@@
 \newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
+\newcommand{\todo}[1]{{\color{red}#1}}
 \newenvironment{referee}{\begin{quote}\it\color{Blue}}{\end{quote}}
@@
 If it is not difficult numerically to retain the poloidal and toroidal magnetic drifts in the
-equations of motion, as it is written in the middle of page 13, why do authors keep these term
-and remove only the component from eqs.(12) and (14)? I understand that it is to make
+equations of motion, as it is written in the middle of page 13, why do authors [not] keep these term
+and remove only the $v_{ma}\cdot\nabla\psi$ component from eqs.(12) and (14)? I understand that it is to make
 the ``DKES trajectories'', but it seems that approximating in (12) and (14) still
@@
 
-Response goes here.
+We do not expect any fundamental complications to arise if the poloidal and toroidal magnetic drifts
+were retained in the kinetic equation. 
+However, we have several reasons for neglecting these magnetic drifts
+ in this manuscript.
+\begin{enumerate}
+\item The ratio of these magnetic drifts to the parallel streaming term we keep is $\sim \rho_* = \rho / L$
+ (the gyroradius to scale length), so neglecting the poloidal and toroidal magnetic drifts is
+equivalent to considering the $\rho_* \to 0$ limit.
+Since our primary interest in this paper is to compare the various options for the electric field
+terms and collision operator, it is reasonable to take this $\rho_* \to 0$ limit for simplicity.
+\item
+Retaining these drifts would make the transport matrix (and figures 1, 2, 4, and 4)
+depend on $\rho_*$ in addition to $\nu'$ and $E_*$. 
+The dependence of the matrix elements on $\nu'$ and $E_*$ is already  complicated
+and interesting to discuss in the $\rho_* \to 0$ limit,
+without the extra complexity of finite-$\rho_*$ effects.
+%\item Including these magnetic drift terms would also introduce dependence on the radial
+%derivatives of the magnetic field
+\item The poloidal and toroidal magnetic drifts would add extra complicating
+terms to the analysis of moments of the kinetic equation in section III.
+%\item If the poloidal and toroidal magnetic drifts are retained without the radial magnetic drift, ????
+%\item The poloidal and magnetic drift terms will take additional time to implement in the code, 
+%and the 
+\end{enumerate}
+
+We have added some text to section II (between (14)-(15)) and a sentence to section IV (between \todo{(26)-(27)}) to clarify these issues.
 
@@
 
-Response goes here.
+\todo{Response goes here.}
 
@@
 
-Response goes here.
+Probably this issue was unclear because the sources actually serve two independent purposes,
+not only to circumvent the conservation problems when $E_r$ is nonzero, but also to eliminate the null
+space of the kinetic equation which is a problem even at $E_r=0$.
+
+First, consider the case $E_r=0$, so there is no difference between the trajectory models.
+Consider the homogeneous solutions of the kinetic equation, i.e. the null space of the linear system
+we solve if no sources or constraints are included. 
+When the linearized Fokker-Planck operator is used, 
+there are 2 linearly independent homogeneous solutions:
+a Maxwellian flux function,
+and a Maxwellian flux function times $v^2$.
+However, if the pure pitch-angle scattering operator or momentum-conserving model
+collision operator are used, there are $N_x$ independent homogeneous solutions,
+since any function of $v$ (independent of $\theta$,$\zeta$, and $\xi$)
+is in the null space.
+
+\todo{Now consider the case of nonzero $E_r$...}
 
@@
 
-The values for $E_r=0$ are now plotted on figures (1)-(2), as $\blacktriangleright$ symbols on the left axis
+Computations for $E_r=0$ are now plotted on figures (1)-(2), as $\blacktriangleright$ symbols on the left axis
 of each sub-figure.
@@
 
-Response goes here.
+We agree that in scenarios with $T_e \gg T_i$, such as a Core Electron Root Confinement regime,
+$E_*$ may be larger than a few percent. (Our original statement
+was meant only to apply to a standard ion-root W7-X regime,
+but this should have been clarified.) The text in the manuscript has now been updated to acknowledge
+the possibility of an electron root and that $E_*$ may not be small.
 
@@
 %This required transformation of $d\Phi_0/d\psi$ can be found in the last line of table 1 in Ref. [49].
+
+The second part of your question is an interesting observation: when the full trajectory model
+is applied to a quasisymmetric field, the source vanishes for any $E_r$, not just the ambipolar one (as in a non-quasisymmetric field.)
+We now highlight this point in Appendix A and in \todo{the new discussion of intrinsic ambipolarity in section III.}
+
 \vspace{0.3in}

### 2014-02-13 doc/sfincsPaper/sfincsPaper.tex

@@
 Third, dropping these terms conveniently eliminates all dependence of the transport matrix (defined in section \ref{sec:transportMatrix})
-on $d T_a/d\psi$ and $dB/d\psi$.
-Fourth, we wish to focus on the effects of the radial electric field.
+on $d T_a/d\psi$, $dB/d\psi$,
+\changed{
+and $\rho_*$.
+Fourth, dropping these terms amounts to taking the limit $\rho_* \to 0$
+(while keeping $d\Phi_0/d\psi$ finite), 
+and this limit is already complicated and interesting to explore without the extra complexity
+of finite-$\rho_*$ corrections.
+Fifth, 
+we wish to focus on the effects of the radial electric field.
 The omitted terms may be important in other situations, but here our primary interest is the treatment of the $d\Phi_0/d\psi$ terms.
+Finally,
+these omitted terms would significantly complicate the analysis in section III,
+in which we will examine moments of the kinetic equation.
+}
+
 Our kinetic equation then becomes
@@
 without increasing the density of the matrix, i.e. without increasing the computational expense
-of the method here.  However, to include radial drifts acting on $f_{a1}$,
+of the method here.  
+\changed{We do not expect any fundamental new complications to arise
+if poloidal and toroidal magnetic drift terms are retained.}
+However, to include radial drifts acting on $f_{a1}$,
 the number of independent variables would increase from 4 to 5 since different flux surfaces
@@
 
-In most plasma scenarios, $E_*$ can be estimated by noting that the
+In \changed{the typical ``ion root'' scenario}, $E_*$ can be estimated by noting that the
 ambipolar electric field arises to bring the ion particle transport
 down to the electron level, and is therefore approximately determined so 
-as to reduce the magnitude of the ``thermodynamic
-force'' appearing as the first component of the vector on the
+as to reduce the magnitude of the thermodynamic
+force appearing as the first component of the vector on the
 right-hand side of (\ref{eq:matrix}). The electric field is thus of
@@
 is of order $E_* \sim \rho_\theta / L_\perp$, where $\rho_{\theta} =
-\rho / (\iota \epsilon)$, with $\epsilon$ the inverse aspect ratio. In
-W7-X, $E_r$ is predicted to be particularly large in regions with
-steep density gradients and will in normal plasma scenarios stay below
-a few tens of kV/m, corresponding to $E_*$ up to a few tenths\cite{Turkin}.
-In most of the plasma, $E_*$ is however expected to be smaller than a few percent.
+\rho / (\iota \epsilon)$ and $\epsilon$ is the inverse aspect ratio,
+\changed{
+and the ratio $\rho_\theta / L_\perp$ is typically $\ll 1$.
+In W7-X, $E_*$ is predicted to be a few percent in normal plasma scenarios \cite{Turkin}.
+The largest $E_r$ in normal W7-X scenarios is predicted to be a few tens of kV/m, in the
+edge where density gradients are steep,
+corresponding to $E_*$ up to a few tenths\cite{Turkin}.
+However, in other previous experiments, scenarios with strong electron heating
+can cause $T_e \gg T_i$,
+giving rise to large positive ``electron root'' electric fields \cite{YokoyamaCERC}.
+In these scenarios, $E_*$ may be $\sim 1$. 
+}
 
@@
 While $\nuii <0$ does not make sense physically, it poses no mathematical or numerical problem.
-If the signs of the collision frequency, $\xi$, and $d\Phi_0/d\psi$ are simultaneously
+In any stellarator (even a non-quasisymmetric and/or non-stellarator-symmetric one),
+if the signs of the collision frequency, $\xi$, and $d\Phi_0/d\psi$ are simultaneously
 reversed in the kinetic equation, the part of $f_{i1}$ driven by $\left< E_{||}B\right>$ remains unchanged,

### 2014-02-18 doc/sfincsPaper/refereeResponse/sfincsPaper_refereeResponse.tex
Added discussion of momentum balance to revised manuscript and referee response

@@
 
-\todo{Response goes here.}
+We have added a new section IV to the paper to discuss momentum balance.
+To summarize, we first discuss how parallel momentum has a different status to mass and energy in that mass and
+energy are conserved, whereas parallel momentum is not conserved due to the mirror force.
+We then compare the $m v_{||}$ moments of the various forms of the drift-kinetic
+equation to the full fluid momentum equation, as you suggest.
+We show the electric field terms in the kinetic perspective
+are associated with a piece of the gyroviscosity in the fluid perspective.  The full trajectory model
+gives the correct viscous force whereas the other trajectory models do not.
+One further noteworthy property of the parallel velocity moments of the kinetic equations
+is that of the trajectory models considered, only the full trajectories preserve intrinsic
+ambipolarity in quasisymmetry when $E_r \ne 0$. This result is now discussed at the end of section IV.
 
@@
 The second part of your question is an interesting observation: when the full trajectory model
-is applied to a quasisymmetric field, the source vanishes for any $E_r$, not just the ambipolar one (as in a non-quasisymmetric field.)
-We now highlight this point in Appendix A and in \todo{the new discussion of intrinsic ambipolarity in section III.}
+is applied to a quasisymmetric field, the source vanishes for any $E_r$, not just at a special ambipolar $E_r$ (as in a non-quasisymmetric field.)
+We now state this fact in the new discussion of intrinsic ambipolarity at the end of section IV.
 

### 2014-02-18 doc/sfincsPaper/sfincsPaper.tex
Added discussion of momentum balance to revised manuscript and referee response

@@
 \newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
+\newcommand{\tens}[1]{\overleftrightarrow{#1}}
 \newcommand{\erf}{\mathrm{erf}}
@@
 These issues are explored in section \ref{sec:sources}.
+\changed{In section \ref{sec:momentum}, we discuss some observations regarding momentum conservation,
+and demonstrate that the electric field terms in the kinetic description
+correspond to a component of gyroviscosity in a fluid description.}
 Details of the numerical implementation are given in section \ref{sec:numerics}.
@@
 Fourth, dropping these terms amounts to taking the limit $\rho_* \to 0$
-(while keeping $d\Phi_0/d\psi$ finite), 
+(while keeping the $d\Phi_0/d\psi$ terms finite), 
 and this limit is already complicated and interesting to explore without the extra complexity
@@
 Finally,
-these omitted terms would significantly complicate the analysis in section III,
+these omitted terms would significantly complicate the analysis in section \ref{sec:sources},
 in which we will examine moments of the kinetic equation.
@@
 
-\section{Conservation properties and sources}
+\section{Particle and energy moment equations, conservation properties, and sources}
 \label{sec:sources}
@@
 We will then describe a modified form of the kinetic equation which robustly produces more sensible results.
+\changed{The issues discussed in this section are related to moment equations for mass and energy; momentum has a different status
+and will be examined in the next section.}
 
@@
 at each grid point in $v$.
+
+\section{Momentum moment equations}
+\label{sec:momentum}
+
+\changed{
+Parallel momentum has a different status to density and energy, in that density and energy are conserved by
+the collisionless motion while parallel momentum is not, due to the mirror force.
+(For example, considering the simple case of a single ion species with $d\Phi_0/d\psi=0$, $F_i$ and $v^2 F_i$ are homogeneous
+solutions to the kinetic equation, whereas $v_{||} F_i$ is not.)
+A consequence is that there does not appear to be a false constraint for $E_r \ne 0$ arising
+from the $\left< \int d^3 v\, v_{||} (\ldots)\right>$ moment of the various forms of the kinetic equation,
+i.e. there is no exact analogue to (\ref{eq:unphysicalMassConservation}), (\ref{eq:unphysicalEnergyConservation1}),
+or (\ref{eq:unphysicalEnergyConservation2}) for momentum.
+When the momentum moment of the various forms of the drift-kinetic equation is taken,
+even if a factor of $B$ or $1/B$ is included in the flux surface average, 
+a collisionless term remains that is not proportional to $d\Phi_0/d\psi$.
+Consequently, for all the trajectory models,
+$d\Phi_0/d\psi=0$ is a well behaved rather than singular
+limit of the momentum moment equation.
+}
+
+\changed{
+Nonetheless, it is interesting to compare the $\int d^3v\, m_a v_{||} (\ldots)$ moment equations
+for each trajectory model to the full parallel momentum fluid equation. This later equation is
+\begin{equation}
+???
+\end{equation}
+First, consider the case of no radial electric field.
+The stress tensor is given to the accuracy needed by $\tens{\Pi}_a \approx \tens{\Pi}_{a||}$
+where $\tens{\Pi}_{a||} = p_{a \bot} \tens{I} + (p_{a||} - p_{a\bot}) \vect{b}\vect{b}$,
+$p_{a||} = m_a \int d^3v\, f_{a1} v_{||}^2$, and $p_{a\bot} = m_a \int d^3v\, f_{a1} v_{\bot}^2/2$.
+Then noting the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the streaming and mirror terms is
+\begin{eqnarray}
+\int d^3v\, m_a v_{||} \left[ v_{||} \vect{b}\cdot\nabla f_{a1} - \frac{1-\xi^2}{2B}v(\vect{b}\cdot\nabla B) \frac{\partial f_{a1}}{\partial \xi} \right]
+& = &
+\vect{b}\cdot\nabla p_{a||1} + \frac{ p_{a\bot 1} - p_{a || 1}}{B} \vect{b}\cdot\nabla B \nonumber \\
+& = &
+ -\vect{b} \cdot \left(\nabla \cdot \tens{\Pi}_{a||} \right),
+\end{eqnarray}
+one can see the drift-kinetic equation (\ref{}) is consistent with the full fluid parallel momentum equation
+at least when $E_r=0$.
+}
+
+\changed{
+Now consider how the situation changes when a radial electric field is introduced. We first
+compute the change to the fluid parallel momentum equation caused by a new contribution
+to the viscosity.
+Examining the $\left( \vect{E} + c^{-1} \vect{v}\times\vect{B}\right) \cdot \nabla_{\vect{v}} f_a$ terms in 
+the full Fokker-Planck equation and integrating in gyrophase, one sees the gyrophase-dependent part of the 
+distribution function $\tilde{f}_a$ will include the following terms proportional to the electric field:
+\begin{equation}
+\tilde{f}_{aE} = -\frac{c}{B} \vect{v}\cdot\vect{b}\times\nabla\Phi \left[ \frac{1}{v} \frac{\partial \bar{f}_a}{\partial v}
+-\frac{\xi}{v^2} \frac{\partial \bar{f}_a}{\partial \xi} \right]
+\label{eq:fTilde}
+\end{equation}
+as reflected (using different indepedent variables) in eq (17) of Ref. \cite{Hazeltine} and eq (6) of Ref. \cite{SimakovCattoSecondOrderDKE}.
+Here, $\bar{f}_a$ is the gyrophase-independent part of the distribution function.
+The associated contribution $\tens{\Pi}_{aE} = m_a \int d^3v\, \tilde{f}_{aE}\vect{v}\vect{v}$
+to the pressure tensor is calculated in equations (27)-(36) of Ref. \cite{SimakovCattoSecondOrderDKE},
+with the result
+\begin{equation}
+\tens{\Pi}_{aE} = m_a n_a V_{a||} \left( \vect{b}\vect{v}_E + \vect{v}_E \vect{b} \right)
+\label{eq:PiE}
+\end{equation}
+where $n_a V_{a||} = \int d^3v\, \bar{f}_a v_{||}$. Note that this contribution to the stress tensor is a part of the gyroviscosity and is off-diagonal.
+Using $\Phi \approx \Phi_0(\psi)$ and $\vect{B} \cdot \left\{ \nabla \cdot \left[ \vect{B}\vect{B}\times\nabla\psi + (\vect{B}\times\nabla\psi) \vect{B} \right] \right\}
+=2B^2\nabla\psi\cdot\nabla\times\vect{B}=0$,
+we then find the contribution to the parallel momentum equation from (\ref{eq:fTilde})-(\ref{eq:PiE}) is
+\begin{equation}
+\vect{b}\cdot(\nabla\cdot\tens{\Pi}_{aE}) = 
+c m_a B (d\Phi_0/d\psi)
+\vect{B}\times\nabla\psi\cdot\nabla\left(n_a V_{a||}/B^3\right).
+\label{eq:PiEParallel}
+\end{equation}
+}
+
+\changed{
+Finally, let us consider the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the radial electric field
+terms in the drift-kinetic equation for various trajectory models, to see if the results agree with (\ref{eq:PiEParallel}).
+For the full trajectories, the moment of the $d\Phi_0/d\psi$ terms in (\ref{eq:kinetic4})-(\ref{eq:fullTrajectories}) is
+\begin{eqnarray}
+\int d^3v\, m_a v_{||} \left[ 
+\frac{c}{B^2} \frac{d\Phi_0}{d\psi} \vect{B}\times\nabla\psi\cdot\nabla f_{a1}
+- (\vma \cdot\nabla\psi) \frac{Z_a e}{2 T_a x_a} \frac{d\Phi_0}{d\psi} \frac{\partial f_{a1}}{\partial x_a}
+\right.
+\nonumber \\
+\left.
++ \xi(1-\xi^2) \frac{c}{2B^3} \frac{d\Phi_0}{d\psi} (\vect{B}\times\nabla\psi\cdot\nabla B)\frac{\partial f_{a}}{\partial \xi}
+\right]
+\nonumber \\
+= 
+c m_a B (d\Phi_0/d\psi)
+\vect{B}\times\nabla\psi\cdot\nabla\left(n_a V_{a||}/B^3\right),
+\label{eq:fullViscosity}
+\end{eqnarray}
+obtained by integrating by parts in $x_a$ and $\xi$.
+Thus, the full-trajectory model agrees with the full fluid parallel momentum equation: (\ref{eq:fullViscosity}) = (\ref{eq:PiEParallel}).
+However, this agreement is
+not shared by the DKES model: the moment of the $d\Phi_0/d\psi$ term in (\ref{eq:DKESTrajectories}) is
+\begin{eqnarray}
+\int d^3v\, m_a v_{||} \left[ 
+\frac{c}{\left< B^2 \right> } \frac{d\Phi_0}{d\psi} \vect{B}\times\nabla\psi\cdot\nabla f_{a1}
+\right]
+= 
+\frac{c m_a}{ \left< B^2 \right>}
+\frac{d\Phi_0}{d\psi}
+\vect{B}\times\nabla\psi\cdot\nabla (n_a V_{a||}),
+\label{eq:DKESViscosity}
+\end{eqnarray}
+which does not equal (\ref{eq:PiEParallel}). The corresponding result for the partial trajectories,
+obtained by replacing $\left< B^2 \right> \to B^2 $ in (\ref{eq:DKESViscosity}), also does not match
+(\ref{eq:PiEParallel}). Thus, the DKES and partial trajectory models do not correctly account for the parallel
+viscous force as the full trajectory model does.
+}
+
+\changed{
+We close this section by noting another an important difference between the trajectory models
+related to the parallel momentum equations.
+Consider the case of \emph{quasisymmetry}, which is the condition \cite{PerAndreiPRL} that
+$\vect{B}\times\nabla\psi\cdot\nabla B = A(\psi) \vect{B}\cdot\nabla B$ for some flux function
+$A(\psi)$.  It was known previously \cite{PerAndreiPRL} that when the $E_r$ terms are not included
+in the trajectories (but retained in the $\partial F_a/\partial \psi$ drive term in (\ref{eq:kinetic4})), the radial neoclassical
+current vanishes for all values of $d \Phi_0/d\psi$
+if and only if the flux surface is quasisymmetric. This property of quasisymmetric flux surfaces
+is known as intrinsic ambipolarity.
+Here, we show that intrinsic ambipolarity persists in quasisymmetric geometry
+when the $E_r$ terms are retained in the full trajectory
+drift-kinetic equation, but not for the DKES or partial trajectory kinetic equations.
+This result follows from the $-\sum_a Z_a \left< \int d^3v\, A v_{||} / \Omega_a (\ldots)\right>$
+moment of the kinetic equations, i.e. a spatially weighted average of the parallel momentum
+moment. For the full trajectories, the $d\Phi_0/d\psi$ terms
+may be integrated by parts, canceling each other to leave
+\begin{equation}
+\sum_a Z_a \left< \int d^3v\, f_{a1} \vma\cdot\nabla\psi\right>=0,
+\label{eq:noCurrent}
+\end{equation}
+meaning the radial current vanishes. However, for the DKES and partial trajectory models, the $d\Phi_0/d\psi$ terms
+do not cancel, leaving an additional term in (\ref{eq:noCurrent}) proportional to $d\Phi_0/d\psi$,
+meaning the radial current is generally nonzero.  Notice that when the full trajectory model is applied in quasisymmetry,
+intrinsic ambipolarity means (\ref{eq:unphysicalEnergyConservation2}) is satisfied even when $d\Phi_0/d\psi \ne 0$,
+so the net heat source vanishes for any radial electric field.
+}
 
@@
 required $\sim 30-50$ minutes to run on 4 nodes of the Edison computer at NERSC.
-At higher collisionality, lower resolution is sufficient, so memory and time requirements are reduced;
-computations with $\nu' > 10^{-2}$ can typically be run on a laptop.
+At higher collisionality, or if fewer harmonics are retained in $B(\theta,\zeta)$, 
+lower resolution is sufficient, so memory and time requirements are reduced;
+\changed{
+for example, in the same W7-X geometry at $\nu'=10^{-2}$, sufficient resolution parameters for convergence
+were $N_\theta=11$, $N_\zeta=64$, $N_\xi=100$, and $N_x=5$,
+and computations required 3 minutes on 1 node of Edison.}
+Computations with $\nu' > 10^{-2}$ can typically be run on a laptop.
 
@@
 \changed{
-A quasisymmetric magnetic field is one satisfying $B(\theta,\zeta) = y(M\theta-N\zeta)$
+Equivalent to the definition at the end of section \ref{sec:sources} \cite{PerAndreiPRL},
+a quasisymmetric magnetic field is one satisfying $B(\theta,\zeta) = y(M\theta-N\zeta)$
 for some periodic function $y$ and integers $M$ and $N$.
@@
 which were solved in Ref.~\cite{AndreiPer2009} by keeping a small number of Laguerre polynomials in kinetic energy.
-When these generalized Spitzer problems are solved numerically keeping many more energy polynomials,
+When these generalized Spitzer problems are solved keeping more energy polynomials,
 we obtain the more accurate coefficients given in ${K}_{2}$ below.

### 2014-02-18 doc/sfincsPaper/sfincsPaper.tex
Finishing discussion of momentum balance in revised manuscript

@@
 As shown in Appendix C of Ref. \cite{usPedestal}, we may choose the gauge
-for the electromagnetic potentials such that $-c^{-1}\vect{b}\cdot\partial\vect{A}/\partial t = \left<E_{||} B\right> B/\left<B^2\right>$
+for the electromagnetic potentials such that 
+\begin{equation}
+-c^{-1}\vect{b}\cdot\partial\vect{A}/\partial t = \left<E_{||} B\right> B/\left<B^2\right>
+\label{eq:gauge}
+\end{equation}
 on the right-hand side of (\ref{eq:kinetic1}).
@@
 Nonetheless, it is interesting to compare the $\int d^3v\, m_a v_{||} (\ldots)$ moment equations
-for each trajectory model to the full parallel momentum fluid equation. This later equation is
-\begin{equation}
-???
-\end{equation}
+for each drift-kinetic trajectory model to the full parallel momentum fluid equation. This later equation,
+the moment of the full Fokker-Planck equation, is
+\begin{equation}
+0 = -\vect{b}\cdot\left( \nabla\cdot\tens{\Pi}_a\right)
++ Z_a e n_a E_{||} + F_{a||}
+\label{eq:fluidParallelMomentum}
+\end{equation}
+where $\tens{\Pi_a} = m_a \int d^3v\, f_a \vect{v}\vect{v}$ is the stress tensor
+and $F_{a||}$ is the parallel component of friction.
 First, consider the case of no radial electric field.
-The stress tensor is given to the accuracy needed by $\tens{\Pi}_a \approx \tens{\Pi}_{a||}$
-where $\tens{\Pi}_{a||} = p_{a \bot} \tens{I} + (p_{a||} - p_{a\bot}) \vect{b}\vect{b}$,
-$p_{a||} = m_a \int d^3v\, f_{a1} v_{||}^2$, and $p_{a\bot} = m_a \int d^3v\, f_{a1} v_{\bot}^2/2$.
-Then noting the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the streaming and mirror terms is
+Recalling
+$f_a = F_a(\psi)\left[ 1 - Z_a e \Phi_1/T_a\right] + f_{a1}$, 
+the stress tensor is given to the accuracy needed by 
+$\tens{\Pi}_a \approx p_a(\psi)\left[1 - Z_a e \Phi_1/T_a\right] \tens{I} + \tens{\Pi}_{a1}$
+where $\tens{\Pi}_{a1} = p_{a 1 \bot} \tens{I} + (p_{a1||} - p_{a1\bot}) \vect{b}\vect{b}$,
+$p_{a1||} = m_a \int d^3v\, f_{a1} v_{||}^2$, and $p_{a1\bot} = m_a \int d^3v\, f_{a1} v_{\bot}^2/2$.
+Notice the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the streaming and mirror terms in (\ref{eq:kinetic4})-(\ref{eq:fullTrajectories}) is
 \begin{eqnarray}
@@
 & = &
-\vect{b}\cdot\nabla p_{a||1} + \frac{ p_{a\bot 1} - p_{a || 1}}{B} \vect{b}\cdot\nabla B \nonumber \\
+\vect{b}\cdot\nabla p_{a1||} + \frac{ p_{a1\bot} - p_{a1 ||}}{B} \vect{b}\cdot\nabla B \nonumber \\
 & = &
- -\vect{b} \cdot \left(\nabla \cdot \tens{\Pi}_{a||} \right),
+\vect{b} \cdot \left(\nabla \cdot \tens{\Pi}_{a1} \right).
 \end{eqnarray}
-one can see the drift-kinetic equation (\ref{}) is consistent with the full fluid parallel momentum equation
+Using this result with (\ref{eq:gauge}), the $m_a v_{||}$ moment of 
+the drift-kinetic equation (\ref{eq:kinetic4})-(\ref{eq:fullTrajectories}) matches
+the full fluid parallel momentum equation (\ref{eq:fluidParallelMomentum})
 at least when $E_r=0$.
@@
 \begin{equation}
-\tilde{f}_{aE} = -\frac{c}{B} \vect{v}\cdot\vect{b}\times\nabla\Phi \left[ \frac{1}{v} \frac{\partial \bar{f}_a}{\partial v}
+\tilde{f}_{aE} = \frac{c}{B} \vect{v}\cdot\vect{b}\times\vect{E} \left[ \frac{1}{v} \frac{\partial \bar{f}_a}{\partial v}
 -\frac{\xi}{v^2} \frac{\partial \bar{f}_a}{\partial \xi} \right]
@@
 \end{equation}
-where $n_a V_{a||} = \int d^3v\, \bar{f}_a v_{||}$. Note that this contribution to the stress tensor is a part of the gyroviscosity and is off-diagonal.
-Using $\Phi \approx \Phi_0(\psi)$ and $\vect{B} \cdot \left\{ \nabla \cdot \left[ \vect{B}\vect{B}\times\nabla\psi + (\vect{B}\times\nabla\psi) \vect{B} \right] \right\}
+where $n_a V_{a||} = \int d^3v\, f_a v_{||}$. Note that this contribution to the stress tensor is a part of the gyroviscosity and is off-diagonal.
+Using $\vect{E} \approx -\nabla \Phi_0(\psi)$ and $\vect{B} \cdot \left\{ \nabla \cdot \left[ \vect{B}\vect{B}\times\nabla\psi + (\vect{B}\times\nabla\psi) \vect{B} \right] \right\}
 =2B^2\nabla\psi\cdot\nabla\times\vect{B}=0$,
@@
 \changed{
-Finally, let us consider the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the radial electric field
+For comparison, let us consider the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the radial electric field
 terms in the drift-kinetic equation for various trajectory models, to see if the results agree with (\ref{eq:PiEParallel}).
@@
 moment of the kinetic equations, i.e. a spatially weighted average of the parallel momentum
-moment. For the full trajectories, the $d\Phi_0/d\psi$ terms
-may be integrated by parts, canceling each other to leave
+moment. For the full trajectories, 
+(\ref{eq:fullViscosity}) vanishes in this spatial average, leaving
 \begin{equation}
@@
 \end{equation}
-meaning the radial current vanishes. However, for the DKES and partial trajectory models, the $d\Phi_0/d\psi$ terms
-do not cancel, leaving an additional term in (\ref{eq:noCurrent}) proportional to $d\Phi_0/d\psi$,
-meaning the radial current is generally nonzero.  Notice that when the full trajectory model is applied in quasisymmetry,
+meaning there is no radial current. However, for the DKES and partial trajectory models, the
+spatial average does not annihilate the $d\Phi_0/d\psi$ terms,
+leaving an additional term in (\ref{eq:noCurrent}) proportional to $d\Phi_0/d\psi$,
+and therefore the radial current is generally nonzero.  
+Consequently, the full trajectory model is the only one of the models that preserves
+intrinsic ambipolarity in quasisymmetry for $E_r \ne 0$.
+Notice that when the full trajectory model is applied in quasisymmetry,
 intrinsic ambipolarity means (\ref{eq:unphysicalEnergyConservation2}) is satisfied even when $d\Phi_0/d\psi \ne 0$,
@@
 - \nabla p_a - \nabla \cdot \vect{\pi}_a + \vect{F}_a, 
+\label{eq:vectorMomentum}
 \end{equation}
@@
 conductivity coefficient $L_{33}$. The flow that arises in response to
-a parallel electric field is determined by the parallel momentum equation
-\begin{equation} 
-n_a Z_a e E_{||} = \vect{b}\cdot \nabla p_a + \vect{b}\cdot(\nabla \cdot \vect{\pi}_a )+ F_{a||}, 
-\end{equation} 
+a parallel electric field is determined by the parallel momentum equation $\vect{b}\cdot$(\ref{eq:vectorMomentum}),
+%\begin{equation} 
+%n_a Z_a e E_{||} = \vect{b}\cdot \nabla p_a + \vect{b}\cdot(\nabla \cdot \vect{\pi}_a )+ F_{a||}, 
+%\end{equation} 
 where the parallel friction force $F_{\|a}$ again vanishes when only a

### 2014-02-20 doc/sfincsPaper/refereeResponse/sfincsPaper_refereeResponse.tex
Added sources to fig 3 in manuscript to address referee's question about source magnitude

@@
 
-We have added some text to section II (between (14)-(15)) and a sentence to section IV (between \todo{(26)-(27)}) to clarify these issues.
+We have added some text to section II (between (15)-(16)) and a sentence to section V (between \todo{(35)-(36)}) to clarify these issues.
 
@@
 
-\todo{Response goes here.}
+The sources normalized to the transport timescale are now plotted in figure 3.
+As discussed in the new text accompanying this figure, 
+the sources normalized by the ion collision time are only a factor $1/0.4=2.5$ larger than the values plotted,
+so the source terms remain much smaller than the collision operator in the kinetic equation.
 

### 2014-02-20 doc/sfincsPaper/sfincsPaper.tex
Added sources to fig 3 in manuscript to address referee's question about source magnitude

@@
 meaning there is no radial current. However, for the DKES and partial trajectory models, the
-spatial average does not annihilate the $d\Phi_0/d\psi$ terms,
+spatial average does not annihilate the $d\Phi_0/d\psi$ term,
 leaving an additional term in (\ref{eq:noCurrent}) proportional to $d\Phi_0/d\psi$,
@@
 \begin{figure}[h!]
-% Figure generated using m20131212_02_PlotSFINCS2SpeciesEScanForPaper.m
+% Figure generated using m20140219_02_PlotSFINCS2SpeciesEScanForPaper_withSources.m
+% ----------------------------------------
+% Previous version of figure generated using m20131212_02_PlotSFINCS2SpeciesEScanForPaper.m
 % SFINCS run on hopper and edison in 
 % /global/scratch2/sd/landrema/20131204-02-sfincs-fortran-hopper-intel
-\includegraphics{m20131212_02_PlotSFINCS2SpeciesEScanForPaper.eps}
+\includegraphics[height=7.5in]{m20140219_02_PlotSFINCS2SpeciesEScanForPaper_withSources.eps}
+%\includegraphics{m20131212_02_PlotSFINCS2SpeciesEScanForPaper.eps}
 \caption{(Color online)
-Fluxes and flows computed for a scenario of steep pressure gradient near the edge of W7-X.
+Fluxes (A)-(B), flow (C), and bootstrap current (D)
+computed for a scenario of steep pressure gradient near the edge of W7-X.
 Magenta dotted line is the ambipolar $E_r$, effectively identical for the three
 trajectory models.
+\changed{
+The ion particle and heat sources in (D)-(E) are normalized
+by a gyro-Bohm transport time $t_{gB}$.}
 \label{fig:2species}}
 \end{figure}
+
+\changed{
+Figures (\ref{fig:2species}).E-F illustrate the ion particle and heat sources computed as part of the calculation.
+As expected, the particle and heat sources are zero for the DKES model,
+and for the full trajectory model, 
+the particle source is always zero and the heat source vanishes at the ambipolar $E_r$.
+Electron sources are negligible.
+The plots show $S_{H\mathrm{p}}$ and $S_{H\mathrm{h}}$ from (\ref{eq:sources}) normalized to a gyro-Bohm transport time scale
+$t_{gB} = L^2/D_{gB}$ with $D_{gB} = (\rho_i / L) cT/(eB_0)$, thereby
+roughly normalizing the numerical sources to the scale of real physical sources arising from 
+the divergence of the turbulent and neoclassical fluxes.
+For this comparison we choose $L = -n / (dn/dr)$ to be the density scale length.
+For the range of electric fields considered, the numerical sources are small on this transport time scale,
+giving confidence in the model.
+For the parameters considered, $\nuii t_{gB} = 0.4$, 
+so dividing the values in figures (\ref{fig:2species}).E-F by this factor,
+the source terms evidently remain much smaller
+than the collision term in the kinetic equation for this calculation.
+}
 

### 2014-02-21 doc/sfincsPaper/refereeResponse/sfincsPaper_refereeResponse.tex
Finished draft of referee response and manuscript corrections

@@
 
-Probably this issue was unclear because the sources actually serve two independent purposes,
+We have expanded the discussion at the end of section III to clarify this issue.
+
+The system of sources and extra constraint equations actually serves two independent purposes,
 not only to circumvent the conservation problems when $E_r$ is nonzero, but also to eliminate the null
-space of the kinetic equation which is a problem even at $E_r=0$.
-
-First, consider the case $E_r=0$, so there is no difference between the trajectory models.
-Consider the homogeneous solutions of the kinetic equation, i.e. the null space of the linear system
-we solve if no sources or constraints are included. 
-When the linearized Fokker-Planck operator is used, 
-there are 2 linearly independent homogeneous solutions:
-a Maxwellian flux function,
-and a Maxwellian flux function times $v^2$.
-However, if the pure pitch-angle scattering operator or momentum-conserving model
-collision operator are used, there are $N_x$ independent homogeneous solutions,
-since any function of $v$ (independent of $\theta$,$\zeta$, and $\xi$)
-is in the null space.
-
-\todo{Now consider the case of nonzero $E_r$...}
+space of the kinetic equation. This latter problem arises even for the DKES trajectories 
+and even at $E_r=0$.  
+When $E_r \ne 0$, the sources are needed to eliminate the conservation problems, and the extra constraints 
+can be thought of as a convenient way to keep the linear system square (number of equations = number of unknowns) upon discretizatiton.
+Even when $E_r = 0$, and even for the DKES trajectories in which sources are not required, the constraints are needed to eliminate
+the null space in the kinetic equation, and the source terms can be thought of as
+a convenient way to keep the linear system square upon discretization.
+Any particle and heat source will be sufficient to solve the first problem (conservation), but to solve the second problem (null space),
+the number of constraints should match the dimensionality of the null space.  This dimension is set by the collision operator (2 for
+Fokker-Planck, $N_x$ for the other two operators.)
+This is why the source/constraint scheme is chosen based on the collision operator.
 

### 2014-02-21 doc/sfincsPaper/sfincsPaper.tex
Finished draft of referee response and manuscript corrections

@@
 
+\changed{
+Note that for both the DKES and full trajectories, the left-hand side of the
+kinetic equation (\ref{eq:kinetic4})
+can be written in the conservative form
+\begin{equation}
+\frac{1}{J}\left[
+\nabla\cdot\left( J \dot{\vect{r}}_a f_{a1} \right) 
++ \frac{\partial}{\partial \xi}\left( J \dot{\xi}_a f_{a1} \right)
++ \frac{\partial}{\partial x_a} \left( J \dot{x}_a f_{a1} \right)
+\right] - C_a
+\label{eq:conservativeForm}
+\end{equation}
+where $J=x_a^2$ is the Jacobian of the transformation between Cartesian velocity coordinates
+and the coordinates $x_a$, $\xi$, and gyrophase. However, for the partial trajectories,
+the left-hand side of (\ref{eq:kinetic4}) is not equivalent
+to (\ref{eq:conservativeForm}).
+}
+
 For all three trajectory models, the quasineutrality equation is effectively decoupled from the kinetic equation (\ref{eq:kinetic4}).
@@
 
+\changed{
+This system of sources and constraints solves not only the problem described above when $E_r \ne 0$,
+but also a different problem that remains even when $E_r=0$ and/or when the DKES trajectories are used:
+the kinetic equation has a null space.
+If the conditions $\left< \int d^3v \,f_{a1}\right>=0$ and $\left< \int d^3v\, v^2 f_{a1}\right>=0$
+were not imposed, any linear combination of $F_{a}$ and $F_{a} v^2$ could be added to one solution
+of the kinetic equation to obtain another solution. Upon discretization, one would obtain a non-invertible
+(or at least very poorly conditioned) linear system, but the imposition of these two extra constraints
+makes the system of equations invertible. 
+}
+
+\changed{
+Such is the case when the full linearized Fokker-Planck 
+collision operator is used, but the situation is different when either the pitch-angle scattering
+operator or momentum-conserving model operator are used instead, for then the kinetic equation has a larger null space:
+any function of $v$ is then a homogeneous solution of the kinetic equation.
+As the dimension of the null space is then equal to $N_x$ (the number of grid points in $x_a$) rather than
+2, it takes $N_x$ rather than 2 constraint equations to eliminate the null space for these collision operators.
+We choose these $N_x$ constraints to be $\left< \int_{-1}^1 d\xi\, f_{a1}\right>=0$ (imposed at each grid point in $x_a$.)
+To keep the linear system square, we must then have $N_x$ rather than 2 unknowns related to the sources. 
+This is accomplished by letting the source be a general function of $x_a$ instead
+of (\ref{eq:sources}) when either the pitch-angle scattering or momentum-conserving model collision operator are used.
+This alternative system of $N_x$ sources and constraints is an equally reasonable solution to the earlier
+conservation problem.
+}
+
+\changed{
+To summarize, the sources and extra constraint equations serve two independent purposes.
+First, when $E_r \ne 0$, the sources are needed to eliminate the conservation problems, and the extra constraints then keep the linear system square (number of equations = number of unknowns) upon discretizatiton.
+Second, even when $E_r = 0$, and even for the DKES trajectories in which sources are not required, the constraints are needed to eliminate the null space in the kinetic equation, and the source terms are a convenient way to keep the linear system square upon discretization.
+The first problem can be solved with either the source (\ref{eq:sources}) or a general speed-dependent source $S_a(x_a)$.
+However, to solve the second problem, the number of constraints should match the dimensionality of the null space.
+For this reason, we apply the source (\ref{eq:sources}) with 2 constraints when the Fokker-Planck operator is used,
+while we apply the general speed-dependent source $S_a(x_a)$ with $N_x$ constraints when either of the other two collision operators is used.
+}
+
+\begin{comment}
 The aforemention mass and energy sources and constraints work when the Fokker-Planck collision operator
@@
 at each grid point in $v$.
+\end{comment}
 

### 2014-02-21 matlab/multiSpecies/SFINCS.m

@@ -203,6 +203,33 @@ constraintScheme = -1;
 %  1 = 2 constraints per species: <n_1> = 0 and <p_1> = 0.
 %  2 = Nx constraints per species: <f_1>=0 at each x.
 
+% To use one of the 4 most common trajectory models, the remaining parameters
+% in this section should be set as follows:
+%
+% Full trajectories:
+%   includeXDotTerm = true
+%   includeElectricFieldTermInXiDot = true
+%   useDKESExBDrift = false
+%   include_fDivVE_term = false
+%
+% Partial trajectories: (non-conservative, as defined in the paper.)
+%   includeXDotTerm = false
+%   includeElectricFieldTermInXiDot = false
+%   useDKESExBDrift = false
+%   include_fDivVE_term = false
+%
+% Conservative partial trajectories: (Not discussed in the paper.)
+%   includeXDotTerm = false
+%   includeElectricFieldTermInXiDot = false
+%   useDKESExBDrift = false
+%   include_fDivVE_term = true
+%
+% DKES trajectories:
+%   includeXDotTerm = false
+%   includeElectricFieldTermInXiDot = false
+%   useDKESExBDrift = true
+%   include_fDivVE_term = false
+
 includeXDotTerm = true;
 %includeXDotTerm = false;
 

### 2014-02-21 matlab/singleSpecies/sfincs.m

@@ -228,6 +228,33 @@ constraintScheme = -1;
 %  1 = 2 constraints: <n_1> = 0 and <p_1> = 0.
 %  2 = Nx constraints: <f>=0 at each x.
 
+% To use one of the 4 most common trajectory models, the remaining parameters
+% in this section should be set as follows:
+%
+% Full trajectories:
+%   includeXDotTerm = true
+%   includeElectricFieldTermInXiDot = true
+%   useDKESExBDrift = false
+%   include_fDivVE_term = false
+%
+% Partial trajectories: (non-conservative, as defined in the paper.)
+%   includeXDotTerm = false
+%   includeElectricFieldTermInXiDot = false
+%   useDKESExBDrift = false
+%   include_fDivVE_term = false
+%
+% Conservative partial trajectories: (Not discussed in the paper.)
+%   includeXDotTerm = false
+%   includeElectricFieldTermInXiDot = false
+%   useDKESExBDrift = false
+%   include_fDivVE_term = true
+%
+% DKES trajectories:
+%   includeXDotTerm = false
+%   includeElectricFieldTermInXiDot = false
+%   useDKESExBDrift = true
+%   include_fDivVE_term = false
+
 includeXDotTerm = true;
 %includeXDotTerm = false;
 

### 2014-02-26 doc/sfincsPaper/sfincsPaper.tex
Made notation for tensors consistent in the manuscript.
Minor modifications to paper

@@
 \newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
-\newcommand{\tens}[1]{\overleftrightarrow{#1}}
+%\newcommand{\tens}[1]{\overleftrightarrow{#1}}
+%\newcommand{\tens}[1]{\overleftrightarrow{\vect{#1}}}
+\newcommand{\tens}[1]{\vect{#1}}
 \newcommand{\erf}{\mathrm{erf}}
@@
 \author{A Moll\'{e}n}
-\affiliation{Department of Applied Physics, Nuclear Engineering,
-  Chalmers University of Technology and Euratom-VR Association,
+\affiliation{Department of Applied Physics, %Nuclear Engineering,
+  Chalmers University of Technology, %and Euratom-VR Association,
   G\"oteborg,
@@
 =-(\vma\cdot\nabla\psi) \left(\frac{\partial F_{a}}{\partial \psi}\right)_{W_{a0}}
-+\frac{Z_a e}{T_a}\vpar\frac{B\left<E_{||} B\right>}{\left< B^2\right>} F_{a}.
++\frac{Z_a e}{T_a}\vpar\frac{B\left<E_{||} B\right>}{\left< B^2\right>} F_{a},
 \label{eq:kinetic3}
@@
 for several reasons.
-First, if the  $\vma$ term in (\ref{eq:rdot1}) were retained, we would need to solve a 5D rather
+First, if the  $\vma$ term in (\ref{eq:rdot1}) was retained, we would need to solve a 5D rather
 than 4D problem due to the radial coupling (i.e. $\psi$ appearing
@@
 =-(\vma\cdot\nabla\psi) \left(\frac{\partial F_{a}}{\partial \psi}\right)_{W_{a0}}
-+\frac{Z_a e}{T_a}\vpar\frac{B\left<E_{||} B\right>}{\left< B^2\right>} F_{a}.
++\frac{Z_a e}{T_a}\vpar\frac{B\left<E_{||} B\right>}{\left< B^2\right>} F_{a},
 \label{eq:kinetic4}
@@
 To summarize, the sources and extra constraint equations serve two independent purposes.
-First, when $E_r \ne 0$, the sources are needed to eliminate the conservation problems, and the extra constraints then keep the linear system square (number of equations = number of unknowns) upon discretizatiton.
+First, when $E_r \ne 0$, the sources are needed to eliminate the conservation problems, and the extra constraints then keep the linear system square (number of equations = number of unknowns) upon discretization.
 Second, even when $E_r = 0$, and even for the DKES trajectories in which sources are not required, the constraints are needed to eliminate the null space in the kinetic equation, and the source terms are a convenient way to keep the linear system square upon discretization.
@@
 \changed{
-We close this section by noting another an important difference between the trajectory models
+We close this section by noting another important difference between the trajectory models
 related to the parallel momentum equations.
-Consider the case of \emph{quasisymmetry}, which is the condition \cite{PerAndreiPRL} that
+Consider the case of \emph{quasisymmetry}, which is the condition %\cite{PerAndreiPRL} 
+that
 $\vect{B}\times\nabla\psi\cdot\nabla B = A(\psi) \vect{B}\cdot\nabla B$ for some flux function
-$A(\psi)$.  It was known previously \cite{PerAndreiPRL} that when the $E_r$ terms are not included
+$A(\psi)$ \cite{PerAndreiPRL}.  It was known previously \cite{PerAndreiPRL} that when the $E_r$ terms are not included
 in the trajectories (but retained in the $\partial F_a/\partial \psi$ drive term in (\ref{eq:kinetic4})), the radial neoclassical
@@
 \changed{
-The ion particle and heat sources in (D)-(E) are normalized
+The ion particle and heat sources in (E)-(F) are normalized
 by a gyro-Bohm transport time $t_{gB}$.}
@@
 m_a n_a \vect{V}_a \cdot \nabla \vect{V}_a = n_a e_a \left( - \nabla \phi + c^{-1}\vect{V}_a \times \vect{B} \right)
-- \nabla p_a - \nabla \cdot \vect{\pi}_a + \vect{F}_a, 
+- \nabla p_a - \nabla \cdot \tens{\pi}_a + \vect{F}_a, 
 \label{eq:vectorMomentum}
@@
   \left< n_a \vect{V}_a \cdot \nabla \psi \right> 
-  = \frac{1}{e_a p'(\psi)} \left< \vect{J} \cdot \left( \vect{F}_a - \nabla \cdot \vect{\pi}_a \right) \right>,
+  = \frac{1}{e_a p'(\psi)} \left< \vect{J} \cdot \left( \vect{F}_a - \nabla \cdot \tens{\pi}_a \right) \right>,
 \end{equation}
@@
 \begin{equation}
-  \left< n_a \vect{V}_a \cdot \nabla \psi \right>_{nc} = \frac{1}{e_ap'} \left< J_\| F_{a\|} + \vect{\pi}:\nabla \vect{J} \right>
+  \left< n_a \vect{V}_a \cdot \nabla \psi \right>_{nc} = \frac{1}{e_ap'} \left< J_\| F_{a\|} + \tens{\pi}_a:\nabla \vect{J} \right>
   \label{PS flux}
 \end{equation}
-where the viscosity tensor is $\vect{\pi}_a =(p_{\|} - p_{\perp})(\vect{b b} - {\bf I}/3)$.
+where the viscosity tensor is $\tens{\pi}_a =(p_{a\|} - p_{a\perp})(\vect{b b} - \tens{I}/3)$.
 The first term in (\ref{PS flux}) is proportional to $\nu$ and
@@
 %\begin{equation} 
-%n_a Z_a e E_{||} = \vect{b}\cdot \nabla p_a + \vect{b}\cdot(\nabla \cdot \vect{\pi}_a )+ F_{a||}, 
+%n_a Z_a e E_{||} = \vect{b}\cdot \nabla p_a + \vect{b}\cdot(\nabla \cdot \tens{\pi}_a )+ F_{a||}, 
 %\end{equation} 
@@
 \begin{equation} 
-n_a Z_a e \left< B  E_{||} \right> = \left< \vect{B}\cdot(\nabla \cdot \vect{\pi}_a) \right> 
+n_a Z_a e \left< B  E_{||} \right> = \left< \vect{B}\cdot(\nabla \cdot \tens{\pi}_a) \right> 
 = \left< (p_\perp - p_\| ) \nabla_\| B \right>.

### 2014-02-27 doc/sfincsPaper/sfincsPaper.tex
Implemented Hakan's suggestions for conclusion of manuscript

@@
 using continuum numerical methods).
-At first glance,  (\ref{eq:fullTrajectories}) and  (\ref{eq:partialTrajectories})
-appear more accurate than  (\ref{eq:DKESTrajectories}).
+Equations  (\ref{eq:fullTrajectories}) and  (\ref{eq:partialTrajectories})
+appear more accurate than  (\ref{eq:DKESTrajectories}),
+\changed{and as we have shown in section \ref{sec:momentum},
+the full trajectory model (\ref{eq:fullTrajectories}) is the only
+one of the three models which gives the correct parallel viscous force 
+and which preserves intrinsic ambipolarity in quasisymmetry.}
 However, as we have shown in section \ref{sec:sources},
@@
 is not well behaved when $E_r \ne 0$,
-with an unphysical constraint forced upon the distribution function.
+with one or two unphysical constraints forced upon the distribution function.
 This analytic property of the kinetic equation must be dealt with
@@
 This $E_r \sim \Erres$ regime can be relevant to experiments with high ratios $\Te / \Ti$
-\cite{Lore, Briesemeister} and strong gradients \cite{Baldzuhn}.
+\cite{YokoyamaCERC, Lore, Briesemeister} and strong gradients \cite{Baldzuhn}.
 In figure \ref{fig:2species}, we find that in the large-$E_r$ region anticipated for the edge of W7-X,

### 2014-02-28 doc/sfincsPaper/refereeResponse/sfincsPaper_refereeResponse.tex
Final version of paper submitted to PoP

@@
 
-We have added some text to section II (between (15)-(16)) and a sentence to section V (between \todo{(35)-(36)}) to clarify these issues.
+We have added some text to section II (between (15)-(16)) and a sentence to section V (between (36)-(37)) to clarify these issues.
 
@@
 
-The system of sources and extra constraint equations actually serves two independent purposes,
+The system of sources and extra constraint equations serves two independent purposes,
 not only to circumvent the conservation problems when $E_r$ is nonzero, but also to eliminate the null
@@
 and even at $E_r=0$.
+
 When $E_r \ne 0$, the sources are needed to eliminate the conservation problems, and the extra constraints 
-can be thought of as a convenient way to keep the linear system square (number of equations = number of unknowns) upon discretizatiton.
+can be thought of as a convenient way to keep the linear system square (number of equations = number of unknowns) upon discretization.
 Even when $E_r = 0$, and even for the DKES trajectories in which sources are not required, the constraints are needed to eliminate
-the null space in the kinetic equation, and the source terms can be thought of as
+the null space in the kinetic equation, and the source terms can then be thought of as
 a convenient way to keep the linear system square upon discretization.
@@
 \begin{itemize}
-\item Albert Moll\'{e}n has recently demonstrated precise agreement between SFINCS and an anlytical calculation
+\item Albert Moll\'{e}n has recently demonstrated precise agreement between SFINCS and an analytical calculation
 of Simakov and Helander for the limit of high collisionality [43]. His results are displayed in a new figure 6, with details given in Appendix B,

### 2014-02-28 doc/sfincsPaper/sfincsPaper.tex
Final version of paper submitted to PoP

@@
 \newcommand{\Sah}{S_{a\mathrm{h}}}
-%\newcommand{\changed}[1]{#1}
-\newcommand{\changed}[1]{\textcolor{red}{#1}}
+\newcommand{\changed}[1]{#1}
+%\newcommand{\changed}[1]{\textcolor{red}{#1}}
 
@@
 
-In this work, we test the validity of several common simplifying assumptions used in numerical neoclassical calculations
-for nonaxisymmetric plasmas.
+In this work, we examine the validity of several common simplifying assumptions used in numerical neoclassical calculations
+for nonaxisymmetric plasmas,
+\changed{
+both by using a new continuum drift-kinetic code
+and by considering analytic properties of the kinetic equation.}
 First, neoclassical phenomena are computed for the LHD and W7-X stellarators using several versions of the drift-kinetic
@@
 tokamak neoclassical codes \cite{Sauter0, Sauter, WongChan, NEOFP, speedGrids}.
-However, due to the numerical challenge of the extra toroidal dimension in stellarators, 
+However, due to the numerical challenge of the extra dimension in stellarators (i.e., the lack of toroidal symmetry), 
 many stellarator neoclassical codes retain only pitch-angle scattering collisions,
@@
 While all of the various options for the particle trajectories have disadvantages,
-SFINCS allows the three options to be compared.
+SFINCS allows the options to be compared.
 As we will show in several calculations for the LHD and W7-X stellarators,
@@
 and demonstrate that the electric field terms in the kinetic description
-correspond to a component of gyroviscosity in a fluid description.}
+correspond to a component of gyroviscosity in a fluid description
+only for the most accurate trajectory model.}
 Details of the numerical implementation are given in section \ref{sec:numerics}.
@@
 in the following manner.
-The kinetic equation becomes well behaved if we introduce small particle and heat sources
+The kinetic equation becomes well behaved if we introduce particle and heat sources
 \begin{equation}
@@
 the sensible requirements that all the flux-surface-averaged density and pressure reside in $F_a$ rather than $f_{a1}$.
-This method of modifying the system of equations has the added benefit 
-(even in the $d\Phi_0/d\psi=0$ limit)
-of eliminating the null solutions $f_{a1} = F_a$ and $f_{a1} = F_a v^2$.
+%This method of modifying the system of equations has the added benefit 
+%(even in the $d\Phi_0/d\psi=0$ limit)
+%of eliminating the null solutions $f_{a1} = F_a$ and $f_{a1} = F_a v^2$.
 When $S_a$ is included in the kinetic equation, new terms proportional to $\Sap$ and/or $\Sah$
@@
 that the three models give nearly identical results. 
-As already mentioned, the need for source terms disappears when the radial electric field equals the value required for ambipolarity, so the source terms do not interfere with the aim of eventually calculating the actual plasma transport.
+And as already mentioned, the source terms for the full trajectory model are both zero when the radial electric field equals the value required 
+for ambipolarity, so for this model the source terms are really a numerical expedient that
+do not affect the transport computations in the end.
 
@@
 \end{equation}
-as reflected (using different indepedent variables) in eq (17) of Ref. \cite{Hazeltine} and eq (6) of Ref. \cite{SimakovCattoSecondOrderDKE}.
+as reflected (using different independent variables) in eq (17) of Ref. \cite{Hazeltine} and eq (6) of Ref. \cite{SimakovCattoSecondOrderDKE}.
 Here, $\bar{f}_a$ is the gyrophase-independent part of the distribution function.
@@
 =2B^2\nabla\psi\cdot\nabla\times\vect{B}=0$,
-we then find the contribution to the parallel momentum equation from (\ref{eq:fTilde})-(\ref{eq:PiE}) is
+we then find the contribution to the parallel momentum equation (\ref{eq:fluidParallelMomentum}) from (\ref{eq:fTilde})-(\ref{eq:PiE}) is
 \begin{equation}
@@
 with (\ref{eq:sources})
-for the three trajectory models (\ref{eq:fullTrajectories})-(\ref{eq:partialTrajectories}),
+for any of the three trajectory models (\ref{eq:fullTrajectories})-(\ref{eq:partialTrajectories}),
 for general nonaxisymmetric nested flux surface geometry, and for an arbitrary number of species.
@@
 the W7-X stellarator \cite{W7X1,W7X2} in its standard configuration.
-In the LHD calculation, only the Boozer harmonics of $B(\theta,\zeta)/B_0$ with amplitude larger than $10^{-2}$
-are retained, as listed in Table 1 of Ref. \cite{BeidlerBigBenchmarking}.
+In the LHD calculation, only the Boozer harmonics of $B(\theta,\zeta)/B_0$ with amplitude $>10^{-2}$
+are retained, as listed in Table 1 of Ref. \cite{BeidlerBigBenchmarking},
+whereas all harmonics with relative amplitude $>4\times 10^{-5}$ are retained for the W7-X calculations.
 For both figures, the Fokker-Planck collision operator is used, and the collisionality

### 2014-04-03 doc/sfincsPaper/sfincsPaper.tex

@@
 for nonaxisymmetric plasmas,
-\changed{
 both by using a new continuum drift-kinetic code
-and by considering analytic properties of the kinetic equation.}
+and by considering analytic properties of the kinetic equation.
 First, neoclassical phenomena are computed for the LHD and W7-X stellarators using several versions of the drift-kinetic
@@
 These issues are explored in section \ref{sec:sources}.
-\changed{In section \ref{sec:momentum}, we discuss some observations regarding momentum conservation,
+In section \ref{sec:momentum}, we discuss some observations regarding momentum conservation,
 and demonstrate that the electric field terms in the kinetic description
 correspond to a component of gyroviscosity in a fluid description
-only for the most accurate trajectory model.}
+only for the most accurate trajectory model.
 Details of the numerical implementation are given in section \ref{sec:numerics}.
@@
 calculations \cite{DKES1, DKES2}.
-The mathematical reason why this term is important at low collisionality is that it has different symmetry properties than other, possible larger, terms in (\ref{eq:kinetic1}). For instance, it 
+The mathematical reason why this term is important at low collisionality is that it has different symmetry properties than other, possibly larger, terms in (\ref{eq:kinetic1}). For instance, it 
 survives if a bounce average is used to annihilate the first term.
@@
 on $d T_a/d\psi$, $dB/d\psi$,
-\changed{
 and $\rho_*$.
@@
 in which we will examine moments of the kinetic equation.
-}
 
@@
 
-\changed{
 Note that for both the DKES and full trajectories, the left-hand side of the
@@
 to (\ref{eq:conservativeForm}).
-}
 
@@
 We will then describe a modified form of the kinetic equation which robustly produces more sensible results.
-\changed{The issues discussed in this section are related to moment equations for mass and energy; momentum has a different status
-and will be examined in the next section.}
+The issues discussed in this section are related to moment equations for mass and energy; momentum has a different status
+and will be examined in the next section.
 
@@
 One motivation for use of the DKES trajectory model is now apparent:
-it is the only model that avoids the imposition of one or more unphysical
+it is the only model (of the three condered here) that avoids the imposition of one or more unphysical
 constraints on the distribution function when $d\Phi_0/d\psi \ne 0$,
@@
 
-\changed{
 This system of sources and constraints solves not only the problem described above when $E_r \ne 0$,
@@
 makes the system of equations invertible.
-}
-
-\changed{
+
 Such is the case when the full linearized Fokker-Planck 
@@
 conservation problem.
-}
-
-\changed{
+
 To summarize, the sources and extra constraint equations serve two independent purposes.
@@
 while we apply the general speed-dependent source $S_a(x_a)$ with $N_x$ constraints when either of the other two collision operators is used.
-}
-
-\begin{comment}
-The aforemention mass and energy sources and constraints work when the Fokker-Planck collision operator
-is used, 
-but not when the pure pitch-angle scattering or momentum-conserving model collision operator is used, since in these cases the kinetic equation has a larger null space
-(any function of $v$ is a solution.)
-For these collision operators, we instead allow a general source $S_a(\psi,v)$,
-introducing $N_x$ unknowns on each flux surface, where $N_x$ is the number of grid points in $v$.
-To supply an equal number of constraints, we impose the equations $\left< \int_{-1}^1 d\xi\, f_{a1}\right>=0$
-at each grid point in $v$.
-\end{comment}
 
@@
 
-\changed{
 Parallel momentum has a different status to density and energy, in that density and energy are conserved by
 the collisionless motion while parallel momentum is not, due to the mirror force.
-(For example, considering the simple case of a single ion species with $d\Phi_0/d\psi=0$, $F_i$ and $v^2 F_i$ are homogeneous
+(For example, considering the case of a single ion species with $d\Phi_0/d\psi=0$, $F_i$ and $v^2 F_i$ are homogeneous
 solutions to the kinetic equation, whereas $v_{||} F_i$ is not.)
@@
 from the $\left< \int d^3 v\, v_{||} (\ldots)\right>$ moment of the various forms of the kinetic equation,
-i.e. there is no exact analogue to (\ref{eq:unphysicalMassConservation}), (\ref{eq:unphysicalEnergyConservation1}),
+i.e. there is no analogue to (\ref{eq:unphysicalMassConservation}), (\ref{eq:unphysicalEnergyConservation1}),
 or (\ref{eq:unphysicalEnergyConservation2}) for momentum.
@@
 limit of the momentum moment equation.
-}
-
-\changed{
+
 Nonetheless, it is interesting to compare the $\int d^3v\, m_a v_{||} (\ldots)$ moment equations
@@
 at least when $E_r=0$.
-}
-
-\changed{
+
 Now consider how the situation changes when a radial electric field is introduced. We first
@@
 \end{equation}
-}
-
-\changed{
+
 For comparison, let us consider the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the radial electric field
@@
 viscous force as the full trajectory model does.
-}
-
-\changed{
+
 We close this section by noting another important difference between the trajectory models
@@
 so the net heat source vanishes for any radial electric field.
-}
 
@@
 of the method here.  
-\changed{We do not expect any fundamental new complications to arise
-if poloidal and toroidal magnetic drift terms are retained.}
+We do not expect any fundamental new complications to arise
+if poloidal and toroidal magnetic drift terms are retained.
 However, to include radial drifts acting on $f_{a1}$,
@@
 
-In \changed{the typical ``ion root'' scenario}, $E_*$ can be estimated by noting that the
+In the typical ``ion root'' scenario, $E_*$ can be estimated by noting that the
 ambipolar electric field arises to bring the ion particle transport
@@
 \rho / (\iota \epsilon)$ and $\epsilon$ is the inverse aspect ratio,
-\changed{
 and the ratio $\rho_\theta / L_\perp$ is typically $\ll 1$.
@@
 In these scenarios, $E_*$ may be $\sim 1$.
-}
 
@@
 of the normalized radial electric field (\ref{eq:EStar}).
-\changed{Results for $E_r=0$ are indicated by the $\blacktriangleright$ symbol to the left of each plot.}
+Results for $E_r=0$ are indicated by the $\blacktriangleright$ symbol to the left of each plot.
 \label{fig:ErComparison_LHD}}
@@
 of the normalized radial electric field (\ref{eq:EStar}).
-\changed{Results for $E_r=0$ are indicated by the $\blacktriangleright$ symbol to the left of each plot.}
+Results for $E_r=0$ are indicated by the $\blacktriangleright$ symbol to the left of each plot.
 \label{fig:ErComparison_W7X}}
@@
 the electrons, since $E_*$ defined using the electron rather than ion thermal speed is always $\ll 1$.
-The vertical magenta dotted line indicates the ambipolar value of $E_r \approx $ 33 kV/m, which is effectively
+The vertical magenta dotted line indicates the ambipolar value of $E_r \approx -33$  kV/m, which is effectively
 identical for the three trajectory models, and comparable to the value predicted in \cite{Turkin}.
@@
 trajectory models.
-\changed{
 The ion particle and heat sources in (E)-(F) are normalized
-by a gyro-Bohm transport time $t_{gB}$.}
+by a gyro-Bohm transport time $t_{gB}$.
 \label{fig:2species}}
@@
 
-\changed{
 Figures (\ref{fig:2species}).E-F illustrate the ion particle and heat sources computed as part of the calculation.
@@
 than the collision term in the kinetic equation for this calculation.
-}
 
@@
 \begin{equation}
-m_a n_a \vect{V}_a \cdot \nabla \vect{V}_a = n_a e_a \left( - \nabla \phi + c^{-1}\vect{V}_a \times \vect{B} \right)
+m_a n_a \vect{V}_a \cdot \nabla \vect{V}_a = n_a e_a \left( - \nabla \Phi + c^{-1}\vect{V}_a \times \vect{B} \right)
 - \nabla p_a - \nabla \cdot \tens{\pi}_a + \vect{F}_a, 
@@
 lower resolution is sufficient, so memory and time requirements are reduced;
-\changed{
 for example, in the same W7-X geometry at $\nu'=10^{-2}$, sufficient resolution parameters for convergence
 were $N_\theta=11$, $N_\zeta=64$, $N_\xi=100$, and $N_x=5$,
-and computations required 3 minutes on 1 node of Edison.}
+and computations required 3 minutes on 1 node of Edison.
 Computations with $\nu' > 10^{-2}$ can typically be run on a laptop.
 
-\changed{
 When the pure pitch-angle scattering collision operator 
@@
 weights and normalized in the same way as (\ref{eq:matrix}).
-}
-
-\changed{
+
 In the short-mean-free-path limit $\nu' \gg 1$, the ion transport and flow can be computed
@@
 in the appropriate $\nu' \gg 1$ limit.
-}
 
@@
 SFINCS computations for three different collision operators are compared.
-\changed{
 Also shown (black crosses) are the transport matrix elements
@@
 demonstrating excellent agreement with SFINCS when the latter is run with the same collision model. 
-}
 \label{fig:collisionComparison_W7X}}
@@
 \caption{(Color online)
-\changed{
 The ion transport matrix elements (defined in (\ref{eq:matrix})) are plotted as functions
@@
 which agree quite well with the SFINCS computations at high collisionality.
-}
 \label{fig:SimakovHelanderComparison}}
@@
 
-Different stellarator transport codes make different assumptions about the
-details of the drift-kinetic equation, and our goal in this work has been
-to examine the consequences of these details.
-In this work, we have compared three versions of the drift-kinetic equation
+In this work, we have examined the impact of several approximations made in 
+stellarator kinetic codes, approximations related to the electric field and the collision operator.
+We have compared three versions of the drift-kinetic equation
 for a stellarator, consisting of (\ref{eq:kinetic4}) with
@@
 Equations  (\ref{eq:fullTrajectories}) and  (\ref{eq:partialTrajectories})
-appear more accurate than  (\ref{eq:DKESTrajectories}),
-\changed{and as we have shown in section \ref{sec:momentum},
+appear to be more accurate than  (\ref{eq:DKESTrajectories}),
+and as we have shown in section \ref{sec:momentum},
 the full trajectory model (\ref{eq:fullTrajectories}) is the only
 one of the three models which gives the correct parallel viscous force 
-and which preserves intrinsic ambipolarity in quasisymmetry.}
+and which preserves intrinsic ambipolarity in quasisymmetry.
 However, as we have shown in section \ref{sec:sources},
@@
 and to Craig Beidler and Peter Catto for other input on this work.
+We also thank the anonymous referee for making several suggestions
+that substantially improved the paper.
+
 \end{acknowledgments}
@@
 discussed analytically in Refs. \cite{Pytte, Boozer83, meQS}.
-\changed{
-Equivalent to the definition at the end of section \ref{sec:sources} \cite{PerAndreiPRL},
+Equivalent to the definition at the end of section \ref{sec:momentum} \cite{PerAndreiPRL},
 a quasisymmetric magnetic field is one satisfying $B(\theta,\zeta) = y(M\theta-N\zeta)$
@@
 This isomorphism holds for all the trajectory models considered in this paper.
-}
-\begin{comment}
-flow
-and heat flux should obey the transformations
-\begin{equation}
-\left< V_{||} B \right> \frac{M-N/\iota}{NI+MG} = C_V
-\end{equation}
-and
-\begin{equation}
-\left< \int d^3v\, f_{\mathrm{i}1} \frac{v^2}{2}\vect{v}_{\mathrm{m}i}\cdot\nabla\psi \right> \frac{\left| M-N/\iota\right| }{(NI+MG)^2} = C_q
-\end{equation}
-respectively, where the quantities $C_V$ and $C_q$ should be independent of $M$ and $N$.
-\end{comment}
-
-\changed{
+
 As $\nuii / (\iota M - N)$ is to be held fixed in this test, while $\iota M - N$ can change sign as $M$ and $N$ are varied, 
@@
 $\propto 1/|\iota M-N|$.
-}
 
 It was verified that the SFINCS code obeyed both versions of these isomorphism transformations
-for various $y$, collisionality regimes, \changed{radial electric fields, trajectory models}, and collision operators.
+for various $y$, collisionality regimes, radial electric fields, trajectory models, and collision operators.
 
@@
 
-\changed{
 From the analytic calculations presented in Ref.~\cite{AndreiPer2009} we can derive expressions
 for $L_{jk}$ of Eq.~(\ref{eq:matrix}) in the Pfirsch-Schl{\"u}ter regime. 
-The transport matrix elements then depend on the function $u$ given by the solution to Eq.~(8) in Ref.~\cite{AndreiPer2009};
+\changed{Note that a pure plasma with singly charged ions is assumed in Ref. \cite{AndreiPer2009}, so we specialize to this case of $Z=1$ in this appendix.}
+The transport matrix elements depend on the function $u$ given by the solution to Eq.~(8) in Ref.~\cite{AndreiPer2009};
 $u$ is proportional to the parallel current divided by $B$. 
@@
 in the absence of radial gradients (i.e. when $E_{||}$ is the only thermodynamic force present). Then $L_{33}$ can be found from the flow $\left< V_\| B \right>$.
-}
-
-\changed{
+
 Expressions for the matrix coefficients in the Pfirsch-Schl{\"u}ter regime are summarized in 
@@
 we obtain the more accurate coefficients given in ${K}_{2}$ below.
-}
-
-\changed{
-%\begin{equation}
-%\begin{array}{c}
+
 \begin{align}
@@
 \end{align}
-}
 

### 2015-05-21 doc/manual/inputParameters.tex
Added user manual

\chapter{Input Parameters}

\newcommand{\param}[5]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries #1}\\{\it Type}: #2\\{\it Default}: #3\\{\it When it matters}: #4\\{\it Meaning}: #5}}
\newcommand{\ssparam}[4]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries #1}\\{\it Type}: #2\\{\it When it matters}: #3\\{\it Meaning}: #4}}
\newcommand{\PETScParam}[2]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries #1}\\{\it Meaning}: #2}}
\newcommand{\myhrule}{{\setlength{\parindent}{0cm} \hrulefill }}

In this chapter we first describe all the parameters which can be included in the {\ttfamily input.namelist} file. 
Then we list some of the command-line flags associated with \PETSc~which can be useful.

\section{The {\ttfamily general} namelist}

\param{RHSMode}
{integer}
{1}
{Always}
{Option related to the number of right-hand sides (i.e. inhomogeneous drive terms) for which the kinetic equation is solved.\\

{\ttfamily RHSMode=1}: Solve for a single right-hand side.\\

{\ttfamily RHSMode=2}: Solve for 3 right-hand sides to get the 3x3 transport matrix. Presently implemented only for 1 species.\\

{\ttfamily RHSMode=3}: Solve for the 2x2 monoenergetic transport coefficients. When this option is chosen, {\ttfamily Nx} is set to 1 and only 1 species is used.
}

\myhrule

\param{outputFileName}
{string}
{``sfincsOutput.h5''}
{Always}
{Name which will be used for the HDF5 output file.  If this parameter is changed from the default value, \sfincsScan~ will not work.}

\myhrule

\param{saveMatlabOutput}
{Boolean}
{false}
{Always}
{If this switch is set to true, Matlab m-files are created which
 store the system matrix, right-hand side, and solution vector.  If an iterative solver is used,
 the preconditioner matrix is also saved.
 PETSc usually generates an error message if you ask to save Matlab output when
the size of the linear system is more then 1400 $\times$ 1400,
 so usually this setting should be false except for very small test problems.
}

\myhrule

\param{MatlabOutputFilename}
{string}
{``sfincsMatrices''}
{Only when {\ttfamily saveMatlabOutput == .true.}}
{Start of the filenames which will be used for Matlab output.}

\myhrule

\param{saveMatricesAndVectorsInBinary}
{Boolean}
{false}
{Always}
{If this switch is set to true, the matrix, right-hand-side, and solution of the
linear system will be saved in PETSc's binary format.  The preconditioner matrix will also
be saved if {\ttfamily tryIterativeSolver == .true.}
}

\myhrule

\param{binaryOutputFilename}
{string}
{``sfincsBinary''}
{Only when {\ttfamily saveMatricesAndVectorsInBinary == .true.}}
{Start of the filenames which will be used for binary output.}

\myhrule

\param{solveSystem}
{Boolean}
{true}
{Always}
{If this parameter is false, the system of equations will not actually be solved.
 Sometimes it can be useful to set this parameter to {\ttfamily .false.} when debugging.
}

\section{Directives for \sfincsScan}

The parameters for \sfincsScan~ begin with the code {\ttfamily !ss} and so are not read by the fortran part of \sfincs.
These parameters matter only when \sfincsScan~is called and are all ignored when \sfincs~is executed directly.
These parameters can appear anywhere in the {\ttfamily input.namelist} file, in any namelist or outside of any namelist.
Note that \sfincsScan~ parameters do not have defaults, unlike fortran namelist parameters.\\

\ssparam{scanType}
{integer}
{Any time \sfincsScan~is called.}
{Which type of scan will be run when \sfincsScan~is called.\\

{\ttfamily scanType=1}:  Convergence scan. (Scan the parameters in the resolutionParameters namelist.)\\

{\ttfamily scanType=2}:  Scan of $E_r$.\\

{\ttfamily scanType=3}:  Scan any one input parameter that takes a numeric value.\\

{\ttfamily scanType=4}:  Scan radius, taking the density and temperature profiles from the {\ttfamily profiles} file.
In this type of scan, the same radial electric field is used at every radius.
See {\ttfamily utils/profiles.XXX} for examples.\\

{\ttfamily scanType=5}:  Scan radius, and at each radius, scan $E_r$.  Density and temperature profiles are again
 taken from the {\ttfamily profiles} file; see {\ttfamily utils/profiles.XXX} for examples.\\

{\ttfamily scanType=21}: Read in a list of requested runs from a file {\ttfamily runspec.dat}. See {\ttfamily utils/sfincsScan\_21} for an example file.
}

\myhrule

\section{\PETSc~commands}

Command-line flags can be used to modify the behavior of any \PETSc~application, including \sfincs.
There are hundreds of \PETSc~options, and a list can be obtained by running with the command-line flag
{\ttfamily -help}. Here we list some of the more useful options.\\

\PETScParam{-help}
{Dumps a list of available command-line options to stdout.}

\myhrule

\PETScParam{-ksp\_view}
{Dumps detailed information to stdout related to the linear solver.}

### 2015-05-21 doc/manual/installation.tex
Added user manual

\chapter{Installation}

\section{Requirements}

To compile \sfincs~you need the \href{http://www.mcs.anl.gov/petsc/}{\PETSc} library (real version, as opposed to complex version) 
and the serial \href{https://www.hdfgroup.org/HDF5/}{\HDF} library.  
\PETSc~is used for iterative solution of large linear and nonlinear systems of equations, and \HDF~is used for saving output.
We have developed and tested \sfincs~ with \PETSc~
versions 3.2 through 3.5.  The commands in \PETSc~often change from version to version, so future versions of \PETSc~may require modifications
to the \sfincs~source code.

Although \sfincs~can be run on a single processor, usually you want to run it in parallel.  
In this case, you need {\ttfamily MPI}, and you need either \href{http://mumps-solver.org/}{\mumps}
or \href{http://crd-legacy.lbl.gov/~xiaoye/SuperLU/}{\superludist}.  (Note that \superludist is a parallel library which is different from the serial library {\ttfamily superlu}).
Both \mumps~and \superludist~are parallelized libraries for direct solution of large sparse linear systems. 
\PETSc~has a built-in serial sparse direct linear solver,
but it sometimes gives an error that there is a ``zero pivot'' when \mumps~and \superludist~have no problem solving the system;
therefore you may want to use \mumps~or \superludist~even for serial runs.  In our experience,
\mumps~requires less memory and time than \superludist~for solving a given linear system.

If you want to load {\ttfamily VMEC wout} files in \netCDF~format, then you need the \netCDF~library.
This library is not required for loading ASCII-format {\ttfamily VMEC wout} files.  If you want to compile \sfincs~without
\netCDF~, then edit {\ttfamily makefile} so that no value is assigned to {\ttfamily USE\_NETCDF}.

The plotting routines \sfincsPlot~and \sfincsScanPlot~ require python 2.X, {\ttfamily numpy}, {\ttfamily scipy}, and {\ttfamily matplotlib}.
These python libraries are not required by the core fortran part of \sfincs.

Although older {\ttfamily MATLAB} versions of \sfincs~are included in the \sfincs~repository,
{\ttfamily MATLAB} is not required for running the fortran version of \sfincs.

\section{Cloning the repository}

The source code for \sfincs~is hosted in a {\ttfamily git} repository at
\url{https://github.com/landreman/sfincs}.
You obtain the \sfincs~source code by cloning the repository. This requires several steps.

\begin{enumerate}
\item Create an account on \url{github.com}, and sign in to {\ttfamily github}.
\item Go to your account settings page, by clicking the wrench icon on the top right.
\item Click on ``SSH keys'' on the left, and add an SSH key for the computer you wish to use. To do this, you may wish to read see the ``generating SSH keys'' guide which is linked to from that page.
\item From a terminal command line in the computer you wish to use, enter {\ttfamily git clone git@github.com:landreman/sfincs.git} to download the repository.
\end{enumerate}

Any time after you have cloned the repository in this way, you can download future updates to the code by entering {\ttfamily git pull} from any subdirectory within your local copy.

\section{Makefiles and environment variables}

To use \sfincs~ you must set the environment variable {\ttfamily SFINCS\_SYSTEM}.
(For example, using the {\ttfamily bash} shell on the {\ttfamily edison}
computer, you would type\\
{\ttfamily export SFINCS\_SYSTEM=edison}\\
 at the command line
or in your {\ttfamily .bashrc} startup script.)
This variable is used in two ways. 
First, {\ttfamily make} uses this variable to looks for the 
appropriate makefile in the {\ttfamily makefiles} directory.  Second, the {\ttfamily SFINCS\_SYSTEM} environment
variable affects the behavior of
\sfincsScan~in several ways, such as determining the command used to submit jobs to the system's queue.

\section{Setting up \sfincs~on a new system}
If you are setting up \sfincs~ on a new system, one for which there is no file {\ttfamily makefiles/makefile.XXX},
there are several things you need to do.

First, copy one of the existing makefiles, and edit it as appropriate.

Second, you will need to edit {\ttfamily utils/sfincsScan}. Look for the {\ttfamily if} block near the top
with sections for {\ttfamily sfincsSystem = edison},{\ttfamily hydra}, and {\ttfamily laptop}. Add an analogous block
for your system to set the command used to submit jobs, and a {\ttfamily nameJobFile} function.

\section{Make test}

### 2015-05-21 doc/manual/overview.tex
Added user manual

\chapter{Overview}

SFINCS is a nifty code \cite{sfincsPaper}.

\section{\sfincs~vs. \sfincsScan}
The fortran part of \sfincs~ solves the kinetic equation for each species
at a single flux surface, a single value of $E_r$, and a single set of resolution parameters.
However, often the goal is to determine the ambipolar $E_r$ at one or more surfaces.
For this task, the \sfincsScan~family of python scripts

\begin{figure}[htbp]
\begin{center}
%\includegraphics[width=1.0\textwidth]{poloidal_cross-sections.pdf}
\mycaption{Trajectories of trapped particles.  (a) In tokamaks and omnigenous stellarators,
all collisionless trajectories are confined.  (b) In contrast, trapped particles in other stellarators
may have a radial drift that does not average to zero.
\label{fig:poloidal_cross_sections}}
\end{center}
\end{figure}

### 2015-05-21 doc/manual/resolution.tex
Added user manual

\chapter{Numerical resolution parameters}

\section{Convergence testing}

\section{Typical resolution requirements}

### 2015-05-22 doc/manual/inputParameters.tex
Added to user manual

@@
 \newcommand{\myhrule}{{\setlength{\parindent}{0cm} \hrulefill }}
+
+\newcommand{\true}{{\ttfamily .true.}}
+\newcommand{\false}{{\ttfamily .false.}}
 
@@
 {Boolean}
-{false}
+{\false}
 {Always}
@@
 {Boolean}
-{true}
+{\true}
 {Always}
@@
 
+\section{The {\ttfamily geometryParameters} namelist}
+
+\param{inputRadialCoordinate}
+{integer}
+{3}
+{When {\ttfamily geometryScheme}==5, 11, or 12}
+{Which radial coordinate to use to specify the flux surface for a single calculation,
+or to specify the range of flux surfaces for a radial scan. See section \ref{sec:radialCoordinates}
+for more information about radial coordinates.\\
+
+{\ttfamily inputRadialCoordinate}==0: Use the flux surface specified by {\ttfamily psiHat\_wish}
+for a single run, and use the range specified by {\ttfamily psiHat\_min} and {\ttfamily psiHat\_max} for radial scans.\\
+
+{\ttfamily inputRadialCoordinate}==1: Use the flux surface specified by {\ttfamily psiN\_wish}
+for a single run, and use the range specified by {\ttfamily psiN\_min} and {\ttfamily psiN\_max} for radial scans.\\
+
+{\ttfamily inputRadialCoordinate}==2: Use the flux surface specified by {\ttfamily rHat\_wish}
+for a single run, and use the range specified by {\ttfamily rHat\_min} and {\ttfamily rHat\_max} for radial scans.\\
+
+{\ttfamily inputRadialCoordinate}==3: Use the flux surface specified by {\ttfamily rN\_wish}
+for a single run, and use the range specified by {\ttfamily rN\_min} and {\ttfamily rN\_max} for radial scans.
+}
+
 \section{Directives for \sfincsScan}
+\label{sec:sfincsScanParams}
 
@@
 {Any time \sfincsScan~is called.}
-{Which type of scan will be run when \sfincsScan~is called.\\
+{Which type of scan will be run when \sfincsScan~is called. \\
 

### 2015-05-22 doc/manual/installation.tex
Added to user manual

@@
 \section{Requirements}
+\label{sec:requirements}
 
 To compile \sfincs~you need the \href{http://www.mcs.anl.gov/petsc/}{\PETSc} library (real version, as opposed to complex version) 
-and the serial \href{https://www.hdfgroup.org/HDF5/}{\HDF} library.  
+and the \href{https://www.hdfgroup.org/HDF5/}{\HDF} library.  Even if you will be running \sfincs~in parallel, you
+only need the serial version of \HDF, not the parallel version.
 \PETSc~is used for iterative solution of large linear and nonlinear systems of equations, and \HDF~is used for saving output.
@@
 This library is not required for loading ASCII-format {\ttfamily VMEC wout} files.  If you want to compile \sfincs~without
-\netCDF~, then edit {\ttfamily makefile} so that no value is assigned to {\ttfamily USE\_NETCDF}.
+\netCDF, then edit \path{sfincs/fortran/version3/makefile} so that no value is assigned to {\ttfamily USE\_NETCDF}.
 
-The plotting routines \sfincsPlot~and \sfincsScanPlot~ require python 2.X, {\ttfamily numpy}, {\ttfamily scipy}, and {\ttfamily matplotlib}.
-These python libraries are not required by the core fortran part of \sfincs.
+The plotting routines \sfincsPlot~and \sfincsScanPlot~require \python~2.X, {\ttfamily numpy}, {\ttfamily scipy}, and {\ttfamily matplotlib}.
+These \python~libraries are not required by the core fortran part of \sfincs.
 
@@
 First, {\ttfamily make} uses this variable to looks for the 
-appropriate makefile in the {\ttfamily makefiles} directory.  Second, the {\ttfamily SFINCS\_SYSTEM} environment
+appropriate makefile in the
+\path{sfincs/fortran/version3/makefiles} directory.  Second, the {\ttfamily SFINCS\_SYSTEM} environment
 variable affects the behavior of
@@
 
+You will probably want to add the directory \path{sfincs/fortran/version3/utils/} to your
+path.  This directory contains the scripts for plotting output and running parameter scans.
+
 \section{Setting up \sfincs~on a new system}
-If you are setting up \sfincs~ on a new system, one for which there is no file {\ttfamily makefiles/makefile.XXX},
+If you are setting up \sfincs~ on a new system, one for which there is no file
+\path{sfincs/fortran/version3/makefiles/makefile.XXX},
 there are several things you need to do.
@@
 
+Third, if you want {\ttfamily make test} to work (see section \ref{sec:maketest}),
+you will need to create files {\ttfamily job.SFINCS\_SYSTEM}
+for each example in the \path{sfincs/fortran/version3/examples/} directory that you want to include
+in the tests.  You may be able to use the same {\ttfamily job.SFINCS\_SYSTEM} file for each example, but for the largest examples,
+you may want to use different numbers of processes or different queues for different examples.
+
+\section{Compiling}
+
+If your system uses ``modules'', make sure you have loaded any required modules.
+(Requirements are discussed in section \ref{sec:requirements}).  
+There may be
+instructions for the specific modules required on your system in the comments in the
+appropriate makefile \\{\ttfamily sfincs/fortran/version3/makefiles/makefile.SFINCS\_SYSTEM} for your system.
+
+Next, to compile, go to the directory {\ttfamily sfincs/fortran/version3/} and run {\ttfamily make}.
+
 \section{Make test}
+\label{sec:maketest}
+
+To test that your \sfincs~executable is working, you can run {\ttfamily make test}
+from the \path{sfincs/fortran/version3/} directory.  Doing so will run
+\sfincs~for some or all of the examples in the {\ttfamily sfincs/fortran/version3/examples/} directories.
+(The runs will be performed in series if no queueing system is available, otherwise the runs will all
+be submitted to the queueing system.)
+After each example completes, several of the output quantities (such as parallel flows and radial fluxes)
+will be checked, using the
+{\ttfamily tests\_small.py} or {\ttfamily tests\_large.py} script in the example's directory.
+
+If you run {\ttfamily make retest} from the \path{sfincs/fortran/version3/} directory,
+no new runs of \sfincs~will be performed, but the {\ttfamily tests\_small.py} or {\ttfamily tests\_large.py} script
+will be run on any existing {\ttfamily sfincsOutput.h5} files in the \path{sfincs/fortran/version3/examples/} directories.

### 2015-05-22 doc/manual/overview.tex
Added to user manual

@@
 
-SFINCS is a nifty code \cite{sfincsPaper}.
+The \sfincs~code is a free, open-source tool for solving neoclassical-type kinetic problems in nonaxisymmetric or axisymmetric plasmas
+with nested toroidal flux surfaces.
+The code solves a drift-kinetic equation for an arbitrary number of species.  Optionally, a quasi-neutrality
+equation can also be solved to obtain the self-consistent variation of the electrostatic potential on a flux surface.
+
+This manual describes ``version 3'' of \sfincs.  There are two older versions of the
+code called singleSpecies and multiSpecies.
+
+This document discusses use and operation of the code.  For more details about the specific equations implemented,
+see the version 3 technical documentation available in the \path{sfincs/docs} directory.
+Ref \cite{sfincsPaper} gives many details and some early physics results.
+
+\section{GMRES/KSP and preconditioning}
+
+At its heart, \sfincs~solves one or more large sparse linear systems $Ax=b$.  Here $b$ is a known right-hand side vector,
+$A$ is a large (often millions $\times$ millions) known sparse matrix, and $x$ is the desired and unknown solution vector.
+The direct way to solve such systems is to $LU$-factorize
+the matrix $A$ into lower- and upper-triangular factors.
+Once the $L$ and $U$ factors are found, the solution of the linear system
+for any right-hand side vector can be rapidly obtained.  However, even if the original matrix is sparse, the $L$ and $U$ factors
+are generally not sparse, and so a very large amount of memory can be required for a direct $LU$-factorization.
+
+An alternative way to solve
+such large linear systems is with a so-called ``Krylov-space'' iterative method, which can dramatically reduce the memory required
+compared to a direct solution.  For the non-symmetric matrices that arise in \sfincs, the preferred Krylov-space
+algorithm is called GMRES (Generalized Minimal RESidual.)  In \sfincs, the \PETSc~library is used to solve the large systems
+of equations. \PETSc~calls its family of linear solvers KSP, so in the output of \sfincs~you will see a ``KSP residual'' reported
+as GMRES iterates towards the solution.
+
+An important element of Krylov methods is preconditioning.  The art of preconditioning is to find a linear operator which has similar eigenvalues
+to the ``true'' matrix you would like to invert (or more precisely, to $LU$-factorize), but which can be inverted faster.
+If a good preconditioner can be found, the number of GMRES iterations is greatly reduced.  Many schemes for preconditioning exist,
+but the version adopted in \sfincs~is to explicity form and $LU$-factorize a preconditioning matrix which is similar
+to the true matrix (but somewhat simpler).  There is a basic trade-off:
+the more similar the preconditioner matrix is to the true matrix, the fewer iterations will be required, but the more time will
+be required to $LU$-factorize the preconditioning operator.  The usual preconditioner matrix in \sfincs~is obtained by dropping all coupling
+between grid points in the speed coordinate and dropping coupling between species.  The preconditioner matrix need not be a physically
+accurate or meaningful operator; as long as GMRES converges, the solution obtained will be independent of the preconditioner to whatever tolerance is specified.
 
 \section{\sfincs~vs. \sfincsScan}
-The fortran part of \sfincs~ solves the kinetic equation for each species
-at a single flux surface, a single value of $E_r$, and a single set of resolution parameters.
-However, often the goal is to determine the ambipolar $E_r$ at one or more surfaces.
-For this task, the \sfincsScan~family of python scripts
+The core fortran part of \sfincs~solves the kinetic equation for each species
+at a single flux surface, a single value of $E_r$, and a single set of other parameters.
+However, often the goal is to determine the ambipolar $E_r$ at one or more surfaces, or to scan some other parameter.
+For this task, the \sfincsScan~family of \python~scripts is available.
+For a full list of the types of scans available, see section \ref{sec:sfincsScanParams}
 
-\begin{figure}[htbp]
-\begin{center}
-%\includegraphics[width=1.0\textwidth]{poloidal_cross-sections.pdf}
-\mycaption{Trajectories of trapped particles.  (a) In tokamaks and omnigenous stellarators,
-all collisionless trajectories are confined.  (b) In contrast, trapped particles in other stellarators
-may have a radial drift that does not average to zero.
-\label{fig:poloidal_cross_sections}}
-\end{center}
-\end{figure}
+\section{Bugs and feedback}
+
+To report any bugs, provide feedback, or ask questions, contact Matt Landreman at
+\href{mailto:matt.landreman@gmail.com}{\nolinkurl{matt.landreman@gmail.com} }
+
+%\begin{figure}[htbp]
+%\begin{center}
+%%\includegraphics[width=1.0\textwidth]{poloidal_cross-sections.pdf}
+%\mycaption{Trajectories of trapped particles.  (a) In tokamaks and omnigenous stellarators,
+%all collisionless trajectories are confined.  (b) In contrast, trapped particles in other stellarators
+%may have a radial drift that does not average to zero.
+%\label{fig:poloidal_cross_sections}}
+%\end{center}
+%\end{figure}

### 2015-05-22 doc/manual/resolution.tex
Added to user manual

@@
 
+Results from \sfincs~should only be believed if you are confident they
+are converged with respect to the numerical resolution parameters
+\Ntheta, \Nzeta, \Nxi, and \Nx.  That is, you want to be sure the
+physics output of the code does not change significantly when any of
+these parameters are increased. The values of \Ntheta, \Nzeta, \Nxi,
+and \Nx~required for convergence depend strongly on the magnetic
+geometry and collisionality, with modest dependence also on the radial
+electric field. It is strongly recommended that you test for
+convergence with respect to \Ntheta, \Nzeta, \Nxi, and \Nx~whenever
+beginning \sfincs~calculations for a new scenario.
+
+Note that ``convergence'' in this sense (convergence with respect to resolution parameters assuming the discretized
+system is solved exactly) is separate from the convergence of GMRES/KSP.
+
+\section{Relatively unimportant resolution parameters}
+
+There are several resolution parameters which are almost never the limiting factor
+for convergence.  You can almost always use {\ttfamily NL=4}, 
+{\ttfamily solverTolerance} = $10^{-5}$ or $10^{-6}$,
+{\ttfamily xMax=5.0},
+and {\ttfamily NxPotentialsPerVth=40.0}.  The latter two of these parameters are in fact ignored
+for the recommended and default {\ttfamily xGridMode} setting, 5.
+
+\section{General suggestions}
+
+The time and memory requirements of the code increase significantly
+when \Ntheta, \Nzeta, \Nxi, or \Nx~are increased. Therefore, you
+probably want to only scan one of these four parameters at a time
+(rather than increasing two or more of them simultaneously) when
+testing for convergence. (This recommended approach is the one taken
+in \sfincsScan~automated convergence scans, discussed in section
+\ref{sec:convergence})
+
+When the mean-free-path is shorter than the parallel length scale of the equilibrium,
+({\ttfamily nuPrime} $\ge$ 1), the parameters required for convergence
+do not depend much on collisionality. In the opposite
+limit in which the mean-free-path is longer than the parallel length scale of the equilibrium,
+({\ttfamily nuPrime} $\le$ 1)
+values of
+\Nzeta~and \Nxi~required for convergence increase dramatically
+as collisionality decreases. The required value of \Ntheta~increases as well, but often
+not quite as dramatically. The \Nx~required for convergence does not depend
+much on collisionality, though typically the required value increases
+slightly with collisionality at high collisionality
+
+The \Nx~required for convergence may need to increase slightly with the number of species.
+Typically you can expect to use \Nx=5-8.
+
+The resolution parameters do not need to vary much with the radial
+electric field as long as the electric field is below about 1/3 of the resonant value.
+(In the notation of \cite{sfincsPaper}, when $E_*<1/3$).
+For almost all experimentally relevant situations (except for HSX where $T_i/T_e$ is extremely small),
+the electric field is far below the resonance, in which case you should not need to vary the resolution
+parameters with the electric field.
+However, if you do approach the $E_r$ resonance, \Nx~will likely need to be increased.
+
 \section{Convergence testing}
+\label{sec:convergence}
 

### 2015-05-22 doc/manual/runs.tex
Added to user manual

\chapter{Specifying and running a computation}

\section{Normalizations}

{\setlength{\parindent}{0cm}
Dimensional quantities in \sfincs~are normalized to ``reference'' values that are denoted by a bar:\\
$\bar{B}$ = reference magnetic field, typically 1 Tesla.\\
$\bar{R}$ = reference length, typically 1 meter.\\
$\bar{n}$ = reference density, typically $10^{19}$ m$^{-3}$, $10^{20}$ m$^{-3}$, or something similar.\\
$\bar{m}$ = reference mass, typically either the mass of hydrogen or deuterium.\\
$\bar{T}$ = reference temperature in energy units, typically 1 eV or 1 keV.\\
$\bar{v} = \sqrt{2 \bar{T} / \bar{m}}$ = reference speed\\
$\bar{\Phi}$ = reference electrostatic potential, typically 1 V or 1 kV.\\
}

You can choose any reference parameters you like, not just the values
suggested here. However, if you use a {\ttfamily vmec} or {\ttfamily .bc} magnetic equilibrium
by choosing {\ttfamily geometryScheme} = 5, 11, or 12, then you MUST use $\bar{B}$ = 1 Tesla and $\bar{R}$ = 1 meter.
The code ``knows'' about the reference values only through
the 3 combinations {\ttfamily Delta}, {\ttfamily alpha}, and {\ttfamily nu\_n}
in the {\ttfamily physicsParameters} namelist.

Normalized quantities are denoted by a ``hat''.  Taking the magnetic field as an example,
$\hat{B}=B/\bar{B}$, where $\hat{B}$ is called {\ttfamily BHat} in the fortran code and \HDF~output file.

\section{Radial coordinates}
\label{sec:radialCoordinates}

A variety of flux-surface label coordinates are used in other codes and in the literature.
One common choice (used in {\ttfamily vmec}) is $\psi_N$, the toroidal flux normalized to its
value at the last closed flux surface.  Another common choice is an ``effective normalized minor radius''
$r_N$, defined by $r_N=\sqrt{\psi_N}$.  For gradients of density, temperature, and electrostatic potential (i.e. the radial
electric field), it is useful to use a dimensional local minor radius $r = r_N a$, where $a$ is some
measure of the plasma effective outer minor radius.  Finally, one could also use $\psi$ directly.
For maximum flexibility, \sfincs~permits any of these four radial coordinates to be used, and different radial
coordinates can be used in different parts of a given computation.  Output quantities which depend
on the radial coordinate, such as radial fluxes, are often given with respect to all radial coordinates.

To specify a run, you can make several independent choices for radial coordinates.  One parameter
you select is {\ttfamily inputRadialCoordinate} in the {\ttfamily geometryParameters} namelist.
This parameter controls which coordinate is used to specify the flux surface. The possible values of {\ttfamily inputRadialCoordinate} are:

0: Use {\ttfamily psiHat}, the toroidal flux (divided by $2\pi$) normalized by $\bar{B}\bar{R}^2$.\\

1: Use {\ttfamily psiN}, the toroidal flux normalized by its value at the last closed flux surface.\\

2: Use {\ttfamily rHat}, defined as {\ttfamily aHat}$\sqrt{\mbox{\ttfamily psiN}}$, where {\ttfamily aHat} is an effective minor radius of the last closed flux surface normalized by $\bar{R}$.\\

3: Use {\ttfamily rN}, defined as $\sqrt{\mbox{\ttfamily psiN}}$.\\

\section{Parallelization}
Choosing the number of nodes and procs.

\section{Issues with running on 1 processor}

### 2015-05-22 doc/manual/inputParameters.tex
Added to user manual

@@
 \chapter{Input Parameters}
+\label{ch:input}
 
@@
 \section{The {\ttfamily geometryParameters} namelist}
+\label{sec:geometryParameters}
+
+\param{geometryScheme}
+{integer}
+{1}
+{Always}
+{How the magnetic geometry is specified.\\
+
+{\ttfamily geometryScheme}==1: Use the following 3-helicity model:\\
+\begin{eqnarray}
+\label{eq:Bmodel}
+B(\theta,\zeta)/\bar{B} 
+&=& (\mbox{\ttfamily B0OverBBar}) [1 + (\mbox{\ttfamily epsilon\_t}) \cos(\theta) \\
+&& + (\mbox{\ttfamily epsilon\_h}) \cos((\mbox{\ttfamily helicity\_l}) \theta - (\mbox{\ttfamily helicity\_n}) \zeta) \nonumber \\
+&&  + (\mbox{\ttfamily epsilon\_antisymm}) \nonumber \\
+&& \times \sin((\mbox{ \ttfamily helicity\_antisymm\_l}) \theta - (\mbox{\ttfamily helicity\_antisymm\_n}) \zeta)] \nonumber
+\end{eqnarray}
+
+{\ttfamily geometryScheme}==2: Use a 3-helicity model of the LHD standard configuration at {\ttfamily rN}=0.5.\\
+
+{\ttfamily geometryScheme}==3: Use a 4-helicity model of the LHD inward-shifted configuration at {\ttfamily rN}=0.5.\\
+
+{\ttfamily geometryScheme}==4: Use a 3-helicity model of the W7-X standard configuration at {\ttfamily rN}=0.5.\\
+
+{\ttfamily geometryScheme}==5: Read the {\ttfamily vmec wout} file specified in {\ttfamily equilibriumFile} below. The file can be
+       either ASCII format or \netCDF~format. (\sfincs~will auto-detect the format.).\\
+
+{\ttfamily geometryScheme}==11: Read the IPP {\ttfamily .bc} format Boozer-coordinate file specified in {\ttfamily equilibriumFile} below. 
+The file is assumed to be stellarator-symmetric.\\
+
+{\ttfamily geometryScheme}==12: Read the IPP {\ttfamily .bc} format Boozer-coordinate file specified in {\ttfamily equilibriumFile} below. 
+The file is assumed to be stellarator-asymmetric.
+}
+
+\myhrule
 
@@
 {3}
-{When {\ttfamily geometryScheme}==5, 11, or 12}
+{When {\ttfamily geometryScheme} == 1, 5, 11, or 12}
 {Which radial coordinate to use to specify the flux surface for a single calculation,
-or to specify the range of flux surfaces for a radial scan. See section \ref{sec:radialCoordinates}
+or to specify the range of flux surfaces for a radial scan. 
+(When {\ttfamily geometryScheme} == 2, 3, or 4, the flux surface used will be {\ttfamily rN}=0.5.)
+See section \ref{sec:radialCoordinates}
 for more information about radial coordinates.\\
@@
 {\ttfamily inputRadialCoordinate}==3: Use the flux surface specified by {\ttfamily rN\_wish}
-for a single run, and use the range specified by {\ttfamily rN\_min} and {\ttfamily rN\_max} for radial scans.
+for a single run, and use the range specified by {\ttfamily rN\_min} and {\ttfamily rN\_max} for radial scans. \\
+
+No matter which option you pick, the value of all 4 radial coordinates used will be saved in the output \HDF~file.
 }
+
+\myhrule
+
+\param{inputRadialCoordinateForGradients}
+{integer}
+{2}
+{Whenever {\ttfamily RHSMode}==1.}
+{Which radial coordinate is used to use to specify the input gradients of density, temperature, and electrostatic potential,
+i.e. which radial coordinate is used in the denominator of these derivatives.
+See section \ref{sec:radialCoordinates}
+for more information about radial coordinates.\\
+
+{\ttfamily inputRadialCoordinateForGradients}==0: Density gradients are specified by {\ttfamily dnHatdpsiHats},
+temperature gradients are specified by {\ttfamily dTHatdpsiHats}, a single $E_r$ is specified by {\ttfamily dPhiHatdpsiHat},
+and the range of an $E_r$ scan is specified by {\ttfamily dPhiHatdpsiHatMin-dPhiHatdpsiHatMax}.\\
+
+{\ttfamily inputRadialCoordinateForGradients}==1: Density gradients are specified by {\ttfamily dnHatdpsiNs},
+temperature gradients are specified by {\ttfamily dTHatdpsiNs}, a single $E_r$ is specified by {\ttfamily dPhiHatdpsiN},
+and the range of an $E_r$ scan is specified by {\ttfamily dPhiHatdpsiNMin-dPhiHatdpsiNMax}.\\
+
+{\ttfamily inputRadialCoordinateForGradients}==2: Density gradients are specified by {\ttfamily dnHatdrHats},
+temperature gradients are specified by {\ttfamily dTHatdrHats}, a single $E_r$ is specified by {\ttfamily dPhiHatdrHat},
+and the range of an $E_r$ scan is specified by {\ttfamily dPhiHatdrHatMin-dPhiHatdrHatMax}.\\
+
+{\ttfamily inputRadialCoordinateForGradients}==3: Density gradients are specified by {\ttfamily dnHatdrNs},
+temperature gradients are specified by {\ttfamily dTHatdrNs}, a single $E_r$ is specified by {\ttfamily dPhiHatdrN},
+and the range of an $E_r$ scan is specified by {\ttfamily dPhiHatdrNMin-dPhiHatdrNMax}.\\
+
+No matter which option you pick, the gradients with respect to all 4 radial coordinates will be saved in the output \HDF~file.
+}
+
+\myhrule
+
+\param{psiHat\_wish}
+{real}
+{-1}
+{Only when {\ttfamily inputRadialCoordinate} == 0 and {\ttfamily geometryScheme} == 1, 5, 11, or 12.}
+{Requested flux surface for the computation.  See section \ref{sec:radialCoordinates}
+for more information about radial coordinates.}
+
+\myhrule
+
+\param{psiN\_wish}
+{real}
+{0.25}
+{Only when {\ttfamily inputRadialCoordinate} == 1 and {\ttfamily geometryScheme} == 1, 5, 11, or 12.}
+{Requested flux surface for the computation.  See section \ref{sec:radialCoordinates}
+for more information about radial coordinates.}
+
+\myhrule
+
+\param{rHat\_wish}
+{real}
+{-1}
+{Only when {\ttfamily inputRadialCoordinate} == 2 and {\ttfamily geometryScheme} == 1, 5, 11, or 12.}
+{Requested flux surface for the computation.  See section \ref{sec:radialCoordinates}
+for more information about radial coordinates.}
+
+\myhrule
+
+\param{rN\_wish}
+{real}
+{0.5}
+{Only when {\ttfamily inputRadialCoordinate} == 3 and {\ttfamily geometryScheme} == 1, 5, 11, or 12.}
+{Requested flux surface for the computation.  See section \ref{sec:radialCoordinates}
+for more information about radial coordinates.}
+
+\myhrule
+
+\param{B0OverBBar}
+{real}
+{1.0}
+{Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily B0OverBBar} will be set according to the requested {\ttfamily geometryScheme}.}
+{Magnitude of the (0,0) Boozer harmonic of the magnetic field strength, normalized by $\bar{B}$.}
+
+\myhrule
+
+\param{GHat}
+{real}
+{??}
+{Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily GHat} will be set according to the requested {\ttfamily geometryScheme}.}
+{$G$ is $(c/2)\times$ the poloidal current outside the flux
+ surface. Equivalently, $G$ is the coefficient of $\nabla\zeta_B$ in the
+ covariant representation of $\vect{B}$ in terms of Boozer coordinates $(\theta_B,\zeta_B)$:
+\begin{equation}
+\label{eq:covariant}
+\vect{B}(\psi,\theta_B,\zeta_B) = \beta(\psi,\theta_B,\zeta_B)\nabla\psi + I(\psi)\nabla\theta_B + G(\psi)\nabla\zeta_B.
+\end{equation}
+{\ttfamily GHat} is $G$ normalized by $\bar{B}\bar{R}$.}
+
+\myhrule
+
+\param{IHat}
+{real}
+{0.0}
+{Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily IHat} will be set according to the requested {\ttfamily geometryScheme}.}
+{$I$ is $(c/2)\times$ the toroidal current inside the flux
+ surface. Equivalently, $I$ is the coefficient of $\nabla\theta_B$ in the
+ covariant representation of $\vect{B}$ in terms of Boozer coordinates $(\theta_B,\zeta_B)$ in (\ref{eq:covariant}).
+{\ttfamily IHat} is $I$ normalized by $\bar{B}\bar{R}$.}
+
+\myhrule
+
+\param{iota}
+{real}
+{0.0}
+{Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily iota} will be set according to the requested {\ttfamily geometryScheme}.}
+{Rotational transform, equivalent to $1/q$ where $q$ is the safety factor.}
+
+\myhrule
+
+\param{epsilon\_t}
+{real}
+{??}
+{Only when {\ttfamily geometryScheme} == 1.}
+{Toroidal variation in $B$, as defined by (\ref{eq:Bmodel}).}
+
+\myhrule
+
+\param{epsilon\_h}
+{real}
+{??}
+{Only when {\ttfamily geometryScheme} == 1.}
+{Helical variation in $B$, as defined by (\ref{eq:Bmodel}).}
+
+\myhrule
+
+\param{epsilon\_antisymm}
+{real}
+{0.0}
+{Only when {\ttfamily geometryScheme} == 1.}
+{Stellarator-antisymmetric variation in $B$, as defined by (\ref{eq:Bmodel}).}
+
+\myhrule
+
+\param{helicity\_l}
+{integer}
+{??}
+{Only when {\ttfamily geometryScheme} == 1.}
+{Poloidal mode number of the helical variation in $B$, as defined by (\ref{eq:Bmodel}).}
+
+\myhrule
+
+\param{helicity\_n}
+{integer}
+{??}
+{Only when {\ttfamily geometryScheme} == 1.}
+{Toroidal mode number of the helical variation in $B$, as defined by (\ref{eq:Bmodel}).}
+
+\myhrule
+
+\param{helicity\_antisymm\_l}
+{integer}
+{??}
+{Only when {\ttfamily geometryScheme} == 1.}
+{Poloidal mode number of the stellarator-antisymmetric variation in $B$, as defined by (\ref{eq:Bmodel}).}
+
+\myhrule
+
+\param{helicity\_antisymm\_n}
+{integer}
+{??}
+{Only when {\ttfamily geometryScheme} == 1.}
+{Toroidal mode number of the stellarator-antisymmetric variation in $B$, as defined by (\ref{eq:Bmodel}).
+Note that you can create an up-down asymmetric tokamak by setting {\ttfamily helicity\_antisymm\_n}=0, {\ttfamily epsilon\_h}=0,
+and {\ttfamily epsilon\_antisymm}$>$0.}
+
+\myhrule
+
+\param{psiAHat}
+{real}
+{??}
+{Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily psiAHat} will be set according to the requested {\ttfamily geometryScheme}.}
+{{\ttfamily psiAHat} = $\psi_a / (\bar{B} \bar{R}^2)$
+where $2 \pi \psi_a$ is the toroidal flux at the last closed flux surface.}
+
+\myhrule
+
+\param{aHat}
+{real}
+{??}
+{Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily aHat} will be set according to the requested {\ttfamily geometryScheme}.}
+{The effective minor radius at the last closed flux surface, in units of $\bar{R}$.
+The code only uses {\ttfamily aBar} for converting between the various radial coordinates in input and output quantities.
+}
+
+\myhrule
+
+\param{equilibriumFile}
+{string}
+{``''}
+{Only when {\ttfamily geometryScheme} == 5, 11, or 12.}
+{Filename from which to load the magnetic equilibrium, either in {\ttfamily vmec wout} ASCII or \netCDF~format, or IPP {\ttfamily .bc} format.}
+
+\myhrule
+
+\param{VMECRadialOption}
+{integer}
+{1}
+{Only when {\ttfamily geometryScheme} == 5.}
+{Controls whether the nearest available flux surface in the {\ttfamily vmec wout} file is used, or whether radial interpolation is applied to the \vmec~data
+to obtain the magnetic field components on the exact surface requested.\\
+
+{\ttfamily VMECRadialOption}=0: Use the exact {\ttfamily XXX\_wish} flux surface requested, by interpolating from the \vmec~radial grid.\\
+
+{\ttfamily VMECRadialOption}=1: Use a surface that may be slightly different from {\ttfamily XXX\_wish} to get the nearest available flux surface from \vmec's HALF grid.  
+   The components of $\vect{B}$ in \vmec~are stored on the half grid, so interpolation is then unnecessary.\\
+
+{\ttfamily VMECRadialOption}=2: Use a surface that may be slightly different from {\ttfamily XXX\_wish} to get the nearest available flux surface from \vmec's FULL grid.
+I'm not sure why you would want this, but the feature is implemented for completeness.
+}
+
+\myhrule
+
+\param{min\_Bmn\_to\_load}
+{real}
+{0.0}
+{Only when {\ttfamily geometryScheme} == 5, 11, or 12.}
+{Filters the magnetic field read from an input file.  Only Fourier modes $(m,n)$ for which $B_{m,n}$ is at least {\ttfamily min\_Bmn\_to\_load} will be included.}
+
+\section{The {\ttfamily speciesParameters}~namelist}
+
+\param{Zs}
+{1D array of reals}
+{1.0}
+{Always}
+{Charges of each species, in units of the proton charge $e$}
+
+\myhrule
+
+\param{mHats}
+{1D array of reals}
+{1.0}
+{Always}
+{Masses of each species, in units of the reference mass $\bar{m}$}
+
+\myhrule
+
+\param{nHats}
+{1D array of reals}
+{1.0}
+{Whenever {\ttfamily RHSMode} == 1}
+{Densities of each species, in units of the reference density $\bar{n}$}
+
+\myhrule
+
+\param{THats}
+{1D array of reals}
+{1.0}
+{Whenever {\ttfamily RHSMode} == 1}
+{Temperatures of each species, in units of the reference temperature $\bar{T}$}
+
+\myhrule
+
+\param{dnHatdpsiHats}
+{1D array of reals}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 0}
+{Radial density gradients of each species, with respect to the radial coordinate $\hat{\psi}$, normalized by the reference density $\bar{n}$.}
+
+\myhrule
+
+\param{dTHatdpsiHats}
+{1D array of reals}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 0}
+{Radial temperature gradients of each species, with respect to the radial coordinate $\hat{\psi}$, normalized by the reference temperature $\bar{T}$.}
+
+\myhrule
+
+\param{dnHatdpsiNs}
+{1D array of reals}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 1}
+{Radial density gradients of each species, with respect to the radial coordinate $\psi_N$, normalized by the reference density $\bar{n}$.}
+
+\myhrule
+
+\param{dTHatdpsiNs}
+{1D array of reals}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 1}
+{Radial temperature gradients of each species, with respect to the radial coordinate $\psi_N$, normalized by the reference temperature $\bar{T}$.}
+
+\myhrule
+
+\param{dnHatdrHats}
+{1D array of reals}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 2}
+{Radial density gradients of each species, with respect to the radial coordinate $\hat{r}$, normalized by the reference density $\bar{n}$.}
+
+\myhrule
+
+\param{dTHatdrHats}
+{1D array of reals}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 2}
+{Radial temperature gradients of each species, with respect to the radial coordinate $\hat{r}$, normalized by the reference temperature $\bar{T}$.}
+
+\myhrule
+
+\param{dnHatdrNs}
+{1D array of reals}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 3}
+{Radial density gradients of each species, with respect to the radial coordinate $r_N$, normalized by the reference density $\bar{n}$.}
+
+\myhrule
+
+\param{dTHatrNs}
+{1D array of reals}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 3}
+{Radial temperature gradients of each species, with respect to the radial coordinate $r_N$, normalized by the reference temperature $\bar{T}$.}
+
+\section{The {\ttfamily physicsParameters}~namelist}
 
@@
 
-\myhrule
-
 \section{\PETSc~commands}
@@
 {Dumps detailed information to stdout related to the linear solver.}
+
+\myhrule
+
+\PETScParam{-ksp\_gmres\_restart {\normalfont \ttfamily$<$integer$>$}}
+{After how many iterations will GMRES restart. Default is 2000. The convergence of GMRES slows every time a restart occurs, but restarts also free up memory.}
+
+\myhrule
+
+\PETScParam{-mat\_mumps\_icntl\_4 {\normalfont \ttfamily$<$integer$>$}}
+{How much diagnostic information will be printed by \mumps. Default is 0. Set to 2 or 3 to print out useful diagnostic information about the memory required for factorizing the preconditioner.}

### 2015-05-22 doc/manual/installation.tex
Added to user manual

@@
 Although \sfincs~can be run on a single processor, usually you want to run it in parallel.  
-In this case, you need {\ttfamily MPI}, and you need either \href{http://mumps-solver.org/}{\mumps}
+In this case, you need {\ttfamily MPI}, and you need at least one
+of the two libraries \href{http://mumps-solver.org/}{\mumps}
 or \href{http://crd-legacy.lbl.gov/~xiaoye/SuperLU/}{\superludist}.  (Note that \superludist is a parallel library which is different from the serial library {\ttfamily superlu}).
-Both \mumps~and \superludist~are parallelized libraries for direct solution of large sparse linear systems. 
+Both \mumps~and \superludist~are parallelized libraries for direct solution of large sparse linear systems,
+which \sfincs~uses to factorize the preconditioning matrix (discussed in secton \ref{sec:gmres}).
 \PETSc~has a built-in serial sparse direct linear solver,
@@
 \item Click on ``SSH keys'' on the left, and add an SSH key for the computer you wish to use. To do this, you may wish to read see the ``generating SSH keys'' guide which is linked to from that page.
-\item From a terminal command line in the computer you wish to use, enter {\ttfamily git clone git@github.com:landreman/sfincs.git} to download the repository.
+\item From a terminal command line in the computer you wish to use, enter\\
+{\ttfamily git clone git@github.com:landreman/sfincs.git}\\
+ to download the repository.
 \end{enumerate}
@@
 This variable is used in two ways. 
-First, {\ttfamily make} uses this variable to looks for the 
+First, {\ttfamily make} uses this variable to look for the 
 appropriate makefile in the
 \path{sfincs/fortran/version3/makefiles} directory.  Second, the {\ttfamily SFINCS\_SYSTEM} environment
-variable affects the behavior of
-\sfincsScan~in several ways, such as determining the command used to submit jobs to the system's queue.
+variable is used by
+\sfincsScan~to determine the command for submitting jobs to the system's queue.
 
@@
 \section{Setting up \sfincs~on a new system}
-If you are setting up \sfincs~ on a new system, one for which there is no file
+If you are setting up \sfincs~on a new system, one for which there is no file
 \path{sfincs/fortran/version3/makefiles/makefile.XXX},

### 2015-05-22 doc/manual/overview.tex
Added to user manual

@@
 
-The \sfincs~code is a free, open-source tool for solving neoclassical-type kinetic problems in nonaxisymmetric or axisymmetric plasmas
+The \sfincs~code is a freely available, open-source tool for solving neoclassical-type kinetic problems in nonaxisymmetric or axisymmetric plasmas
 with nested toroidal flux surfaces.
@@
 
-This document discusses use and operation of the code.  For more details about the specific equations implemented,
+This document discusses the practical use and operation of the code.  For more details about the specific equations implemented,
 see the version 3 technical documentation available in the \path{sfincs/docs} directory.
@@
 
+Often, the limiting factor for \sfincs~is the ability of \mumps~or \superludist~to factorize the preconditioner matrix.
+You may therefore find it useful to see the control parameters and error codes in the \mumps~user manual:
+\url{http://mumps.enseeiht.fr/doc/userguide_5.0.0.pdf}.
+
 \section{GMRES/KSP and preconditioning}
+\label{sec:gmres}
 
@@
 
-\section{Bugs and feedback}
+\section{Input and Output}
+
+The input parameters for a \sfincs~computation are specified in a file named {\ttfamily input.namelist}.
+This file contains both information for the fortran part of \sfincs~(in standard fortran namelist format),
+as well as special lines beginning with {\ttfamily !ss} which are read by \sfincsScan.
+The variables which can be specified in {\ttfamily input.namelist} are detailed in chapter \ref{ch:input}.
+For scans over minor radius, an additional file named {\ttfamily profiles} is used to specify the profiles of density and temperature
+for each species, as well as the range of radial electric field to consider.
+
+The output from a single \sfincs~computation is saved in \HDF~format in the file {\ttfamily sfincsOutput.h5}.
+To browse this file you can enter {\ttfamily h5dump sfincsOutput.h5|less} from the command line.
+Every array saved in this file is annotated with strings that describe the array dimensions,
+for example \Ntheta$\times$ \Nzeta. One of the array dimensions may be {\ttfamily iteration},
+which can either indicate the iteration of the Newton solver for a nonlinear calculation,
+or which right-hand side vector was used when computing a transport matrix.
+Many of the variables in the output file are also annotated with
+text that describes their meaning and normalization.
+
+\section{Questions, Bugs, and Feedback}
 

### 2015-05-22 doc/manual/runs.tex
Added to user manual

@@
 $\bar{T}$ = reference temperature in energy units, typically 1 eV or 1 keV.\\
-$\bar{v} = \sqrt{2 \bar{T} / \bar{m}}$ = reference speed\\
+$\bar{v} = \sqrt{2 \bar{T} / \bar{m}}$ = thermal speed at the reference temperature and mass\\
 $\bar{\Phi}$ = reference electrostatic potential, typically 1 V or 1 kV.\\
@@
 A variety of flux-surface label coordinates are used in other codes and in the literature.
-One common choice (used in {\ttfamily vmec}) is $\psi_N$, the toroidal flux normalized to its
+One common choice (used in \vmec) is $\psi_N$, the toroidal flux normalized to its
 value at the last closed flux surface.  Another common choice is an ``effective normalized minor radius''
@@
 For maximum flexibility, \sfincs~permits any of these four radial coordinates to be used, and different radial
-coordinates can be used in different parts of a given computation.  Output quantities which depend
+coordinates can be used in different aspects of a given computation.  Output quantities which depend
 on the radial coordinate, such as radial fluxes, are often given with respect to all radial coordinates.
+In \sfincs, the four radial coordinates are named as follows:\\
 
-To specify a run, you can make several independent choices for radial coordinates.  One parameter
-you select is {\ttfamily inputRadialCoordinate} in the {\ttfamily geometryParameters} namelist.
-This parameter controls which coordinate is used to specify the flux surface. The possible values of {\ttfamily inputRadialCoordinate} are:
+{\setlength{\parindent}{0cm}
 
-0: Use {\ttfamily psiHat}, the toroidal flux (divided by $2\pi$) normalized by $\bar{B}\bar{R}^2$.\\
+{\ttfamily psiHat} = $\hat\psi$ is the toroidal flux (divided by $2\pi$), normalized by $\bar{B}\bar{R}^2$.\\
 
-1: Use {\ttfamily psiN}, the toroidal flux normalized by its value at the last closed flux surface.\\
+{\ttfamily psiN} = $\psi_N$ is the toroidal flux normalized by its value at the last closed flux surface.\\
 
-2: Use {\ttfamily rHat}, defined as {\ttfamily aHat}$\sqrt{\mbox{\ttfamily psiN}}$, where {\ttfamily aHat} is an effective minor radius of the last closed flux surface normalized by $\bar{R}$.\\
+{\ttfamily rHat} = $\hat{r}$ is defined as {\ttfamily aHat}$\sqrt{\mbox{\ttfamily psiN}}$, where {\ttfamily aHat} is an effective minor radius of the last closed flux surface normalized by $\bar{R}$.\\
 
-3: Use {\ttfamily rN}, defined as $\sqrt{\mbox{\ttfamily psiN}}$.\\
+{\ttfamily rN} = $r_N$ is defined as $\sqrt{\mbox{\ttfamily psiN}}$.\\
+
+}
+
+These four radial coordinates are identified by the numbers 0, 1, 2, and 3 respectively.
+
+When setting up a run, you can make several independent choices for radial coordinates.  One parameter
+you select is {\ttfamily inputRadialCoordinateForGradients} in the {\ttfamily geometryParameters} namelist.
+This parameter controls which coordinate is used to specify the gradients. The possible values of {\ttfamily inputRadialCoordinateForGradients} are:\\
+
+{\setlength{\parindent}{0cm}
+
+0: Use derivatives with respect to {\ttfamily psiHat}: Density gradients are specified using {\ttfamily dnHatdpsiHats}, 
+temperature gradients are specified using {\ttfamily dTHatdpsiHats},  a single $E_r$ is specified using
+{\ttfamily dPhiHatdpsiHat}, and a range of $E_r$ for a scan is specified using {\ttfamily dPhiHatdpsiHatMin}-{\ttfamily dPhiHatdpsiHatMax}.
+\\
+
+1: Use derivatives with respect to {\ttfamily psiN}: Density gradients are specified using {\ttfamily dnHatdpsiNs}, 
+temperature gradients are specified using {\ttfamily dTHatdpsiNs},  a single $E_r$ is specified using
+{\ttfamily dPhiHatdpsiN}, and a range of $E_r$ for a scan is specified using {\ttfamily dPhiHatdpsiNMin}-{\ttfamily dPhiHatdpsiNMax}.
+\\
+
+2: Use derivatives with respect to {\ttfamily rHat}: Density gradients are specified using {\ttfamily dnHatdrHats}, 
+temperature gradients are specified using {\ttfamily dTHatdrHats},  a single $E_r$ is specified using
+{\ttfamily dPhiHatdrHat}, and a range of $E_r$ for a scan is specified using {\ttfamily dPhiHatdrHatMin}-{\ttfamily dPhiHatdrHatMax}.
+\\
+
+3: Use derivatives with respect to {\ttfamily rN}: Density gradients are specified using {\ttfamily dnHatdrNs}, 
+temperature gradients are specified using {\ttfamily dTHatdrNs},  a single $E_r$ is specified using
+{\ttfamily dPhiHatdrN}, and a range of $E_r$ for a scan is specified using {\ttfamily dPhiHatdrNMin}-{\ttfamily dPhiHatdrNMax}.
+\\
+
+}
+
+Another choice involving radial coordinates is how to specify the flux surface for the computation.
+This choice is made using the parameter {\ttfamily inputRadialCoordinate} in the {\ttfamily geometryParameters}
+namelist, which is again an integer from 0 to 3, and this parameter need not be the same as {\ttfamily inputRadialCoordinateForGradients}.
+An extra complication with specifying the flux surface is that the magnetic equilibrium file will contain data on a finite number of surfaces,
+and you may wish to use one of these surfaces.  For this reason, the parameters for specifying the flux surface have {\ttfamily \_wish}
+appended to the name. In other words, the allowed values for {\ttfamily inputRadialCoordinate} are:\\
+
+{\setlength{\parindent}{0cm}
+
+0: Specify the flux surface using {\ttfamily psiHat\_wish}.\\
+
+1: Specify the flux surface using {\ttfamily psiN\_wish}.\\
+
+2: Specify the flux surface using {\ttfamily rHat\_wish}.\\
+
+3: Specify the flux surface using {\ttfamily rN\_wish}.\\
+
+}
+
+When using {\ttfamily geometryScheme} == 11 or 12, \sfincs~will always shift the ``wish'' value so it matches an available surface in the magnetic equilibrium file.
+For {\ttfamily geometryScheme} == 5, the {\ttfamily VMECRadialOption} parameter lets you can choose whether to shift to the nearest surface in the magnetic equilibrium file,
+or to interpolate the \vmec~data onto the exact value of radius you specify.
+
+If you perform a radial scan, then there is a third choice you can make: which radial coordinate to use in the {\ttfamily profiles} file.
+This choice is made with an integer 0, 1, 2, or 3 in the first non-comment line of the {\ttfamily profiles} file.
+The radial coordinate used in the {\ttfamily profiles} file need not be the same as either
+{\ttfamily inputRadialCoordinate} or {\ttfamily inputRadialCoordinateForGradients}.
+Note however that the maximum and minimum radial electric field specified in the {\ttfamily profiles}
+file must be defined as the derivative of the electrostatic potential with respect to the radial coordinate {\ttfamily inputRadialCoordinateForGradients}.
+
+For more details about the behavior of {\ttfamily inputRadialCoordinate}, {\ttfamily inputRadialCoordinateForGradients}, and {\ttfamily VMECRadialOption},
+see section \ref{sec:geometryParameters}.
+
+\section{Trajectory models}
+
+As discussed in \cite{sfincsPaper},
+one of the capabilities of \sfincs~is to compare various models for the terms in the kinetic equation involving $E_r$.
+These variations of the kinetic equation are called ``trajectory models'' in \cite{sfincsPaper}.
+The relevant terms in the kinetic equation can be turned off and on by certain Boolean parameters in the {\ttfamily physicsParameters} namelist.
+The models described in \cite{sfincsPaper} are selected as follows:\\
+
+{\setlength{\parindent}{0cm}
+
+\underline{Full trajectories:}\\
+{\ttfamily 
+includeXDotTerm = .true.\\
+includeElectricFieldTermInXiDot = .true.\\
+useDKESExBDrift = .false.\\
+}
+
+\underline{Partial trajectories:}\\
+{\ttfamily
+includeXDotTerm = .false.\\
+includeElectricFieldTermInXiDot = .false.\\
+useDKESExBDrift = .false.\\
+}
+
+\underline{DKES trajectories:}\\
+{\ttfamily
+includeXDotTerm = .false.\\
+includeElectricFieldTermInXiDot = .false.\\
+useDKESExBDrift = .true.\\
+}
+}
 

### 2015-05-23 doc/manual/inputParameters.tex
Added to user manual

@@
 Then we list some of the command-line flags associated with \PETSc~which can be useful.
+Note that all parameters in {\ttfamily input.namelist}, both for \sfincs~and \sfincsScan,
+are case-insensitive.
 
@@
 {Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily B0OverBBar} will be set according to the requested {\ttfamily geometryScheme}.}
-{Magnitude of the (0,0) Boozer harmonic of the magnetic field strength, normalized by $\bar{B}$.}
+{Magnitude of the (0,0) Boozer harmonic of the magnetic field strength (equivalent to $\left< B^3\right>/\left<B^2\right>$), normalized by $\bar{B}$.}
 
@@
 {real}
-{??}
+{3.7481}
 {Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily GHat} will be set according to the requested {\ttfamily geometryScheme}.}
@@
 {real}
-{0.0}
+{0.4542}
 {Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily iota} will be set according to the requested {\ttfamily geometryScheme}.}
-{Rotational transform, equivalent to $1/q$ where $q$ is the safety factor.}
+{Rotational transform (rationalized), equivalent to $1/q$ where $q$ is the safety factor.}
 
@@
 {real}
-{??}
+{-0.07053}
 {Only when {\ttfamily geometryScheme} == 1.}
@@
 {real}
-{??}
+{0.05067}
 {Only when {\ttfamily geometryScheme} == 1.}
@@
 {integer}
-{??}
+{2}
 {Only when {\ttfamily geometryScheme} == 1.}
@@
 {integer}
-{??}
+{10}
 {Only when {\ttfamily geometryScheme} == 1.}
@@
 {integer}
-{??}
+{1}
 {Only when {\ttfamily geometryScheme} == 1.}
@@
 {integer}
-{??}
+{0}
 {Only when {\ttfamily geometryScheme} == 1.}
@@
 {real}
-{??}
+{0.15596}
 {Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily psiAHat} will be set according to the requested {\ttfamily geometryScheme}.}
@@
 {real}
-{??}
+{0.5585}
 {Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily aHat} will be set according to the requested {\ttfamily geometryScheme}.}
@@
 
+\param{Delta}
+{real}
+{4.5694e-3}
+{Always}
+{Roughly speaking, {\ttfamily Delta} is $\rho_*$ at the reference parameters. The precise definition is
+\begin{eqnarray}
+\mbox{\ttfamily Delta} 
+&=& \frac{c \bar{m} \bar{v}}{e \bar{B} \bar{R}} \;\;\; \mbox{(Gaussian units)} \\
+&=&  \frac{\bar{m} \bar{v}}{e \bar{B} \bar{R}} \;\;\; \mbox{(SI units)}, \nonumber
+\end{eqnarray}
+where $c$ is the speed of light,
+$e$ is the proton mass,
+and quantities with a bar are the normalization reference parameters discussed in section \ref{sec:normalizations}.
+The default value {\ttfamily Delta} = 4.5694e-3 corresponds to $\bar{B}$ = 1 Tesla, $\bar{R}$ = 1 meter,
+$\bar{m}$ = proton mass, and $\bar{T}$ = 1 keV.}
+
+\myhrule
+
+\param{alpha}
+{real}
+{1.0}
+{Whenever {\ttfamily RHSMode} == 1 and $E_r$ is nonzero.}
+{{\ttfamily alpha} $= e \bar{\Phi}/ \bar{T}$ (both Gaussian and SI units) where $e$ is the proton mass,
+and $\bar{\Phi}$ and $\bar{T}$ are the normalization reference parameters discussed in section \ref{sec:normalizations}.
+The default value {\ttfamily alpha} = 1.0 corresponds to $\bar{T}$ = 1 keV and $\bar{\Phi}$ = 1 kV.
+The default value {\ttfamily alpha} = 1.0 also corresponds to $\bar{T}$ = 1 eV and $\bar{\Phi}$ = 1 V.
+}
+
+\myhrule
+
+\param{EParallelHat}
+{real}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1}
+{Inductive parallel electric field:
+\begin{equation}
+\mbox{\ttfamily EParallelHat} = \left< \vect{E}\cdot\vect{B}\right> \frac{\bar{R}}{\bar{\Phi}\bar{B}}
+\end{equation}
+(in both Gaussian and SI units) where $\left< \ldots \right>$ denotes a flux surface average,
+$\vect{E}$ and $\vect{B}$ are the electric and magnetic field vectors, and 
+quantities with a bar are the normalization reference parameters discussed in section \ref{sec:normalizations}.}
+
+\myhrule
+
+\param{dPhiHatdpsiHat}
+{real}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 0}
+{The derivative of the electrostatic potential with respect to the radial coordinate $\hat{\psi}$,
+i.e. the radial electric field up to a constant.
+Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
+will be used, depending on\\
+{\ttfamily inputRadialCoordinateForGradients}.
+}
+
+\myhrule
+
+\param{dPhiHatdpsiN}
+{real}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 1}
+{The derivative of the electrostatic potential with respect to the radial coordinate $\psi_N$,
+i.e. the radial electric field up to a constant.
+Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
+will be used, depending on\\
+{\ttfamily inputRadialCoordinateForGradients}.
+}
+
+\myhrule
+
+\param{dPhiHatdrHat}
+{real}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 2}
+{The derivative of the electrostatic potential with respect to the radial coordinate $\hat{r}$,
+i.e. the radial electric field up to a constant.
+Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
+will be used, depending on\\
+{\ttfamily inputRadialCoordinateForGradients}.
+}
+
+\myhrule
+
+\param{dPhiHatdrN}
+{real}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 3}
+{The derivative of the electrostatic potential with respect to the radial coordinate $r_N$,
+i.e. the radial electric field up to a constant.
+Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
+will be used, depending on\\
+{\ttfamily inputRadialCoordinateForGradients}.
+}
+
+\myhrule
+
+\param{nu\_n}
+{real}
+{8.330e-3}
+{Whenever {\ttfamily RHSMode} == 1}
+{Dimensionless collisionality at the reference parameters:
+\begin{equation}
+\mbox{\ttfamily nu\_n} = \bar{\nu} \frac{\bar{R}}{\bar{v}},
+\end{equation}
+where $\bar{R}$ and $\bar{v}$ are the normalization reference parameters discussed in section \ref{sec:normalizations},
+and $\bar{\nu}$ is the dimensional collision frequency at the reference parameters. This frequency is defined as
+\begin{eqnarray}
+\bar{\nu}
+&=& \frac{4\sqrt{2\pi} \bar{n} e^4 \ln\Lambda}{3 (4\pi\epsilon_0)^2 \sqrt{\bar{m}} \bar{T}^{3/2}} \;\;\; \mbox{(SI units}) \\
+&=& \frac{4\sqrt{2\pi} \bar{n} e^4 \ln\Lambda}{3 \sqrt{\bar{m}} \bar{T}^{3/2}} \;\;\; \mbox{(Gaussian units}) \nonumber
+\end{eqnarray}
+where $e$ is the proton charge, $\bar{n}$, $\bar{m}$, and $\bar{T}$ are the normalization reference parameters discussed in section \ref{sec:normalizations},
+and $\ln\Lambda$ is the Coulomb logarithm.
+The default value {\ttfamily nu\_n} = 8.330e-3 corresponds to $\bar{R}$ = 1 meter,
+$\bar{m}$ = proton mass, $\bar{n}$ = $10^{20}$ m$^{-3}$, $\bar{T}$ = 1 keV, and $\ln\Lambda = 17$.
+}
+
+\myhrule
+
+\param{nuPrime}
+{real}
+{1.0}
+{Only when {\ttfamily RHSMode} == 3.}
+{Dimensionless collisionality used in place of {\ttfamily nHats}, {\ttfamily THats}, {\ttfamily mHats}, {\ttfamily Zs}, and {\ttfamily nu\_n} for computing monoenergetic transport coefficients.
+See section \ref{sec:monoenergetic} for more details.}
+
+\myhrule
+
+\param{EStar}
+{real}
+{0.0}
+{Only when {\ttfamily RHSMode} == 3.}
+{Normalized radial electric field used in place of {\ttfamily dPhiHatdXXX} for computing monoenergetic transport coefficients.
+See section \ref{sec:monoenergetic} for more details.}
+
+\myhrule
+
+\param{collisionOperator}
+{integer}
+{0}
+{Always}
+{Which collision operator to use:\\
+
+{\setlength{\parindent}{0cm}
+
+{\ttfamily collisionOperator} = 0: Full linearized Fokker-Planck operator.\\
+
+{\ttfamily collisionOperator} = 1: Pitch-angle scattering operator (with no momentum-conserving field term).
+}}
+
+\myhrule
+
+\param{constraintScheme}
+{integer}
+{-1}
+{Always}
+{Controls a small number of extra rows and columns of the system matrix
+which (1) eliminate the null space of the matrix, and (2) ensure that a steady-state solution
+to the kinetic equation exists even when phase-space volume and/or energy are not conserved.
+These issues are detailed in section III of Ref \cite{sfincsPaper}.
+\\
+
+{\setlength{\parindent}{0cm}
+
+{\ttfamily constraintScheme} = -1: Automatic. If {\ttfamily collisionOperator}==0 then {\ttfamily constraintScheme} will be set to 1, 
+otherwise {\ttfamily constraintScheme} will be set to 2.\\
+
+{\ttfamily constraintScheme} = 0: No constraints.\\
+
+{\ttfamily constraintScheme} = 1: 2 constraints per species: $\left<n_1\right>=0$ and $\left<p_1\right>=0$.\\
+
+{\ttfamily constraintScheme} = 2: \Nx~constraints per species: $\left< f(L=0)\right>=0$ at each $x$.\\
+
+You should set {\ttfamily constraintScheme} to -1 unless you know what you are doing.
+}}
+
+\myhrule
+
+\param{includeXDotTerm}
+{Boolean}
+{\true}
+{Whenever {\ttfamily RHSMode} $<3$ and the radial electric field is nonzero.}
+{Whether or not to include the term in the kinetic equation corresponding to a change
+in speed proportional to the radial electric field.  This term is given by $\dot{x}$ in
+equation (17) of \cite{sfincsPaper}.}
+
+\myhrule
+
+\param{includeElectricFieldTermInXiDot}
+{Boolean}
+{\true}
+{Whenever {\ttfamily RHSMode} $<3$ and the radial electric field is nonzero.}
+{Whether or not to include the term in the kinetic equation corresponding to a change
+in pitch angle $\xi$ proportional to the radial electric field.  This term is given by the last line of
+equation (17) of \cite{sfincsPaper}.}
+
+\myhrule
+
+\param{useDKESExBDrift}
+{Boolean}
+{\false}
+{Whenever {\ttfamily RHSMode} $<3$ and the radial electric field is nonzero.}
+{If true, the $\vect{E}\times\vect{B}$ drift term 
+multiplying $\partial f/\partial \theta$ and  $\partial f/\partial \zeta$
+is taken to be $\vect{E}\times\vect{B}\cdot \nabla (\theta \;\mbox{or} \;\zeta) / \left< B^2 \right>$
+instead of 
+ $\vect{E}\times\vect{B}\cdot \nabla (\theta \;\mbox{or} \;\zeta) /B^2$.
+}
+
+\myhrule
+
+\param{include\_fDivVE\_term}
+{Boolean}
+{\false}
+{Never}
+{Obsolete}
+
+\myhrule
+
+\param{includePhi1}
+{Boolean}
+{\false}
+{Whenever {\ttfamily RHSMode} == 1.}
+{If false, no terms involving $\Phi_1 = \Phi-\left<\Phi\right>$ are included in the kinetic equation,
+and the quasineutrality equation is not solved.  If true, then terms involving $\Phi_1$ are included
+in the kinetic equation, and the quasineutrality equation is solved at each point on the flux surface.
+In this latter case, many more quantities are computed and saved in the output file, such as radial fluxes
+associated with the radial $\vect{E}\times\vect{B}$ drift.}
+
+\myhrule
+
+\param{nonlinear}
+{Boolean}
+{\false}
+{Whenever {\ttfamily RHSMode} == 1.}
+{If true, certain terms that are nonlinear in the unknowns will be included in the kinetic equation.
+Newton's method will be used to solve the nonlinear system, meaning that the usual linear solve in \sfincs~must
+be iterated several times.
+Running with {\ttfamily nonlinear}=\true requires {\ttfamily includePhi1}=\true
+}
+
+\myhrule
+
+\param{includeTemperatureEquilibrationTerm}
+{Boolean}
+{\false}
+{Whenever {\ttfamily RHSMode} == 1.}
+{When true, the term $C_{ab}[ f_{Ma}, f_{Mb}]$ is included in the kinetic equation,
+i.e. collisions between the leading-order Maxwellians of different species.  This term is nonzero
+when the temperature is not the same for all species.  The resulting contribution to the non-Maxwellian
+distribution function is isotropic and so does not directly give any parallel or radial transport.}
+
+\myhrule
+
+\param{magneticDriftScheme}
+{integer}
+{0}
+{Whenever {\ttfamily RHSMode} == 1.}
+{This variable controls the poloidal and magnetic drifts, and does not affect the radial magnetic drift.
+THIS FUNCTIONALITY IS WORK IN PROGRESS, AND RESULTS ARE BUGGY WHEN THIS VARIABLE IS NONZERO.\\
+
+{\ttfamily magneticDriftScheme} = 0:  No poloidal or toroidal magnetic drift.\\
+
+{\ttfamily magneticDriftScheme} = 1: Use the grad-B and curvature drift, plus the parallel velocity correction $v_{\perp}^2/(2 \Omega_c) \vect{b} \vect{b} \cdot \nabla \times \vect{b}$.\\
+
+{\ttfamily magneticDriftScheme} = 2:   Use the magnetic drift $v_m = (v_{||} / \Omega_c) \nabla (v_{||} \vect{b})$.
+}
+
+\section{The {\ttfamily resolutionParameters}~namelist}
+
+In this namelist, there are 4 parameters you definitely need to be aware of and adjust: \Ntheta, \Nzeta, \Nxi, and \Nx.  
+See chapter \ref{ch:resolution} for details. You may or may not
+need to adjust {\ttfamily solverTolerance}.  The other parameters in this namelist almost never need to be adjusted.
+
+\myhrule
+
+\param{Ntheta}
+{integer}
+{15}
+{Always}
+{Number of grid points in the poloidal angle. 
+This parameter should be odd; see {\ttfamily forceOddNthetaAndNzeta} in this namelist.
+Memory and time requirements DO depend strongly on this parameter.
+For stellarator calculations, this parameter can usually be in the range 15-25.
+For tokamak calculations at low collisionality, the value of this parameter may need to be higher.}
+
+\myhrule
+
+\param{Nzeta}
+{integer}
+{15}
+{Always}
+{Number of grid points in the toroidal angle (per identical segment of the stellarator.)  
+This parameter should be odd; see {\ttfamily forceOddNthetaAndNzeta} in this namelist.
+Memory and time requirements DO depend strongly on this parameter.
+Set this parameter to 1 for a tokamak calculation.
+For stellarator calculations, the value of this parameter required for convergence depends strongly on
+the collisionality. At high collisionality, this parameter can be several 10s, depending
+on the complexity of $B(\theta,\zeta)$. At low collisionality, this parameter may need to be many 10s or
+even $>$ 100 for convergence.
+}
+
+\myhrule
+
+\param{Nxi}
+{integer}
+{16}
+{Always}
+{Number of Legendre polynomials used to represent the pitch-angle dependence of the distribution function.
+Memory and time requirements DO depend strongly on this parameter.
+The value of this parameter required for convergence depends strongly on
+the collisionality. At high collisionality, this parameter can be as low
+as 5. At low collisionality, this parameter may need to be many 10s or
+even $>$ 100 for convergence.}
+
+\myhrule
+
+\param{Nx}
+{integer}
+{5}
+{Always}
+{Number of grid points in energy used to represent the distribution function.
+Memory and time requirements DO depend strongly on this parameter.
+This parameter almost always needs to be at least 5.
+Usually a value in the range 5-8 is plenty for convergence, though in exceptional circumstances
+you may need to go up to 10-15.
+}
+
+\myhrule
+
+\param{solverTolerance}
+{real}
+{1e-6}
+{Whenever {\ttfamily useIterativeLinearSolver} == \true}
+{
+Tolerance used to define convergence of the Krylov solver.
+This parameter does not affect memory requirements but it does affect the
+time required for solution somewhat.
+Occasionally you may want to ease this tolerance to 1e-5 so fewer iterations of the Krylov solver are needed.
+}
+
+\myhrule
+
+\param{NL}
+{integer}
+{4}
+{Whenever {\ttfamily collisionOperator} == 0.}
+{Number of Legendre polynomials used to represent the Rosenbluth
+potentials. This number can basically always be 4,
+since results barely change when {\ttfamily NL} is increased above this value.
+Memory and time requirements do NOT depend strongly on this parameter.
+}
+
+\myhrule
+
+\param{NxPotentialsPerVth}
+{real}
+{40.0}
+{Only when {\ttfamily collisionOperator} == 0 and {\ttfamily xGridScheme} $<$ 5.
+Since {\ttfamily xGridScheme} = 5 is recommended, this parameter is basically obsolete.}
+{
+Number of grid points in energy used to represent the Rosenbluth potentials
+for the original implementation of the Fokker-Planck operator described in \cite{speedGrids}.
+Memory and time requirements do NOT depend strongly on this parameter.}
+
+\myhrule
+
+\param{xMax}
+{real}
+{5.0}
+{Only when {\ttfamily collisionOperator} == 0 and {\ttfamily xGridScheme} $<$ 5.
+Since {\ttfamily xGridScheme} = 5 is recommended, this parameter is basically obsolete.}
+{
+Maximum normalized speed for the Rosenbluth potential grid
+for the original implementation of the Fokker-Planck operator described in \cite{speedGrids}.
+Memory and time requirements do NOT depend strongly on this parameter.}
+
+\myhrule
+
+\param{forceOddNthetaAndNzeta}
+{Boolean}
+{\true}
+{Always}
+{
+If true, 1 is added to \Ntheta~any time a run is attempted with even \Ntheta,
+and 1 is added to \Nzeta~any time a run is attempted with even \Nzeta.
+When false, the even and odd grid points are effectively decoupled so results are unstable.
+This parameter should be true unless you know what you are doing.
+}
+
 \section{Directives for \sfincsScan}
@@
 In this type of scan, the same radial electric field is used at every radius.
-See {\ttfamily utils/profiles.XXX} for examples.\\
+See \path{sfincs/fortran/utils/profiles.XXX} for examples.\\
 
 {\ttfamily scanType=5}:  Scan radius, and at each radius, scan $E_r$.  Density and temperature profiles are again
- taken from the {\ttfamily profiles} file; see {\ttfamily utils/profiles.XXX} for examples.\\
+ taken from the {\ttfamily profiles} file; see \path{sfincs/fortran/utils/profiles.XXX} for examples.\\
 
-{\ttfamily scanType=21}: Read in a list of requested runs from a file {\ttfamily runspec.dat}. See {\ttfamily utils/sfincsScan\_21} for an example file.
+{\ttfamily scanType=21}: Read in a list of requested runs from a file {\ttfamily runspec.dat}. See \path{sfincs/fortran/utils/sfincsScan_21} for an example file.
 }

### 2015-05-23 doc/manual/overview.tex
Added to user manual

@@
 \url{http://mumps.enseeiht.fr/doc/userguide_5.0.0.pdf}.
+
+\section{Geometry options}
+In \sfincs, a variety of options are available for the magnetic field geometry.  The geometry can be read directly
+from a {\ttfamily vmec wout}
+file, or from the {\ttfamily .bc} format Boozer-coordinate data files used at the Max Planck Institute for Plasma Physics (IPP).
+A general analytic model for the magnetic field is also available, given by equation (\ref{eq:Bmodel}),
+as are several analytic models for LHD and W7-X in which 3 or 4 Fourier components are retained.
+For more details about geometry options in \sfincs, see section \ref{sec:geometryParameters}.
 

### 2015-05-23 doc/manual/resolution.tex
Added to user manual

@@
 \chapter{Numerical resolution parameters}
+\label{ch:resolution}
 
@@
 There are several resolution parameters which are almost never the limiting factor
-for convergence.  You can almost always use {\ttfamily NL=4}, 
-{\ttfamily solverTolerance} = $10^{-5}$ or $10^{-6}$,
+for convergence, and so which almost never need to be adjusted.
+These parameters and good values for them are {\ttfamily NL=4}, 
+{\ttfamily solverTolerance} = $10^{-6}$,
 {\ttfamily xMax=5.0},

### 2015-05-23 doc/manual/runs.tex
Added to user manual

@@
 \section{Normalizations}
+\label{sec:normalizations}
 
@@
 \section{Issues with running on 1 processor}
+
+\section{Monoenergetic transport coefficients}
+\label{sec:monoenergetic}
+
+By setting {\ttfamily RHSMode=3}, \sfincs~can be run in a mode
+where it solves the same kinetic equation (prior to discretization) as {\ttfamily dkes}
+and other monoenergetic codes.
+When {\ttfamily RHSMode=3}, the values of {\ttfamily Zs}, {\ttfamily THats}, {\ttfamily nHats},
+{\ttfamily mHats}, {\ttfamily nu\_n}, and {\ttfamily dPhiHatdXXX} are all ignored.
+Instead, the collisionality is set by {\ttfamily nuPrime}, and the radial electric field is set
+by {\ttfamily EStar}.  The first of these quantities is the dimensionless collisionality
+\begin{equation}
+\mbox{\ttfamily nuPrime} = \frac{(G+\iota I) \nu}{v B_0}
+\end{equation}
+where $G$ and $I$ are defined in (\ref{eq:covariant}), $\iota=1/q$ is the rotational transform,
+$v$ is the speed at which the monoenergetic calculation is being performed, and $B_0$ is the (0,0) Fourier harmonic of $B$
+with respect to the Boozer poloidal and toroidal angles.
+\todo{How is $\nu$ defined here?}
+The normalized radial electric field is
+\begin{equation}
+\mbox{\ttfamily EStar} = \frac{cG}{\iota v B_0} \frac{d\Phi}{d\psi}
+\end{equation}
+(Gaussian units).
+When {\ttfamily RHSMode} == 1, {\ttfamily nuPrime} and {\ttfamily EStar} are ignored.
+\todo{Should be change the behavior of RHSMode=2 so it uses nuPrime and EStar instead of nu\_n?}

### 2015-05-23 fortran/version3/globalVariables.F90

@@ -23,12 +23,12 @@ module globalVariables
   ! ********************************************************
   ! ********************************************************
 
-  logical :: saveMatlabOutput, saveMatricesAndVectorsInBinary
-  character(len=200) :: MatlabOutputFilename
-  character(len=200) :: binaryOutputFilename
-  character(len=200) :: outputFilename
-  logical :: solveSystem
-  integer :: RHSMode, whichRHS
+  logical :: saveMatlabOutput = .false., saveMatricesAndVectorsInBinary = .false.
+  character(len=200) :: MatlabOutputFilename = "sfincsMatrices"
+  character(len=200) :: binaryOutputFilename = "sfincsBinary"
+  character(len=200) :: outputFilename = "sfincsOutput.h5"
+  logical :: solveSystem = .true.
+  integer :: RHSMode = 1, whichRHS
   logical :: isAParallelDirectSolverInstalled
 
   ! ********************************************************
@@ -39,18 +39,18 @@ module globalVariables
   ! ********************************************************
   ! ********************************************************
 
-  integer :: geometryScheme
-  PetscScalar ::  GHat, IHat, iota, B0OverBBar, psiAHat
-  PetscScalar :: epsilon_t, epsilon_h, epsilon_antisymm
-  integer :: NPeriods, helicity_l, helicity_n, helicity_antisymm_l, helicity_antisymm_n
-  character(len=200) :: equilibriumFile
-  PetscScalar :: min_Bmn_to_load, dGdpHat, aHat
-  PetscScalar :: psiHat_wish, psiN_wish, rHat_wish, rN_wish
+  integer :: geometryScheme = 1
+  PetscScalar ::  GHat = 3.7481d+0, IHat = 0.0, iota = 0.4542d+0, B0OverBBar = 1.0, psiAHat = 0.15596d+0
+  PetscScalar :: epsilon_t = -0.07053d+0, epsilon_h = 0.05067d+0, epsilon_antisymm = 0.0
+  integer :: NPeriods = 0, helicity_l = 2, helicity_n = 10, helicity_antisymm_l = 1, helicity_antisymm_n = 0
+  character(len=200) :: equilibriumFile = ""
+  PetscScalar :: min_Bmn_to_load = 0.0, dGdpHat, aHat = 0.5585d+0
[…]

### 2015-05-23 doc/manual/inputParameters.tex
Added to user manual

@@
 \section{The {\ttfamily general} namelist}
+
+The default values are usually best for the parameters in this namelist.
+
+\myhrule
 
@@
 \label{sec:geometryParameters}
+
+The parameters in this namelist define the magnetic geometry, and so you will almost certainly want to modify some of these parameters.
+
+\myhrule
 
@@
 
+This namelist defines which species are included in the calculation, along with the density and temperature and gradients thereof.
+You will definitely want to set the parameters in this namelist.
+
+\myhrule
+
 \param{Zs}
@@
 \section{The {\ttfamily physicsParameters}~namelist}
+
+The parameters in this namelist determine which terms are included or excluded in the kinetic equation.
+You will want to be aware of most of these parameters.
+
+\myhrule
 
@@
 
+\section{The {\ttfamily otherNumericalParameters}~namelist}
+
+The parameters in this namelist are advanced, and the default values are best for routine use of the code.
+
+\myhrule
+
+\param{thetaDerivativeScheme}
+{integer}
+{2}
+{Always}
+{Discretization scheme for the poloidal angle coordinate theta.\\
+
+{\ttfamily thetaDerivativeScheme} = 0: Fourier spectral collocation.  The differentiation matrix in theta is dense.\\
+
+{\ttfamily thetaDerivativeScheme} = 1: Finite differences with a 3 point stencil.  (The differentiation matrix in theta is tridiagonal, aside from the corners.)\\
+
+{\ttfamily thetaDerivativeScheme} = 2: Finite differences with a 5 point stencil.  (The differentiation matrix in theta is pendadiagonal, aside from the corners.).\\
+
+The best value for this parameter is usually 2.}
+
+\myhrule
+
+\param{zetaDerivativeScheme}
+{integer}
+{2}
+{Always}
+{Discretization scheme for the toroidal angle coordinate zeta.\\
+
+{\ttfamily zetaDerivativeScheme} = 0: Fourier spectral collocation.  The differentiation matrix in zeta is dense.\\
+
+{\ttfamily zetaDerivativeScheme} = 1: Finite differences with a 3 point stencil.  (The differentiation matrix in zeta is tridiagonal, aside from the corners.)\\
+
+{\ttfamily zetaDerivativeScheme} = 2: Finite differences with a 5 point stencil.  (The differentiation matrix in zeta is pendadiagonal, aside from the corners.).\\
+
+The best value for this parameter is usually 2.}
+
+\myhrule
+
+\param{xGridScheme}
+{integer}
+{5}
+{Whenever {\ttfamily RHSMode} is 1 or 2.}
+{Discretization scheme for the speed coordinate $x$.\\
+
+{\ttfamily xGridScheme} = 1: New orthogonal polynomials with no point at $x=0$. Original treatment of Rosenbluth potentials.\\
+
+{\ttfamily xGridScheme} = 2: New orthogonal polynomials with a point at $x=0$. Original treatment of Rosenbluth potentials.\\
+
+{\ttfamily xGridScheme} = 3: Uniform finite differences on [0, {\ttfamily xMax}], forcing $f=0$ at {\ttfamily xMax}. 2-point stencil for interpolating to other grids.\\
+
+{\ttfamily xGridScheme} = 4: Uniform finite differences on [0, {\ttfamily xMax}], forcing $f=0$ at {\ttfamily xMax}. 4-point stencil for interpolating to other grids.\\
+
+{\ttfamily xGridScheme} = 5: New orthogonal polynomials with no point at $x=0$. New treatment of Rosenbluth potentials.\\
+
+{\ttfamily xGridScheme} = 6: New orthogonal polynomials with a point at $x=0$. New treatment of Rosenbluth potentials.\\
+
+The recommended value for this parameter is 5.  
+When {\ttfamily xGridScheme} = 5 or 6, then the following quantities do not matter: 
+{\ttfamily NxPotentialsPerVth}, {\ttfamily xMax}, and {\ttfamily xPotentialsGridScheme}.}
+
+\myhrule
+
+\param{xGrid\_k}
+{integer}
+{0}
+{Whenever {\ttfamily RHSMode} is 1 or 2 and {\ttfamily xGridScheme} = 1, 2, 5, or 6.}
+{For {\ttfamily xGridScheme} = 1, 2, 5, or 6, the distribution function will be represented in terms of polynomials $P_n(x)$
+that are orthogonal under the weight $\int_0^\infty dx\; x^k \exp(-x^2) P_n(x) P_m(x)  \propto  \delta_{n,m}$
+where $k$ is an exponent set by the parameter {\ttfamily xGrid\_k} here.
+A good value to use is 0, 1, or 2.}
+
+\myhrule
+
+\param{xPotentialsGridScheme}
+{integer}
+{2}
+{Whenever {\ttfamily RHSMode} is 1 or 2 and {\ttfamily xGridScheme} is $<$5.
+Since the recommended setting for {\ttfamily xGridScheme} is 5, this parameter is rarely relevant.}
+{When an explicit grid is used for the Rosenbluth potentials, which grid and interpolation scheme to use.\\
+
+{\ttfamily xPotentialsGridScheme} = 1: Uniform grid. 5-point stencil for derivatives. 2-point stencil for interpolating to other grids.\\
+
+{\ttfamily xPotentialsGridScheme} = 2: Uniform grid. 5-point stencil for derivatives. 4-point stencil for interpolating to other grids.\\
+
+{\ttfamily xPotentialsGridScheme} = 3: Use same grid as for distribution function, so no interpolation needed for the self-collision operator.
+  You must set {\ttfamily xGridScheme} = 3 or 4 to use this setting.  Use 2-point stencil for interpolating to other species' grids.\\
+
+{\ttfamily xPotentialsGridScheme} = 4: Same as option 3, except use a 4-point stencil for interpolating to other species' grids.\\
+
+The recommended setting is {\ttfamily xPotentialsGridScheme} = 2.}
+
+\myhrule
+
+\param{useIterativeLinearSolver}
+{Boolean}
+{\true}
+{Always}
+{If false, a sparse direct solver
+will be used. The direct solver is faster for small (i.e. low-resolution) problems and always yields a solution
+(as long as there is sufficient memory).
+For large (high resolution) problems, the iterative solver will usually be faster and will use much
+ less memory, but it may not always converge.
+}
+
+\myhrule
+
+\param{whichParallelSolverToFactorPreconditioner}
+{integer}
+{1}
+{Always}
+{Which software package is used to $LU$-factorize the preconditioner matrix.\\
+
+{\ttfamily whichParallelSolverToFactorPreconditioner} = 1: Use \mumps~if it is available, otherwise use \superludist.\\
+
+{\ttfamily whichParallelSolverToFactorPreconditioner} = 2: Force use of \superludist~ even if \mumps~is available.
+}
+
+\myhrule
+
+\param{PESCPreallocationStrategy}
+{integer}
+{1}
+{Always}
+{This setting changes the estimated number of nonzeros (nnz) used for allocating memory for the system matrix and preconditioner.\\
+
+{\ttfamily PESCPreallocationStrategy} = 0: Old method with high estimated nnz. This method involves relatively simpler code but uses WAY more memory than necessary.\\
+
+{\ttfamily PESCPreallocationStrategy} = 1: New method with lower, more precise estimated nnz.  This method should use much less memory.\\
+
+Use {\ttfamily PETSCPreallocationStrategy} = 1 unless you know what you are doing.}
+
+\section{The {\ttfamily preconditionerOptions}~namelist}
+
+This namelist controls how elements are removed from the ``real'' matrix in order to obtain
+the preconditioner matrix. The default values are usually best, but if you find that there are more than 100 iterations
+of GMRES/KSP, it may be worth adjusting these settings.  As long as KSP converges, these parameters
+should have no impact (to several digits) on the physical outputs such as parallel flows and radial fluxes.
+Therefore, do not worry about (for example) ``dropping coupling between species'' in the first
+parameter below, since full inter-species coupling will be retained in the real equations that are being solved.
+
+\myhrule
+
+\param{preconditioner\_species}
+{integer}
+{1}
+{Whenever {\ttfamily useIterativeLinearSolver} = \true~and there are 2 or more species.}
+{\\
+{\ttfamily preconditioner\_species} = 0: Keep all coupling between species.\\
+
+{\ttfamily preconditioner\_species} = 1: Drop all coupling between species.\\
+
+The default value of 1 is recommended, except perhaps at high collisionality where 0 may be preferable.}
+
+\myhrule
+
+\param{preconditioner\_x}
+{integer}
+{1}
+{Whenever {\ttfamily useIterativeLinearSolver} = \true~and {\ttfamily RHSMode} = 1 or 2.}
+{\\
+{\ttfamily preconditioner\_x} = 0: Keep full $x$ coupling.\\
+
+{\ttfamily preconditioner\_x} = 1: Drop everything off-diagonal in $x$.\\
+
+{\ttfamily preconditioner\_x} = 2: Keep only upper-triangular part in $x$.\\
+
+{\ttfamily preconditioner\_x} = 3: Keep only the tridiagonal terms in $x$.\\
+
+{\ttfamily preconditioner\_x} = 4: Keep only the diagonal and superdiagonal in $x$.\\
+
+The default value of 1 is strongly recommended, except perhaps at high collisionality where 0 may be preferable.}
+
+\myhrule
+
+\param{preconditioner\_x\_min\_L}
+{integer}
+{0}
+{Whenever {\ttfamily useIterativeLinearSolver} = \true~and {\ttfamily RHSMode} = 1 or 2 and {\ttfamily preconditioner\_x} $>$ 0.}
+{The $x$ structure of the matrix will only be simplified for Legendre index $L$ is $\ge$ this value.
+Set {\ttfamily preconditioner\_x\_min\_L} = 0 to simplify the matrix for every $L$.
+Recommended values are 0, 1, or 2.}
+
+\myhrule
+
+\param{preconditioner\_theta}
+{integer}
+{0}
+{Whenever {\ttfamily useIterativeLinearSolver} = \true}
+{\\
+{\ttfamily preconditioner\_theta} = 0: Keep full $\theta$ coupling.\\
+
+{\ttfamily preconditioner\_theta} = 1: Use a 3-point finite difference stencil for $d/d\theta$.\\
+
+{\ttfamily preconditioner\_theta} = 2: Drop all $\theta$ coupling.\\
+
+{\ttfamily preconditioner\_theta} = 3: Replace $d/d\theta$ with the identity matrix.\\
+
+The default value of 0 is strongly recommended.}
+
+\myhrule
+
+\param{preconditioner\_theta\_min\_L}
+{integer}
+{0}
+{Whenever {\ttfamily useIterativeLinearSolver} = \true~and {\ttfamily preconditioner\_theta} $>$ 0.}
+{The $\theta$ structure of the matrix will only be simplified for Legendre index $L$ is $\ge$ this value.
+Set {\ttfamily preconditioner\_theta\_min\_L} = 0 to simplify the matrix for every $L$.
+}
+
+\myhrule
+
+\param{preconditioner\_zeta}
+{integer}
+{0}
+{Whenever {\ttfamily useIterativeLinearSolver} = \true}
+{\\
+{\ttfamily preconditioner\_zeta} = 0: Keep full $\zeta$ coupling.\\
+
+{\ttfamily preconditioner\_zeta} = 1: Use a 3-point finite difference stencil for $d/d\zeta$.\\
+
+{\ttfamily preconditioner\_zeta} = 2: Drop all $\zeta$ coupling.\\
+
+{\ttfamily preconditioner\_zeta} = 3: Replace $d/d\zeta$ with the identity matrix.\\
+
+The default value of 0 is strongly recommended.}
+
+\myhrule
+
+\param{preconditioner\_zeta\_min\_L}
+{integer}
+{0}
+{Whenever {\ttfamily useIterativeLinearSolver} = \true~and {\ttfamily preconditioner\_zeta} $>$ 0.}
+{The $\zeta$ structure of the matrix will only be simplified for Legendre index $L$ is $\ge$ this value.
+Set {\ttfamily preconditioner\_zeta\_min\_L} = 0 to simplify the matrix for every $L$.
+}
+
+\myhrule
+
+\param{preconditioner\_xi}
+{integer}
+{1}
+{Whenever {\ttfamily useIterativeLinearSolver} = \true}
+{\\
+{\ttfamily preconditioner\_xi} = 0: Keep full $\xi$ coupling.\\
+
+{\ttfamily preconditioner\_xi} = 1: Drop terms that are $\pm 2$ rows from the diagonal in $\xi$, 
+      so the preconditioner matrix becomes tridiagonal in $\xi$.
+(Normally the preconditioner matrix is pentadiagonal in $\xi$.)\\
+
+Either a setting of 0 or 1 can be good for this parameter.}
+
+\myhrule
+
+\param{reusePreconditioner}
+{Boolean}
+{\true}
+{Only when {\ttfamily nonlinear} = \true}
+{If true, the nonlinear term will not be included in the preconditioner matrix, meaning the preconditioner matrix is the same
+at every iteration, and so the preconditioner matrix only needs to be $LU$-factorized once.  If false, the preconditioner matrix
+for the Jacobian will be different at each iteration of the Newton solve, so the preconditioner needs to be $LU$-factorized at
+each iteration.  The nonlinear term also introduces a lot of nonzeros into the preconditioner matrix, so setting
+{\ttfamily reusePreconditioner =}\true~not only dramatically reduces the time required for a nonlinear calculation, but also the memory required.}
+
+\section{The {\ttfamily export\_f}~namelist}
+
+This namelist controls whether and how the distribution function is saved in {\ttfamily sfincsOutput.h5}.
+For each of the 4 coordinates $(\theta, \zeta, x, \xi)$, the distribution function can be given with the same discretization
+used for solving the kinetic equation, or you can interpolate to a different grid/discretization.
+For all available settings, the distribution function will be reported on a tensor product grid in the 4 coordinates.
+
+\myhrule
+
+\param{export\_full\_f}
+{Boolean}
+{\false}
+{Always}
+{Whether or not to save the full distribution function (the sum of the leading-order Maxwellian and the departure from it)
+in the output file.}
+
+\myhrule
+
+\param{export\_delta\_f}
+{Boolean}
+{\false}
+{Always}
+{Whether or not to save the departure from a Maxwellian distribution function in the output file.}
+
+\myhrule
+
+\param{export\_f\_theta\_option}
+{integer}
+{2}
+{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true}
+{Controls which grid in $\theta$ is used for exporting the distribution function.\\
+
+{\ttfamily export\_f\_theta\_option} = 0: Report the distribution function on the original $\theta$ grid (with \Ntheta~points) used for solving the kinetic equation.\\
+
+{\ttfamily export\_f\_theta\_option} = 1: Interpolate to a different grid, specified by {\ttfamily export\_f\_theta}. Linear interpolation will be used.
+No sorting of the requested values is performed.\\
+
+{\ttfamily export\_f\_theta\_option} = 2: Do not interpolate. Use the
+values of the $\theta$ grid that are closest to the values requested
+in {\ttfamily export\_f\_theta}.  Values of $\theta$ will be in
+increasing order. If multiple requested values are close to the same
+grid point, the number of points returned will be less than the number
+of points requested. \\
+
+For all of these options, you can see {\ttfamily export\_f\_theta} in {\ttfamily sfincsOutput.h5} for the actual grid used in the end.
+}
+
+\myhrule
+
+\param{export\_f\_zeta\_option}
+{integer}
+{2}
+{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true}
+{Controls which grid in $\zeta$ is used for exporting the distribution function.\\
+
+{\ttfamily export\_f\_zeta\_option} = 0: Report the distribution function on the original $\zeta$ grid (with \Nzeta~points) used for solving the kinetic equation.\\
+
+{\ttfamily export\_f\_zeta\_option} = 1: Interpolate to a different grid, specified by {\ttfamily export\_f\_zeta}. Linear interpolation will be used.
+No sorting of the requested values is performed.\\
+
+{\ttfamily export\_f\_zeta\_option} = 2: Do not interpolate. Use the
+values of the $\zeta$ grid that are closest to the values requested
+in {\ttfamily export\_f\_zeta}.  Values of $\zeta$ will be in
+increasing order. If multiple requested values are close to the same
+grid point, the number of points returned will be less than the number
+of points requested. \\
+
+For all of these options, you can see {\ttfamily export\_f\_zeta} in {\ttfamily sfincsOutput.h5} for the actual grid used in the end.
+}
+
+\myhrule
+
+\param{export\_f\_theta}
+{1D array of reals}
+{0.0}
+{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true, and {\ttfamily export\_f\_theta\_option} $>0$.}
+{Values of $\theta$ on which you want to save the distribution function.  modulo$(\ldots, 2\pi)$ will be applied.  See {\ttfamily export\_f\_theta\_option} for details}
+
+\myhrule
+
+\param{export\_f\_zeta}
+{1D array of reals}
+{0.0}
+{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true, and {\ttfamily export\_f\_zeta\_option} $>0$.}
+{Values of $\zeta$ on which you want to save the distribution function.  modulo$(\ldots, 2\pi/\mbox{\ttfamily NPeriods})$ will be applied.  See {\ttfamily export\_f\_zeta\_option} for details}
+
+\myhrule
+
+\param{export\_f\_xi\_option}
+{integer}
+{1}
+{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true}
+{Controls which discretization in $\xi$ is used for exporting the distribution function.\\
+
+{\ttfamily export\_f\_xi\_option} = 0: Report the distribution function as amplitudes of \Nxi~Legendre polynomials, as used internally by \sfincs~for solving the kinetic equation.\\
+
+{\ttfamily export\_f\_xi\_option} = 1: Report the distribution function on the values of $\xi$ specified by {\ttfamily export\_f\_xi}. No sorting of the requested values is performed.
+}
+
+\myhrule
+
+\param{export\_f\_xi}
+{1D array of reals}
+{0.0}
+{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true, and {\ttfamily export\_f\_xi\_option} = 1.}
+{Values of $\xi$ on which you want to save the distribution function.  Values must lie in the range $[-1,1]$.}
+
+\myhrule
+
+\param{export\_f\_x\_option}
+{integer}
+{0}
+{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true}
+{Controls which grid in $x = v/\sqrt{2T/m}$ is used for exporting the distribution function.\\
+
+{\ttfamily export\_f\_x\_option} = 0: Report the distribution function on the original $x$ grid (with \Nx~points) used for solving the kinetic equation.\\
+
+{\ttfamily export\_f\_x\_option} = 1: Interpolate to a different grid, specified by {\ttfamily export\_f\_x}.  Polynomial spectral interpolation will be used.
+  No sorting of the requested values is performed.\\
+
+{\ttfamily export\_f\_x\_option} = 2: Do not interpolate. Use the values of the internal $x$ grid that are closest to the values requested in {\ttfamily export\_f\_x}.
+  Values of $x$ will be in increasing order. If multiple requested values are close to the same grid point, the number of
+  points returned will be less than the number of points requested.\\
+
+For all of these options, you can see {\ttfamily export\_f\_x} in {\ttfamily sfincsOutput.h5} for the actual grid used in the end.
+}
+
+\myhrule
+
+\param{export\_f\_x}
+{1D array of reals}
+{1.0}
+{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true, and {\ttfamily export\_f\_x\_option} $>0$.}
+{Values of $x$ on which you want to save the distribution function.  Values must be $\ge 0$.}
+
 \section{Directives for \sfincsScan}
@@
 
-{\ttfamily scanType=1}:  Convergence scan. (Scan the parameters in the resolutionParameters namelist.)\\
-
-{\ttfamily scanType=2}:  Scan of $E_r$.\\
-
-{\ttfamily scanType=3}:  Scan any one input parameter that takes a numeric value.\\
-
-{\ttfamily scanType=4}:  Scan radius, taking the density and temperature profiles from the {\ttfamily profiles} file.
+{\ttfamily scanType} = 1:  Convergence scan. (Scan the parameters in the resolutionParameters namelist.)\\
+
+{\ttfamily scanType} = 2:  Scan of $E_r$.\\
+
+{\ttfamily scanType} = 3:  Scan any one input parameter that takes a numeric value.\\
+
+{\ttfamily scanType} = 4:  Scan radius, taking the density and temperature profiles from the {\ttfamily profiles} file.
 In this type of scan, the same radial electric field is used at every radius.
@@
 
-{\ttfamily scanType=5}:  Scan radius, and at each radius, scan $E_r$.  Density and temperature profiles are again
+{\ttfamily scanType} = 5:  Scan radius, and at each radius, scan $E_r$.  Density and temperature profiles are again
  taken from the {\ttfamily profiles} file; see \path{sfincs/fortran/utils/profiles.XXX} for examples.\\
 
-{\ttfamily scanType=21}: Read in a list of requested runs from a file {\ttfamily runspec.dat}. See \path{sfincs/fortran/utils/sfincsScan_21} for an example file.
+{\ttfamily scanType} = 21: Read in a list of requested runs from a file {\ttfamily runspec.dat}. See \path{sfincs/fortran/utils/sfincsScan_21} for an example file.
 }

### 2015-05-23 fortran/version3/export_f.F90

@@ -10,16 +10,16 @@ module export_f
 #include <finclude/petscsysdef.h>
 
   integer, parameter :: max_N_export_f = 500 ! Max grid size in each one of the 4 coordinates (theta, zeta, xi, x)
-  PetscScalar, dimension(max_N_export_f) :: export_f_theta
-  PetscScalar, dimension(max_N_export_f) :: export_f_zeta
-  PetscScalar, dimension(max_N_export_f) :: export_f_xi
-  PetscScalar, dimension(max_N_export_f) :: export_f_x
+  PetscScalar, dimension(max_N_export_f) :: export_f_theta = 0
+  PetscScalar, dimension(max_N_export_f) :: export_f_zeta  = 0
+  PetscScalar, dimension(max_N_export_f) :: export_f_xi    = 0
+  PetscScalar, dimension(max_N_export_f) :: export_f_x     = 1
   integer :: N_export_f_theta, N_export_f_zeta, N_export_f_xi, N_export_f_x
   PetscScalar, dimension(:,:), allocatable :: map_theta_to_export_f_theta
   PetscScalar, dimension(:,:), allocatable :: map_zeta_to_export_f_zeta
   PetscScalar, dimension(:,:), allocatable :: map_x_to_export_f_x
   PetscScalar, dimension(:,:), allocatable :: map_xi_to_export_f_xi
-  integer :: export_f_theta_option=0, export_f_zeta_option=0, export_f_xi_option=0, export_f_x_option=0
+  integer :: export_f_theta_option=2, export_f_zeta_option=2, export_f_xi_option=1, export_f_x_option=0
   logical :: export_full_f = .false., export_delta_f = .false.
 
   PetscScalar, dimension(:,:,:,:,:), allocatable :: delta_f, full_f

### 2015-05-23 fortran/version3/globalVariables.F90

@@ -115,7 +115,7 @@ module globalVariables
   ! 2 = 4th order dinite differences
 
   integer :: xPotentialsGridScheme = 2, xPotentialsInterpolationScheme
-  integer :: xGridScheme = 1, xInterpolationScheme
+  integer :: xGridScheme = 5, xInterpolationScheme
   logical :: pointAtX0
 
   integer :: Ntheta = 15
@@ -135,19 +135,19 @@ module globalVariables
 
   logical :: useIterativeLinearSolver=.true.
 
-  integer :: whichParallelSolverToFactorPreconditioner
+  integer :: whichParallelSolverToFactorPreconditioner = 1
   ! Options for whichParallelSolverToFactorPreconditioner:
   ! 1 = use mumps if it is detected, otherwise use superlu_dist
   ! 2 = force use of superlu_dist, if it is available
 
   integer :: preconditioner_x=1, preconditioner_x_min_L=0, preconditioner_zeta=0
-  integer :: preconditioner_theta=0, preconditioner_xi=0, preconditioner_species=1
+  integer :: preconditioner_theta=0, preconditioner_xi=1, preconditioner_species=1
   integer :: preconditioner_theta_min_L=0, preconditioner_zeta_min_L=0
   logical :: reusePreconditioner=.true.
 
   integer :: constraintScheme=-1
 
-  integer :: PETSCPreallocationStrategy
+  integer :: PETSCPreallocationStrategy=1
 
   ! ********************************************************
   !

### 2015-05-27 doc/manual/inputParameters.tex
Added to user manual, better control of mumps cntl 1

@@
 
+(All the variables in this formula are discussed later in this namelist.)\\
+
 {\ttfamily geometryScheme}==2: Use a 3-helicity model of the LHD standard configuration at {\ttfamily rN}=0.5.\\
@@
 {\ttfamily geometryScheme}==12: Read the IPP {\ttfamily .bc} format Boozer-coordinate file specified in {\ttfamily equilibriumFile} below. 
-The file is assumed to be stellarator-asymmetric.
+The file is assumed to be stellarator-\emph{asymmetric}.
 }
@@
 or to specify the range of flux surfaces for a radial scan. 
-(When {\ttfamily geometryScheme} == 2, 3, or 4, the flux surface used will be {\ttfamily rN}=0.5.)
+(Regardless of the value of this parameter, when {\ttfamily geometryScheme} == 2, 3, or 4, the flux surface used will be {\ttfamily rN}=0.5.)
 See section \ref{sec:radialCoordinates}
@@
 You will definitely want to set the parameters in this namelist.
+Note that only one of the four parameters 
+{\ttfamily dnHatdpsiHats}, 
+{\ttfamily dnHatdpsiNs}, 
+{\ttfamily dnHatdrHats}, or
+{\ttfamily dnHatdrNs}
+will be used, depending on the value of {\ttfamily inputRadialCoordinateForGradients}
+in the {\ttfamily geometryParameters} namelist.
+Similarly, only one of the four parameters
+{\ttfamily dTHatdpsiHats}, 
+{\ttfamily dTHatdpsiNs}, 
+{\ttfamily dTHatdrHats}, or
+{\ttfamily dTHatdrNs}
+will be used.
 

### 2015-05-27 doc/manual/installation.tex
Added to user manual, better control of mumps cntl 1

@@
 
+To eliminate the need to set the environment variable and path 
+as described above
+at each login session, you may find it convenient to set both
+in your startup script, such as  {\ttfamily .bashrc}.  In this startup script
+you may also want to load any modules needed by \sfincsScan~such as \python, {\ttfamily numpy},
+{\ttfamily scipy}, and {\ttfamily matplotlib} (if your computing system uses modules).
+
 \section{Setting up \sfincs~on a new system}
@@
 Second, you will need to edit {\ttfamily utils/sfincsScan}. Look for the {\ttfamily if} block near the top
-with sections for {\ttfamily sfincsSystem = edison},{\ttfamily hydra}, and {\ttfamily laptop}. Add an analogous block
+with sections for {\ttfamily sfincsSystem = edison}, {\ttfamily hydra}, and {\ttfamily laptop}. Add an analogous block
 for your system to set the command used to submit jobs, and a {\ttfamily nameJobFile} function.
@@
 will be run on any existing {\ttfamily sfincsOutput.h5} files in the \path{sfincs/fortran/version3/examples/} directories.
+
+The {\ttfamily make test} functionality relies on system-specific commands in the
+\path{sfincs/fortran/version3/makefiles/makefile.SFINCS_SYSTEM} file.  If you experience
+problems with  {\ttfamily make test}, it is likely that this file needs modification.

### 2015-05-27 doc/manual/overview.tex
Added to user manual, better control of mumps cntl 1

@@
 with nested toroidal flux surfaces.
-The code solves a drift-kinetic equation for an arbitrary number of species.  Optionally, a quasi-neutrality
-equation can also be solved to obtain the self-consistent variation of the electrostatic potential on a flux surface.
+As with other neoclassical codes, the input information used by \sfincs~is the equilibrium magnetic geometry together with the density, radial density gradient, temperature,
+and radial temperature gradient of each species.  The code then solves a drift-kinetic equation for each species,
+yielding the (gyro-angle averaged) distribution function.  Moments of the distribution function are computed such as the parallel
+flow, bootstrap current, radial particle flux, radial heat flux, and variation of the density over a flux surface.  
+These moments are all saved in the output file, and if you wish, you can also save the distribution function itself.
+Optionally, a quasi-neutrality
+equation can be solved at the same time as the drift-kinetic equations, yielding the self-consistent variation of the electrostatic potential on a flux surface.
 
-This manual describes ``version 3'' of \sfincs.  There are two older versions of the
-code called singleSpecies and multiSpecies.
+The kinetic equations solved in \sfincs~have four independent variables: poloidal angle $\theta$, toroidal angle $\zeta$,
+normalized speed $x = v / v_{thermal}$, and pitch angle $\xi = v_{||}/v$.  The third velocity coordinate (gyro-angle) does not appear
+since gyro-averaged equations are solved.  The flux surface label (radius) coordinate is only a parameter, rather than a full independent variable,
+since a radially local approximation is made.
 
@@
 
-Often, the limiting factor for \sfincs~is the ability of \mumps~or \superludist~to factorize the preconditioner matrix.
+Often, the limiting factor for \sfincs~is the ability of the libraries \mumps~or \superludist~to factorize the preconditioner matrix, discussed in section \ref{sec:gmres}.
 You may therefore find it useful to see the control parameters and error codes in the \mumps~user manual:
 \url{http://mumps.enseeiht.fr/doc/userguide_5.0.0.pdf}.
+
+This manual describes ``version 3'' of \sfincs.  
+To preserve previous versions of the code that have been used for publications, two older versions of the
+code called singleSpecies and multiSpecies are also present in the repository.  For both of these older versions,
+both MATLAB and fortran editions exist which are independent of each other.  At present, version 3 exists
+only in a fortran edition.
+
+\section{Features}
+
+\begin{itemize}
+
+\item
+Both self-species and inter-species collisions are treated using the most accurate linear operator available, the full linearized Fokker-Planck collision operator,
+with no approximation of the field-particle term or expansion in mass ratio.  This collision operator conserves mass, momentum, and energy.
+
+\item
+Realistic experimental geometry can be simulated using an interface to \vmec.  Analytic model equilibria can also be used.
+
+\item
+Full coupling in the speed (or equivalently, kinetic energy) coordinate is retained, i.e. no monoenergetic approximation is made.  However,
+if desired, \sfincs~can also be run in monoenergetic mode to compare with older codes.
+
+\item
+The code is formulated to permit solution of a wide variety of kinetic equations, whether or not phase space volume and/or energy are conserved, so individual terms can be turned on or off to examine their effect.
+
+\item
+A variety of models for terms involving the radial electric field are available to allow comparison between models.
+
+\item
+The code takes advantage of modern algorithms (GMRES) and parallelized libraries (\PETSc, \superludist, and \mumps).
+
+\item
+Efficient representation of velocity space is achieved using a pseudospectral method based upon non-classical orthogonal polynomials. \cite{speedGrids}
+
+\item
+Optional nonlinear terms in the kinetic equation (involving both the non-Maxwellian distribution function and poloidal/toroidal electric field) can be included using Newton's method.
+
+\end{itemize}
+
+\section{Limitations}
+
+\begin{itemize}
+
+\item
+The \sfincs~code is radially local, in the sense that it approximates 
+the radial derivative of the distribution function $\partial f/\partial \psi$ by
+the derivative of a Maxwellian flux function $\partial f_M(\psi,x)/\partial \psi$.
+This approximation is important for reducing the otherwise 5D space of independent variables $(\psi,\theta,\zeta,x,\xi)$
+to a 4D space $(\theta,\zeta,x,\xi)$.
+As a result, \sfincs~cannot compute certain finite-orbit-width effects that occur when the radial extent of the particle orbits
+between bounces or transits is not small compared to the scale of radial variation in the equilibrium.
+Such finite orbit width effects are significant near the magnetic axis, and in strong transport barriers, such as the pedestal of a tokamak H-mode.
+
+\item
+Turbulence is neglected.  There are good theoretical reasons to expect that the neoclassical effects computed by \sfincs~should
+decouple from turbulence, as detailed in \cite{AbelReview}.  However, this argument relies on an expansion in $\rho_* \ll 1$, and so may break down
+in some circumstances when $\rho_*$ is not sufficiently small.
+
+\item
+It is assumed that nested toroidal magnetic surfaces exist. Thus, the code cannot accurately model regions of stochastic field,
+magnetic islands, or open field lines.
+
+\end{itemize}
 
@@
 as are several analytic models for LHD and W7-X in which 3 or 4 Fourier components are retained.
+The primary switch for controlling the magnetic geometry in \sfincs~is the {\ttfamily geometryScheme} parameter
+in the {\ttfamily geometryParameters} input namelist.
 For more details about geometry options in \sfincs, see section \ref{sec:geometryParameters}.
@@
 For this task, the \sfincsScan~family of \python~scripts is available.
+Using these scripts, it is also possible to scan other variables in the input file,
+and in particular, to scan the resolution parameters to ensure
+the physical output quantities are numerically converged.
 For a full list of the types of scans available, see section \ref{sec:sfincsScanParams}
@@
 
-To report any bugs, provide feedback, or ask questions, contact Matt Landreman at
+We enthusiastically welcome any contributions to the code or documentation.
+For write permission to the repository, or to report any bugs, provide feedback, or ask questions, contact Matt Landreman at
 \href{mailto:matt.landreman@gmail.com}{\nolinkurl{matt.landreman@gmail.com} }

### 2015-05-27 doc/manual/inputParameters.tex
Added profiles examples

@@
 
+The resolution parameters each have 3 associated sfincsScan parameters
+which are used for convergence scans ({\ttfamily scanType} = 1): {\ttfamily ...MinFactor}, {\ttfamily ...MaxFactor}, and {\ttfamily ...NumRuns}.
+The first two of these set the range by which the associated resolution parameter is scaled
+in a convergence scan. The {\ttfamily ...NumRuns} parameter sets the number of values tried in a
+convergence scan. The code attempts to space the values evenly in a logarithmic
+sense, as in Matlab's 'logspace' function. For example, the following settings\\
+{\ttfamily
+    Nxi = 20\\
+!ss NxiMinFactor = 0.5\\
+!ss NxiMaxFactor = 2.0\\
+!ss NxiNumRuns = 3\\
+}
+would mean the values {\ttfamily Nxi} = 10, 20, and 40 would be tried in a convergence scan.
+If you don't want to scan a variable in a convergence scan, set the associated
+{\ttfamily ...NumRuns} parameter to 0, or do not include this parameter in the input file.
+For each resolution parameter ({\ttfamily Ntheta}, {\ttfamily Nzeta}, {\ttfamily Nxi}, etc.), the value itself is read by
+Fortran and so should not be preceded by {\ttfamily !ss}.  However the {\ttfamily ...MinFactor}, {\ttfamily ...MaxFactor}, and {\ttfamily ...NumRuns}
+quantities are read by \sfincsScan~and so must be preceded by {\ttfamily !ss}
+
+\myhrule
+
 \ssparam{scanType}

### 2015-05-27 doc/manual/runs.tex
edited section 5.6 of user manual

@@
 $v$ is the speed at which the monoenergetic calculation is being performed, and $B_0$ is the (0,0) Fourier harmonic of $B$
-with respect to the Boozer poloidal and toroidal angles.
-\todo{How is $\nu$ defined here?}
+with respect to the Boozer poloidal and toroidal angles. The collision
+frequency $\nu$ is here the value of $\nu_\mathrm{ii}$ one would have if
+$v$ were the thermal speed. That is, in SI units,
+\begin{equation}
+  \nu=\frac{4}{3\sqrt{\pi}}\frac{n Z^4e^4\ln \Lambda}{4\pi\epsilon_0^2m^2v^3}.
+\end{equation}
+%
 The normalized radial electric field is
@@
 \todo{Should be change the behavior of RHSMode=2 so it uses nuPrime and EStar instead of nu\_n?}
+
+The two parameters {\ttfamily nuPrime} and {\ttfamily EStar} are
+related to the corresponding DKES parameters {\ttfamily CMUL} and
+{\ttfamily EFIELD} by
+\begin{eqnarray}
+  \mbox{\ttfamily
+    CMUL}&\equiv&\frac{\nu_\mathrm{D}}{v}=\frac{3\sqrt{\pi}}{4}\left(\mathrm{erf}(1)-\mathrm{Ch}(1)\right)
+  \frac{B_0}{G+\iota I}\mbox{\ttfamily nuPrime},\\
+  \mbox{\ttfamily EFIELD}&\equiv&-\left[\frac{d\Phi}{dr}\right]_\mathrm{DKES}\frac{1}{vB_0}=-\frac{\iota}{G}\left[\frac{d\Psi}{dr}\right]_\mathrm{DKES}\mbox{\ttfamily EStar},
+\end{eqnarray}
+where $\mathrm{Ch}$ is the Chandrasekhar function and $\nu_\mathrm{D}$
+is the actual pitch-angle deflection frequency of the particle.
+\todo{These expressions are in SI units. Should there be some
+  factors of $c$ to adhere to the Gaussian standard used here?
+  Probably not.}

### 2015-05-28 doc/manual/inputParameters.tex
Added to user manual

@@
 \section{The {\ttfamily resolutionParameters}~namelist}
+\label{sec:resolutionParameters}
 
@@
 
-The resolution parameters each have 3 associated sfincsScan parameters
+\myhrule
+
+\ssparam{scanType}
+{integer}
+{Any time \sfincsScan~is called.}
+{Which type of scan will be run when \sfincsScan~is called. \\
+
+{\ttfamily scanType} = 1:  Resolution convergence scan. (Scan the parameters in the resolutionParameters namelist.)\\
+
+{\ttfamily scanType} = 2:  Scan of $E_r$.\\
+
+{\ttfamily scanType} = 3:  Scan any one input parameter that takes a numeric value.\\
+
+{\ttfamily scanType} = 4:  Scan radius, taking the density and temperature profiles from the {\ttfamily profiles} file.
+In this type of scan, the same radial electric field is used at every radius.
+See \path{sfincs/fortran/utils/profiles.XXX} for examples.\\
+
+{\ttfamily scanType} = 5:  Scan radius, and at each radius, scan $E_r$.  Density and temperature profiles are again
+ taken from the {\ttfamily profiles} file; see \path{sfincs/fortran/utils/profiles.XXX} for examples.
+In this type of scan, \sfincsScan~creates a subdirectory for each value of minor radius, and a
+{\ttfamily scanType} = 2 scan is run in each of these subdirectories.
+\\
+
+{\ttfamily scanType} = 21: Read in a list of requested runs from a file {\ttfamily runspec.dat}. See \path{sfincs/fortran/utils/sfincsScan_21} for an example file.
+}
+
+\subsection{Parameters related only to {\ttfamily scanType} = 1 (resolution convergence scans).}
+
+The resolution parameters discussed in section \ref{sec:resolutionParameters}
+each have 3 associated \sfincsScan~parameters
 which are used for convergence scans ({\ttfamily scanType} = 1): {\ttfamily ...MinFactor}, {\ttfamily ...MaxFactor}, and {\ttfamily ...NumRuns}.
@@
 
-\ssparam{scanType}
-{integer}
-{Any time \sfincsScan~is called.}
-{Which type of scan will be run when \sfincsScan~is called. \\
-
-{\ttfamily scanType} = 1:  Convergence scan. (Scan the parameters in the resolutionParameters namelist.)\\
-
-{\ttfamily scanType} = 2:  Scan of $E_r$.\\
-
-{\ttfamily scanType} = 3:  Scan any one input parameter that takes a numeric value.\\
-
-{\ttfamily scanType} = 4:  Scan radius, taking the density and temperature profiles from the {\ttfamily profiles} file.
-In this type of scan, the same radial electric field is used at every radius.
-See \path{sfincs/fortran/utils/profiles.XXX} for examples.\\
-
-{\ttfamily scanType} = 5:  Scan radius, and at each radius, scan $E_r$.  Density and temperature profiles are again
- taken from the {\ttfamily profiles} file; see \path{sfincs/fortran/utils/profiles.XXX} for examples.\\
-
-{\ttfamily scanType} = 21: Read in a list of requested runs from a file {\ttfamily runspec.dat}. See \path{sfincs/fortran/utils/sfincsScan_21} for an example file.
-}
+\ssparam{NthetaMaxFactor}
+{real}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Maximum factor by which \Ntheta~will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{NthetaMinFactor}
+{real}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Minimum factor by which \Ntheta~will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{NthetaNumRuns}
+{integer}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Maximum number of values of \Ntheta~which will be used in a convergence scan.  Only odd integers can be used
+for \Ntheta, so the actual number of \Ntheta~values used in the scan may be less than {\ttfamily NthetaNumRuns}.}
+
+\myhrule
+
+\ssparam{NzetaMaxFactor}
+{real}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Maximum factor by which \Nzeta~will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{NzetaMinFactor}
+{real}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Minimum factor by which \Nzeta~will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{NzetaNumRuns}
+{integer}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Maximum number of values of \Nzeta~which will be used in a convergence scan.  Only odd integers can be used
+for \Nzeta, so the actual number of \Nzeta~values used in the scan may be less than {\ttfamily NzetaNumRuns}.}
+
+\myhrule
+
+\ssparam{NxiMaxFactor}
+{real}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Maximum factor by which \Nxi~will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{NxiMinFactor}
+{real}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Minimum factor by which \Nxi~will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{NxiNumRuns}
+{integer}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Maximum number of values of \Nxi~which will be used in a convergence scan.  Only integers can be used
+for \Nxi, so the actual number of \Nxi~values used in the scan may be less than {\ttfamily NxiNumRuns}.}
+
+\myhrule
+
+\ssparam{NxMaxFactor}
+{real}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Maximum factor by which \Nx~will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{NxMinFactor}
+{real}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Minimum factor by which \Nx~will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{NxNumRuns}
+{integer}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Maximum number of values of \Nx~which will be used in a convergence scan.  Only integers can be used
+for \Nx, so the actual number of \Nx~values used in the scan may be less than {\ttfamily NxNumRuns}.}
+
+\myhrule
+
+\ssparam{solverToleranceMaxFactor}
+{real}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Maximum factor by which {\ttfamily solverTolerance} will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{solverToleranceMinFactor}
+{real}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Minimum factor by which {\ttfamily solverTolerance} will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{solverToleranceNumRuns}
+{integer}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Number of values of {\ttfamily solverTolerance} which will be used in a convergence scan.}
+
+\myhrule
+
+\ssparam{NLMaxFactor}
+{real}
+{Only when {\ttfamily collisionOperator} = 0 and \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Maximum factor by which \NL~will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{NLMinFactor}
+{real}
+{Only when {\ttfamily collisionOperator} = 0 and \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Minimum factor by which \NL~will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{NLNumRuns}
+{integer}
+{Only when {\ttfamily collisionOperator} = 0 and \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Maximum number of values of \NL~which will be used in a convergence scan.  Only integers can be used
+for \NL, so the actual number of \NL~values used in the scan may be less than {\ttfamily NLNumRuns}.}
+
+\myhrule
+
+\ssparam{NxPotentialsPerVthMaxFactor}
+{real}
+{Only when {\ttfamily collisionOperator} = 0, {\ttfamily xGridScheme} $<$ 5, and \sfincsScan~is run with {\ttfamily scanType} = 1.
+Since the recommended value of {\ttfamily xGridScheme} is 5, this parameter is basically obsolete.}
+{Maximum factor by which {\ttfamily NxPotentialsPerVth} will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{NxPotentialsPerVthMinFactor}
+{real}
+{Only when {\ttfamily collisionOperator} = 0, {\ttfamily xGridScheme} $<$ 5, and \sfincsScan~is run with {\ttfamily scanType} = 1.
+Since the recommended value of {\ttfamily xGridScheme} is 5, this parameter is basically obsolete.}
+{Minimum factor by which {\ttfamily NxPotentialsPerVth} will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{NxPotentialsPerVthNumRuns}
+{integer}
+{Only when {\ttfamily collisionOperator} = 0, {\ttfamily xGridScheme} $<$ 5, and \sfincsScan~is run with {\ttfamily scanType} = 1.
+Since the recommended value of {\ttfamily xGridScheme} is 5, this parameter is basically obsolete.}
+{Number of values of {\ttfamily NxPotentialsPerVth} which will be used in a convergence scan.}
+
+\myhrule
+
+\ssparam{xMaxMaxFactor}
+{real}
+{Only when {\ttfamily collisionOperator} = 0, {\ttfamily xGridScheme} $<$ 5, and \sfincsScan~is run with {\ttfamily scanType} = 1.
+Since the recommended value of {\ttfamily xGridScheme} is 5, this parameter is basically obsolete.}
+{Maximum factor by which {\ttfamily xMax} will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{xMaxMinFactor}
+{real}
+{Only when {\ttfamily collisionOperator} = 0, {\ttfamily xGridScheme} $<$ 5, and \sfincsScan~is run with {\ttfamily scanType} = 1.
+Since the recommended value of {\ttfamily xGridScheme} is 5, this parameter is basically obsolete.}
+{Minimum factor by which {\ttfamily xMax} will be multiplied in a convergence scan.}
+
+\myhrule
+
+\ssparam{xMaxNumRuns}
+{integer}
+{Only when {\ttfamily collisionOperator} = 0, {\ttfamily xGridScheme} $<$ 5, and \sfincsScan~is run with {\ttfamily scanType} = 1.
+Since the recommended value of {\ttfamily xGridScheme} is 5, this parameter is basically obsolete.}
+{Number of values of {\ttfamily xMax} which will be used in a convergence scan.}
+
+\subsection{Parameters related only to {\ttfamily scanType} = 2 (scans of radial electric field).}
+
+In this scan of the radial electric field, the values of electric field used
+will always be uniformly (linearly) spaced.  
+Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHatMax}, {\ttfamily dPhiHatdpsiNMax}, {\ttfamily dPhiHatdrHatMax}, and {\ttfamily dPhiHatdrNMax}
+will be used, depending on\\
+{\ttfamily inputRadialCoordinateForGradients}.
+Similarly, exactly 1 of the 4 variables
+{\ttfamily dPhiHatdpsiHatMin}, {\ttfamily dPhiHatdpsiNMin}, {\ttfamily dPhiHatdrHatMin}, and {\ttfamily dPhiHatdrNMin}
+will be used.
+
+\myhrule
+
+\ssparam{NErs}
+{integer}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 2.}
+{Number of values of radial electric field to consider in a scan.}
+
+\myhrule
+
+\ssparam{dPhiHatdpsiHatMax}
+{real}
+{Only when {\ttfamily inputRadialCoordinateForGradients} = 0 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
+{Maximum value of {\ttfamily dPhiHatdpsiHat} to use in the scan.}
+
+\myhrule
+
+\ssparam{dPhiHatdpsiHatMin}
+{real}
+{Only when {\ttfamily inputRadialCoordinateForGradients} = 0 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
+{Minimum value of {\ttfamily dPhiHatdpsiHat} to use in the scan.}
+
+\myhrule
+
+\ssparam{dPhiHatdpsiNMax}
+{real}
+{Only when {\ttfamily inputRadialCoordinateForGradients} = 1 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
+{Maximum value of {\ttfamily dPhiHatdpsiN} to use in the scan.}
+
+\myhrule
+
+\ssparam{dPhiHatdpsiNMin}
+{real}
+{Only when {\ttfamily inputRadialCoordinateForGradients} = 1 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
+{Minimum value of {\ttfamily dPhiHatdpsiN} to use in the scan.}
+
+\myhrule
+
+\ssparam{dPhiHatdrHatMax}
+{real}
+{Only when {\ttfamily inputRadialCoordinateForGradients} = 2 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
+{Maximum value of {\ttfamily dPhiHatdrHat} to use in the scan.}
+
+\myhrule
+
+\ssparam{dPhiHatdrHatMin}
+{real}
+{Only when {\ttfamily inputRadialCoordinateForGradients} = 2 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
+{Minimum value of {\ttfamily dPhiHatdrHat} to use in the scan.}
+
+\myhrule
+
+\ssparam{dPhiHatdrNMax}
+{real}
+{Only when {\ttfamily inputRadialCoordinateForGradients} = 3 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
+{Maximum value of {\ttfamily dPhiHatdrN} to use in the scan.}
+
+\myhrule
+
+\ssparam{dPhiHatdrNMin}
+{real}
+{Only when {\ttfamily inputRadialCoordinateForGradients} = 3 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
+{Minimum value of {\ttfamily dPhiHatdrN} to use in the scan.}
+
+\subsection{Parameters related only to {\ttfamily scanType} = 3 (scans of an arbitrary input parameter).}
+
+\ssparam{scanVariable}
+{string. Must be of the fortran namelist parameters that takes an integer or real value. Case-insensitive.}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 3.}
+{Name of the variable to scan in a {\ttfamily scanType} = 3 scan.}
+
+\myhrule
+
+\ssparam{scanVariableMax}
+{real}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 3.}
+{Maximum value of {\ttfamily scanVariable} to use in a {\ttfamily scanType} = 3 scan.}
+
+\myhrule
+
+\ssparam{scanVariableMin}
+{real}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 3.}
+{Minimum value of {\ttfamily scanVariable} to use in a {\ttfamily scanType} = 3 scan.}
+
+\myhrule
+
+\ssparam{scanVariableN}
+{integer}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 3.}
+{Number of values of {\ttfamily scanVariable} to use in a {\ttfamily scanType} = 3 scan.}
+
+\myhrule
+
+\ssparam{scanVariableScale}
+{string. Must be `linear', `lin', 'logarithmic', or 'log'}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 3.}
+{Whether to space the values of {\ttfamily scanVariable} in a linear or logarithmic manner.
+The settings `linear' and `lin' have identical behavior. 
+The settings `logarithmic' and `log' have identical behavior.}
+
+\myhrule
+
+\subsection{Parameters related only to {\ttfamily scanType} = 4 or 5 (radial scans).}
+
+Notice that exactly 1 of the 4 variables {\ttfamily psiHat\_max}, {\ttfamily psiN\_max}, {\ttfamily rHat\_max}, and {\ttfamily rN\_max}
+will be used, depending on
+{\ttfamily inputRadialCoordinate}.
+Similarly, exactly 1 of the 4 variables
+{\ttfamily psiHat\_min}, {\ttfamily psiN\_min}, {\ttfamily rHat\_min}, and {\ttfamily rN\_min}
+will be used.
+
+\myhrule
+
+\ssparam{profilesScheme}
+{integer}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
+{How to specify the profiles of density, temperature, and (when  {\ttfamily scanType} = 5)
+the range of radial electric field to consider.\\
+
+{\ttfamily profilesScheme} = 1: Read a `{\ttfamily profiles}' file which contains the input profiles on a grid in one of the 4 available radial coordinates.\\
+
+{\ttfamily profilesScheme} = 2: Read a `{\ttfamily profiles}' file which contains the input profiles expressed as polynomials in one of the 4 available radial coordinates.
+}
+
+\myhrule
+
+\ssparam{Nradius}
+{integer}
+{Only when \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
+{Maximum number of values of minor radius to consider in the scan.
+Depending on {\ttfamily geometryScheme} and {\ttfamily VMECRadialOption},
+it may be that only surfaces available in the magnetic equilibrium file will be
+used, in which case fewer than {\ttfamily Nradius} radii may be used.}
+
+\myhrule
+
+\ssparam{psiHat\_max}
+{real}
+{Only when {\ttfamily inputRadialCoordinate} = 0 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
+{Maximum value of {\ttfamily psiHat} to use in the scan.}
+
+\myhrule
+
+\ssparam{psiHat\_min}
+{real}
+{Only when {\ttfamily inputRadialCoordinate} = 0 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
+{Minimum value of {\ttfamily psiHat} to use in the scan.}
+
+\myhrule
+
+\ssparam{psiN\_max}
+{real}
+{Only when {\ttfamily inputRadialCoordinate} = 1 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
+{Maximum value of {\ttfamily psiN} to use in the scan.}
+
+\myhrule
+
+\ssparam{psiN\_min}
+{real}
+{Only when {\ttfamily inputRadialCoordinate} = 1 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
+{Minimum value of {\ttfamily psiN} to use in the scan.}
+
+\myhrule
+
+\ssparam{rHat\_max}
+{real}
+{Only when {\ttfamily inputRadialCoordinate} = 2 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
+{Maximum value of {\ttfamily rHat} to use in the scan.}
+
+\myhrule
+
+\ssparam{rHat\_min}
+{real}
+{Only when {\ttfamily inputRadialCoordinate} = 2 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
+{Minimum value of {\ttfamily rHat} to use in the scan.}
+
+\myhrule
+
+\ssparam{rN\_max}
+{real}
+{Only when {\ttfamily inputRadialCoordinate} = 3 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
+{Maximum value of {\ttfamily rN} to use in the scan.}
+
+\myhrule
+
+\ssparam{rN\_min}
+{real}
+{Only when {\ttfamily inputRadialCoordinate} = 3 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
+{Minimum value of {\ttfamily rN} to use in the scan.}
 

### 2015-05-31 doc/manual/runs.tex
Added to user manual

@@
 
+There is not a significant difference in computational cost between these models.
+
+\section{Quasineutrality and variation of the electrostatic potential on the flux surface}
+
+One choice you should consider in setting up a computation is whether or not
+to include variation on the flux surface of the electrostatic potential, $\Phi_1(\theta,\zeta)$.
+Such variation does occur to some degree in a real plasma, but it is 
+neglected in analytical theory and in many codes such as \dkes.  It can be proved
+that including $\Phi_1$ has no effect on the particle or heat fluxes, parallel flows, or bootstrap current
+when $E_r=0$, but generally there can be some difference when $E_r \ne 0$.
+(There is a subtlety in showing that the heat flux is the same with and without $\Phi_1$,
+discussed in the notes 20150325-01 in \path{sfincs/doc/}.)
+
+In \sfincs, you can choose whether or not to
+include $\Phi_1$ using the paramter {\ttfamily includePhi1} in the {\ttfamily physicsParameters}
+namelist.  If and only if $\Phi_1$ is included, a quasineutrality equation is solved
+at each point on the flux surface.  Due to these extra unknowns ($\Phi_1$) and extra equations
+(quasineutrality), the system matrix is slightly larger when {\ttfamily includePhi1} is \true.
+Specifically, the number of rows and columns are each increased by \Ntheta$\times$\Nzeta$+1$.  This increase is miniscule compared
+to the number of rows and columns associated with the kinetic equation, which depends not only on real space
+but also on velocity space and species.  Thus, there is very little extra computational cost associated
+with {\ttfamily includePhi1}.  You may wish to set {\ttfamily includePhi1} = \true~when
+using \sfincs~to model an experiment, and set  {\ttfamily includePhi1} = \false~when
+comparing \sfincs~with analytic theory or with another code that does not include $\Phi_1$.
+
 \section{Parallelization}

### 2015-05-31 doc/manual/inputParameters.tex
Example input files pared down to non-default parameters, since documentation is now in user manual.

@@
 {4.5694e-3}
-{Always}
+{Whenever {\ttfamily RHSMode} == 1.}
 {Roughly speaking, {\ttfamily Delta} is $\rho_*$ at the reference parameters. The precise definition is
@@
 The default value {\ttfamily alpha} = 1.0 also corresponds to $\bar{T}$ = 1 eV and $\bar{\Phi}$ = 1 V.
-}
-
-\myhrule
-
-\param{EParallelHat}
-{real}
-{0.0}
-{Whenever {\ttfamily RHSMode} == 1}
-{Inductive parallel electric field:
-\begin{equation}
-\mbox{\ttfamily EParallelHat} = \left< \vect{E}\cdot\vect{B}\right> \frac{\bar{R}}{\bar{\Phi}\bar{B}}
-\end{equation}
-(in both Gaussian and SI units) where $\left< \ldots \right>$ denotes a flux surface average,
-$\vect{E}$ and $\vect{B}$ are the electric and magnetic field vectors, and 
-quantities with a bar are the normalization reference parameters discussed in section \ref{sec:normalizations}.}
-
-\myhrule
-
-\param{dPhiHatdpsiHat}
-{real}
-{0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 0}
-{The derivative of the electrostatic potential with respect to the radial coordinate $\hat{\psi}$,
-i.e. the radial electric field up to a constant.
-Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
-will be used, depending on\\
-{\ttfamily inputRadialCoordinateForGradients}.
-}
-
-\myhrule
-
-\param{dPhiHatdpsiN}
-{real}
-{0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 1}
-{The derivative of the electrostatic potential with respect to the radial coordinate $\psi_N$,
-i.e. the radial electric field up to a constant.
-Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
-will be used, depending on\\
-{\ttfamily inputRadialCoordinateForGradients}.
-}
-
-\myhrule
-
-\param{dPhiHatdrHat}
-{real}
-{0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 2}
-{The derivative of the electrostatic potential with respect to the radial coordinate $\hat{r}$,
-i.e. the radial electric field up to a constant.
-Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
-will be used, depending on\\
-{\ttfamily inputRadialCoordinateForGradients}.
-}
-
-\myhrule
-
-\param{dPhiHatdrN}
-{real}
-{0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 3}
-{The derivative of the electrostatic potential with respect to the radial coordinate $r_N$,
-i.e. the radial electric field up to a constant.
-Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
-will be used, depending on\\
-{\ttfamily inputRadialCoordinateForGradients}.
 }
@@
 
+\param{EParallelHat}
+{real}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1}
+{Inductive parallel electric field:
+\begin{equation}
+\mbox{\ttfamily EParallelHat} = \left< \vect{E}\cdot\vect{B}\right> \frac{\bar{R}}{\bar{\Phi}\bar{B}}
+\end{equation}
+(in both Gaussian and SI units) where $\left< \ldots \right>$ denotes a flux surface average,
+$\vect{E}$ and $\vect{B}$ are the electric and magnetic field vectors, and 
+quantities with a bar are the normalization reference parameters discussed in section \ref{sec:normalizations}.}
+
+\myhrule
+
+\param{dPhiHatdpsiHat}
+{real}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 0}
+{The derivative of the electrostatic potential with respect to the radial coordinate $\hat{\psi}$,
+i.e. the radial electric field up to a constant.
+Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
+will be used, depending on\\
+{\ttfamily inputRadialCoordinateForGradients}.
+}
+
+\myhrule
+
+\param{dPhiHatdpsiN}
+{real}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 1}
+{The derivative of the electrostatic potential with respect to the radial coordinate $\psi_N$,
+i.e. the radial electric field up to a constant.
+Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
+will be used, depending on\\
+{\ttfamily inputRadialCoordinateForGradients}.
+}
+
+\myhrule
+
+\param{dPhiHatdrHat}
+{real}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 2}
+{The derivative of the electrostatic potential with respect to the radial coordinate $\hat{r}$,
+i.e. the radial electric field up to a constant.
+Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
+will be used, depending on\\
+{\ttfamily inputRadialCoordinateForGradients}.
+}
+
+\myhrule
+
+\param{dPhiHatdrN}
+{real}
+{0.0}
+{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 3}
+{The derivative of the electrostatic potential with respect to the radial coordinate $r_N$,
+i.e. the radial electric field up to a constant.
+Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
+will be used, depending on\\
+{\ttfamily inputRadialCoordinateForGradients}.
+}
+
+\myhrule
+
 \param{collisionOperator}
@@
 \PETScParam{-mat\_mumps\_icntl\_4 {\normalfont \ttfamily$<$integer$>$}}
-{How much diagnostic information will be printed by \mumps. Default is 0. Set to 2 or 3 to print out useful diagnostic information about the memory required for factorizing the preconditioner.}
+{How much diagnostic information will be printed by \mumps. Default is 3, causing extensive diagnostic information to be printed 
+to standard output about the memory required for factorizing the preconditioner.
+Set this parameter to 0 to suppress this output from mumps.}

### 2015-05-31 fortran/version3/globalVariables.F90

@@ -63,11 +63,13 @@ module globalVariables
   integer, parameter :: maxNumSpecies = 100
   integer, parameter :: speciesNotInitialized = -9999
   integer :: NSpecies = -9999
-  PetscScalar, dimension(maxNumSpecies) :: Zs = 1, mHats = 1.0, NHats = 1.0, THats = 1.0
-  PetscScalar, dimension(maxNumSpecies) :: dNHatdpsiHats = 0, dTHatdpsiHats = 0
-  PetscScalar, dimension(maxNumSpecies) :: dNHatdpsiNs   = 0, dTHatdpsiNs   = 0
-  PetscScalar, dimension(maxNumSpecies) :: dNHatdrHats   = 0, dTHatdrHats   = 0
-  PetscScalar, dimension(maxNumSpecies) :: dNHatdrNs     = 0, dTHatdrNs     = 0
+  ! For all the array variables declared in the next 5 lines, default values
+  ! are set in readInput.F90
+  PetscScalar, dimension(maxNumSpecies) :: Zs, mHats, NHats, THats
+  PetscScalar, dimension(maxNumSpecies) :: dNHatdpsiHats, dTHatdpsiHats
+  PetscScalar, dimension(maxNumSpecies) :: dNHatdpsiNs,   dTHatdpsiNs
+  PetscScalar, dimension(maxNumSpecies) :: dNHatdrHats,   dTHatdrHats
+  PetscScalar, dimension(maxNumSpecies) :: dNHatdrNs,     dTHatdrNs
 
   ! ********************************************************
   ! ********************************************************

### 2015-06-03 doc/manual/inputParameters.tex
scan21 comments

@@
 
-{\ttfamily scanType} = 21: Read in a list of requested runs from a file {\ttfamily runspec.dat}. See \path{sfincs/fortran/utils/sfincsScan_21} for an example file.
+{\ttfamily scanType} = 21: Read in a list of requested runs from a
+file {\ttfamily runspec.dat}. See
+\path{sfincs/fortran/utils/sfincsScan_21} for an example file. If the
+file has a different name than {\ttfamily runspec.dat}, for instance
+``thefilename.dat'', this name can be
+specified by adding the line\\
+{\ttfamily
+!ss runSpecFile = thefilename.dat\\
+}
 }

### 2015-06-04 doc/manual/inputParameters.tex
Added hyperlinks to manual

--- 2015-05-31 doc/manual/inputParameters.tex
@@
 
-\newcommand{\param}[5]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries #1}\\{\it Type}: #2\\{\it Default}: #3\\{\it When it matters}: #4\\{\it Meaning}: #5}}
-\newcommand{\ssparam}[4]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries #1}\\{\it Type}: #2\\{\it When it matters}: #3\\{\it Meaning}: #4}}
+\newcommand{\param}[5]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries \hypertarget{#1}{#1}}\\{\it Type}: #2\\{\it Default}: #3\\{\it When it matters}: #4\\{\it Meaning}: #5}}
+\newcommand{\ssparam}[4]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries \hypertarget{#1}{#1}}\\{\it Type}: #2\\{\it When it matters}: #3\\{\it Meaning}: #4}}
 \newcommand{\PETScParam}[2]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries #1}\\{\it Meaning}: #2}}
@@
 
-{\ttfamily RHSMode=1}: Solve for a single right-hand side.\\
-
-{\ttfamily RHSMode=2}: Solve for 3 right-hand sides to get the 3x3 transport matrix. Presently implemented only for 1 species.\\
-
-{\ttfamily RHSMode=3}: Solve for the 2x2 monoenergetic transport coefficients. When this option is chosen, {\ttfamily Nx} is set to 1 and only 1 species is used.
+{\ttfamily RHSMode} = 1: Solve for a single right-hand side.\\
+
+{\ttfamily RHSMode} = 2: Solve for 3 right-hand sides to get the 3$\times$3 transport matrix. Presently implemented only for 1 species.\\
+
+{\ttfamily RHSMode} = 3: Solve for the 2$\times$2 monoenergetic transport coefficients. When this option is chosen, \Nx~is set to 1 and only 1 species is used.
 }
@@
 {``sfincsMatrices''}
-{Only when {\ttfamily saveMatlabOutput == .true.}}
+{Only when \parlink{saveMatlabOutput} == \true.}
 {Start of the filenames which will be used for Matlab output.}
@@
 linear system will be saved in PETSc's binary format.  The preconditioner matrix will also
-be saved if {\ttfamily tryIterativeSolver == .true.}
-}
+be saved if \parlink{useIterativeLinearSolver} == \true.}
 
@@
 {``sfincsBinary''}
-{Only when {\ttfamily saveMatricesAndVectorsInBinary == .true.}}
+{Only when \parlink{saveMatricesAndVectorsInBinary} == \true.}
 {Start of the filenames which will be used for binary output.}
@@
 {If this parameter is false, the system of equations will not actually be solved.
- Sometimes it can be useful to set this parameter to {\ttfamily .false.} when debugging.
+ Sometimes it can be useful to set this parameter to \false~when debugging.
 }
@@
 B(\theta,\zeta)/\bar{B} 
-&=& (\mbox{\ttfamily B0OverBBar}) [1 + (\mbox{\ttfamily epsilon\_t}) \cos(\theta) \\
-&& + (\mbox{\ttfamily epsilon\_h}) \cos((\mbox{\ttfamily helicity\_l}) \theta - (\mbox{\ttfamily helicity\_n}) \zeta) \nonumber \\
-&&  + (\mbox{\ttfamily epsilon\_antisymm}) \nonumber \\
-&& \times \sin((\mbox{ \ttfamily helicity\_antisymm\_l}) \theta - (\mbox{\ttfamily helicity\_antisymm\_n}) \zeta)] \nonumber
+&=& (\mbox{\parlink{B0OverBBar}}) [1 + (\mbox{\parlink{epsilon\_t}}) \cos(\theta) \\
+&& + (\mbox{\parlink{epsilon\_h}}) \cos((\mbox{\parlink{helicity\_l}}) \theta - (\mbox{\parlink{helicity\_n}}) \zeta) \nonumber \\
+&&  + (\mbox{\parlink{epsilon\_antisymm}}) \nonumber \\
+&& \times \sin((\mbox{\parlink{helicity\_antisymm\_l}}) \theta - (\mbox{\parlink{helicity\_antisymm\_n}}) \zeta)] \nonumber
 \end{eqnarray}
@@
 
-{\ttfamily geometryScheme}==5: Read the {\ttfamily vmec wout} file specified in {\ttfamily equilibriumFile} below. The file can be
+{\ttfamily geometryScheme}==5: Read the {\ttfamily vmec wout} file specified in \parlink{equilibriumFile} below. The file can be
        either ASCII format or \netCDF~format. (\sfincs~will auto-detect the format.).\\
 
-{\ttfamily geometryScheme}==11: Read the IPP {\ttfamily .bc} format Boozer-coordinate file specified in {\ttfamily equilibriumFile} below. 
+{\ttfamily geometryScheme}==11: Read the IPP {\ttfamily .bc} format Boozer-coordinate file specified in \parlink{equilibriumFile} below. 
 The file is assumed to be stellarator-symmetric.\\
 
-{\ttfamily geometryScheme}==12: Read the IPP {\ttfamily .bc} format Boozer-coordinate file specified in {\ttfamily equilibriumFile} below. 
+{\ttfamily geometryScheme}==12: Read the IPP {\ttfamily .bc} format Boozer-coordinate file specified in \parlink{equilibriumFile} below. 
 The file is assumed to be stellarator-\emph{asymmetric}.
@@
 {3}
-{When {\ttfamily geometryScheme} == 1, 5, 11, or 12}
+{When \parlink{geometryScheme} == 1, 5, 11, or 12}
 {Which radial coordinate to use to specify the flux surface for a single calculation,
 or to specify the range of flux surfaces for a radial scan. 
-(Regardless of the value of this parameter, when {\ttfamily geometryScheme} == 2, 3, or 4, the flux surface used will be {\ttfamily rN}=0.5.)
+(Regardless of the value of this parameter, when \parlink{geometryScheme} == 2, 3, or 4, the flux surface used will be {\ttfamily rN}=0.5.)
 See section \ref{sec:radialCoordinates}
@@
 
-{\ttfamily inputRadialCoordinate}==0: Use the flux surface specified by {\ttfamily psiHat\_wish}
-for a single run, and use the range specified by {\ttfamily psiHat\_min} and {\ttfamily psiHat\_max} for radial scans.\\
-
-{\ttfamily inputRadialCoordinate}==1: Use the flux surface specified by {\ttfamily psiN\_wish}
-for a single run, and use the range specified by {\ttfamily psiN\_min} and {\ttfamily psiN\_max} for radial scans.\\
-
-{\ttfamily inputRadialCoordinate}==2: Use the flux surface specified by {\ttfamily rHat\_wish}
-for a single run, and use the range specified by {\ttfamily rHat\_min} and {\ttfamily rHat\_max} for radial scans.\\
-
-{\ttfamily inputRadialCoordinate}==3: Use the flux surface specified by {\ttfamily rN\_wish}
-for a single run, and use the range specified by {\ttfamily rN\_min} and {\ttfamily rN\_max} for radial scans. \\
+{\ttfamily inputRadialCoordinate}==0: Use the flux surface specified by \parlink{psiHat\_wish}
+for a single run, and use the range specified by \parlink{psiHat\_min} and \parlink{psiHat\_max} for radial scans.\\
+
+{\ttfamily inputRadialCoordinate}==1: Use the flux surface specified by \parlink{psiN\_wish}
+for a single run, and use the range specified by \parlink{psiN\_min} and \parlink{psiN\_max} for radial scans.\\
+
+{\ttfamily inputRadialCoordinate}==2: Use the flux surface specified by \parlink{rHat\_wish}
+for a single run, and use the range specified by \parlink{rHat\_min} and \parlink{rHat\_max} for radial scans.\\
+
+{\ttfamily inputRadialCoordinate}==3: Use the flux surface specified by \parlink{rN\_wish}
+for a single run, and use the range specified by \parlink{rN\_min} and \parlink{rN\_max} for radial scans. \\
 
@@
 {2}
-{Whenever {\ttfamily RHSMode}==1.}
+{Whenever \parlink{RHSMode}==1.}
 {Which radial coordinate is used to use to specify the input gradients of density, temperature, and electrostatic potential,
@@
 
-{\ttfamily inputRadialCoordinateForGradients}==0: Density gradients are specified by {\ttfamily dnHatdpsiHats},
-temperature gradients are specified by {\ttfamily dTHatdpsiHats}, a single $E_r$ is specified by {\ttfamily dPhiHatdpsiHat},
-and the range of an $E_r$ scan is specified by {\ttfamily dPhiHatdpsiHatMin-dPhiHatdpsiHatMax}.\\
-
-{\ttfamily inputRadialCoordinateForGradients}==1: Density gradients are specified by {\ttfamily dnHatdpsiNs},
-temperature gradients are specified by {\ttfamily dTHatdpsiNs}, a single $E_r$ is specified by {\ttfamily dPhiHatdpsiN},
-and the range of an $E_r$ scan is specified by {\ttfamily dPhiHatdpsiNMin-dPhiHatdpsiNMax}.\\
-
-{\ttfamily inputRadialCoordinateForGradients}==2: Density gradients are specified by {\ttfamily dnHatdrHats},
-temperature gradients are specified by {\ttfamily dTHatdrHats}, a single $E_r$ is specified by {\ttfamily dPhiHatdrHat},
-and the range of an $E_r$ scan is specified by {\ttfamily dPhiHatdrHatMin-dPhiHatdrHatMax}.\\
-
-{\ttfamily inputRadialCoordinateForGradients}==3: Density gradients are specified by {\ttfamily dnHatdrNs},
-temperature gradients are specified by {\ttfamily dTHatdrNs}, a single $E_r$ is specified by {\ttfamily dPhiHatdrN},
-and the range of an $E_r$ scan is specified by {\ttfamily dPhiHatdrNMin-dPhiHatdrNMax}.\\
+{\ttfamily inputRadialCoordinateForGradients}==0: Density gradients are specified by \parlink{dnHatdpsiHats},
+temperature gradients are specified by \parlink{dTHatdpsiHats}, a single $E_r$ is specified by \parlink{dPhiHatdpsiHat},
+and the range of an $E_r$ scan is specified by \parlink{dPhiHatdpsiHatMin}-\parlink{dPhiHatdpsiHatMax}.\\
+
+{\ttfamily inputRadialCoordinateForGradients}==1: Density gradients are specified by \parlink{dnHatdpsiNs},
+temperature gradients are specified by \parlink{dTHatdpsiNs}, a single $E_r$ is specified by \parlink{dPhiHatdpsiN},
+and the range of an $E_r$ scan is specified by \parlink{dPhiHatdpsiNMin}-\parlink{dPhiHatdpsiNMax}.\\
+
+{\ttfamily inputRadialCoordinateForGradients}==2: Density gradients are specified by \parlink{dnHatdrHats},
+temperature gradients are specified by \parlink{dTHatdrHats}, a single $E_r$ is specified by \parlink{dPhiHatdrHat},
+and the range of an $E_r$ scan is specified by \parlink{dPhiHatdrHatMin}-\parlink{dPhiHatdrHatMax}.\\
+
+{\ttfamily inputRadialCoordinateForGradients}==3: Density gradients are specified by \parlink{dnHatdrNs},
+temperature gradients are specified by \parlink{dTHatdrNs}, a single $E_r$ is specified by \parlink{dPhiHatdrN},
+and the range of an $E_r$ scan is specified by \parlink{dPhiHatdrNMin}-\parlink{dPhiHatdrNMax}.\\
 
@@
 {-1}
-{Only when {\ttfamily inputRadialCoordinate} == 0 and {\ttfamily geometryScheme} == 1, 5, 11, or 12.}
+{Only when \parlink{inputRadialCoordinate} == 0 and \parlink{geometryScheme} == 1, 5, 11, or 12.}
 {Requested flux surface for the computation.  See section \ref{sec:radialCoordinates}
@@
 {0.25}
-{Only when {\ttfamily inputRadialCoordinate} == 1 and {\ttfamily geometryScheme} == 1, 5, 11, or 12.}
+{Only when \parlink{inputRadialCoordinate} == 1 and \parlink{geometryScheme} == 1, 5, 11, or 12.}
 {Requested flux surface for the computation.  See section \ref{sec:radialCoordinates}
@@
 {-1}
-{Only when {\ttfamily inputRadialCoordinate} == 2 and {\ttfamily geometryScheme} == 1, 5, 11, or 12.}
+{Only when \parlink{inputRadialCoordinate} == 2 and \parlink{geometryScheme} == 1, 5, 11, or 12.}
 {Requested flux surface for the computation.  See section \ref{sec:radialCoordinates}
@@
 {0.5}
-{Only when {\ttfamily inputRadialCoordinate} == 3 and {\ttfamily geometryScheme} == 1, 5, 11, or 12.}
+{Only when \parlink{inputRadialCoordinate} == 3 and \parlink{geometryScheme} == 1, 5, 11, or 12.}
 {Requested flux surface for the computation.  See section \ref{sec:radialCoordinates}
@@
 {1.0}
-{Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily B0OverBBar} will be set according to the requested {\ttfamily geometryScheme}.}
+{Only when \parlink{geometryScheme} == 1. Otherwise, {\ttfamily B0OverBBar} will be set according to the requested \parlink{geometryScheme}.}
 {Magnitude of the (0,0) Boozer harmonic of the magnetic field strength (equivalent to $\left< B^3\right>/\left<B^2\right>$), normalized by $\bar{B}$.}
@@
 {3.7481}
-{Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily GHat} will be set according to the requested {\ttfamily geometryScheme}.}
+{Only when \parlink{geometryScheme} == 1. Otherwise, {\ttfamily GHat} will be set according to the requested \parlink{geometryScheme}.}
 {$G$ is $(c/2)\times$ the poloidal current outside the flux
@@
 {0.0}
-{Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily IHat} will be set according to the requested {\ttfamily geometryScheme}.}
+{Only when \parlink{geometryScheme} == 1. Otherwise, {\ttfamily IHat} will be set according to the requested \parlink{geometryScheme}.}
 {$I$ is $(c/2)\times$ the toroidal current inside the flux
@@
 {0.4542}
-{Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily iota} will be set according to the requested {\ttfamily geometryScheme}.}
+{Only when \parlink{geometryScheme} == 1. Otherwise, {\ttfamily iota} will be set according to the requested \parlink{geometryScheme}.}
 {Rotational transform (rationalized), equivalent to $1/q$ where $q$ is the safety factor.}
@@
 {-0.07053}
-{Only when {\ttfamily geometryScheme} == 1.}
+{Only when \parlink{geometryScheme} == 1.}
 {Toroidal variation in $B$, as defined by (\ref{eq:Bmodel}).}
@@
 {0.05067}
-{Only when {\ttfamily geometryScheme} == 1.}
+{Only when \parlink{geometryScheme} == 1.}
 {Helical variation in $B$, as defined by (\ref{eq:Bmodel}).}
@@
 {0.0}
-{Only when {\ttfamily geometryScheme} == 1.}
+{Only when \parlink{geometryScheme} == 1.}
 {Stellarator-antisymmetric variation in $B$, as defined by (\ref{eq:Bmodel}).}
@@
 {2}
-{Only when {\ttfamily geometryScheme} == 1.}
+{Only when \parlink{geometryScheme} == 1.}
 {Poloidal mode number of the helical variation in $B$, as defined by (\ref{eq:Bmodel}).}
@@
 {10}
-{Only when {\ttfamily geometryScheme} == 1.}
+{Only when \parlink{geometryScheme} == 1.}
 {Toroidal mode number of the helical variation in $B$, as defined by (\ref{eq:Bmodel}).}
@@
 {1}
-{Only when {\ttfamily geometryScheme} == 1.}
+{Only when \parlink{geometryScheme} == 1.}
 {Poloidal mode number of the stellarator-antisymmetric variation in $B$, as defined by (\ref{eq:Bmodel}).}
@@
 {0}
-{Only when {\ttfamily geometryScheme} == 1.}
+{Only when \parlink{geometryScheme} == 1.}
 {Toroidal mode number of the stellarator-antisymmetric variation in $B$, as defined by (\ref{eq:Bmodel}).
-Note that you can create an up-down asymmetric tokamak by setting {\ttfamily helicity\_antisymm\_n}=0, {\ttfamily epsilon\_h}=0,
-and {\ttfamily epsilon\_antisymm}$>$0.}
+Note that you can create an up-down asymmetric tokamak by setting \parlink{helicity\_antisymm\_n}=0, \parlink{epsilon\_h}=0,
+and \parlink{epsilon\_antisymm}$>$0.}
 
@@
 {0.15596}
-{Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily psiAHat} will be set according to the requested {\ttfamily geometryScheme}.}
+{Only when \parlink{geometryScheme} == 1. Otherwise, {\ttfamily psiAHat} will be set according to the requested \parlink{geometryScheme}.}
 {{\ttfamily psiAHat} = $\psi_a / (\bar{B} \bar{R}^2)$
@@
 {0.5585}
-{Only when {\ttfamily geometryScheme} == 1. Otherwise, {\ttfamily aHat} will be set according to the requested {\ttfamily geometryScheme}.}
+{Only when \parlink{geometryScheme} == 1. Otherwise, {\ttfamily aHat} will be set according to the requested \parlink{geometryScheme}.}
 {The effective minor radius at the last closed flux surface, in units of $\bar{R}$.
@@
 {``''}
-{Only when {\ttfamily geometryScheme} == 5, 11, or 12.}
+{Only when \parlink{geometryScheme} == 5, 11, or 12.}
 {Filename from which to load the magnetic equilibrium, either in {\ttfamily vmec wout} ASCII or \netCDF~format, or IPP {\ttfamily .bc} format.}
@@
 {1}
-{Only when {\ttfamily geometryScheme} == 5.}
+{Only when \parlink{geometryScheme} == 5.}
 {Controls whether the nearest available flux surface in the {\ttfamily vmec wout} file is used, or whether radial interpolation is applied to the \vmec~data
@@
 {0.0}
-{Only when {\ttfamily geometryScheme} == 5, 11, or 12.}
+{Only when \parlink{geometryScheme} == 5, 11, or 12.}
 {Filters the magnetic field read from an input file.  Only Fourier modes $(m,n)$ for which $B_{m,n}$ is at least {\ttfamily min\_Bmn\_to\_load} will be included.}
@@
 Note that only one of the four parameters 
-{\ttfamily dnHatdpsiHats}, 
-{\ttfamily dnHatdpsiNs}, 
-{\ttfamily dnHatdrHats}, or
-{\ttfamily dnHatdrNs}
-will be used, depending on the value of {\ttfamily inputRadialCoordinateForGradients}
-in the {\ttfamily geometryParameters} namelist.
+\parlink{dnHatdpsiHats}, 
+\parlink{dnHatdpsiNs}, 
+\parlink{dnHatdrHats}, or
+\parlink{dnHatdrNs}
+will be used, depending on the value of \parlink{inputRadialCoordinateForGradients}
+in the {\ttfamily \hyperref[sec:geometryParameters]{geometryParameters}} namelist.
 Similarly, only one of the four parameters
-{\ttfamily dTHatdpsiHats}, 
-{\ttfamily dTHatdpsiNs}, 
-{\ttfamily dTHatdrHats}, or
-{\ttfamily dTHatdrNs}
+\parlink{dTHatdpsiHats}, 
+\parlink{dTHatdpsiNs}, 
+\parlink{dTHatdrHats}, or
+\parlink{dTHatdrNs}
 will be used.
@@
 {1.0}
-{Whenever {\ttfamily RHSMode} == 1}
+{Whenever \parlink{RHSMode} == 1}
 {Densities of each species, in units of the reference density $\bar{n}$}
@@
 {1.0}
-{Whenever {\ttfamily RHSMode} == 1}
+{Whenever \parlink{RHSMode} == 1}
 {Temperatures of each species, in units of the reference temperature $\bar{T}$}
@@
 {0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 0}
+{Whenever \parlink{RHSMode} == 1 and \parlink{inputRadialCoordinateForGradients} == 0}
 {Radial density gradients of each species, with respect to the radial coordinate $\hat{\psi}$, normalized by the reference density $\bar{n}$.}
@@
 {0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 0}
+{Whenever \parlink{RHSMode} == 1 and \parlink{inputRadialCoordinateForGradients} == 0}
 {Radial temperature gradients of each species, with respect to the radial coordinate $\hat{\psi}$, normalized by the reference temperature $\bar{T}$.}
@@
 {0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 1}
+{Whenever \parlink{RHSMode} == 1 and \parlink{inputRadialCoordinateForGradients} == 1}
 {Radial density gradients of each species, with respect to the radial coordinate $\psi_N$, normalized by the reference density $\bar{n}$.}
@@
 {0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 1}
+{Whenever \parlink{RHSMode} == 1 and \parlink{inputRadialCoordinateForGradients} == 1}
 {Radial temperature gradients of each species, with respect to the radial coordinate $\psi_N$, normalized by the reference temperature $\bar{T}$.}
@@
 {0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 2}
+{Whenever \parlink{RHSMode} == 1 and \parlink{inputRadialCoordinateForGradients} == 2}
 {Radial density gradients of each species, with respect to the radial coordinate $\hat{r}$, normalized by the reference density $\bar{n}$.}
@@
 {0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 2}
+{Whenever \parlink{RHSMode} == 1 and \parlink{inputRadialCoordinateForGradients} == 2}
 {Radial temperature gradients of each species, with respect to the radial coordinate $\hat{r}$, normalized by the reference temperature $\bar{T}$.}
@@
 {0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 3}
+{Whenever \parlink{RHSMode} == 1 and \parlink{inputRadialCoordinateForGradients} == 3}
 {Radial density gradients of each species, with respect to the radial coordinate $r_N$, normalized by the reference density $\bar{n}$.}
@@
 
-\param{dTHatrNs}
+\param{dTHatdrNs}
 {1D array of reals}
 {0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 3}
+{Whenever \parlink{RHSMode} == 1 and \parlink{inputRadialCoordinateForGradients} == 3}
 {Radial temperature gradients of each species, with respect to the radial coordinate $r_N$, normalized by the reference temperature $\bar{T}$.}
@@
 \section{The {\ttfamily physicsParameters}~namelist}
+\label{sec:physicsParameters}
 
@@
 {4.5694e-3}
-{Whenever {\ttfamily RHSMode} == 1.}
+{Whenever \parlink{RHSMode} == 1.}
 {Roughly speaking, {\ttfamily Delta} is $\rho_*$ at the reference parameters. The precise definition is
@@
 {1.0}
-{Whenever {\ttfamily RHSMode} == 1 and $E_r$ is nonzero.}
+{Whenever \parlink{RHSMode} == 1 and $E_r$ is nonzero.}
 {{\ttfamily alpha} $= e \bar{\Phi}/ \bar{T}$ (both Gaussian and SI units) where $e$ is the proton mass,
@@
 {8.330e-3}
-{Whenever {\ttfamily RHSMode} == 1}
+{Whenever \parlink{RHSMode} == 1}
 {Dimensionless collisionality at the reference parameters:
@@
 {1.0}
-{Only when {\ttfamily RHSMode} == 3.}
-{Dimensionless collisionality used in place of {\ttfamily nHats}, {\ttfamily THats}, {\ttfamily mHats}, {\ttfamily Zs}, and {\ttfamily nu\_n} for computing monoenergetic transport coefficients.
+{Only when \parlink{RHSMode} == 3.}
+{Dimensionless collisionality used in place of \parlink{nHats}, \parlink{THats}, \parlink{mHats}, \parlink{Zs}, and \parlink{nu\_n} for computing monoenergetic transport coefficients.
 See section \ref{sec:monoenergetic} for more details.}
@@
 {0.0}
-{Only when {\ttfamily RHSMode} == 3.}
+{Only when \parlink{RHSMode} == 3.}
 {Normalized radial electric field used in place of {\ttfamily dPhiHatdXXX} for computing monoenergetic transport coefficients.
@@
 {0.0}
-{Whenever {\ttfamily RHSMode} == 1}
+{Whenever \parlink{RHSMode} == 1}
 {Inductive parallel electric field:
@@
 {0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 0}
+{Whenever \parlink{RHSMode} == 1 and \parlink{inputRadialCoordinateForGradients} == 0}
 {The derivative of the electrostatic potential with respect to the radial coordinate $\hat{\psi}$,
 i.e. the radial electric field up to a constant.
-Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
+Notice that exactly 1 of the 4 variables \parlink{dPhiHatdpsiHat}, \parlink{dPhiHatdpsiN}, \parlink{dPhiHatdrHat}, and \parlink{dPhiHatdrN}
 will be used, depending on\\
-{\ttfamily inputRadialCoordinateForGradients}.
+\parlink{inputRadialCoordinateForGradients}.
 }
@@
 {0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 1}
+{Whenever \parlink{RHSMode} == 1 and \parlink{inputRadialCoordinateForGradients} == 1}
 {The derivative of the electrostatic potential with respect to the radial coordinate $\psi_N$,
 i.e. the radial electric field up to a constant.
-Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
+Notice that exactly 1 of the 4 variables \parlink{dPhiHatdpsiHat}, \parlink{dPhiHatdpsiN}, \parlink{dPhiHatdrHat}, and \parlink{dPhiHatdrN}
 will be used, depending on\\
-{\ttfamily inputRadialCoordinateForGradients}.
+\parlink{inputRadialCoordinateForGradients}.
 }
@@
 {0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 2}
+{Whenever \parlink{RHSMode} == 1 and \parlink{inputRadialCoordinateForGradients} == 2}
 {The derivative of the electrostatic potential with respect to the radial coordinate $\hat{r}$,
 i.e. the radial electric field up to a constant.
-Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
+Notice that exactly 1 of the 4 variables \parlink{dPhiHatdpsiHat}, \parlink{dPhiHatdpsiN}, \parlink{dPhiHatdrHat}, and \parlink{dPhiHatdrN}
 will be used, depending on\\
-{\ttfamily inputRadialCoordinateForGradients}.
+\parlink{inputRadialCoordinateForGradients}.
 }
@@
 {0.0}
-{Whenever {\ttfamily RHSMode} == 1 and {\ttfamily inputRadialCoordinateForGradients} == 3}
+{Whenever \parlink{RHSMode} == 1 and \parlink{inputRadialCoordinateForGradients} == 3}
 {The derivative of the electrostatic potential with respect to the radial coordinate $r_N$,
 i.e. the radial electric field up to a constant.
-Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHat}, {\ttfamily dPhiHatdpsiN}, {\ttfamily dPhiHatdrHat}, and {\ttfamily dPhiHatdrN}
+Notice that exactly 1 of the 4 variables \parlink{dPhiHatdpsiHat}, \parlink{dPhiHatdpsiN}, \parlink{dPhiHatdrHat}, and \parlink{dPhiHatdrN}
 will be used, depending on\\
-{\ttfamily inputRadialCoordinateForGradients}.
+\parlink{inputRadialCoordinateForGradients}.
 }
@@
 
-{\ttfamily constraintScheme} = -1: Automatic. If {\ttfamily collisionOperator}==0 then {\ttfamily constraintScheme} will be set to 1, 
+{\ttfamily constraintScheme} = -1: Automatic. If \parlink{collisionOperator}==0 then {\ttfamily constraintScheme} will be set to 1, 
 otherwise {\ttfamily constraintScheme} will be set to 2.\\
@@
 {\true}
-{Whenever {\ttfamily RHSMode} $<3$ and the radial electric field is nonzero.}
+{Whenever \parlink{RHSMode} $<3$ and the radial electric field is nonzero.}
 {Whether or not to include the term in the kinetic equation corresponding to a change
@@
 {\false}
-{Whenever {\ttfamily RHSMode} $<3$ and the radial electric field is nonzero.}
+{Whenever \parlink{RHSMode} $<3$ and the radial electric field is nonzero.}
 {If true, the $\vect{E}\times\vect{B}$ drift term 
@@
 {\false}
-{Whenever {\ttfamily RHSMode} == 1.}
+{Whenever \parlink{RHSMode} == 1.}
 {If false, no terms involving $\Phi_1 = \Phi-\left<\Phi\right>$ are included in the kinetic equation,
@@
 {\false}
-{Whenever {\ttfamily RHSMode} == 1.}
+{Whenever \parlink{RHSMode} == 1.}
 {If true, certain terms that are nonlinear in the unknowns will be included in the kinetic equation.
@@
 be iterated several times.
-Running with {\ttfamily nonlinear}=\true requires {\ttfamily includePhi1}=\true
+Running with {\ttfamily nonlinear}=\true requires \parlink{includePhi1}=\true
 }
@@
 {\false}
-{Whenever {\ttfamily RHSMode} == 1.}
+{Whenever \parlink{RHSMode} == 1.}
 {When true, the term $C_{ab}[ f_{Ma}, f_{Mb}]$ is included in the kinetic equation,
@@
 {0}
-{Whenever {\ttfamily RHSMode} == 1.}
+{Whenever \parlink{RHSMode} == 1.}
 {This variable controls the poloidal and magnetic drifts, and does not affect the radial magnetic drift.
@@
 {Number of grid points in the poloidal angle. 
-This parameter should be odd; see {\ttfamily forceOddNthetaAndNzeta} in this namelist.
+This parameter should be odd; see \parlink{forceOddNthetaAndNzeta} in this namelist.
 Memory and time requirements DO depend strongly on this parameter.
@@
 {Number of grid points in the toroidal angle (per identical segment of the stellarator.)  
-This parameter should be odd; see {\ttfamily forceOddNthetaAndNzeta} in this namelist.
+This parameter should be odd; see \parlink{forceOddNthetaAndNzeta} in this namelist.
 Memory and time requirements DO depend strongly on this parameter.
@@
 {1e-6}
-{Whenever {\ttfamily useIterativeLinearSolver} == \true}
+{Whenever \parlink{useIterativeLinearSolver} == \true}
 {
@@
 {4}
-{Whenever {\ttfamily collisionOperator} == 0.}
+{Whenever \parlink{collisionOperator} == 0.}
 {Number of Legendre polynomials used to represent the Rosenbluth
 potentials. This number can basically always be 4,
-since results barely change when {\ttfamily NL} is increased above this value.
+since results barely change when \parlink{NL} is increased above this value.
 Memory and time requirements do NOT depend strongly on this parameter.
@@
 {40.0}
-{Only when {\ttfamily collisionOperator} == 0 and {\ttfamily xGridScheme} $<$ 5.
-Since {\ttfamily xGridScheme} = 5 is recommended, this parameter is basically obsolete.}
+{Only when \parlink{collisionOperator} == 0 and \parlink{xGridScheme} $<$ 5.
+Since \parlink{xGridScheme} = 5 is recommended, this parameter is basically obsolete.}
 {
@@
 {5.0}
-{Only when {\ttfamily collisionOperator} == 0 and {\ttfamily xGridScheme} $<$ 5.
-Since {\ttfamily xGridScheme} = 5 is recommended, this parameter is basically obsolete.}
+{Only when \parlink{collisionOperator} == 0 and \parlink{xGridScheme} $<$ 5.
+Since \parlink{xGridScheme} = 5 is recommended, this parameter is basically obsolete.}
 {
@@
 {5}
-{Whenever {\ttfamily RHSMode} is 1 or 2.}
+{Whenever \parlink{RHSMode} is 1 or 2.}
 {Discretization scheme for the speed coordinate $x$.\\
@@
 
-{\ttfamily xGridScheme} = 3: Uniform finite differences on [0, {\ttfamily xMax}], forcing $f=0$ at {\ttfamily xMax}. 2-point stencil for interpolating to other grids.\\
-
-{\ttfamily xGridScheme} = 4: Uniform finite differences on [0, {\ttfamily xMax}], forcing $f=0$ at {\ttfamily xMax}. 4-point stencil for interpolating to other grids.\\
+{\ttfamily xGridScheme} = 3: Uniform finite differences on [0, \parlink{xMax}], forcing $f=0$ at \parlink{xMax}. 2-point stencil for interpolating to other grids.\\
+
+{\ttfamily xGridScheme} = 4: Uniform finite differences on [0, \parlink{xMax}], forcing $f=0$ at \parlink{xMax}. 4-point stencil for interpolating to other grids.\\
 
@@
 When {\ttfamily xGridScheme} = 5 or 6, then the following quantities do not matter: 
-{\ttfamily NxPotentialsPerVth}, {\ttfamily xMax}, and {\ttfamily xPotentialsGridScheme}.}
+\parlink{NxPotentialsPerVth}, \parlink{xMax}, and \parlink{xPotentialsGridScheme}.}
 
@@
 {0}
-{Whenever {\ttfamily RHSMode} is 1 or 2 and {\ttfamily xGridScheme} = 1, 2, 5, or 6.}
-{For {\ttfamily xGridScheme} = 1, 2, 5, or 6, the distribution function will be represented in terms of polynomials $P_n(x)$
+{Whenever \parlink{RHSMode} is 1 or 2 and \parlink{xGridScheme} = 1, 2, 5, or 6.}
+{For \parlink{xGridScheme} = 1, 2, 5, or 6, the distribution function will be represented in terms of polynomials $P_n(x)$
 that are orthogonal under the weight $\int_0^\infty dx\; x^k \exp(-x^2) P_n(x) P_m(x)  \propto  \delta_{n,m}$
@@
 {2}
-{Whenever {\ttfamily RHSMode} is 1 or 2 and {\ttfamily xGridScheme} is $<$5.
-Since the recommended setting for {\ttfamily xGridScheme} is 5, this parameter is rarely relevant.}
+{Whenever \parlink{RHSMode} is 1 or 2 and \parlink{xGridScheme} is $<$5.
+Since the recommended setting for \parlink{xGridScheme} is 5, this parameter is rarely relevant.}
 {When an explicit grid is used for the Rosenbluth potentials, which grid and interpolation scheme to use.\\
@@
 {\ttfamily xPotentialsGridScheme} = 3: Use same grid as for distribution function, so no interpolation needed for the self-collision operator.
-  You must set {\ttfamily xGridScheme} = 3 or 4 to use this setting.  Use 2-point stencil for interpolating to other species' grids.\\
+  You must set \parlink{xGridScheme} = 3 or 4 to use this setting.  Use 2-point stencil for interpolating to other species' grids.\\
 
@@
 {1}
-{Whenever {\ttfamily useIterativeLinearSolver} = \true~and there are 2 or more species.}
+{Whenever \parlink{useIterativeLinearSolver} = \true~and there are 2 or more species.}
 {\\
@@
 {1}
-{Whenever {\ttfamily useIterativeLinearSolver} = \true~and {\ttfamily RHSMode} = 1 or 2.}
+{Whenever \parlink{useIterativeLinearSolver} = \true~and \parlink{RHSMode} = 1 or 2.}
 {\\
@@
 {0}
-{Whenever {\ttfamily useIterativeLinearSolver} = \true~and {\ttfamily RHSMode} = 1 or 2 and {\ttfamily preconditioner\_x} $>$ 0.}
+{Whenever \parlink{useIterativeLinearSolver} = \true~and \parlink{RHSMode} = 1 or 2 and \parlink{preconditioner\_x} $>$ 0.}
 {The $x$ structure of the matrix will only be simplified for Legendre index $L$ is $\ge$ this value.
@@
 {0}
-{Whenever {\ttfamily useIterativeLinearSolver} = \true}
+{Whenever \parlink{useIterativeLinearSolver} = \true}
 {\\
@@
 {0}
-{Whenever {\ttfamily useIterativeLinearSolver} = \true~and {\ttfamily preconditioner\_theta} $>$ 0.}
+{Whenever \parlink{useIterativeLinearSolver} = \true~and \parlink{preconditioner\_theta} $>$ 0.}
 {The $\theta$ structure of the matrix will only be simplified for Legendre index $L$ is $\ge$ this value.
@@
 {0}
-{Whenever {\ttfamily useIterativeLinearSolver} = \true~and {\ttfamily preconditioner\_zeta} $>$ 0.}
+{Whenever \parlink{useIterativeLinearSolver} = \true~and \parlink{preconditioner\_zeta} $>$ 0.}
 {The $\zeta$ structure of the matrix will only be simplified for Legendre index $L$ is $\ge$ this value.
@@
 {1}
-{Whenever {\ttfamily useIterativeLinearSolver} = \true}
+{Whenever \parlink{useIterativeLinearSolver} = \true}
 {\\
@@
 {\true}
-{Only when {\ttfamily nonlinear} = \true}
+{Only when \parlink{nonlinear} = \true}
 {If true, the nonlinear term will not be included in the preconditioner matrix, meaning the preconditioner matrix is the same
@@
 {2}
-{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true}
+{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true}
 {Controls which grid in $\theta$ is used for exporting the distribution function.\\
@@
 
-{\ttfamily export\_f\_theta\_option} = 1: Interpolate to a different grid, specified by {\ttfamily export\_f\_theta}. Linear interpolation will be used.
+{\ttfamily export\_f\_theta\_option} = 1: Interpolate to a different grid, specified by \parlink{export\_f\_theta}. Linear interpolation will be used.
 No sorting of the requested values is performed.\\
@@
 values of the $\theta$ grid that are closest to the values requested
-in {\ttfamily export\_f\_theta}.  Values of $\theta$ will be in
+in \parlink{export\_f\_theta}.  Values of $\theta$ will be in
 increasing order. If multiple requested values are close to the same
@@
 
-For all of these options, you can see {\ttfamily export\_f\_theta} in {\ttfamily sfincsOutput.h5} for the actual grid used in the end.
+For all of these options, you can see \parlink{export\_f\_theta} in {\ttfamily sfincsOutput.h5} for the actual grid used in the end.
 }
@@
 {2}
-{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true}
+{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true}
 {Controls which grid in $\zeta$ is used for exporting the distribution function.\\
@@
 
-{\ttfamily export\_f\_zeta\_option} = 1: Interpolate to a different grid, specified by {\ttfamily export\_f\_zeta}. Linear interpolation will be used.
+{\ttfamily export\_f\_zeta\_option} = 1: Interpolate to a different grid, specified by \parlink{export\_f\_zeta}. Linear interpolation will be used.
 No sorting of the requested values is performed.\\
@@
 values of the $\zeta$ grid that are closest to the values requested
-in {\ttfamily export\_f\_zeta}.  Values of $\zeta$ will be in
+in \parlink{export\_f\_zeta}.  Values of $\zeta$ will be in
 increasing order. If multiple requested values are close to the same
@@
 
-For all of these options, you can see {\ttfamily export\_f\_zeta} in {\ttfamily sfincsOutput.h5} for the actual grid used in the end.
+For all of these options, you can see \parlink{export\_f\_zeta} in {\ttfamily sfincsOutput.h5} for the actual grid used in the end.
 }
@@
 {0.0}
-{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true, and {\ttfamily export\_f\_theta\_option} $>0$.}
-{Values of $\theta$ on which you want to save the distribution function.  modulo$(\ldots, 2\pi)$ will be applied.  See {\ttfamily export\_f\_theta\_option} for details}
+{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true, and \parlink{export\_f\_theta\_option} $>0$.}
+{Values of $\theta$ on which you want to save the distribution function.  modulo$(\ldots, 2\pi)$ will be applied.  See \parlink{export\_f\_theta\_option} for details}
 
@@
 {0.0}
-{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true, and {\ttfamily export\_f\_zeta\_option} $>0$.}
-{Values of $\zeta$ on which you want to save the distribution function.  modulo$(\ldots, 2\pi/\mbox{\ttfamily NPeriods})$ will be applied.  See {\ttfamily export\_f\_zeta\_option} for details}
+{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true, and \parlink{export\_f\_zeta\_option} $>0$.}
+{Values of $\zeta$ on which you want to save the distribution function.  modulo$(\ldots, 2\pi/\mbox{\ttfamily NPeriods})$ will be applied.  See \parlink{export\_f\_zeta\_option} for details}
 
@@
 {1}
-{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true}
+{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true}
 {Controls which discretization in $\xi$ is used for exporting the distribution function.\\
@@
 
-{\ttfamily export\_f\_xi\_option} = 1: Report the distribution function on the values of $\xi$ specified by {\ttfamily export\_f\_xi}. No sorting of the requested values is performed.
+{\ttfamily export\_f\_xi\_option} = 1: Report the distribution function on the values of $\xi$ specified by \parlink{export\_f\_xi}. No sorting of the requested values is performed.
 }
@@
 {0.0}
-{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true, and {\ttfamily export\_f\_xi\_option} = 1.}
+{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true, and \parlink{export\_f\_xi\_option} = 1.}
 {Values of $\xi$ on which you want to save the distribution function.  Values must lie in the range $[-1,1]$.}
@@
 {0}
-{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true}
+{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true}
 {Controls which grid in $x = v/\sqrt{2T/m}$ is used for exporting the distribution function.\\
@@
 
-{\ttfamily export\_f\_x\_option} = 1: Interpolate to a different grid, specified by {\ttfamily export\_f\_x}.  Polynomial spectral interpolation will be used.
+{\ttfamily export\_f\_x\_option} = 1: Interpolate to a different grid, specified by \parlink{export\_f\_x}.  Polynomial spectral interpolation will be used.
   No sorting of the requested values is performed.\\
 
-{\ttfamily export\_f\_x\_option} = 2: Do not interpolate. Use the values of the internal $x$ grid that are closest to the values requested in {\ttfamily export\_f\_x}.
+{\ttfamily export\_f\_x\_option} = 2: Do not interpolate. Use the values of the internal $x$ grid that are closest to the values requested in \parlink{export\_f\_x}.
   Values of $x$ will be in increasing order. If multiple requested values are close to the same grid point, the number of
@@
 
-For all of these options, you can see {\ttfamily export\_f\_x} in {\ttfamily sfincsOutput.h5} for the actual grid used in the end.
+For all of these options, you can see \parlink{export\_f\_x} in {\ttfamily sfincsOutput.h5} for the actual grid used in the end.
 }
@@
 {1.0}
-{Whenever {\ttfamily export\_full\_f} or {\ttfamily export\_delta\_f} is \true, and {\ttfamily export\_f\_x\_option} $>0$.}
+{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true, and \parlink{export\_f\_x\_option} $>0$.}
 {Values of $x$ on which you want to save the distribution function.  Values must be $\ge 0$.}
@@
 \subsection{Parameters related only to {\ttfamily scanType} = 1 (resolution convergence scans).}
+\label{sec:scanType1Parameters}
 
@@
 each have 3 associated \sfincsScan~parameters
-which are used for convergence scans ({\ttfamily scanType} = 1): {\ttfamily ...MinFactor}, {\ttfamily ...MaxFactor}, and {\ttfamily ...NumRuns}.
+which are used for convergence scans (\parlink{scanType} = 1): {\ttfamily ...MinFactor}, {\ttfamily ...MaxFactor}, and {\ttfamily ...NumRuns}.
 The first two of these set the range by which the associated resolution parameter is scaled
@@
 }
-would mean the values {\ttfamily Nxi} = 10, 20, and 40 would be tried in a convergence scan.
+would mean the values \Nxi = 10, 20, and 40 would be tried in a convergence scan.
 If you don't want to scan a variable in a convergence scan, set the associated
 {\ttfamily ...NumRuns} parameter to 0, or do not include this parameter in the input file.
-For each resolution parameter ({\ttfamily Ntheta}, {\ttfamily Nzeta}, {\ttfamily Nxi}, etc.), the value itself is read by
+For each resolution parameter (\Ntheta, \Nzeta, \Nxi, etc.), the value itself is read by
 Fortran and so should not be preceded by {\ttfamily !ss}.  However the {\ttfamily ...MinFactor}, {\ttfamily ...MaxFactor}, and {\ttfamily ...NumRuns}
@@
 {real}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 1.}
 {Maximum factor by which \Ntheta~will be multiplied in a convergence scan.}
@@
 {real}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 1.}
 {Minimum factor by which \Ntheta~will be multiplied in a convergence scan.}
@@
 {integer}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 1.}
 {Maximum number of values of \Ntheta~which will be used in a convergence scan.  Only odd integers can be used
@@
 {real}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 1.}
 {Maximum factor by which \Nzeta~will be multiplied in a convergence scan.}
@@
 {real}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 1.}
 {Minimum factor by which \Nzeta~will be multiplied in a convergence scan.}
@@
 {integer}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 1.}
 {Maximum number of values of \Nzeta~which will be used in a convergence scan.  Only odd integers can be used
@@
 {real}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 1.}
 {Maximum factor by which \Nxi~will be multiplied in a convergence scan.}
@@
 {real}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 1.}
 {Minimum factor by which \Nxi~will be multiplied in a convergence scan.}
@@
 {integer}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 1.}
 {Maximum number of values of \Nxi~which will be used in a convergence scan.  Only integers can be used
@@
 {real}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 1.}
 {Maximum factor by which \Nx~will be multiplied in a convergence scan.}
@@
 {real}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 1.}
 {Minimum factor by which \Nx~will be multiplied in a convergence scan.}
@@
 {integer}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 1.}
 {Maximum number of values of \Nx~which will be used in a convergence scan.  Only integers can be used
@@
 {real}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
-{Maximum factor by which {\ttfamily solverTolerance} will be multiplied in a convergence scan.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 1.}
+{Maximum factor by which \parlink{solverTolerance} will be multiplied in a convergence scan.}
 
@@
 {real}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
-{Minimum factor by which {\ttfamily solverTolerance} will be multiplied in a convergence scan.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 1.}
+{Minimum factor by which \parlink{solverTolerance} will be multiplied in a convergence scan.}
 
@@
 {integer}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 1.}
-{Number of values of {\ttfamily solverTolerance} which will be used in a convergence scan.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 1.}
+{Number of values of \parlink{solverTolerance} which will be used in a convergence scan.}
 
@@
 {real}
-{Only when {\ttfamily collisionOperator} = 0 and \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Only when \parlink{collisionOperator} = 0 and \sfincsScan~is run with \parlink{scanType} = 1.}
 {Maximum factor by which \NL~will be multiplied in a convergence scan.}
@@
 {real}
-{Only when {\ttfamily collisionOperator} = 0 and \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Only when \parlink{collisionOperator} = 0 and \sfincsScan~is run with \parlink{scanType} = 1.}
 {Minimum factor by which \NL~will be multiplied in a convergence scan.}
@@
 {integer}
-{Only when {\ttfamily collisionOperator} = 0 and \sfincsScan~is run with {\ttfamily scanType} = 1.}
+{Only when \parlink{collisionOperator} = 0 and \sfincsScan~is run with \parlink{scanType} = 1.}
 {Maximum number of values of \NL~which will be used in a convergence scan.  Only integers can be used
@@
 {real}
-{Only when {\ttfamily collisionOperator} = 0, {\ttfamily xGridScheme} $<$ 5, and \sfincsScan~is run with {\ttfamily scanType} = 1.
-Since the recommended value of {\ttfamily xGridScheme} is 5, this parameter is basically obsolete.}
-{Maximum factor by which {\ttfamily NxPotentialsPerVth} will be multiplied in a convergence scan.}
+{Only when \parlink{collisionOperator} = 0, \parlink{xGridScheme} $<$ 5, and \sfincsScan~is run with \parlink{scanType} = 1.
+Since the recommended value of \parlink{xGridScheme} is 5, this parameter is basically obsolete.}
+{Maximum factor by which \parlink{NxPotentialsPerVth} will be multiplied in a convergence scan.}
 
@@
 {real}
-{Only when {\ttfamily collisionOperator} = 0, {\ttfamily xGridScheme} $<$ 5, and \sfincsScan~is run with {\ttfamily scanType} = 1.
-Since the recommended value of {\ttfamily xGridScheme} is 5, this parameter is basically obsolete.}
-{Minimum factor by which {\ttfamily NxPotentialsPerVth} will be multiplied in a convergence scan.}
+{Only when \parlink{collisionOperator} = 0, \parlink{xGridScheme} $<$ 5, and \sfincsScan~is run with \parlink{scanType} = 1.
+Since the recommended value of \parlink{xGridScheme} is 5, this parameter is basically obsolete.}
+{Minimum factor by which \parlink{NxPotentialsPerVth} will be multiplied in a convergence scan.}
 
@@
 {integer}
-{Only when {\ttfamily collisionOperator} = 0, {\ttfamily xGridScheme} $<$ 5, and \sfincsScan~is run with {\ttfamily scanType} = 1.
-Since the recommended value of {\ttfamily xGridScheme} is 5, this parameter is basically obsolete.}
-{Number of values of {\ttfamily NxPotentialsPerVth} which will be used in a convergence scan.}
+{Only when \parlink{collisionOperator} = 0, \parlink{xGridScheme} $<$ 5, and \sfincsScan~is run with \parlink{scanType} = 1.
+Since the recommended value of \parlink{xGridScheme} is 5, this parameter is basically obsolete.}
+{Number of values of \parlink{NxPotentialsPerVth} which will be used in a convergence scan.}
 
@@
 {real}
-{Only when {\ttfamily collisionOperator} = 0, {\ttfamily xGridScheme} $<$ 5, and \sfincsScan~is run with {\ttfamily scanType} = 1.
-Since the recommended value of {\ttfamily xGridScheme} is 5, this parameter is basically obsolete.}
-{Maximum factor by which {\ttfamily xMax} will be multiplied in a convergence scan.}
+{Only when \parlink{collisionOperator} = 0, \parlink{xGridScheme} $<$ 5, and \sfincsScan~is run with \parlink{scanType} = 1.
+Since the recommended value of \parlink{xGridScheme} is 5, this parameter is basically obsolete.}
+{Maximum factor by which \parlink{xMax} will be multiplied in a convergence scan.}
 
@@
 {real}
-{Only when {\ttfamily collisionOperator} = 0, {\ttfamily xGridScheme} $<$ 5, and \sfincsScan~is run with {\ttfamily scanType} = 1.
-Since the recommended value of {\ttfamily xGridScheme} is 5, this parameter is basically obsolete.}
-{Minimum factor by which {\ttfamily xMax} will be multiplied in a convergence scan.}
+{Only when \parlink{collisionOperator} = 0, \parlink{xGridScheme} $<$ 5, and \sfincsScan~is run with \parlink{scanType} = 1.
+Since the recommended value of \parlink{xGridScheme} is 5, this parameter is basically obsolete.}
+{Minimum factor by which \parlink{xMax} will be multiplied in a convergence scan.}
 
@@
 {integer}
-{Only when {\ttfamily collisionOperator} = 0, {\ttfamily xGridScheme} $<$ 5, and \sfincsScan~is run with {\ttfamily scanType} = 1.
-Since the recommended value of {\ttfamily xGridScheme} is 5, this parameter is basically obsolete.}
-{Number of values of {\ttfamily xMax} which will be used in a convergence scan.}
+{Only when \parlink{collisionOperator} = 0, \parlink{xGridScheme} $<$ 5, and \sfincsScan~is run with \parlink{scanType} = 1.
+Since the recommended value of \parlink{xGridScheme} is 5, this parameter is basically obsolete.}
+{Number of values of \parlink{xMax} which will be used in a convergence scan.}
 
@@
 will always be uniformly (linearly) spaced.  
-Notice that exactly 1 of the 4 variables {\ttfamily dPhiHatdpsiHatMax}, {\ttfamily dPhiHatdpsiNMax}, {\ttfamily dPhiHatdrHatMax}, and {\ttfamily dPhiHatdrNMax}
+Notice that exactly 1 of the 4 variables \parlink{dPhiHatdpsiHatMax}, \parlink{dPhiHatdpsiNMax}, \parlink{dPhiHatdrHatMax}, and \parlink{dPhiHatdrNMax}
 will be used, depending on\\
-{\ttfamily inputRadialCoordinateForGradients}.
+\parlink{inputRadialCoordinateForGradients}.
 Similarly, exactly 1 of the 4 variables
-{\ttfamily dPhiHatdpsiHatMin}, {\ttfamily dPhiHatdpsiNMin}, {\ttfamily dPhiHatdrHatMin}, and {\ttfamily dPhiHatdrNMin}
+\parlink{dPhiHatdpsiHatMin}, \parlink{dPhiHatdpsiNMin}, \parlink{dPhiHatdrHatMin}, and \parlink{dPhiHatdrNMin}
 will be used.
@@
 {integer}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 2.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 2.}
 {Number of values of radial electric field to consider in a scan.}
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinateForGradients} = 0 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
-{Maximum value of {\ttfamily dPhiHatdpsiHat} to use in the scan.}
+{Only when \parlink{inputRadialCoordinateForGradients} = 0 and \sfincsScan~is run with \parlink{scanType} = 2.}
+{Maximum value of \parlink{dPhiHatdpsiHat} to use in the scan.}
 
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinateForGradients} = 0 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
-{Minimum value of {\ttfamily dPhiHatdpsiHat} to use in the scan.}
+{Only when \parlink{inputRadialCoordinateForGradients} = 0 and \sfincsScan~is run with \parlink{scanType} = 2.}
+{Minimum value of \parlink{dPhiHatdpsiHat} to use in the scan.}
 
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinateForGradients} = 1 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
-{Maximum value of {\ttfamily dPhiHatdpsiN} to use in the scan.}
+{Only when \parlink{inputRadialCoordinateForGradients} = 1 and \sfincsScan~is run with \parlink{scanType} = 2.}
+{Maximum value of \parlink{dPhiHatdpsiN} to use in the scan.}
 
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinateForGradients} = 1 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
-{Minimum value of {\ttfamily dPhiHatdpsiN} to use in the scan.}
+{Only when \parlink{inputRadialCoordinateForGradients} = 1 and \sfincsScan~is run with \parlink{scanType} = 2.}
+{Minimum value of \parlink{dPhiHatdpsiN} to use in the scan.}
 
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinateForGradients} = 2 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
-{Maximum value of {\ttfamily dPhiHatdrHat} to use in the scan.}
+{Only when \parlink{inputRadialCoordinateForGradients} = 2 and \sfincsScan~is run with \parlink{scanType} = 2.}
+{Maximum value of \parlink{dPhiHatdrHat} to use in the scan.}
 
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinateForGradients} = 2 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
-{Minimum value of {\ttfamily dPhiHatdrHat} to use in the scan.}
+{Only when \parlink{inputRadialCoordinateForGradients} = 2 and \sfincsScan~is run with \parlink{scanType} = 2.}
+{Minimum value of \parlink{dPhiHatdrHat} to use in the scan.}
 
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinateForGradients} = 3 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
-{Maximum value of {\ttfamily dPhiHatdrN} to use in the scan.}
+{Only when \parlink{inputRadialCoordinateForGradients} = 3 and \sfincsScan~is run with \parlink{scanType} = 2.}
+{Maximum value of \parlink{dPhiHatdrN} to use in the scan.}
 
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinateForGradients} = 3 and \sfincsScan~is run with {\ttfamily scanType} = 2.}
-{Minimum value of {\ttfamily dPhiHatdrN} to use in the scan.}
+{Only when \parlink{inputRadialCoordinateForGradients} = 3 and \sfincsScan~is run with \parlink{scanType} = 2.}
+{Minimum value of \parlink{dPhiHatdrN} to use in the scan.}
 
@@
 {string. Must be of the fortran namelist parameters that takes an integer or real value. Case-insensitive.}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 3.}
-{Name of the variable to scan in a {\ttfamily scanType} = 3 scan.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 3.}
+{Name of the variable to scan in a \parlink{scanType} = 3 scan.}
 
@@
 {real}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 3.}
-{Maximum value of {\ttfamily scanVariable} to use in a {\ttfamily scanType} = 3 scan.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 3.}
+{Maximum value of \parlink{scanVariable} to use in a \parlink{scanType} = 3 scan.}
 
@@
 {real}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 3.}
-{Minimum value of {\ttfamily scanVariable} to use in a {\ttfamily scanType} = 3 scan.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 3.}
+{Minimum value of \parlink{scanVariable} to use in a \parlink{scanType} = 3 scan.}
 
@@
 {integer}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 3.}
-{Number of values of {\ttfamily scanVariable} to use in a {\ttfamily scanType} = 3 scan.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 3.}
+{Number of values of \parlink{scanVariable} to use in a \parlink{scanType} = 3 scan.}
 
@@
 {string. Must be `linear', `lin', 'logarithmic', or 'log'}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 3.}
-{Whether to space the values of {\ttfamily scanVariable} in a linear or logarithmic manner.
+{Only when \sfincsScan~is run with \parlink{scanType} = 3.}
+{Whether to space the values of \parlink{scanVariable} in a linear or logarithmic manner.
 The settings `linear' and `lin' have identical behavior. 
@@
 
-Notice that exactly 1 of the 4 variables {\ttfamily psiHat\_max}, {\ttfamily psiN\_max}, {\ttfamily rHat\_max}, and {\ttfamily rN\_max}
+Notice that exactly 1 of the 4 variables \parlink{psiHat\_max}, \parlink{psiN\_max}, \parlink{rHat\_max}, and \parlink{rN\_max}
 will be used, depending on
-{\ttfamily inputRadialCoordinate}.
+\parlink{inputRadialCoordinate}.
 Similarly, exactly 1 of the 4 variables
-{\ttfamily psiHat\_min}, {\ttfamily psiN\_min}, {\ttfamily rHat\_min}, and {\ttfamily rN\_min}
+\parlink{psiHat\_min}, \parlink{psiN\_min}, \parlink{rHat\_min}, and \parlink{rN\_min}
 will be used.
@@
 {integer}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
-{How to specify the profiles of density, temperature, and (when  {\ttfamily scanType} = 5)
+{Only when \sfincsScan~is run with \parlink{scanType} = 4 or 5.}
+{How to specify the profiles of density, temperature, and (when  \parlink{scanType} = 5)
 the range of radial electric field to consider.\\
@@
 {integer}
-{Only when \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
+{Only when \sfincsScan~is run with \parlink{scanType} = 4 or 5.}
 {Maximum number of values of minor radius to consider in the scan.
-Depending on {\ttfamily geometryScheme} and {\ttfamily VMECRadialOption},
+Depending on \parlink{geometryScheme} and \parlink{VMECRadialOption},
 it may be that only surfaces available in the magnetic equilibrium file will be
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinate} = 0 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
-{Maximum value of {\ttfamily psiHat} to use in the scan.}
+{Only when \parlink{inputRadialCoordinate} = 0 and \sfincsScan~is run with \parlink{scanType} = 4 or 5.}
+{Maximum value of \parlink{psiHat} to use in the scan.}
 
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinate} = 0 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
-{Minimum value of {\ttfamily psiHat} to use in the scan.}
+{Only when \parlink{inputRadialCoordinate} = 0 and \sfincsScan~is run with \parlink{scanType} = 4 or 5.}
+{Minimum value of \parlink{psiHat} to use in the scan.}
 
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinate} = 1 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
-{Maximum value of {\ttfamily psiN} to use in the scan.}
+{Only when \parlink{inputRadialCoordinate} = 1 and \sfincsScan~is run with \parlink{scanType} = 4 or 5.}
+{Maximum value of \parlink{psiN} to use in the scan.}
 
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinate} = 1 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
-{Minimum value of {\ttfamily psiN} to use in the scan.}
+{Only when \parlink{inputRadialCoordinate} = 1 and \sfincsScan~is run with \parlink{scanType} = 4 or 5.}
+{Minimum value of \parlink{psiN} to use in the scan.}
 
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinate} = 2 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
-{Maximum value of {\ttfamily rHat} to use in the scan.}
+{Only when \parlink{inputRadialCoordinate} = 2 and \sfincsScan~is run with \parlink{scanType} = 4 or 5.}
+{Maximum value of \parlink{rHat} to use in the scan.}
 
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinate} = 2 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
-{Minimum value of {\ttfamily rHat} to use in the scan.}
+{Only when \parlink{inputRadialCoordinate} = 2 and \sfincsScan~is run with \parlink{scanType} = 4 or 5.}
+{Minimum value of \parlink{rHat} to use in the scan.}
 
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinate} = 3 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
-{Maximum value of {\ttfamily rN} to use in the scan.}
+{Only when \parlink{inputRadialCoordinate} = 3 and \sfincsScan~is run with \parlink{scanType} = 4 or 5.}
+{Maximum value of \parlink{rN} to use in the scan.}
 
@@
 {real}
-{Only when {\ttfamily inputRadialCoordinate} = 3 and \sfincsScan~is run with {\ttfamily scanType} = 4 or 5.}
-{Minimum value of {\ttfamily rN} to use in the scan.}
+{Only when \parlink{inputRadialCoordinate} = 3 and \sfincsScan~is run with \parlink{scanType} = 4 or 5.}
+{Minimum value of \parlink{rN} to use in the scan.}
 

### 2015-06-04 doc/manual/overview.tex
Added hyperlinks to manual

@@
 \item
+The electrostatic potential can either be taken to be constant or non-constant on a flux surface.
+
+\item
 Optional nonlinear terms in the kinetic equation (involving both the non-Maxwellian distribution function and poloidal/toroidal electric field) can be included using Newton's method.
@@
 as are several analytic models for LHD and W7-X in which 3 or 4 Fourier components are retained.
-The primary switch for controlling the magnetic geometry in \sfincs~is the {\ttfamily geometryScheme} parameter
-in the {\ttfamily geometryParameters} input namelist.
+The primary switch for controlling the magnetic geometry in \sfincs~is the \parlink{geometryScheme} parameter
+in the {\ttfamily \hyperref[sec:geometryParameters]{geometryParameters}} input namelist.
 For more details about geometry options in \sfincs, see section \ref{sec:geometryParameters}.
@@
 \href{mailto:matt.landreman@gmail.com}{\nolinkurl{matt.landreman@gmail.com} }
-
-%\begin{figure}[htbp]
-%\begin{center}
-%%\includegraphics[width=1.0\textwidth]{poloidal_cross-sections.pdf}
-%\mycaption{Trajectories of trapped particles.  (a) In tokamaks and omnigenous stellarators,
-%all collisionless trajectories are confined.  (b) In contrast, trapped particles in other stellarators
-%may have a radial drift that does not average to zero.
-%\label{fig:poloidal_cross_sections}}
-%\end{center}
-%\end{figure}

### 2015-06-04 doc/manual/resolution.tex
Added to user manual
Added hyperlinks to manual

@@
 and \Nx~required for convergence depend strongly on the magnetic
-geometry and collisionality, with modest dependence also on the radial
-electric field. It is strongly recommended that you test for
+geometry and collisionality. It is strongly recommended that you test for
 convergence with respect to \Ntheta, \Nzeta, \Nxi, and \Nx~whenever
@@
 for convergence, and so which almost never need to be adjusted.
-These parameters and good values for them are {\ttfamily NL=4}, 
-{\ttfamily solverTolerance} = $10^{-6}$,
-{\ttfamily xMax=5.0},
-and {\ttfamily NxPotentialsPerVth=40.0}.  The latter two of these parameters are in fact ignored
-for the recommended and default {\ttfamily xGridMode} setting, 5.
+These parameters and good values for them are \NL = 4, 
+\parlink{solverTolerance} = $10^{-6}$,
+\parlink{xMax} = 5.0,
+and \parlink{NxPotentialsPerVth} = 40.0.  
+(These values are set as the defaults.)
+The latter two of these parameters are in fact ignored
+for the recommended and default \parlink{xGridScheme} setting, 5.
 
@@
 When the mean-free-path is shorter than the parallel length scale of the equilibrium,
-({\ttfamily nuPrime} $\ge$ 1), the parameters required for convergence
+%({\ttfamily nuPrime} $\ge$ 1), 
+the parameters required for convergence
 do not depend much on collisionality. In the opposite
 limit in which the mean-free-path is longer than the parallel length scale of the equilibrium,
-({\ttfamily nuPrime} $\le$ 1)
+%({\ttfamily nuPrime} $\le$ 1)
 values of
@@
 not quite as dramatically. The \Nx~required for convergence does not depend
-much on collisionality, though typically the required value increases
-slightly with collisionality at high collisionality
+nearly as much on collisionality.  The reason is that
+at low collisionality, a boundary layer develops in the distribution function
+along the boundary between trapped and passing (untrapped) particles.
+The location of this boundary depends on $\theta$, $\zeta$, and $\xi$,
+and so high resolution is required in these coordinates to resolve the boundary layer.
+But the location of the trapped-passing boundary is independent
+of $x$, and hence the resolution requried in $x$ is not particularly high.
 
+Typically you can expect to use \Nx = 5-8, with 5 being sufficient at low collisionality
+and 8 being required at high collisionality.
 The \Nx~required for convergence may need to increase slightly with the number of species.
-Typically you can expect to use \Nx=5-8.
 
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=1.0\textwidth]{sfincsConvergenceScan.pdf}
+\mycaption{Plot generated by \sfincsScanPlot showing a resolution convergence scan for the example
+{\ttfamily geometryScheme4\_2species\_noEr}.
+\label{fig:convergence}}
+\end{center}
+\end{figure}
+
+To check how well the results of \sfincs~are converged with respect to the resolution
+parameters, run \sfincsScan~with \parlink{scanType} = 1.
+In this type of scan, a ``base case'' is first run at the
+values of \Ntheta, \Nzeta, \Nxi, and \Nx~specified in the input file.
+Then, each of these parameters will be varied in turn, holding the other parameters fixed.
+The range of \Ntheta~in the scan is specified by the parameters \parlink{NthetaMinFactor},
+\parlink{NthetaMaxFactor}, and \parlink{NthetaNumRuns}. 
+Each of these three parameters is read by \sfincsScan~rather than by \sfincs~itself,
+and so it must be prefaced by {\ttfamily !ss} in the input namelist file.
+The ranges of \Nzeta, \Nxi, and \Nx~are set by parameters with analogous names,
+as detailed in section \ref{sec:scanType1Parameters}.
+
+For example, suppose you initially run \sfincsScan~with \Ntheta = 15, 
+\parlink{NthetaMinFactor} = 0.7, \parlink{NthetaMaxFactor} = 1.5, and 
+\parlink{NthetaNumRuns} = 5.
+This will generate \sfincs~runs at \Ntheta = 11, 13, 15 (the base case), 19, and 23.
+Notice that \sfincsScan~automatically ensures that only odd values are used.
+(Only odd values of \Ntheta~and \Nzeta~are used internally in \sfincs.)
+The maximum and minimum values for the scan, 11 and 23, are the nearest odd integers
+to \Ntheta$\times$\parlink{NthetaMinFactor} and \Ntheta$\times$\parlink{NthetaMaxFactor} respectively.
+Notice also that there are `gaps' at  \Ntheta = 17 and 21 since \sfincsScan~attempts to space the values
+logarithmically rather than uniformly.
+
+In addition to the 4 resolution parameters above, \sfincsScan~also allows the parameters
+\parlink{solverTolerance}, \parlink{xMax}, and \parlink{NxPotentialsPerVth}
+to be scanned.  These latter two parameters are not used for the default \parlink{xGridScheme}
+so it is usually unnecessary to scan them.  It is also usually unnecessary to scan \parlink{solverTolerance}
+since the default value is usually robust.
+
+A directory will be created for each run.  Each directory will contain
+a copy of {\ttfamily input.namelist} in which \Ntheta, \Nzeta, \Nx, or \Nxi~has been
+altered appropriately by \sfincsScan. Each directory will also contain a job file for the run.
+
+You do not need to scan all variables. For example, if you do not wish to scan \Nx,
+you can either set \parlink{NxNumRuns} = 0, or you can not specify
+\parlink{NxNumRuns} in the {\ttfamily input.namelist} file.
+
+Once the individual runs in the scan have begun to finish, you can plot the results by running \sfincsScanPlot.
+You can plot the results before all the runs in the scan finish, although at least the base case
+run must be completed.  Different quantities will be displayed depending on
+\parlink{RHSMode} and \parlink{includePhi1}.
+
+Typical convergence behavior is illustrated in figure \ref{fig:convergence},
+showing the results of \sfincsScan~and \sfincsScanPlot~for the example
+{\ttfamily geometryScheme4\_2species\_noEr}.
+In this figure, a very large number of values are included for the resolution
+parameters, more than you would likely include in routine use of the code.
+The red vertical lines emphasize the ``base case'' resolution parameters.
+For all parameters scanned, the physical output quantities (particle flux, heat flux, etc)
+are essentially unchanged when any of the resolution parameters
+are increased beyond the base case value.  Thus, the results in the base case are converged,
+and so we can believe the results.
+Observe in the figure that the output quantities tend to oscillate when \Nxi~is incremented by 1 and insufficient
+\Nxi~is used. Results for even and odd \Nxi~converge to the same value, as one would hope.
+
+If you wish to add more runs to the scan at a later time,
+you can alter the relevant \sfincsScan~parameters in the original {\ttfamily input.namelist}~file
+and run \sfincsScan~again from the original directory.  The \sfincsScan~code will automatically detect which runs
+have already been submitted, so duplicate runs will not be generated.
+For example, consider the scan of \Ntheta~described previously in this section,
+and suppose you wish to fill in the gaps at \Ntheta = 17 and 21 in the original scan.
+To do this, you could set \parlink{NthetaNumRuns} to a large number like 100
+and re-run \sfincsScan.  Since \sfincsScan~intelligently avoids duplication,
+the only new runs generated will be  \Ntheta = 17 and 21.
+
+If you are ultimately intending to scan $E_r$, you probably only need to run a convergence scan
+at a single value of $E_r$.  This is because the resolution requirements of \sfincs~are not sensitive to
+$E_r$, as discussed above.
+Also, if you are ultimately intending to run \sfincs~at a range of minor radii,
+it is reasonably to only run a convergence scan at a single radius close to the magnetic axis.
+This is because the typically peaked shape of the temperature profile means that
+collisionality is lowest near the axis.  Since the resolution requirements of \sfincs~are most
+demanding at low collisionality, then if the code is converged at the radius of lowest collisionality,
+it should be converged at all radii.
+
 \section{Typical resolution requirements}

### 2015-06-04 doc/manual/runs.tex
Example input files pared down to non-default parameters, since documentation is now in user manual.
Added hyperlinks to manual

@@
 suggested here. However, if you use a {\ttfamily vmec} or {\ttfamily .bc} magnetic equilibrium
-by choosing {\ttfamily geometryScheme} = 5, 11, or 12, then you MUST use $\bar{B}$ = 1 Tesla and $\bar{R}$ = 1 meter.
+by choosing \parlink{geometryScheme} = 5, 11, or 12, then you MUST use $\bar{B}$ = 1 Tesla and $\bar{R}$ = 1 meter.
 The code ``knows'' about the reference values only through
-the 3 combinations {\ttfamily Delta}, {\ttfamily alpha}, and {\ttfamily nu\_n}
-in the {\ttfamily physicsParameters} namelist.
+the 3 combinations \parlink{Delta}, \parlink{alpha}, and \parlink{nu\_n}
+in the {\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}} namelist.
 
@@
 
-{\ttfamily psiHat} = $\hat\psi$ is the toroidal flux (divided by $2\pi$), normalized by $\bar{B}\bar{R}^2$.\\
+{\ttfamily \hypertarget{psiHat}{psiHat}} = $\hat\psi$ is the toroidal flux (divided by $2\pi$), normalized by $\bar{B}\bar{R}^2$.\\
 
-{\ttfamily psiN} = $\psi_N$ is the toroidal flux normalized by its value at the last closed flux surface.\\
+{\ttfamily \hypertarget{psiN}{psiN}} = $\psi_N$ is the toroidal flux normalized by its value at the last closed flux surface.\\
 
-{\ttfamily rHat} = $\hat{r}$ is defined as {\ttfamily aHat}$\sqrt{\mbox{\ttfamily psiN}}$, where {\ttfamily aHat} is an effective minor radius of the last closed flux surface normalized by $\bar{R}$.\\
+{\ttfamily \hypertarget{rHat}{rHat}} = $\hat{r}$ is defined as {\ttfamily aHat}$\sqrt{\mbox{\ttfamily psiN}}$, where {\ttfamily aHat} is an effective minor radius of the last closed flux surface normalized by $\bar{R}$.\\
 
-{\ttfamily rN} = $r_N$ is defined as $\sqrt{\mbox{\ttfamily psiN}}$.\\
+{\ttfamily \hypertarget{rN}{rN}} = $r_N$ is defined as $\sqrt{\mbox{\ttfamily psiN}}$.\\
 
@@
 When setting up a run, you can make several independent choices for radial coordinates.  One parameter
-you select is {\ttfamily inputRadialCoordinateForGradients} in the {\ttfamily geometryParameters} namelist.
-This parameter controls which coordinate is used to specify the gradients. The possible values of {\ttfamily inputRadialCoordinateForGradients} are:\\
+you select is \parlink{inputRadialCoordinateForGradients} in the 
+{\ttfamily \hyperref[sec:geometryParameters]{geometryParameters}} namelist.
+This parameter controls which coordinate is used to specify the gradients. The possible values of \parlink{inputRadialCoordinateForGradients} are:\\
 
@@
 
-0: Use derivatives with respect to {\ttfamily psiHat}: Density gradients are specified using {\ttfamily dnHatdpsiHats}, 
-temperature gradients are specified using {\ttfamily dTHatdpsiHats},  a single $E_r$ is specified using
-{\ttfamily dPhiHatdpsiHat}, and a range of $E_r$ for a scan is specified using {\ttfamily dPhiHatdpsiHatMin}-{\ttfamily dPhiHatdpsiHatMax}.
+0: Use derivatives with respect to {\ttfamily psiHat}: Density gradients are specified using \parlink{dnHatdpsiHats}, 
+temperature gradients are specified using \parlink{dTHatdpsiHats},  a single $E_r$ is specified using
+\parlink{dPhiHatdpsiHat}, and a range of $E_r$ for a scan is specified using \parlink{dPhiHatdpsiHatMin}-\parlink{dPhiHatdpsiHatMax}.
 \\
 
-1: Use derivatives with respect to {\ttfamily psiN}: Density gradients are specified using {\ttfamily dnHatdpsiNs}, 
-temperature gradients are specified using {\ttfamily dTHatdpsiNs},  a single $E_r$ is specified using
-{\ttfamily dPhiHatdpsiN}, and a range of $E_r$ for a scan is specified using {\ttfamily dPhiHatdpsiNMin}-{\ttfamily dPhiHatdpsiNMax}.
+1: Use derivatives with respect to {\ttfamily psiN}: Density gradients are specified using \parlink{dnHatdpsiNs}, 
+temperature gradients are specified using \parlink{dTHatdpsiNs},  a single $E_r$ is specified using
+\parlink{dPhiHatdpsiN}, and a range of $E_r$ for a scan is specified using \parlink{dPhiHatdpsiNMin}-\parlink{dPhiHatdpsiNMax}.
 \\
 
-2: Use derivatives with respect to {\ttfamily rHat}: Density gradients are specified using {\ttfamily dnHatdrHats}, 
-temperature gradients are specified using {\ttfamily dTHatdrHats},  a single $E_r$ is specified using
-{\ttfamily dPhiHatdrHat}, and a range of $E_r$ for a scan is specified using {\ttfamily dPhiHatdrHatMin}-{\ttfamily dPhiHatdrHatMax}.
+2: Use derivatives with respect to {\ttfamily rHat}: Density gradients are specified using \parlink{dnHatdrHats}, 
+temperature gradients are specified using \parlink{dTHatdrHats},  a single $E_r$ is specified using
+\parlink{dPhiHatdrHat}, and a range of $E_r$ for a scan is specified using \parlink{dPhiHatdrHatMin}-\parlink{dPhiHatdrHatMax}.
 \\
 
-3: Use derivatives with respect to {\ttfamily rN}: Density gradients are specified using {\ttfamily dnHatdrNs}, 
-temperature gradients are specified using {\ttfamily dTHatdrNs},  a single $E_r$ is specified using
-{\ttfamily dPhiHatdrN}, and a range of $E_r$ for a scan is specified using {\ttfamily dPhiHatdrNMin}-{\ttfamily dPhiHatdrNMax}.
+3: Use derivatives with respect to {\ttfamily rN}: Density gradients are specified using \parlink{dnHatdrNs}, 
+temperature gradients are specified using \parlink{dTHatdrNs},  a single $E_r$ is specified using
+\parlink{dPhiHatdrN}, and a range of $E_r$ for a scan is specified using \parlink{dPhiHatdrNMin}-\parlink{dPhiHatdrNMax}.
 \\
@@
 Another choice involving radial coordinates is how to specify the flux surface for the computation.
-This choice is made using the parameter {\ttfamily inputRadialCoordinate} in the {\ttfamily geometryParameters}
-namelist, which is again an integer from 0 to 3, and this parameter need not be the same as {\ttfamily inputRadialCoordinateForGradients}.
+This choice is made using the parameter \parlink{inputRadialCoordinate} in the {\ttfamily \hyperref[sec:geometryParameters]{geometryParameters}}
+namelist, which is again an integer from 0 to 3, and this parameter need not be the same as \parlink{inputRadialCoordinateForGradients}.
 An extra complication with specifying the flux surface is that the magnetic equilibrium file will contain data on a finite number of surfaces,
 and you may wish to use one of these surfaces.  For this reason, the parameters for specifying the flux surface have {\ttfamily \_wish}
-appended to the name. In other words, the allowed values for {\ttfamily inputRadialCoordinate} are:\\
+appended to the name. In other words, the allowed values for \parlink{inputRadialCoordinate} are:\\
 
@@
 
-0: Specify the flux surface using {\ttfamily psiHat\_wish}.\\
+0: Specify the flux surface using \parlink{psiHat\_wish}.\\
 
-1: Specify the flux surface using {\ttfamily psiN\_wish}.\\
+1: Specify the flux surface using \parlink{psiN\_wish}.\\
 
-2: Specify the flux surface using {\ttfamily rHat\_wish}.\\
+2: Specify the flux surface using \parlink{rHat\_wish}.\\
 
-3: Specify the flux surface using {\ttfamily rN\_wish}.\\
+3: Specify the flux surface using \parlink{rN\_wish}.\\
 
@@
 
-When using {\ttfamily geometryScheme} == 11 or 12, \sfincs~will always shift the ``wish'' value so it matches an available surface in the magnetic equilibrium file.
-For {\ttfamily geometryScheme} == 5, the {\ttfamily VMECRadialOption} parameter lets you can choose whether to shift to the nearest surface in the magnetic equilibrium file,
+When using \parlink{geometryScheme} == 11 or 12, \sfincs~will always shift the ``wish'' value so it matches an available surface in the magnetic equilibrium file.
+For \parlink{geometryScheme} == 5, the \parlink{VMECRadialOption} parameter lets you can choose whether to shift to the nearest surface in the magnetic equilibrium file,
 or to interpolate the \vmec~data onto the exact value of radius you specify.
@@
 The radial coordinate used in the {\ttfamily profiles} file need not be the same as either
-{\ttfamily inputRadialCoordinate} or {\ttfamily inputRadialCoordinateForGradients}.
+\parlink{inputRadialCoordinate} or \parlink{inputRadialCoordinateForGradients}.
 Note however that the maximum and minimum radial electric field specified in the {\ttfamily profiles}
-file must be defined as the derivative of the electrostatic potential with respect to the radial coordinate {\ttfamily inputRadialCoordinateForGradients}.
+file must be defined as the derivative of the electrostatic potential with respect to the radial coordinate \parlink{inputRadialCoordinateForGradients}.
 
-For more details about the behavior of {\ttfamily inputRadialCoordinate}, {\ttfamily inputRadialCoordinateForGradients}, and {\ttfamily VMECRadialOption},
+For more details about the behavior of \parlink{inputRadialCoordinate}, \parlink{inputRadialCoordinateForGradients}, and \parlink{VMECRadialOption},
 see section \ref{sec:geometryParameters}.
@@
 These variations of the kinetic equation are called ``trajectory models'' in \cite{sfincsPaper}.
-The relevant terms in the kinetic equation can be turned off and on by certain Boolean parameters in the {\ttfamily physicsParameters} namelist.
+The relevant terms in the kinetic equation can be turned off and on by certain Boolean parameters in the {\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}} namelist.
 The models described in \cite{sfincsPaper} are selected as follows:\\
@@
 {\ttfamily 
-includeXDotTerm = .true.\\
-includeElectricFieldTermInXiDot = .true.\\
-useDKESExBDrift = .false.\\
+\parlink{includeXDotTerm} = .true.\\
+\parlink{includeElectricFieldTermInXiDot} = .true.\\
+\parlink{useDKESExBDrift} = .false.\\
 }
@@
 {\ttfamily
-includeXDotTerm = .false.\\
-includeElectricFieldTermInXiDot = .false.\\
-useDKESExBDrift = .false.\\
+\parlink{includeXDotTerm} = .false.\\
+\parlink{includeElectricFieldTermInXiDot} = .false.\\
+\parlink{useDKESExBDrift} = .false.\\
 }
@@
 {\ttfamily
-includeXDotTerm = .false.\\
-includeElectricFieldTermInXiDot = .false.\\
-useDKESExBDrift = .true.\\
+\parlink{includeXDotTerm} = .false.\\
+\parlink{includeElectricFieldTermInXiDot} = .false.\\
+\parlink{useDKESExBDrift} = .true.\\
 }
@@
 In \sfincs, you can choose whether or not to
-include $\Phi_1$ using the paramter {\ttfamily includePhi1} in the {\ttfamily physicsParameters}
+include $\Phi_1$ using the paramter {\ttfamily \hyperlink{includePhi1}{includePhi1}} in the 
+{\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}}
 namelist.  If and only if $\Phi_1$ is included, a quasineutrality equation is solved
 at each point on the flux surface.  Due to these extra unknowns ($\Phi_1$) and extra equations
-(quasineutrality), the system matrix is slightly larger when {\ttfamily includePhi1} is \true.
+(quasineutrality), the system matrix is slightly larger when \parlink{includePhi1} is \true.
 Specifically, the number of rows and columns are each increased by \Ntheta$\times$\Nzeta$+1$.  This increase is miniscule compared
@@
 but also on velocity space and species.  Thus, there is very little extra computational cost associated
-with {\ttfamily includePhi1}.  You may wish to set {\ttfamily includePhi1} = \true~when
-using \sfincs~to model an experiment, and set  {\ttfamily includePhi1} = \false~when
+with \parlink{includePhi1}.  You may wish to set \parlink{includePhi1} = \true~when
+using \sfincs~to model an experiment, and set \parlink{includePhi1} = \false~when
 comparing \sfincs~with analytic theory or with another code that does not include $\Phi_1$.
@@
 
-By setting {\ttfamily RHSMode=3}, \sfincs~can be run in a mode
-where it solves the same kinetic equation (prior to discretization) as {\ttfamily dkes}
-and other monoenergetic codes.
-When {\ttfamily RHSMode=3}, the values of {\ttfamily Zs}, {\ttfamily THats}, {\ttfamily nHats},
-{\ttfamily mHats}, {\ttfamily nu\_n}, and {\ttfamily dPhiHatdXXX} are all ignored.
-Instead, the collisionality is set by {\ttfamily nuPrime}, and the radial electric field is set
-by {\ttfamily EStar}.  The first of these quantities is the dimensionless collisionality
+By setting \parlink{RHSMode} = 3, \sfincs~can be run in a mode
+where it solves the same kinetic equation (prior to discretization) as 
+\dkes~and other monoenergetic codes.
+When \parlink{RHSMode} = 3, the values of \parlink{Zs}, \parlink{THats}, \parlink{nHats},
+\parlink{mHats}, \parlink{nu\_n}, and {\ttfamily dPhiHatdXXX} are all ignored.
+Instead, the collisionality is set by \parlink{nuPrime}, and the radial electric field is set
+by \parlink{EStar}.  The first of these quantities is the dimensionless collisionality
 \begin{equation}
-\mbox{\ttfamily nuPrime} = \frac{(G+\iota I) \nu}{v B_0}
+\mbox{\parlink{nuPrime}} = \frac{(G+\iota I) \nu}{v B_0}
 \end{equation}
@@
 \begin{equation}
-\mbox{\ttfamily EStar} = \frac{cG}{\iota v B_0} \frac{d\Phi}{d\psi}
+\mbox{\parlink{EStar}} = \frac{cG}{\iota v B_0} \frac{d\Phi}{d\psi}
 \end{equation}
@@
 
-The two parameters {\ttfamily nuPrime} and {\ttfamily EStar} are
+The two parameters \parlink{nuPrime} and \parlink{EStar} are
 related to the corresponding DKES parameters {\ttfamily CMUL} and

### 2015-06-04 doc/manual/inputParameters.tex

@@
 
-{\ttfamily scanType} = 21: Read in a list of requested runs from a file {\ttfamily runspec.dat}. See \path{sfincs/fortran/utils/sfincsScan_21} for an example file.
+{\ttfamily scanType} = 21: Read in a list of requested runs from a
+file {\ttfamily runspec.dat}. See
+\path{sfincs/fortran/utils/sfincsScan_21} for an example file. If the
+file has a different name than {\ttfamily runspec.dat}, for instance
+{\ttfamily thefilename.dat}, this name can be
+specified by adding the line\\
+{\ttfamily
+!ss runSpecFile = thefilename.dat\\
+}
 }

### 2015-06-06 doc/manual/resolution.tex
Added convergence scan figure to manual

@@
 \includegraphics[width=1.0\textwidth]{sfincsConvergenceScan.pdf}
-\mycaption{Plot generated by \sfincsScanPlot showing a resolution convergence scan for the example
+\mycaption{Plot generated by \sfincsScanPlot~showing a resolution convergence scan for the example
 {\ttfamily geometryScheme4\_2species\_noEr}.
@@
 \parlink{solverTolerance}, \parlink{xMax}, and \parlink{NxPotentialsPerVth}
-to be scanned.  These latter two parameters are not used for the default \parlink{xGridScheme}
+to be scanned.  The last two of these parameters are not used for the default \parlink{xGridScheme}
 so it is usually unnecessary to scan them.  It is also usually unnecessary to scan \parlink{solverTolerance}
@@
 {\ttfamily geometryScheme4\_2species\_noEr}.
-In this figure, a very large number of values are included for the resolution
-parameters, more than you would likely include in routine use of the code.
+In this figure, a very large number of runs are included in the scan,
+more than you would likely include in routine use of the code.
 The red vertical lines emphasize the ``base case'' resolution parameters.
 For all parameters scanned, the physical output quantities (particle flux, heat flux, etc)
-are essentially unchanged when any of the resolution parameters
-are increased beyond the base case value.  Thus, the results in the base case are converged,
+do not visibly change (on the scale of the plots) when any of the resolution parameters
+are increased beyond the base case value.  Thus, the results in the base case are well converged,
 and so we can believe the results.
@@
 \Nxi~is used. Results for even and odd \Nxi~converge to the same value, as one would hope.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=1.0\textwidth]{sfincsConvergenceScan2.pdf}
+\mycaption{Plot generated by \sfincsScanPlot~showing a typical resolution convergence scan for W7-X,
+as you might generate during routine use of \sfincs.
+\label{fig:convergence2}}
+\end{center}
+\end{figure}
+
+For routine use of the code, it is not necessary to include as many
+runs in the convergence scan as shown in figure \ref{fig:convergence},
+and so a more typical routine scan would look like figure \ref{fig:convergence2}.
+This scan is for a pure hydrogen plasma in the W7-X standard configuration, using \parlink{rN} = 0.19,
+$n_e$ = $10^{20}$ m$^{-3}$, $T_i$ = 3.6 keV, and $T_e$ = 5.5 keV.
+Notice the vertical scales in figure \ref{fig:convergence2} have suppressed zeros.
+For all of the physical output quantities, the results change by no more than a few percent when each resolution
+parameter is increased by 50\%. Hence we can conclude that 
+the results are suitably converged at the base case resolution parameters.
+The \sfincs~resolution parameters and \sfincsScan~parameters used for this scan were as follows:\\
+{\ttfamily
+\Ntheta = 19 \\
+!ss \parlink{NthetaMinFactor} = 0.7 \\
+!ss \parlink{NthetaMaxFactor} = 1.5 \\
+!ss \parlink{NthetaNumRuns} = 6 \\
+\Nzeta = 125 \\
+!ss \parlink{NzetaMinFactor} = 0.7 \\
+!ss \parlink{NzetaMaxFactor} = 1.5 \\
+!ss \parlink{NzetaNumRuns} = 8 \\
+\Nxi = 140 \\
+!ss \parlink{NxiMinFactor} = 0.7 \\
+!ss \parlink{NxiMaxFactor} = 1.5 \\
+!ss \parlink{NxiNumRuns} = 8 \\
+\Nx = 5 \\
+!ss \parlink{NxMinFactor} = 1 \\
+!ss \parlink{NxMaxFactor} = 1.6 \\
+!ss \parlink{NxNumRuns} = 4 \\
+}
+These \sfincsScan~parameters are good for routine convergence tests, although the \sfincs~parameters proper
+(\Ntheta, \Nzeta, \Nxi, and \Nx) should be tailored to your application.
 
@@
 have already been submitted, so duplicate runs will not be generated.
-For example, consider the scan of \Ntheta~described previously in this section,
+For example, consider the scan of \Ntheta~described at the start of this section,
 and suppose you wish to fill in the gaps at \Ntheta = 17 and 21 in the original scan.

### 2015-06-06 doc/manual/runs.tex
Added convergence scan figure to manual

@@
 $\bar{R}$ = reference length, typically 1 meter.\\
-$\bar{n}$ = reference density, typically $10^{19}$ m$^{-3}$, $10^{20}$ m$^{-3}$, or something similar.\\
+$\bar{n}$ = reference density, typically $10^{20}$ m$^{-3}$, $10^{19}$ m$^{-3}$, or something similar.\\
 $\bar{m}$ = reference mass, typically either the mass of hydrogen or deuterium.\\
@@
 see section \ref{sec:geometryParameters}.
+
+\section{Poloidal and toroidal angles}
 

### 2015-06-07 doc/manual/resolution.tex
Now displaying spread percentages in sfincsScanPlot_1

@@
 The red vertical lines emphasize the ``base case'' resolution parameters.
+In each figure the spread is printed, defined as the maximum value $-$ minimum value,
+divided by the value half-way between maximum and minimum.
+If some runs have resolution below the base resolution,
+the spread is also computed and printed
+excluding runs below the base case resolution. (This is the second percentage printed in each plot,
+and is usually more important than the first spread percentage.)
 For all parameters scanned, the physical output quantities (particle flux, heat flux, etc)
-do not visibly change (on the scale of the plots) when any of the resolution parameters
-are increased beyond the base case value.  Thus, the results in the base case are well converged,
+do not visibly change on the scale of the plots when any of the resolution parameters
+are increased beyond the base case value.  Indeed, the spread for each output quantity is $\le$ 3.5\%
+when any resolution parameter is increased.
+Thus, the results in the base case are well converged,
 and so we can believe the results.
@@
 Notice the vertical scales in figure \ref{fig:convergence2} have suppressed zeros.
-For all of the physical output quantities, the results change by no more than a few percent when each resolution
+For all of the physical output quantities, the spread beyond the base case is $\le$ 3.4\%,
+i.e. results change by no more than this percentage when each resolution
 parameter is increased by 50\%. Hence we can conclude that 

### 2015-06-15 doc/manual/inputParameters.tex
First benchmarks with matlab version 3 successful

@@
 {Always}
-{If this switch is set to true, the matrix, right-hand-side, and solution of the
-linear system will be saved in PETSc's binary format.  The preconditioner matrix will also
+{If this switch is set to true, the matrix, right-hand-side, and solution vector of the
+linear system will be saved in PETSc's binary format.  
+The preconditioner matrix will also
 be saved if \parlink{useIterativeLinearSolver} == \true.}
+These matrices and vectors are not very interesting for routine use of the code,
+only for code development and debugging.
+Regardless of how this parameter is set,
+the physically interesting input and output quantities will be saved in a
+separate \HDF~file.
 
@@
 {Only when \parlink{saveMatricesAndVectorsInBinary} == \true.}
-{Start of the filenames which will be used for binary output.}
+{Start of the filenames which will be used for binary output of the system
+matrices, right-hand-side vectors, and solution vectors.
+These matrices and vectors are not very interesting for routine use of the code,
+only for code development and debugging.
+Regardless of how this parameter is set,
+the physically interesting input and output quantities will be saved in a
+separate \HDF~file.
+}
 

### 2015-06-15 fortran/version3/evaluateResidual.F90

@@ -22,6 +22,8 @@
     PetscScalar :: dPhiHatdpsiHatToUseInRHS
     PetscReal :: norm
     integer :: ixMin
+    PetscViewer :: viewer
+    character(len=200) :: filename
 
     if (masterProc) then
        print *,"evaluateResidual called."
@@ -182,5 +184,33 @@
     scalar = -1
     call VecAXPY(residualVec, scalar, rhs, ierr)
     call VecDestroy(rhs, ierr)
+
+    if (saveMatlabOutput) then
+       write (filename,fmt="(a,i3.3,a)") trim(MatlabOutputFilename) // "_iteration_", iterationForResidualOutput, &
+            "_residual.m"
+       if (masterProc) then
+          print *,"Saving residual in matlab format: ",trim(filename)
+       end if
+       call PetscViewerASCIIOpen(MPIComm, trim(filename), viewer, ierr)
+       call PetscViewerSetFormat(viewer, PETSC_VIEWER_ASCII_MATLAB, ierr)
+
+       call PetscObjectSetName(residualVec, "residual", ierr)
+       call VecView(residualVec, viewer, ierr)
+
+       call PetscViewerDestroy(viewer, ierr)
+    end if
+
+    if (saveMatricesAndVectorsInBinary) then
+       write (filename,fmt="(a,i3.3,a)") trim(binaryOutputFilename) // "_iteration_", iterationForResidualOutput, &
+            "_residual"
+       if (masterProc) then
+          print *,"Saving residual in binary format: ",trim(filename)
+       end if
+       call PetscViewerBinaryOpen(MPIComm, trim(filename), FILE_MODE_WRITE, viewer, ierr)
+       call VecView(residualVec, viewer, ierr)
+       call PetscViewerDestroy(viewer, ierr)
+    end if
+
+    iterationForResidualOutput = iterationForResidualOutput + 1
 
   end subroutine evaluateResidual

### 2015-06-15 fortran/version3/populateMatrix.F90

@@ -53,8 +53,8 @@
     double precision :: myMatInfo(MAT_INFO_SIZE)
     integer :: NNZ, NNZAllocated, NMallocs
     PetscScalar :: CHat_element, dfMdx
-    character(len=100) :: whichMatrixName, filename
-    PetscViewer :: MatlabOutput
+    character(len=200) :: whichMatrixName, filename
+    PetscViewer :: viewer
     integer :: ithetaRow, ithetaCol, izetaRow, izetaCol, ixMin, ixMinCol
     VecScatter :: vecScatterContext
     Vec :: vecOnEveryProc
@@ -1363,8 +1363,8 @@
                          end do
                       end if
                       
+                      !   if (.false.) then
                       if (L < NL) then
-                         !   if (.false.) then
                          ! Add Rosenbluth potential terms.
 
                          if (xGridScheme==5 .or. xGridScheme==6) then
@@ -1861,13 +1861,24 @@
        if (masterProc) then
           print *,"Saving matrix in matlab format: ",trim(filename)
        end if
-       call PetscViewerASCIIOpen(MPIComm, trim(filename) , MatlabOutput, ierr)
-       call PetscViewerSetFormat(MatlabOutput, PETSC_VIEWER_ASCII_MATLAB, ierr)
+       call PetscViewerASCIIOpen(MPIComm, trim(filename), viewer, ierr)
+       call PetscViewerSetFormat(viewer, PETSC_VIEWER_ASCII_MATLAB, ierr)
 
        call PetscObjectSetName(matrix, "matrix", ierr)
-       call MatView(matrix, MatlabOutput, ierr)
+       call MatView(matrix, viewer, ierr)
 
-       call PetscViewerDestroy(MatlabOutput, ierr)
+       call PetscViewerDestroy(viewer, ierr)
+    end if
+
+    if (saveMatricesAndVectorsInBinary) then
+       write (filename,fmt="(a,i3.3,a,i1)") trim(binaryOutputFilename) // "_iteration_", iterationForMatrixOutput, &
+            "_whichMatrix_",whichMatrix
+       if (masterProc) then
+          print *,"Saving matrix in binary format: ",trim(filename)
+       end if
+       call PetscViewerBinaryOpen(MPIComm, trim(filename), FILE_MODE_WRITE, viewer, ierr)
+       call MatView(matrix, viewer, ierr)
+       call PetscViewerDestroy(viewer, ierr)
     end if
 
     if (useStateVec) then

### 2015-06-15 matlab/version3/src/sfincs_compareToFortran.m

@@ -11,11 +11,45 @@ global dBHat_sup_theta_dpsiHat dBHat_sup_theta_dzeta
 global dBHat_sup_zeta_dpsiHat dBHat_sup_zeta_dtheta
 global psiHat psiN rHat rN
 
+global FSADensityPerturbation FSABFlow FSAPressurePerturbation
+global particleFlux_vm0_psiHat particleFlux_vm_psiHat particleFlux_vE0_psiHat particleFlux_vE_psiHat
+global momentumFlux_vm0_psiHat momentumFlux_vm_psiHat momentumFlux_vE0_psiHat momentumFlux_vE_psiHat
+global heatFlux_vm0_psiHat heatFlux_vm_psiHat heatFlux_vE0_psiHat heatFlux_vE_psiHat
+global jHat FSABjHat FSABjHatOverB0 FSABjHatOverRootFSAB2
+global totalDensity totalPressure velocityUsingFSADensity velocityUsingTotalDensity MachUsingFSAThermalSpeed
+
 %filename = 'sfincsOutput.h5';
+
+succeeded = false;
+try
+    h5read(filename,'/BHat');
+    succeeded = true;
+catch
+end
+
+if ~succeeded
+    % Try changing / to \ in filename, in case we are on a Windows
+    % filesystem.
+    originalFilename = filename;
+    filename = strrep(filename,'/','\')
+    try
+        h5read(filename,'/BHat');
+    catch
+        fprintf('Unable to compare to fortran result: unable to open %s\n',originalFilename)
+        return
+    end
+end
+
 
 fprintf('Comparing matlab results to fortran results in %s\n',filename)
 
-small = 1e-12;
+% First compare 'input' quantities that should agree to within roundoff
+% error.
+
+tolerance = 1e-12;
+comparisonType = 1;
+% 1 = max(abs(difference)) < tolerance
+% 2 = max(abs(difference) ./ mean) < tolerance
 
 compare('theta')
 compare('zeta')
@@ -58,17 +92,38 @@ compare('dBHat_sup_theta_dzeta')
 compare('dBHat_sup_zeta_dtheta')
 compare('dBHat_sup_zeta_dpsiHat')
 
+% Now compare 'output' quantities that will differ beyond the solver
+% tolerance.
+
+tolerance = 0.003;
+comparisonType = 2;
+
+compare('FSABFlow')
+compare('FSABjHat')
+compare('particleFlux_vm_psiHat')
+compare('heatFlux_vm_psiHat')
+
[…]

### 2015-06-15 matlab/version3/src/sfincs_createGrids.m

@@ -1,12 +1,12 @@
 function sfincs_createGrids()
 
-global Ntheta Nzeta Nx NxPotentialsPerVth xMax xGrid_k
+global Ntheta Nzeta Nx Nxi NL NxPotentialsPerVth xMax solverTolerance xGrid_k
 global constraintScheme collisionOperator NPeriods pointAtX0
 global thetaDerivativeScheme zetaDerivativeScheme forceOddNthetaAndNzeta
 global theta ddtheta thetaWeights ddtheta_preconditioner preconditioner_theta
 global zeta ddzeta zetaWeights ddzeta_preconditioner preconditioner_zeta
 global x xWeights ddx d2dx2 ddx_preconditioner xGridScheme xPotentialsGridScheme
-global theta2D zeta2D RHSMode preconditioner_x
+global theta2D zeta2D RHSMode preconditioner_x xMaxPotentials
 global xInterpolationScheme xPotentialsInterpolationScheme NxPotentials
 global xPotentials ddxPotentials d2dx2Potentials interpolateXToXPotentials
 
@@ -30,6 +30,18 @@ if forceOddNthetaAndNzeta
         Ntheta=Ntheta+1;
     end
 end
+
+fprintf('---- Numerical parameters: ----\n')
+fprintf('            Ntheta = %d\n',Ntheta)
+fprintf('             Nzeta = %d\n',Nzeta)
+fprintf('               Nxi = %d\n',Nxi)
+fprintf('                NL = %d\n',NL)
+fprintf('                Nx = %d\n',Nx)
+if xGridScheme<5
+    fprintf('NxPotentialsPerVth = %d\n',NxPotentialsPerVth)
+    fprintf('              xMax = %d\n',xMax)
+end
+fprintf('   solverTolerance = %d\n',solverTolerance)
 
 sfincs_computeMatrixSize()
 
@@ -135,8 +147,6 @@ switch xPotentialsGridScheme
         error('Invalid xPotentialsInterpolationScheme')
 end
 
-weight = @(xx) (xx.^xGrid_k) .* exp(-xx.*xx);
-            
 % Set pointAtX0, x, xWeights, ddx, d2dx2
 if RHSMode==3
     % Monoenergetic calculation
@@ -160,10 +170,10 @@ else
                 end
                 pointAtX0 = true;
             end
[…]

### 2015-06-15 matlab/version3/src/sfincs_diagnostics.m

@@ -1,8 +1,21 @@
 function sfincs_diagnostics()
 
-global stateVector f0
-global Ntheta Nzeta Nspecies Nx Nxi
+global stateVector f0 constraintScheme sources force0RadialCurrentInEquilibrium
+global Ntheta Nzeta Nspecies Nx thetaWeights zetaWeights RHSMode
 global Phi1Hat dPhi1Hatdtheta dPhi1Hatdzeta ddtheta ddzeta includePhi1
+global BLOCK_F BLOCK_QN BLOCK_PHI1_CONSTRAINT BLOCK_DENSITY_CONSTRAINT BLOCK_PRESSURE_CONSTRAINT BLOCK_F_CONSTRAINT
+global BHat DHat BHat_sub_theta BHat_sub_zeta dBHatdtheta dBHatdzeta dBHat_sub_zeta_dtheta dBHat_sub_theta_dzeta
+global THats Zs mHats nHats x xWeights Delta alpha
+global B0OverBBar FSABHat2 VPrimeHat
+
+global FSADensityPerturbation FSABFlow FSAPressurePerturbation
+global particleFlux_vm0_psiHat particleFlux_vm_psiHat particleFlux_vE0_psiHat particleFlux_vE_psiHat
+global momentumFlux_vm0_psiHat momentumFlux_vm_psiHat momentumFlux_vE0_psiHat momentumFlux_vE_psiHat
+global heatFlux_vm0_psiHat heatFlux_vm_psiHat heatFlux_vE0_psiHat heatFlux_vE_psiHat
+global jHat FSABjHat FSABjHatOverB0 FSABjHatOverRootFSAB2
+global totalDensity totalPressure velocityUsingFSADensity velocityUsingTotalDensity MachUsingFSAThermalSpeed
+
+fprintf('Computing diagnostics.\n')
 
 if includePhi1
     for itheta = 1:Ntheta
@@ -12,8 +25,216 @@ if includePhi1
     
     dPhi1Hatdtheta = ddtheta * Phi1Hat;
     dPhi1Hatdzeta = (ddzeta * (Phi1Hat'))';
+    
+    index = sfincs_indices(1,1,1,1,1,BLOCK_PHI1_CONSTRAINT);
+    lambda = stateVector(index);
+else
+    lambda = 0;
+end
+
+switch constraintScheme
+    case 0
+        sources = [];
+    case 1
+        sources = zeros(Nspecies,2);
[…]

### 2015-06-15 matlab/version3/src/sfincs_evaluateResidual.m

@@ -1,12 +1,14 @@
 function residual = sfincs_evaluateResidual()
 
 global stateVector matrixSize RHSMode pointAtX0 dPhiHatdpsiHat includeTemperatureEquilibrationTerm f0
-global x Nspecies Ntheta Nzeta Nx Delta alpha
-global Zs THats mHats nHats dnHatdpsiHats dTHatdpsiHats
+global x Nspecies Ntheta Nzeta Nx Delta alpha BLOCK_F
+global Zs THats mHats nHats dnHatdpsiHats dTHatdpsiHats EParallelHat
+global BHat DHat FSABHat2 BHat_sub_zeta BHat_sub_theta
+global dBHatdtheta dBHatdzeta
 
 fprintf('Evaluating residual.\n')
 
-if norm(stateVector)<1e-15
+if norm(stateVector)>1e-100
     % Part of the residual comes from multiplying the state vector by a big
     % matrix.
     whichMatrix = 3;
@@ -40,6 +42,8 @@ end
 
 x2 = x.*x;
 expx2 = exp(-x2);
+sqrtpi = sqrt(pi);
+
 allZeta = 1:Nzeta;
 rhs = zeros(matrixSize,1);
 for ispecies = 1:Nspecies
@@ -51,9 +55,9 @@ for ispecies = 1:Nspecies
     sqrtmHat = sqrt(mHat);
     
     spatialFactor = Delta*nHat*mHat*sqrtmHat ...
-        /(2*pi*sqrtpi*Z*(BHat.^3)*sqrtTHat) ...
-        *(BHat_sub_zeta*dBHatdtheta - BHat_sub_theta*dBHatdzeta) ...
-        * DHat;
+        ./(2*pi*sqrtpi*Z*(BHat.^3)*sqrtTHat) ...
+        .*(BHat_sub_zeta.*dBHatdtheta - BHat_sub_theta.*dBHatdzeta) ...
+        .* DHat;
     
     for ix = ixMin:Nx
         xPartOfRHS = x2(ix)*expx2(ix)*( dnHatdpsiHats(ispecies)/nHat ...

### 2015-06-15 matlab/version3/src/sfincs_main.m

@@ -1,5 +1,7 @@
 function sfincs_main()
 
+startTime = tic;
+
 fprintf('***************************************************************************\n')
 fprintf('SFINCS: Stellarator Fokker-Plank Iterative Neoclassical Conservative Solver\n')
 fprintf('Version 3 - Matlab serial edition\n')
@@ -40,5 +42,8 @@ sfincs_setInputRadialCoordinate()
 sfincs_solver()
 
 %sfincs_finalizeHDF5()
+
+fprintf('Total execution time: %g seconds.\n',toc(startTime))
+fprintf('Good bye!\n')
 
 end

### 2015-06-15 matlab/version3/src/sfincs_solveLinearSystem.m

@@ -1,8 +1,8 @@
-function soln = sfincs_solveLinearSystem(matrix, rhs, preconditioner_L, preconditioner_U, preconditioner_P, preconditioner_Q)
+function soln = sfincs_solveLinearSystem(matrix, rhs)
 
 global GMRES_restart GMRES_maxIterations solverTolerance useIterativeLinearSolver
 
-if useIterativeLinearSolver
+if ~useIterativeLinearSolver
     tic
     fprintf('Beginning direct linear solve... ')
     soln = matrix \ rhs;
@@ -13,7 +13,7 @@ else
     fprintf('Applying GMRES... ')
     x0 = zeros(size(rhs));
     [soln,fl0,rr0,it0,rv0]=gmres(matrix,rhs,GMRES_restart,solverTolerance,GMRES_maxIterations/GMRES_restart,@preconditioner, [], x0);
-    fprintf('Done. Took %g seconds.\n',toc)
+    fprintf('Done. Took %g seconds, %d iterations.\n',toc, numel(rv0))
     switch fl0
         case 0
             fprintf('Converged!\n')
@@ -47,6 +47,7 @@ else
 end
 
     function solnVector = preconditioner(rhsVector)
+        global preconditioner_L preconditioner_U preconditioner_P preconditioner_Q
         solnVector = preconditioner_Q * (preconditioner_U \ (preconditioner_L \ (preconditioner_P * rhsVector)));
     end
 

### 2015-06-19 doc/manual/inputParameters.tex
bugfix in matlab version3

@@
 
-{\ttfamily magneticDriftScheme} = 1: Use the grad-B and curvature drift, plus the parallel velocity correction $v_{\perp}^2/(2 \Omega_c) \vect{b} \vect{b} \cdot \nabla \times \vect{b}$.\\
-
-{\ttfamily magneticDriftScheme} = 2:   Use the magnetic drift $v_m = (v_{||} / \Omega_c) \nabla (v_{||} \vect{b})$.
+{\ttfamily magneticDriftScheme} = 1: Use the magnetic drift $v_m = (v_{||} / \Omega_c) \nabla \times (v_{||} \vect{b})$.\\
+
+{\ttfamily magneticDriftScheme} = 2: Use the grad-B and curvature drift, plus the parallel velocity correction $v_{\perp}^2/(2 \Omega_c) \vect{b} \vect{b} \cdot \nabla \times \vect{b}$.
 }

### 2015-06-19 matlab/version3/src/sfincs_compareToFortran.m

@@ -94,11 +94,11 @@ compare('dBHat_sub_theta_dzeta')
 compare('dBHat_sub_zeta_dtheta')
 compare('dBHat_sub_zeta_dpsiHat')
 
-compare('dBHat_sup_theta_dpsiHat')
+%compare('dBHat_sup_theta_dpsiHat')
 compare('dBHat_sup_theta_dzeta')
 
 compare('dBHat_sup_zeta_dtheta')
-compare('dBHat_sup_zeta_dpsiHat')
+%compare('dBHat_sup_zeta_dpsiHat')
 
 % Now compare 'output' quantities that will differ beyond the solver
 % tolerance.

### 2015-06-19 matlab/version3/src/sfincs_populateMatrix.m

@@ -142,7 +142,7 @@ for ispecies = 1:Nspecies
                     % Diagonal in L
                     colIndices = rowIndices;
                     addSparseBlock(rowIndices, colIndices, ExBTerm + nonlinearTerm ...
-                        + x2(ix)*(magneticDriftTerm1+magneticDriftTerm2)*(L-1)*L/((2*L-3)*(2*L-1))...
+                        + x2(ix)*(magneticDriftTerm1*2*(3*L*L+3*L-2)+magneticDriftTerm2*(2*L*L+2*L-1))/((2*L+3)*(2*L-1))...
                         + x2(ix)*magneticDriftTerm3*(-2)*L*(L+1)/((2*L+3)*(2*L-1)))
     
                     % Super-diagonal in L
@@ -197,6 +197,7 @@ for ispecies = 1:Nspecies
             ExBTermSpatialPart = ExBTermSpatialPart ./ (BHat.*BHat);
         end
         
+        
         magneticDriftFactor = Delta*THat*DHat./(2*Z*(BHat.^3));
         if magneticDriftScheme>0
             magneticDriftSpatialPart1 = magneticDriftFactor .* (-BHat_sub_theta.*dBHatdpsiHat + BHat_sub_psi.*dBHatdtheta);
@@ -210,7 +211,11 @@ for ispecies = 1:Nspecies
         else
             magneticDriftSpatialPart3 = zeros(Ntheta,Nzeta);
         end
-        
+        %{
+        magneticDriftSpatialPart1 = zeros(Ntheta,Nzeta);
+        magneticDriftSpatialPart2 = zeros(Ntheta,Nzeta);
+        magneticDriftSpatialPart3 = zeros(Ntheta,Nzeta);
+        %}
         if nonlinear
             nonlinearTermSpatialPart = alpha*Delta*DHat.*BHat_sub_psi.*dPhi1Hatdtheta./(2*BHat.*BHat);
         else
@@ -238,7 +243,7 @@ for ispecies = 1:Nspecies
                     % Diagonal in L
                     colIndices = rowIndices;
                     addSparseBlock(rowIndices, colIndices, ExBTerm + nonlinearTerm ...
-                        + x2(ix)*(magneticDriftTerm1+magneticDriftTerm2)*(L-1)*L/((2*L-3)*(2*L-1))...
[…]

### 2015-07-01 doc/manual/runs.tex
Tweaks for compatibility with petsc versions prior to 3.5

@@
 
-\section{Poloidal and toroidal angles}
-
 \section{Trajectory models}
@@
   Probably not.}
+
+\section{Poloidal and toroidal angles}
+
+If you are interested in any of the output quantities that vary on a flux surface,
+such as the density or electrostatic potential, then it is important to know
+how the poloidal and toroidal angles ($\theta$ and $\zeta$) in \sfincs~are defined.
+The definitions of the poloidal and toroidal angles in 
+\sfincs~depend on the input parameter \parlink{geometryScheme}. When a \vmec~equilibrium is imported by setting
+\parlink{geometryScheme} = 5, then \sfincs~will use the same poloidal and toroidal angles
+as \vmec.  The toroidal angle in this case is the normal cylindrical coordinate. Note that field lines
+are not straight in these \vmec~coordinates.
+For any other setting of \parlink{geometryScheme}, \sfincs~will use Boozer coordinates.

### 2015-07-01 fortran/version3/evaluateJacobian.F90

@@ -4,7 +4,7 @@
 
 #if (PETSC_VERSION_MAJOR < 3 || (PETSC_VERSION_MAJOR==3 && PETSC_VERSION_MINOR < 5))
        ! Syntax for PETSc versions up through 3.4
-  subroutine evaluateJacobian(mysnes, stateVec, jacobian, jacobianPC, userContext, flag, ierr)
+  subroutine evaluateJacobian(mysnes, stateVec, jacobian, jacobianPC, flag, userContext, ierr)
 #else
        ! Syntax for PETSc version 3.5 and later
   subroutine evaluateJacobian(mysnes, stateVec, jacobian, jacobianPC, userContext, ierr)
@@ -30,7 +30,8 @@
     end if
 
 #if (PETSC_VERSION_MAJOR < 3 || (PETSC_VERSION_MAJOR==3 && PETSC_VERSION_MINOR < 5))
-    ! Syntax for PETSc versions up through 3.4
+    ! Syntax for PETSc versions up through 3.4.
+    ! For PETSc versions 3.5 and later, the code related to 'reusePreconditioner' is implemented in solver.F90.
     if (reusePreconditioner) then
        flag = SAME_PRECONDITIONER
     else

### 2015-07-01 fortran/version3/solver.F90

@@ -29,6 +29,8 @@ module solver
     Mat :: factorMat
     PetscInt :: mumps_which_cntl
     PetscReal :: mumps_value
+    PetscReal :: atol, rtol, stol
+    integer :: maxit, maxf
 
     external evaluateJacobian, evaluateResidual, diagnosticsMonitor
 
@@ -99,6 +101,11 @@ module solver
           if (masterProc) then
              print *,"We will use mumps to factorize the preconditioner."
           end if
+#if (PETSC_VERSION_MAJOR < 3 || (PETSC_VERSION_MAJOR==3 && PETSC_VERSION_MINOR < 5))
+          ! The functions MatMumpsSetICNTL were introduced in PETSc 3.5.
+          ! For earlier versions, we can achieve a similar result with the following hack:
+          call PetscOptionsInsertString("-mat_mumps_cntl_1 1e-6 -mat_mumps_icntl_4 2", ierr)
+#endif
        case (2)
           call PCFactorSetMatSolverPackage(preconditionerContext, MATSOLVERSUPERLU_DIST, ierr)
           if (masterProc) then
@@ -148,21 +155,39 @@ module solver
 #endif
     end if
 
-    ! Set the algorithm to use for the nonlinear solver:
+    ! In older versions of PETSC (either <3.5 or <3.4, I'm not certain)
+    ! the monitor is never called when snes type = SNESKSPONLY.
+    ! Therefore, it is preferable to always have snes type = SNESNEWTONLS but set the 
+    ! number of iterations to 1 for a linear run.
+    call SNESSetType(mysnes, SNESNEWTONLS, ierr)
     if (nonlinear) then
-       ! SNESNEWTONLS = Newton's method with an optional line search.
-       ! As of PETSc version 3.5, this is the default algorithm, but I'll set it manually here anyway to be safe.
-       call SNESSetType(mysnes, SNESNEWTONLS, ierr)
        if (masterProc) then
           print *,"Since this is a nonlinear run, we will use Newton's method."
        end if
     else
-       ! SNESKSPONLY = Only do 1 linear step.
-       call SNESSetType(mysnes, SNESKSPONLY, ierr)
+       call SNESGetTolerances(mysnes, atol, rtol, stol, maxit, maxf, ierr)
+       maxit = 1
+       call SNESSetTolerances(mysnes, atol, rtol, stol, maxit, maxf, ierr)
        if (masterProc) then
[…]

### 2015-07-07 doc/manual/inputParameters.tex
Added ExBDerivativeScheme option for upwinding

@@
 {Whenever \parlink{RHSMode} == 1.}
-{This variable controls the poloidal and magnetic drifts, and does not affect the radial magnetic drift.
-THIS FUNCTIONALITY IS WORK IN PROGRESS, AND RESULTS ARE BUGGY WHEN THIS VARIABLE IS NONZERO.\\
+{This variable controls the poloidal and magnetic drifts, and does not affect the radial magnetic drift.\\
 
@@
 The best value for this parameter is usually 2.}
+
+\myhrule
+
+\param{ExBDerivativeScheme}
+{integer}
+{0}
+{Whenever the radial electric field is nonzero}
+{Options for controlling upwinding of the $\vect{E}\times\vect{B}$ drift terms.  Note the options are different than for \parlink{thetaDerivativeScheme}
+and \parlink{zetaDerivativeScheme}.\\
+
+{\ttfamily ExBDerivativeScheme} = 0: Use same differentiation matrices as for the parallel streaming term.  This option is best except for very
+large radial electric fields, when the electric field is as large as the resonance or larger.\\
+
+{\ttfamily ExBDerivativeScheme} = 1: Treat the $\vect{E}\times\vect{B}$ drift terms using a 4 point upwinded stencil (1 point on 1 side, 2 points on the other.)
+}
 

### 2015-07-07 fortran/version3/createGrids.F90

@@ -201,6 +201,8 @@
     allocate(theta(Ntheta))
     allocate(thetaWeights(Ntheta))
     allocate(ddtheta(Ntheta,Ntheta))
+    allocate(ddtheta_ExB_plus(Ntheta,Ntheta))
+    allocate(ddtheta_ExB_minus(Ntheta,Ntheta))
     allocate(d2dtheta2(Ntheta,Ntheta))
     allocate(theta_preconditioner(Ntheta))
     allocate(thetaWeights_preconditioner(Ntheta))
@@ -221,7 +223,27 @@
        stop
     end select
 
+!    scheme = 90 ! Delete this line!
     call uniformDiffMatrices(Ntheta, zero, two*pi, scheme, theta, thetaWeights, ddtheta, d2dtheta2)
+!!$    print *,"Here comes ddtheta:"
+!!$    do i=1,Ntheta
+!!$       print *,ddtheta(i,:)
+!!$    end do
+!!$    print *,"Here comes d2dtheta2:"
+!!$    do i=1,Ntheta
+!!$       print *,d2dtheta2(i,:)
+!!$    end do
+!!$    stop
+
+
+    ! Create upwinded matrices for ExB terms:
+    scheme = 80
+    call uniformDiffMatrices(Ntheta, zero, two*pi, scheme, theta_preconditioner, &
+         thetaWeights_preconditioner, ddtheta_ExB_plus, d2dtheta2_preconditioner)
+    scheme = 90
+    call uniformDiffMatrices(Ntheta, zero, two*pi, scheme, theta_preconditioner, &
+         thetaWeights_preconditioner, ddtheta_ExB_minus, d2dtheta2_preconditioner)
+
 
     ! If needed, also make a sparser differentiation matrix for the preconditioner:
     select case(preconditioner_theta)
@@ -271,6 +293,8 @@
     allocate(zeta(Nzeta))
     allocate(zetaWeights(Nzeta))
     allocate(ddzeta(Nzeta,Nzeta))
+    allocate(ddzeta_ExB_plus(Nzeta,Nzeta))
+    allocate(ddzeta_ExB_minus(Nzeta,Nzeta))
     allocate(d2dzeta2(Nzeta,Nzeta))
     allocate(zeta_preconditioner(Nzeta))
     allocate(zetaWeights_preconditioner(Nzeta))
@@ -297,8 +321,18 @@
        zetaWeights = 2*pi
        ddzeta = 0
        d2dzeta2 = 0 ! d2dzeta2 is not actually used.
[…]

### 2015-07-07 fortran/version3/populateMatrix.F90

@@ -360,10 +360,20 @@
           allocate(colIndices(Ntheta))
           do L=0,(Nxi-1)
 
-             if (whichMatrix>0 .or. L < preconditioner_theta_min_L) then
-                ddthetaToUse = ddtheta
+             if (ExBDerivativeScheme==0) then
+                if (whichMatrix>0 .or. L < preconditioner_theta_min_L) then
+                   ddthetaToUse = ddtheta
+                else
+                   ddthetaToUse = ddtheta_preconditioner
+                end if
              else
-                ddthetaToUse = ddtheta_preconditioner
+                ! Assume DHat has the same sign everywhere. (Should be true even for VMEC coordinates.)
+                ! Assume BHat_sub_zeta has the same sign everywhere. (True for Boozer but in principle could be violated for VMEC coordinates?)
+                if (factor*DHat(1,1)*BHat_sub_zeta(1,1) > 0) then
+                   ddthetaToUse = ddtheta_ExB_plus
+                else
+                   ddthetaToUse = ddtheta_ExB_minus
+                end if
              end if
 
              do izeta=izetaMin,izetaMax
@@ -414,10 +424,20 @@
           allocate(colIndices(Nzeta))
           do L=0,(Nxi-1)
 
-             if (whichMatrix>0 .or. L < preconditioner_zeta_min_L) then
-                ddzetaToUse = ddzeta
+             if (ExBDerivativeScheme==0) then
+                if (whichMatrix>0 .or. L < preconditioner_zeta_min_L) then
+                   ddzetaToUse = ddzeta
+                else
+                   ddzetaToUse = ddzeta_preconditioner
+                end if
              else
-                ddzetaToUse = ddzeta_preconditioner
+                ! Assume DHat has the same sign everywhere. (Should be true even for VMEC coordinates.)
+                ! Assume BHat_sub_theta has the same sign everywhere. (True for Boozer but could be violated for VMEC coordinates?)
+                if (factor*DHat(1,1)*BHat_sub_theta(1,1) > 0) then
+                   ddzetaToUse = ddzeta_ExB_plus
+                else
+                   ddzetaToUse = ddzeta_ExB_minus
+                end if
              end if
 
              do itheta=ithetaMin, ithetaMax

### 2015-07-07 fortran/version3/writeHDF5Output.F90

@@ -177,6 +177,7 @@ contains
        call writeHDF5Field("geometryScheme", geometryScheme, "")
        call writeHDF5Field("thetaDerivativeScheme", thetaDerivativeScheme, "")
        call writeHDF5Field("zetaDerivativeScheme", zetaDerivativeScheme, "")
+       call writeHDF5Field("ExBDerivativeScheme", ExBDerivativeScheme, "")
        call writeHDF5Field("xGridScheme", xGridScheme, "")
        call writeHDF5Field("xGrid_k", xGrid_k, "Exponent of x in the orthogonality relation for the speed polynomials")
        call writeHDF5Field("xPotentialsGridScheme", xPotentialsGridScheme, "")

### 2015-07-21 doc/manual/overview.tex

@@
 To preserve previous versions of the code that have been used for publications, two older versions of the
-code called singleSpecies and multiSpecies are also present in the repository.  For both of these older versions,
-both MATLAB and fortran editions exist which are independent of each other.  At present, version 3 exists
-only in a fortran edition.
+code called singleSpecies and multiSpecies are also present in the repository.  
+For all versions, both MATLAB and fortran editions exist which are independent of each other.
+The MATLAB editions exist primarily for debugging the fortran editions.
+The same algorithms are implemented in the two different languages, and for identical input parameters, 
+the matrices and output quantities from the different editions should agree to several significant digits (roughly within the solver tolerance.)
+Any significant differences between the matrices and output of the MATLAB and fortran editions can be used to identify a bug.
+The MATLAB versions are serial whereas the fortran versions
+are parallelized, so the fortran versions are significantly faster and can access much more memory.
+For realistic experimental geometry and collisionality, the resolution (and hence memory) requirements will mean
+you will need to use the fortran edition.
 

### 2015-07-26 doc/manual/inputParameters.tex
Added ExBDerivativeScheme 2 and 3. Added many options for x grid including upwinding.

@@
 
+\param{ExBDerivativeScheme}
+{integer}
+{0}
+{Whenever the radial electric field is nonzero}
+{Options for controlling upwinding of the $\vect{E}\times\vect{B}$ drift terms.  Note the options are different than for \parlink{thetaDerivativeScheme}
+and \parlink{zetaDerivativeScheme}.  This option exists because when the radial electric field is large enough to be comparable to the resonance,
+such that the coefficient in front of the $\partial f/\partial \theta$ or $\partial f/\partial\zeta$ terms vanishes near the thermal speed of one of the species,
+the distribution function can develop unphysical grid-scale structure.  This option allows upwinding of these spatial derivatives, which eliminates the grid-scale oscillations.\\
+
+{\ttfamily ExBDerivativeScheme} = 0: Use same differentiation matrices for the $\vect{E}\times\vect{B}$ drift terms as for the parallel streaming term.\\
+
+{\ttfamily ExBDerivativeScheme} = 1: Treat the $\vect{E}\times\vect{B}$ drift terms using a 4 point upwinded stencil (1 point on 1 side, 2 points on the other.)\\
+
+{\ttfamily ExBDerivativeScheme} = 2: Treat the $\vect{E}\times\vect{B}$ drift terms using a 5 point upwinded stencil (1 point on 1 side, 3 points on the other.)\\
+
+{\ttfamily ExBDerivativeScheme} = 3: Treat the $\vect{E}\times\vect{B}$ drift terms using a 6 point upwinded stencil (2 points on 1 side, 3 points on the other.)\\
+
+The different settings require slightly different amounts of memory for factorization due to the different numbers of nonzeros. Option 1 requires the least memory, then 0, then 2,
+and 3 requires the most.  The difference in memory required appears to be small, around 10\%.
+On the other hand, the denser the stencil, the more accurate the result. Hence, in order from least to most accurate, the options are 1, 0, 2, 3.
+You should use the default {\ttfamily ExBDerivativeScheme} = 0 except for very large radial electric fields, when the electric field is as large as the resonance or larger.
+In this case, {\ttfamily ExBDerivativeScheme} = 3 is recommended.
+}
+
+\myhrule
+
 \param{xGridScheme}
@@
 
-The recommended value for this parameter is 5.  
+{\ttfamily xGridScheme} = 7: Chebyshev grid on [0, \parlink{xMax}], forcing $f=0$ at \parlink{xMax}. Original treatment of Rosenbluth potentials.\\
+
+{\ttfamily xGridScheme} = 8: Chebyshev grid on [0, \parlink{xMax}], with no boundary condition imposed at \parlink{xMax}. Original treatment of Rosenbluth potentials.\\
+
+The recommended value for this parameter is the default, 5.  
 When {\ttfamily xGridScheme} = 5 or 6, then the following quantities do not matter: 

### 2015-07-26 fortran/version3/createGrids.F90

@@ -25,6 +25,8 @@
     PetscScalar, dimension(:), allocatable :: xWeights_plus1
     PetscScalar, dimension(:,:), allocatable :: ddx_plus1, d2dx2_plus1
     PetscScalar, dimension(:,:), allocatable :: interpolateXToXPotentials_plus1, extrapMatrix
+    PetscScalar, dimension(:), allocatable :: x_subset, xWeights_subset
+    PetscScalar, dimension(:,:), allocatable :: ddx_subset, d2dx2_subset
 
 
     DM :: myDM
@@ -235,15 +237,37 @@
 !!$    end do
 !!$    stop
 
-
     ! Create upwinded matrices for ExB terms:
-    scheme = 80
-    call uniformDiffMatrices(Ntheta, zero, two*pi, scheme, theta_preconditioner, &
-         thetaWeights_preconditioner, ddtheta_ExB_plus, d2dtheta2_preconditioner)
-    scheme = 90
-    call uniformDiffMatrices(Ntheta, zero, two*pi, scheme, theta_preconditioner, &
-         thetaWeights_preconditioner, ddtheta_ExB_minus, d2dtheta2_preconditioner)
-
+    select case (ExBDerivativeScheme)
+    case (0)
+       ! It should not matter what ddtheta_ExB_plus and ddtheta_ExB_minus are in this case.
+       ddtheta_ExB_plus = 0
+       ddtheta_ExB_minus = 0
+    case (1)
+       scheme = 80
+       call uniformDiffMatrices(Ntheta, zero, two*pi, scheme, theta_preconditioner, &
+            thetaWeights_preconditioner, ddtheta_ExB_plus, d2dtheta2_preconditioner)
+       scheme = 90
+       call uniformDiffMatrices(Ntheta, zero, two*pi, scheme, theta_preconditioner, &
+            thetaWeights_preconditioner, ddtheta_ExB_minus, d2dtheta2_preconditioner)
+    case (2)
+       scheme = 100
+       call uniformDiffMatrices(Ntheta, zero, two*pi, scheme, theta_preconditioner, &
+            thetaWeights_preconditioner, ddtheta_ExB_plus, d2dtheta2_preconditioner)
+       scheme = 110
[…]

### 2015-07-26 fortran/version3/populateMatrix.F90

@@ -30,7 +30,9 @@
     PetscScalar :: Z, nHat, THat, mHat, sqrtTHat, sqrtMHat, speciesFactor, speciesFactor2
     PetscScalar :: T32m, factor, LFactor, temp, temp1, temp2, xDotFactor, xDotFactor2, stuffToAdd
     PetscScalar, dimension(:), allocatable :: xb, expxb2
-    PetscScalar, dimension(:,:), allocatable :: thetaPartOfTerm, localThetaPartOfTerm, xPartOfXDot
+    PetscScalar, dimension(:,:), allocatable :: thetaPartOfTerm, localThetaPartOfTerm
+    PetscScalar, dimension(:,:), allocatable :: xPartOfXDot_plus, xPartOfXDot_minus, xPartOfXDot
+    PetscScalar, dimension(:,:), allocatable :: ddxToUse_plus, ddxToUse_minus
     integer :: i, j, ix, ispecies, itheta, izeta, L, ixi, index, ix_row, ix_col
     integer :: rowIndex, colIndex
     integer :: ell, iSpeciesA, iSpeciesB
@@ -174,6 +176,8 @@
     ! *********************************************************
 
     allocate(ddxToUse(Nx,Nx))
+    allocate(ddxToUse_plus(Nx,Nx))
+    allocate(ddxToUse_minus(Nx,Nx))
     allocate(d2dx2ToUse(Nx,Nx))
     allocate(ddthetaToUse(Ntheta, Ntheta))
     allocate(ddzetaToUse(Nzeta, Nzeta))
@@ -805,6 +809,8 @@
        if (includeXDotTerm .and. (whichMatrix .ne. 2)) then
 
           allocate(xPartOfXDot(Nx,Nx))
+          allocate(xPartOfXDot_plus(Nx,Nx))
+          allocate(xPartOfXDot_minus(Nx,Nx))
           allocate(rowIndices(Nx))
           allocate(colIndices(Nx))
           !factor = alpha*Delta/(4*psiAHat)*dPhiHatdpsiN
@@ -817,14 +823,18 @@
                 ixMinCol = 1
              end if
 
+             ! Upwind in x
              if (whichMatrix==0 .and. L >= preconditioner_x_min_L) then
-                ddxToUse = ddx_preconditioner
+                ddxToUse_plus = ddx_xDot_preconditioner_plus
+                ddxToUse_minus = ddx_xDot_preconditioner_minus
              else
-                ddxToUse = ddx
[…]

### 2015-07-26 fortran/version3/readInput.F90

@@ -57,7 +57,7 @@ contains
 
     namelist / otherNumericalParameters /  &
          useIterativeLinearSolver, thetaDerivativeScheme, zetaDerivativeScheme, ExBDerivativeScheme, &
-         xGridScheme, whichParallelSolverToFactorPreconditioner, &
+         xDotDerivativeScheme, xGridScheme, whichParallelSolverToFactorPreconditioner, &
          PETSCPreallocationStrategy, xPotentialsGridScheme, xGrid_k
 
     namelist / preconditionerOptions / preconditioner_x, preconditioner_x_min_L, preconditioner_zeta, &

### 2015-07-26 doc/manual/inputParameters.tex
Added magneticDriftDerivativeScheme to user manual

@@
 {Options for controlling upwinding of the $\vect{E}\times\vect{B}$ drift terms.  Note the options are different than for \parlink{thetaDerivativeScheme}
-and \parlink{zetaDerivativeScheme}.  This option exists because when the radial electric field is large enough to be comparable to the resonance,
+and \parlink{zetaDerivativeScheme}, and are analogous to the options for \parlink{magneticDriftDerivativeScheme}.
+This option exists because when the radial electric field is large enough to be comparable to the resonance,
 such that the coefficient in front of the $\partial f/\partial \theta$ or $\partial f/\partial\zeta$ terms vanishes near the thermal speed of one of the species,
@@
 The different settings require slightly different amounts of memory for factorization due to the different numbers of nonzeros. Option 1 requires the least memory, then 0, then 2,
-and 3 requires the most.  The difference in memory required appears to be small, around 10\%.
+and 3 requires the most.  The difference in memory required appears to be small, around 10-20\% between settings 1 and 3.
 On the other hand, the denser the stencil, the more accurate the result. Hence, in order from least to most accurate, the options are 1, 0, 2, 3.
+Option 1 is substantially less accurate than the other options.
 You should use the default {\ttfamily ExBDerivativeScheme} = 0 except for very large radial electric fields, when the electric field is as large as the resonance or larger.
 In this case, {\ttfamily ExBDerivativeScheme} = 3 is recommended.
+}
+
+\myhrule
+
+\param{magneticDriftDerivativeScheme}
+{integer}
+{3}
+{Whenever \parlink{magneticDriftScheme} is nonzero}
+{Options for controlling upwinding of the magnetic drift terms.  Note the options are different than for \parlink{thetaDerivativeScheme}
+and \parlink{zetaDerivativeScheme}, and are analogous to the options for \parlink{ExBDerivativeScheme}.
+This option exists because when magnetic drift terms involving $\partial f/\partial \theta$ and $\partial f/\partial\zeta$
+are implemented using the same centered difference stencil as the parallel streaming term,
+the distribution function can develop unphysical grid-scale structure.  This option allows upwinding of these spatial derivatives, which eliminates the grid-scale oscillations.\\
+
+{\ttfamily magneticDriftDerivativeScheme} = 0: Use same differentiation matrices for the magnetic drift terms as for the parallel streaming term.\\
+
+{\ttfamily magneticDriftDerivativeScheme} = 1: Treat the magnetic drift terms using a 4 point upwinded stencil (1 point on 1 side, 2 points on the other.)\\
+
+{\ttfamily magneticDriftDerivativeScheme} = 2: Treat the magnetic drift terms using a 5 point upwinded stencil (1 point on 1 side, 3 points on the other.)\\
+
+{\ttfamily magneticDriftDerivativeScheme} = 3: Treat the magnetic drift terms using a 6 point upwinded stencil (2 points on 1 side, 3 points on the other.)\\
+
+The different settings require slightly different amounts of memory for factorization due to the different numbers of nonzeros. Option 1 requires the least memory, then 0, then 2,
+and 3 requires the most.  The difference in memory required appears to be small, around 10-20\% between settings 1 and 3.
+On the other hand, the denser the stencil, the more accurate the result. 
+Hence, in order from least to most accurate, the options are 1, 0, 2, 3. Option 1 is substantially less accurate than the other options.
+The default setting of 3 is robust and recommended in most cases.
 }

### 2015-07-27 doc/manual/runs.tex
Default magneticDriftDerivativeScheme now 3. Bugfix in sfincsScan_3

@@
 
+\section{Poloidal and toroidal magnetic drifts}
+
+You can choose to either include or not include the poloidal and toroidal magnetic drifts.
+These drifts are turned off by default.  To turn them on, all you need to do is
+set \parlink{magneticDriftScheme} = 1 in the {\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}} namelist.
+If on, you must use VMEC
+geometry (\parlink{geometryScheme} = 5).  The magnetic drift terms introduce nonzeros in the system matrix,
+and therefore increase the memory required for factorization; a typical increase is 30-40\%.
+
 \section{Parallelization}
-Choosing the number of nodes and procs.
+[Coming soon...] Choosing the number of nodes and procs.
 

### 2015-07-27 fortran/version3/globalVariables.F90

@@ -117,7 +117,7 @@ module globalVariables
   ! 2 = 4th order dinite differences
 
   integer :: ExBDerivativeScheme = 0
-  integer :: magneticDriftDerivativeScheme = 0
+  integer :: magneticDriftDerivativeScheme = 3
   integer :: xDotDerivativeScheme = 0
 
   integer :: xPotentialsGridScheme = 2, xPotentialsInterpolationScheme

### 2015-07-27 doc/manual/runs.tex
Added examples with magnetic drifts

@@
 set \parlink{magneticDriftScheme} = 1 in the {\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}} namelist.
-If on, you must use VMEC
-geometry (\parlink{geometryScheme} = 5).  The magnetic drift terms introduce nonzeros in the system matrix,
+(Setting \parlink{magneticDriftScheme} = 2 uses a slightly different parallel magnetic drift, which gives
+indistinguishable results to setting 1 for all cases examined so far, and which is in fact exactly identical to setting 1
+in the limit of vanishing plasma beta.)
+If the poloidal/toroidal magnetic drifts are turned on, you must use VMEC
+geometry (\parlink{geometryScheme} = 5), since the magnetic drifts
+depend on various derivatives of the components of the magnetic field which
+are not available in the simplified geometry models.
+
+The magnetic drift terms introduce nonzeros in the system matrix,
 and therefore increase the memory required for factorization; a typical increase is 30-40\%.
+These magnetic drift terms often have only a minor effect, except perhaps when the radial
+electric field is near 0.
+
+If the electrostatic potential is not constant on flux surfaces and poloidal/toroidal magnetic drifts
+are included, certain terms exist in the kinetic equation which have not yet been implemented in the code.
+Thus, at present it is not strictly correct to simultaneously set  \parlink{magneticDriftScheme} > 0 
+and  \parlink{includePhi1} = \true.
 

### 2015-07-27 fortran/version3/examples/filteredW7XNetCDF_2species_magneticDrifts_noEr/tests_small.py

@@ -1,0 +1,24 @@
+#!/usr/bin/env python
+
+# This python script checks the sfincsOutput.h5 file for an example to 
+# see if the results are close to expected values.  This script may be
+# run directly, and it is also called when "make test" is run from the
+# main SFINCS directory.
+
+execfile('../testsCommon.py')
+
+desiredTolerance = 0.001
+
+numFailures = 0
+
+# Species 1
+numFailures += shouldBe("FSABFlow[0,0;;;]", -3.829380699567663E-003, desiredTolerance)
+numFailures += shouldBe("particleFlux_vm_psiHat[0,0;;;]", -3.046431720535173E-007, desiredTolerance)
+numFailures += shouldBe("heatFlux_vm_psiHat[0,0;;;]", -1.462349694439989E-006, desiredTolerance)
+
+# Species 2
+numFailures += shouldBe("FSABFlow[1,0;;;]", -3.594147540383717E-004, desiredTolerance)
+numFailures += shouldBe("particleFlux_vm_psiHat[1,0;;;]", -6.257597194755813E-009, desiredTolerance)
+numFailures += shouldBe("heatFlux_vm_psiHat[1,0;;;]", -3.411992535031735E-008, desiredTolerance)
+
+exit(numFailures > 0)

### 2015-07-28 doc/manual/runs.tex
Bugfix in radialScans

@@
 The magnetic drift terms introduce nonzeros in the system matrix,
-and therefore increase the memory required for factorization; a typical increase is 30-40\%.
-These magnetic drift terms often have only a minor effect, except perhaps when the radial
+and therefore increase the memory required for factorization; a typical increase is 20-40\%.
+These magnetic drift terms typically have a minor effect except when the radial
 electric field is near 0.
@@
 are included, certain terms exist in the kinetic equation which have not yet been implemented in the code.
-Thus, at present it is not strictly correct to simultaneously set  \parlink{magneticDriftScheme} > 0 
+Thus, at present it is not strictly correct to simultaneously set  \parlink{magneticDriftScheme} $> 0$ 
 and  \parlink{includePhi1} = \true.

### 2015-08-02 doc/manual/installation.tex
Added profiles examples
Updated make test to remove distinction between small and large examples

@@
 
-Next, to compile, go to the directory {\ttfamily sfincs/fortran/version3/} and run {\ttfamily make}.
+Next, to compile, go to the directory {\ttfamily sfincs/fortran/version3/} and run {\ttfamily make -j}.
+(The {\ttfamily -j} flag means compiliation will be done in parallel, i.e. faster.)
 
@@
 will be checked, using the
-{\ttfamily tests\_small.py} or {\ttfamily tests\_large.py} script in the example's directory.
+{\ttfamily tests.py} script in the example's directory.
 
 If you run {\ttfamily make retest} from the \path{sfincs/fortran/version3/} directory,
-no new runs of \sfincs~will be performed, but the {\ttfamily tests\_small.py} or {\ttfamily tests\_large.py} script
+no new runs of \sfincs~will be performed, but the {\ttfamily tests.py} script
 will be run on any existing {\ttfamily sfincsOutput.h5} files in the \path{sfincs/fortran/version3/examples/} directories.
 
-The {\ttfamily make test} functionality relies on system-specific commands in the
-\path{sfincs/fortran/version3/makefiles/makefile.SFINCS_SYSTEM} file.  If you experience
-problems with  {\ttfamily make test}, it is likely that this file needs modification.
+The {\ttfamily make test} functionality relies on several environment variables set in the
+\path{sfincs/fortran/version3/makefiles/makefile.SFINCS_SYSTEM} file, as well as on the
+\path{job.SFINCS_SYSTEM} files in each example subdirectory.  If you experience
+problems with  {\ttfamily make test}, there is a good chance that one of these files needs modification.

### 2015-08-03 doc/manual/inputParameters.tex
Additions to user manual

@@
 \section{The {\ttfamily otherNumericalParameters}~namelist}
+\label{sec:otherNumericalParameters}
 
@@
 \PETScParam{-ksp\_gmres\_restart {\normalfont \ttfamily$<$integer$>$}}
-{After how many iterations will GMRES restart. Default is 2000. The convergence of GMRES slows every time a restart occurs, but restarts also free up memory.}
+{After how many iterations will GMRES restart. Default is 2000. The convergence of GMRES slows every time a restart occurs, but restarts also free up memory.
+The memory required by GMRES is typically quite small compared to the memory required for the $LU$ factorization.}
+
+\myhrule
+
+\PETScParam{-pc\_factor\_mat\_solver\_package {\normalfont \ttfamily$<$packagename$>$}}
+{Which sparse direct solver package is used to $LU$-factorize the preconditioner matrix.
+%Options are given by the items in quotation marks  
+%{\href{http://www.mcs.anl.gov/petsc/petsc-current/docs/manualpages/Mat/MatSolverPackage.html#MatSolverPackage}{here.}}
+This command-line flag overrides the related namelist parameter\\
+\parlink{whichParallelSolverToFactorPreconditioner}.
+See section \ref{sec:solvers} for further information about the available packages.
+}
+
+\section{\mumps~commands}
+\label{sec:mumpsControlParameters}
+
+The \mumps~solver package has many control parameters which are documented in the manual,
+available \href{http://mumps-solver.org/}{here}.
+In \sfincs, as in any \PETSc~application, these control parameters can be set
+using the command-line flags {\ttfamily -mat\_mumps\_cntl\_X YYYY} (for floating point parameters)
+and {\ttfamily -mat\_mumps\_icntl\_X YYYY} (for integer control parameters). 
+Here, {\ttfamily X} is the numeric index of the control parameter,
+and {\ttfamily YYYY} is the desired setting.
+Here we list some of the more useful options.\\
 
@@
 Set this parameter to 0 to suppress this output from mumps.}
+
+\myhrule
+
+\PETScParam{-mat\_mumps\_icntl\_14 {\normalfont \ttfamily$<$integer$>$}}
+{Percentage margin allowed for increase of certain arrays during the $LU$ factorization.
+The default value set by \sfincs~is 50 (higher than the original default value in \mumps.)
+If \sfincs~exits with the \mumps~error {\ttfamily INFO(1)=-9}, then further increasing this parameter may help.}
+
+\myhrule
+
+\PETScParam{-mat\_mumps\_icntl\_22 1}
+{Turns on the out-of-core solve capability,
+which reduces the memory required at the cost of speed.
+See section \ref{sec:parallelization} for further details.}

### 2015-08-03 doc/manual/resolution.tex
Additions to user manual

@@
 
-\section{Typical resolution requirements}
+\section{Examples of resolution requirements}
+
+To estimate the appropriate resolution parameters for various circumstances, you can look at
+the examples in the \path{sfincs/fortran/version3/examples/} directory. 
+Some other examples of appropriate resolution parameters are given in the following sections.
+
+\subsection{W7-X with anticipated experimental density and temperature}
+\label{sec:w7xresolution}
+
+The following parameters have been extensively tested with W7-X geometry for densities near $10^{20}$ m$^{-3}$
+and temperatures up to 6 keV, and found to give convergence to $\sim 3$\% (as shown in figure \ref{fig:convergence2}):\\
+{\ttfamily
+\Ntheta = 19 \\
+\Nzeta = 125 \\
+\Nxi = 140 \\
+\Nx = 5 \\
+}
+Note that at lower temperature and/or higher density, the collisionality will be higher,
+so comparable convergence could be achieved at lower \parlink{Nzeta} and \parlink{Nxi}.
+Conversely, at higher temperatures and/or lower desntiy, the collisionality will be lower,
+so comparable convergence would likely require higher  \parlink{Nzeta} and \parlink{Nxi}.
+
+\subsection{Figure 3 of the original SFINCS paper}
+
+For figure 3 in Ref \cite{sfincsPaper}, corresponding to a pure plasma in W7-X with $n = 6.6\times 10^{19}$ m$^3$ and $T_i = T_e = 1$ keV,
+the calculations used\\
+{\ttfamily
+\Ntheta = 19 \\
+\Nzeta = 59 \\
+\Nxi = 60 \\
+\Nx = 5. \\
+}
+Note the lower values of  \parlink{Nzeta} and \parlink{Nxi} used compared to section \ref{sec:w7xresolution},
+which were sufficient since the temperatures were lower.
+
+\subsection{Figure 4 of the original SFINCS paper}
+
+For figure 4 in Ref \cite{sfincsPaper}, corresponding to a collisionality scan in LHD,
+the following resolution parameters were used:
+
+\centerline{
+\begin{tabular}{c|c|c|c|c}
+\parlink{nuPrime} & \Ntheta & \Nzeta & \Nxi & \Nx \\
+\hline
+0.001 & 85 & 41 & 103 & 5 \\
+0.01 & 21 & 25 & 70 & 5 \\
+0.1 & 15 & 13 & 37 & 5 \\
+0.3 & 15 & 13 & 34 & 6 \\
+1 & 13 & 13 & 13 & 8 \\
+10 & 15 & 13 & 13 & 8 \\
+100 & 15 & 13 & 13 & 8
+\end{tabular}
+}
+
+\subsection{Figure 5 of the original SFINCS paper}
+
+For figure 5 in Ref \cite{sfincsPaper}, corresponding to a collisionality scan in W7-X,
+the following resolution parameters were used:
+
+\centerline{
+\begin{tabular}{c|c|c|c|c}
+\parlink{nuPrime} & \Ntheta & \Nzeta & \Nxi & \Nx \\
+\hline
+0.001 & 29 & 83 & 180 & 5 \\
+0.01  & 11 & 64 & 100 & 5 \\
+0.1   & 11 & 37 & 37  & 5 \\
+0.3   & 11 & 29 & 30  & 5 \\
+1     & 13 & 31 & 24  & 6 \\
+10    & 13 & 35 & 12  & 7 \\
+100   & 11 & 37 & 13  & 8
+\end{tabular}
+}

### 2015-08-26 doc/manual/runs.tex

@@
 The magnetic drift terms introduce nonzeros in the system matrix,
-and therefore increase the memory required for factorization; a typical increase is 20-40\%.
-These magnetic drift terms typically have a minor effect except when the radial
+and therefore increase the memory and time required for factorization.
+The change is small; a typical increase in both memory and time is 20-40\%.
+These magnetic drift terms typically have a minor effect on the physics outputs except when the radial
 electric field is near 0.
@@
 
-\section{Parallelization}
-[Coming soon...] Choosing the number of nodes and procs.
+\section{Sparse direct solver packages}
+\label{sec:solvers}
 
-\section{Issues with running on 1 processor}
+As discussed briefly in section \ref{sec:gmres}, the most computationally demanding step in 
+\sfincs~is the direct $LU$-factorization of a very large sparse nonsymmetric real matrix. The \PETSc~library
+which \sfincs~uses has interfaces to a large number of other packages for direct factorization of such matrices,
+making it possible to choose among the various solver packages with just a command-line flag ({\ttfamily -pc\_factor\_mat\_solver\_package}).  Some lists of the direct solvers
+available in \PETSc~can be found \href{http://www.mcs.anl.gov/petsc/petsc-current/docs/manualpages/Mat/MatSolverPackage.html#MatSolverPackage}{here}
+or in the ``direct solvers''-``LU'' section of \href{http://www.mcs.anl.gov/petsc/documentation/linearsolvertable.html}{this page}.
+It is important for the $LU$-solver package to be one that is efficiently parallelized, 
+in order to be able to solve problems at the high resolutions required for experimentally relevant collisionality and magnetic geometry.
+The recommended choice of $LU$ solver is \href{http://mumps-solver.org/}{\mumps} (which is the default), and another good option is \href{http://crd-legacy.lbl.gov/~xiaoye/SuperLU/}{\superludist}.
+(The {\ttfamily PARDISO} library available in the Intel Math Kernel Library is probably suitable as well, though we have not investigated it yet.)
+In side-to-side comparisons, we find \mumps~systematically uses substantially less memory and time than \superludist~for factorization.
+In principle, other solver packages for asymmetric matrices that are interfaced to \PETSc~could be used as well, such as {\ttfamily UMFPACK}, {\ttfamily PASTIX}, etc.
+
+There are two ways to choose between solver packages.
+One method is the \sfincs~parameter \parlink{whichParallelSolverToFactorPreconditioner} in the 
+{\ttfamily \hyperref[sec:otherNumericalParameters]{otherNumericalParameters}} namelist.
+This parameter only allows you to choose between \mumps~ and \superludist, not other solver packages interfaced to \PETSc.
+Another way to choose between solver packages is the command-line flag {\ttfamily -pc\_factor\_mat\_solver\_package},
+followed by one of the options in quotation marks \href{http://www.mcs.anl.gov/petsc/petsc-current/docs/manualpages/Mat/MatSolverPackage.html#MatSolverPackage}{here}.
+The command-line flag overrides the namelist parameter.
+
+The physics outputs of the code should be independent of the solver package used to several significant digits.
+In principle, different solver packages solving the same linear system should find the identical solution.
+However there will be small differences in the solutions associated with roundoff error.
+
+Note that {\ttfamily superlu} and \superludist~are distinct libraries. The former is serial while the latter is parallelized.
+Therefore there is no reason to use {\ttfamily superlu}; \superludist~is always preferable.
+
+The \mumps~package has a large number of control parameters, which are documented in the \mumps~manual which can be downloaded \href{http://mumps-solver.org/}{here}.
+You do not need to be aware of most of these control parameters.
+However, several parameters which may be useful are discussed in section \ref{sec:mumpsControlParameters}.
+It is also worth being aware of the section of the \mumps~manual on error messages. 
+This section is useful for interpreting the {\ttfamily INFO(1)} and {\ttfamily INFO(2)} error codes that
+are reported if \sfincs~exits with an error associated with \mumps.
+
+The \superludist~package has many fewer options than \mumps. 
+You can find of list of the options by running \sfincs~with the {\ttfamily -help} command-line flag when
+using \superludist, and searching the output for lines containing {\ttfamily superlu\_dist}.
+The \superludist~options are also documented in the package's manual, available \href{http://crd-legacy.lbl.gov/~xiaoye/SuperLU/#superlu_dist}{here}.
+We have not found any advantage in adjusting
+any of the \superludist~options.
+
+The \PETSc~library includes a built-in sparse direct solver which works on only a single processor.
+You can select this solver using the command-line flag \\
+\centerline{\ttfamily -pc\_factor\_mat\_solver\_package petsc}\\
+This solver could potentially be useful if you are running on a system that does not have \mumps~or \superludist~installed,
+and you are only considering problems that require sufficiently little memory (e.g. tokamaks) that parallelization is not required.
+However, this solver is less robust than \mumps~or \superludist, sometimes exiting with an error message that there is a zero pivot
+even though  \mumps~and \superludist~can solve the same system with no problem. Therefore, even if you plan to use only a single processor,
+we still recommend that you install  \mumps~or \superludist.
+
+\section{Parallelization: Choosing the number of nodes \& processors}
+\label{sec:parallelization}
+
+Usually the limiting factor for \sfincs~is not time but memory.
+(The time required depends on the resolution used, but jobs for experimentally relevant W7-X parameters
+typically take under 10 minutes.)
+Therefore, when considering how many nodes to request for a \sfincs~job,
+the first issue to consider is ensuring you have requested sufficient total memory.
+A good way to determine the memory required (assuming you are using the default solver \mumps)
+is to first run \sfincs~on 1 node using the parameters of interest
+and look for the following line in standard output:\\
+\centerline{\ttfamily ** TOTAL     space in MBYTES for IC factorization         :       1072}\\
+The number at the end will generally be different; it depends not only on the
+resolution and number of species used, but also increases with the number of processors
+as discussed below.  Make sure the number of nodes requested times the number of megabytes per node exceeds this number.
+This line in standard output is generated by \mumps, so if you are using a different solver, this information is not printed,
+and you may need to determine the number of nodes by trial-and-error.
+Since some memory on each node is used by the operating system and by \sfincs~functions other than the solver,
+you may need to use a slightly higher number of nodes than this estimate suggests.
+For experimentally relevant W7-X and HSX parameters, we typically use 2-6 nodes with 64 GB each.
+Problems with lower resolution requirements, such as tokamaks, often can be run on a single node.
+
+The `IC' in the above line stands for `in core', meaning the $L$ and $U$ factors are stored
+in memory rather than on disk.  In \mumps, one can also choose to do an `out of core' (OOC) solve,
+in which case substantially less memory is typically required.  The price you pay for this memory savings is time, since
+disk access is slow compared to memory access.  Due to this slowdown, we have not used the OOC capability much, but
+you might find it useful in some circumstances.
+To see how much memory would be required for an OOC solve,
+look for the following line in standard output:\\
+\centerline{\ttfamily ** TOTAL     space in MBYTES for OOC factorization        :       128}\\
+(The number at the end will generally be different.)
+To invoke out-of-core mode, you must take two steps. First, use the following command-line flag when calling \sfincs:\\
+\centerline{\ttfamily -mat\_mumps\_icntl\_22 1}\\
+Second, you must set the environment variable {\ttfamily MUMPS\_OOC\_TMPDIR} to some reasonable directory
+before calling \sfincs. This environment variable could be set for example in the batch job file.
+Temporary files containing the $L$ and $U$ factors will be stored in the directory indicated.
+
+Another issue to consider is how many processors to use.  It is not always best to use the maximum number of processes
+available on the number of nodes you have chosen. The reason is that as the preconditioner matrix is divided among
+more and more processes, the $LU$ factorization becomes less efficient, requiring more memory and more communication.
+If you examine the \mumps~IC and OOC memory requirements indicated above, you will find they increase somewhat 
+as the number of processes increase.
+One needs to find a balance between speed (favoring many processes) and memory requirements (favoring few processes).
+While it is almost always better overall to use 2 processes compared to 1 process, it is not always better to use 128 processes
+compared to 64 processes.  The sweet spot is often in the range of 16-64 processes.
+To determine how to request fewer processes than the maximum available on a given number of nodes,
+see the documentation for your computing system.
+
+If more memory is required than is available, the system will usually terminate your job with an out-of-memory (OOM) error.
+When this occurs, you need to either increase the number of nodes requested or decrease the number of processors
+requested.
+
+Most of the time, \sfincs~is run via \sfincsScan~as some parameter is scanned, such as the radial electric field.
+In this case, the scan is ``embarassingly parallel'' in the sense that each job in the scan is completely independent
+of the other jobs.  Even if each individual job requires only 1 or a small number of nodes, it is still useful
+to run \sfincs~on a computing system with many nodes so the scan can be carried out in parallel.
+
+%\section{Issues with running on 1 processor}
 

### 2015-09-23 doc/manual/inputParameters.tex
Implemented includeRadialExBDrive

@@
 
+\param{includeRadialExBDrive}
+{Boolean}
+{\false}
+{Whenever \parlink{RHSMode} == 1.}
+{If true, the radial $\vect{E}\times\vect{B}$ term $(\vect{v}_{E} \cdot\nabla\psi)f_{Ms} [(1/n_s)(dn_s/d\psi) + (x_s^2-3/2)(1/T_s)(dT_s/d\psi)]$
+will be included in the kinetic equation. This is one of the terms considered in Ref \cite{Regana2013} which
+should be unimportant for the main ions but which may be important for impurities.
+Use of this option requires \parlink{includePhi1}=\true, since the radial $\vect{E}\times\vect{B}$ drift
+arises due to $\Phi_1$.
+}
+
+\myhrule
+
 \param{nonlinear}
@@
 {Whenever \parlink{RHSMode} == 1.}
-{If true, certain terms that are nonlinear in the unknowns will be included in the kinetic equation.
+{If true, the term
+$-(Z_s e/m) (\nabla_{||} \Phi_1) (\partial f_{s1}/\partial v_{||})_\mu$ will be included
+in the kinetic equation.  This is one of the terms considered in Ref \cite{Regana2013} which
+should be unimportant for the main ions but which may be important for impurities.
+This term is nonlinear in the unknowns $f_{s1}$ and $\Phi_1$.
 Newton's method will be used to solve the nonlinear system, meaning that the usual linear solve in \sfincs~must

### 2015-09-23 fortran/version3/populateMatrix.F90

@@ -1035,6 +1035,59 @@
        end if
 
        ! *********************************************************
+       ! Add the radial ExB drive term:
+       ! (v_E dot grad psi) f_M [(1/n)(dn/dpsi) + (x^2 - 3/2)(1/T)(dT/dpsi)]
+       ! This is the new linear term considered by Garcia-Regana et al PPCF (2013).
+       ! *********************************************************
+
+       if ((whichMatrix .ne. 2) .and. includeRadialExBDrive) then
+          L=0
+          do ix = ixMin,Nx
+
+             ! d Phi_1 / d theta term:
+             itheta = -1 ! Just so a runtime error occurs if I use itheta by mistake
+             do ithetaRow = ithetaMin, ithetaMax
+                do izeta = izetaMin, izetaMax
+                   rowIndex = getIndex(ispecies, ix, L+1, ithetaRow, izeta, BLOCK_F)
+
+                   factor = -alpha*Delta*DHat(ithetaRow,izeta)*BHat_sub_zeta(ithetaRow,izeta) &
+                        /(two*BHat(ithetaRow,izeta)*BHat(ithetaRow,izeta)) &
+                        * nHat * (mHat*sqrtMHat)/(THat*sqrtTHat*pi*sqrtpi) * expx2(ix) & ! This line is f_M
+                        * (dNHatdpsiHats(ispecies)/nHat + (x2(ix)-3/two)*dTHatdpsiHats(ispecies)/THat)
+                   
+                   do ithetaCol = 1,Ntheta
+                      colIndex = getIndex(1,1,1,ithetaCol, izeta, BLOCK_QN)
+                      call MatSetValueSparse(matrix, rowIndex, colIndex,&
+                           factor*ddtheta(ithetaRow, ithetaCol), ADD_VALUES, ierr)
+                   end do
+                end do
+             end do
+
+             ! d Phi_1 / d zeta term:
+             izeta = -1 ! Just so a runtime error occurs if I use itheta by mistake
+             ithetaRow = -1 ! Just so a runtime error occurs if I use itheta by mistake
+             ithetaCol = -1 ! Just so a runtime error occurs if I use itheta by mistake
+             do itheta = ithetaMin, ithetaMax
+                do izetaRow = izetaMin, izetaMax
+                   rowIndex = getIndex(ispecies, ix, L+1, itheta, izetaRow, BLOCK_F)
+
+                   factor = alpha*Delta*DHat(ithetaRow,izeta)*BHat_sub_theta(ithetaRow,izeta) &
[…]

### 2015-09-23 fortran/version3/validateInput.F90

@@ -289,6 +289,15 @@ subroutine validateInput()
      stop
   end if
 
+  if (includeRadialExBDrive .and. (.not. includePhi1)) then
+     if (masterProc) then
+        print *,"Error! You requested a calculation including the radial ExB drive term"
+        print *,"(includeRadialExBDrive=.true.) but you set includePhi1=.false."
+        print *,"These options are inconsistent since the radial ExB drive term involves Phi1."
+     end if
+     stop
+  end if
+
   if (magneticDriftScheme<0) then
      if (masterProc) then
         print *,"Error! magneticDriftScheme must be >= 0."

### 2015-10-27 doc/manual/inputParameters.tex

@@
 in speed proportional to the radial electric field.  This term is given by $\dot{x}$ in
-equation (17) of \cite{sfincsPaper}.}
-
+equation (17) of \cite{sfincsPaper}:}
+\begin{equation}
+-(\vect{v}_{ma} \cdot\nabla r) \frac{Z_s e}{2 T_s x_s} \frac{d\Phi_0}{dr} \frac{\partial f_{a1}}{\partial x_s}
+\end{equation}
 \myhrule
@@
 in pitch angle $\xi$ proportional to the radial electric field.  This term is given by the last line of
-equation (17) of \cite{sfincsPaper}.}
+equation (17) of \cite{sfincsPaper}:}
+\begin{equation}
+\frac{(1-\xi^2)\xi}{2B^3} \frac{d\Phi_0}{dr}(\vect{B}\times\nabla r\cdot\nabla B)
+ \frac{\partial f_{s1}}{\partial \xi}
+\end{equation}
 
@@
 
-\param{PESCPreallocationStrategy}
+\param{PETSCPreallocationStrategy}
 {integer}
@@
 
-{\ttfamily PESCPreallocationStrategy} = 0: Old method with high estimated nnz. This method involves relatively simpler code but uses WAY more memory than necessary.\\
-
-{\ttfamily PESCPreallocationStrategy} = 1: New method with lower, more precise estimated nnz.  This method should use much less memory.\\
+{\ttfamily PETSCPreallocationStrategy} = 0: Old method with high estimated nnz. This method involves relatively simpler code but uses WAY more memory than necessary.\\
+
+{\ttfamily PETSCPreallocationStrategy} = 1: New method with lower, more precise estimated nnz.  This method should use much less memory.\\
 

### 2015-12-17 doc/manual/inputParameters.tex

@@
 
+\param{preconditioner\_magnetic\_drifts\_max\_L}
+{integer}
+{2}
+{Whenever \parlink{useIterativeLinearSolver} = \true and \parlink{magneticDriftScheme}$>0$}
+{Maximum Legendre mode number for which the poloidal and toroidal magnetic drift terms are included in the preconditioner.
+As this parameter is increased, more memory is required for factorization, but fewer KSP iterations are required.
+Setting this parameter to $\ge$ \parlink{Nxi} means that the magnetic drifts are always included in the preconditioner.
+Setting this parameter to $<0$ means the magnetic drifts are never included in the preconditioner.
+The default value should be good.}
+
+\myhrule
+
 \param{reusePreconditioner}

### 2016-01-07 doc/manual/inputParameters.tex
Added constraintScheme 3 and 4 to try out alternate sources.

@@
 
-{\ttfamily constraintScheme} = 1: 2 constraints per species: $\left<n_1\right>=0$ and $\left<p_1\right>=0$.\\
+{\ttfamily constraintScheme} = 1: 2 constraints per species: $\left<n_1\right>=0$ and $\left<p_1\right>=0$.
+The particle and heat sources have the form $S = (a_2 x^2 + a_0) e^{-x^2}$.
+The $a_2$ and $a_0$ coefficients are determined so that one source term provides particles but not energy,
+whereas the other source term provides energy but not particles. \\
 
 {\ttfamily constraintScheme} = 2: \Nx~constraints per species: $\left< f(L=0)\right>=0$ at each $x$.\\
+
+{\ttfamily constraintScheme} = 3: Same as {\ttfamily constraintScheme} = 1,
+except the particle and heat sources have the form $S = (a_4 x^4 + a_0) e^{-x^2}$.\\
+
+{\ttfamily constraintScheme} = 4: Same as {\ttfamily constraintScheme} = 1,
+except the particle and heat sources have the form $S = (a_4 x^4 + a_2 x^2) e^{-x^2}$.\\
 

### 2016-01-07 fortran/version3/createGrids.F90

@@ -1237,7 +1237,8 @@
 
     select case (constraintScheme)
     case (0)
-    case (1)
+       ! No allocation needed in this case.
+    case (1,3,4)
        allocate(sources(Nspecies,2))
     case (2)
        allocate(sources(Nspecies,Nx))

### 2016-01-07 fortran/version3/preallocateMatrix.F90

@@ -82,7 +82,7 @@ subroutine preallocateMatrix(matrix, whichMatrix)
   
   select case (constraintScheme)
   case (0)
-  case (1)
+  case (1, 3, 4)
      ! The rows for the constraints have more nonzeros:
      !predictedNNZsForEachRow((Nspecies*Nx*Ntheta*Nzeta*Nxi+1):matrixSize) = Ntheta*Nzeta*Nx + 1
      !predictedNNZsForEachRow((Nspecies*Nx*Ntheta*Nzeta*Nxi+1):matrixSize) = Ntheta*Nzeta*Nx

### 2016-01-07 fortran/version3/writeHDF5Output.F90

@@ -802,18 +802,21 @@ contains
     maxDimForIterationSpeciesSources(2)   = Nspecies
     dimForIterationSpeciesSourcesChunk(2) = Nspecies
 
-    if (constraintScheme==1) then
+    select case (constraintScheme)
+    case (0)
+       ! No action needed since we will not write sources to the file
+       ! when constraintScheme==0
+    case (1,3,4)
        dimForIterationSpeciesSources(3)      = 2
        maxDimForIterationSpeciesSources(3)   = 2
        dimForIterationSpeciesSourcesChunk(3) = 2
-    else
-       ! This block of code will also be run when constraintScheme==0,
-       ! but it does not matter since we will not write sources to the file
-       ! when constraintScheme==0
+    case (2)
        dimForIterationSpeciesSources(3)      = Nx
        maxDimForIterationSpeciesSources(3)   = Nx
        dimForIterationSpeciesSourcesChunk(3) = Nx
-    end if
+    case default
+       stop "Invalid constraintScheme!"
+    end select
 
     call h5screate_simple_f(rank, dimForIterationSpeciesSources, dspaceIDForIterationSpeciesSources, &
          HDF5Error, maxDimForIterationSpeciesSources)

### 2016-02-06 doc/manual/inputParameters.tex
You can now directly specify Er instead of dPhiHatdXXX. To do so set inputRadialCoordinateForGradients=4.

@@
 {integer}
-{2}
+{4}
 {Whenever \parlink{RHSMode}==1.}
@@
 
-No matter which option you pick, the gradients with respect to all 4 radial coordinates will be saved in the output \HDF~file.
+{\ttfamily inputRadialCoordinateForGradients}==4: 
+Same as {\ttfamily inputRadialCoordinateForGradients}==2, except \parlink{Er} is used instead of \parlink{dPhiHatdrHat}.
+Thus, density gradients are specified by \parlink{dnHatdrHats},
+temperature gradients are specified by \parlink{dTHatdrHats}, a single $E_r$ is specified by \parlink{Er},
+and the range of an $E_r$ scan is specified by \parlink{ErMin}-\parlink{ErMax}.\\
+
+No matter which option you pick, the gradients with respect to all radial coordinates will be saved in the output \HDF~file.
 }
@@
 i.e. the radial electric field up to a constant.
-Notice that exactly 1 of the 4 variables \parlink{dPhiHatdpsiHat}, \parlink{dPhiHatdpsiN}, \parlink{dPhiHatdrHat}, and \parlink{dPhiHatdrN}
+Notice that exactly 1 of the 5 variables \parlink{dPhiHatdpsiHat}, \parlink{dPhiHatdpsiN}, \parlink{dPhiHatdrHat}, \parlink{dPhiHatdrN}, or \parlink{Er}
 will be used, depending on\\
@@
 i.e. the radial electric field up to a constant.
-Notice that exactly 1 of the 4 variables \parlink{dPhiHatdpsiHat}, \parlink{dPhiHatdpsiN}, \parlink{dPhiHatdrHat}, and \parlink{dPhiHatdrN}
+Notice that exactly 1 of the 5 variables \parlink{dPhiHatdpsiHat}, \parlink{dPhiHatdpsiN}, \parlink{dPhiHatdrHat}, \parlink{dPhiHatdrN}, or \parlink{Er}
+will be used, depending on\\
+\parlink{inputRadialCoordinateForGradients}.
+}
+
+\myhrule
+
+\param{Er}
+{real}
+{0.0}
+{Whenever \parlink{RHSMode} == 1 and \parlink{inputRadialCoordinateForGradients} == 4}
+{The derivative of the electrostatic potential with respect to the radial coordinate $\hat{r}$, multiplied by $-1$,
+i.e. the radial electric field.
+Notice that exactly 1 of the 5 variables \parlink{dPhiHatdpsiHat}, \parlink{dPhiHatdpsiN}, \parlink{dPhiHatdrHat}, \parlink{dPhiHatdrN}, or \parlink{Er}
 will be used, depending on\\
@@
 will always be uniformly (linearly) spaced.  
-Notice that exactly 1 of the 4 variables \parlink{dPhiHatdpsiHatMax}, \parlink{dPhiHatdpsiNMax}, \parlink{dPhiHatdrHatMax}, and \parlink{dPhiHatdrNMax}
+Notice that exactly 1 of the 5 variables \parlink{dPhiHatdpsiHatMax}, \parlink{dPhiHatdpsiNMax}, \parlink{dPhiHatdrHatMax}, \parlink{dPhiHatdrNMax}, or \parlink{ErMax}
 will be used, depending on\\
 \parlink{inputRadialCoordinateForGradients}.
-Similarly, exactly 1 of the 4 variables
-\parlink{dPhiHatdpsiHatMin}, \parlink{dPhiHatdpsiNMin}, \parlink{dPhiHatdrHatMin}, and \parlink{dPhiHatdrNMin}
+Similarly, exactly 1 of the 5 variables
+\parlink{dPhiHatdpsiHatMin}, \parlink{dPhiHatdpsiNMin}, \parlink{dPhiHatdrHatMin}, \parlink{dPhiHatdrNMin}, or \parlink{ErMin}
 will be used.
@@
 {Minimum value of \parlink{dPhiHatdrN} to use in the scan.}
+
+\myhrule
+
+\ssparam{ErMax}
+{real}
+{Only when \parlink{inputRadialCoordinateForGradients} = 4 and \sfincsScan~is run with \parlink{scanType} = 2.}
+{Maximum value of \parlink{Er} to use in the scan.}
+
+\myhrule
+
+\ssparam{ErMin}
+{real}
+{Only when \parlink{inputRadialCoordinateForGradients} = 4 and \sfincsScan~is run with \parlink{scanType} = 2.}
+{Minimum value of \parlink{Er} to use in the scan.}
 
@@
 See section \ref{sec:parallelization} for further details.}
+
+\myhrule
+
+\PETScParam{-mat\_mumps\_icntl\_28 2}
+{Uses one of the parallelized libraries ParMETIS or PT-SCOTCH for analyzing the matrix,
+instead of the default serial algorithm.}

### 2016-02-06 doc/manual/runs.tex
You can now directly specify Er instead of dPhiHatdXXX. To do so set inputRadialCoordinateForGradients=4.

@@
 
+4: Same as option 2, except the radial electric field is specified using \parlink{Er}. Thus, derivatives with respect to {\ttfamily rHat} will be used:
+Density gradients are specified using \parlink{dnHatdrHats}, 
+temperature gradients are specified using \parlink{dTHatdrHats},  a single $E_r$ is specified using
+\parlink{Er}, and a range of $E_r$ for a scan is specified using \parlink{ErMin}-\parlink{ErMax}.
+\\
 }
+
+The most common choice is the default, 4.
 
@@
 Note however that the maximum and minimum radial electric field specified in the {\ttfamily profiles}
-file must be defined as the derivative of the electrostatic potential with respect to the radial coordinate \parlink{inputRadialCoordinateForGradients}.
+file must be given in terms of the electric field variable selected by \parlink{inputRadialCoordinateForGradients}.
 
@@
 \section{Trajectory models}
+\label{sec:trajectoryModels}
 
@@
 \section{Quasineutrality and variation of the electrostatic potential on the flux surface}
+\label{sec:qn}
 
@@
 \section{Poloidal and toroidal magnetic drifts}
+\label{sec:magneticDrifts}
 

### 2016-02-06 doc/manual/inputParameters.tex
Examples updated to use Er instead of dPhiHatdXXX

@@
 {Whenever \parlink{RHSMode} == 1 and \parlink{inputRadialCoordinateForGradients} == 4}
-{The derivative of the electrostatic potential with respect to the radial coordinate $\hat{r}$, multiplied by $-1$,
-i.e. the radial electric field.
+{The derivative of the normalized electrostatic potential $\hat{\Phi}$ with respect to the radial coordinate $\hat{r}$, multiplied by $-1$,
+i.e. {\ttfamily Er}$=-$\parlink{dPhiHatdrHat}.
 Notice that exactly 1 of the 5 variables \parlink{dPhiHatdpsiHat}, \parlink{dPhiHatdpsiN}, \parlink{dPhiHatdrHat}, \parlink{dPhiHatdrN}, or \parlink{Er}
@@
 
-\param{ExBDerivativeScheme}
+\param{ExBDerivativeSchemeTheta}
 {integer}
@@
 such that the coefficient in front of the $\partial f/\partial \theta$ or $\partial f/\partial\zeta$ terms vanishes near the thermal speed of one of the species,
-the distribution function can develop unphysical grid-scale structure.  This option allows upwinding of these spatial derivatives, which eliminates the grid-scale oscillations.\\
-
-{\ttfamily ExBDerivativeScheme} = 0: Use same differentiation matrices for the $\vect{E}\times\vect{B}$ drift terms as for the parallel streaming term.\\
-
-{\ttfamily ExBDerivativeScheme} = 1: Treat the $\vect{E}\times\vect{B}$ drift terms using a 4 point upwinded stencil (1 point on 1 side, 2 points on the other.)\\
-
-{\ttfamily ExBDerivativeScheme} = 2: Treat the $\vect{E}\times\vect{B}$ drift terms using a 5 point upwinded stencil (1 point on 1 side, 3 points on the other.)\\
-
-{\ttfamily ExBDerivativeScheme} = 3: Treat the $\vect{E}\times\vect{B}$ drift terms using a 6 point upwinded stencil (2 points on 1 side, 3 points on the other.)\\
+the distribution function can develop unphysical grid-scale structure.  This option allows upwinding of the $\partial f/\partial\theta$
+spatial derivative, which eliminates the grid-scale oscillations.
+The parameter \parlink{ExBDerivativeSchemeZeta} serves the same function for the $\partial f/\partial\zeta$ derivative.
+\\
+
+{\ttfamily ExBDerivativeSchemeTheta} = 0: Use same differentiation matrices for the $\vect{E}\times\vect{B}$ drift terms as for the parallel streaming term.\\
+
+{\ttfamily ExBDerivativeSchemeTheta} = 1: Treat the $\vect{E}\times\vect{B}$ drift terms using a 4 point upwinded stencil (1 point on 1 side, 2 points on the other.)\\
+
+{\ttfamily ExBDerivativeSchemeTheta} = 2: Treat the $\vect{E}\times\vect{B}$ drift terms using a 5 point upwinded stencil (1 point on 1 side, 3 points on the other.)\\
+
+{\ttfamily ExBDerivativeSchemeTheta} = 3: Treat the $\vect{E}\times\vect{B}$ drift terms using a 6 point upwinded stencil (2 points on 1 side, 3 points on the other.)\\
 
@@
 Option 1 is substantially less accurate than the other options.
-You should use the default {\ttfamily ExBDerivativeScheme} = 0 except for very large radial electric fields, when the electric field is as large as the resonance or larger.
-In this case, {\ttfamily ExBDerivativeScheme} = 3 is recommended.
-}
+You should use the default {\ttfamily ExBDerivativeSchemeTheta} = 0 except for very large radial electric fields, when the electric field is as large as the resonance or larger.
+In this case, {\ttfamily ExBDerivativeSchemeTheta} = 3 is recommended.
+}
+
+\myhrule
+
+\param{ExBDerivativeSchemeZeta}
+{integer}
+{0}
+{Whenever the radial electric field is nonzero}
+{This parameter behaves exactly like \parlink{ExBDerivativeSchemeTheta}, except it applies to the $\zeta$ coordinate instead of $\theta$.}
 
@@
 {Options for controlling upwinding of the magnetic drift terms.  Note the options are different than for \parlink{thetaDerivativeScheme}
-and \parlink{zetaDerivativeScheme}, and are analogous to the options for \parlink{ExBDerivativeScheme}.
+and \parlink{zetaDerivativeScheme}, and are analogous to the options for \parlink{ExBDerivativeSchemeTheta}.
 This option exists because when magnetic drift terms involving $\partial f/\partial \theta$ and $\partial f/\partial\zeta$

### 2016-02-06 fortran/version3/validateInput.F90

@@ -628,31 +628,45 @@ subroutine validateInput()
      print *,line
   end if
 
-  if (ExBDerivativeScheme<0) then
+  if (ExBDerivativeSchemeTheta<0) then
      if (masterProc) then
-        print *,"Error! ExBDerivativeScheme cannot be less than 0."
+        print *,"Error! ExBDerivativeSchemeTheta cannot be less than 0."
      end if
      stop
   end if
   
-  if (ExBDerivativeScheme>3) then
+  if (ExBDerivativeSchemeTheta>3) then
      if (masterProc) then
-        print *,"Error! ExBDerivativeScheme cannot be more than 3."
+        print *,"Error! ExBDerivativeSchemeTheta cannot be more than 3."
      end if
      stop
   end if
   
-  if (ExBDerivativeScheme>0 .and. preconditioner_theta>0) then
+  if (ExBDerivativeSchemeZeta<0) then
      if (masterProc) then
-        print *,"Error! The implementation of ExBDerivativeScheme>0 does not presently allow"
+        print *,"Error! ExBDerivativeSchemeZeta cannot be less than 0."
+     end if
+     stop
+  end if
+  
+  if (ExBDerivativeSchemeZeta>3) then
+     if (masterProc) then
+        print *,"Error! ExBDerivativeSchemeZeta cannot be more than 3."
+     end if
+     stop
+  end if
+  
+  if (ExBDerivativeSchemeTheta>0 .and. preconditioner_theta>0) then
+     if (masterProc) then
+        print *,"Error! The implementation of ExBDerivativeSchemeTheta>0 does not presently allow"
         print *,"       preconditioning in theta (preconditioner_theta>0)."
      end if
      stop
   end if
   
-  if (ExBDerivativeScheme>0 .and. preconditioner_zeta>0) then
+  if (ExBDerivativeSchemeZeta>0 .and. preconditioner_zeta>0) then
      if (masterProc) then
-        print *,"Error! The implementation of ExBDerivativeScheme>0 does not presently allow"
+        print *,"Error! The implementation of ExBDerivativeSchemeZeta>0 does not presently allow"
         print *,"       preconditioning in zeta (preconditioner_zeta>0)."
      end if
      stop

### 2016-02-06 fortran/version3/writeHDF5Output.F90

@@ -182,7 +182,8 @@ contains
        call writeHDF5Field("geometryScheme", geometryScheme, "")
        call writeHDF5Field("thetaDerivativeScheme", thetaDerivativeScheme, "")
        call writeHDF5Field("zetaDerivativeScheme", zetaDerivativeScheme, "")
-       call writeHDF5Field("ExBDerivativeScheme", ExBDerivativeScheme, "")
+       call writeHDF5Field("ExBDerivativeSchemeTheta", ExBDerivativeSchemeTheta, "")
+       call writeHDF5Field("ExBDerivativeSchemeZeta", ExBDerivativeSchemeZeta, "")
        call writeHDF5Field("magneticDriftDerivativeScheme", magneticDriftDerivativeScheme, "")
        call writeHDF5Field("xGridScheme", xGridScheme, "")
        call writeHDF5Field("xGrid_k", xGrid_k, "Exponent of x in the orthogonality relation for the speed polynomials")

### 2016-07-20 doc/manual/version3/equations.tex
Beginning version 4

\chapter{Kinetic equations}

The \sfincs~code is capable of solving many different variants of the drift-kinetic equation.
In this section we summarize the most common drift-kinetic equations
which can be solved with the code, giving the associated input parameters
(which are all in the  {\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}} namelist.)
For other choices of these input parameters, 
other combinations of terms in the drift-kinetic equation can be used than those
given here.
For more information, see section \ref{sec:trajectoryModels}, section \ref{sec:qn}, section \ref{sec:magneticDrifts},
and the version 3 technical documentation.
The following terms are always included in \sfincs:
parallel streaming, the magnetic mirror force, the collision operator,
and the inhomogeneous drive term from radial gradients.
Other terms generally can be turned on and off as desired using parameters in the
{\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}} namelist.

Throughout this section, gradients and other partial derivatives are taken
at constant $x_a = v/\sqrt{2T_a/m_a}$ and at constant $\xi = v_{||} / v$, unless denoted otherwise with subscripts.
We use $s$ to denote the species, $r$ to denote any radial coordinate, (expressions are independent
of the specific radial coordinate used,)
$C_a$ to denote the collision operator, and $S_a$ to denote the source-sink term.
A more detailed discussion of the equations implemented in the code can be found
in the technical documentation for \sfincs~version 3, available in the {\ttfamily /doc} directory
of the repository.

\section{Default: Full $\vect{E}\times\vect{B}$ trajectories; no poloidal or toroidal magnetic drifts; flux function potential}

\begin{eqnarray}
&&\left(v_{||}\vect{b} + \frac{d\Phi_0}{dr} \frac{1}{B^2} \vect{B}\times\nabla r \right) \cdot \nabla f_{s1} \\
&&+ \left[ - \frac{(1-\xi^2)v}{2B} \nabla_{||} B
+\frac{(1-\xi^2)\xi}{2B^3} \frac{d\Phi_0}{dr}\vect{B}\times\nabla r\cdot\nabla B \right]
 \frac{\partial f_{s1}}{\partial \xi} \nonumber \\
&&-(\vect{v}_{ma} \cdot\nabla r) \frac{Z_s e}{2 T_s x_s} \frac{d\Phi_0}{dr} \frac{\partial f_{a1}}{\partial x_s} \nonumber \\
&&+ (\vect{v}_{ma} \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM}
 = C_a + S_a \nonumber
\end{eqnarray}
Note that this equation is equivalent to the following one, in which the independent variables
are $(\mu,v_{||})$ instead of $(\xi,x_s)$:
\begin{eqnarray}
&&\left(v_{||}\vect{b} + \frac{d\Phi_0}{dr} \frac{1}{B^2} \vect{B}\times\nabla r \right) \cdot (\nabla f_{s1})_{\mu, v_{||}} \\
&&+ \left[ - \mu \nabla_{||} B
-\frac{v_{||}}{B^2} \frac{d\Phi_0}{dr} \vect{b}\times\nabla B \times \nabla r \right]
\left( \frac{\partial f_{a1}}{\partial v_{||}} \right)_{\mu} \nonumber \\
&& + (\vect{v}_{ma} \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM}
=C_a + S_a \nonumber
\end{eqnarray}
These equivalent forms of the kinetic equation are selected using \\
\parlink{includeXDotTerm} = \true  \;\;\; (Default) \\
\parlink{includeElectricFieldTermInXiDot} = \true \;\;\; (Default) \\
\parlink{useDKESExBDrift} = \false \;\;\; (Default) \\
\parlink{magneticDriftScheme} = 0 \;\;\; (Default) \\
\parlink{includePhi1} = \false \;\;\; (Default) \\
\parlink{includeRadialExBDrive} = \false \;\;\; (Default) \\
\parlink{nonlinear} = \false \;\;\; (Default)

\section{DKES $\vect{E}\times\vect{B}$ trajectories; no poloidal or toroidal magnetic drifts; flux function potential}

This form of the kinetic equation is useful for benchmarking with DKES and other codes
that use the same equation.

\begin{eqnarray}
&&\left(v_{||}\vect{b} + \frac{d\Phi_0}{dr} \frac{1}{\left< B^2\right>} \vect{B}\times\nabla r \right) \cdot \nabla f_{s1} \\
&& - \frac{(1-\xi^2)v}{2B} (\nabla_{||} B)
 \frac{\partial f_{s1}}{\partial \xi} \nonumber \\
&&+ (\vect{v}_{ma} \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM}
 = C_a + S_a \nonumber
\end{eqnarray}
This form of the kinetic equation is selected using \\
\parlink{includeXDotTerm} = \false  \;\;\; (Not default) \\
\parlink{includeElectricFieldTermInXiDot} = \false \;\;\; (Not default) \\
\parlink{useDKESExBDrift} = \true \;\;\; (Not default) \\
\parlink{magneticDriftScheme} = 0 \;\;\; (Default) \\
\parlink{includePhi1} = \false \;\;\; (Default) \\
\parlink{includeRadialExBDrive} = \false \;\;\; (Default) \\
\parlink{nonlinear} = \false \;\;\; (Default)

\section{Full $\vect{E}\times\vect{B}$ trajectories; including poloidal and toroidal magnetic drifts; flux function potential}

\begin{eqnarray}
&&\left(v_{||}\vect{b} + \frac{d\Phi_0}{dr} \frac{1}{B^2} \vect{B}\times\nabla r  + \vect{v}_{ms}\right) \cdot \nabla \theta \frac{\partial f_{s1}}{\partial \theta} \\
&&\left(v_{||}\vect{b} + \frac{d\Phi_0}{dr} \frac{1}{B^2} \vect{B}\times\nabla r  + \vect{v}_{ms}\right) \cdot \nabla \zeta \frac{\partial f_{s1}}{\partial \zeta} \nonumber \\
&&+ \left\{ - \frac{(1-\xi^2)v}{2B} \nabla_{||} B
+\frac{(1-\xi^2)\xi}{2B^3} \frac{d\Phi_0}{dr}\vect{B}\times\nabla r\cdot\nabla B \right. \nonumber \\
&& \left. \hspace{0.25in} -\frac{T_s x_s^2 (\nabla r\cdot\nabla\theta\times\nabla\zeta)}{Z_s e B^3} (1-\xi^2)\xi
\left[ \left( \frac{\partial B_r}{\partial \zeta} - \frac{\partial B_\zeta}{\partial r}\right)\frac{\partial B}{\partial\theta}
+ \left(\frac{\partial B_\theta}{\partial r} - \frac{\partial B_r}{\partial\theta}\right) \frac{\partial B}{\partial\zeta}\right]
\right\}
 \frac{\partial f_{s1}}{\partial \xi} \nonumber \\
&&-(\vect{v}_{ma} \cdot\nabla r) \frac{Z_s e}{2 T_s x_s} \frac{d\Phi_0}{dr} \frac{\partial f_{a1}}{\partial x_s} \nonumber \\
&&+ (\vect{v}_{ma} \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM}
 = C_a + S_a \nonumber
\end{eqnarray}
Notice the magnetic drifts affect the coefficients of $\partial f_{s1}/\partial \theta$, $\partial f_{s1}/\partial \zeta$, 
and $\partial f_{s1}/\partial \xi$,
but there is no change to the coefficient of $\partial f_{s1}/\partial x_s$.
This form of the kinetic equation is selected using \\
\parlink{includeXDotTerm} = \true  \;\;\; (Default) \\
\parlink{includeElectricFieldTermInXiDot} = \true \;\;\; (Default) \\
\parlink{useDKESExBDrift} = \false \;\;\; (Default) \\
\parlink{magneticDriftScheme} = 1 \;\;\; (Not default) \\
\parlink{includePhi1} = \false \;\;\; (Default) \\
\parlink{includeRadialExBDrive} = \false \;\;\; (Default) \\
\parlink{nonlinear} = \false \;\;\; (Default)

\section{Full $\vect{E}\times\vect{B}$ trajectories; no poloidal or toroidal magnetic drifts; leading $\Phi_1$ term}

In this form of the drift-kinetic equation, we include the largest term involving $\Phi_1$,
associated with the adiabatic response.
However, other terms considered in Ref \cite{Regana2013} are not included.

\begin{eqnarray}
&&\left(v_{||}\vect{b} + \frac{d\Phi_0}{dr} \frac{1}{B^2} \vect{B}\times\nabla r \right) \cdot \nabla f_{s1} \\
&&+ \left[ - \frac{(1-\xi^2)v}{2B} \nabla_{||} B
+\frac{(1-\xi^2)\xi}{2B^3} \frac{d\Phi_0}{dr}\vect{B}\times\nabla r\cdot\nabla B \right]
 \frac{\partial f_{s1}}{\partial \xi} \nonumber \\
&&-(\vect{v}_{ma} \cdot\nabla r) \frac{Z_s e}{2 T_s x_s} \frac{d\Phi_0}{dr} \frac{\partial f_{a1}}{\partial x_s} \nonumber \\
&&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
&&+ (\vect{v}_{ma} \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM}
 = C_a + S_a \nonumber
\end{eqnarray}
Note that this equation is equivalent to the following one, in which the independent variables
are $(\mu,v_{||})$ instead of $(\xi,x_s)$:
\begin{eqnarray}
&&\left(v_{||}\vect{b} + \frac{d\Phi_0}{dr} \frac{1}{B^2} \vect{B}\times\nabla r \right) \cdot (\nabla f_{s1})_{\mu, v_{||}} \\
&&+ \left[ - \mu \nabla_{||} B
-\frac{v_{||}}{B^2} \frac{d\Phi_0}{dr} \vect{b}\times\nabla B \times \nabla r \right]
\left( \frac{\partial f_{a1}}{\partial v_{||}} \right)_{\mu} \nonumber \\
&&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
&& + (\vect{v}_{ma} \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM}
=C_a + S_a \nonumber
\end{eqnarray}
These equivalent forms of the kinetic equation are selected using \\
\parlink{includeXDotTerm} = \true  \;\;\; (Default) \\
\parlink{includeElectricFieldTermInXiDot} = \true \;\;\; (Default) \\
\parlink{useDKESExBDrift} = \false \;\;\; (Default) \\
\parlink{magneticDriftScheme} = 0 \;\;\; (Default) \\
\parlink{includePhi1} = \true \;\;\; (Not default) \\
\parlink{includeRadialExBDrive} = \false \;\;\; (Default) \\
\parlink{nonlinear} = \false \;\;\; (Default)

\section{Full $\vect{E}\times\vect{B}$ trajectories; no poloidal or toroidal magnetic drifts; Garc\'{i}a-Rega\~{n}a $\Phi_1$ terms}

This form of the drift-kinetic equation is nearly identical to equation (11) in Ref \cite{Regana2013}.
The one difference (which should be small) is that at the end of (11), Garc\'{i}a-Rega\~{n}a
has a term $\propto \vect{v}_m \cdot \nabla \Phi_1$, which is not (yet) in \sfincs.

\begin{eqnarray}
&&\left(v_{||}\vect{b} + \frac{d\Phi_0}{dr} \frac{1}{B^2} \vect{B}\times\nabla r \right) \cdot \nabla f_{s1} \\
&&+ \left[ - \frac{(1-\xi^2)v}{2B} \nabla_{||} B
+\frac{(1-\xi^2)\xi}{2B^3} \frac{d\Phi_0}{dr}\vect{B}\times\nabla r\cdot\nabla B 
-\frac{Z_s e}{v m_s}(1-\xi^2)(\nabla_{||}\Phi_1)
\right]
 \frac{\partial f_{s1}}{\partial \xi} \nonumber \\
&&+ \left[
-(\vect{v}_{ma} \cdot\nabla r) \frac{Z_s e}{2 T_s x_s} \frac{d\Phi_0}{dr} 
-\frac{Z_s e \xi}{v_s m_s}
\right] \frac{\partial f_{a1}}{\partial x_s} \nonumber \\
&&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
&& + (\vect{v}_E \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr}  + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM} \nonumber \\
&&+ (\vect{v}_{ma} \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM}
 = C_a + S_a \nonumber
\end{eqnarray}
Note that this equation is equivalent to the following one, in which the independent variables
are $(\mu,v_{||})$ instead of $(\xi,x_s)$:
\begin{eqnarray}
&&\left(v_{||}\vect{b} + \frac{d\Phi_0}{dr} \frac{1}{B^2} \vect{B}\times\nabla r \right) \cdot (\nabla f_{s1})_{\mu, v_{||}} \\
&&+ \left[ -\frac{Z_s e}{m_s} \nabla_{||} \Phi_1
- \mu \nabla_{||} B
-\frac{v_{||}}{B^2} \frac{d\Phi_0}{dr} \vect{b}\times\nabla B \times \nabla r \right]
\left( \frac{\partial f_{a1}}{\partial v_{||}} \right)_{\mu} \nonumber \\
&&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
&& + (\vect{v}_E \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr}  + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM} \nonumber \\
&& + (\vect{v}_{ma} \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM}
=C_a + S_a \nonumber
\end{eqnarray}
These equivalent forms of the kinetic equation are selected using \\
\parlink{includeXDotTerm} = \true  \;\;\; (Default) \\
\parlink{includeElectricFieldTermInXiDot} = \true \;\;\; (Default) \\
\parlink{useDKESExBDrift} = \false \;\;\; (Default) \\
\parlink{magneticDriftScheme} = 0 \;\;\; (Default) \\
\parlink{includePhi1} = \true \;\;\; (Not default) \\
\parlink{includeRadialExBDrive} = \true \;\;\; (Not default) \\
\parlink{nonlinear} = \true \;\;\; (Not default)

### 2016-07-20 fortran/version4/ChebyshevGrid.F90

@@ -1,0 +1,125 @@
+subroutine ChebyshevGrid(N, xMin, xMax, x, weights, D)
+  ! Creates a Chebyshev grid, together with the associated weights
+  ! for Clenshaw-Curtis integration, and the spectral
+  ! differentiation matrix.
+  ! Based on the Matlab function 
+  ! m20120124_06_multiChebyshevWeightsAndDifferentiation.m
+  ! 
+  ! Created by Matt Landreman, 
+  ! Massachusetts Institute of Technology, Plasma Science & Fusion Center, 2012.
+  !
+  ! Inputs:
+  ! N = number of grid points desired.
+  ! xMin = minimum of interval
+  ! xMax = maximum of interval
+  !
+  ! Outputs:
+  ! x(N) = grid points (abscissae)
+  ! weights(N) = Clenshaw-Curtis integration weights
+  ! D(N,N) = differentiation matrix
+
+  use GlobalVariables, only: prec
+
+  implicit none
+
+  integer, intent(in) :: N
+  real(prec), intent(in) :: xMin, xMax
+  real(prec), intent(out) :: x(N), weights(N), D(N,N)
+
+  integer :: i, j, N1, M
+  real(prec), parameter :: pi = 3.1415926535897932384626433d+0
+  real(prec), allocatable :: c(:), bigX(:,:), dX(:,:), sumD(:), cw(:), cc(:), f(:)
+
+  if (xMax <= xMin) then
+     print *,"Error! xMax should be larger than xMin."
+     stop
+  end if
+
+  if (N<1) then
+     print *,"Error! N must be at least 1."
+     stop
+  end if
+
+  ! ****************************************************
+  ! First, build abscissae and differentiation matrix
+  ! ****************************************************
+
+  N1 = N-1
+  x = [( cos(pi*i/N1), i=0,N1 )]
+
+  allocate(c(N))
+  allocate(sumD(N))
+  allocate(bigX(N,N))
+  allocate(dX(N,N))
+
+  c = 1
+  c(1) = 2
+  c(N) = 2
+  do i=2, N, 2
+     c(i) = -c(i)
+  end do
+
+  do i=1,N
+     bigX(i,:)=x(i)
+  end do
+
+  dX = bigX - transpose(bigX)
+
+  do i=1,N
+     dX(i,i) = 1
+  end do
+
+  do i=1,N
+     do j=1,N
[…]

### 2016-07-20 fortran/version4/ChebyshevInterpolationMatrix.F90

@@ -1,0 +1,86 @@
+subroutine ChebyshevInterpolationMatrix(N, M, xk, x, matrix)
+  ! Interpolates from Chebyshev points xk to a new grid x.
+  ! Based on the Matlab function chebint included with DMSuite.
+  !
+  ! Inputs:
+  ! N = number of Chebyshev points.
+  ! M = number of points on the new grid.
+  ! xk(N) = Chebyshev points.  (This subroutine will not check that they are indeed Chebyshev points!!)
+  ! x(M) = grid points onto which we will interpolate.
+  !
+  ! Outputs:
+  ! matrix(M,N) = interpolation matrix
+
+  use globalVariables, only: prec
+
+  implicit none
+
+  integer, intent(in) :: N, M
+  real(prec), intent(in) :: xk(N), x(M)
+  real(prec), intent(out) :: matrix(M,N)
+
+  integer :: i, j
+  real(prec), allocatable :: w(:), D(:,:), denominator(:), xScaled(:), xkScaled(:)
+  real(prec) :: temp, xMin, xMax, xMid
+
+  if (N<1) then
+     print *,"Error! N must be at least 1."
+     stop
+  end if
+
+  if (M<1) then
+     print *,"Error! M must be at least 1."
+     stop
+  end if
+
+  allocate(xScaled(M))
+  allocate(xkScaled(N))
+  xMin = minval(xk)
+  xMax = maxval(xk)
+  xMid = (xMin+xMax)*(0.5d+0)
+  xScaled = -2*(x-xMid)/(xMax-xMin)
+  xkScaled = -2*(xk-xMid)/(xMax-xMin)
+
+!!$  print *,"xScaled:",xScaled
+!!$  print *,"xkScaled:",xkScaled
+
+  allocate(w(N))
+  do i=1,N
+     w(i) = (-1) ** (i-1)
+  end do
+  w(1) = w(1) / (2.0d+0)
+  w(N) = w(N) / (2.0d+0)
+  
+  allocate(D(M,N))
+  do i=1,M
+     do j=1,N
+        temp = xScaled(i) - xkScaled(j)
+        if (abs(temp)<1.0d-15) then
+           D(i,j) = 1/(temp+1.0d-15)
+        else
+           D(i,j) = 1/temp
+        end if
+     end do
+  end do
+
+  allocate(denominator(M))
+  denominator = matmul(D,w)
+
+!!$  print *,"w:",w
[…]

### 2016-07-20 fortran/version4/FourierTransformMod.F90

@@ -1,0 +1,115 @@
+! We put this subroutine in a module because otherwise there is a segmentation fault when passing allocatable arrays as parameters.
+module FourierTransformMod
+
+  use globalVariables, only: dp
+
+  implicit none
+
+  private
+  real(dp), dimension(:,:), allocatable :: sinmtheta, cosmtheta, sinnzeta, cosnzeta
+
+  public :: FourierTransform, initFourierTrig
+
+contains
+
+  subroutine FourierTransform(original, vect)
+
+    use globalVariables, only: Ntheta, Nzeta, theta, zeta, xm, xn, NFourier, NPeriods
+    
+    implicit none
+    
+    real(dp), dimension(:,:), intent(in) :: original
+    real(dp), dimension(:), intent(out) :: vect
+    
+    real(dp) :: angle
+    integer :: itheta, izeta, factor
+    integer :: imn, m, n
+    integer :: tic, toc, countrate
+
+    ! ***************************************************************
+    ! Validate input
+    ! ***************************************************************
+   
+    if (xm(1) .ne. 0) stop "xm(1) should be 0."
+    if (xn(1) .ne. 0) stop "xn(1) should be 0."
+    if (size(vect) .ne. NFourier*2-1) stop "Size of vect is not correct"
+    if (size(original, 1) .ne. Ntheta) stop "Size of first dimension of original is not correct"
+    if (size(original, 2) .ne. Nzeta)  stop "Size of second dimension of original is not correct"
+    
+    ! ***************************************************************
+    ! Main calculation
+    ! ***************************************************************
+    
+    vect = 0
+    call system_clock(tic,countrate)
+
+    ! Rather than use an FFT, let's just use a 'slow' but transparent algorithm. Speed is not crucial for this subroutine.
+    do imn = 2,NFourier
+       m = xm(imn)
+       n = xn(imn)/NPeriods
+       do itheta = 1,Ntheta
+          do izeta = 1,Nzeta
+             !angle = m*theta(itheta) - n*NPeriods*zeta(izeta)
+             
+             !! cos(m theta - n zeta) = cos(m theta) cos(n zeta) + sin(m theta) sin(n zeta)
[…]

### 2016-07-20 fortran/version4/chooseFourierModes.F90

@@ -1,0 +1,108 @@
+subroutine chooseFourierModes()
+
+  use rankMod
+  use petscsys
+  use FourierTransformMod
+  use globalVariables, only: NFourier, NFourier2, xm, xn, mmax, nmax, NPeriods, prec
+
+  implicit none
+
+  integer :: NFourier_original
+  integer :: m, n, index, imn
+  real(prec), dimension(:,:), allocatable :: patternWithExpectedSpectrum
+  real(prec), dimension(:), allocatable :: FourierVector, amplitudes
+  integer, dimension(:), allocatable :: xn_sorted, xm_sorted, permutation
+
+  allocate(xn_sorted(NFourier))
+  allocate(xm_sorted(NFourier))
+
+  ! Temporarily set up (xm,xn) to include all possible modes.
+  NFourier_original = NFourier
+  NFourier = mmax*(2*nmax+1) + nmax + 1
+  NFourier2 = NFourier*2-1
+  allocate(xm(NFourier))
+  allocate(xn(NFourier))
+
+  xm=0
+  xn=0
+  ! Handle the m=0 modes
+  do n=1,nmax
+     xn(n+1)=n
+  end do
+  ! Handle the m>0 modes
+  index = nmax+1
+  do m=1,mmax
+     do n=-nmax,nmax
+        index = index + 1
+        xn(index) = n
+        xm(index) = m
+     end do
+  end do
+  xn = xn*NPeriods
+!!$  print *,"Initialized xm,xn to all non-constant modes."
+!!$  print *,"xm=",xm
+!!$  print *,"xn=",xn
+!!$
+!!$  print *,"B:"
+!!$  do n=1,Ntheta
+!!$     print *,B(n,:)
+!!$  end do
+
+  ! Guess a function that we expect will have a similar Fourier spectrum to the distribution function:
+  allocate(patternWithExpectedSpectrum(Ntheta,Nzeta))
+  patternWithExpectedSpectrum = (B ** 6) + 1/(B*B)
+
+  allocate(FourierVector(NFourier2))
+  call initFourierTrig()
+  call FourierTransform(patternWithExpectedSpectrum, FourierVector)
+  deallocate(patternWithExpectedSpectrum)
+
+  allocate(amplitudes(NFourier))
+  amplitudes=0
+  do imn=2,NFourier
+     amplitudes(imn) = sqrt(FourierVector(imn)**2 + FourierVector(imn+NFourier-1)**2)
[…]

### 2016-07-20 fortran/version4/createGrids.F90

@@ -1,0 +1,864 @@
+#include "PETScVersions.F90"
+#if (PETSC_VERSION_MAJOR < 3 || (PETSC_VERSION_MAJOR==3 && PETSC_VERSION_MINOR < 6))
+#include <finclude/petscdmdadef.h>
+#else
+#include <petsc/finclude/petscdmdadef.h>
+#endif
+
+  subroutine createGrids()
+
+    use globalVariables
+    use polynomialDiffMatrices
+    use xGrid
+    use petscdmda
+    use geometry
+    use indices
+    use export_f
+
+    implicit none
+
+    PetscErrorCode :: ierr
+    integer :: i, j, itheta, izeta, scheme
+    real(prec), dimension(:,:), allocatable :: d2dtheta2, d2dzeta2
+    real(prec), dimension(:), allocatable :: theta_preconditioner, thetaWeights_preconditioner
+    real(prec), dimension(:,:), allocatable :: d2dtheta2_preconditioner
+    real(prec), dimension(:), allocatable :: zeta_preconditioner, zetaWeights_preconditioner
+    real(prec), dimension(:,:), allocatable :: d2dzeta2_preconditioner
+    real(prec), dimension(:), allocatable :: xWeightsPotentials
+
+    real(prec), dimension(:), allocatable :: xWeights_plus1
+    real(prec), dimension(:,:), allocatable :: ddx_plus1, d2dx2_plus1
+    real(prec), dimension(:,:), allocatable :: interpolateXToXPotentials_plus1, extrapMatrix
+    real(prec), dimension(:), allocatable :: x_subset, xWeights_subset
+    real(prec), dimension(:,:), allocatable :: ddx_subset, d2dx2_subset
+
+
+    DM :: myDM
+    integer, parameter :: bufferLength = 200
+    character(len=bufferLength) :: procAssignments
+
+    integer :: tag, dummy(1)
+    integer :: status(MPI_STATUS_SIZE)
+
+
+    ! *******************************************************************************
+    ! Do a few sundry initialization tasks:
+    ! *******************************************************************************
+
+    if (constraintScheme < 0) then
+       if (collisionOperator == 0) then
+          constraintScheme = 1
+       else
+          constraintScheme = 2
+       end if
+    end if
+
+    if (masterProc) then
[…]

### 2016-07-20 fortran/version4/diagnostics.F90

@@ -1,0 +1,995 @@
+#include "PETScVersions.F90"
+#if (PETSC_VERSION_MAJOR < 3 || (PETSC_VERSION_MAJOR==3 && PETSC_VERSION_MINOR < 6))
+#include <finclude/petscsnesdef.h>
+#else
+#include <petsc/finclude/petscsnesdef.h>
+#endif
+
+! This next subroutine is called as a "Monitor" of SNES, set in solver.F90 using SNESSetMonitor.
+  subroutine diagnosticsMonitor(mysnes, iterationNum, residual, userContext, ierr)
+
+    use globalVariables, only: masterProc, iterationForMatrixOutput
+    use petscsnes
+
+    implicit none
+
+    SNES :: mysnes
+    PetscInt :: iterationNum
+    PetscReal :: residual
+    integer :: userContext(*)
+    PetscErrorCode :: ierr
+    KSP :: myKSP
+
+    Vec :: soln
+
+    if (masterProc) then
+       if (iterationNum > 0) then
+          print "(a,i4,a)",    "--------- Completed iteration ",iterationNum," of SNES -----------------------------------"
+       end if
+       print "(a,es14.7,a)","--------- Residual function norm: ",dble(residual)," -----------------------------"
+    end if
+
+    if (iterationNum==0) then
+       return
+    end if
+
+
+    call SNESGetKSP(mysnes, myKSP, ierr)
+    call checkIfKSPConverged(myKSP)
+
+    iterationForMatrixOutput = iterationNum
+    call SNESGetSolution(mysnes, soln, ierr)
+    call diagnostics(soln, iterationNum)
+
+  end subroutine diagnosticsMonitor
+
+  ! *******************************************************************************************
+  ! *******************************************************************************************
+
+  subroutine checkIfKSPConverged(myKSP)
+
+    use globalVariables, only: integerToRepresentTrue, integerToRepresentFalse, useIterativeLinearSolver, masterProc
+    use petscksp
+
+    implicit none
+
+    KSP :: myKSP
+    KSPConvergedReason :: reason
+    integer :: ierr
+    integer :: didLinearCalculationConverge
+
+    if (useIterativeLinearSolver) then
+       call KSPGetConvergedReason(myKSP, reason, ierr)
+       if (reason>0) then
+          if (masterProc) then
[…]

### 2016-07-20 fortran/version4/indices.F90

@@ -1,0 +1,308 @@
+  ! *********************************************************
+  ! For constraintScheme==0,
+  ! *********************************************************
+  
+  ! Order of the rows of the matrix and of the RHS:
+  ! --------------------------------
+  ! for iSpecies = 1:Nspecies
+  !   for ix = 1:Nx
+  !     for L = 0:(NL-1)
+  !       for iFourier = 1:NFourier2
+  !         Enforce the drift-kinetic equation
+  ! for iFourier = 1:NFourier2
+  !   Enforce quasineutrality
+  ! Force <\tilde{\phi}> = 0
+  
+  ! Order of the vector of unknowns & of columns in the matrix:
+  ! --------------------------------
+  ! for iSpecies = 1:Nspecies
+  !   for ix = 1:Nx
+  !     for L = 0:(NL-1)
+  !       for iFourier = 1:NFourier2
+  !         Distribution function
+  ! for iFourier = 1:NFourier2
+  !   \tilde{\phi}
+  ! lambda
+
+
+  
+  ! *********************************************************
+  ! For constraintScheme==1, 3, or 4:
+  ! *********************************************************
+  
+  ! Order of the rows of the matrix and of the RHS:
+  ! --------------------------------
+  ! for iSpecies = 1:Nspecies
+  !   for ix = 1:Nx
+  !     for L = 0:(NL-1)
+  !       for iFourier = 1:NFourier2
+  !         Enforce the drift-kinetic equation
+  ! for iFourier = 1:NFourier
+  !   Enforce quasineutrality
+  ! Force <\tilde{\phi}> = 0
+  ! for iSpecies = 1:Nspecies
+  !   Force <n_1> = 0
+  !   Force <p_1> = 0
+  
+  
+  ! Order of the vector of unknowns & of columns in the matrix:
+  ! --------------------------------
+  ! for iSpecies = 1:Nspecies
+  !   for ix = 1:Nx
+  !     for L = 0:(NL-1)
+  !       for iFourier = 1:NFourier2
+  !           Enforce the drift-kinetic equation
+  ! for iFourier = 1:NFourier2
+  !     \tilde{\phi}
+  ! lambda
+  ! for iSpecies = 1:Nspecies
+  !   particle source
+  !   energy source
+
+
+  ! *********************************************************
+  ! For constraintScheme==2,
+  ! *********************************************************
+
+  ! Order of the rows of the matrix and of the RHS:
[…]

### 2016-07-20 fortran/version4/polynomialDiffMatrices.F90

@@ -1,0 +1,185 @@
+! This module computes the first and second spectral differentiation matrices.
+! Any weight and grid points are allowed.  However, the method is poorly
+! conditioned unless the grid points correspond to Gaussian integration
+! abscissae for the given weight.  No warning is given in the case of
+! poor conditioning.  Presently, the weight function is
+! set to exp(-x^2), but you are free to change it.  When changing the
+! weight, you must also change the functions dweightdxOverWeight
+! and d2weightdx2OverWeight appropriately.
+!
+! This algorithm closely follows poldif.m for MATLAB
+! Part of DMSuite
+! written by J.A.C. Weideman, S.C. Reddy 1998
+! Available here
+! http://www.mathworks.com/matlabcentral/fileexchange/29
+! or here
+! http://dip.sun.ac.za/~weideman/research/differ.html
+!
+! Matt Landreman
+! Massachusetts Institute of Technology
+! Plasma Science & Fusion Center
+! November, 2012
+!
+module polynomialDiffMatrices
+
+  implicit none
+
+  public :: makeXPolynomialDiffMatrices
+  private :: weight, dweightdxOverWeight, d2weightdx2OverWeight
+
+contains
+
+  ! ------------------------------------------------------------------
+
+  function weight(x)
+
+    use globalVariables, only: prec
+    use xGrid, only: xGrid_k
+
+    implicit none
+
+    real(prec), intent(in) :: x
+    real(prec) :: weight
+
+    weight = exp(-x*x)*(x ** xGrid_k)
+
+  end function weight
+
+  ! ------------------------------------------------------------------
+
+  function dweightdxOverWeight(x)
+
+    use xGrid, only: xGrid_k
+
+    implicit none
+
+    real(prec), intent(in) :: x
+    real(prec) :: dweightdxOverWeight
+
+    if (abs(x)<1e-12) then
+       ! Handle possible point at x=0, avoiding divide-by-0:
+       dweightdxOverWeight = 0.0d+0
+    else
+       !dweightdxOverWeight = -2*x
+       dweightdxOverWeight = xGrid_k / x - 2*x
+    end if
+
[…]

### 2016-07-20 fortran/version4/populateMatrix.F90

@@ -1,0 +1,2361 @@
+! For compilers that do not include the error function erf(x), the line
+! below should be un-commented:
+!#define USE_GSL_ERF
+  
+#include "PETScVersions.F90"
+#if (PETSC_VERSION_MAJOR < 3 || (PETSC_VERSION_MAJOR==3 && PETSC_VERSION_MINOR < 6))
+#include <finclude/petscmatdef.h>
+#else
+#include <petsc/finclude/petscmatdef.h>
+#endif
+
+  subroutine populateMatrix(matrix, whichMatrix, stateVec)
+
+    use petscmat
+    use globalVariables
+    use sparsify
+    use indices
+    use xGrid, only: xGrid_k
+
+    implicit none
+
+    Mat :: matrix
+    integer, intent(in) :: whichMatrix
+    Vec :: stateVec ! stateVec is ignored when includePhi1=false or when evaluating the residual.
+
+    ! Allowed values for whichMatrix:
+    ! 0 = preconditioner for Jacobian
+    ! 1 = Jacobian
+    ! 2 = matrix which multiplies f0 when evaluating the residual
+    ! 3 = matrix which multiplies f1 when evaluating the residual
+
+    PetscErrorCode :: ierr
+    real(prec) :: Z, nHat, THat, mHat, sqrtTHat, sqrtMHat, speciesFactor, speciesFactor2
+    real(prec) :: T32m, factor, LFactor, temp, temp1, temp2, xDotFactor, xDotFactor2, stuffToAdd
+    real(prec) :: factor1, factor2, factorJ1, factorJ2, factorJ3, factorJ4, factorJ5  !!Added by AM 2016-03
+    real(prec), dimension(:), allocatable :: xb, expxb2
+    real(prec), dimension(:,:), allocatable :: thetaPartOfTerm, localThetaPartOfTerm
+    real(prec), dimension(:,:), allocatable :: xPartOfXDot_plus, xPartOfXDot_minus, xPartOfXDot
+    real(prec), dimension(:,:), allocatable :: ddxToUse_plus, ddxToUse_minus
+    integer :: i, j, ix, ispecies, itheta, izeta, L, ixi, index, ix_row, ix_col
+    integer :: rowIndex, colIndex
+    integer :: rowIndex2, L2 !!Added by AM 2016-03
+    integer :: ell, iSpeciesA, iSpeciesB, maxL
+    integer, dimension(:), allocatable :: rowIndices, colIndices
[…]

### 2016-07-20 fortran/version4/radialCoordinates.F90

@@ -1,0 +1,246 @@
+module radialCoordinates
+
+  use globalVariables
+
+  implicit none
+
+contains
+
+  ! -----------------------------------------------------------------------------------
+
+  subroutine setInputRadialCoordinateWish
+
+    ! This subroutine takes the selected "wish" radial coordinate, and uses it to over-write
+    ! the other "wish" radial coordinates.
+
+    implicit none
+
+    if (aHat<0) then
+       print *,"Error! aHat cannot be <0."
+       stop
+    end if
+
+    ! First, set psiHat_wish itself:
+
+    select case (inputRadialCoordinate)
+    case (0)
+       ! Selected input radial coordinate is psiHat.
+       ! Nothing to do here.
+
+       if (masterProc) then
+          print *,"Selecting the flux surface to use based on psiHat_wish = ",psiHat_wish
+       end if
+       if (psiHat_wish / psiAHat < 0) then
+          print *,"Error! psiHat_wish/psiAHat cannot be <0."
+          stop
+       end if
+       if (psiHat_wish / psiAHat > 1) then
+          print *,"Error! psiHat_wish/psiAHat cannot be >1."
+          stop
+       end if
+
+    case (1)
+       ! Selected input radial coordinate is psiN.
+       psiHat_wish = psiN_wish * psiAHat
+
+       if (masterProc) then
+          print *,"Selecting the flux surface to use based on psiN_wish = ",psiN_wish
+       end if
+       if (psiN_wish<0) then
+          print *,"Error! psiN_wish cannot be <0."
+          stop
+       end if
+       if (psiN_wish>1) then
+          print *,"Error! psiN_wish cannot be >1."
+          stop
+       end if
+
+    case (2)
+       ! Selected input radial coordinate is rHat.
+       psiHat_wish = psiAHat * rHat_wish * rHat_wish / (aHat * aHat)
+
+       if (masterProc) then
+          print *,"Selecting the flux surface to use based on rHat_wish = ",rHat_wish
+       end if
+       if (rHat_wish<0) then
+          print *,"Error! rHat_wish cannot be <0."
+          stop
+       end if
+       if (rHat_wish>aHat) then
+          print *,"Error! rHat_wish cannot be > aHat."
+          stop
+       end if
+
+    case (3)
[…]

### 2016-07-20 fortran/version4/rankMod.F90

@@ -1,0 +1,209 @@
+! We put this subroutine in a module because otherwise there is a segmentation fault when passing allocatable arrays as parameters.
+module rankMod
+  implicit none
+contains
+
+  Subroutine rank (XDONT, IRNGT)
+  ! Downloaded from http://www.fortran-2000.com/rank/mrgrnk.f90
+  ! On output, the integer array IRNGT gives the order of the elements of the double-precision input array XDONT,
+  ! ordered from largest to smallest.
+! __________________________________________________________
+!   MRGRNK = Merge-sort ranking of an array
+!   For performance reasons, the first 2 passes are taken
+!   out of the standard loop, and use dedicated coding.
+! __________________________________________________________
+! __________________________________________________________
+      Integer, Parameter :: kdp = selected_real_kind(15)
+      Real (kind=kdp), Dimension (:), Intent (In) :: XDONT
+      Integer, Dimension (:), Intent (Out) :: IRNGT
+! __________________________________________________________
+      Real (kind=kdp) :: XVALA, XVALB
+!
+      Integer, Dimension (SIZE(IRNGT)) :: JWRKT
+      Integer :: LMTNA, LMTNC, IRNG1, IRNG2
+      Integer :: NVAL, IIND, IWRKD, IWRK, IWRKF, JINDA, IINDA, IINDB
+!
+      NVAL = Min (SIZE(XDONT), SIZE(IRNGT))
+      Select Case (NVAL)
+      Case (:0)
+         Return
+      Case (1)
+         IRNGT (1) = 1
+         Return
+      Case Default
+         Continue
+      End Select
+!
+!  Fill-in the index array, creating ordered couples
+!
+      Do IIND = 2, NVAL, 2
+         If (XDONT(IIND-1) <= XDONT(IIND)) Then
+            IRNGT (IIND-1) = IIND - 1
+            IRNGT (IIND) = IIND
+         Else
+            IRNGT (IIND-1) = IIND
+            IRNGT (IIND) = IIND - 1
+         End If
+      End Do
+      If (Modulo(NVAL, 2) /= 0) Then
+         IRNGT (NVAL) = NVAL
+      End If
+!
[…]

### 2016-07-20 fortran/version4/readVMEC/pisa_io.f90

@@ -1,0 +1,220 @@
+!> Module io_module provides utilities and definitions for input and
+!! output operations.
+!!
+!! @author Joachim Geiger
+!! @version 1.1
+module pisa_io
+implicit none
+
+integer, parameter :: eof=-1      !< end of file reached
+integer, parameter :: eol=-2      !< end of line reached
+logical, parameter :: l_show=.false.  !< test flag
+
+private :: l_show
+
+contains
+
+!----------------------------------------------------------------------
+!>find a unit number available for i/o action
+!!taken from the book "Object oriented programming via Fortran 90/95"
+!! by Ed Akin, p. 305
+!* * * * * * * * * * * * * * * * * * * * * * * * * * * * *
+!! History:
+!! - 1.0: loop through all allowed units was only done if last_unit > 0
+!!        and not if the unit was already open.
+!! - 1.1: added loop through all allowed units in case the unit was open!
+!! - 1.2: logical variable open changed to is_open (keyword).
+!!        change integer variable count to icount (keyword).
+!!
+function get_next_io_unit () result (next)
+  implicit none
+  integer :: next   !< the next available unit number
+  integer, parameter :: min_unit = 10, max_unit = 999
+  integer, save      :: last_unit = 0   !< initalize
+  integer            :: icount          !< number of failures
+  logical            :: is_open         !< file status
+
+  icount = 0 ; next = min_unit - 1
+  if ( last_unit > 0 ) then ! check next in line
+    next = last_unit + 1
+    inquire (unit=next, opened=is_open)
+    if ( .not. is_open )then
+      last_unit = next ! found it
+      return
+    end if ! found it
+  end if ! last_unit
+! next unit was not open => loop over units
+  do ! forever
+    next = next + 1
+    inquire (unit=next, opened=is_open)
+    if ( .not. is_open ) then
+      last_unit = next     ! found it
+      exit ! the unit loop
+    end if
+    if ( next == max_unit ) then ! attempt reset 3 times
+      last_unit = 0
+      icount     = icount + 1
+      if ( icount <= 3 ) next = min_unit - 1
+    end if ! reset try
+    if ( next > max_unit ) then ! abort
+      print *,'ERROR: max unit exceeded in get_next_io_unit'
+      stop    'ERROR: max unit exceeded in get_next_io_unit'
+    end if ! abort
+  end do ! over unit numbers
[…]

### 2016-07-20 fortran/version4/sfincs.F90

@@ -1,0 +1,128 @@
+! Main program
+
+#include "PETScVersions.F90"
+#if (PETSC_VERSION_MAJOR < 3 || (PETSC_VERSION_MAJOR==3 && PETSC_VERSION_MINOR < 6))
+#include <finclude/petscsysdef.h>
+#else
+#include <petsc/finclude/petscsysdef.h>
+#endif
+
+program sfincs
+
+  use globalVariables
+  use writeHDF5Output
+  use readInput
+  use petscsysdef
+  use solver
+  use geometry
+  use radialCoordinates
+
+  implicit none
+
+  PetscErrorCode ierr
+  PetscLogDouble :: startTime, time1
+
+  call PetscInitialize(PETSC_NULL_CHARACTER, ierr)
+
+  call MPI_COMM_SIZE(PETSC_COMM_WORLD, numProcs, ierr)
+  call MPI_COMM_RANK(PETSC_COMM_WORLD, myRank, ierr)
+  masterProc = (myRank==0)
+
+  ! In the future, if we want to divide the processors into sub-communicators, this next line would change:
+  MPIComm = PETSC_COMM_WORLD
+
+  if (masterProc) then
+     print *,"****************************************************************************"
+     print *,"SFINCS: Stellarator Fokker-Plank Iterative Neoclassical Conservative Solver"
+     print *,"Version 4"
+#if defined(PETSC_USE_REAL_SINGLE)
+     print *,"Using single precision."
+#else
+     print *,"Using double precision."
+#endif
+     if (numProcs==1) then
+        print *,"Serial job (1 process) detected."
+     else
+        print "(a, i4, a)", " Parallel job (",numProcs," processes) detected."
+     end if
+  end if
+
+  call PetscTime(time1, ierr)
+  startTime = time1
+
+  call readNamelistInput()
+  call validateInput()
+
+  ! If running with >1 proc,
+  ! make sure either superlu_dist or mumps is installed, and pick which one
+  ! of these packages to use:
+  call chooseParallelDirectSolver()
+
+  if (masterProc) then
+     print *,"---- Physics parameters: ----"
+     print *,"Number of particle species = ", Nspecies
+     print *,"Delta (rho* at reference parameters)          = ", Delta
+     print *,"alpha (e Phi / T at reference parameters)     = ", alpha
+     print *,"nu_n (collisionality at reference parameters) = ", nu_n
+     if (includePhi1) then
+        print *,"Nonlinear run"
+        if (includePhi1InKineticEquation) then
[…]

### 2016-07-20 fortran/version4/uniformDiffMatrices.F90

@@ -1,0 +1,888 @@
+subroutine uniformDiffMatrices(N, xMin, xMax, scheme, x, weights, ddx, d2dx2)
+  ! Finite difference and spectral differentiation matrices and integration
+  ! weights for a uniform grid.
+  !
+  ! Created by Matt Landreman, 
+  ! Massachusetts Institute of Technology, Plasma Science & Fusion Center, 2012.
+  !
+  ! Inputs:
+  !   N = number of grid points.
+  !   xMin = minimum value in the domain.
+  !   xMax = maximum value in the domain.
+  !   scheme = switch for controlling order of accuracy for differentiation
+  !            and handling of endpoints.
+  !
+  ! Options for scheme:
+  ! 0 =  The domain [xMin, xMax] is assumed to be periodic. A 3-point stencil
+  !      is used everywhere. A grid point will be placed at xMin but not 
+  !      xMax.
+  ! 1 =  Same as scheme=0, except a grid point will be placed at xMax but not
+  !      xMin.
+  ! 2 =  The domain [xMin, xMax] is assumed to be non-periodic. A 3-point 
+  !      stencil is used everywhere.  The first and last row of the
+  !      differentiation matrices will use one-sided differences, so they
+  !      will each have a non-tridiagonal element.
+  ! 3 =  The same as scheme=2, except that the first differentiation matrix
+  !      will use a 2-point 1-sided stencil for the first and last elements
+  !      so the matrix is strictly tri-diagonal.  The 2nd derivative matrix
+  !      is the same as for option 2, since it is not possible to compute 
+  !      the 2nd derivative with only a 2-point stencil.
+  ! 10 = The domain [xMin, xMax] is assumed to be periodic. A 5-point stencil
+  !      is used everywhere. A grid point will be placed at xMin but not 
+  !      xMax.  This option is like scheme=0 but more accurate.
+  ! 11 = Same as scheme=10, except a grid point will be placed at xMax but
+  !      not xMin.  This option is like scheme=1 but more accurate.
+  ! 12 = The domain [xMin, xMax] is assumed to be non-periodic. A 5-point 
+  !      stencil is used everywhere.  The first two and last two rows of 
[…]

### 2016-07-20 fortran/version4/xGrid.F90

@@ -1,0 +1,528 @@
+  ! This module computes the nodes and weights for Gaussian quadrature for any
+  ! weight function on any interval.  The algorithm used is the Stieltjes
+  ! procedure described in section 4.6.3 of Numerical Recipes, 3rd edition,
+  ! together with the Jacobi matrix procedure of section 4.6.2.  See also
+  ! www.nr.com/webnotes?3
+  ! The module is presently configured to use the weight exp(-x^2)*(x^k) and to
+  ! use the interval [0, Inf), which yields the grid and integration weights
+  ! dicussed in the following paper:
+  ! Landreman & Ernst, Journal of Computational Physics 243, 130 (2013).
+  ! This grid is useful for velocity space.
+  ! You can change the "weight" function in the module and the integration
+  ! domain in the makeXGrid subroutine if you wish.
+  !
+  ! The makeXGrid subroutine calls LAPACK, which is
+  ! automatically included in any program that is linked to the PETSc 
+  ! libraries.
+  !
+  ! The makeXGrid subroutine also calls QUADPACK for numerical
+  ! integration.  You can download QUADPACK here:
+  ! http://www.netlib.org/quadpack/
+  !
+  ! Matt Landreman
+  ! Massachusetts Institute of Technology
+  ! Plasma Science & Fusion Center
+  ! November, 2012
+  !
+
+#define workspaceSize 5000
+
+  ! Call QUADPACK routines to evaluate integrals:
+  ! qage = single-precision quadrature on finite domain
+  ! dqage = double-precision quadrature on finite domain
+  ! qagie = single-precision quadrature on (semi-)infinite domain
+  ! dqagie = double-precision quadrature on (semi-)infinite domain
+
+#define integrate dqage
+#define integrate_semiinf dqagie
+#define zero 0.0d+0
+
+! QUADPACK gives ier.ne.0 when it cannot achieve the requested tolerance.
+! When this occurs, it probably just means it is not feasbible to get as many digits
+! of precision as we requested, so we can safely ignore the warnings.
+!#define showQuadpackWarnings .true.
+#define showQuadpackWarnings .false.
+
+
+#define line "******************************************************************"
+
+  module xGrid
+
+    use globalVariables, only: prec
+
+    implicit none
+
+    private
[…]

### 2016-07-26 doc/manual/version3/inputParameters.tex
Updated manual with new parameters related to Phi1 calculations

--- 2016-02-06 doc/manual/inputParameters.tex
@@
 {Radial temperature gradients of each species, with respect to the radial coordinate $r_N$, normalized by the reference temperature $\bar{T}$.}
+
+\myhrule
+
+\param{withAdiabatic}
+{Boolean}
+{\false}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true}
+{If this parameter is \true, an adiabatic species is added into the quasineutrality equation (but has no other effect).}
+
+\myhrule
+
+\param{adiabaticZ}
+{real}
+{-1.0}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and \parlink{withAdiabatic} == \true}
+{Charge of adiabatic species, in units of the proton charge $e$.}
+
+\myhrule
+
+\param{adiabaticMHat}
+{real}
+{5.44617e-4}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and \parlink{withAdiabatic} == \true}
+{Mass of adiabatic species, in units of the reference mass $\bar{m}$.}
+
+\myhrule
+
+\param{adiabaticNHat}
+{real}
+{1.0}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and \parlink{withAdiabatic} == \true}
+{Density of adiabatic species, in units of the reference density $\bar{n}$.}
+
+\myhrule
+
+\param{adiabaticTHat}
+{real}
+{1.0}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and \parlink{withAdiabatic} == \true}
+{Temperature of adiabatic species, in units of the reference temperature $\bar{T}$.}
 
@@
 and the quasineutrality equation is not solved.  If true, then terms involving $\Phi_1$ are included
-in the kinetic equation, and the quasineutrality equation is solved at each point on the flux surface.
+in the kinetic equation if \parlink{includePhi1inKineticEquation} = \true, and the quasineutrality equation is solved at each point on the flux surface (make sure that your input densities fulfill quasineutrality, see \parlink{quasineutralityOption}).
 In this latter case, many more quantities are computed and saved in the output file, such as radial fluxes
-associated with the radial $\vect{E}\times\vect{B}$ drift.}
-
-\myhrule
-
-\param{includeRadialExBDrive}
+associated with the radial $\vect{E}\times\vect{B}$ drift.\newline
+%%Text added by AM 2016-07
+If true, the system becomes nonlinear in the unknowns $f_{s1}$ and $\Phi_1$. 
+Newton's method will be used to solve the nonlinear system, meaning that the usual linear solve in \sfincs~must be iterated several times.
+}
+
+\myhrule
+
+\param{includePhi1inKineticEquation}
 {Boolean}
-{\false}
-{Whenever \parlink{RHSMode} == 1.}
-{If true, the radial $\vect{E}\times\vect{B}$ term $(\vect{v}_{E} \cdot\nabla\psi)f_{Ms} [(1/n_s)(dn_s/d\psi) + (x_s^2-3/2)(1/T_s)(dT_s/d\psi)]$
-will be included in the kinetic equation. This is one of the terms considered in Ref \cite{Regana2013} which
-should be unimportant for the main ions but which may be important for impurities.
-Use of this option requires \parlink{includePhi1}=\true, since the radial $\vect{E}\times\vect{B}$ drift
-arises due to $\Phi_1$.
-}
-
-\myhrule
-
-\param{nonlinear}
-{Boolean}
-{\false}
-{Whenever \parlink{RHSMode} == 1.}
-{If true, the term
-$-(Z_s e/m) (\nabla_{||} \Phi_1) (\partial f_{s1}/\partial v_{||})_\mu$ will be included
-in the kinetic equation.  This is one of the terms considered in Ref \cite{Regana2013} which
-should be unimportant for the main ions but which may be important for impurities.
-This term is nonlinear in the unknowns $f_{s1}$ and $\Phi_1$.
-Newton's method will be used to solve the nonlinear system, meaning that the usual linear solve in \sfincs~must
-be iterated several times.
-Running with {\ttfamily nonlinear}=\true requires \parlink{includePhi1}=\true
-}
+{\true}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true}
+{If true, the terms containing $\Phi_1$ will be included in the kinetic equation. 
+If false, and \parlink{includePhi1} = \true, then $\Phi_1$ is only included in the quasineutrality equation. 
+}
+
+\myhrule
+
+\param{quasineutralityOption}
+{integer}
+{1}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true}
+{This variable controls which quasineutrality equation is used, and must be set to either 1 or 2.
+Make sure that your input densities fulfill quasineutrality (also accounting for the adiabatic species if \parlink{withAdiabatic}=\true).\\
+
+{\ttfamily quasineutralityOption} = 1: Full quasineutrality equation. All species (kinetic and adiabatic if applicable) are included in the quasineutrality equation.\\
+
+{\ttfamily quasineutralityOption} = 2: {\ttfamily EUTERPE} quasineutrality.
+\sfincs~must be run with an adiabatic species, \parlink{withAdiabatic}=\true, and only the adiabatic species and the first kinetic species are included in the quasineutrality equation (their input densities must fulfill quasineutrality).
+This is the quasineutrality considered in Ref \cite{Regana2013}.
+}
+
+%%The includeRadialExBDrive parameter has been removed by AM 2016-07
+%\myhrule
+%
+%\param{includeRadialExBDrive}
+%{Boolean}
+%{\false}
+%{Whenever \parlink{RHSMode} == 1.}
+%{If true, the radial $\vect{E}\times\vect{B}$ term $(\vect{v}_{E} \cdot\nabla\psi)f_{Ms} [(1/n_s)(dn_s/d\psi) + (x_s^2-3/2)(1/T_s)(dT_s/d\psi)]$
+%will be included in the kinetic equation. This is one of the terms considered in Ref \cite{Regana2013} which
+%should be unimportant for the main ions but which may be important for impurities.
+%Use of this option requires \parlink{includePhi1}=\true, since the radial $\vect{E}\times\vect{B}$ drift
+%arises due to $\Phi_1$.
+%}
+
+%%The nonlinear parameter has been removed by AM 2016-07
+%\myhrule
+%
+%\param{nonlinear}
+%{Boolean}
+%{\false}
+%{Whenever \parlink{RHSMode} == 1.}
+%{If true, the term
+%$-(Z_s e/m) (\nabla_{||} \Phi_1) (\partial f_{s1}/\partial v_{||})_\mu$ will be included
+%in the kinetic equation.  This is one of the terms considered in Ref \cite{Regana2013} which
+%should be unimportant for the main ions but which may be important for impurities.
+%This term is nonlinear in the unknowns $f_{s1}$ and $\Phi_1$.
+%Newton's method will be used to solve the nonlinear system, meaning that the usual linear solve in \sfincs~must
+%be iterated several times.
+%Running with {\ttfamily nonlinear}=\true requires \parlink{includePhi1}=\true
+%}
 

### 2016-08-16 doc/manual/version4/inputParameters.tex
v4 Added FourierOption and FourierFactor

--- 2016-02-06 doc/manual/inputParameters.tex
@@
 {Always}
-{If this switch is set to true, the matrix, right-hand-side, and solution vector of the
+{If this switch is set to true, the matrix, preconditioner matrix, right-hand-side, and solution vector of the
 linear system will be saved in PETSc's binary format.  
-The preconditioner matrix will also
-be saved if \parlink{useIterativeLinearSolver} == \true.}
 These matrices and vectors are not very interesting for routine use of the code,
@@
 separate \HDF~file.
+}
 
@@
 {integer}
-{1}
+{0}
 {Only when \parlink{geometryScheme} == 5.}
@@
 
-\param{Ntheta}
-{integer}
-{15}
-{Always}
-{Number of grid points in the poloidal angle. 
-This parameter should be odd; see \parlink{forceOddNthetaAndNzeta} in this namelist.
+\param{NFourier}
+{integer}
+{10}
+{Always}
+{Number of Fourier modes in the poloidal and toroidal angles used to represent the distribution function.
 Memory and time requirements DO depend strongly on this parameter.
-For stellarator calculations, this parameter can usually be in the range 15-25.
-For tokamak calculations at low collisionality, the value of this parameter may need to be higher.}
-
-\myhrule
-
-\param{Nzeta}
-{integer}
-{15}
-{Always}
-{Number of grid points in the toroidal angle (per identical segment of the stellarator.)  
-This parameter should be odd; see \parlink{forceOddNthetaAndNzeta} in this namelist.
-Memory and time requirements DO depend strongly on this parameter.
-Set this parameter to 1 for a tokamak calculation.
-For stellarator calculations, the value of this parameter required for convergence depends strongly on
-the collisionality. At high collisionality, this parameter can be several 10s, depending
-on the complexity of $B(\theta,\zeta)$. At low collisionality, this parameter may need to be many 10s or
-even $>$ 100 for convergence.
+The value of this parameter required for convergence depends strongly on
+the collisionality, with higher values required at low collisionality.
+}
+
+\myhrule
+
+\param{mmax}
+{integer}
+{32}
+{Always}
+{Determines the maximum poloidal mode number $m$ which can be included: $m \le${\ttfamily mmax}. Memory and time requirements do not depend strongly on this parameter.
+}
+
+\myhrule
+
+\param{nmax}
+{integer}
+{32}
+{Always}
+{Determines the maximum toroidal mode number $n$ which can be included: $|n|/${\ttfamily Nperiod}$\le${\ttfamily nmax}. Memory and time requirements do not depend strongly on this parameter.
+To force a calculation to be axisymmetric, set {\ttfamily nmax} = 0.
 }
@@
 as 5. At low collisionality, this parameter may need to be many 10s or
-even $>$ 100 for convergence.}
+even $>$ 200 for convergence.}
 
@@
 {1e-6}
-{Whenever \parlink{useIterativeLinearSolver} == \true}
+{Always}
 {
@@
 
-\myhrule
-
-\param{forceOddNthetaAndNzeta}
-{Boolean}
-{\true}
-{Always}
-{
-If true, 1 is added to \Ntheta~any time a run is attempted with even \Ntheta,
-and 1 is added to \Nzeta~any time a run is attempted with even \Nzeta.
-When false, the even and odd grid points are effectively decoupled so results are unstable.
-This parameter should be true unless you know what you are doing.
-}
-
 \section{The {\ttfamily otherNumericalParameters}~namelist}
@@
 
-\param{thetaDerivativeScheme}
+\param{FourierOption}
 {integer}
 {2}
-{Always}
-{Discretization scheme for the poloidal angle coordinate theta.\\
-
-{\ttfamily thetaDerivativeScheme} = 0: Fourier spectral collocation.  The differentiation matrix in theta is dense.\\
-
-{\ttfamily thetaDerivativeScheme} = 1: Finite differences with a 3 point stencil.  (The differentiation matrix in theta is tridiagonal, aside from the corners.)\\
-
-{\ttfamily thetaDerivativeScheme} = 2: Finite differences with a 5 point stencil.  (The differentiation matrix in theta is pendadiagonal, aside from the corners.).\\
-
-The best value for this parameter is usually 2.}
-
-\myhrule
-
-\param{zetaDerivativeScheme}
-{integer}
-{2}
-{Always}
-{Discretization scheme for the toroidal angle coordinate zeta.\\
-
-{\ttfamily zetaDerivativeScheme} = 0: Fourier spectral collocation.  The differentiation matrix in zeta is dense.\\
-
-{\ttfamily zetaDerivativeScheme} = 1: Finite differences with a 3 point stencil.  (The differentiation matrix in zeta is tridiagonal, aside from the corners.)\\
-
-{\ttfamily zetaDerivativeScheme} = 2: Finite differences with a 5 point stencil.  (The differentiation matrix in zeta is pendadiagonal, aside from the corners.).\\
-
-The best value for this parameter is usually 2.}
-
-\myhrule
-
-\param{ExBDerivativeSchemeTheta}
-{integer}
-{0}
-{Whenever the radial electric field is nonzero}
-{Options for controlling upwinding of the $\vect{E}\times\vect{B}$ drift terms.  Note the options are different than for \parlink{thetaDerivativeScheme}
-and \parlink{zetaDerivativeScheme}, and are analogous to the options for \parlink{magneticDriftDerivativeScheme}.
-This option exists because when the radial electric field is large enough to be comparable to the resonance,
-such that the coefficient in front of the $\partial f/\partial \theta$ or $\partial f/\partial\zeta$ terms vanishes near the thermal speed of one of the species,
-the distribution function can develop unphysical grid-scale structure.  This option allows upwinding of the $\partial f/\partial\theta$
-spatial derivative, which eliminates the grid-scale oscillations.
-The parameter \parlink{ExBDerivativeSchemeZeta} serves the same function for the $\partial f/\partial\zeta$ derivative.
-\\
-
-{\ttfamily ExBDerivativeSchemeTheta} = 0: Use same differentiation matrices for the $\vect{E}\times\vect{B}$ drift terms as for the parallel streaming term.\\
-
-{\ttfamily ExBDerivativeSchemeTheta} = 1: Treat the $\vect{E}\times\vect{B}$ drift terms using a 4 point upwinded stencil (1 point on 1 side, 2 points on the other.)\\
-
-{\ttfamily ExBDerivativeSchemeTheta} = 2: Treat the $\vect{E}\times\vect{B}$ drift terms using a 5 point upwinded stencil (1 point on 1 side, 3 points on the other.)\\
-
-{\ttfamily ExBDerivativeSchemeTheta} = 3: Treat the $\vect{E}\times\vect{B}$ drift terms using a 6 point upwinded stencil (2 points on 1 side, 3 points on the other.)\\
-
-The different settings require slightly different amounts of memory for factorization due to the different numbers of nonzeros. Option 1 requires the least memory, then 0, then 2,
-and 3 requires the most.  The difference in memory required appears to be small, around 10-20\% between settings 1 and 3.
-On the other hand, the denser the stencil, the more accurate the result. Hence, in order from least to most accurate, the options are 1, 0, 2, 3.
-Option 1 is substantially less accurate than the other options.
-You should use the default {\ttfamily ExBDerivativeSchemeTheta} = 0 except for very large radial electric fields, when the electric field is as large as the resonance or larger.
-In this case, {\ttfamily ExBDerivativeSchemeTheta} = 3 is recommended.
-}
-
-\myhrule
-
-\param{ExBDerivativeSchemeZeta}
-{integer}
-{0}
-{Whenever the radial electric field is nonzero}
-{This parameter behaves exactly like \parlink{ExBDerivativeSchemeTheta}, except it applies to the $\zeta$ coordinate instead of $\theta$.}
-
-\myhrule
-
-\param{magneticDriftDerivativeScheme}
-{integer}
-{3}
-{Whenever \parlink{magneticDriftScheme} is nonzero}
-{Options for controlling upwinding of the magnetic drift terms.  Note the options are different than for \parlink{thetaDerivativeScheme}
-and \parlink{zetaDerivativeScheme}, and are analogous to the options for \parlink{ExBDerivativeSchemeTheta}.
-This option exists because when magnetic drift terms involving $\partial f/\partial \theta$ and $\partial f/\partial\zeta$
-are implemented using the same centered difference stencil as the parallel streaming term,
-the distribution function can develop unphysical grid-scale structure.  This option allows upwinding of these spatial derivatives, which eliminates the grid-scale oscillations.\\
-
-{\ttfamily magneticDriftDerivativeScheme} = 0: Use same differentiation matrices for the magnetic drift terms as for the parallel streaming term.\\
-
-{\ttfamily magneticDriftDerivativeScheme} = 1: Treat the magnetic drift terms using a 4 point upwinded stencil (1 point on 1 side, 2 points on the other.)\\
-
-{\ttfamily magneticDriftDerivativeScheme} = 2: Treat the magnetic drift terms using a 5 point upwinded stencil (1 point on 1 side, 3 points on the other.)\\
-
-{\ttfamily magneticDriftDerivativeScheme} = 3: Treat the magnetic drift terms using a 6 point upwinded stencil (2 points on 1 side, 3 points on the other.)\\
-
-The different settings require slightly different amounts of memory for factorization due to the different numbers of nonzeros. Option 1 requires the least memory, then 0, then 2,
-and 3 requires the most.  The difference in memory required appears to be small, around 10-20\% between settings 1 and 3.
-On the other hand, the denser the stencil, the more accurate the result. 
-Hence, in order from least to most accurate, the options are 1, 0, 2, 3. Option 1 is substantially less accurate than the other options.
-The default setting of 3 is robust and recommended in most cases.
-}
+{Whenever \parlink{nmax}$>$0, i.e. whenever the calculation is nonaxisymmetric.}
+{Option for choosing which spatial Fourier modes will be included.\\
+
+{\ttfamily FourierOption} = 0: Include all Fourier modes satisfying $|n|/${\ttfamily Nperiod}$\le $\parlink{nmax} and $|m| \le$\parlink{mmax}.
+In this case, the input value of \parlink{NFourier} will be ignored, and \parlink{NFourier} will be set to 
+\parlink{mmax}$\times(2\times$\parlink{nmax}$+1)+$\parlink{nmax}$+1$.\\
+
+{\ttfamily FourierOption} = 1: Include the first \parlink{NFourier} modes ranked by $m^2+[n/(${\ttfamily Nperiod}$\times$\parlink{FourierFactor}$)]^2$.\\
+
+{\ttfamily FourierOption} = 2: Include the first \parlink{NFourier} modes ranked by the spectrum
+of $B^\alpha$ for various powers $\alpha$.
+}
+
+\myhrule
+
+\param{FourierFactor}
+{real}
+{1.0}
+{Whenever \parlink{FourierOption}=1 and \parlink{nmax}$>0$.}
+{Scales the aspect ratio of the ellipse in $(m,n)$ space for the spatial Fourier modes to include.
+The modes are ranked by $m^2+[n/(${\ttfamily Nperiod}$\times$\parlink{FourierFactor}$)]^2$.}
 
@@
 
-\param{useIterativeLinearSolver}
-{Boolean}
-{\true}
-{Always}
-{If false, a sparse direct solver
-will be used. The direct solver is faster for small (i.e. low-resolution) problems and always yields a solution
-(as long as there is sufficient memory).
-For large (high resolution) problems, the iterative solver will usually be faster and will use much
- less memory, but it may not always converge.
-}
-
-\myhrule
-
 \param{whichParallelSolverToFactorPreconditioner}
@@
 {1}
-{Whenever \parlink{useIterativeLinearSolver} = \true~and there are 2 or more species.}
+{Whenever there are 2 or more species.}
 {\\
@@
 {1}
-{Whenever \parlink{useIterativeLinearSolver} = \true~and \parlink{RHSMode} = 1 or 2.}
+{Whenever \parlink{RHSMode} = 1 or 2.}
 {\\
@@
 {0}
-{Whenever \parlink{useIterativeLinearSolver} = \true~and \parlink{RHSMode} = 1 or 2 and \parlink{preconditioner\_x} $>$ 0.}
+{Whenever \parlink{RHSMode} = 1 or 2 and \parlink{preconditioner\_x} $>$ 0.}
 {The $x$ structure of the matrix will only be simplified for Legendre index $L$ is $\ge$ this value.
@@
 
-\param{preconditioner\_theta}
+\param{preconditioner\_Fourier}
 {integer}
 {0}
-{Whenever \parlink{useIterativeLinearSolver} = \true}
+{Always}
 {\\
-{\ttfamily preconditioner\_theta} = 0: Keep full $\theta$ coupling.\\
-
-{\ttfamily preconditioner\_theta} = 1: Use a 3-point finite difference stencil for $d/d\theta$.\\
-
-{\ttfamily preconditioner\_theta} = 2: Drop all $\theta$ coupling.\\
-
-{\ttfamily preconditioner\_theta} = 3: Replace $d/d\theta$ with the identity matrix.\\
+{\ttfamily preconditioner\_Fourier} = 0: Keep full coupling in the toroidal and poloidal angles.\\
+
+{\ttfamily preconditioner\_Fourier} = 1: Drop some coupling in the toroidal and poloidal angles.\\
 
@@
 
-\param{preconditioner\_theta\_min\_L}
+\param{preconditioner\_Fourier\_min\_L}
 {integer}
 {0}
-{Whenever \parlink{useIterativeLinearSolver} = \true~and \parlink{preconditioner\_theta} $>$ 0.}
-{The $\theta$ structure of the matrix will only be simplified for Legendre index $L$ is $\ge$ this value.
-Set {\ttfamily preconditioner\_theta\_min\_L} = 0 to simplify the matrix for every $L$.
-}
-
-\myhrule
-
-\param{preconditioner\_zeta}
-{integer}
-{0}
-{Whenever \parlink{useIterativeLinearSolver} = \true}
-{\\
-{\ttfamily preconditioner\_zeta} = 0: Keep full $\zeta$ coupling.\\
-
-{\ttfamily preconditioner\_zeta} = 1: Use a 3-point finite difference stencil for $d/d\zeta$.\\
-
-{\ttfamily preconditioner\_zeta} = 2: Drop all $\zeta$ coupling.\\
-
-{\ttfamily preconditioner\_zeta} = 3: Replace $d/d\zeta$ with the identity matrix.\\
-
-The default value of 0 is strongly recommended.}
-
-\myhrule
-
-\param{preconditioner\_zeta\_min\_L}
-{integer}
-{0}
-{Whenever \parlink{useIterativeLinearSolver} = \true~and \parlink{preconditioner\_zeta} $>$ 0.}
-{The $\zeta$ structure of the matrix will only be simplified for Legendre index $L$ is $\ge$ this value.
-Set {\ttfamily preconditioner\_zeta\_min\_L} = 0 to simplify the matrix for every $L$.
+{Whenever \parlink{preconditioner\_Fourier} $>$ 0.}
+{The Fourier structure of the matrix will only be simplified for Legendre index $L$ is $\ge$ this value.
+Set {\ttfamily preconditioner\_Fourier\_min\_L} = 0 to simplify the matrix for every $L$.
 }
@@
 {1}
-{Whenever \parlink{useIterativeLinearSolver} = \true}
+{Always}
 {\\
@@
 Either a setting of 0 or 1 can be good for this parameter.}
-
-\myhrule
-
-\param{preconditioner\_magnetic\_drifts\_max\_L}
-{integer}
-{2}
-{Whenever \parlink{useIterativeLinearSolver} = \true and \parlink{magneticDriftScheme}$>0$}
-{Maximum Legendre mode number for which the poloidal and toroidal magnetic drift terms are included in the preconditioner.
-As this parameter is increased, more memory is required for factorization, but fewer KSP iterations are required.
-Setting this parameter to $\ge$ \parlink{Nxi} means that the magnetic drifts are always included in the preconditioner.
-Setting this parameter to $<0$ means the magnetic drifts are never included in the preconditioner.
-The default value should be good.}
 

### 2016-08-16 fortran/version4/chooseFourierModes.F90

@@ -4,7 +4,7 @@ subroutine chooseFourierModes()
   use petscsys
   use FourierTransformMod
   use globalVariables, only: NFourier, NFourier2, xm, xn, mmax, nmax, NPeriods, prec, BHat, masterProc, &
-       Ntheta, Nzeta, predictedAmplitudes
+       Ntheta, Nzeta, predictedAmplitudes, FourierOption, FourierFactor
 
   implicit none
 
@@ -44,32 +44,62 @@ subroutine chooseFourierModes()
 
   else
 
-     allocate(xn_sorted(NFourier))
-     allocate(xm_sorted(NFourier))
+     select case (FourierOption)
+     case (0)
+        ! Include all (m,n) such that m<=mmax and |n|<=nmax.
+        ! NFourier should already have been set to mmax*(2*nmax+1)+nmax+1 in validateInput.
+        ! But let's do it again here to be safe:
+        NFourier = mmax*(2*nmax+1)+nmax+1
+        NFourier2 = NFourier*2-1
 
-     ! Temporarily set up (xm,xn) to include all possible modes.
-     NFourier_original = NFourier
-     NFourier = mmax*(2*nmax+1) + nmax + 1
-     NFourier2 = NFourier*2-1
-     allocate(xm(NFourier))
-     allocate(xn(NFourier))
+        allocate(xm(NFourier))
+        allocate(xn(NFourier))
      
-     xm=0
-     xn=0
-     ! Handle the m=0 modes
-     do n=1,nmax
-        xn(n+1)=n
-     end do
-     ! Handle the m>0 modes
-     index = nmax+1
-     do m=1,mmax
-        do n=-nmax,nmax
-           index = index + 1
-           xn(index) = n
-           xm(index) = m
+        xm=0
+        xn=0
+        ! Handle the m=0 modes
+        do n=1,nmax
+           xn(n+1)=n
         end do
-     end do
-     xn = xn*NPeriods
+        ! Handle the m>0 modes
+        index = nmax+1
+        do m=1,mmax
+           do n=-nmax,nmax
+              index = index + 1
+              xn(index) = n
+              xm(index) = m
+           end do
+        end do
+        xn = xn*NPeriods
+
+     case (1,2)
+        ! Pick certain (m,n) pairs to include
+        allocate(xn_sorted(NFourier))
+        allocate(xm_sorted(NFourier))
+
[…]

### 2016-08-16 fortran/version4/globalVariables.F90

@@ -170,6 +170,8 @@ module globalVariables
   !
   ! ********************************************************
 
+  integer :: FourierOption = 2
+  real(prec) :: FourierFactor = 1.0
   real(prec), dimension(:,:,:), allocatable :: FourierAmplitudeVsL
   integer, dimension(:), allocatable :: Nxi_for_x, min_x_for_L, L_to_save
   integer, dimension(:), allocatable :: xm, xn

### 2016-08-16 fortran/version4/readInput.F90

@@ -62,7 +62,7 @@ contains
 
     namelist / otherNumericalParameters /  &
          xGridScheme, whichParallelSolverToFactorPreconditioner, &
-         PETSCPreallocationStrategy, xPotentialsGridScheme, xGrid_k, Nxi_for_x_option
+         PETSCPreallocationStrategy, xPotentialsGridScheme, xGrid_k, Nxi_for_x_option, FourierOption, FourierFactor
 
     namelist / preconditionerOptions / preconditioner_x, preconditioner_x_min_L, &
          preconditioner_Fourier, preconditioner_xi, preconditioner_species, reusePreconditioner, &

### 2016-09-09 doc/manual/version3/inputParameters.tex

@@
 
+\myhrule
+
+\param{Nxi\_for\_x\_option}
+{integer}
+{1}
+{Always}
+{This setting controls how the number of Legendre polynomials depends on the speed $x$ for representing the distribution function.
+You can see the actual number of Legendre polynomials used for each $x$ grid point by examining the {\ttfamily Nxi\_for\_x} variable
+in the output file.\\
+
+{\ttfamily Nxi\_for\_x\_option} = 0: Use the same number (\parlink{Nxi}) of Legendre polynomials for all $x$.  This approach is
+simplest, but it is somewhat inefficient since small values of $x$ represent higher collisionality, so there is weaker pitch-angle dependence
+of the distribution function.\\
+
+{\ttfamily Nxi\_for\_x\_option} = 1: The number of Legendre polynomials will be a linear function of $x$ up to $x=2$,
+at which point \parlink{Nxi} polynomials will be used. For $x>2$, the number of Legendre polynomials used will remain at \parlink{Nxi}.\\
+
+{\ttfamily Nxi\_for\_x\_option} = 2: The number of Legendre polynomials will be a quadratic function of $x$ up to $x=2$,
+at which point \parlink{Nxi} polynomials will be used. For $x>2$, the number of Legendre polynomials used will remain at \parlink{Nxi}.\\
+
+}
+
 \section{The {\ttfamily preconditionerOptions}~namelist}

### 2016-09-15 doc/manual/version3/inputParameters.tex
added more magnetic drift options

@@
 
-{\ttfamily magneticDriftScheme} = 2: Use the grad-B and curvature drift, plus the parallel velocity correction $v_{\perp}^2/(2 \Omega_c) \vect{b} \vect{b} \cdot \nabla \times \vect{b}$.
+{\ttfamily magneticDriftScheme} = 2: Use the grad-B and curvature
+drift, plus the parallel velocity correction $v_{\perp}^2/(2 \Omega_c)
+\vect{b} \vect{b} \cdot \nabla \times \vect{b}$.\\
+
+{\ttfamily magneticDriftScheme} = 3: Magnetic drift as defined in
+Eq.~(67) in \cite{Kasilov2014}.\\
+
+{\ttfamily magneticDriftScheme} = 4: Magnetic drift as defined in
+Eq.~(16) in \cite{Martitsch2016}. (This only works for .bc magnetic
+geometry input files so far)\\
+
 }

### 2016-09-19 doc/manual/version3/inputParameters.tex
adding more magnetic drift options

@@
 {\ttfamily magneticDriftScheme} = 4: Magnetic drift as defined in
-Eq.~(16) in \cite{Martitsch2016}. (This only works for .bc magnetic
+Eq.~(16) in \cite{Martitsch2016}.\\ (This {\ttfamily
+  magneticDriftScheme} only works for .bc magnetic
+geometry input files so far)\\
+
+{\ttfamily magneticDriftScheme} = 5: Magnetic drift as defined in
+Eqs.~(46--48) in \cite{Sugama2016}. The function $\alpha$
+ is chosen as $\alpha=(1-\xi^2)/(1+\xi^2)$ for the terms
+in Eq.~(48) proportional to the normal curvature, which are the terms
+that need
+to be regularized at $\xi=\pm 1$.\\
+(This {\ttfamily magneticDriftScheme} only works for .bc magnetic
 geometry input files so far)\\

### 2016-09-19 fortran/version3/validateInput.F90

@@ -460,11 +460,20 @@ subroutine validateInput()
      stop
   end if
 
-  if (magneticDriftScheme>4) then
+  if (magneticDriftScheme>5) then
      if (masterProc) then
-        print *,"Error! magneticDriftScheme must be <= 4."
+        print *,"Error! magneticDriftScheme must be <= 5."
      end if
      stop
+  end if
+
+  if (magneticDriftScheme>2) then
+     if (.not.( (geometryScheme == 11) .or. (geometryScheme == 12) )) then
+        if (masterProc) then
+           print *,"Error! magneticDriftScheme 3, 4 and 5 have only been implemented for geometryScheme 11 and 12."
+        end if
+        stop
        
+     end if
   end if
 
   if (magneticDriftScheme>0 .and. includePhi1) then

### 2017-01-20 doc/manual/version3/inputParameters.tex
mentioned that radius interpolation is possible also for .bc files

@@
 {1}
-{Only when \parlink{geometryScheme} == 5.}
-{Controls whether the nearest available flux surface in the {\ttfamily vmec wout} file is used, or whether radial interpolation is applied to the \vmec~data
+{Only when \parlink{geometryScheme} == 5, 11 or 12.}
+{Controls whether the nearest available flux surface in the {\ttfamily
+    vmec wout} file or .bc file is used, or whether radial interpolation is applied to the data
 to obtain the magnetic field components on the exact surface requested.\\
 
-{\ttfamily VMECRadialOption}=0: Use the exact {\ttfamily XXX\_wish} flux surface requested, by interpolating from the \vmec~radial grid.\\
-
-{\ttfamily VMECRadialOption}=1: Use a surface that may be slightly different from {\ttfamily XXX\_wish} to get the nearest available flux surface from \vmec's HALF grid.  
+{\ttfamily VMECRadialOption}=0: Use the exact {\ttfamily XXX\_wish} flux surface requested, by interpolating from the radial grid.\\
+
+{\ttfamily VMECRadialOption}=1: Use a surface that may be slightly
+different from {\ttfamily XXX\_wish} to get the nearest available flux
+surface from \vmec's HALF grid (or the .bc file grid).  
    The components of $\vect{B}$ in \vmec~are stored on the half grid, so interpolation is then unnecessary.\\

### 2017-01-20 doc/manual/version3/inputParameters.tex
small update

@@
 {Controls whether the nearest available flux surface in the {\ttfamily
-    vmec wout} file or .bc file is used, or whether radial interpolation is applied to the data
+    vmec wout} file or {\ttfamily .bc} file is used, or whether radial interpolation is applied to the data
 to obtain the magnetic field components on the exact surface requested.\\
@@
 different from {\ttfamily XXX\_wish} to get the nearest available flux
-surface from \vmec's HALF grid (or the .bc file grid).  
+surface from \vmec's HALF grid (or the {\ttfamily .bc} file grid).  
    The components of $\vect{B}$ in \vmec~are stored on the half grid, so interpolation is then unnecessary.\\

### 2017-02-02 doc/manual/version3/inputParameters.tex
added xDotDerivativeScheme to manual

--- 2016-09-19 doc/manual/version3/inputParameters.tex
@@
 
+\param{xDotDerivativeScheme}
+{integer}
+{0}
+{Whenever \parlink{includeXDotTerm} is \true.}
+{Changes the differentiation matrix used for the collisionless $\dot{x} \propto E_r$ term. This parameter
+does not affect the differentiation matrix used for the energy scattering part of the collision operator.\\
+
+{\ttfamily xDotDerivativeScheme} = 0: Use the same differentiation matrix for the $\dot{x}$ term as for the collision operator.\\
+
+{\ttfamily xDotDerivativeScheme} = 11: Use a low-order tri-diagonal differentiation matrix for the $\dot{x}$ term.\\
+
+Other values of {\ttfamily xDotDerivativeScheme} have been implemented (see {\ttfamily createGrids.F90}) but are
+not recommended. The recommended setting is almost always {\ttfamily xDotDerivativeScheme} = 0.
+The one exception is sometimes when \parlink{magneticDriftScheme}$>0$, in which
+case the convergence with respect to \parlink{Nx} is sometimes poor for values of $E_r$ near the peak in the radial fluxes.
+In this case, {\ttfamily xDotDerivativeScheme} = 11 may resolve the problem.
+}
+
+\myhrule
+
 \param{useIterativeLinearSolver}

### 2017-04-08 doc/manual/version3/inputParameters.tex
Minor improvements to magnetic drifts

--- 2017-01-20 doc/manual/version3/inputParameters.tex
@@
 to be regularized at $\xi=\pm 1$.\\
-(This {\ttfamily magneticDriftScheme} only works for .bc magnetic
-geometry input files so far)\\
+
+{\ttfamily magneticDriftScheme} = 6: Same as {\ttfamily magneticDriftScheme} 5,
+except that no regularization is applied to the magnetic drift terms multiplying $\partial f/\partial \theta$
+and $\partial f/\partial\zeta$. Regularization is still applied to the magnetic drift term multiplying $\partial f/\partial \xi$.\\
+
+{\ttfamily magneticDriftScheme} = 7: Same as {\ttfamily magneticDriftScheme} 1,
+except the contribution to $\dot{\xi}$ from the magnetic drifts is dropped.\\
+
+{\ttfamily magneticDriftScheme} = 8: Same as {\ttfamily magneticDriftScheme} 1,
+except the magnetic shear term for axisymmetry is added to the coefficient of $\partial f/\partial \zeta$.
+There is no change to the coefficient of $\partial f/\partial\theta.$ This option only makes sense for straight field line coordinates
+and nearly axisymmetric geometry.\\
+
+{\ttfamily magneticDriftScheme} = 9: Same as {\ttfamily magneticDriftScheme} 1,
+except the magnetic shear term for axisymmetry is added to the coefficient of $\partial f/\partial \theta$.
+There is no change to the coefficient of $\partial f/\partial\zeta.$ This option only makes sense for straight field line coordinates
+and nearly axisymmetric geometry.\\
 
@@
 The recommended setting is {\ttfamily xPotentialsGridScheme} = 2.}
+
+\myhrule
+
+\param{xDotDerivativeScheme}
+{integer}
+{0}
+{Whenever \parlink{includeXDotTerm} is \true.}
+{Changes the differentiation matrix used for the collisionless $\dot{x} \propto E_r$ term. This parameter
+does not affect the differentiation matrix used for the energy scattering part of the collision operator.\\
+
+{\ttfamily xDotDerivativeScheme} = 0: Use the same differentiation matrix for the $\dot{x}$ term as for the collision operator.\\
+
+{\ttfamily xDotDerivativeScheme} = 11: Use a low-order tri-diagonal differentiation matrix for the $\dot{x}$ term.\\
+
+Other values of {\ttfamily xDotDerivativeScheme} have been implemented (see {\ttfamily createGrids.F90}) but are
+not recommended. The recommended setting is almost always {\ttfamily xDotDerivativeScheme} = 0.
+The one exception is sometimes when \parlink{magneticDriftScheme}$>0$, in which
+case the convergence with respect to \parlink{Nx} is sometimes poor for values of $E_r$ near the peak in the radial fluxes.
+In this case, {\ttfamily xDotDerivativeScheme} = 11 may resolve the problem.
+}
 

### 2017-04-08 fortran/version3/populateMatrix.F90

@@ -517,7 +517,7 @@
              do izeta = izetaMin, izetaMax                
                 do ithetaRow = ithetaMin, ithetaMax
                    select case (magneticDriftScheme)
-                   case (1,2)
+                   case (1,2,7,8)
                       geometricFactor1 = (BHat_sub_zeta(ithetaRow,izeta)*dBHatdpsiHat(ithetaRow,izeta) &
                       - BHat_sub_psi(ithetaRow,izeta)*dBHatdzeta(ithetaRow,izeta))
 
                     
@@ -532,6 +532,14 @@
                    case (6)
                       geometricFactor1 = BHat_sub_zeta(ithetaRow,izeta) * gradpsidotgradB_overgpsipsi(ithetaRow,izeta)
                       geometricFactor2 = BHat_sub_zeta(ithetaRow,izeta) * 2.0 * pPrimeHat / BHat(ithetaRow,izeta) !unregularized
+                   case (9)
+                      geometricFactor1 = (BHat_sub_zeta(ithetaRow,izeta)*dBHatdpsiHat(ithetaRow,izeta) &
+                      - BHat_sub_psi(ithetaRow,izeta)*dBHatdzeta(ithetaRow,izeta))
+
                     
+                      geometricFactor2 = 2.0 * BHat(ithetaRow,izeta) &
+                      * (dBHat_sub_psi_dzeta(ithetaRow,izeta) - dBHat_sub_zeta_dpsiHat(ithetaRow,izeta) &
+                      + diotadpsiHat*GHat*BHat_sup_theta(itheta,izetaRow)/(DHat(itheta,izetaRow)*iota*iota) )
+                      ! Need to add shear term here!
                    case default
                       stop "Invalid magneticDriftScheme in d/dtheta term"
                    end select
@@ -637,7 +645,7 @@
                 do izetaRow = izetaMin, izetaMax
 
                    select case (magneticDriftScheme)
-                   case (1,2)
+                   case (1,2,7,9)
                       geometricFactor1 = (BHat_sub_psi(itheta,izetaRow)*dBHatdtheta(itheta,izetaRow) &
                            - BHat_sub_theta(itheta,izetaRow)*dBHatdpsiHat(itheta,izetaRow))
                       
@@ -652,7 +660,8 @@
                       
                       geometricFactor2 = 2.0 * BHat(itheta,izetaRow) &
[…]

### 2017-04-08 fortran/version3/readVMEC/read_NEMEC_file.f90

@@ -174,6 +174,7 @@
       allocate(hoverr(vmec%ns),fjcuru(vmec%ns),fjcurv(vmec%ns))         
       allocate(hspecw(vmec%ns))
       allocate(vmec%phips(vmec%ns))
+      allocate(vmec%presf(vmec%ns))
 
       !allocate(fsve(0:nsin),hsve(1:nsin))
 
@@ -332,7 +333,13 @@
       vmec%bsubsmns(:,:,1) = hbsubsmns(:,:,2)*1.5_DP - hbsubsmns(:,:,3)*0.5_DP
       vmec%bsubsmnc(:,:,vmec%ns) = hbsubsmnc(:,:,vmec%ns)*1.5_DP - hbsubsmnc(:,:,vmec%ns-1)*0.5_DP
       vmec%bsubsmns(:,:,vmec%ns) = hbsubsmns(:,:,vmec%ns)*1.5_DP - hbsubsmns(:,:,vmec%ns-1)*0.5_DP
+
+      ! Convert pressure from the half to full mesh
+      vmec%presf(2:vmec%ns-1) = (hpres(2:vmec%ns-1) + hpres(3:vmec%ns)) / 2
+      vmec%presf(1) = hpres(2)*1.5_DP - hpres(3)*0.5_DP
+      vmec%presf(vmec%ns) = hpres(vmec%ns)*1.5_DP - hpres(vmec%ns-1)*0.5_DP
       
+
     end subroutine read_nemec_file
 
 !-----------------------------------------------------------------------------!

### 2017-04-08 fortran/version3/validateInput.F90

@@ -460,9 +460,9 @@ subroutine validateInput()
      stop
   end if
 
-  if (magneticDriftScheme>6) then
+  if (magneticDriftScheme>9) then
      if (masterProc) then
-        print *,"Error! magneticDriftScheme must be <= 6."
+        print *,"Error! magneticDriftScheme must be <= 9."
      end if
      stop
   end if
@@ -477,9 +477,9 @@ subroutine validateInput()
   end if
 
   if (magneticDriftScheme == 5 .or. magneticDriftScheme == 6) then
-     if (.not.( (geometryScheme == 11) .or. (geometryScheme == 12) .or. (geometryScheme == 5) )) then
+     if (.not.( (geometryScheme == 11) .or. (geometryScheme == 12) .or. (geometryScheme == 5) .or. (geometryScheme == 6))) then
         if (masterProc) then
-           print *,"Error! magneticDriftSchemes 5 and 6 have only been implemented for geometryScheme 5, 11, and 12."
+           print *,"Error! magneticDriftSchemes 5 and 6 have only been implemented for geometryScheme 5, 6, 11, and 12."
         end if
         stop
        
      end if

### 2017-07-13 doc/manual/version3/inputParameters.tex
Switched from readVMEC to mini_libstell

@@
 I'm not sure why you would want this, but the feature is implemented for completeness.
+}
+
+\myhrule
+
+\param{VMEC\_Nyquist\_option}
+{integer}
+{1}
+{Only when \parlink{geometryScheme} == 5.}
+{In VMEC, some quantities are known at poloidal and toroidal mode numbers given by the arrays {\ttfamily xm} and
+{\ttfamily xn}, whereas other quantities are known at a larger range of mode numbers {\ttfamily xm\_nyq} and {\ttfamily xn\_nyq}.
+The sfincs parameter {\ttfamily VMEC\_Nyquist\_option}
+controls whether or not to include mode numbers that exist in the {\ttfamily xm\_nyq} and {\ttfamily xn\_nyq}
+arrays but not in the {\ttfamily xm} and {\ttfamily xn} arrays.
+This this parameter only affects high-frequency components of the magnetic field, results of sfincs
+should not be very sensitive to this parameter.
+\\
+
+{\ttfamily VMEC\_Nyquist\_option}=1: Only include poloidal and toroidal mode numbers that exist
+in the {\ttfamily xm} and {\ttfamily xn} arrays.  This option was always used prior to July 2017.\\
+
+{\ttfamily VMEC\_Nyquist\_option}=2: Also include the larger poloidal and toroidal mode numbers that exist
+in the {\ttfamily xm\_nyq} and {\ttfamily xn\_nyq} arrays, where available.
 }

### 2017-07-13 fortran/version3/readVMEC/handle_err.f90

@@ -1,11 +1,0 @@
-subroutine ezcdf_error(status,nam3,nam1,nam2)
-  include "netcdf.inc"
-  INTEGER, intent(in) :: status
-  character*(*), intent(in) :: nam1, nam2, nam3
-  if (status .ne. nf_noerr) then
-     WRITE(*,10) nam1,nam2,nam3
-10   format('% ',a,'--E-- A netCDF error has occurred in: ' ,a/      &
-          &          'while processing: ',a)
-     print *, nf_strerror(status)
-  endif
-end subroutine ezcdf_error

### 2017-07-13 fortran/version3/validateInput.F90

@@ -214,6 +214,11 @@ subroutine validateInput()
      print *,line
   end if
 
+  if (VMEC_Nyquist_option < 1 .or. VMEC_Nyquist_option > 2) then
+     if (masterProc) print *,"Error! VMEC_Nyquist_option must be either 1 or 2."
+     stop
+  end if
+
   ! species namelist:
 
   flag = .false.

### 2017-07-24 doc/manual/version3/inputParameters.tex
Initial commit. Added sensitivity inputs to manual, globalVariables, readInput, and validateInput.

--- 2017-05-03 doc/manual/version3/inputParameters.tex
@@
 
+\section{The \texttt{sensitivityOptions} namelist}
+
+\param{sensitivityOption}
+{integer}
+{0}
+{When {\ttfamily geometryScheme == 5} and {\ttfamily includePhi1 == .false.}}. \\
+{\ttfamily sensitivityOption} == 0: No sensitivity output computed. \\
+{\ttfamily sensitivityOption} == 1: Sensitivity computed for VMEC harmonics using adjoint equation.
+The quantities of interest for sensitivity computations are specified by {\ttfamily AdjointRHSOption} and {\ttfamily AdjointRHSSpeciesOption}.
+\\
+
+\myhrule
+
+\param{AdjointRHSOption}
+{integer}
+{0}
+{When {\ttfamily geometryScheme == 5} and {\ttfamily sensitivityOption == 1}} \\
+{\ttfamily AdjointRHSOption == 0}: Sensitivity is computed for bootstrap current, particle flux, and heat flux. \\
+{\ttfamily AdjointRHSOption == 1}: Sensitivity is computed for just bootstrap current. \\
+{\ttfamily AdjointRHSOption == 2}: Sensitivity is computed for just particle flux. \\
+{\ttfamily AdjointRHSOption == 3}: Sensitivity is computed for just heat flux.
+
+\myhrule
+
+\param{AdjointRHSSpeciesOption}
+{integer}
+{-1}
+{When {\ttfamily geometryScheme == 5} and {\ttfamily sensitivityOption == 1} and {\ttfamily AdjointRHSOption /= 0}} \\
+{\ttfamily AdjointRHSSpeciesOption == -1}: Sensitivity computed for output quantities specified by {\ttfamily AdjointRHSOption} for all species. \\
+{\ttfamily AdjointRHSSpeciesOption >= 0}: Sensitivity computed for output quantities specified by {\ttfamily AdjointRHSOption} for species {\ttfamily AdjointRHSSpeciesOption}. \\
+
+\myhrule
+
 \section{\PETSc~commands}

### 2017-07-24 fortran/version3/geometry.F90

@@ -95,6 +95,15 @@ contains
        NPeriods = vmec%nfp
        psiAHat = vmec%phi(vmec%ns)/(2*pi)
        aHat = vmec%Aminor_p
+       ! Check that sensitivity is computed with stellarator symmetry
+       if (sensitivityOption > 1) then
+          if (vmec%iasym > 0) then
+            if (masterProc) then
+              print *,"Error! sensitivityOption>0 must be used with VMEC geometry."
+            endif
+            stop
+          endif
+       endif
 
     case (6)
        ! Read Erika Strumberger's NEMEC format used at IPP.

### 2017-07-24 fortran/version3/readInput.F90

@@ -48,7 +48,8 @@ contains
 
     namelist / physicsParameters / Delta, alpha, nu_n, EParallelHat, EParallelHatSpec, & !!EParallelHatSpec added by HM 2017-02
          collisionOperator, constraintScheme, includeXDotTerm, &
-!!         includeElectricFieldTermInXiDot, useDKESExBDrift, include_fDivVE_term, nonlinear, &
 !!Commented by AM 2016-02
+!!         includeElectricFieldTermInXiDot, useDKESExBDrift, include_fDivVE_term, nonlinear, &
+ !!Commented by AM 2016-02
          includeElectricFieldTermInXiDot, useDKESExBDrift, include_fDivVE_term, & !!Added by AM 2016-02
          dPhiHatdpsiHat, dPhiHatdpsiN, dPhiHatdrHat, dPhiHatdrN, Er, &
          includeTemperatureEquilibrationTerm, includePhi1, &
@@ -84,6 +85,9 @@ contains
     namelist / export_f / export_full_f, export_delta_f, export_f_theta, export_f_zeta, export_f_x, export_f_xi, &
          export_f_theta_option, export_f_zeta_option, export_f_xi_option, export_f_x_option
 
+    namelist / sensitivityOptions / sensitivityOption, adjointRHSOption, &
+      adjointRHSSpeciesOption
+
     Zs = speciesNotInitialized
     mHats = speciesNotInitialized
     nHats = speciesNotInitialized
@@ -96,7 +100,8 @@ contains
     dTHatdrHats = speciesNotInitialized
     dNHatdrNs = speciesNotInitialized
     dTHatdrNs = speciesNotInitialized
-
	
+
+	
 
     export_f_theta = speciesNotInitialized
     export_f_zeta = speciesNotInitialized

### 2017-07-24 fortran/version3/validateInput.F90

@@ -35,7 +35,8 @@ subroutine validateInput()
      stop
   end if
   
-  !!if (RHSMode == 2 .and. nonlinear) then
 !!Commented by AM 2016-02
+  !!if (RHSMode == 2 .and. nonlinear) then
+ !!Commented by AM 2016-02
   if (RHSMode == 2 .and. includePhi1) then !!Added by AM 2016-02
      if (masterProc) then
         print *,"Error! RHSMode cannot be 2 for a nonlinear calculation."
@@ -61,19 +62,23 @@ subroutine validateInput()
         Nxi_for_x_option=0
      end if
 
-     !!if (nonlinear) then
 !!Commented by AM 2016-02
+     !!if (nonlinear) then
+ !!Commented by AM 2016-02
      if (includePhi1) then !!Added by AM 2016-02
         if (masterProc) then
            print *,line
            print *,line
-!!           print *,"**   WARNING: You asked for RHSMode=3 (monoenergetic transport matrix) with nonlinear = .true., which is incompatble."
 !!Commented by AM 2016-02
-!!           print *,"**            Setting nonlinear = .false."
 !!Commented by AM 2016-02
+!!           print *,"**   WARNING: You asked for RHSMode=3 (monoenergetic transport matrix) with nonlinear = .true., which is incompatble."
+ !!Commented by AM 2016-02
+!!           print *,"**            Setting nonlinear = .false."
+ !!Commented by AM 2016-02
            print *,"**   WARNING: You asked for RHSMode=3 (monoenergetic transport matrix) with includePhi1 = .true., which is incompatble." !!Added by AM 2016-02
            print *,"**            Setting includePhi1 = .false." !!Added by AM 2016-02
            print *,line
            print *,line
         end if
-!!        nonlinear = .false.
 !!Commented by AM 2016-02
+!!        nonlinear = .false.
+ !!Commented by AM 2016-02
         includePhi1 = .false. !!Added by AM 2016-02
      end if
 
@@ -443,9 +448,12 @@ subroutine validateInput()
 !!$!!  if (includeRadialExBDrive .and. (.not. includePhi1)) then !!Commented by AM 2016-03
 !!$  if (includePhi1InKineticEquation .and. (.not. includePhi1)) then !!Added by AM 2016-03
 !!$     if (masterProc) then
-!!$!!        print *,"Error! You requested a calculation including the radial ExB drive term"
 !!Commented by AM 2016-03
-!!$!!        print *,"(includeRadialExBDrive=.true.) but you set includePhi1=.false."
[…]

### 2017-07-26 doc/adjointDocs/Adjoint SFINCS Manual.tex
Started adjoint docs.

\documentclass[11pt]{amsart}
\usepackage{geometry}                
\geometry{letterpaper}                   
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{placeins}
\usepackage{setspace}
\onehalfspacing
\usepackage{fullpage}
\usepackage{commath}
\usepackage{bm}
\usepackage{breqn}
\usepackage{bm}
\usepackage{mathrsfs}
% My definitions
\def\bhat{\ensuremath{\mathbf{\hat{b}}}}
\def\vpar{\ensuremath{v_{||}}}
\def\vperp{\ensuremath{v_{\perp}}}
\newcommand{\partder}[2]{\dfrac{\partial #1}{\partial #2}} % partial der.
\newcommand{\der}[2]{\dfrac{d #1}{d #2}} % partial der.
\DeclareMathOperator{\sign}{sign}
\DeclareMathOperator{\Res}{Res}
% Definitions for code snipets
\usepackage{listings}
\usepackage{color}
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\lstset{frame=tb,
  language=Matlab,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=none,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=true,
  tabsize=3
}
\begin{document}
\author{Elizabeth Paul}
\date{July 25th, 2017}
\title{Manual for Adjoint SFINCS}
\maketitle

\section{Input Arguments}
As described in the \texttt{SFINCSUserManual}, the sensitivity will be computed when \texttt{sensitivityOption > 1}. There is also a flag to determine the desired fluxes to compute sensitivity of, \texttt{adjointRHSOption}, and for which species the sensitivity of the fluxes will be computed, \texttt{adjointRHSSpeciesOption}. Adjoint SFINCS must also be run with the following input parameters. 
\begin{itemize}
\item \texttt{geometryScheme == 5} (must also be stellarator symmetric)
\item \texttt{includePhi1 == .false.}
\item \texttt{EParallelHat == 0}
\item \texttt{useDKESExBDrift == .false.} 
\item \texttt{magneticDriftScheme == 0}
%\item \texttt{force0RadialCurrentInEquilibrium = .true.}
\item \texttt{includeXDotTerm == .true.} 
\item \texttt{constraintScheme == 1}
\item \texttt{collisionOperator == 0}
\end{itemize}

\section{Computing Geometrical Factors}
We need to compute the derivatives of the components of $\bm{B}$ $\left( \hat{B}, \hat{B}^{\theta}, \hat{B}^{\zeta}, \hat{B}_{\theta}, \hat{B}_{\zeta}, \hat{D}, \partder{\hat{B}}{\theta}, \partder{\hat{B}}{\zeta} \right)$, with respect to their Fourier components ($\hat{B}_{m,n}$, $\hat{B}_{m,n}^{\theta}$, $\hat{B}_{m,n}^{\zeta}$, $\hat{B}^{m,n}_{\theta}$, $\hat{B}^{m,n}_{\zeta}$, $\hat{g}_{m,n}$). The arrays \texttt{dBHatdFourier}, \texttt{dBHat_sub_thetadFourier}, \texttt{dBHat_sub_zetadFourier}, \texttt{dBHat_sup_thetadFourier}, \texttt{dBHat_sup_zetadFourier}, \texttt{dDHatdFourier}, \texttt{dBHatdthetadFourier}, and \texttt{dBHatdzetadFourier} contain these derivatives. These arrays are allocated and populated in \texttt{computeBHat_VMEC}. I also allocate and populate arrays \texttt{ns} and \texttt{ms} which keep track of which modes are kept in the Fourier spectrum.

\section{Computing Inner Products}
The subroutines \texttt{populatedMatrixdLambda} and \texttt{populatedRHSdLambda} populate $\partder{\mathbb{L}}{\lambda}$ and $\partder{\mathbb{S}}{\lambda}$ for a given Fourier component ($\hat{B}, \hat{B}^{\theta}, \hat{B}^{\zeta}, \hat{B}_{\theta}, \hat{B}_{\zeta}, \hat{D}$) and mode number. These will be called by a subroutine which performs an inner product.

\end{document}

### 2017-07-30 fortran/version3/populateMatrix.F90
populateMatrix modified to populate adjoint jacobian and preconditioner. Corresponding discussion added to adjoint docs.

@@ -103,9 +103,15 @@
        ! This matrix is quite similar to the Jacobian matrix (whichMatrix=1), since most terms in the system of equations are linear.
        ! However there are a few differences related to the nonlinear terms.
        whichMatrixName = "residual f1"
+    case (4)
+       whichMatrixName = "adjoint Jacobian preconditioner"
+       ! The preconditioner for the matrix that is used for the adjoint solve for sensitivity computations
+    case (5)
+       whichMatrixName = "adjoint Jacobian"
+       ! The matrix that is used for the adjoint solve for sensitivity computations
     case default
        if (masterProc) then
-          print *,"Error! whichMatrix must be 0, 1, 2, or 3."
+          print *,"Error! whichMatrix must be 0, 1, 2, 3, 4, or 5."
        end if
        stop
     end select
@@ -124,7 +130,7 @@
     ! (unlike mumps or superlu_dist), then if we're using this solver
     ! add some values to the diagonals of the preconditioner.  By trial-and-error, I found it works
     ! best to shift the diagonal of the quasineutrality and constraint blocks but not for the kinetic-equation block.
-    if ((.not. isAParallelDirectSolverInstalled) .and. masterProc .and. whichMatrix==0) then
+    if ((.not. isAParallelDirectSolverInstalled) .and. masterProc .and. ((whichMatrix==0) .or. (whichMatrix == 5))) then
        print *,"Since PETSc's built-in solver is being used instead of superlu_dist or mumps, and this"
        print *,"   fragile solver often gives spurious error messages about zero pivots, the diagonal"
        print *,"   of the preconditioner is being shifted."
@@ -164,7 +170,8 @@
        end if
     end if
 
-!!    useStateVec = (nonlinear .and. (whichMatrix==0 .or. whichMatrix==1))
 !!Commented by AM 2016-02
+!!    useStateVec = (nonlinear .and. (whichMatrix==0 .or. whichMatrix==1))
+ !!Commented by AM 2016-02
     useStateVec = (includePhi1 .and. (whichMatrix==0 .or. whichMatrix==1)) !!Added by AM 2016-02
     if (useStateVec) then
        ! We need delta f to evaluate the Jacobian, so send a copy to every proc:
[…]

### 2017-07-31 doc/manual/version3/inputParameters.tex
Edited input variables to logicals. Made associated edits to manual and writehdf5 including a writeHDF5Booleans procedure

--- 2017-05-03 doc/manual/version3/inputParameters.tex
@@
 
-{\ttfamily RHSMode} = 3: Solve for the 2$\times$2 monoenergetic transport coefficients. When this option is chosen, \Nx~is set to 1 and only 1 species is used.
+{\ttfamily RHSMode} = 3: Solve for the 2$\times$2 monoenergetic transport coefficients. When this option is chosen, \Nx~is set to 1 and only 1 species is used. \\
+
+{\ttfamily RHSMode} = 4: Solve the adjoint equation at constant $E_r$.
+This must be used with {\ttfamily geometryScheme == 5}, stellarator symmetric geometry, {\ttfamily includePhi1 = .false.}, {\ttfamily EParallelHat == 0}, {\ttfamily useDKESExBDrift == .false.}, {\ttfamily magneticDriftScheme == 0}, {\ttfamily includeXDotTerm == .true.}, {\ttfamily constraintScheme == 1}, and {\ttfamily collisionOperator == 0}.
+\\
+
+{\ttfamily RHSMode} = 5: Solve the adjoint equation at the ambipolar $E_r$. The same input parameters must be used as those detailed for {\ttfamily RHSMode == 4}. 
 }
@@
 
+\section{The \texttt{sensitivityOptions} namelist}
+
+\param{adjointBootstrapOption}
+{Boolean}
+{{\ttfamily .false. }}
+{When {\ttfamily RHSMode == 4} or {\ttfamily RHSMode == 5}. }
+{If this parameter is {\ttfamily .true.}, the sensitivity of the bootstrap current will be computed.}
+
+\myhrule
+
+\param{adjointRadialCurrentOption}
+{Boolean}
+{{\ttfamily .false. }}
+{When {\ttfamily RHSMode == 4} or {\ttfamily RHSMode == 5}. }
+{If this parameter is {\ttfamily .true.}, the sensitivity of the radial current will be computed.}
+
+\myhrule
+
+\param{adjointTotalHeatFluxOption}
+{Boolean}
+{{\ttfamily .false. }}
+{When {\ttfamily RHSMode == 4} or {\ttfamily RHSMode == 5}. }
+{If this parameter is {\ttfamily .true.}, the sensitivity of the species-summed heat flux will be computed.}
+
+\myhrule
+
+\param{adjointHeatFluxOption}
+{1D array of booleans}
+{0}
+{When {\ttfamily RHSMode == 4} or {\ttfamily RHSMode == 5}} 
+{This array is of the same length of {\ttfamily nHats} and other species parameters. If parameter is {\ttfamily .true.}, the sensitivity of the heat flux will be computed for this species.}
+
+\myhrule
+
+\param{adjointParticleFluxOption}
+{1D array of booleans}
+{0}
+{When {\ttfamily RHSMode == 4} or {\ttfamily RHSMode == 5}}
+{This array is of the same length of {\ttfamily nHats} and other species parameters. If parameter is {\ttfamily .true.}, the sensitivity of the particle flux will be computed for this species.}
+
 \section{\PETSc~commands}

### 2017-07-31 fortran/version3/populateAdjointRHS.f90

@@ -1,0 +1,8 @@
+
+subroutine populateAdjointRHS(rhs, whichRHS)
+  integer, intent(in) ::
+  Vec :: stateVec
+
+
+
+end subroutine populateAdjointRHS

### 2017-07-31 fortran/version3/populateMatrix.F90

@@ -104,11 +104,13 @@
        ! However there are a few differences related to the nonlinear terms.
        whichMatrixName = "residual f1"
     case (4)
+      whichMatrixName = "adjoint Jacobian"
+      ! The matrix that is used for the adjoint solve for sensitivity computations
+      ! This is the same matrix used to evaluate the residual for the adjoint solve
+    case (5)
        whichMatrixName = "adjoint Jacobian preconditioner"
        ! The preconditioner for the matrix that is used for the adjoint solve for sensitivity computations
-    case (5)
-       whichMatrixName = "adjoint Jacobian"
-       ! The matrix that is used for the adjoint solve for sensitivity computations
+
     case default
        if (masterProc) then
           print *,"Error! whichMatrix must be 0, 1, 2, 3, 4, or 5."
@@ -337,7 +339,7 @@
 
              do itheta=ithetaMin, ithetaMax
                 do izeta=1,Nzeta
-                   if ((whichMatrix == 4) .or (whichMatrix == 5)) then
+                   if (whichMatrix == 4 .or. whichMatrix == 5) then
                       zetaPartOfTerm(izeta,:) = -sqrtTHat/sqrtMHat * BHat_sup_zeta(itheta,izeta) &
                         * ddzetaToUse(izeta,:) / BHat(itheta,izeta)
                    else
@@ -401,7 +403,7 @@
           do L=0,(Nxi-1)
 
              if (ExBDerivativeSchemeTheta==0) then
-                if (((whichMatrix>0) .and (whichMatrix .ne. 5)) .or. L < preconditioner_theta_min_L) then
+                if ((whichMatrix>0 .and. whichMatrix .ne. 5) .or. L < preconditioner_theta_min_L) then
                    ddthetaToUse = ddtheta
                 else
                    ddthetaToUse = ddtheta_preconditioner
@@ -655,7 +657,7 @@
        izeta = -1  ! So izeta is not used in place of izetaRow or izetaCol by mistake.
        ithetaRow = -1 ! So ithetaRow is not used in place of itheta by mistake.
        ithetaCol = -1 ! So ithetaCol is not used in place of itheta by mistake.
-       if ((whichMatrix .ne. 2) .and. (magneticDriftScheme>0) .and. (whichMarix .ne. 4)) then
+       if ((whichMatrix .ne. 2) .and. (magneticDriftScheme>0) .and. (whichMatrix .ne. 4)) then
[…]

### 2017-07-31 fortran/version3/readInput.F90

@@ -19,6 +19,7 @@ contains
     integer :: NDNHatdpsiNs,   NDTHatdpsiNs
     integer :: NDNHatdrHats,   NDTHatdrHats
     integer :: NDNHatdrNs,     NDTHatdrNs
+    !logical, dimension(maxNumSpecies) :: adjointHeatFluxOptionCopy, adjointParticleFluxOptionCopy
 
     namelist / general / saveMatlabOutput, MatlabOutputFilename, &
          outputFilename, solveSystem, RHSMode, &
@@ -85,8 +86,8 @@ contains
     namelist / export_f / export_full_f, export_delta_f, export_f_theta, export_f_zeta, export_f_x, export_f_xi, &
          export_f_theta_option, export_f_zeta_option, export_f_xi_option, export_f_x_option
 
-    namelist / sensitivityOptions / sensitivityOption, adjointRHSOption, &
-      adjointRHSSpeciesOption
+    namelist / sensitivityOptions / adjointBootstrapOption, adjointRadialCurrentOption, &
+      adjointTotalHeatFluxOption, adjointHeatFluxOption, adjointParticleFluxOption
 
     Zs = speciesNotInitialized
     mHats = speciesNotInitialized
@@ -165,6 +166,8 @@ contains
           print *,"Successfully read parameters from speciesParameters namelist in ", trim(filename), "."
        end if
 
+
+
        read(fileUnit, nml=physicsParameters, iostat=didFileAccessWork)
        if (didFileAccessWork /= 0) then
           print *,"Proc ",myRank,": Error!  I was able to open the file ", trim(filename), &
@@ -217,8 +220,6 @@ contains
           print *,"Successfully read parameters from export_f namelist in ", trim(filename), "."
        end if
     end if
-
-    close(unit = fileUnit)
 
     ! Validate species parameters                                                                                                                                                                            
 
@@ -426,6 +427,25 @@ contains
     end select
 
     Nspecies = NZs
+
+    ! We have to wait until we know Nspecies to read adjoint namelist
+     if (RHSMode>3) then
+       allocate(adjointHeatFluxOption(Nspecies))
+       allocate(adjointParticleFluxOption(NSpecies))
+       adjointHeatFluxOption = .false.
+       adjointParticleFluxOption = .false.
+       read(fileUnit, nml=sensitivityOptions, iostat=didFileAccessWork)
+       if (didFileAccessWork /= 0) then
+          print *,"Proc ",myRank,": Error!  I was able to open the file ", trim(filename), &
[…]

### 2017-08-08 doc/manual/version3/inputParameters.tex
Updated user manual. Made sure adjoint inputs are initialized.

@@
 
+\myhrule
+
+\param{nMaxAdjoint}
+{integer}
+{0}
+{When {\ttfamily RHSMode == 4} or {\ttfamily RHSMode == 5}}
+{Sensitivity of Fourier harmonics of $\bm{B}$ will be computed for $\abs{n} \leq $ \texttt{nMaxAdjoint}.}
+
+\myhrule
+
+\param{mMaxAdjoint}
+{integer}
+{0}
+{When {\ttfamily RHSMode == 4} or {\ttfamily RHSMode == 5}}
+{Sensitivity of Fourier harmonics of $\bm{B}$ will be computed for $m \leq$ \texttt{mMaxAdjoint}.}
+
+\myhrule
+
 \section{\PETSc~commands}

### 2017-08-08 fortran/version3/globalVariables.F90

@@ -96,6 +96,7 @@ module globalVariables
   logical :: adjointBootstrapOption = .false.
   logical :: adjointRadialCurrentOption = .false.
   logical :: adjointTotalHeatFluxOption = .false.
+  ! These are initialized to .false. in readInput
   logical, dimension(:), allocatable :: adjointHeatFluxOption, adjointParticleFluxOption
 
   ! ********************************************************
@@ -379,7 +380,8 @@ module globalVariables
   ! ********************************************************
 
   integer, dimension(:), allocatable :: ns, ms
-  integer :: nMaxAdjoint, mMaxAdjoint, NModesAdjoint, NLambdas
+  integer :: nMaxAdjoint = 0, mMaxAdjoint = 0
+  integer :: NModesAdjoint = 0, NLambdas = 6
   PetscScalar, dimension(:,:), allocatable :: dRadialCurrentdLambda,dTotalHeatFluxdLambda,dBootstrapdLambda
   PetscScalar, dimension(:,:,:), allocatable :: dParticleFluxdLambda, dHeatFluxdLambda
   !! The following are for testing innerProduct()

### 2017-08-15 doc/manual/version3/inputParameters.tex

--- 2017-07-24 doc/manual/version3/inputParameters.tex
@@
 
-\section{The \texttt{sensitivityOptions} namelist}
-
-\param{sensitivityOption}
-{integer}
-{0}
-{When {\ttfamily geometryScheme == 5} and {\ttfamily includePhi1 == .false.}}. \\
-{\ttfamily sensitivityOption} == 0: No sensitivity output computed. \\
-{\ttfamily sensitivityOption} == 1: Sensitivity computed for VMEC harmonics using adjoint equation.
-The quantities of interest for sensitivity computations are specified by {\ttfamily AdjointRHSOption} and {\ttfamily AdjointRHSSpeciesOption}.
-\\
-
-\myhrule
-
-\param{AdjointRHSOption}
-{integer}
-{0}
-{When {\ttfamily geometryScheme == 5} and {\ttfamily sensitivityOption == 1}} \\
-{\ttfamily AdjointRHSOption == 0}: Sensitivity is computed for bootstrap current, particle flux, and heat flux. \\
-{\ttfamily AdjointRHSOption == 1}: Sensitivity is computed for just bootstrap current. \\
-{\ttfamily AdjointRHSOption == 2}: Sensitivity is computed for just particle flux. \\
-{\ttfamily AdjointRHSOption == 3}: Sensitivity is computed for just heat flux.
-
-\myhrule
-
-\param{AdjointRHSSpeciesOption}
-{integer}
-{-1}
-{When {\ttfamily geometryScheme == 5} and {\ttfamily sensitivityOption == 1} and {\ttfamily AdjointRHSOption /= 0}} \\
-{\ttfamily AdjointRHSSpeciesOption == -1}: Sensitivity computed for output quantities specified by {\ttfamily AdjointRHSOption} for all species. \\
-{\ttfamily AdjointRHSSpeciesOption >= 0}: Sensitivity computed for output quantities specified by {\ttfamily AdjointRHSOption} for species {\ttfamily AdjointRHSSpeciesOption}. \\
-
-\myhrule
-
 \section{\PETSc~commands}

### 2017-09-08 doc/poloidal_variation_in_collision_operator_2017-09-08.tex
Added the skeleton to implementing the option to include poloidal variations in the collision operator.

\documentclass[12pt]{article}
\usepackage{url} 
%\usepackage[dvips]{graphicx}
\usepackage[pdftex]{graphicx}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{bm}
\usepackage{float}
\usepackage{color}
\usepackage[dvipsnames]{xcolor}
\usepackage{wrapfig}
\usepackage{multicol}
\usepackage[colorlinks=true,
		linkcolor=red,
		citecolor=blue,
		urlcolor=blue]{hyperref}

\usepackage{titlesec}
\usepackage[titletoc]{appendix}
%\usepackage{appendix}
\setcounter{secnumdepth}{4}

\titleformat{\paragraph}
{\normalfont\normalsize\bfseries}{\theparagraph}{1em}{}
\titlespacing*{\paragraph}
{0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}

\setlength {\parindent} { 10mm} 
\setlength{\textheight}{230mm} 
\setlength{\textwidth}{160mm} 
\setlength{\oddsidemargin}{0mm}
\setlength{\topmargin}{-10mm} 
% newcommands
\newcommand{\p}{\partial}
\newcommand{\g}[1]{\mbox{\boldmath $#1$}}
\newcommand{\vi}{\g V_{\! \! i}}
\newcommand{\ps}{Pfirsch-Schl\"{u}ter} 
\newcommand{\lp}{\left(}
\newcommand{\rp}{\right)}
\newcommand{\ca}[1]{\mbox{\cal $#1$}}
\newcommand{\be}{\begin{displaymath}}
\newcommand{\ee}{\end{displaymath}}
\newcommand{\bn}{\begin{equation}}
\newcommand{\en}{\end{equation}}
\newcommand{\mygtrsim}{\mathrel{\mbox{\raisebox{-1mm}{$\stackrel{>}{\sim}$}}}}
\newcommand{\mylsim}{\mathrel{\mbox{\raisebox{-1mm}{$\stackrel{<}{\sim}$}}}}
\newcommand{\vek}{\bf}
\newcommand{\ten}{\sf}
\newcommand{\bfm}[1]{\mbox{\boldmath$#1$}}
\newcommand{\lang}{\left\langle}
\newcommand{\rang}{\right\rangle}
\newcommand{\vo}[1]{\left|\begin {array}{l} \mbox{} \\ \mbox{} \\$#1$ \end
{array}\right .}  
\newcommand{\von}[2]{\left |\begin {array}{l}
\mbox{}\\$#1$\\$#2$ \end {array}\right .}
\newcommand{\simgt}{\:{\raisebox{-1.5mm}{$\stackrel
{\textstyle{>}}{\sim}$}}\:}
\newcommand{\simlt}{\:{\raisebox{-1.5mm}{$\stackrel
{\textstyle{<}}{\sim}$}}\:}
%\renewcommand {\baselinestretch} {1.67}
%\pagestyle{empty}
\newcommand{\todo}[1]{\textbf{\textcolor{red}{TODO: #1}}}
\newcommand{\remark}[1]{\textbf{\textcolor{red}{REMARK: #1}}}

\title{Implementation of poloidal density variation in collision operator}

\pagestyle{fancy}
\fancyhead{}
\chead{Stefan Buller %850227-2019
\\ Implementation of poloidal density variation in collision operator}
\cfoot{\thepage}
\renewcommand{\headrulewidth}{1pt}
\renewcommand{\footrulewidth}{1pt}
\setlength{\headheight}{28pt}
\setlength{\footskip}{25pt}

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\mE}{\mathcal{E}}
\newcommand{\energy}{\mathcal{E}}
\newcommand{\mK}{\mathcal{K}}
\newcommand{\mN}{\mathcal{N}}
\newcommand{\mD}{\mathcal{D}}
\newcommand{\ord}{\mathcal{O}}
\newcommand{\Tpe}{T_\perp}
\newcommand{\Tpa}{T_\|}
\newcommand{\vpe}{v_\perp}
\newcommand{\vpa}{v_\|}
\newcommand{\kpe}{k_\perp}
\newcommand{\kpa}{k_\|}
\newcommand{\Bv}{\mathbf{B}}
\newcommand{\Ev}{\mathbf{E}}
\newcommand{\bv}{\mathbf{b}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\cd}{\cdot}
\newcommand{\na}{\nabla}
\newcommand{\btheta}{\bar{\theta}}
\newcommand{\phit}{\tilde{\phi}}
\newcommand{\oert}{\tilde{\omega}_{Er}}

\begin{document}
\titlepage

\maketitle

These notes are concerned with the implementation of the \texttt{poloidalVariationInCollisionOperator} option in SFINCS version 3.

\section*{$\Phi_1$ implementation in SFINCS}
In a previous set of notes (\emph{$\Phi_1$ implementation}), poloidal density variation to the lowest order was implemented by linearizing around a poloidally varying Maxwellian $f_0$, rather than a flux-function Maxwellian $f_M$.

Poloidal variations can occur to lowest order if the potential varies on a flux-surface, in which case the lowest order distribution function becomes
\begin{equation}
f_0(\psi,\theta,\zeta) = f_M(\psi) e^{-Ze\Phi_1(\theta,\zeta)/T}.
\end{equation}
[We will in the future suppress the $\psi$ dependence, as SFINCS only solves for a single flux-surface.]

In the \emph{$\Phi_1$ implementation} notes, the equations were modified to linearize around this $f_0$ rather than $f_M$. However, the linearization in the collision operator was not treated, which is the subject of these notes.

\section*{$\Phi_1$ implementation in the linearized collision operator}
The linearized collision operator can be written as
\begin{equation}
C^{L:f_0}_{ab} = C_{ab}\{f_{a0},f_{b0}\} +  C_{ab}\{f_{a1},f_{b0}\} +  C_{ab}\{f_{a0},f_{b1}\},
\end{equation}
where we use the superscript $L:f_0$ to indicate that the operator has been linearized around $f_0$ rather than $f_M$.

As $f_0$ and $f_M$ have the same velocity space-structure, the terms in the above linearization can easily be expressed in terms of the terms in the linearization around $f_M$:
\begin{equation}
C^{L:f_0}_{ab} =  C_{ab}\{f_{aM},f_{bM}\} e^{-(Z_a/T_a + Z_b/T_b) e\Phi_1(\theta,\zeta)} +  C_{ab}\{f_{a1},f_{b0}\} e^{-Z_b e\Phi_1(\theta,\zeta)/T_b} +  C_{ab}\{f_{aM},f_{b1}\} e^{-Z_a e\Phi_1(\theta,\zeta)/T_a}.
\end{equation}
Effectively, from the explicit expressions for $C^{L:f_M}$, one can obtain $C^{L:f_0}$ by substituting the density of each species by:
\begin{equation}
n_a \to n_a e^{-\frac{Z_a e\Phi_1(\theta,\zeta)}{T_a}}. \label{eq:denssub}
\end{equation}

\subsection*{Changes to the Jacobian}
As the old collision operator only depend on $f_{1}$ (neglecting the temperature equilibration term $C_{ab}\{f_{aM},f_{bM}\}$ for the moment), its contribution to the Jacobian was
\begin{equation}
J^{L:f_M} = \frac{\delta C^{L:f_M}}{\delta f_{1}}.
\end{equation}
Furthermore, as the linearized collision operator is linear in $f_{1}$, we have
\begin{equation}
\frac{\delta C^{L:f_M}}{\delta f_{1}} =  C^{L:f_M}[.],
\end{equation}
where the notation $C^{L:f_M}[.]$ implies that the operator does not act on anything, i.e. the matrix used to represent $C^{L:f_M}[.]$ enters as is into the Jacobian.

In the system linearized around $f_0$, we have essentially the same $f_1$ dependence, if we substitute the density as in \eqref{eq:denssub}
\begin{equation}
\frac{\delta C^{L:f_0}}{\delta f_{1}} =  C^{L:f_M}[.] \left\{n_a \to n_a e^{-\frac{Z_a e\Phi_1(\theta,\zeta)}{T_a}}\right\},
\end{equation}
however, as $\Phi_1$ is an unknown, we get new non-zero elements in the Jacobian due to $\frac{\delta C^{L:f_0}}{\delta \Phi_{1}}$.
As $C^{L:f_0}$ only depends on $\Phi_1$ through $n_a e^{-\frac{Z_a e\Phi_1(\theta,\zeta)}{T_a}}$, we can also get the new expressions through as a density substitution
\begin{equation}
n_a e^{-\frac{Z_a e\Phi_1}{T_a}} \to  -\frac{Z_a e}{T_a} n_a e^{-\frac{Z_a e\Phi_1}{T_a}}.
\end{equation}
Thus
\begin{equation}
\frac{\delta C^{L:f_0}}{\delta \Phi_{1}} =  C^{L:f_M}[f_1] \left\{n_a \to -\frac{Z_a e}{T_a} n_a e^{-\frac{Z_a e\Phi_1}{T_a}}\right\}.
\end{equation}
Note that the collision operator here acts on $f_1$.

In addition, if we include temperature equilibration, we get an additional contribtution to the Jacobian
\begin{equation}
\frac{\delta C^{L:f_0}}{\delta \Phi_{1}} =  -e\frac{Z_a T_b + Z_b T_a}{T_a T_b} e^{\left(-\frac{Z_a T_b + Z_b T_a}{T_a T_b} e\Phi_1\right)} C_{ab}\{f_{aM},f_{bM}\}.
\end{equation}
This term, like all the other in this document, is merely an extra factor to an expression already calculated in the code.

\section*{Changes to the code}
The collision operator is linearized around $f_0$ if \texttt{poloidalVariationInCollisionOperator} is set to true in the input file. This switch adds the extra factor $e^{-\frac{Z_a e\Phi_1}{T_a}}$ to the densities in the collision operator as calculated in the code, and reuses the calculations to evaluate $\frac{\delta C^{L:f_0}}{\delta \Phi_{1}}$. If \texttt{includeTemperatureEquilibrationTerm} is set to true, the contribution from this term is also included in the Jacobian.

\end{document}

### 2017-09-08 fortran/version3/globalVariables.F90

@@ -113,6 +113,7 @@ module globalVariables
   logical :: include_fDivVE_term = .false.
   logical :: includeTemperatureEquilibrationTerm = .false.
   logical :: includePhi1 = .false.
+  logical :: poloidalVariationInCollisionOperator = .false.
 !!  logical :: includeRadialExBDrive = .false. !!Commented by AM 2016-02
   logical :: includePhi1InKineticEquation = .true. !!Added by AM 2016-03
 

### 2017-09-28 doc/PoloidalVariationInCollisionOperator_code.tex
added documentation

\documentclass[12pt]{article}
\usepackage{url} 
%\usepackage[dvips]{graphicx}
\usepackage[pdftex]{graphicx}
\usepackage[latin1]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{bm}
\usepackage{float}
\usepackage{listings}
\usepackage{color}

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstset{frame=tb,
  language=Fortran,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=none,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=true,
  tabsize=3,
  frame=none
}\usepackage[dvipsnames]{xcolor}
\usepackage{wrapfig}
\usepackage{multicol}
\usepackage[colorlinks=true,
		linkcolor=red,
		citecolor=blue,
		urlcolor=blue]{hyperref}

\usepackage{titlesec}
\usepackage[titletoc]{appendix}
%\usepackage{appendix}
\setcounter{secnumdepth}{4}

\titleformat{\paragraph}
{\normalfont\normalsize\bfseries}{\theparagraph}{1em}{}
\titlespacing*{\paragraph}
{0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}

\setlength {\parindent} { 10mm} 
\setlength{\textheight}{230mm} 
\setlength{\textwidth}{160mm} 
\setlength{\oddsidemargin}{0mm}
\setlength{\topmargin}{-10mm} 
% newcommands
\newcommand{\p}{\partial}
\newcommand{\g}[1]{\mbox{\boldmath $#1$}}
\newcommand{\vi}{\g V_{\! \! i}}
\newcommand{\ps}{Pfirsch-Schl\"{u}ter} 
\newcommand{\lp}{\left(}
\newcommand{\rp}{\right)}
\newcommand{\ca}[1]{\mbox{\cal $#1$}}
\newcommand{\be}{\begin{displaymath}}
\newcommand{\ee}{\end{displaymath}}
\newcommand{\bn}{\begin{equation}}
\newcommand{\en}{\end{equation}}
\newcommand{\mygtrsim}{\mathrel{\mbox{\raisebox{-1mm}{$\stackrel{>}{\sim}$}}}}
\newcommand{\mylsim}{\mathrel{\mbox{\raisebox{-1mm}{$\stackrel{<}{\sim}$}}}}
\newcommand{\vek}{\bf}
\newcommand{\ten}{\sf}
\newcommand{\bfm}[1]{\mbox{\boldmath$#1$}}
\newcommand{\lang}{\left\langle}
\newcommand{\rang}{\right\rangle}
\newcommand{\vo}[1]{\left|\begin {array}{l} \mbox{} \\ \mbox{} \\$#1$ \end
{array}\right .}  
\newcommand{\von}[2]{\left |\begin {array}{l}
\mbox{}\\$#1$\\$#2$ \end {array}\right .}
\newcommand{\simgt}{\:{\raisebox{-1.5mm}{$\stackrel
{\textstyle{>}}{\sim}$}}\:}
\newcommand{\simlt}{\:{\raisebox{-1.5mm}{$\stackrel
{\textstyle{<}}{\sim}$}}\:}
%\renewcommand {\baselinestretch} {1.67}
%\pagestyle{empty}
\newcommand{\todo}[1]{\textbf{\textcolor{red}{TODO: #1}}}
\newcommand{\remark}[1]{\textbf{\textcolor{red}{REMARK: #1}}}

\title{Implementation of poloidal density variation in collision operator}

\pagestyle{fancy}
\fancyhead{}
\chead{Aylwin Iantchenko %850227-2019
\\ Implementation of poloidal density variation in collision operator}
\cfoot{\thepage}
\renewcommand{\headrulewidth}{1pt}
\renewcommand{\footrulewidth}{1pt}
\setlength{\headheight}{28pt}
\setlength{\footskip}{25pt}

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\mE}{\mathcal{E}}
\newcommand{\energy}{\mathcal{E}}
\newcommand{\mK}{\mathcal{K}}
\newcommand{\mN}{\mathcal{N}}
\newcommand{\mD}{\mathcal{D}}
\newcommand{\ord}{\mathcal{O}}
\newcommand{\Tpe}{T_\perp}
\newcommand{\Tpa}{T_\|}
\newcommand{\vpe}{v_\perp}
\newcommand{\vpa}{v_\|}
\newcommand{\kpe}{k_\perp}
\newcommand{\kpa}{k_\|}
\newcommand{\Bv}{\mathbf{B}}
\newcommand{\Ev}{\mathbf{E}}
\newcommand{\bv}{\mathbf{b}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\cd}{\cdot}
\newcommand{\na}{\nabla}
\newcommand{\btheta}{\bar{\theta}}
\newcommand{\phit}{\tilde{\phi}}
\newcommand{\oert}{\tilde{\omega}_{Er}}

% Simplify equations
\newcommand{\eq}[1]{\begin{align*}\begin{gathered}#1\end{gathered}\end{align*}}
\newcommand{\eqre}[1]{\begin{align}\begin{gathered}#1\end{gathered}\end{align}}

\begin{document}
\titlepage

\maketitle

The aim with this document is to explain the modifications in \texttt{SFINCS} that have been done, in order to include poloidal density variation in the Fokker-Planck collision operator.

As is explained in ref.~\cite{ref:PolVarColl_stefan} poloidal density variation can be included by modifying the lowest order distribution function 
\eq{
f_{Ms}(\psi) \rightarrow f_{Ms}(\psi)e^{-Z_se\Phi_1(\theta,\zeta)/T_s},
}
where $f_M(\psi)$ is the flux-function Maxwellian, $\Phi_1$ are first order variation of the electrostatic potential, and $Z_s,T_s$ are the charge and temperature of species $s$ respectively. To implement this change one may simply modify the species density accordingly
\eq{
n_s \rightarrow n_se^{-Ze\Phi_1(\theta,\zeta)/T},
}
in the original equations for the various terms in the collision operator, described in ref.~\cite{ref:Coll1}.

The linearised collision operator takes the form
\eqre{
C_{ab}^{L:f0} =  C_{ab}\left\{f_{aM},f_{bM}\right\}e^{-\left(Z_a/T_a + Z_b/T_b\right)e\Phi_1(\theta,\zeta)} + C_{ab}\left\{f_{a1},f_{bM}\right\}e^{-Z_b/T_be\Phi_1(\theta,\zeta)} + \\ + C_{ab}\left\{f_{aM},f_{b1}\right\}e^{-Z_a/T_ae\Phi_1(\theta,\zeta)},
}
which describes collisions between species $a$ and $b$. In normalised \texttt{SFINCS} units this becomes
\eqre{
\label{eq:CollMod}
\hat C_{ab}^{L:f0} = \hat C_{ab}\left\{f_{aM},f_{bM}\right\}e^{-\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha\hat\Phi_1(\theta,\zeta)} + \hat C_{ab}\left\{f_{a1},f_{bM}\right\}e^{-\hat Z_b/\hat T_b \alpha\hat\Phi_1(\theta,\zeta)} + \\ + \hat C_{ab}\left\{f_{aM},f_{b1}\right\}e^{-\hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta)}.
}

\noindent
The first term on the RHS of Eq.~\eqref{eq:CollMod} is the \textit{temperature equilibration term}. Since this term does not include $f_{1s}$ it has to be treated differently compared to the other two cases.

The contribution to the Jacobian are $\partial \hat C_{ab}^{L:f0}/\partial \hat f_{1s}$ and $\partial \hat C_{ab}^{L:f0}/\partial \hat \Phi_1$. The first is the same as in Eq.~\eqref{eq:CollMod} except that the collision operator does not act on the distribution function. In the second case we get

\eqre{
\label{eq:Jac1}
\hat C_{ab}^{L:f0} = -\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha \hat C_{ab}\left\{f_{aM},f_{bM}\right\}e^{-\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha\hat\Phi_1(\theta,\zeta)}  \\ -\hat Z_b/\hat T_b \alpha\ \hat C_{ab}\left\{f_{a1},f_{bM}\right\}e^{-\hat Z_b/\hat T_b \alpha\hat\Phi_1(\theta,\zeta)}  - \hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta) \hat C_{ab}\left\{f_{aM},f_{b1}\right\}e^{-\hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta)}.
}

\noindent
Apart from the different prefactor, the term in Eq.~\eqref{eq:Jac1} differs from term in $\partial \hat C_{ab}^{L:f0}/\partial \hat f_{1s}$ by acting on the first order distribution function $f_{1s}$. For example, in the first case, terms such as $\partial f_{1s}/\partial x_s$ become 
\eq{
\frac{1}{\partial f_{1s}}\frac{\partial f_{1s}}{\partial x_s} = \frac{\partial}{\partial x_s},
}
whereas in the second case, when take the derivative with respect to $\Phi_1$, we have to actually calculate the derivative $\partial f_{1s}/\partial x_s$.

%We will now go through the various changes in the code that have been done to implement Eq.~\eqref{eq:CollMod}.

\section*{Implementation in the Code}
To implement Eq.~\eqref{eq:CollMod} we need to modify the evaluation of the residual and the jacobian. The residual is evaluated in \texttt{evaluateResidual.F90} and the jacobian in \texttt{evaluateJacobian.F90}. Both these two routines call the functions in \texttt{PopulateMatrix.F90}, where the actual assembly of the residual and jacobian matrices are done. Therefore, we only need to modify \texttt{PopulateMatrix.F90}.

\subsection*{Residual}
\label{sec:Res}
In the residual we have to include the extra pre-factor appearing in Eq.~\eqref{eq:CollMod}. This is done by modifying the last block in the calculation of the collision operator, just before the terms are saved into the main matrix. We introduce the pre-factor \texttt{preFactor} which is normally equal to one, but $e^{-\hat Z_a/\hat T_a \alpha\hat\Phi_1(\theta,\zeta)}$ when poloidal density variation is included (by setting the flag \texttt{poloidalVariationInCollisionOperator = .true.}).

\begin{lstlisting}
  preFactor = 1.0 ! Initiate the preFactor used to multiply CHat before saving into the Main matrix
  if (poloidalVariationInCollisionOperator) then 
     preFactor = exp(-Zs(iSpeciesA)*alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesA)) 
\end{lstlisting}
This factor then multiplies the collision operator, just before saving into the main matrix
\begin{lstlisting}
  call MatSetValueSparse(matrix, rowIndex, colIndex, &
       -nu_n*preFactor*CHat(ix_row,ix_col), ADD_VALUES, ierr) ! Multiply CHat with preFactor before saving
\end{lstlisting}
\subsubsection*{Temperature equilibration term}
In the original version of the code, the temperature equilibration term is calculated in \\ \texttt{evaluateResidual.F90} by using \texttt{whichMatrix = 2} when calling \texttt{PopulateMatrix.F90}. The output is then multiplied with the first order distribution function $f_{0s}$. Since the original version of \texttt{SFINCS} already includes $\Phi_1$ in the definition of $f_{0s}$ (in the subroutine \texttt{init\_f0()} inside \texttt{PopulateMatrix.F90}), the same routines as were presented above can be used. In the end, when calling \texttt{PopulateMatrix} we include the pre-factor $e^{-\hat Z_a/\hat T_a \alpha\hat\Phi_1(\theta,\zeta)}$. Then by multiplying with $f_{0b}$ in \texttt{evaluateResidual.F90} we get the desired total pre-factor $e^{-\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha\hat\Phi_1(\theta,\zeta)}$.

\subsection*{Jacobian}
The $\partial C/\partial f_{1s}$ contribution in the jacobian is at this point already taking the density variations into account if we implement the modifications explained in Section~\ref{sec:Res}. However, we need to add the calculation of $\partial C/\partial \Phi_1$.

\subsubsection*{The $\partial C/\partial \Phi_1$ term}
\label{sec:Jac1}
To implement the derivative with respect to $\Phi_1$ we add another block right after the calculation of the residual terms. This block should be evaluated whenever the jacobian (\texttt{whichmatrix =1}) or the preconditioner (\texttt{whichmatrix =0}) is calculated, with \texttt{poloidalVariationInCollisionOperator = .true.}

\begin{lstlisting}
 if (poloidalVariationInCollisionOperator .and. (whichMatrix == 1 .or. whichMatrix == 0)) then
\end{lstlisting}

We begin with treating the other two terms appearing in Eq.~\eqref{eq:Jac1} and will then look at the temperature equilibration term. For the second two terms in Eq.~\eqref{eq:Jac1} we first have to implement the correct pre-factor.

\begin{lstlisting}
 preFactor = -Zs(iSpeciesA)*alpha/Thats(iSpeciesA)*exp(-Zs(iSpeciesA)&
 *alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesA))\end{lstlisting}

\noindent
The second change is including the first order distribution function (and calculating derivatives thereof). For this purpose we first have to initiate this distribution function, by reading the appropriate terms in the current state vector

\begin{lstlisting}
 index = getIndex(iSpeciesB,ix,L+1,itheta,izeta,BLOCK_F) ! f1b from statevector
 f1b(ix) = stateArray(index + 1)
\end{lstlisting}
\noindent
What is left is to multiply this vector with the collision operator matrix \texttt{CHat} containing terms such as $\partial/\partial x_s$.

 \begin{lstlisting}
 CHatTimesf = matmul(CHat,f1b)
 \end{lstlisting}
\noindent
The output is then saved into the main matrix in the $\partial \Hat C/\partial \hat \Phi_1$ block,

 \begin{lstlisting}
 call MatSetValue(matrix, rowIndex, colIndex, & 
 -nu_n*preFactor*CHatTimesf(ix_col), ADD_VALUES, ierr) 
\end{lstlisting}
\noindent
using the approptiate row and column indices.

 \begin{lstlisting}
rowIndex=getIndex(iSpeciesA,ix_row,L+1,itheta,izeta,BLOCK_F)
do ix_col = max(ixMinCol,min_x_for_L(L)),Nx
   ! Get column index for the d/dPhi1 terms
   colIndex=getIndex(1,1,1,itheta,izeta,BLOCK_QN)
\end{lstlisting}

\subsubsection*{Temperature equilibration}
\label{sec:Jac2}
If the temperature equilibration terms should be included, we have to add additional terms into the main matrix, when calculating  $\partial \Hat C/\partial \hat \Phi_1$. To begin with we need to include the different pre-factor (since in this case $f_0$ from \texttt{init\_f0()} is not used).

\begin{lstlisting}
if (includeTemperatureEquilibrationTerm) then
   ! Get the correct preFactor
   preFactor = (-Zs(iSpeciesA)*alpha/Thats(iSpeciesA) -Zs(iSpeciesB)*alpha/  & 
    Thats(iSpeciesB)) 
   *exp(-Zs(iSpeciesA)*alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesA)-Zs(iSpeciesB) & 
   *alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesB))
\end{lstlisting}
\noindent
The second change originates from the fact that the collision operator in this temperature equilibration case, should multiply the Maxwellian for species $b$ and not $f_{1b}$ (as was previously the case in Section~\ref{sec:Jac2}).  First, we have to initiate this distribution function
 \begin{lstlisting}
 if (includeTemperatureEquilibrationTerm) then
    fM(ix) = expxb2(ix)*nhats(iSpeciesB)*mhats(iSpeciesB)*sqrt(mhats(iSpeciesB)) &
     /(pi*sqrtpi*Thats(iSpeciesB)*sqrt(Thats(iSpeciesB)))
     end if   
\end{lstlisting}
\noindent
and then use it to multiply to the total collision operator.
 \begin{lstlisting}
  CHatTimesf = matmul(CHat,fM)
 \end{lstlisting}
\noindent
Then we add this term into the main matrix, using the the same procedure as was explained in Section~\ref{sec:Jac1}

\newpage

\begin{thebibliography}{99}

\bibitem{ref:PolVarColl_stefan} S.~Buller, {\em Poloidal variation in collision operator} (2017).

\bibitem{ref:Coll1} M.~Landreman, {\em Technical Documentation for version 3 of SFINCS} (2014).

%\bibitem{simakov} A.~N.~Simakov, P.~Helander,
%  {\em Phys. Plasmas} {\bf 16}, 042503 (2009).

%\bibitem{nonAxis} P.~Helander, Theory of plasma confinement in non-axisymmetric magnetic fields (2013).

%\bibitem{MH} L.~Råde and B.~Westergren, Mathematics Handbook for Science and Engineering, $5^{\mathrm{th}}$ edition, 2004. %\vspace{-5mm}

%\bibitem{Abra} M.~Abramowitz and I.~A.~Stegun, Handbook of Mathematical Functions, $10^{\mathrm{th}}$ printing, 1972.

\end{thebibliography}

\end{document}

### 2017-10-05 doc/PoloidalVariationInCollisionOperator_code.tex
updated documentation

@@
 \title{Implementation of poloidal density variation in collision operator}
+\author{Aylwin Iantchenko}
 
@@
 
-The aim with this document is to explain the modifications in \texttt{SFINCS} that have been done, in order to include poloidal density variation in the Fokker-Planck collision operator.
+The aim of this document is to explain the modifications in \texttt{SFINCS} that have been done, in order to include poloidal density variation in the collision operator (in both full Fokker-Planck and pure pitch angle scattering operator options).
 
-As is explained in ref.~\cite{ref:PolVarColl_stefan} poloidal density variation can be included by modifying the lowest order distribution function 
+As is explained in Ref.~\cite{ref:PolVarColl_stefan} poloidal density variation can be included by modifying the lowest order distribution function 
 \eq{
-f_{Ms}(\psi) \rightarrow f_{Ms}(\psi)e^{-Z_se\Phi_1(\theta,\zeta)/T_s},
+f_{Ms}(\psi) \rightarrow f_{0s} = f_{Ms}(\psi)e^{-Z_se\Phi_1(\theta,\zeta)/T_s},
 }
@@
 To implement this change one may simply modify the species density accordingly
-\eq{
+\eqre{
+\label{eq:densMod}
 n_s \rightarrow n_se^{-Ze\Phi_1(\theta,\zeta)/T},
 }
-in the original equations for the various terms in the collision operator, described in ref.~\cite{ref:Coll1}.
-
-The linearised collision operator takes the form
+in the original equations for the various terms in the collision operator, which are described in Ref.~\cite{ref:Coll1}. The linearised collision operator takes the new form
 \eqre{
@@
 The first term on the RHS of Eq.~\eqref{eq:CollMod} is the \textit{temperature equilibration term}.
-Since this term does not include $f_{1s}$ it has to be treated differently compared to the other two cases.
+Since this term does not include $f_{1s}$ it has to be treated differently compared to the other two cases, as will be seen shortly.
+In the code, the calculation of the residual has to be modified to include the factors specified in Eq.~\eqref{eq:CollMod}.
 
@@
 \label{eq:Jac1}
-\hat C_{ab}^{L:f0} = -\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha \hat C_{ab}\left\{f_{aM},f_{bM}\right\}e^{-\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha\hat\Phi_1(\theta,\zeta)}  \\ -\hat Z_b/\hat T_b \alpha\ \hat C_{ab}\left\{f_{a1},f_{bM}\right\}e^{-\hat Z_b/\hat T_b \alpha\hat\Phi_1(\theta,\zeta)}  - \hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta) \hat C_{ab}\left\{f_{aM},f_{b1}\right\}e^{-\hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta)}.
+\frac{\partial \hat C_{ab}^{L:f0}}{\partial \hat \Phi_1} = -\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha \hat C_{ab}\left\{f_{aM},f_{bM}\right\}e^{-\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha\hat\Phi_1(\theta,\zeta)}  \\ -\hat Z_b/\hat T_b \alpha\ \hat C_{ab}\left\{f_{a1},f_{bM}\right\}e^{-\hat Z_b/\hat T_b \alpha\hat\Phi_1(\theta,\zeta)}  - \hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta) \hat C_{ab}\left\{f_{aM},f_{b1}\right\}e^{-\hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta)}.
 }
@@
 \noindent
-Apart from the different prefactor, the term in Eq.~\eqref{eq:Jac1} differs from term in $\partial \hat C_{ab}^{L:f0}/\partial \hat f_{1s}$ by acting on the first order distribution function $f_{1s}$.
+Apart from the different pre-factor, the term in Eq.~\eqref{eq:Jac1} differs from term in $\partial \hat C_{ab}^{L:f0}/\partial \hat f_{1s}$ by acting on the first order distribution function $f_{1s}$.
 For example, in the first case, terms such as $\partial f_{1s}/\partial x_s$ become
@@
 }
-whereas in the second case, when take the derivative with respect to $\Phi_1$, we have to actually calculate the derivative $\partial f_{1s}/\partial x_s$.
+whereas in the second case, when we take the derivative with respect to $\Phi_1$, we have to include the derivative $\partial f_{1s}/\partial x_s$ (and $\partial f_{0s}/\partial x_s$ for temperature equilibration), as well.
 
@@
 \section*{Implementation in the Code}
-To implement Eq.~\eqref{eq:CollMod} we need to modify the evaluation of the residual and the jacobian.
-The residual is evaluated in \texttt{evaluateResidual.F90} and the jacobian in \texttt{evaluateJacobian.F90}.
-Both these two routines call the functions in \texttt{PopulateMatrix.F90}, where the actual assembly of the residual and jacobian matrices are done.
-Therefore, we only need to modify \texttt{PopulateMatrix.F90}.
+To implement Eq.~\eqref{eq:CollMod} and Eq.~\eqref{eq:Jac1} we need to modify the evaluation of the residual and the Jacobian.
+The residual is evaluated in \texttt{evaluateResidual.F90} and the Jacobian in \texttt{evaluateJacobian.F90}.
+Both these two routines call the functions in \texttt{PopulateMatrix.F90}, where the actual assembly of the residual and Jacobian matrices are done.
+As a consequence of this structure, we only need to modify \texttt{PopulateMatrix.F90} in order to include poloidal density variation in the collision operator.
 
@@
 In the residual we have to include the extra pre-factor appearing in Eq.~\eqref{eq:CollMod}.
-This is done by modifying the last block in the calculation of the collision operator, just before the terms are saved into the main matrix.
-We introduce the pre-factor \texttt{preFactor} which is normally equal to one, but $e^{-\hat Z_a/\hat T_a \alpha\hat\Phi_1(\theta,\zeta)}$ when poloidal density variation is included (by setting the flag \texttt{poloidalVariationInCollisionOperator = .true.}).
+Since this factor is species specific, we have to keep track of which species density is appearing in the equations.
+The pitch-angle scattering and the $\hat C^E_{ab}\left \{f_{a1}\right \}$ parts of the collision operator (see Ref.
+~\cite{ref:Coll1}) both uses the density $\texttt{nHats(iSpeciesB)}$ while the other contributions to the total collision operator \texttt{CHat} uses \texttt{nHats(iSpeciesA)} instead.
+
+Rather than creating many if-statements in the current implementation of the collision operator, for sake of clarity we separate the two implementations with one if-statement directly before the various contributions to the collision operator are calculated.
+If \texttt{poloidalVariationInCollisionOperator = .false.} the first block is executed, which contains the original code.
 
 \begin{lstlisting}
-  preFactor = 1.0 ! Initiate the preFactor used to multiply CHat before saving into the Main matrix
-  if (poloidalVariationInCollisionOperator) then 
+if (.not. poloidalVariationInCollisionOperator) then   !! Added by AI (2017-09).
+\end{lstlisting}
+
+\noindent Otherwise, if \texttt{poloidalVariationInCollisionOperator = .true.}, \texttt{includePhi1 = .true.} and \texttt{includePhi1InKineticEquation = .true.} we proceed to the new block where poloidal density variation is included.
+
+\begin{lstlisting}
+else if (includePhi1 .and. includePhi1InKineticEquation) then! Do with poloidal density variation
+\end{lstlisting}
+
+\noindent
+Since we now will have a \texttt{Phi1Hat} dependence in the various terms of the collision operator we will have to iterate over \texttt{itheta} and \texttt{izeta} when calculating the various contributions.
+Before calculating each term, we define a pre-factor (\texttt{PreFactor}) as is specified in Eq.~\eqref{eq:CollMod}, taking the correct species index into account.
+
+\begin{lstlisting}
+ do itheta=ithetaMin,ithetaMax
+    do izeta=izetaMin,izetaMax
+       ! Generate preFactor for nHats(iSpeciesA) terms
        preFactor =  exp(-Zs(iSpeciesA)*alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesA))
 \end{lstlisting}
-This factor then multiplies the collision operator, just before saving into the main matrix
-\begin{lstlisting}
-  call MatSetValueSparse(matrix, rowIndex, colIndex, &
-       -nu_n*preFactor*CHat(ix_row,ix_col), ADD_VALUES, ierr) ! Multiply CHat with preFactor before saving
-\end{lstlisting}
+
+\noindent
+Because of the \texttt{itheta}, \texttt{izeta} dependence we have redefined \texttt{CECD} and \texttt{nuDHat} to their equivalents \texttt{CECDpol} and \texttt{nuDHatpol} with space for their \texttt{itheta} and \texttt{izeta} components.
+
+The rest of this block is the same as in the original code, except that we do the second iteration over \texttt{itheta}, \texttt{izeta} slightly earlier, and include the appropriate \texttt{PreFactor} in front of additional contributions to the total collision operator \texttt{CHat}.
+
 \subsubsection*{Temperature equilibration term}
@@
 The output is then multiplied with the first order distribution function $f_{0s}$.
-Since the original version of \texttt{SFINCS} already includes $\Phi_1$ in the definition of $f_{0s}$ (in the subroutine \texttt{init\_f0()} inside \texttt{PopulateMatrix.F90}), the same routines as were presented above can be used.
-In the end, when calling \texttt{PopulateMatrix} we include the pre-factor $e^{-\hat Z_a/\hat T_a \alpha\hat\Phi_1(\theta,\zeta)}$.
-Then by multiplying with $f_{0b}$ in \texttt{evaluateResidual.F90} we get the desired total pre-factor $e^{-\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha\hat\Phi_1(\theta,\zeta)}$.
+Since the original version of \texttt{SFINCS} already includes $\Phi_1$ in the definition of $f_{0s}$ (in the subroutine \texttt{init\_f0()} inside \texttt{PopulateMatrix.F90}), the same routines as were presented above can be used also if \texttt{includeTemperatureEquilibrationTerm = .true.}.
 
 \subsection*{Jacobian}
-The $\partial C/\partial f_{1s}$ contribution in the jacobian is at this point already taking the density variations into account if we implement the modifications explained in Section~\ref{sec:Res}.
-However, we need to add the calculation of $\partial C/\partial \Phi_1$.
+ If the modifications explained in the previous section are implemented, the $\partial C/\partial f_{1s}$ contribution to the Jacobian is at this point already taking the density variations into account.
+We need to add the calculation of $\partial C/\partial \Phi_1$.
 
@@
 \label{sec:Jac1}
-To implement the derivative with respect to $\Phi_1$ we add another block right after the calculation of the residual terms.
-This block should be evaluated whenever the jacobian (\texttt{whichmatrix =1}) or the preconditioner (\texttt{whichmatrix =0}) is calculated, with \texttt{poloidalVariationInCollisionOperator = .true.}
+We begin with treating the other two terms appearing in Eq.~\eqref{eq:Jac1} and will then look at the temperature equilibration term.
+
+To implement the derivative with respect to $\Phi_1$ we need to redo the calculation of the various contributions to the collision operator, but including a new pre-factor, \texttt{PreFactorJ} as specified in Eq.~\eqref{eq:Jac1}.
+This should only be done when the Jacobian (\texttt{whichmatrix =1}) or the preconditioner (\texttt{whichmatrix =0}) is calculated.
 
 \begin{lstlisting}
- if (poloidalVariationInCollisionOperator .and. (whichMatrix == 1 .or. whichMatrix == 0)) then
+ if (whichMatrix == 1 .or. whichMatrix == 0) then
+    ! Generate preFactorJ for nHats(iSpeciesB) terms              
+    preFactorJ =  (-Zs(iSpeciesA)*alpha/Thats(iSpeciesA)) &
+    *exp(-Zs(iSpeciesA)*alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesA))
 \end{lstlisting}
 
-We begin with treating the other two terms appearing in Eq.~\eqref{eq:Jac1} and will then look at the temperature equilibration term.
-For the second two terms in Eq.~\eqref{eq:Jac1} we first have to implement the correct pre-factor.
+\noindent
+The result is saved in the new matrices \texttt{nuDHatpolJ}, \texttt{CECDpolJ} etc.
+Same calculations are then repeated for these terms (except for the different pre-factor), and the result is saved in \texttt{CHatJ}.
+
+In contrary to the residual (and the $\partial C/\partial f_{1s}$ terms ), when calculating $\partial C/\partial \Phi_1$ we have to include the first order distribution function before saving the result into the main matrix.
+For this purpose we first have to initiate this distribution function, by reading the appropriate terms in the current state vector.
 
 \begin{lstlisting}
- preFactor = -Zs(iSpeciesA)*alpha/Thats(iSpeciesA)*exp(-Zs(iSpeciesA)&
- *alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesA))\end{lstlisting}
+ do ix= max(ixMinCol,min_x_for_L(L)),Nx
+    ! Generate f1b from state vector
+    index = getIndex(iSpeciesB,ix,L+1,itheta,izeta,BLOCK_F) 
+    f1b(ix) = stateArray(index + 1)
+\end{lstlisting}
 
 \noindent
-The second change is including the first order distribution function (and calculating derivatives thereof).
-For this purpose we first have to initiate this distribution function, by reading the appropriate terms in the current state vector
+We multiply the terms in the collision operator with this distribution function.
 
  \begin{lstlisting}
- index = getIndex(iSpeciesB,ix,L+1,itheta,izeta,BLOCK_F) ! f1b from statevector
- f1b(ix) = stateArray(index + 1)
+  CHatTimesf = matmul(CHatJ,f1b)
  \end{lstlisting}
 \noindent
-What is left is to multiply this vector with the collision operator matrix \texttt{CHat} containing terms such as $\partial/\partial x_s$.
+The result is then saved into the $\partial \Hat C/\partial \hat \Phi_1$ block of the main matrix,
 
  \begin{lstlisting}
- CHatTimesf = matmul(CHat,f1b)
+do ix_row=max(ixMin,min_x_for_L(L)),Nx
+   rowIndex=getIndex(iSpeciesA,ix_row,L+1,itheta,izeta,BLOCK_F)
+   ! Get column index for the d/dPhi1 terms
+   colIndex=getIndex(1,1,1,itheta,izeta,BLOCK_QN)
+   ! Save into the main matrix, note that here we only use ix_row since CHatTimesf is now a vector
+   call MatSetValue(matrix, rowIndex, colIndex, & 
+   -nu_n*CHatTimesf(ix_row), ADD_VALUES, ierr) 
+   ! need to use MatSetValue, otherwise petsc gives error
+end do ! ix_row
 \end{lstlisting}
 \noindent
-The output is then saved into the main matrix in the $\partial \Hat C/\partial \hat \Phi_1$ block,
-
- \begin{lstlisting}
- call MatSetValue(matrix, rowIndex, colIndex, & 
- -nu_n*preFactor*CHatTimesf(ix_col), ADD_VALUES, ierr) 
-\end{lstlisting}
-\noindent
-using the approptiate row and column indices.
-
- \begin{lstlisting}
-rowIndex=getIndex(iSpeciesA,ix_row,L+1,itheta,izeta,BLOCK_F)
-do ix_col = max(ixMinCol,min_x_for_L(L)),Nx
-   ! Get column index for the d/dPhi1 terms
-   colIndex=getIndex(1,1,1,itheta,izeta,BLOCK_QN)
-\end{lstlisting}
 
@@
 If the temperature equilibration terms should be included, we have to add additional terms into the main matrix, when calculating  $\partial \Hat C/\partial \hat \Phi_1$.
-To begin with we need to include the different pre-factor (since in this case $f_0$ from \texttt{init\_f0()} is not used).
+These terms are different from the other contributions since now we need to multiply the output with $f_{0s} = f_{Ms}(\psi)e^{-Z_se\Phi_1(\theta,\zeta)/T_s}$ (instead of the first order distribution function $f_{1s}$).
+Because of this, we need to generate the Maxwellian distribution function for \texttt{iSpeciesB}.
 
 \begin{lstlisting}
-if (includeTemperatureEquilibrationTerm) then
-   ! Get the correct preFactor
-   preFactor = (-Zs(iSpeciesA)*alpha/Thats(iSpeciesA) -Zs(iSpeciesB)*alpha/  & 
-    Thats(iSpeciesB)) 
-   *exp(-Zs(iSpeciesA)*alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesA)-Zs(iSpeciesB) & 
-   *alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesB))
-\end{lstlisting}
-\noindent
-The second change originates from the fact that the collision operator in this temperature equilibration case, should multiply the Maxwellian for species $b$ and not $f_{1b}$ (as was previously the case in Section~\ref{sec:Jac2}).
-First, we have to initiate this distribution function
- \begin{lstlisting}
- if (includeTemperatureEquilibrationTerm) then
-    fM(ix) = expxb2(ix)*nhats(iSpeciesB)*mhats(iSpeciesB)*sqrt(mhats(iSpeciesB)) &
-     /(pi*sqrtpi*Thats(iSpeciesB)*sqrt(Thats(iSpeciesB)))
+if (includeTemperatureEquilibrationTerm .and. L==0) then
+   fM(ix) = sqrt(mhats(iSpeciesB)/Thats(iSpeciesB))*mhats(iSpeciesB)/Thats(iSpeciesB)&
+   * nhats(iSpeciesB)/(pi*sqrtpi)*expxb2(ix)
 end if   
 \end{lstlisting}
+
 \noindent
-and then use it to multiply to the total collision operator.
+Because of the \texttt{Phi1Hat} dependence in the zeroth order distribution function we get a different contribution to the Jacobian, as specified in the first term of Eq.~\eqref{eq:Jac1}.
+In the code, we need to use both the residual term \texttt{CHat} and the Jacobian terms \texttt{CHatJ} to obtain the correct result.
+
 \begin{lstlisting}
-  CHatTimesf = matmul(CHat,fM)
+  CHatTimesf = matmul(CHatJ + CHat*(-Zs(iSpeciesB)*alpha/Thats(iSpeciesB))&
+   *exp(-Zs(iSpeciesB)*alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesB)),fM)
 \end{lstlisting}
+
 \noindent
-Then we add this term into the main matrix, using the the same procedure as was explained in Section~\ref{sec:Jac1}
+What is left is to add this term into the main matrix, using the the same procedure as was explained in the previous section.
+
+\subsection*{Pure pitch angle scattering operator}
+The implementation of poloidal density variation in the case the pure pitch angle scattering collision operator is used (corresponding to the case \texttt{collisionoperator = 1}), is slightly easier because of the existing loop over \texttt{itheta}, \texttt{izeta} and since we are working with a scalar \texttt{CHat\_element}.
+Also, since $L > 0$ always for this term, no temperature equilibration has to be included.
+The only block we need to change is when we save the collision operator into the main matrix.
+
+We introduce the factor \texttt{preFactor} which is in normal cases equal to one, but when \texttt{poloidalVariationInCollisionOperator = .true.} it is equal to the exponential pre-factor containing $\Phi_1$ in Eq.~\eqref{eq:CollMod}.
+
+Before saving into the main matrix, we multiply the collision operator with this factor.
+
+ \begin{lstlisting}
+ call MatSetValue(matrix, index, colIndex, & 
+   -nu_n*CHat_element*preFactor, ADD_VALUES, ierr)
+  \end{lstlisting}
+
+\subsection*{Jacobian, the $\partial C/\partial \Phi_1$ term}
+Just as before, the   $\partial C/\partial f_1$ terms are the same as in the residual.
+What is different is the $\partial C/\partial \Phi_1$  term.
+The procedure to calculate this term is however very similar, to the case with the residual.
+The only difference is the different pre-factor which now also is multiplied with the distribution function from the state vector.
+
+ \begin{lstlisting}
+  preFactor = -Zs(iSpeciesA)*alpha/Thats(iSpeciesA)*exp(-Zs(iSpeciesA) &
+            *alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesA))*stateArray(index + 1)
+  \end{lstlisting}
+
+\section*{Current state of the project}
+Presently, the modifications discussed in the previous section have been implemented into the ``poloidalVariationInCollisionOperator'' branch of \texttt{SFINCS}.
+What remains is to benchmark the output with theoretical predictions, using for example the model described in Ref.~\cite{ref:Fulop}.
+
+Using the new implementation, for ``reasonable'' input values of impurity charge and main ion temperature gradient, no particular poloidal asymmetries have been observed, and the result is fairly similar if the original code is used instead.
+Attempts to benchmark with the model described in Ref.~\cite{ref:Fulop} suggests that much larger ion temperature gradients and impurity charge has to be used, in order to observe poloidal asymmetries in the perturbed density.
+The problem is, at this high gradients and impurity charge, it becomes difficult to converge.
+In general, it seems that a slightly higher \texttt{Ntheta} has to be used, when including poloidal variations in the collision operator.
 
@@
 
-\bibitem{ref:PolVarColl_stefan} S.~Buller, {\em Poloidal variation in collision operator} (2017).
+\bibitem{ref:PolVarColl_stefan} S.~Buller, \href{https://github.com/landreman/sfincs/blob/poloidalVariationInCollisionOperator/doc/poloidal_variation_in_collision_operator_2017-09-08.pdf}{ \em Poloidal variation in collision operator} (2017).
 
-\bibitem{ref:Coll1} M.~Landreman, {\em Technical Documentation for version 3 of SFINCS} (2014).
+\bibitem{ref:Coll1} M.~Landreman, \href{https://github.com/landreman/sfincs/blob/poloidalVariationInCollisionOperator/doc/20150507-01\%20Technical\%20documentation\%20for\%20version\%203\%20of\%20SFINCS.pdf}{\em Technical Documentation for version 3 of SFINCS} (2014).
 
-%\bibitem{simakov} A.~N.~Simakov, P.~Helander,
-%  {\em Phys. Plasmas} {\bf 16}, 042503 (2009).
-
-%\bibitem{nonAxis} P.~Helander, Theory of plasma confinement in non-axisymmetric magnetic fields (2013).
-
-%\bibitem{MH} L.~Råde and B.~Westergren, Mathematics Handbook for Science and Engineering, $5^{\mathrm{th}}$ edition, 2004. %\vspace{-5mm}
-
-%\bibitem{Abra} M.~Abramowitz and I.~A.~Stegun, Handbook of Mathematical Functions, $10^{\mathrm{th}}$ printing, 1972.
-
+\bibitem{ref:Fulop} T.~F\"{u}l\"{o}p, P.~Helander \href{http://aip.scitation.org/doi/abs/10.1063/1.873593}{\em Phys. Plasmas}{ \bf 6}, 3066 (1999) 
 \end{thebibliography}

### 2017-11-27 doc/PoloidalVariationInCollisionOperator_code.tex
Updated documentation

@@
 
+\usepackage{siunitx}
+
 \title{Implementation of poloidal density variation in collision operator}
@@
 
+\usepackage{tikzPackages}
+
 \begin{document}
@@
 }
-where $f_M(\psi)$ is the flux-function Maxwellian, $\Phi_1$ are first order variation of the electrostatic potential, and $Z_s,T_s$ are the charge and temperature of species $s$ respectively.
+where $f_M(\psi)$ is the flux-function Maxwellian, $\Phi_1$ is first order variation of the electrostatic potential, and $Z_s,T_s$ are the charge and temperature of species $s$ respectively.
 To implement this change one may simply modify the species density accordingly
@@
 Since this term does not include $f_{1s}$ it has to be treated differently compared to the other two cases, as will be seen shortly.
-In the code, the calculation of the residual has to be modified to include the factors specified in Eq.~\eqref{eq:CollMod}.
+In the code, the calculation of the residual (see Ref.~\cite{ref:Phi1Implem} for definition of the residual and Jacobian in this context) has to be modified to include the factors specified in Eq.~\eqref{eq:CollMod}.
 
-The contribution to the Jacobian are $\partial \hat C_{ab}^{L:f0}/\partial \hat f_{1s}$ and $\partial \hat C_{ab}^{L:f0}/\partial \hat \Phi_1$.
-The first is the same as in Eq.~\eqref{eq:CollMod} except that the collision operator does not act on the distribution function.
+The contribution to the Jacobian are $\delta \hat C_{ab}^{L:f0}/\delta \hat f_{1s}$ and $\delta \hat C_{ab}^{L:f0}/\delta\hat \Phi_1$.
+The first contribution is the same as in Eq.~\eqref{eq:CollMod} (without the temperature equilibration term) except that the collision operator does not act on the distribution function.
 In the second case we get
@@
 \label{eq:Jac1}
-\frac{\partial \hat C_{ab}^{L:f0}}{\partial \hat \Phi_1} = -\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha \hat C_{ab}\left\{f_{aM},f_{bM}\right\}e^{-\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha\hat\Phi_1(\theta,\zeta)}  \\ -\hat Z_b/\hat T_b \alpha\ \hat C_{ab}\left\{f_{a1},f_{bM}\right\}e^{-\hat Z_b/\hat T_b \alpha\hat\Phi_1(\theta,\zeta)}  - \hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta) \hat C_{ab}\left\{f_{aM},f_{b1}\right\}e^{-\hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta)}.
+\frac{\delta \hat C_{ab}^{L:f0}}{\delta \hat \Phi_1} = -\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha \hat C_{ab}\left\{f_{aM},f_{bM}\right\}e^{-\left(\hat Z_a/\hat T_a + \hat Z_b/\hat T_b\right) \alpha\hat\Phi_1(\theta,\zeta)}  \\ -\hat Z_b/\hat T_b \alpha\ \hat C_{ab}\left\{f_{a1},f_{bM}\right\}e^{-\hat Z_b/\hat T_b \alpha\hat\Phi_1(\theta,\zeta)}  - \hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta) \hat C_{ab}\left\{f_{aM},f_{b1}\right\}e^{-\hat Z_a/\hat T_a\alpha\hat\Phi_1(\theta,\zeta)}.
 }
@@
 \noindent
-Apart from the different pre-factor, the term in Eq.~\eqref{eq:Jac1} differs from term in $\partial \hat C_{ab}^{L:f0}/\partial \hat f_{1s}$ by acting on the first order distribution function $f_{1s}$.
+Apart from the different pre-factor, the term in Eq.~\eqref{eq:Jac1} differs from term in $\delta \hat C_{ab}^{L:f0}/\delta \hat f_{1s}$ by acting on the first order distribution function $f_{1s}$.
 For example, in the first case, terms such as $\partial f_{1s}/\partial x_s$ become
 \eq{
-\frac{1}{\partial f_{1s}}\frac{\partial f_{1s}}{\partial x_s} = \frac{\partial}{\partial x_s},
+\frac{\partial}{\partial f_{1s}}\frac{\partial f_{1s}}{\partial x_s} = \frac{\partial}{\partial x_s},
 }
-whereas in the second case, when we take the derivative with respect to $\Phi_1$, we have to include the derivative $\partial f_{1s}/\partial x_s$ (and $\partial f_{0s}/\partial x_s$ for temperature equilibration), as well.
+whereas in the second case, when we take the derivative with respect to $\Phi_1$, we have to include the derivative $\partial f_{1s}/\partial x_s$ (and $\partial f_{0s}/\partial x_s$ for temperature equilibration, which again is not present in the first contribution), as well.
 
@@
 The residual is evaluated in \texttt{evaluateResidual.F90} and the Jacobian in \texttt{evaluateJacobian.F90}.
-Both these two routines call the functions in \texttt{PopulateMatrix.F90}, where the actual assembly of the residual and Jacobian matrices are done.
+Both these two routines call the functions in \texttt{PopulateMatrix.F90}, where the actual assembly of the matrices used to compute the residual (by multiplying the matrix with the state-vector) and Jacobian, is done.
 As a consequence of this structure, we only need to modify \texttt{PopulateMatrix.F90} in order to include poloidal density variation in the collision operator.
@@
 \noindent
-Because of the \texttt{itheta}, \texttt{izeta} dependence we have redefined \texttt{CECD} and \texttt{nuDHat} to their equivalents \texttt{CECDpol} and \texttt{nuDHatpol} with space for their \texttt{itheta} and \texttt{izeta} components.
+When required, this pre-factor is overwritten using the charge and temperature of \texttt{iSpeciesB} instead. 
+Because of the \texttt{itheta}, \texttt{izeta} dependence we have replaced \texttt{CECD} and \texttt{nuDHat} to their equivalents \texttt{CECDpol} and \texttt{nuDHatpol} with space for their \texttt{itheta} and \texttt{izeta} components.
 
@@
 In the original version of the code, the temperature equilibration term is calculated in \\ \texttt{evaluateResidual.F90} by using \texttt{whichMatrix = 2} when calling \texttt{PopulateMatrix.F90}.
-The output is then multiplied with the first order distribution function $f_{0s}$.
-Since the original version of \texttt{SFINCS} already includes $\Phi_1$ in the definition of $f_{0s}$ (in the subroutine \texttt{init\_f0()} inside \texttt{PopulateMatrix.F90}), the same routines as were presented above can be used also if \texttt{includeTemperatureEquilibrationTerm = .true.}.
+The output is then multiplied with $f_{0s}$.
+Since the original version of \texttt{SFINCS} already includes $\Phi_1$ in the definition of $f_{0s}$ (in the subroutine \texttt{init\_f0()} inside \texttt{PopulateMatrix.F90}), the same routines as were presented above can be used also if\\ \texttt{includeTemperatureEquilibrationTerm = .true.}.
 
 \subsection*{Jacobian}
- If the modifications explained in the previous section are implemented, the $\partial C/\partial f_{1s}$ contribution to the Jacobian is at this point already taking the density variations into account.
-We need to add the calculation of $\partial C/\partial \Phi_1$.
+ If the modifications explained in the previous section are implemented, the $\delta C/\delta f_{1s}$ contribution to the Jacobian is at this point already taking the density variations into account.
+We need to add the calculation of $\delta C/\delta \Phi_1$.
 
-\subsubsection*{The $\partial C/\partial \Phi_1$ term}
+\subsubsection*{The $\delta C/\delta \Phi_1$ term}
 \label{sec:Jac1}
-We begin with treating the other two terms appearing in Eq.~\eqref{eq:Jac1} and will then look at the temperature equilibration term.
+We begin with treating the two latter terms appearing in Eq.~\eqref{eq:Jac1} and will then look at the temperature equilibration term.
 
@@
 \noindent
+Again, the same is done using the charge and temperature of \texttt{iSpeciesB} instead, whenever required. 
 The result is saved in the new matrices \texttt{nuDHatpolJ}, \texttt{CECDpolJ} etc.
@@
 
-In contrary to the residual (and the $\partial C/\partial f_{1s}$ terms ), when calculating $\partial C/\partial \Phi_1$ we have to include the first order distribution function before saving the result into the main matrix.
+In contrary to the residual (and the $\delta C/\delta f_{1s}$ terms ), when calculating $\delta C/\delta \Phi_1$ we have to include the first order distribution function before saving the result into the main matrix.
 For this purpose we first have to initiate this distribution function, by reading the appropriate terms in the current state vector.
@@
 \noindent
-The result is then saved into the $\partial \Hat C/\partial \hat \Phi_1$ block of the main matrix,
+The result is then saved into the $\delta \Hat C/\delta \hat \Phi_1$ block of the main matrix,
 
@@
 \label{sec:Jac2}
-If the temperature equilibration terms should be included, we have to add additional terms into the main matrix, when calculating  $\partial \Hat C/\partial \hat \Phi_1$.
+If the temperature equilibration terms should be included, we have to add additional terms into the main matrix, when calculating  $\delta \Hat C/\delta \hat \Phi_1$.
 These terms are different from the other contributions since now we need to multiply the output with $f_{0s} = f_{Ms}(\psi)e^{-Z_se\Phi_1(\theta,\zeta)/T_s}$ (instead of the first order distribution function $f_{1s}$).
@@
 \begin{lstlisting}
-  CHatTimesf = matmul(CHatJ + CHat*(-Zs(iSpeciesB)*alpha/Thats(iSpeciesB))&
-   *exp(-Zs(iSpeciesB)*alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesB)),fM)
+  CHatTimesf = matmul((CHatJ + CHat*(-Zs(iSpeciesB)*alpha/Thats(iSpeciesB))&
+   )*exp(-Zs(iSpeciesB)*alpha*Phi1Hat(itheta,izeta)/Thats(iSpeciesB)),fM)
 \end{lstlisting}
@@
 \noindent
+Written in terms of operators we have that \texttt{CHat} is  $\hat C_{ab}\left\{f_{aM},\cdot\right\}e^{-\hat Z_a/\hat T_a \alpha\hat\Phi_1(\theta,\zeta)}$ and \texttt{CHatJ} is $\hat C_{ab}\left\{f_{aM},\cdot\right\}\left ( -\hat Z_a/\hat T_a \alpha \right ) e^{-\hat Z_a/\hat T_a \alpha\hat\Phi_1(\theta,\zeta)}$.
+Here $\cdot$ means that the operator does not act on anything.
 What is left is to add this term into the main matrix, using the the same procedure as was explained in the previous section.
@@
 
-\subsection*{Jacobian, the $\partial C/\partial \Phi_1$ term}
-Just as before, the   $\partial C/\partial f_1$ terms are the same as in the residual.
-What is different is the $\partial C/\partial \Phi_1$  term.
+\subsection*{Jacobian, the $\delta C/\delta \Phi_1$ term}
+Just as before, the   $\delta C/\delta f_1$ terms are the same as in the residual.
+What is different is the $\delta C/\delta \Phi_1$  term.
 The procedure to calculate this term is however very similar, to the case with the residual.
@@
 Presently, the modifications discussed in the previous section have been implemented into the ``poloidalVariationInCollisionOperator'' branch of \texttt{SFINCS}.
-What remains is to benchmark the output with theoretical predictions, using for example the model described in Ref.~\cite{ref:Fulop}.
+What remains is to benchmark the output with theoretical predictions, using for example the model described in Ref.~\cite{ref:Fulop}\footnote{Note that in this model the collision operator is not linearised around a poloidally varying zeroth order distribution function (as is done here), but the poloidal variation enters through the perturbed distribution function in the ion-impurity collision operator, when computing the parallel friction force.}.
+Note that in this case only tokamak cases are considered.
 
-Using the new implementation, for ``reasonable'' input values of impurity charge and main ion temperature gradient, no particular poloidal asymmetries have been observed, and the result is fairly similar if the original code is used instead.
+The first step is to test the model with the output from the original \texttt{SFINCS} code, without including poloidal density variation.
+By setting the temperature gradient to zero we remove the effect from the collision operator and expect therefore the output from \texttt{SFINCS} to agree reasonably well with the predictions from the model.
+Example of typical \texttt{SFINCS}-input used to get the best agreement so far is:
+ \begin{lstlisting}
+ &speciesParameters
+ Zs = 1.0, 20.0
+ mHats = 1.0, 20.0
+ nHats = 4.0, 0.012
+ THats = 2.0, 2.0
+ dnHatdrNs = -40.0, -0.12
+ dTHatdrNs = -0.0, -0.0
+
+ &geometryParameters
+ inputRadialCoordinate = 3
+ inputRadialCoordinateForGradients = 3
+
+ rN_wish = 0.3
+
+ geometryScheme = 1
+ B0OverBBar = 60.0
+ GHat = 180.0
+ IHat = 0.9
+ iota = 0.5
+ epsilon_t = 0.1
+ epsilon_h = 0
+ epsilon_antisymm = 0
+ helicity_l = 1
+ helicity_n = 1
+
+ psiAHat = 30.0
+ aHat = 1.0
+/
+  \end{lstlisting}
+\noindent
+For this input, the orderings of various parameters described in Ref.~\cite{ref:Fulop} are reasonably satisfied (we have $\Delta\sim 0.3, \delta \sim 0.15, \epsilon \sim 0.1, n_zZ^2/n_i \sim 1.2$ and $\nu^*\sim 0.11$).
+If we use this input for varying magnetic field strength $B_0$ (as an example) and calculate the sine and cosine parts of the perturbed impurity density, we get Fig.~\ref{fig:1} and Fig.~\ref{fig:2} respectively.
+Solid lines correspond to the case when \texttt{SFINCS} output is used to calculate the perturbed density while dashed lines correspond to the analytical model.
+The two outputs show good agreement for the cosine parts while the sine is slightly deviating.
+Same scan using the new code (where poloidal density variation is included) gives exactly the same result.
+Some convergence issues appear when using the new implementation with high impurity charge (in this regime for $Z_s > 20$).
+No poloidal asymmetries have been observed so far, when using input to \texttt{SFINCS} which satisfies the orderings described in  Ref.~\cite{ref:Fulop} (which is required if we want to compare the \texttt{SFINCS} output with the model).
+
+\begin{figure}[H]
+ \setlength\figureheight{0.3\textheight}
+ \setlength\figurewidth{0.9\textwidth}
+    \centering
+    \input{figures/out.tex}
+    \caption{Sine part of the perturbed density for a scan in the magnetic field strength $B_0$.
+Solid line shows result calculated from \texttt{SFINCS} output while the dashed line is the prediction from the model.}
+            \label{fig:1}
+      \end{figure}
+
+  \begin{figure}[H]
+ \setlength\figureheight{0.3\textheight}
+ \setlength\figurewidth{0.9\textwidth}
+    \centering
+        \input{figures/out2.tex}
+       \caption{Cosine part of the perturbed density for a scan in the magnetic field strength $B_0$.
+Solid line shows result calculated from \texttt{SFINCS} output while the dashed line is the prediction from the model}
+                  \label{fig:2}
+  \end{figure}
+
+%For vaious scans, the result is a good agreement between the cosine parts of the perturbed density, while the sine part deviates somewhat.
+Below is an example of input parameters used for a scan in "X".
+As sen in Fig.
+
+%Test with the code using the new implementation, when poloidal density varation is included gives identical results as when poloidal density variations are neglected.
+The new code has a problem converging for large impurity charge.
+
+ %Using the new implementation, for ``reasonable'' input values of impurity charge and main ion temperature gradient, no particular poloidal asymmetries have been observed, and the result is fairly similar if the original code is used instead.
 Attempts to benchmark with the model described in Ref.~\cite{ref:Fulop} suggests that much larger ion temperature gradients and impurity charge has to be used, in order to observe poloidal asymmetries in the perturbed density.
@@
 In general, it seems that a slightly higher \texttt{Ntheta} has to be used, when including poloidal variations in the collision operator.
+
+ % a test comment
 
@@
 
+\bibitem{ref:Phi1Implem} A.~Moll\'en, \href{https://github.com/landreman/sfincs/blob/master/doc/Phi1_implementation_2016-01.pdf}{ \em Implementation of $\Phi_1$ in SFINCS} (2016).
+
 \bibitem{ref:Coll1} M.~Landreman, \href{https://github.com/landreman/sfincs/blob/poloidalVariationInCollisionOperator/doc/20150507-01\%20Technical\%20documentation\%20for\%20version\%203\%20of\%20SFINCS.pdf}{\em Technical Documentation for version 3 of SFINCS} (2014).
 
-\bibitem{ref:Fulop} T.~F\"{u}l\"{o}p, P.~Helander \href{http://aip.scitation.org/doi/abs/10.1063/1.873593}{\em Phys. Plasmas}{ \bf 6}, 3066 (1999) 
+\bibitem{ref:Fulop} T.~F\"{u}l\"{o}p, P.~Helander \href{http://aip.scitation.org/doi/abs/10.1063/1.873593}{\em Phys. Plasmas}{ \bf 6}, 3066 (1999) % fix this
+
+.
+
+%\bibitem{simakov} A.~N.~Simakov, P.~Helander,
+%  {\em Phys. Plasmas} {\bf 16}, 042503 (2009).
+
+%\bibitem{nonAxis} P.~Helander, Theory of plasma confinement in non-axisymmetric magnetic fields (2013).
+
+%\bibitem{MH} L.~Råde and B.~Westergren, Mathematics Handbook for Science and Engineering, $5^{\mathrm{th}}$ edition, 2004. %\vspace{-5mm}
+
+%\bibitem{Abra} M.~Abramowitz and I.~A.~Stegun, Handbook of Mathematical Functions, $10^{\mathrm{th}}$ printing, 1972.
+
 \end{thebibliography}
@@
 \end{document}
+
+% Some notes:
+Basically the structure is the following: the densoty might not always be corresponding to the row index (since we might have cross terms) but the distribution function is ALWAYS the column index, which is SpeciesB, that is why we use  f1b in the end

### 2017-12-05 doc/PoloidalVariationInCollisionOperator_code.tex
updated documentation

@@
 Rather than creating many if-statements in the current implementation of the collision operator, for sake of clarity we separate the two implementations with one if-statement directly before the various contributions to the collision operator are calculated.
-If \texttt{poloidalVariationInCollisionOperator = .false.} the first block is executed, which contains the original code.
+If \texttt{Phi1} is included and \texttt{poloidalVariationInCollisionOperator = .true.} the contribution from the collision operator is calculated taking poloidal density variation into account.
 
 \begin{lstlisting}
-if (.not. poloidalVariationInCollisionOperator) then   !! Added by AI (2017-09).
+ if (poloidalVariationInCollisionOperator .and. includePhi1 .and. includePhi1InKineticEquation) then
 \end{lstlisting}
 
-\noindent Otherwise, if \texttt{poloidalVariationInCollisionOperator = .true.}, \texttt{includePhi1 = .true.} and \texttt{includePhi1InKineticEquation = .true.} we proceed to the new block where poloidal density variation is included.
-
-\begin{lstlisting}
-else if (includePhi1 .and. includePhi1InKineticEquation) then! Do with poloidal density variation
-\end{lstlisting}
+Otherwise the second block is executed, which contains the original code.
 
@@
 \end{document}
-
-% Some notes:
-Basically the structure is the following: the densoty might not always be corresponding to the row index (since we might have cross terms) but the distribution function is ALWAYS the column index, which is SpeciesB, that is why we use  f1b in the end

### 2018-01-18 doc/manual/version3/inputParameters.tex
Added testingInnerProduct. adjointRHS construction and innerProduct now pass test.

--- 2017-08-08 doc/manual/version3/inputParameters.tex
@@
 {\ttfamily RHSMode} = 4: Solve the adjoint equation at constant $E_r$.
-This must be used with {\ttfamily geometryScheme == 5}, stellarator symmetric geometry, {\ttfamily includePhi1 = .false.}, {\ttfamily EParallelHat == 0}, {\ttfamily useDKESExBDrift == .false.}, {\ttfamily magneticDriftScheme == 0}, {\ttfamily includeXDotTerm == .true.}, {\ttfamily constraintScheme == 1}, and {\ttfamily collisionOperator == 0}.
+This must be used with {\ttfamily geometryScheme == 5}, stellarator symmetric geometry, {\ttfamily includePhi1 = .false.}, {\ttfamily EParallelHat == 0}, {\ttfamily useDKESExBDrift == .false.}, {\ttfamily magneticDriftScheme == 0}, {\ttfamily includeXDotTerm == .true.}, {\ttfamily constraintScheme == 1} of -1, {\ttfamily geometryScheme == 5}, and {\ttfamily collisionOperator == 0}.
 \\

### 2018-01-18 fortran/version3/globalVariables.F90

@@ -1,3 +1,5 @@
+
+!> File containing global variables.
 module globalVariables
 
   implicit none
@@ -379,7 +381,10 @@ module globalVariables
   !
   ! ********************************************************
 
-  integer, dimension(:), allocatable :: ns, ms
+  !> @var ns Values of n to use for Fourier derivatives.
+  !> @var ms Values of m to sue for Fourier derivatives.
+  integer, dimension(:), allocatable :: ns
+  integer, dimension(:), allocatable :: ms !> Values of n and m to use for Fourier derivatives.
   integer :: nMaxAdjoint = 0, mMaxAdjoint = 0
   integer :: NModesAdjoint = 0, NLambdas = 6
   PetscScalar, dimension(:,:), allocatable :: dRadialCurrentdLambda,dTotalHeatFluxdLambda,dBootstrapdLambda

### 2018-01-18 fortran/version3/populatedRHSdLambda.F90

@@ -5,12 +5,10 @@
 #include <petsc/finclude/petscvecdef.h>
 #endif
 
-! See evaluateResidual.f90 (where rhs is constructed). Much of the code has been copied
-
-! dRHSdLambda is a Vec which should be allocated by the calling subroutine
-! whichLambda corresponds to the component of B that is going to vary
-! whichMode corresponds to the index imn of ns and ms of the desired Fourier mode
-
+  !> Populates \f$\partial \mathbb{S}/\partial \lambda\f$ to compute explicit dependence of integrated quantiteis on geometry. See evaluateResidual.f90 (where \f$\mathbb{S}\f$ is constructed). Much of the code has been copied.
+  !! @param dMatrixdLambda Matrix to be populated. Should be allocated by calling subroutine.
+  !! @param whichLambda Indicates which component of magnetic field derivative is respect to. If = 0 \f$E_r\f$, = 1 \f$\hat{B}\f$, = 2 \f$\hat{B}^{\theta}\f$, = 3 \f$\hat{B}^{\zeta}\f$, = 4 \f$\hat{B}_{\theta}\f$, = 5 \f$\hat{B}_{\zeta}\f$, = 6 \f$\hat{D}\f$
+  !! @param whichMode Indicates index of ms and ns for derivative.
   subroutine populatedRHSdLambda(dRHSdLambda, whichLambda, whichMode)
 
     use petscvec

### 2018-01-18 fortran/version3/testingInnerProduct.F90

@@ -1,0 +1,109 @@
+#include "../PETScVersions.F90"
+#if (PETSC_VERSION_MAJOR < 3 || (PETSC_VERSION_MAJOR==3 && PETSC_VERSION_MINOR < 6))
+#include <finclude/petscvecdef.h>
+#else
+#include <petsc/finclude/petscvecdef.h>
+#endif
+
+!> This subroutine is used to test the innerProduct subroutine. The inner product
+!! of the forward solution with the corresponding adjoint RHS should yield the same
+!! result as heatFlux_vm_rHat, particleFlux_vm_rHat, or radial current using
+!! FSABVelocityUsingFSADensityOverRootFSAB2.
+!! @param forwardSolution Solution to forward equation
+!! @param whichAdjointRHS Indicates which integrated quantity is being differentiated. If = 1 (particle flux), = 2 (heat flux), = 3 (bootstrap current).
+!! @param whichSpecies Indicates species used for inner product. If = 0, corresponds to a species-summed quantity. If nonzero, indicates number of species.
+subroutine testingInnerProduct(forwardSolution,whichAdjointRHS,whichSpecies)
+
+  use globalVariables
+  use petscvec
+  use adjointDiagnostics
+
+  implicit none
+
+  integer :: whichSpecies, whichAdjointRHS
+  Vec :: forwardSolution
+  integer :: ispecies
+  PetscErrorCode :: ierr
+  PetscScalar :: result
+  PetscScalar :: quantityToCompare
+  Vec :: adjointRHSVec
+  VecScatter :: VecScatterContext
+  Vec :: adjointRHSOnProc0, forwardSolutionOnProc0
+  PetscScalar, pointer :: adjointRHSArray(:), forwardSolutionArray(:)
+  PetscScalar :: innerProductResult, residual
+
+  if (masterProc) then
+    print "(a,i4,a,i4,a)","--------- Computing innerProduct test with species ", whichSpecies," and whichAdjointRHS ", whichAdjointRHS,"-----------------------------------"
+  end if
+
+  !> Allocate adjointRHSVec
+  call VecCreateMPI(MPIComm, PETSC_DECIDE, matrixSize, adjointRHSVec, ierr)
+  call VecSet(adjointRHSVec, zero, ierr)
+
+  !> Populate RHS vec
+  if (masterProc) then
+    print*, "populating adjointRHS"
+  end if
+  call populateAdjointRHS(adjointRHSVec, whichAdjointRHS, whichSpecies)
[…]

### 2018-01-18 fortran/version3/validateInput.F90

@@ -1176,9 +1176,9 @@ subroutine validateInput()
       stop
     end if
     ! Check constraintScheme
-    if (constraintScheme /= -1) then
+    if (constraintScheme /= -1 .and. constraintScheme /= 1) then
       if (masterProc) then
-        print *,"Error! RHSMode>3 must be used with constraintScheme=1."
+        print *,"Error! RHSMode>3 must be used with constraintScheme=1 or constraintScheme=-1."
       end if
       stop
     end if

### 2018-02-09 doc/manual/version3/equations.tex
Working on updating the manual

@@
 associated with the adiabatic response.
-However, other terms considered in Ref \cite{Regana2013} are not included.
+However, other terms considered in Ref \cite{Regana2017} are not included.
 
@@
 
-This form of the drift-kinetic equation is nearly identical to equation (11) in Ref \cite{Regana2013}.
+This form of the drift-kinetic equation is nearly identical to equation (11) in Ref \cite{Regana2017}.
 The one difference (which should be small) is that at the end of (11), Garc\'{i}a-Rega\~{n}a

### 2018-02-09 doc/manual/version3/inputParameters.tex
Working on updating the manual

--- 2017-08-15 doc/manual/version3/inputParameters.tex
@@
 {Filters the magnetic field read from an input file.  Only Fourier modes $(m,n)$ for which $B_{m,n}$ is at least {\ttfamily min\_Bmn\_to\_load} will be included.}
+
+\myhrule
+
+\param{rippleScale}
+{real}
+{1.0}
+{\todo{Only when ?}}
+{\todo{Write description}}
+
+\myhrule
+
+\param{EParallelHatSpec\_bcdatFile}
+{string}
+{``''}
+{\todo{Only when ?}}
+{\todo{Write description}}
 
@@
 {Temperature of adiabatic species, in units of the reference temperature $\bar{T}$.}
+
+\myhrule
+
+\param{withNBIspec}
+{Boolean}
+{\false}
+{\todo{Whenever ?}}
+{\todo{Write description}}
+
+\myhrule
+
+\param{NBIspecZ}
+{real}
+{1.0}
+{\todo{Whenever \parlink{RHSMode} == 1 and \parlink{withNBIspec} == \true}}
+{\todo{Charge of NBI heated species, in units of the proton charge $e$.}}
+
+\myhrule
+
+\param{NBIspecNHat}
+{real}
+{0.0}
+{\todo{Whenever \parlink{RHSMode} == 1 and \parlink{withNBIspec} == \true}}
+{\todo{Density of NBI heated species, in units of the reference density $\bar{n}$.}}
 
@@
 
+\param{includePhi1InCollisionOperator}
+{Boolean}
+{\false}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and\\ \parlink{includePhi1inKineticEquation} == \true}
+{If true, the terms containing $\Phi_1$ will be included in the collision operator. 
+If false or \parlink{includePhi1inKineticEquation} == \false, and \parlink{includePhi1} = \true, then $\Phi_1$ is not included in the collision operator. 
+The default is false since it remains to show if this effect is important, and the memory and time requirements can increase significantly with this parameter.
+}
+
+\myhrule
+
 \param{quasineutralityOption}
@@
 \sfincs~must be run with an adiabatic species, \parlink{withAdiabatic}=\true, and only the adiabatic species and the first kinetic species are included in the quasineutrality equation (their input densities must fulfill quasineutrality).
-This is the quasineutrality considered in Ref \cite{Regana2013}.
+This is the quasineutrality considered in Ref \cite{Regana2017}.
 }
@@
 
+\myhrule
+
+\param{EParallelHatSpec}
+{1D array of reals}
+{0.0}
+{\todo{Whenever \parlink{RHSMode} == 1} ?}
+{\todo{Write description}}
+
 \section{The {\ttfamily resolutionParameters}~namelist}
@@
 \section{The {\ttfamily preconditionerOptions}~namelist}
+\label{sec:preconditionerOptions}
 
@@
 {\true}
-{Only when \parlink{nonlinear} = \true}
+{Only when \parlink{includePhi1} = \true}
 {If true, the nonlinear term will not be included in the preconditioner matrix, meaning the preconditioner matrix is the same
@@
 \section{\PETSc~commands}
+\label{sec:PETScCommands}
 

### 2018-02-09 doc/manual/version3/outputParameters.tex
Working on updating the manual

\chapter{Output Parameters}
\label{ch:output}

In this chapter we describe output parameters which are available in the \sfincs~output file \parlink{outputFileName} after a successful \sfincs~run. 
The output is in the \HDF~format, and the output can be looked at e.g. by using the {\ttfamily h5dump} command on Linux. 
Note that all parameters in the output file are case-sensitive.

\section{Convergence flags}

\section{Output fluxes and flows}

%The default values are usually best for the parameters in this namelist.
%
%\myhrule
%
%\param{outputFileName}
%{string}
%{``sfincsOutput.h5''}
%{Always}
%{Name which will be used for the HDF5 output file.  If this parameter is changed from the default value, \sfincsScan~ will not work.}
%
%\myhrule

### 2018-02-09 doc/manual/version3/runs.tex
Working on updating the manual

--- 2016-02-06 doc/manual/runs.tex
@@
 
-The most common choice is the default, 4.
+The most common choice is the default, 4.  The quantity \parlink{Er} in both the input and output files is defined as 
+$- d\hat{\Phi} / d \hat{r} = -(\bar{R} / \bar{\Phi}) d \Phi / d r$.
 
@@
 
-\section{Quasineutrality and variation of the electrostatic potential on the flux surface}
+%\section{Quasineutrality and variation of the electrostatic potential on the flux surface}
+\section{Nonlinear calculations, includePhi1 = {\ttfamily .true.}}
 \label{sec:qn}
@@
 In \sfincs, you can choose whether or not to
-include $\Phi_1$ using the paramter {\ttfamily \hyperlink{includePhi1}{includePhi1}} in the 
+include $\Phi_1$ using the parameter {\ttfamily \hyperlink{includePhi1}{includePhi1}} in the 
 {\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}}
-namelist.  If and only if $\Phi_1$ is included, a quasineutrality equation is solved
+namelist. 
+Furthermore, the parameters \parlink{includePhi1inKineticEquation}, \parlink{includePhi1InCollisionOperator} and \parlink{quasineutralityOption} can be used to control how $\Phi_1$ is included in the system of equations.
+If and only if $\Phi_1$ is included, a quasineutrality equation is solved
 at each point on the flux surface.  Due to these extra unknowns ($\Phi_1$) and extra equations
@@
 to the number of rows and columns associated with the kinetic equation, which depends not only on real space
-but also on velocity space and species.  Thus, there is very little extra computational cost associated
-with \parlink{includePhi1}.  You may wish to set \parlink{includePhi1} = \true~when
+but also on velocity space and species.  
+%Thus, there is very little extra computational cost associated
+%with \parlink{includePhi1}. 
+Thus, there is typically not a substantial increase in memory requirements associated with \parlink{includePhi1} (however, setting the preconditioner option \parlink{reusePreconditioner} = \false~can lead to a significant increase in memory requirements).
+The time requirements will normally increase when \parlink{includePhi1} is \true~because the system of equations is then nonlinear. 
+The nonlinear equations are solved with a Newton method using the Scalable Nonlinear Equations Solvers (SNES) in \PETSc. SNES internally employs KSP for the solution of its linear systems. 
+Note that for a nonlinear calculation the \sfincs~output file \parlink{outputFileName} will contain the results of each nonlinear iteration in the relevant output parameters.
+The final result of the calculation is stored in the last position of the arrays.
+The resolution parameters required for numerical convergence in nonlinear calculations are often similar to the parameters in the corresponding linear calculations (based on experience). 
+A sanity check to control that the nonlinear calculation is not incorrect (however it is not a guarantee for numerical convergence), is to check that the total flux-surface averaged particle flux for the lowest order distribution function $\displaystyle \left\langle \int d^3 v \, f_{s0} \left(\bm{v}_{\mathrm{d} s} + \bm{v}_{E}\right) \cdot \nabla \psi \right\rangle$ of each plasma species $s$ comes out as 0.
+This can e.g.
+be done by comparing {\ttfamily particleFlux\_vm0\_psiHat} to {\ttfamily particleFlux\_vE0\_psiHat} in the \sfincs~output file \parlink{outputFileName} and see that they are equal but with opposite sign.
+
+You may wish to set \parlink{includePhi1} = \true~when
 using \sfincs~to model an experiment, and set \parlink{includePhi1} = \false~when
 comparing \sfincs~with analytic theory or with another code that does not include $\Phi_1$. 
+In Ref~\cite{Mollen2018} there are examples of \sfincs~calculations with \parlink{includePhi1} = \true~for experimental data.
+
+\subsection{\PETSc~commands}
+In this section we summarize experiences with the Scalable Nonlinear Equations Solvers (SNES) in \PETSc, 
+and list some of the command-line flags associated with \PETSc~which can be useful for nonlinear calculations. 
+(See section~\ref{sec:PETScCommands} for general useful options in \sfincs.)
+
+If the nonlinear solver fails to converge with the default settings, a first useful thing to try is setting \parlink{reusePreconditioner} = \false~in the {\ttfamily \hyperref[sec:preconditionerOptions]{preconditionerOptions}} namelist (note that memory requirements might increase).
+Secondly, it can be useful to change the SNES linesearch type (see -snes\_linesearch\_type flag below) and to increase the absolute convergence tolerance (see -ksp\_atol flag below) for the KSP iterations and the relative convergence tolerance (see -snes\_rtol flag below) for the SNES iterations.
+Moreover, note that in each nonlinear iteration the nested linear iterations will start with the current residual.
+If the calculation progresses towards nonlinear convergence this implies that the starting residual norm in each nonlinear iteration is smaller and smaller.
+Therefore in a calculation with \parlink{includePhi1} = \true~it is sometimes useful to use a less strict requirement on \parlink{solverTolerance} in the {\ttfamily \hyperref[sec:resolutionParameters]{resolutionParameters}} namelist, since in the last nonlinear iterations the residual will otherwise end up being unnecessarily small causing long run times (this problem is also avoided by appropriate -ksp\_atol and -snes\_rtol flags).\\
+Some useful \PETSc~flags taken from Ref~\cite{PETSc2017}:\\
+
+\subsubsection{KSP}
+
+KSP is intended for solving nonsingular linear systems of the form $A x = b$. 
+The residual at iteration $k$ is $r_k \equiv b - A x_k$ and convergence is detected if
+\[
+\left\|r_k\right\|_2 < \mathrm{max} \left(\text{\ttfamily rtol} \cdot \left\|b\right\|_2, \text{\ttfamily atol}\right).
+\]
+Divergence is detected if
+\[
+\left\|r_k\right\|_2 > \text{\ttfamily dtol} \cdot \left\|b\right\|_2.
+\]
+\\
+
+\myhrule
+
+\PETScParam{-ksp\_atol {\normalfont \ttfamily$<$atol$>$}}
+{The absolute convergence tolerance absolute size of the (possibly preconditioned) residual norm in the KSP iterations. 
+The default is {\ttfamily 1e-50}, which in practice implies that absolute convergence will never be fulfilled. 
+Setting this to a value of the order {\ttfamily 1e-10} can be useful in a nonlinear calculation.}
+
+\myhrule
+
+\PETScParam{-ksp\_rtol {\normalfont \ttfamily$<$rtol$>$}}
+{The relative convergence tolerance, relative decrease in the (possibly preconditioned) residual norm in the KSP iterations. 
+The default in \PETSc~is {\ttfamily 1e-5}, 
+but the value in \sfincs~is set by \parlink{solverTolerance} in the {\ttfamily \hyperref[sec:resolutionParameters]{resolutionParameters}} namelist. This can be overwritten by using the flag.}
+
+\myhrule
+
+\PETScParam{-ksp\_divtol {\normalfont \ttfamily$<$dtol$>$}}
+{The divergence tolerance, amount (possibly preconditioned) residual norm can increase before KSP concludes that the method is diverging.
+The default is {\ttfamily 1e5}, which is usually a good setting.}
+
+\myhrule
+
+\PETScParam{-ksp\_max\_it {\normalfont \ttfamily$<$maxits$>$}}
+{Maximum number of iterations to use in KSP. 
+The default is {\ttfamily 1e4}, which is usually a good setting.}
+
+\subsubsection{SNES}
+
+The SNES class includes methods for solving systems of nonlinear equations of the form $\bm{F} \left(\bm{x}\right) = 0$, 
+where $\bm{F}~:~\Re^n \rightarrow \Re^n$. 
+\PETSc's default method for solving the nonlinear equation is Newton's method. The general
+form of the $n$-dimensional Newton's method for solving the system is 
+\[
+\bm{x}_{k + 1} = \bm{x}_{k} - \bm{J} \left(\bm{x}_{k}\right)^{-1} \bm{F} \left(\bm{x}_{k}\right),~~k = 0, 1, \ldots
+\]
+where $\bm{x}_{0}$ is an initial approximation to the solution and $\bm{J} \left(\bm{x}_{k}\right) = \bm{F}^\prime \left(\bm{x}\right)$, the Jacobian, is nonsingular at each iteration.
+In practice, the Newton iteration is implemented by the following two
+steps: (Approximately) solve $\bm{J} \left(\bm{x}_{k}\right) \Delta \bm{x}_{k} = - \bm{F} \left(\bm{x}_{k}\right)$,  update $\bm{x}_{k + 1} = \bm{x}_{k} + \Delta \bm{x}_{k}$. 
+In the output it is useful to pay attention to the reason why the SNES iterations have converged/diverged. 
+SNES iterations typically only finish successfully due to one of the following reasons (see SNES flags): {\ttfamily SNES\_CONVERGED\_FNORM\_ABS}, {\ttfamily SNES\_CONVERGED\_FNORM\_RELATIVE},\\ {\ttfamily SNES\_CONVERGED\_SNORM\_RELATIVE}.
+In the \sfincs~output file \parlink{outputFileName} there is also a flag {\ttfamily didNonlinearCalculationConverge} which specifies if the nonlinear calculation converged.\\
+
+\myhrule
+
+\PETScParam{-snes\_type  {\normalfont \ttfamily$<$type$>$}}
+{SNES includes several Newton-like nonlinear solvers based on line search techniques and trust region methods.
+A table of \PETSc~nonlinear solvers can be found in Ref~\cite{PETSc2017}.
+In \sfincs~the (default) method SNESNEWTONLS ({\ttfamily \bfseries -snes\_type} {\normalfont \ttfamily newtonls}) seems to work the best and we have not yet had any success with the other solvers.
+}
+
+\myhrule
+
+\PETScParam{-snes\_linesearch\_type  {\normalfont \ttfamily$<$type$>$}}
+{This flag is only used when {\ttfamily \bfseries -snes\_type} {\normalfont \ttfamily newtonls}.
+The different options are {\normalfont \ttfamily basic}, {\normalfont \ttfamily bt}, {\normalfont \ttfamily l2}, {\normalfont \ttfamily cp}, {\normalfont \ttfamily nleqerr}, {\normalfont \ttfamily shell}.
+The default is {\normalfont \ttfamily bt}, but in \sfincs~the experience is that {\normalfont \ttfamily l2} or {\normalfont \ttfamily cp} can often work better. 
+Depending on the chosen method there are other flags that can be used to control how the search is performed: {\ttfamily \bfseries -snes\_linesearch\_alpha}, {\ttfamily \bfseries -snes\_linesearch\_damping}, {\ttfamily \bfseries -snes\_linesearch\_maxstep}, {\ttfamily \bfseries -snes\_linesearch\_max\_it}, {\ttfamily \bfseries -snes\_linesearch\_minlambda}, {\ttfamily \bfseries -snes\_linesearch\_order}.
+We have not yet had any success with other values than the default for all these flags.
+}
+
+\myhrule
+
+\PETScParam{-snes\_atol {\normalfont \ttfamily$<$atol$>$}}
+{The absolute convergence tolerance absolute size of the residual norm in the SNES iterations. 
+The default is {\ttfamily ?}, but usually this flag can be ignored. 
+If $\left\|\bm{F} \left(\bm{x}_{k}\right)\right\|_2 \leq \text{\ttfamily atol}$ then the output flag {\ttfamily SNES\_CONVERGED\_FNORM\_ABS} is specified.}
+
+\myhrule
+
+\PETScParam{-snes\_rtol {\normalfont \ttfamily$<$rtol$>$}}
+{The relative convergence tolerance, relative decrease in the residual norm in the SNES iterations. 
+The default in \PETSc~is {\ttfamily ?}, setting this to {\ttfamily 1e-6} can be a good option. 
+If $\left\|\bm{F} \left(\bm{x}_{k}\right)\right\|_2 \leq \text{\ttfamily rtol} \cdot \left\|\bm{F} \left(\bm{x}_{0}\right)\right\|_2$ then the output flag {\ttfamily SNES\_CONVERGED\_FNORM\_RELATIVE} is specified.}
+
+\myhrule
+
+\PETScParam{-snes\_stol {\normalfont \ttfamily$<$stol$>$}}
+{The convergence tolerance in terms of the norm of the change in the solution between steps in the SNES iterations.
+The default is {\ttfamily ?}, but usually this flag can be ignored. 
+If $\left\|\Delta \bm{x}\right\|_2 \leq \text{\ttfamily stol} \cdot \left\|\bm{x}\right\|_2$ then the output flag {\ttfamily SNES\_CONVERGED\_SNORM\_RELATIVE} is specified.}
+
+\myhrule
+
+\PETScParam{-snes\_max\_it {\normalfont \ttfamily$<$maxit$>$}}
+{The maximum number of iterations before SNES stops. 
+The default is {\ttfamily 50}, which is usually a good setting. (Sometimes convergence is slow in SNES but after more than 50 iterations it is unlikely that convergence will be reached.)}
+
+\myhrule
+
+\PETScParam{-snes\_view}
+{Dumps detailed information to stdout related to the nonlinear solver.}
+
+\myhrule
+
+\PETScParam{-snes\_linesearch\_monitor}
+{Dumps detailed information to stdout at every iteration of the nonlinear solver to display the iteration's progress when a line search is performed.}
 

### 2018-02-15 doc/manual/version3/outputParameters.tex
Working on manual

@@
 
-In this chapter we describe output parameters which are available in the \sfincs~output file \parlink{outputFileName} after a successful \sfincs~run. 
+\newcommand{\outputparam}[4]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries \hypertarget{#1}{#1}}\\{\it Type}: #2\\{\it Available when}: #3\\{\it Meaning}: #4}}
+\newcommand{\outputparamWithUnderscore}[5]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries \hypertarget{#2}{#1}}\\{\it Type}: #3\\{\it Available when}: #4\\{\it Meaning}: #5}}
+\newcommand{\underscoreparlink}[2]{{\ttfamily \hyperlink{#2}{#1}}}
+
+In this chapter we describe output parameters which are available in the \sfincs~output file \parlink{outputFileName} after a %successful 
+\sfincs~run. 
 The output is in the \HDF~format, and the output can be looked at e.g. by using the {\ttfamily h5dump} command on Linux. 
 Note that all parameters in the output file are case-sensitive. 
+Moreover, many of the input parameters in chapter~\ref{ch:input} also exist in the output file and we refrain from repeating them here.
 
-\section{Convergence flags}
+\section{Convergence and test parameters}
 
-\section{Output fluxes and flows}
+\myhrule
+
+\outputparam{finished}
+{integer}
+{\sfincs~has finished.}
+{If this variable exists, then \sfincs~reached the end of all requested computations and exited gracefully. 
+Note that this does not guarantee that a nonlinear calculation (\parlink{includePhi1} == \true) has converged.}
+
+\myhrule
+
+\outputparam{integerToRepresentTrue}
+{integer}
+{Always}
+{Since \HDF~does not have a Boolean datatype, this integer value is used to represent \true.}
+
+\myhrule
+
+\outputparam{integerToRepresentFalse}
+{integer}
+{Always}
+{Since \HDF~does not have a Boolean datatype, this integer value is used to represent \false.}
+
+\myhrule
+
+\outputparam{didNonlinearCalculationConverge}
+{integer}
+{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true, \parlink{includePhi1} == \true and \parlink{finished} available.}
+{If this variable is \true~the nonlinear iteration has converged. Interpret this Boolean data by comparing to \parlink{integerToRepresentTrue}/\parlink{integerToRepresentFalse}.}
+
+\myhrule
+
+\outputparam{lambda}
+{real}
+{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true and \parlink{includePhi1} == \true.}
+{Lagrange multiplier associated with the constraint that $\left\langle \Phi_1\right\rangle = 0$, which is included in the quasineutrality equation. %Should be within machine precision of 0.
+This parameter should come out as negligible in quasineutrality, e.g. its size should be significantly smaller than the maximum charge density of the plasma species $Z_s n_s$.
+\todo{Is this correct?}}
+
+\myhrule
+
+\outputparam{NIterations}
+{integer}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{Number of iterations in the nonlinear calculation. (In a linear calculation this will be 1.)}
+
+\myhrule
+
+\section{Radial fluxes and flows}
+
+A detailed description of the most important output fluxes and flows are given in the Technical Documentation for \sfincs. 
+Here we give a concise version of some of the quantities. 
+Transforming the dimension of the output quantities into SI units is typically done with the normalizations 
+$\bar{B}$, $\bar{R}$, $\bar{n}$, $\bar{m}$, $\bar{T}$, $\bar{v}$, $\bar{\Phi}$ (see section~\ref{sec:normalizations}).
+
+\subsubsection{Radial fluxes}
+
+The radial fluxes have the form 
+\[
+\left\langle \bm{X} \cdot \nabla Y \right\rangle
+\]
+where $Y$ is a flux-surface label and $\left< \ldots \right>$ denotes a flux surface average.
+In \sfincs~there are 4 possible labels: $\hat{\psi}$, $\psi_N$, $\hat{r}$ and $r_N$ (see section~\ref{sec:radialCoordinates}), indicated by the suffix {\ttfamily \_psiHat}, {\ttfamily \_psiN}, {\ttfamily \_rHat} or {\ttfamily \_rN} in the output variable name.
+Below we will only present the description of the {\ttfamily XXX\_psiHat} variables, but the description of the other variables are readily obtained by noting that 
+all flux-surface labels are dimensionless which implies that the normalization factor will be the same.
+E.g.
+when $\displaystyle \left\langle \bm{X} \cdot \nabla \hat{\psi} \right\rangle = \text{\ttfamily NormalizationFactor} \cdot \text{\ttfamily XXX\_psiHat}$ then also $\displaystyle \left\langle \bm{X} \cdot \nabla \hat{r} \right\rangle = \text{\ttfamily NormalizationFactor} \cdot \text{\ttfamily XXX\_rHat}$ applies with the same {\ttfamily NormalizationFactor}.
+
+For the radial flux of each quantity (particle, momentum and energy), and for each radial coordinate, there is an output quantity associated with the zeroth-order $f_{s0}$ or with the full distribution function $f_s = f_{s0} + f_{s1}$, and there is also a quantity associated with the radial magnetic drift, with the radial $\bm{E} \times \bm{B}$-drift or with the total drift $\bm{v}_{ds} = \bm{v}_{ms} + \bm{v}_{E}$.
+Here we only present all possible combinations for the particle flux, but the pattern is the same for the momentum and energy flux.
+
+\myhrule
+
+\outputparamWithUnderscore{particleFlux\_vd\_psiHat}
+{particleFlux_vd_psiHat}
+{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true and \parlink{includePhi1} == \true.}
+{$\displaystyle \text{\ttfamily particleFlux\_vd\_psiHat} = \frac{\bar{R}}{\bar{n} \bar{v}} \left\langle \int d^3v f_s \, \bm{v}_{ds} \cdot \nabla \hat{\psi}\right\rangle$ for all plasma species.
+The final result is in the last index corresponding to the last nonlinear iteration.}
+
+\myhrule
+
+\outputparamWithUnderscore{particleFlux\_vm\_psiHat}
+{particleFlux_vm_psiHat}
+{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{$\displaystyle \text{\ttfamily particleFlux\_vm\_psiHat} = \frac{\bar{R}}{\bar{n} \bar{v}} \left\langle \int d^3v f_s \, \bm{v}_{ms} \cdot \nabla \hat{\psi}\right\rangle$ for all plasma species.
+The final result is in the last index corresponding to the last nonlinear iteration.}
+
+\myhrule
+
+\outputparamWithUnderscore{particleFlux\_vE\_psiHat}
+{particleFlux_vE_psiHat}
+{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true and \parlink{includePhi1} == \true.}
+{$\displaystyle \text{\ttfamily particleFlux\_vE\_psiHat} = \frac{\bar{R}}{\bar{n} \bar{v}} \left\langle \int d^3v f_s \, \bm{v}_{E} \cdot \nabla \hat{\psi}\right\rangle$ for all plasma species.
+The final result is in the last index corresponding to the last nonlinear iteration.}
+
+\myhrule
+
+\outputparamWithUnderscore{particleFlux\_vm0\_psiHat}
+{particleFlux_vm0_psiHat}
+{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{$\displaystyle \text{\ttfamily particleFlux\_vm0\_psiHat} = \frac{\bar{R}}{\bar{n} \bar{v}} \left\langle \int d^3v f_{s0} \, \bm{v}_{ms} \cdot \nabla \hat{\psi}\right\rangle$ for all plasma species.
+The final result is in the last index corresponding to the last nonlinear iteration.
+This variable should cancel \underscoreparlink{particleFlux\_vE0\_psiHat}{particleFlux_vE0_psiHat} in the last nonlinear iteration since $f_{s0}$ cannot drive radial transport (i.e.
+in a linear calculations this quantity should be 0 to high precision unless there is a radial current in the magnetic equilibrium).}
+
+\myhrule
+
+\outputparamWithUnderscore{particleFlux\_vE0\_psiHat}
+{particleFlux_vE0_psiHat}
+{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true and \parlink{includePhi1} == \true.}
+{$\displaystyle \text{\ttfamily particleFlux\_vE0\_psiHat} = \frac{\bar{R}}{\bar{n} \bar{v}} \left\langle \int d^3v f_{s0} \, \bm{v}_{E} \cdot \nabla \hat{\psi}\right\rangle$ for all plasma species.
+The final result is in the last index corresponding to the last nonlinear iteration.
+This variable should cancel \underscoreparlink{particleFlux\_vm0\_psiHat}{particleFlux_vm0_psiHat} in the last nonlinear iteration since $f_{s0}$ cannot drive radial transport.}
+
+\myhrule
+
+\outputparamWithUnderscore{particleFlux\_vd1\_psiHat}
+{particleFlux_vd1_psiHat}
+{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true and \parlink{includePhi1} == \true.}
+{$\displaystyle \text{\ttfamily particleFlux\_vd1\_psiHat} = \text{\ttfamily particleFlux\_vm\_psiHat}~+~\text{\ttfamily particleFlux\_vE0\_psiHat}$.
+The rationale for defining this quantity is that (at least for $E_r = 0$) it gives the particle flux that would be expected if \parlink{includePhi1} == \false.}
+
+\myhrule
+
+\outputparamWithUnderscore{particleFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}}
+{particleFlux_vXN_Y}
+{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{See above for various $\textcolor{magenta}{XN}$.}
+{For different options of $\textcolor{magenta}{XN}$ see output {\ttfamily particleFlux} parameters defined above. 
+For different options of $\textcolor{magenta}{Y}$ see flux-surface labels (section~\ref{sec:radialCoordinates}). 
+$\displaystyle \text{\ttfamily particleFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}} \left\langle \int d^3v f_{s\textcolor{magenta}{N}} \, \bm{v}_{\textcolor{magenta}{X}s} \cdot \nabla \textcolor{magenta}{Y}\right\rangle$ for all plasma species.
+The final result is in the last index corresponding to the last nonlinear iteration.}
+
+\myhrule
+
+%\outputparamWithUnderscore{momentumFlux\_vm\_psiHat}
+%{momentumFluxvmpsiHat}
+%{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+%{Always}
+%{$\displaystyle \text{\ttfamily momentumFlux\_vm\_psiHat} = \frac{\bar{R}}{\bar{n} \bar{v}^2 \bar{B} \bar{m}} \left\langle \int d^3v f_s \, B m_s v_{\|} \bm{v}_{ms} \cdot \nabla \hat{\psi}\right\rangle$ for all plasma species.
+The final result is in the last index corresponding to the last nonlinear iteration.}
+%
+%\myhrule
+
+\outputparamWithUnderscore{momentumFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}}
+{momentumFlux_vXN_Y}
+{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{See above for various $\textcolor{magenta}{XN}$.}
+{For different options of $\textcolor{magenta}{XN}$ see  
+\underscoreparlink{particleFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}}{particleFlux_vXN_Y}.
+For different options of $\textcolor{magenta}{Y}$ see flux-surface labels (section~\ref{sec:radialCoordinates}). 
+$\displaystyle \text{\ttfamily momentumFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}^2 \bar{B} \bar{m}} \left\langle \int d^3v f_{s\textcolor{magenta}{N}} \, B m_s v_{\|} \bm{v}_{\textcolor{magenta}{X}s} \cdot \nabla \textcolor{magenta}{Y}\right\rangle$ for all plasma species.
+The final result is in the last index corresponding to the last nonlinear iteration.}
+
+\myhrule
+
+\outputparamWithUnderscore{heatFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}}
+{heatFlux_vXN_Y}
+{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{See above for various $\textcolor{magenta}{XN}$.}
+{For different options of $\textcolor{magenta}{XN}$ see  
+\underscoreparlink{particleFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}}{particleFlux_vXN_Y}.
+For different options of $\textcolor{magenta}{Y}$ see flux-surface labels (section~\ref{sec:radialCoordinates}). 
+$\displaystyle \text{\ttfamily heatFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}^3 \bar{m}} \left\langle \int d^3v f_{s\textcolor{magenta}{N}} \, \frac{m_s v^2}{2} \bm{v}_{\textcolor{magenta}{X}s} \cdot \nabla \textcolor{magenta}{Y}\right\rangle$ for all plasma species.
+The final result is in the last index corresponding to the last nonlinear iteration.}
+
+\myhrule
+
+\outputparam{flow}
+{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{$\displaystyle \text{\ttfamily flow} =  \frac{1}{\bar{n} \bar{v}} \int d^3v \, v_{\|} f_s$.}
+
+\myhrule
+
+\outputparam{velocityUsingFSADensity}
+{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{{\ttfamily velocityUsingFSADensity} =  \parlink{flow}~/~\parlink{nHats}.}
+
+\myhrule
+
+\outputparam{velocityUsingTotalDensity}
+{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{{\ttfamily velocityUsingTotalDensity} =  \parlink{flow}~/~\parlink{totalDensity}.}
+
+\myhrule
+
+\outputparam{MachUsingFSAThermalSpeed}
+{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{{\ttfamily MachUsingFSAThermalSpeed} =  (\parlink{mHats}~/~\parlink{THats})${}^{1/2}$~$\cdot$~\parlink{flow}~/~\parlink{nHats}.}
+
+\myhrule
+
+\outputparam{FSABFlow}
+{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{$\displaystyle \text{\ttfamily FSABFlow} =  \frac{1}{\bar{v} \bar{B} \bar{n}} \left\langle B \int d^3v \, v_{\|} f_s \right\rangle$.}
+
+\myhrule
+
+\outputparam{FSABVelocityUsingFSADensity}
+{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{{\ttfamily FSABVelocityUsingFSADensity} =  \parlink{FSABFlow}~/~\parlink{nHats}.}
+
+\myhrule
+
+\outputparam{FSABVelocityUsingFSADensityOverB0}
+{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{{\ttfamily FSABVelocityUsingFSADensityOverB0} =  ($\bar{B}$~/~$B_0$)~$\cdot$~\parlink{FSABFlow}~/~\parlink{nHats}.}
+
+\myhrule
+
+\outputparam{FSABVelocityUsingFSADensityOverRootFSAB2}
+{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{{\ttfamily FSABVelocityUsingFSADensityOverRootFSAB2} =  (1~/~\parlink{nHats})~$\cdot$~\parlink{FSABFlow}~/~(\parlink{FSABHat2})${}^{1/2}$.}
+
+\myhrule
+
+\outputparam{jHat}
+{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{$\displaystyle \text{\ttfamily jHat} = \frac{\bm{j} \cdot \bm{b}}{e \bar{n} \bar{v}} = \frac{1}{e \bar{n} \bar{v}} \sum_s Z_s e \int d^3v \, v_{\|} f_s = \sum_s Z_s \cdot \text{\parlink{flow}}_s$.}
+
+\myhrule
+
+\outputparam{FSABjHat}
+{$\text{\ttfamily NIterations}$ array of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{$\displaystyle \text{\ttfamily FSABjHat} =  \sum_s Z_s \cdot \text{\parlink{FSABFlow}}_s$.}
+
+\myhrule
+
+\outputparam{FSABjHatOverB0}
+{$\text{\ttfamily NIterations}$ array of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{$\displaystyle \text{\ttfamily FSABjHatOverB0} =  \frac{\bar{B}}{B_0} \text{\parlink{FSABjHat}}$.}
+
+\myhrule
+
+\outputparam{FSABjHatOverRootFSAB2}
+{$\text{\ttfamily NIterations}$ array of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{$\displaystyle \text{\ttfamily FSABjHatOverRootFSAB2} =  \frac{\bar{\text{\parlink{FSABjHat}}}}{\left(\text{\parlink{FSABHat2}}\right)^{1/2}}$.}
+
+\myhrule
+
+\section{Distributions}
+
+\myhrule
+
+\outputparam{densityPerturbation}
+{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{$\displaystyle \text{\ttfamily densityPerturbation} =  \frac{1}{\bar{n}} \int d^3v f_{s1}$.}
+
+\myhrule
+
+\outputparam{totalDensity}
+{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{{\ttfamily totalDensity} = \parlink{nHats}~+~\parlink{densityPerturbation}.}
+
+\myhrule
+
+\outputparam{FSADensityPerturbation}
+{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{$\displaystyle \text{\ttfamily FSADensityPerturbation} =  \frac{1}{\bar{n}} \left\langle \int d^3v f_{s1} \right\rangle$. 
+Should be within machine precision of 0.}
+
+\myhrule
+
+\outputparam{pressurePerturbation}
+{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{$\displaystyle \text{\ttfamily densityPerturbation} =  \frac{1}{\bar{n} \bar{T}} \frac{m_s}{3} \int d^3v \, v^2 f_{s1}$.}
+
+\myhrule
+
+\outputparam{totalPressure}
+{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{{\ttfamily totalPressure} = \parlink{nHats} $\cdot$ \parlink{THats}~+~\parlink{pressurePerturbation}.}
+
+\myhrule
+
+\outputparam{FSAPressurePerturbation}
+{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{$\displaystyle \text{\ttfamily FSADensityPerturbation} =  \frac{1}{\bar{n} \bar{T}} \frac{m_s}{3} \left\langle \int d^3v \, v^2 f_{s1} \right\rangle$. 
+Should be within machine precision of 0.}
+
+\myhrule
+
+\outputparam{delta\_f}
+{$M_x \times M_{\xi} \times M_{\zeta} \times M_{\theta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{export\_delta\_f} == \true.}
+{?}
+
+\myhrule
+
+\section{Miscellaneous}
+
+\myhrule
+
+\outputparam{VPrimeHat}
+{real}
+{Always}
+{$\displaystyle \text{\ttfamily VPrimeHat} =  \int_{0}^{2 \pi} d \theta \int_{0}^{2 \pi} d \zeta \frac{\bar{B}}{\bar{R}} \frac{1}{\nabla \psi \cdot \nabla \theta \times \nabla \zeta}$.}
+
+\myhrule
+
+\outputparam{FSABHat2}
+{real}
+{Always}
+{$\displaystyle \text{\ttfamily FSABHat2} =  \frac{1}{\bar{B}^2} \left\langle B^2 \right\rangle$.}
+
+\myhrule
 

### 2018-02-15 doc/manual/version3/equations.tex
Working on manual

@@
 Throughout this section, gradients and other partial derivatives are taken
-at constant $x_a = v/\sqrt{2T_a/m_a}$ and at constant $\xi = v_{||} / v$, unless denoted otherwise with subscripts.
+at constant $x_s = v/\sqrt{2T_s/m_s}$ and at constant $\xi = v_{||} / v$, unless denoted otherwise with subscripts.
 We use $s$ to denote the species, $r$ to denote any radial coordinate, (expressions are independent
 of the specific radial coordinate used,)
-$C_a$ to denote the collision operator, and $S_a$ to denote the source-sink term.
+$C_s$ to denote the collision operator, and $S_s$ to denote the source-sink term.
 A more detailed discussion of the equations implemented in the code can be found
@@
  \frac{\partial f_{s1}}{\partial \xi} \nonumber \\
-&&-(\vect{v}_{ma} \cdot\nabla r) \frac{Z_s e}{2 T_s x_s} \frac{d\Phi_0}{dr} \frac{\partial f_{a1}}{\partial x_s} \nonumber \\
-&&+ (\vect{v}_{ma} \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM}
- = C_a + S_a \nonumber
+&&-(\vect{v}_{ms} \cdot\nabla r) \frac{Z_s e}{2 T_s x_s} \frac{d\Phi_0}{dr} \frac{\partial f_{s1}}{\partial x_s} \nonumber \\
+&&+ (\vect{v}_{ms} \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM}
+ = C_s + S_s \nonumber
 \end{eqnarray}
@@
 -\frac{v_{||}}{B^2} \frac{d\Phi_0}{dr} \vect{b}\times\nabla B \times \nabla r \right]
-\left( \frac{\partial f_{a1}}{\partial v_{||}} \right)_{\mu} \nonumber \\
-&& + (\vect{v}_{ma} \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM}
-=C_a + S_a \nonumber
+\left( \frac{\partial f_{s1}}{\partial v_{||}} \right)_{\mu} \nonumber \\
+&& + (\vect{v}_{ms} \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM}
+=C_s + S_s \nonumber
 \end{eqnarray}
@@
 \parlink{magneticDriftScheme} = 0 \;\;\; (Default) \\
-\parlink{includePhi1} = \false \;\;\; (Default) \\
-\parlink{includeRadialExBDrive} = \false \;\;\; (Default) \\
-\parlink{nonlinear} = \false \;\;\; (Default)
+\parlink{includePhi1} = \false \;\;\; (Default) %\\
+%\parlink{includeRadialExBDrive} = \false \;\;\; (Default) \\
+%\parlink{nonlinear} = \false \;\;\; (Default)
 
@@
  \frac{\partial f_{s1}}{\partial \xi} \nonumber \\
-&&+ (\vect{v}_{ma} \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM}
- = C_a + S_a \nonumber
+&&+ (\vect{v}_{ms} \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM}
+ = C_s + S_s \nonumber
 \end{eqnarray}
@@
 \parlink{magneticDriftScheme} = 1 \;\;\; (Not default) \\
-\parlink{includePhi1} = \false \;\;\; (Default) \\
-\parlink{includeRadialExBDrive} = \false \;\;\; (Default) \\
-\parlink{nonlinear} = \false \;\;\; (Default)
+\parlink{includePhi1} = \false \;\;\; (Default) %\\
+%\parlink{includeRadialExBDrive} = \false \;\;\; (Default) \\
+%\parlink{nonlinear} = \false \;\;\; (Default)
 
@@
 However, other terms considered in Ref \cite{Regana2017} are not included.
+\todo{This is not valid anymore}
 
@@
  \frac{\partial f_{s1}}{\partial \xi} \nonumber \\
-&&-(\vect{v}_{ma} \cdot\nabla r) \frac{Z_s e}{2 T_s x_s} \frac{d\Phi_0}{dr} \frac{\partial f_{a1}}{\partial x_s} \nonumber \\
+&&-(\vect{v}_{ms} \cdot\nabla r) \frac{Z_s e}{2 T_s x_s} \frac{d\Phi_0}{dr} \frac{\partial f_{s1}}{\partial x_s} \nonumber \\
 &&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
-&&+ (\vect{v}_{ma} \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM}
- = C_a + S_a \nonumber
+&&+ (\vect{v}_{ms} \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM}
+ = C_s + S_s \nonumber
 \end{eqnarray}
@@
 -\frac{v_{||}}{B^2} \frac{d\Phi_0}{dr} \vect{b}\times\nabla B \times \nabla r \right]
-\left( \frac{\partial f_{a1}}{\partial v_{||}} \right)_{\mu} \nonumber \\
+\left( \frac{\partial f_{s1}}{\partial v_{||}} \right)_{\mu} \nonumber \\
 &&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
-&& + (\vect{v}_{ma} \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM}
-=C_a + S_a \nonumber
+&& + (\vect{v}_{ms} \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM}
+=C_s + S_s \nonumber
 \end{eqnarray}
@@
 \parlink{magneticDriftScheme} = 0 \;\;\; (Default) \\
-\parlink{includePhi1} = \true \;\;\; (Not default) \\
-\parlink{includeRadialExBDrive} = \false \;\;\; (Default) \\
-\parlink{nonlinear} = \false \;\;\; (Default)
+\parlink{includePhi1} = \true \;\;\; (Not default) %\\
+%\parlink{includeRadialExBDrive} = \false \;\;\; (Default) \\
+%\parlink{nonlinear} = \false \;\;\; (Default)
 
@@
 &&+ \left[
--(\vect{v}_{ma} \cdot\nabla r) \frac{Z_s e}{2 T_s x_s} \frac{d\Phi_0}{dr} 
+-(\vect{v}_{ms} \cdot\nabla r) \frac{Z_s e}{2 T_s x_s} \frac{d\Phi_0}{dr} 
 -\frac{Z_s e \xi}{v_s m_s}
-\right] \frac{\partial f_{a1}}{\partial x_s} \nonumber \\
+\right] \frac{\partial f_{s1}}{\partial x_s} \nonumber \\
 &&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
-&& + (\vect{v}_E \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr}  + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM} \nonumber \\
-&&+ (\vect{v}_{ma} \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM}
- = C_a + S_a \nonumber
+&& + (\vect{v}_E \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr}  + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM} \nonumber \\
+&&+ (\vect{v}_{ms} \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM}
+ = C_s + S_s \nonumber
 \end{eqnarray}
@@
 -\frac{v_{||}}{B^2} \frac{d\Phi_0}{dr} \vect{b}\times\nabla B \times \nabla r \right]
-\left( \frac{\partial f_{a1}}{\partial v_{||}} \right)_{\mu} \nonumber \\
+\left( \frac{\partial f_{s1}}{\partial v_{||}} \right)_{\mu} \nonumber \\
 &&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
-&& + (\vect{v}_E \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr}  + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM} \nonumber \\
-&& + (\vect{v}_{ma} \cdot \nabla r) \left[ \frac{1}{n_a} \frac{dn_a}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{aM}
-=C_a + S_a \nonumber
+&& + (\vect{v}_E \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr}  + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM} \nonumber \\
+&& + (\vect{v}_{ms} \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM}
+=C_s + S_s \nonumber
 \end{eqnarray}
@@
 \parlink{magneticDriftScheme} = 0 \;\;\; (Default) \\
-\parlink{includePhi1} = \true \;\;\; (Not default) \\
-\parlink{includeRadialExBDrive} = \true \;\;\; (Not default) \\
-\parlink{nonlinear} = \true \;\;\; (Not default)
+\parlink{includePhi1} = \true \;\;\; (Not default) %\\
+%\parlink{includeRadialExBDrive} = \true \;\;\; (Not default) \\
+%\parlink{nonlinear} = \true \;\;\; (Not default)

### 2018-02-15 doc/manual/version3/outputParameters.tex
Working on manual

@@
 
-\outputparam{delta\_f}
+\outputparamWithUnderscore{delta\_f}
+{delta_f}
 {$M_x \times M_{\xi} \times M_{\zeta} \times M_{\theta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
 {\parlink{export\_delta\_f} == \true.}
-{?}
+{First order distribution function $f_{s1} = f_{s} - f_{s0}$ for each species, normalized by $\bar{n} / \bar{v}^3$. 
+The phase-space grid $\left(M_x, M_{\xi}, M_{\zeta}, M_{\theta}\right)$} is specified by the input parameters in the {\ttfamily \hyperref[sec:exportfParameters]{export\_f}} namelist.
+
+\myhrule
+
+\outputparamWithUnderscore{full\_f}
+{full_f}
+{$M_x \times M_{\xi} \times M_{\zeta} \times M_{\theta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
+{\parlink{export\_full\_f} == \true.}
+{Full distribution function $f_{s}$ for each species, normalized by $\bar{n} / \bar{v}^3$. 
+The phase-space grid $\left(M_x, M_{\xi}, M_{\zeta}, M_{\theta}\right)$} is specified by the input parameters in the {\ttfamily \hyperref[sec:exportfParameters]{export\_f}} namelist.
 

### 2018-03-05 doc/manual/version3/inputParameters.tex
Working on manual
v3 manual update

@@
 {\false}
-{\todo{Whenever ?}}
-{\todo{Write description}}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and \parlink{quasineutralityOption} == 1}
+{When this is \true, an extra species (here called NBIspec)
+  is added to the quasineutrality condition, but not as a species for
+  which the DKE is solved. This is to mimic that fast ions influece
+  the other species mostly via the parallel electric field, an not
+  much via collisions. The NBIspec is assumed to have a density
+  distribution on the flux suface proporional to $B$ (which is the
+  case if all particles have zero pitch angle).}
 
@@
 {1.0}
-{\todo{Whenever \parlink{RHSMode} == 1 and \parlink{withNBIspec} == \true}}
-{\todo{Charge of NBI heated species, in units of the proton charge $e$.}}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and \parlink{quasineutralityOption} == 1}
+{Charge of the fast ion species, in units of the proton charge $e$.}
 
@@
 {0.0}
-{\todo{Whenever \parlink{RHSMode} == 1 and \parlink{withNBIspec} == \true}}
-{\todo{Density of NBI heated species, in units of the reference density $\bar{n}$.}}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and \parlink{quasineutralityOption} == 1}
+{Density of the fast ion species, in units of the reference density $\bar{n}$.}
 
@@
 \section{The {\ttfamily export\_f}~namelist}
+\label{sec:exportfParameters}
 

### 2018-04-17 doc/PoloidalVariationInCollisionOperator_code.tex

@@
 \noindent
-Apart from the different pre-factor, the term in Eq.~\eqref{eq:Jac1} differs from term in $\delta \hat C_{ab}^{L:f0}/\delta \hat f_{1s}$ by acting on the first order distribution function $f_{1s}$.
+Apart from the different pre-factor, the term in Eq.~\eqref{eq:Jac1} differs from the term in $\delta \hat C_{ab}^{L:f0}/\delta \hat f_{1s}$ by acting on the first order distribution function $f_{1s}$.
 For example, in the first case, terms such as $\partial f_{1s}/\partial x_s$ become
@@
 \subsection*{Jacobian}
- If the modifications explained in the previous section are implemented, the $\delta C/\delta f_{1s}$ contribution to the Jacobian is at this point already taking the density variations into account.
+ If the modifications explained in the previous section are implemented, the $\delta C/\delta f_{1s}$ contribution to the Jacobian is at this point already taking the density variation into account.
 We need to add the calculation of $\delta C/\delta \Phi_1$.
@@
 Presently, the modifications discussed in the previous section have been implemented into the ``poloidalVariationInCollisionOperator'' branch of \texttt{SFINCS}.
-What remains is to benchmark the output with theoretical predictions, using for example the model described in Ref.~\cite{ref:Fulop}\footnote{Note that in this model the collision operator is not linearised around a poloidally varying zeroth order distribution function (as is done here), but the poloidal variation enters through the perturbed distribution function in the ion-impurity collision operator, when computing the parallel friction force.}.
-Note that in this case only tokamak cases are considered.
+What remains is to verify that the new implementation is correct.
+First, we try to make simple tests and compare the output from the new code (with poloidal density variation included in the collision operator), with the original code (without poloidal density variation in the collision operator).
+This is discussed in the first subsection below.
+Next, we attempt to benchmark the output from the code with theoretical predictions.
+This is done in the second subsection.
+
+Tokamak geometry has been used in all cases considered here.
+
+\subsection*{Tests of new code}
+First, for simple scenarios (such as one species cases, or two species but with low impurity charge, $Z \le 4$), comparison between the new and old version of the code gives exactly the same results.
+The time it takes to run a simulation is practically the same.
+
+Next, we try to increase the effect of poloidal density variation in the collision operator, by increasing the impurity charge.
+For $Z \le 16$ this leads to a much longer computation time, but eventually the solver converges and we get a solution.
+When attempting to run with $Z = 17$, the nonlinear solver diverges.
+It is possible to get a solution by using \texttt{reusePreconditioner = .false.} in the input to \texttt{SFINCS} and by changing the relative and absolute tolerances used in the iterations.
+This can be done by providing the flags \texttt{-ksp\_atol} and \texttt{-snes\_rtol} when running the \texttt{SFINCS} simulation.
+Example of values that can be used are $1e-8$ for \texttt{-ksp\_atol} and $1.0e-6$ for \texttt{-snes\_rtol}.
+With these options, the nonlinear solver should eventually converge.
+Note that the other input parameters, such as geometry, species and resolution parameters, all have a strong impact on whether the solver will converge or not.
+Typical input parameters used here will be presented in the next subsection, further below.
+
+Even for these more complicated cases, with high impurity charge, there is still practically no difference between using the new code and the original version.
+This is demonstrated in Fig.~\ref{fig:0} where the perturbed impurity density is shown for $Z=1$ and $Z=17$ respectively.
+Both cases have been computed using the new and original version of the code.
+As is seen in the figure, the output from the old and new version of the code is almost identical.
+
+\begin{figure}
+ \setlength\figureheight{0.4\textheight}
+ \setlength\figurewidth{0.9\textwidth}
+    \centering
+    \input{figures/outdn.tex}
+    \caption{Normalised perturbed impurity density for impurity charge $Z = 1$ (orange) and $Z = 17$ (violet) respectively.
+The result is shown for the case when the original code is used (solid lines) and  when the new code is used instead, with poloidal density variation included in the collision operator(dashed line with circles).}
+            \label{fig:0}
+      \end{figure}
+
+Other cases, when changing for example impurity density, have also been studied.
+All have the same conclusion that poloidal density variation does not seem to matter in the collision operator, at least when Tokamak geometry is considered.
+Further work with Stellarators might be more interesting and should be done.
+
+%Note that depending on the other input parameters used, the value of the charge   
+%First we test the new version of the code for simple scenarios, such as plasmas with only one species, or with two species but with low impurity charge ($Z \le 6$).
+For such cases, the new version of the code successfully generates an output which is almost identical to the output from the original code.
+The time it takes to run a simulation is practically the same.
+
+%Next, in order to increase the effect of poloidal density variation in the collision operator, we try to increase the impurity charge.
+Now the code runs much slower (compared to when running the same cases with the original code) and it is much more difficult for the (nonlinear) solver to find a solution, which divereges.
+It is possible to make the nonlinear solver converge by using specific input.
+First, set the parameters \texttt{reusePreconditione = .false.} in the input file to \texttt{SFINCS}.
+Secondly, you should modify the relative and absolute tolerance used in the solver-iterations.
+This can be done by running \texttt{SFINCS} with the flags \texttt{-ksp\_atol} and \texttt{-snes\_rtol}.
+Good values are $1e-8$ and $1.0e-6$.
+By playing around with these input parameters it is eventually possible to make the code converge, although at a significant increased computation time.
+Still, even at higher charge number is is practically no difference between using the new version of the code and the old.
+This is demonstrated in Fig.~\ref{fig:0} where we show the perturbed impurity density as a function of poloidal angle, for two cases.
+One has impurity charge $z = 1$ while the other has $z = 17$.
+Both cases have been run using the original code and the new implementation, where poloidal density variation has been included.
+
+%Other simulations, for example when increasing the impurity density while maintaining high impurity charge also take time to run, and in the end does not show any particular difference when compared with running the simulation with the original code.
+This shows that poloidal density varation seems to play a small role in simple Tokamak scenarious.
+Perhaps, Stellarator cases may be required to observe a bigger difference.
+
+\subsection*{Comparing with model}
+The second approach we take to validate the code, is to benchmark the \texttt{SFINCS} output with theoretical predictions, using for example the model described in Ref.~\cite{ref:Fulop}\footnote{Note that in this model the collision operator is not linearised around a poloidally varying zeroth order distribution function (as is done here), but the poloidal variation enters through the perturbed distribution function in the ion-impurity collision operator, when computing the parallel friction force.}.
+
+%Note that in this case only Tokamak cases are considered.
 
@@
 Example of typical \texttt{SFINCS}-input used to get the best agreement so far is:
+\newpage
  \begin{lstlisting}
@@
 The two outputs show good agreement for the cosine parts while the sine is slightly deviating.
-Same scan using the new code (where poloidal density variation is included) gives exactly the same result.
-Some convergence issues appear when using the new implementation with high impurity charge (in this regime for $Z_s > 20$).
-No poloidal asymmetries have been observed so far, when using input to \texttt{SFINCS} which satisfies the orderings described in  Ref.~\cite{ref:Fulop} (which is required if we want to compare the \texttt{SFINCS} output with the model).
 
@@
        \caption{Cosine part of the perturbed density for a scan in the magnetic field strength $B_0$.
-Solid line shows result calculated from \texttt{SFINCS} output while the dashed line is the prediction from the model}
+Solid line shows result calculated from \texttt{SFINCS} output while the dashed line is the prediction from the model.}
                   \label{fig:2}
   \end{figure}
+
+Scans in $\epsilon$ shows a similar result, but a scan in impurity charge and density show more disagreement between model predictions and output from \texttt{SFINCS}.
+Similar tests using the new version of the code (with nonzero temperature gradients) show the same result.
+
+It appears to be challenging to get good agreement between the model and \texttt{SFINCS}.
+A possible explanation is the difficulty to get asymptotic in the various quantities required for the model to be valid, while maintaining a sufficiently large density perturbation to make a meaningful comparison.
+As mentioned, here we use $\Delta\sim 0.3, \delta \sim 0.15, \epsilon \sim 0.1, n_zZ^2/n_i \sim 1.2$ and $\nu^*\sim 0.11$ which might not be good enough for the theoretical model to be valid.
+
+Comparison between \texttt{SFINCS} and the model has been done using a series of python scripts available on \href{https://github.com/landreman/sfincs/commit/520091f6b9ed06cf0b7b1caaa7e0fcfc477f906f}{GitHub}.
+Further work is required and welcome.
+
+%Scan in other parameters, such as impurity density or $\epsilon$ show even worse disagreement.
+
+%It seems to be very difficult to get a proper agreement between this model and \texttt{SFINCS} which might be because it is difficult to get asymptotic in the various quantities, such as for example $\epsilon \ll 1$, while maintaining sufficiently large density perturbations.
 

### 2018-04-17 doc/manual/version3/inputParameters.tex

--- 2018-02-09 doc/manual/version3/inputParameters.tex
@@
 {``''}
-{\todo{Only when ?}}
-{\todo{Write description}}
+{Only when \parlink{EParallelHatSpec} is non-zero
+  and \parlink{RHSmode}==1 \\and \parlink{EParallelHatSpec\_bcdatFile}
+  is not == ``''}
+{When set this loads a file which looks like a bc file, but with the
+  part which usually contains Fourier coefficients for
+  $R$,$Z$,$B$,\ldots instead contains the coefficients for a function
+  which will multiply the ``effective parallel electric fields'' \parlink{EParallelHatSpec}
+  (otherwise just proportional to $B$).}
 
@@
 
-\param{withNBIspec}
-{Boolean}
-{\false}
-{\todo{Whenever ?}}
-{\todo{Write description}}
+\param{withNBIspec} {Boolean} {\false} {Whenever \parlink{RHSMode} ==
+  1 and \parlink{includePhi1} == \true\\
+  ~and \parlink{quasineutralityOption} == 1} {When this is \true, an
+  extra species (here called NBIspec) is added to the quasineutrality
+  condition, but not as a species for which the DKE is solved. This is
+  to mimic how the fast ions influence the other species via the
+  $\phi_1$ variation on the flux surface. (The influence via
+  collisions can be modelled using \parlink{EParallelHatSpec}). The
+  NBIspec is assumed to have a density distribution on the flux suface
+  proportional to $B$ (which is the case if all particles have zero
+  pitch angle).}
 
@@
 {1.0}
-{\todo{Whenever \parlink{RHSMode} == 1 and \parlink{withNBIspec} == \true}}
-{\todo{Charge of NBI heated species, in units of the proton charge $e$.}}
+{Whenever \parlink{withNBIspec}== \true~ and \parlink{RHSMode} == 1 \\
+  and \parlink{includePhi1} == \true~ and \parlink{quasineutralityOption} == 1}
+{Charge of the fast ion species, in units of the proton charge $e$, see \parlink{withNBIspec}.}
 
@@
 {0.0}
-{\todo{Whenever \parlink{RHSMode} == 1 and \parlink{withNBIspec} == \true}}
-{\todo{Density of NBI heated species, in units of the reference density $\bar{n}$.}}
+{Whenever \parlink{withNBIspec}== \true~ and \parlink{RHSMode} == 1 \\
+and \parlink{includePhi1} == \true~ and \parlink{quasineutralityOption} == 1}
+{Density of the fast ion species, in units of the reference density
+  $\bar{n}$, see \parlink{withNBIspec}. \todo{Implement possibility to load a bcdat file with the
+  fast ion density distribution on the flux surfaces}}
 
@@
 {0.0}
-{\todo{Whenever \parlink{RHSMode} == 1} ?}
-{\todo{Write description}}
+{Whenever it is set to non-zero and \parlink{RHSmode}==1}
+{This is used to introduce parallel forces due to collisions with
+  a fast ion species (which are not included in the DKE solve). The
+  collision operator for collisions with a fast species has the same
+  structure as parallel electric field term in the DKE, and the input parameter
+  here is thus given in terms of a corresponding ``effective'' electric field. Since the
+forces from these collisions are different on the different species,
+this input must be given as an array of values corresponding to the different
+species. These effective electric fields are added to the physical one
+\parlink{EParallelHat} in the code. The electric field strength is
+proportional to $B$ unless explicitly specified using an input file \parlink{EParallelHatSpec\_bcdatFile}.}
 

### 2018-04-17 doc/manual/version3/outputParameters.tex

@@
 {\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
-{{\ttfamily totalDensity} = \parlink{nHats}~+~\parlink{densityPerturbation}.}
+{{\ttfamily totalDensity} = \parlink{nHats}$\cdot \exp \left(- \frac{Z e \Phi_1}{T}  \right)$~+~\parlink{densityPerturbation}.}
 
@@
 {\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
-{{\ttfamily totalPressure} = \parlink{nHats} $\cdot$ \parlink{THats}~+~\parlink{pressurePerturbation}.}
+{{\ttfamily totalPressure} = \parlink{nHats}$\cdot \exp \left(- \frac{Z e \Phi_1}{T}  \right)$ $\cdot$ \parlink{THats}~+~\parlink{pressurePerturbation}.}
 

### 2018-04-17 doc/manual/version3/inputParameters.tex
added sfincsScan22

@@
 {\ttfamily
-!ss runSpecFile = thefilename.dat\\
-}
+!ss runSpecFile = thefilename.dat\\}
+
+{\ttfamily scanType} = 22: Similar to {\ttfamily scanType}=5 but the
+radial scan is specified with a runspec file like for {\ttfamily
+  scanType}=21. For each line in the {\ttfamily runspec.dat} file a
+directory is created, in which a {\ttfamily scanType}=2 $E_r$ scan is started.\\
 }

### 2018-07-13 doc/manual/version3/inputParameters.tex
Edit to rippleScale. Added entry to manual.

@@
 {1.0}
-{\todo{Only when ?}}
-{\todo{Write description}}
+{Only when \parlink{geometryScheme}==5. \parlink{helicity\_l} and \parlink{helicity\_n} must be specified.}
+{Scales the values of components of VMEC geometry (modes of $B$, $B^{u}$, $B_{u}$, $B^{v}$, $B_{v}$, $B_{s}$, and $\sqrt{g}$) that do not satisfy a desired symmetry.
+If \parlink{helicity\_n}$==0$, modes for which $n \neq 0$ or $m \neq$ \parlink{helicity\_l} are multiplied by \parlink{rippleScale}.
+If \parlink{helicity\_n}$\neq0$, modes for which $m/n \neq$ \parlink{helicity\_l}/\parlink{helicity\_n} or $n==0$ are multiplied by \parlink{rippleScale}.
+All other modes remain unchanged.}
 

### 2018-12-14 doc/manual/version3/inputParameters.tex
Added adjointOptions, geometryscheme=13, and ambipolarSolve things to manual.

@@
 
-{\ttfamily RHSMode} = 3: Solve for the 2$\times$2 monoenergetic transport coefficients. When this option is chosen, \Nx~is set to 1 and only 1 species is used.
+{\ttfamily RHSMode} = 3: Solve for the 2$\times$2 monoenergetic transport coefficients. When this option is chosen, \Nx~is set to 1 and only 1 species is used. \\
+
+{\ttfamily RHSMode} = 4: Solve the adjoint equation to compute derivatives at constant $E_r$.
+Options related to the adjoint solve are detailed in section \ref{sec:adjointOptions}.
+{\ttfamily RHSMode} = 4 must be used with \parlink{includePhi1}=\false, \\ \parlink{EParallelHat}\texttt{=0}, \parlink{magneticDriftScheme}\texttt{>0}, \parlink{constraintScheme}\texttt{=-1} or 1, and \\ \parlink{collisionOperator}\texttt{=0}.
+\\
+
+{\ttfamily RHSMode} = 5: Solve the adjoint equation to compute derivatives at the ambipolar $E_r$.
+Options related to the adjoint solve are detailed in section \ref{sec:adjointOptions}.
+{\ttfamily RHSMode} = 5 must be used with \parlink{includePhi1}=\false, \parlink{EParallelHat}\texttt{=0}, \parlink{magneticDriftScheme}\texttt{>0}, \parlink{constraintScheme}\texttt{=-1} or 1, and \\ \parlink{collisionOperator}\texttt{=0}.\\
+
 }
@@
 
+\myhrule
+
+\param{ambipolarSolve}
+{Boolean}
+{\false}
+{When \parlink{RHSMode}==1, 4, or 5.}
+{When \true, a root finding method will be used to find the ambipolar \parlink{Er}.
+All output parameters will be included at the ambipolar \parlink{Er}.
+Options related to this parameter are \parlink{NEr\_ambipolarSolve}, \parlink{Er\_search\_tolerance\_dx}, \parlink{Er\_search\_tolerance\_f}, \parlink{ambipolarSolveOption}, \parlink{Er\_min}, and \parlink{Er\_max}.}
+
+\myhrule
+
+\param{NEr\_ambipolarSolve}
+{integer}
+{20}
+{When \parlink{ambipolarSolve} == \true.}
+{Maximum number of solves to allow while finding the ambipolar \parlink{Er}.}
+
+\myhrule
+
+\param{Er\_search\_tolerance\_dx}
+{real}
+{1.d-8}
+{When \parlink{ambipolarSolve} == \true and \parlink{ambipolarSolveOption}/=2.}
+{Tolerance used for ambipolar solve.
+If step size in \parlink{Er} falls below \parlink{Er\_search\_tolerance\_dx}, the ambipolar solve completes successfully.
+Note that this is not used for \parlink{ambipolarSolveOption}==2.}
+
+\myhrule
+
+\param{Er\_search\_tolerance\_f}
+{real}
+{1.d-10}
+{When \parlink{ambipolarSolve} == \true.}
+{Tolerance used for ambipolar solve. If the radial current,
+\begin{gather}
+J_r = \sum_s \texttt{Zs}(s) \texttt{particleFlux\_vm\_rN}(s),
+\end{gather}
+falls below \parlink{Er\_search\_tolerance\_f}, the ambipolar solve completes successfully.}
+
+\myhrule
+
+\param{ambipolarSolveOption}
+{integer}
+{1}
+{When \parlink{ambipolarSolve} == \true.}
+{Indicates which root solving algorithm to use for ambipolar solve.
+If \parlink{ambipolarSolveOption}==1 a Newton bisection method is used, \parlink{ambipolarSolveOption}==2 a Brent method is used, and if \parlink{ambipolarSolveOption}==3 a Newton method is used.
+If \parlink{ambipolarSolveOption}==1 or 3, an adjoint method will used to compute the $\partial J_r/\partial E_r$, so must be used with \parlink{includePhi1}==\false, \parlink{EParallelHat}==0, \parlink{magneticDriftScheme}>0, \parlink{constraintScheme}==-1 or 1, and \parlink{collisionOperator}==0.}
+
+\myhrule
+
+\param{Er\_min}
+{real}
+{-100}
+{When \parlink{ambipolarSolve} == \true and \parlink{ambipolarSolveOption} /= 3. }
+{Minimum value of \parlink{Er} used to bracket the ambipolar root. }
+
+\myhrule
+
+\param{Er\_max}
+{real}
+{100}
+{When \parlink{ambipolarSolve} == \true and \parlink{ambipolarSolveOption} /= 3. }
+{Maximum  value of \parlink{Er} used to bracket the ambipolar root. }
+
+\myhrule
+
 \section{The {\ttfamily geometryParameters} namelist}
@@
 The file is assumed to be stellarator-\emph{asymmetric}.
+
+{\ttfamily geometryScheme}==13: Read Boozer geometry from the {\ttfamily geometryParamters} namelist.
+Here the geometry parameters are specified by \parlink{boozer\_bmnc}, \parlink{GHat}, \parlink{IHat}, and \parlink{iota}.
+This has been implemented for use with STELLOPT for direct optimization of the Boozer spectrum.
 }
@@
 {3.7481}
-{Only when \parlink{geometryScheme} == 1. Otherwise, {\ttfamily GHat} will be set according to the requested \parlink{geometryScheme}.}
+{Only when \parlink{geometryScheme} == 1 or 13. Otherwise, {\ttfamily GHat} will be set according to the requested \parlink{geometryScheme}.}
 {$G$ is $(c/2)\times$ the poloidal current outside the flux
@@
 {0.0}
-{Only when \parlink{geometryScheme} == 1. Otherwise, {\ttfamily IHat} will be set according to the requested \parlink{geometryScheme}.}
+{Only when \parlink{geometryScheme} == 1 or 13. Otherwise, {\ttfamily IHat} will be set according to the requested \parlink{geometryScheme}.}
 {$I$ is $(c/2)\times$ the toroidal current inside the flux
@@
 {0.4542}
-{Only when \parlink{geometryScheme} == 1. Otherwise, {\ttfamily iota} will be set according to the requested \parlink{geometryScheme}.}
+{Only when \parlink{geometryScheme} == 1 or 13. Otherwise, {\ttfamily iota} will be set according to the requested \parlink{geometryScheme}.}
 {Rotational transform (rationalized), equivalent to $1/q$ where $q$ is the safety factor.}
@@
   (otherwise just proportional to $B$).}
+
+\myhrule
+
+\param{boozer\_bmnc}
+{2D array of reals}
+{0}
+{Only when \parlink{geometryScheme} == 13. }
+{Fourier harmonics describing the magnetic field in Boozer coordinates,
+\begin{gather}
+B(\theta_B,\zeta_B) = \sum_{m,n}\mathrm{\texttt{boozer\_bmnc}}(m,n) \cos(m\theta_B-n\mathrm{\texttt{Nperiods}} \zeta_B),
+\end{gather}
+The first index of \texttt{ Boozer\_bmnc} corresponds to $m \in $[0,\parlink{mmax\_boozer}], and the second corresponds to $n \in $[-\parlink{nmax\_boozer},\parlink{nmax\_boozer}].
+Here \parlink{Nperiods} can be specified in the input file.
+}
+
+\myhrule
+
+\param{mmax\_boozer}
+{integer}
+{40}
+{Only when \parlink{geometryScheme} == 13.}
+{Boozer geometry is described for $m \leq $ \texttt{mmax\_boozer}. }
+
+\myhrule
+
+\param{nmax\_boozer}
+{integer}
+{40}
+{Only when \parlink{geometryScheme} == 13.}
+{Boozer geometry is described for $|n| \leq $ \texttt{nmax\_boozer}. }
+
+\myhrule
+
+\param{Nperiods}
+{integer}
+{0}
+{Only when \parlink{geometryScheme} == 13.}
+{Boozer geometry will assume number of periods is \texttt{Nperiods}.}
 
@@
 
+\section{The \texttt{adjointOptions} namelist}
+\label{sec:adjointOptions}
+
+\param{adjointBootstrapOption}
+{Boolean}
+{\false}
+{When \parlink{RHSMode} == 4 or 5}
+{If this parameter is \true, the derivative of the bootstrap current (\parlink{FSABjHatOverRootFSAB2}) is computed with respect to geometry parameters.
+If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
+Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
+
+\myhrule
+
+\param{adjointRadialCurrentOption}
+{Boolean}
+{\false}
+{When \parlink{RHSMode} == 4 or 5}
+{If this parameter is \true the derivative of the radial current ($J_r$) is computed with respect to geometry parameters,
+\begin{gather}
+J_r = \sum_s \texttt{Zs}(s) \texttt{particleFlux\_vm\_rN}(s),
+\end{gather}
+where \parlink{Zs} and \parlink{particleFlux\_vm\_rN} can be found elsewhere in this document.
+If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
+Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
+
+\myhrule
+
+\param{adjointTotalHeatFluxOption}
+{Boolean}
+{\false}
+{When \parlink{RHSMode} == 4 or 5}
+{If this parameter is \true, the derivative of the total heat flux ($Q$) is computed with respect to geometry parameters,
+\begin{gather}
+Q = \sum_s \texttt{Zs}(s) \texttt{heatFlux\_vm\_rN}(s),
+\end{gather}
+where \parlink{Zs} and \parlink{heatFlux\_vm\_rN} can be found elsewhere in this document.
+If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
+Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
+
+\myhrule
+
+\param{adjointHeatFluxOption}
+{1D array of booleans}
+{\false}
+{When \parlink{RHSMode} == 4 or 5}
+{This array is of the same length of {\ttfamily nHats} and other species parameters.
+If parameter is \true, the derivative of the heat flux (\parlink{heatFlux\_vm\_rN}) will be computed for this species with respect to geometry parameters.
+If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
+Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
+
+\myhrule
+
+\param{adjointParticleFluxOption}
+{1D array of booleans}
+{\false}
+{When \parlink{RHSMode} == 4 or 5}
+{This array is of the same length of {\ttfamily nHats} and other species parameters.
+If parameter is {\ttfamily .true.}, the derivative of the particle flux (\parlink{particleFlux\_vm\_rN}) will be computed for this species with respect to geometry parameters.
+If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
+Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
+
+\myhrule
+
+\param{adjointParallelFlowOption}
+{1D array of booleans}
+{\false}
+{When \parlink{RHSMode} == 4 or 5}
+{This array is of the same length of {\ttfamily nHats} and other species parameters.
+If parameter is {\ttfamily .true.}, the derivative of the parallel flow (\parlink{FSABVelocityUsingFSADensityOverRootFSAB2}) will be computed for this species with respect to geometry parameters.
+If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
+Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
+
+\myhrule
+
+\param{nMaxAdjoint}
+{integer}
+{0}
+{When \parlink{RHSMode} == 4 or 5}
+{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \parlink{adjointParallelFlowOption} will computed for Fourier modes $|n| < $ \texttt{nMaxAdjoint}.
+}
+
+\myhrule
+
+\param{mMaxAdjoint}
+{integer}
+{0}
+{When \parlink{RHSMode} == 4 or 5}
+{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \parlink{adjointParallelFlowOption} will computed for Fourier modes $m < $ \texttt{mMaxAdjoint}.
+}
+
+\myhrule
+
+\param{nMinAdjoint}
+{integer}
+{0}
+{When \parlink{RHSMode} == 4 or 5}
+{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \parlink{adjointParallelFlowOption} will computed for Fourier modes $|n| > $ \texttt{nMaxAdjoint}.
+}
+
+\myhrule
+
+\param{mMinAdjoint}
+{integer}
+{0}
+{When \parlink{RHSMode} == 4 or 5}
+{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \parlink{adjointParallelFlowOption} will computed for Fourier modes $m > $ \texttt{mMaxAdjoint}.
+}
+
+\myhrule
+
+\param{debugAdjoint}
+{boolean}
+{\false}
+{When \parlink{RHSMode} == 4 or 5. }
+{If \texttt{.true.}, then derivatives computed with the adjoint method will be compared with forward difference derivatives with step size set by \parlink{deltaLambda}.}
+
+\myhrule
+
+\param{deltaLambda}
+{real}
+{1.d-4}
+{When \parlink{deltaLamda} == \texttt{.true.}}
+{The step size used for forward difference derivatives will be \texttt{deltaLambda} multiplied by the quantity that is being varied.
+For example, the step size for a Boozer harmonic, $B_{mn}$, would be $\Delta B_{mn} = $\texttt{deltaLambda}$B_{mn}$.}
+
+\myhrule
+
 \section{Directives for \sfincsScan}

### 2019-01-17 doc/classical_radial_fluxes_2019-01-17.tex
Adds a radial heat flux calculation to the classicalTransport module. Unifies the classical heat flux and particle flux by writing both in terms of Braginskii matrices. Adds documentation of the classical flux calculation.

\documentclass[12pt, a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath}
%\usepackage{relsize}

%\usepackage{fixmath}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[usenames,dvipsnames]{color}
%\usepackage[small,font=it]{caption}
\usepackage{amssymb}
%\usepackage{icomma}
%\usepackage{hyperref}
%\usepackage{mcode}
\usepackage{verbatim}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{units}
\usepackage{fancyhdr}
\pagestyle{fancy}
\lhead{\footnotesize \parbox{11cm}{Braginskii matrix approach to classical transport from the linearized Fokker-Planck collision operator}}
\rhead{\footnotesize \parbox{2cm}{Stefan Buller}}
\renewcommand\headheight{24pt}
%\usepackage{dirtytalk} \say{} command for quotations

\makeatletter
\newenvironment{tablehere}
  {\def\@captype{table}}
  {}

\newenvironment{figurehere}
  {\def\@captype{figure}}
  {}

\newsavebox\myboxA
\newsavebox\myboxB
\newlength\mylenA

\newcommand*\obar[2][0.75]{% OverBAR, adds bar over an element
    \sbox{\myboxA}{$\m@th#2$}%
    \setbox\myboxB\null% Phantom box
    \ht\myboxB=\ht\myboxA%
    \dp\myboxB=\dp\myboxA%
    \wd\myboxB=#1\wd\myboxA% Scale phantom
    \sbox\myboxB{$\m@th\overline{\copy\myboxB}$}%  Overlined phantom
    \setlength\mylenA{\the\wd\myboxA}%   calc width diff
    \addtolength\mylenA{-\the\wd\myboxB}%
    \ifdim\wd\myboxB<\wd\myboxA%
       \rlap{\hskip 0.5\mylenA\usebox\myboxB}{\usebox\myboxA}%
    \else
        \hskip -0.5\mylenA\rlap{\usebox\myboxA}{\hskip 0.5\mylenA\usebox\myboxB}%
    \fi}

\makeatother

%\def\equationautorefname{ekvation}
%\def\tableautorefname{tabell}
%\def\figureautorefname{figur}
%\def\sectionautorefname{sektion}
%\def\subsectionautorefname{sektion}

\DeclareMathOperator\erf{erf}
%\DeclareMathOperator\mod{mod}

\newcommand{\ordo}[1]{{\cal O}\left( #1 \right)}
\DeclareMathOperator{\sgn}{sgn}
%\renewcommand{\vec}[1]{\boldsymbol{#1}}
\newcommand{\im}{\ensuremath{\mathrm{i}}}
\newcommand{\e}{\ensuremath{\mathrm{e}}}
\newcommand{\E}{\ensuremath{\mathcal{E}}}
\newcommand{\p}{\ensuremath{\partial}}
\newcommand{\Z}{\ensuremath{\mathbb{Z}}}

\newcommand{\bra}[1]{\langle #1 \mid}
\newcommand{\ket}[1]{\mid #1 \rangle}
\newcommand\matris[4]{\ensuremath{\begin{pmatrix} #1 & #2 \\ #3 & #4\end{pmatrix}}}
\renewcommand{\d}{\ensuremath{\mathrm{d}}}
%\DeclareMathOperator*{\sgn}{sgn}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\lang}{\left\langle}
\newcommand{\rang}{\right\rangle}

%to get leftrightarrow over tensors of rank-2.
\def\shrinkage{2.1mu}
\def\vecsign{\mathchar"017E}
\def\dvecsign{\smash{\stackon[-1.95pt]{\mkern-\shrinkage\vecsign}{\rotatebox{180}{$\mkern-\shrinkage\vecsign$}}}}
\def\dvec#1{\def\useanchorwidth{T}\stackon[-4.2pt]{#1}{\,\dvecsign}}
\usepackage{stackengine}
\stackMath
\def\perfect{\textsc{Perfect}}

\lstset{language=[90]Fortran,
  basicstyle=\ttfamily,
  keywordstyle=\color{red},
  commentstyle=\color{green},
  morecomment=[l]{!\ }% Comment only with space after !
   frame=single,
   breaklines=true,
   postbreak=\raisebox{0ex}[0ex][0ex]{\ensuremath{\color{blue}\hookrightarrow\space}}
}

\usepackage[
    backend=biber,
    style=alphabetic,
    citestyle=authoryear,
    natbib=true,
    url=false,
    doi=false,
    isbn=false,
    eprint=false,
    sorting=none,
    maxcitenames=2,
    maxbibnames=3
]{biblatex}

\AtEveryBibitem{\clearfield{month}}
\AtEveryCitekey{\clearfield{month}}
\AtEveryBibitem{\clearfield{issue}}
\AtEveryCitekey{\clearfield{issue}}

\bibliography{../../../plasma-bib}

%=================================================================
\begin{document}
\section{Drift-kinetic classical transport}
These notes derive the classical transport under the same set of assumptions as in the linearized drift-kinetic equation solved by the $\delta f$-code neoclassical code \textsc{Sfincs} (and probably most other neoclassical codes).

\subsection{Motivation}
For a mass-ratio expanded ion-impurity collision operator, we found that the classical transport of collisional impurities in W7-X can be comparable to the neoclassical transport. Thus, it would be of interest to have a general numerical tool to calculate the classical transport alongside the neoclassical transport from a drift-kinetic solver such as \textsc{Sfincs}.

\subsection{Gyrophase dependent part of $f$}
In many ways, calculating the classical transport is simpler than calculating the neoclassical transport, as the gyrophase dependent part of the distribution function is smaller than the gyrophase-independent in the expansion parameter $\rho/L$, and the gyrophase dependent part to required order is given entirely by the zeroth order gyrophase independent distribution. When $f_{a0}$ is a Maxwellian $f_{aM}$, the gyrophase dependent part is given by
\begin{equation}
\tilde{f}_{a1} = -\vec{\rho}_a(\gamma) \cdot \nabla f_{Ma} = -\vec{\rho}(\gamma) \cdot \nabla \psi  \frac{\p f_{Ma}}{\p \psi}, \label{eq:ftilde}
\end{equation}
where $\gamma$ is the gyrophase, $\psi$ a flux-label and $\vec{\rho}$ the gyrophase vector
\begin{equation}
\vec{\rho}_a = \frac{1}{\Omega_a}\vec{b} \times \vec{v}_\perp,
\end{equation}
with $\vec{v}_\perp$ the velocity perpendicular to the magnetic field, $\Omega_a=Z_a eB/m_a$ the gyrofrequency, $\vec{b}$ the unit vector in the direction of the magnetic field $\vec{B}$; and $\psi$ a flux-label, which acts as our radial coordinate.

\autoref{eq:ftilde} can be massaged a bit to yield (we drop the tilde since we will only consider the gyrophase dependent part of $f_{a1}$ here)
\begin{equation}
f_{a1} = \frac{1}{\Omega_a}(\vec{b}\times \nabla \psi) \cdot  \vec{v}_\perp\frac{\p f_{Ma}}{\p \psi},
\end{equation}
and we introduce the local coordinate system $\vec{v}_\perp = v_2 \hat{e}_2 + v_3 \hat{e}_3$, where $\hat{e}_2$, $\hat{e}_3$ form an orthonormal triplet with $\vec{b}$. There is some freedom in how we align our $\hat{e}_2$, $\hat{e}_3$ coordinates, and we choose these vectors so that $\hat{e}_2$ is entirely in the $(\vec{b}\times \nabla \psi)$ direction. Thus
\begin{equation}
f_{a1} = \frac{1}{\Omega_a}|\vec{b}\times \nabla \psi| v_2 \frac{\p f_{Ma}}{\p \psi}.
\end{equation}

Finally, we evaluate $\frac{\p f_{Ma}}{\p \psi}$. For heavy impurities, the lowest order distribution may vary on the flux-surface in response to a flux-surface variation of the electrostatic potential, in which case, we write
\begin{equation}
f_{Ma} = \eta_a(\psi) \left(\frac{m_a}{2\pi T_a}\right)^{3/2} \exp{\left(-\frac{m_av^2}{2T_a} - \frac{Z_a e \tilde{\Phi}}{T_a}\right)},
\end{equation}
where the \emph{pseudo-density} $\eta_a$ is related to the ordinary density by
\begin{equation}
\eta_a = n_a \e^{Z_a e \tilde{\Phi}/T_a}.
\end{equation}
The gradient thus becomes
\begin{equation}
\frac{\p f_{Ma}}{\p \psi} = f_{Ma}\left(\frac{\d \ln\eta_a}{\d \psi} + \frac{Z_a e}{T_a} \frac{\p \tilde{\Phi}}{\p \psi}  + \frac{Z_a e \tilde{\Phi}}{T_a} \frac{\d \ln T_a}{\d \psi}
 + \left[\frac{m_a v^2}{2T_a} - \frac{3}{2}\right]  \frac{\d \ln T_a}{\d \psi}\right). \label{eq:dpsif0}
\end{equation}
To simplify notation, we will introduce the quantities
\begin{align}
\alpha_{1a} \equiv & \frac{\d \ln\eta_a}{\d \psi} + \frac{Z_a e}{T_a} \frac{\p \tilde{\Phi}}{\p \psi}  + \frac{Z_a e \tilde{\Phi}}{T_a} \frac{\d \ln T_a}{\d \psi}
  - \frac{3}{2}  \frac{\d \ln T_a}{\d \psi} \\
  \alpha_{2a} \equiv & \frac{\d \ln T_a}{\d \psi},
\end{align}
so that $\frac{\p f_{Ma}}{\p \psi} = f_{Ma}(\alpha_{a1} + \alpha_{2a} x_a^2)$,
where
\begin{equation}
x_a^2 \equiv \frac{m_a v^2}{2T_a} \equiv \frac{v^2}{v_{Ta}^2}.
\end{equation}
Thus
\begin{equation}
f_{a1} = \frac{|\vec{b}\times \nabla \psi| }{\Omega_a}v_2 f_{Ma}(\alpha_{1a} + \alpha_{2a} x_a^2).
\end{equation}

\subsection{Radial fluxes}
The classical radial flux of particle and heat is given by
\begin{align}
  \Gamma_a \equiv \lang \vec{\Gamma}_a \cdot \nabla \psi \rang^{\text{C}} \equiv& \lang \frac{\vec{b} \times \nabla \psi}{Z_a e B} \cdot\vec{R}_a \rang,\label{eq:Gamma} \\
  Q_a \equiv \lang \vec{Q}_a \cdot \nabla \psi \rang^{\text{C}} \equiv& \lang \frac{\vec{b} \times \nabla \psi}{Z_a e B} \cdot\vec{G}_a \rang,\label{eq:Q}
\end{align}
where we have introduced the \emph{friction force} and \emph{energy-weighted friction force}
\begin{align}
  \vec{R}_a &= \int \d^3 v m_a \vec{v} C[f_{a}],\\
  \vec{G}_a &= \int \d^3 v \frac{m_a v^2}{2}  m_a \vec{v} C[f_{a}].
\end{align}
Here, $C[f_a]$ is the Fokker-Planck collision operator accounting for the effects of all species acting on $f_a$
\begin{equation}
    C[f_a] = \sum_b C_{ab}\left[f_{a},f_{b}\right].
    \end{equation}
    An important property of the Fokker-Planck operator is that it preserves the gyrophase dependence of $f_a$. Thus, only the gyrophase dependent part of $f_{a}$ will contribute to the friction force in the $\vec{b} \times \nabla\psi$ direction.
    As the gyrophase dependent part of $f_a$ acts as a small correction to $f_{Ma}$, we can linearize the Fokker-Planck operator around $f_{Ma}$ and write
\begin{equation}
  \begin{aligned}
    C[f_a] =& \sum_b C_{ab}\left[f_{Ma} + f_{1a},f_{Mb}+f_{1b}\right] \\
    \approx & \sum_b C_{ab}[f_{Ma},f_{Mb}] + C_{ab}[f_{Ma},f_{b1}] + C_{ab}[f_{a1},f_{Mb}],
      \end{aligned}
    \end{equation}
    where the gyrophase-independent $C_{ab}[f_{Ma},f_{Mb}]$ part will not contribute to the classical transport. With this result, we write the relevant components of $\vec{R}_a$ and $\vec{G}_a$ as
    \begin{align}
  \vec{b} \times \nabla \psi \cdot \vec{R}_a &= m_a  |\vec{b} \times \nabla \psi|\sum_b \int \!\d^3 v \, v_2 \left(C_{ab}[f_{a1},f_{Mb}] + C_{ab}[f_{Ma},f_{b1}]\right),\\
  \vec{b} \times \nabla \psi \cdot \vec{G}_a &= m_a T_a |\vec{b} \times \nabla \psi| \sum_b \int \!\d^3 v\, v_2 x_a^2  \left(C_{ab}[f_{a1},f_{Mb}] + C_{ab}[f_{Ma},f_{b1}] \right).
    \end{align}
    Inserting our expressions for $f_{1}$, we thus get
    \begin{align}
      &\begin{aligned}
        (\vec{b} \times \nabla \psi) \cdot \vec{R}_a = m_a  |\vec{b} \times \nabla \psi|^2\sum_b \int \!\d^3 v \, v_2 &\left( \frac{\alpha_{1a}}{\Omega_a}  C_{ab}[v_2 f_{Ma},f_{Mb}] + \frac{\alpha_{2a}}{\Omega_a} C_{ab}[v_2 x_a^2 f_{Ma} ,f_{Mb}] \right.\\
          &\left.+ \frac{\alpha_{1b}}{\Omega_b}C_{ab}[f_{Ma},v_2 f_{Mb}] + \frac{\alpha_{2b}}{\Omega_b}C_{ab}[f_{Ma},v_2 x_b^2 f_{Mb} ]\right),
      \end{aligned}\\
      &\begin{aligned}
        (\vec{b} \times \nabla \psi) \cdot \vec{G}_a = m_a T_a |\vec{b} \times \nabla \psi|^2 \sum_b \int \!\d^3 v\, v_2 x_a^2  &\left( \frac{\alpha_{1a} }{\Omega_a}C_{ab}[v_2 f_{Ma},f_{Mb}]
          + \frac{\alpha_{2a}}{\Omega_a}C_{ab}[v_2 x_a^2 f_{Ma} ,f_{Mb}]
        \right.\\
        &\left.+ \frac{\alpha_{1b}}{\Omega_b}C_{ab}[f_{Ma},v_2 f_{Ma}]
          + \frac{\alpha_{2b}}{\Omega_b}C_{ab}[f_{Ma},v_2 x_b^2f_{Ma}] 
        \right).
      \end{aligned}
    \end{align}
    These friction-force projections can be conveniently expressed in terms of Braginskii matrix elements
    \begin{align}
  M_{ab}^{jk} = \frac{\tau_{ab}}{n_a} \int v_2 L_j^{(3/2)}(x_a^2) C_{ab}\left[\frac{m_a v_2}{T_a} L_k^{(3/2)}(x_a^2) f_{a0},f_{b0}\right] \\
  N_{ab}^{jk} = \frac{\tau_{ab}}{n_a} \int v_2 L_j^{(3/2)}(x_a^2) C_{ab}\left[f_{a0},\frac{m_b v_2}{T_b} L_k^{(3/2)}(x_b^2)  f_{b0}\right],
\end{align}
where
\begin{equation}
\tau_{ab} = \frac{12 \pi^{3/2} \epsilon_0^2}{\sqrt{2} e^4 \ln \Lambda} \frac{m_a^{1/2} T_a^{3/2}}{Z_a^2 Z_b^2 n_b},
\end{equation}
and the relevant Sonine polynomials are
\begin{align}
  L_0^{(3/2)}(x_a^2) = 1 \\
  L_1^{(3/2)}(x_a^2) = \frac{5}{2} - x_a^2,
\end{align}
so that
\begin{align}
  x_a^2 - \frac{5}{2} =& -L_1^{(3/2)}(x_a^2)  \label{eq:qmotiv}\\
  x_a^2  =& \frac{5}{2}L_0^{(3/2)}(x_a^2) -L_1^{(3/2)}(x_a^2)  \\
  1 = &L_0^{(3/2)}(x_a^2).
\end{align}
\autoref{eq:qmotiv} makes it convenient to calculate the conductive heat flux $q_a = Q_a - \frac{5}{2} T_a \Gamma_a$ rather than directly calculating $Q_a$. With this in mind, we write
\begin{align}
      &\begin{aligned}
        (\vec{b} \times \nabla \psi) \cdot \vec{R}_a = m_a  |\vec{b} \times \nabla \psi|^2\sum_b \frac{n_a}{\tau_{ab}}&\left[ \frac{\alpha_{1a}}{\Omega_a}  \frac{T_a}{m_a}M_{ab}^{00} + \frac{\alpha_{2a}}{\Omega_a}\frac{T_a}{m_a} \left(\frac{5}{2}M_{ab}^{00} - M_{ab}^{01}\right)  \right.\\
          &\left.+ \frac{\alpha_{1b}}{\Omega_b}\frac{T_b}{m_b}N_{ab}^{00}+ \frac{\alpha_{2b}}{\Omega_b}\frac{T_b}{m_b}\left(\frac{5}{2} N_{ab}^{00} - N_{ab}^{01}\right)\right],
      \end{aligned}\\
      &\begin{aligned}
        (\vec{b} \times \nabla \psi) \cdot \left(\vec{G}_a - T_a \frac{5}{2}\vec{R}_a\right) = -m_a T_a |\vec{b} \times \nabla \psi|^2 \sum_b \frac{n_a}{\tau_{ab}} &\left[ \frac{\alpha_{1a} }{\Omega_a} \frac{T_a}{m_a}M_{ab}^{10}
          + \frac{\alpha_{2a}}{\Omega_a}\frac{T_a}{m_a}\left(\frac{5}{2}M_{ab}^{10} - M_{ab}^{11}\right) 
        \right.\\
        &\left.+ \frac{\alpha_{1b}}{\Omega_b}\frac{T_b}{m_b} N_{ab}^{10}
          + \frac{\alpha_{2b}}{\Omega_b}\frac{T_b}{m_b}\left(\frac{5}{2}N_{ab}^{10} - N_{ab}^{11}\right) 
        \right].
      \end{aligned}
\end{align}
The classical particle-flux and heat-flux thus become

\begin{align}
  \hspace*{-18ex}\Gamma_a =&  \frac{m_a}{Z_a  e^2}\lang |\nabla \psi|^2 \sum_b \frac{n_a n_b}{B^2 \tau_{ab} n_b}\left[ \frac{\alpha_{1a} T_a}{Z_a}  M_{ab}^{00} + \frac{\alpha_{2a} T_a}{Z_a} \left(\frac{5}{2}M_{ab}^{00} - M_{ab}^{01}\right) + \frac{\alpha_{1b} T_b}{Z_b}N_{ab}^{00}+ \frac{\alpha_{2b} T_b}{Z_b}\left(\frac{5}{2} N_{ab}^{00} - N_{ab}^{01}\right)\right] \rang \\
  \hspace*{-18ex}q_a =&  -\frac{T_a m_a}{Z_a e^2}\lang |\nabla \psi|^2 \sum_b \frac{n_a n_b}{B^2 \tau_{ab} n_b }\left[ \frac{\alpha_{1a} T_a}{Z_a}  M_{ab}^{10} + \frac{\alpha_{2a} T_a}{Z_a} \left(\frac{5}{2}M_{ab}^{10} - M_{ab}^{11}\right) + \frac{\alpha_{1b} T_b}{Z_b}N_{ab}^{10}+ \frac{\alpha_{2b} T_b}{Z_b}\left(\frac{5}{2} N_{ab}^{10} - N_{ab}^{11}\right)\right] \rang.
\end{align}

The values for the matrix-components can be found in the appendix of Helander+Sigmar. We can simplify the above expressions slightly by noting the symmetry properties
\begin{align}
  M_{ab}^{jk} &= M_{ab}^{kj}, \\
  N_{ab}^{jk} &= \frac{T_a v_{Ta}}{T_b v_{Tb}} N_{ba}^{kj}, \\
  N_{ab}^{j0} &=-M_{ab}^{j0} = -M_{ab}^{0j},  \\
  N_{ab}^{0j} &= \frac{T_a v_{Ta}}{T_b v_{Tb}} N_{ba}^{j0} =-\frac{T_a v_{Ta}}{T_b v_{Tb}} M_{ba}^{j0}  =-\frac{T_a^{3/2} m_b^{1/2}}{T_b^{3/2} m_a^{1/2}} M_{ba}^{0j}, \\
\end{align}
where the last property follows from applying the other 3. Thus, we only need
\begin{align}
  M_{ab}^{00} =& - \frac{1+\frac{m_a}{m_b}}{\left(1+\frac{m_a T_b}{m_b T_a}\right)^{3/2}} = - \frac{\left(1+\frac{m_a}{m_b}\right)\left(1+\frac{m_a T_b}{m_b T_a}\right)}{\left(1+\frac{m_a T_b}{m_b T_a}\right)^{5/2}}\\
  M_{ab}^{01} =& - \frac{3}{2} \frac{1+\frac{m_a}{m_b}}{\left(1+\frac{m_a T_b}{m_b T_a}\right)^{5/2}} \\
  M_{ba}^{01} =& - \frac{3}{2} \frac{1+\frac{m_b}{m_a}}{\left(1+\frac{m_b T_a}{m_a T_b}\right)^{5/2}} =- \frac{3}{2} \frac{\left(1+\frac{m_a}{m_b}\right)\frac{m_b}{m_a}\left(\frac{m_a T_b}{m_b T_a}\right)^{5/2}}{\left(1+\frac{m_a T_b}{m_b T_a}\right)^{5/2}} \\
  M_{ab}^{11} =& - \frac{\frac{13}{4} + 4 \frac{m_a T_b}{m_b T_a} + \frac{15}{2}\left(\frac{m_a T_b}{m_b T_a}\right)^2 }{\left(1+\frac{m_a T_b}{m_b T_a}\right)^{5/2}} \\
  N_{ab}^{11} =& \frac{27}{4}  \frac{\frac{m_a}{m_b}}{\left(1+\frac{m_a T_b}{m_b T_a}\right)^{5/2}}. \\
\end{align}

The property $N_{ab}^{j0} =-M_{ab}^{j0} = -M_{ab}^{0j}$ also means that the radial electric field does not contribute to the classical heat or particle flux.

Writing out the $\alpha_1$ and $\alpha_2$ and employing the relevant symmetry properties, we obtain the particle flux
\begin{equation}
\begin{aligned}
 \hspace*{-10ex} \Gamma_a =  \frac{m_a}{Z_a e^2}\sum_b \frac{1}{\tau_{ab} n_b}&\left[\lang  n_a n_b\frac{ |\nabla \psi|^2}{B^2} \rang M_{ab}^{00} \left(\frac{ T_a}{Z_a}  \frac{\d \ln\eta_a}{\d \psi} -\frac{ T_b}{Z_b}  \frac{\d \ln\eta_b}{\d \psi}\right) \right.\\
    &\left.+    \lang n_a n_b\frac{  |\nabla \psi|^2}{B^2} e\tilde{\Phi} \rang M_{ab}^{00} \left(\frac{\d \ln T_a}{\d \psi} -\frac{\d \ln T_b}{\d \psi}\right) \right.\\
   &\left. + \lang n_a n_b \frac{|\nabla \psi|^2 }{B^2}\rang\left(  \left(M_{ab}^{00} - M_{ab}^{01}\right)\frac{ T_a}{Z_a}\frac{\d \ln T_a}{\d \psi} 
    - \left(M_{ab}^{00} - \frac{m_a T_b}{m_b T_a} M_{ab}^{01}\right)\frac{ T_b}{Z_b}\frac{\d \ln T_b}{\d \psi}\right)\right] 
\end{aligned}
\end{equation}
and the heat-flux
\begin{equation}
\begin{aligned}
  \hspace*{-10ex} q_a =  -\frac{T_a m_a}{Z_a e^2} \sum_b \frac{1}{\tau_{ab} n_b} &\left[
    \lang n_a n_b\frac{|\nabla \psi|^2}{B^2 } \rang M_{ab}^{01}\left(\frac{T_a}{Z_a}  \frac{\d \ln\eta_a}{\d \psi} - \frac{T_b}{Z_b}\frac{\d \ln\eta_b}{\d \psi}\right) \right.\\
    &\left.+\lang n_a n_b\frac{|\nabla \psi|^2}{B^2 } e \tilde{\Phi} \rang M_{ab}^{01} \left(\frac{\d \ln T_a}{\d \psi}-\frac{\d \ln T_b}{\d \psi}\right) \right.\\
    &\left.+ \lang n_a n_b\frac{|\nabla \psi|^2}{B^2 } \rang \left(\left(M_{ab}^{01} - M_{ab}^{11}\right)\frac{ T_a}{Z_a} \frac{\d \ln T_a}{\d \psi}
    - \left(M_{ab}^{01} + N_{ab}^{11}\right)\frac{T_b}{Z_b}\frac{\d \ln T_b}{\d \psi}\right) \right]
\end{aligned}
\end{equation}

\section{SFINCS implementation}
\textsc{Sfincs} uses normalized quantities (denoted with a hat), where the normalized radial fluxes are defined in terms of the physical fluxes as
\begin{align}
  \hat{\Gamma}_a &= \frac{\bar{R}}{\bar{n}\bar{v}} \frac{1}{Z_a e} \lang B^{-2} \left(\vec{B} \times \nabla \hat{\psi}\right) \cdot \vec{R}_a \rang, \\
  \hat{Q}_a &= \frac{\bar{R}}{\bar{n}\bar{v} \bar{T}} \frac{1}{Z_a e} \lang B^{-2} \left(\vec{B} \times \nabla \hat{\psi}\right) \cdot \vec{G}_a \rang,
\end{align}
where $\hat{\psi} = \psi/(\bar{B}\hat{R}^2)$ is just a new choice of radial coordinate rather than a normalization -- see the \textsc{Sfincs} manual for more details. We also define the dimensionless parameters $\alpha = e\bar{\Phi}/\bar{T}$ and $\Delta = \frac{\sqrt{2\bar{m}\bar{T}}}{e\bar{R}\bar{B}}$, and the dimensionless collisionality $\nu_n$
\begin{align}
  \bar{\nu} &\equiv \frac{\sqrt{2}}{12 \pi^{3/2}} \frac{\bar{n} e^4 \ln \Lambda}{\epsilon_0^2 \bar{m}^{1/2} \bar{T}^{3/2}} \\
  \nu_n &\equiv \bar{\nu} \frac{\bar{R} \sqrt{\bar{m}}}{\sqrt{2\bar{T}}} =\frac{1}{12 \pi^{3/2}} \frac{\bar{n} \bar{R} e^4 \ln \Lambda}{\epsilon_0^2 \bar{T}^{2}},
\end{align}
so that
\begin{equation}
\frac{1}{\tau_{ab} n_b} = \frac{Z_a^2 Z_b^2}{\hat{m}_a^{1/2} \hat{T}_a^{3/2}} \frac{\sqrt{2} \bar{T}^{1/2}}{\bar{n} \bar{R} \bar{m}^{1/2}} \nu_n = \frac{Z_a^2 Z_b^2}{\hat{m}_a^{1/2} \hat{T}_a^{3/2}} \frac{\sqrt{2} \bar{T}^{1/2}\bar{m}^{1/2} \bar{B}}{\bar{n} \bar{R} \bar{B} \bar{m}} \nu_n = \frac{Z_a^2 Z_b^2}{\hat{m}_a^{1/2} \hat{T}_a^{3/2}} \frac{e\bar{B}}{\bar{n}\bar{m}} \Delta \nu_n.
\end{equation}
With this, the normalized particle and conductive heat-flux becomes
\begin{equation}
\begin{aligned}
  \hspace*{-10ex} \hat{\Gamma}_a =   \frac{\Delta^2 \nu_n}{2}\frac{Z_a\hat{m}_a^{1/2}}{\hat{T}_a^{3/2}}\sum_b Z_b^2 &\left[
    \lang  \hat{n}_a \hat{n}_b\frac{ |\bar{R}\nabla \hat{\psi}|^2}{\hat{B}^2} \rang M_{ab}^{00} \left(\frac{ \hat{T}_a}{Z_a}  \frac{\d \ln\eta_a}{\d \hat{\psi}} -\frac{ \hat{T}_b}{Z_b}  \frac{\d \ln\eta_b}{\d \hat{\psi}}\right) \right.\\
    &\left.+    \lang \hat{n}_a \hat{n}_b\frac{  |\bar{R}\nabla \hat{\psi}|^2}{\hat{B}^2} \hat{\Phi}_1 \rang \alpha  M_{ab}^{00} \left(\frac{\d \ln T_a}{\d \hat{\psi}} -\frac{\d \ln T_b}{\d \hat{\psi}}\right) \right.\\
   &\left. + \lang \hat{n}_a \hat{n}_b \frac{| \bar{R} \nabla \hat{\psi}|^2 }{\hat{B}^2}\rang\left(  \left(M_{ab}^{00} - M_{ab}^{01}\right)\frac{ \hat{T}_a}{Z_a}\frac{\d \ln T_a}{\d \hat{\psi}} 
    - \left(M_{ab}^{00} - \frac{\hat{m}_a \hat{T}_b}{\hat{m}_b \hat{T}_a} M_{ab}^{01}\right)\frac{ \hat{T}_b}{Z_b}\frac{\d \ln T_b}{\d \hat{\psi}}\right)\right] 
\end{aligned}
\end{equation}

\begin{equation}
\begin{aligned}
  \hspace*{-10ex} \hat{q}_a =  -\frac{\Delta^2 \nu_n}{2} \frac{Z_a\hat{m}_a^{1/2}}{\hat{T}_a^{1/2}} \sum_b Z_b^2 &\left[
    \lang \hat{n}_a \hat{n}_b\frac{|\bar{R}\nabla \hat{\psi}|^2}{\hat{B}^2 } \rang M_{ab}^{01}\left(\frac{\hat{T}_a}{Z_a}  \frac{\d \ln\eta_a}{\d \hat{\psi}} - \frac{\hat{T}_b}{Z_b}\frac{\d \ln\eta_b}{\d \hat{\psi}}\right) \right.\\
    &\left.+\lang \hat{n}_a \hat{n}_b\frac{|\bar{R}\nabla \hat{\psi}|^2}{\hat{B}^2 }  \hat{\Phi}_1 \rang \alpha M_{ab}^{01} \left(\frac{\d \ln T_a}{\d \hat{\psi}}-\frac{\d \ln T_b}{\d \hat{\psi}}\right) \right.\\
    &\left.+ \lang \hat{n}_a \hat{n}_b\frac{|\bar{R}\nabla \hat{\psi}|^2}{\hat{B}^2 } \rang \left(\left(M_{ab}^{01} - M_{ab}^{11}\right)\frac{ \hat{T}_a}{Z_a} \frac{\d \ln T_a}{\d \hat{\psi}}
    - \left(M_{ab}^{01} + N_{ab}^{11}\right)\frac{\hat{T}_b}{Z_b}\frac{\d \ln T_b}{\d \hat{\psi}}\right) \right].
\end{aligned}
\end{equation}
Finally, using $\hat{n}_a = \hat{\eta}_a \e^{-Z_a \alpha \hat{\Phi}_1/\hat{T}_a}$, we get
\begin{equation}
\hat{n}_a\hat{n}_b = \hat{\eta}_a \hat{\eta}_b \exp{\left(-\alpha\hat{\Phi}_1\left[\frac{Z_a}{\hat{T}_a} + \frac{Z_b}{\hat{T}_b}\right]\right)},
\end{equation}
where $\hat{\eta}$ is the ``density'' in \textsc{Sfincs}, which is a flux-function. In the code, we define
\begin{align}
  G_{ab}^{(1)} =& \lang \exp{\left(-\alpha\hat{\Phi}_1\left[\frac{Z_a}{\hat{T}_a} + \frac{Z_b}{\hat{T}_b}\right]\right)}\frac{|\bar{R}\nabla \hat{\psi}|^2}{\hat{B}^2 } \rang \\
  G_{ab}^{(2)} =& \lang \exp{\left(-\alpha\hat{\Phi}_1\left[\frac{Z_a}{\hat{T}_a} + \frac{Z_b}{\hat{T}_b}\right]\right)}\frac{|\bar{R}\nabla \hat{\psi}|^2}{\hat{B}^2 } \hat{\Phi}_1 \rang,
\end{align}
so that
\begin{equation}
\begin{aligned}
  \hspace*{-10ex} \hat{\Gamma}_a =   \frac{\Delta^2 \nu_n}{2}\frac{Z_a\hat{m}_a^{1/2} \hat{\eta}_a}{\hat{T}_a^{3/2}}\sum_b Z_b^2 \hat{\eta}_b&\left[
    G_{ab}^{(1)} M_{ab}^{00} \left(\frac{ \hat{T}_a}{Z_a}  \frac{\d \ln\eta_a}{\d \hat{\psi}} -\frac{ \hat{T}_b}{Z_b}  \frac{\d \ln\eta_b}{\d \hat{\psi}}\right) \right.\\
    &\left.+    G_{ab}^{(2)} \alpha  M_{ab}^{00} \left(\frac{\d \ln T_a}{\d \hat{\psi}} -\frac{\d \ln T_b}{\d \hat{\psi}}\right) \right.\\
   &\left. +G_{ab}^{(1)}\left(  \left(M_{ab}^{00} - M_{ab}^{01}\right)\frac{ \hat{T}_a}{Z_a}\frac{\d \ln T_a}{\d \hat{\psi}} 
    - \left(M_{ab}^{00} - \frac{\hat{m}_a \hat{T}_b}{\hat{m}_b \hat{T}_a} M_{ab}^{01}\right)\frac{ \hat{T}_b}{Z_b}\frac{\d \ln T_b}{\d \hat{\psi}}\right)\right],
\end{aligned}
\end{equation}

\begin{equation}
\begin{aligned}
  \hspace*{-10ex} \hat{q}_a =  -\frac{\Delta^2 \nu_n}{2} \frac{Z_a\hat{m}_a^{1/2} \hat{\eta}_a}{\hat{T}_a^{1/2}} \sum_b Z_b^2 \hat{\eta}_b &\left[
    G_{ab}^{(1)}  M_{ab}^{01}\left(\frac{\hat{T}_a}{Z_a}  \frac{\d \ln\eta_a}{\d \hat{\psi}} - \frac{\hat{T}_b}{Z_b}\frac{\d \ln\eta_b}{\d \hat{\psi}}\right) \right.\\
    &\left.+G_{ab}^{(2)} \alpha M_{ab}^{01} \left(\frac{\d \ln T_a}{\d \hat{\psi}}-\frac{\d \ln T_b}{\d \hat{\psi}}\right) \right.\\
    &\left.+ G_{ab}^{(1)} \left(\left(M_{ab}^{01} - M_{ab}^{11}\right)\frac{ \hat{T}_a}{Z_a} \frac{\d \ln T_a}{\d \hat{\psi}}
    - \left(M_{ab}^{01} + N_{ab}^{11}\right)\frac{\hat{T}_b}{Z_b}\frac{\d \ln T_b}{\d \hat{\psi}}\right) \right],
\end{aligned}
\end{equation}
and
\begin{equation}
\hat{Q}_a = \hat{q}_a + \frac{5}{2} \hat{T}_a \hat{\Gamma}_a. 
\end{equation}

\paragraph{Thanks to Sarah Newton} for bringing to my attention that the classical transport can be calculated using the Braginskii matrices.

\end{document}

### 2019-01-17 fortran/version3/classicalTransport.F90

@@ -6,28 +6,36 @@ module classicalTransport
 
 contains
  
-  subroutine calculateClassicalParticleFlux(classicalPF)
+  subroutine calculateClassicalParticleFlux(classicalPF, classicalHF)
     use globalVariables, only: Nspecies, Ntheta, Nzeta, Delta, nu_n, mHats, THats, Zs, thetaWeights, zetaWeights, DHat, gpsipsi, BHat,alpha, Phi1Hat, VPrimeHat, NHats, dNHatdpsiHats, dTHatdpsiHats
     implicit None
     integer :: itheta, izeta, iSpeciesA, iSpeciesB
-    PetscScalar :: geometry1, geometry2, y
-    PetscScalar, dimension(Nspecies), intent(out) :: classicalPF
+    PetscScalar :: geometry1, geometry2, xab2, Mab00, Mab01, Mab11, Nab11 !, y, classicalPF_OLD
+    PetscScalar, dimension(Nspecies), intent(out) :: classicalPF, classicalHF
 
     do ispeciesA=1,Nspecies
+       
        classicalPF(ispeciesA) = 0
+       classicalHF(ispeciesA) = 0
+       ! classicalPF_OLD = 0
 
        do ispeciesB=1,Nspecies
-          if (ispeciesA == ispeciesB) then
-             ! term is exactly zero
-             cycle
-          end if
           
+          xab2 = mHats(ispeciesA)*THats(ispeciesB)/(mHats(ispeciesB)*THats(ispeciesA))
+
+          ! Braginskii matrix elements
+          Mab00 = - (1+mHats(ispeciesA)/mHats(ispeciesB)) * (1 + xab2)
+          Mab01 = - 1.5 * (1+mHats(ispeciesA)/mHats(ispeciesB))
+          Mab11 = -(13. + 16. * xab2 + 30. * xab2**2)/4.
+          Nab11 = 27. * mHats(ispeciesA)/(4.*mHats(ispeciesA))
+          Mab00 = Mab00/((1+xab2)**(2.5))
+          Mab01 = Mab01/((1+xab2)**(2.5))
+          Mab11 = Mab11/((1+xab2)**(2.5))
+          Nab11 = Nab11/((1+xab2)**(2.5))
           
[…]

### 2019-01-17 fortran/version3/diagnostics.F90

@@ -235,7 +235,7 @@
     call extractPhi1(solutionWithDeltaF)
 
     ! Calculate the classical transport:
-    call calculateClassicalParticleFlux(classicalParticleFlux_psiHat)
+    call calculateClassicalParticleFlux(classicalParticleFlux_psiHat,classicalHeatFlux_psiHat)
 
     ! The solution vector contains the departure from a Maxwellian, not the "full f" distribution function.
     ! Form the full f:
@@ -687,6 +687,7 @@
        heatFlux_vd_psiN = ddpsiN2ddpsiHat * heatFlux_vd_psiHat
        heatFlux_withoutPhi1_psiN = ddpsiN2ddpsiHat * heatFlux_withoutPhi1_psiHat
        classicalParticleFlux_psiN = ddpsiN2ddpsiHat * classicalParticleFlux_psiHat
+       classicalHeatFlux_psiN = ddpsiN2ddpsiHat * classicalHeatFlux_psiHat
 
        particleFlux_vm0_rHat = ddrHat2ddpsiHat * particleFlux_vm0_psiHat
        particleFlux_vm_rHat = ddrHat2ddpsiHat * particleFlux_vm_psiHat
@@ -708,6 +709,7 @@
        heatFlux_vd_rHat = ddrHat2ddpsiHat * heatFlux_vd_psiHat
        heatFlux_withoutPhi1_rHat = ddrHat2ddpsiHat * heatFlux_withoutPhi1_psiHat
        classicalParticleFlux_rHat = ddrHat2ddpsiHat * classicalParticleFlux_psiHat
+       classicalHeatFlux_rHat = ddrHat2ddpsiHat * classicalHeatFlux_psiHat
 
        particleFlux_vm0_rN = ddrN2ddpsiHat * particleFlux_vm0_psiHat
        particleFlux_vm_rN = ddrN2ddpsiHat * particleFlux_vm_psiHat
@@ -729,6 +731,7 @@
        heatFlux_vd_rN = ddrN2ddpsiHat * heatFlux_vd_psiHat
        heatFlux_withoutPhi1_rN = ddrN2ddpsiHat * heatFlux_withoutPhi1_psiHat
        classicalParticleFlux_rHat = ddrN2ddpsiHat * classicalParticleFlux_psiHat
+       classicalHeatFlux_rHat = ddrN2ddpsiHat * classicalHeatFlux_psiHat
 
[…]

### 2019-01-17 fortran/version3/sfincs_main.F90

@@ -146,10 +146,13 @@ module sfincs_main
 
     ! calculate the classical transport without Phi1
     ! this can be done without solving the system, and so is done here
-    call calculateClassicalParticleFlux(classicalParticleFluxNoPhi1_psiHat)
+    call calculateClassicalParticleFlux(classicalParticleFluxNoPhi1_psiHat,classicalHeatFluxNoPhi1_psiHat)
     classicalParticleFluxNoPhi1_psiN = ddpsiN2ddpsiHat * classicalParticleFluxNoPhi1_psiHat
     classicalParticleFluxNoPhi1_rHat = ddrHat2ddpsiHat * classicalParticleFluxNoPhi1_psiHat
     classicalParticleFluxNoPhi1_rN = ddrN2ddpsiHat * classicalParticleFluxNoPhi1_psiHat
+    classicalHeatFluxNoPhi1_psiN = ddpsiN2ddpsiHat * classicalHeatFluxNoPhi1_psiHat
+    classicalHeatFluxNoPhi1_rHat = ddrHat2ddpsiHat * classicalHeatFluxNoPhi1_psiHat
+    classicalHeatFluxNoPhi1_rN = ddrN2ddpsiHat * classicalHeatFluxNoPhi1_psiHat
 
     ! Create HDF5 data structures, and save the quantities that will not change
     ! at each iteration of the solver (i.e. save all quantities except diagnostics.)

### 2019-01-17 fortran/version3/writeHDF5Output.F90

@@ -317,6 +317,11 @@ contains
        call writeHDF5Field("classicalParticleFluxNoPhi1_psiN", classicalParticleFluxNoPhi1_psiN, dspaceIDForSpecies, dimForSpecies, "Classical particle flux projected onto \nabla psiN, calculcated, with Phi1 = 0.")
        call writeHDF5Field("classicalParticleFluxNoPhi1_rHat", classicalParticleFluxNoPhi1_rHat, dspaceIDForSpecies, dimForSpecies, "Classical particle flux projected onto \nabla rHat, calculcated, with Phi1 = 0.")
        call writeHDF5Field("classicalParticleFluxNoPhi1_rN", classicalParticleFluxNoPhi1_rN, dspaceIDForSpecies, dimForSpecies, "Classical particle flux projected onto \nabla rN, calculcated, with Phi1 = 0.")
+       !!Added by SB 2019-01!!
+       call writeHDF5Field("classicalHeatFluxNoPhi1_psiHat", classicalHeatFluxNoPhi1_psiHat, dspaceIDForSpecies, dimForSpecies, "Classical particle flux projected onto \nabla psiHat, calculcated with Phi1 = 0.")
+       call writeHDF5Field("classicalHeatFluxNoPhi1_psiN", classicalHeatFluxNoPhi1_psiN, dspaceIDForSpecies, dimForSpecies, "Classical particle flux projected onto \nabla psiN, calculcated, with Phi1 = 0.")
+       call writeHDF5Field("classicalHeatFluxNoPhi1_rHat", classicalHeatFluxNoPhi1_rHat, dspaceIDForSpecies, dimForSpecies, "Classical particle flux projected onto \nabla rHat, calculcated, with Phi1 = 0.")
+       call writeHDF5Field("classicalHeatFluxNoPhi1_rN", classicalHeatFluxNoPhi1_rN, dspaceIDForSpecies, dimForSpecies, "Classical particle flux projected onto \nabla rN, calculcated, with Phi1 = 0.")
        !!!!!!!!!!!!!!!!!!!!!!!
 
        call writeHDF5Field("dPhiHatdpsiHat", dPhiHatdpsiHat, "")
@@ -587,6 +592,10 @@ contains
        call writeHDF5ExtensibleField(iterationNum, "classicalParticleFlux_psiN", classicalParticleFlux_psiHat, ARRAY_ITERATION_SPECIES, "")
[…]

### 2019-02-12 doc/manual/version3/inputParameters.tex
Changed discreteAdjointOption to a boolean. Added to manual.

@@
 Options related to the adjoint solve are detailed in section \ref{sec:adjointOptions}.
-{\ttfamily RHSMode} = 4 must be used with \parlink{includePhi1}=\false, \\ \parlink{EParallelHat}\texttt{=0}, \parlink{magneticDriftScheme}\texttt{>0}, \parlink{constraintScheme}\texttt{=-1} or 1, and \\ \parlink{collisionOperator}\texttt{=0}.
+{\ttfamily RHSMode} = 4 must be used with \\ \parlink{includePhi1}=\false, \parlink{EParallelHat}\texttt{=0}, \parlink{magneticDriftScheme}\texttt{>0}, \\ \parlink{constraintScheme}\texttt{=-1} or 1, and \parlink{collisionOperator}\texttt{=0}.
 \\
@@
 Options related to the adjoint solve are detailed in section \ref{sec:adjointOptions}.
-{\ttfamily RHSMode} = 5 must be used with \parlink{includePhi1}=\false, \parlink{EParallelHat}\texttt{=0}, \parlink{magneticDriftScheme}\texttt{>0}, \parlink{constraintScheme}\texttt{=-1} or 1, and \\ \parlink{collisionOperator}\texttt{=0}.\\
+{\ttfamily RHSMode} = 5 must be used with \\ \parlink{includePhi1}=\false, \parlink{EParallelHat}\texttt{=0}, \parlink{magneticDriftScheme}\texttt{>0}, \\ \parlink{constraintScheme}\texttt{=-1} or 1, and \parlink{collisionOperator}\texttt{=0}.\\
 
@@
 All output parameters will be included at the ambipolar \parlink{Er}.
-Options related to this parameter are \parlink{NEr\_ambipolarSolve}, \parlink{Er\_search\_tolerance\_dx}, \parlink{Er\_search\_tolerance\_f}, \parlink{ambipolarSolveOption}, \parlink{Er\_min}, and \parlink{Er\_max}.}
+Options related to this parameter are \parlink{NEr\_ambipolarSolve}, \parlink{Er\_search\_tolerance\_dx}, \parlink{Er\_search\_tolerance\_f}, \\ \parlink{ambipolarSolveOption}, \parlink{Er\_min}, and \parlink{Er\_max}.}
 
@@
 {1}
-{When \parlink{ambipolarSolve} == \true.}
-{Indicates which root solving algorithm to use for ambipolar solve.
-If \parlink{ambipolarSolveOption}==1 a Newton bisection method is used, \parlink{ambipolarSolveOption}==2 a Brent method is used, and if \parlink{ambipolarSolveOption}==3 a Newton method is used.
-If \parlink{ambipolarSolveOption}==1 or 3, an adjoint method will used to compute the $\partial J_r/\partial E_r$, so must be used with \parlink{includePhi1}==\false, \parlink{EParallelHat}==0, \parlink{magneticDriftScheme}>0, \parlink{constraintScheme}==-1 or 1, and \parlink{collisionOperator}==0.}
+{When \parlink{ambipolarSolve} == \true}
+{Indicates which root solving algorithm to use for ambipolar solve. \\
+\parlink{ambipolarSolveOption}==1 a Newton bisection method is used \\ 
+\parlink{ambipolarSolveOption}==2 a Brent method is used \\
+\parlink{ambipolarSolveOption}==3 a Newton method is used.
+
+If \parlink{ambipolarSolveOption}==1 or 3, an adjoint method will used to compute the $\partial J_r/\partial E_r$, so must be used with \parlink{includePhi1}==\false, \parlink{EParallelHat}==0, \parlink{magneticDriftScheme}$>0$, \parlink{constraintScheme}==-1 or 1, and \parlink{collisionOperator}==0.}
 
@@
 {\ttfamily geometryScheme}==13: Read Boozer geometry from the {\ttfamily geometryParamters} namelist.
-Here the geometry parameters are specified by \parlink{boozer\_bmnc}, \parlink{GHat}, \parlink{IHat}, and \parlink{iota}.
+Here the geometry parameters are specified by \parlink{Nperiods}, \parlink{boozer\_bmnc}, \parlink{GHat}, \parlink{IHat}, and \parlink{iota}.
 This has been implemented for use with STELLOPT for direct optimization of the Boozer spectrum.
@@
 \end{gather}
-The first index of \texttt{ Boozer\_bmnc} corresponds to $m \in $[0,\parlink{mmax\_boozer}], and the second corresponds to $n \in $[-\parlink{nmax\_boozer},\parlink{nmax\_boozer}].
-Here \parlink{Nperiods} can be specified in the input file.
-}
-
-\myhrule
-
-\param{mmax\_boozer}
-{integer}
-{40}
-{Only when \parlink{geometryScheme} == 13.}
-{Boozer geometry is described for $m \leq $ \texttt{mmax\_boozer}. }
-
-\myhrule
-
-\param{nmax\_boozer}
-{integer}
-{40}
-{Only when \parlink{geometryScheme} == 13.}
-{Boozer geometry is described for $|n| \leq $ \texttt{nmax\_boozer}. }
+The first index of \texttt{ Boozer\_bmnc} corresponds to $m \in $[0,40], and the second corresponds to $n \in $[-40,40]. Here \parlink{Nperiods} can be specified in the input file.
+}
 
@@
 {If this parameter is \true, the derivative of the bootstrap current (\parlink{FSABjHatOverRootFSAB2}) is computed with respect to geometry parameters.
-If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
+If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bsubzetamnc}, and \texttt{gmnc} which are found in the wout file.
 Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
@@
 where \parlink{Zs} and \parlink{particleFlux\_vm\_rN} can be found elsewhere in this document.
-If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
+If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bsubzetamnc}, and \texttt{gmnc} which are found in the wout file.
 Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
@@
 where \parlink{Zs} and \parlink{heatFlux\_vm\_rN} can be found elsewhere in this document.
-If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
+If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bsubzetamnc}, and \texttt{gmnc} which are found in the wout file.
 Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
@@
 If parameter is \true, the derivative of the heat flux (\parlink{heatFlux\_vm\_rN}) will be computed for this species with respect to geometry parameters.
-If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
+If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bsubzetamnc}, and \texttt{gmnc} which are found in the wout file.
 Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
@@
 If parameter is {\ttfamily .true.}, the derivative of the particle flux (\parlink{particleFlux\_vm\_rN}) will be computed for this species with respect to geometry parameters.
-If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
+If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bsubzetamnc}, and \texttt{gmnc} which are found in the wout file.
 Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
@@
 If parameter is {\ttfamily .true.}, the derivative of the parallel flow (\parlink{FSABVelocityUsingFSADensityOverRootFSAB2}) will be computed for this species with respect to geometry parameters.
-If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
+If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bsubzetamnc}, and \texttt{gmnc} which are found in the wout file.
 Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
+
+\myhrule
+
+\param{discreteAdjointOption}
+{boolean}
+{\true}
+{When \parlink{RHSMode} == 4 or 5}
+{When \parlink{RHSMode} is \true, the discrete adjoint technique is used.
+If \false, the continuous technique is used.
+Both should give the same result within discretization error, but the discrete approach should be slightly faster.
+}
 
@@
 {When \parlink{RHSMode} == 4 or 5}
-{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \parlink{adjointParallelFlowOption} will computed for Fourier modes $|n| < $ \texttt{nMaxAdjoint}.
+{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \\ \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \\ \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \\ \parlink{adjointParallelFlowOption} will computed for Fourier modes $|n| \leq $ \texttt{nMaxAdjoint}.
 }
@@
 {When \parlink{RHSMode} == 4 or 5}
-{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \parlink{adjointParallelFlowOption} will computed for Fourier modes $m < $ \texttt{mMaxAdjoint}.
+{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \\ \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \\ \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \\ \parlink{adjointParallelFlowOption} will computed for Fourier modes $m \leq $ \texttt{mMaxAdjoint}.
 }
@@
 {When \parlink{RHSMode} == 4 or 5}
-{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \parlink{adjointParallelFlowOption} will computed for Fourier modes $|n| > $ \texttt{nMaxAdjoint}.
+{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \\ \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \\ \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \\ \parlink{adjointParallelFlowOption} will computed for Fourier modes $|n| \geq $ \texttt{nMaxAdjoint}.
 }
@@
 {When \parlink{RHSMode} == 4 or 5}
-{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \parlink{adjointParallelFlowOption} will computed for Fourier modes $m > $ \texttt{mMaxAdjoint}.
+{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \\ \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \\ \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \\ \parlink{adjointParallelFlowOption} will computed for Fourier modes $m \geq $ \texttt{mMaxAdjoint}.
 }

### 2019-02-12 fortran/version3/adjointDiagnostics.F90

@@ -115,8 +115,7 @@ module adjointDiagnostics
 			this_zetaWeights => zetaWeights
 			this_DHat => DHat
 
-      select case (discreteAdjointOption)
-        case (0) ! continuous
+			if (discreteAdjointOption .eqv. .false.) then
           ! Scatter deltaF to master proc
           call VecScatterCreateToZero(deltaF, VecScatterContext, deltaFOnProc0, ierr)
           call VecScatterBegin(VecScatterContext, deltaF, deltaFOnProc0, INSERT_VALUES, SCATTER_FORWARD, ierr)
@@ -183,9 +182,9 @@ module adjointDiagnostics
             end do ! ispecies
           end if ! masterProc
 
-        case (1) ! discrete
+			else ! discrete
           call VecDot(deltaF, deltaG, result, ierr)
-      end select
+			end if
     end subroutine
 
     !> Evaluates the term in the sensitivity derivative of the heat flux

### 2019-02-12 fortran/version3/solver.F90

@@ -106,7 +106,7 @@ module solver
        end if
 
        call SNESGetKSP(mysnes, KSPInstance, ierr)
-			 if (discreteAdjointOption==0) then
+			 if (discreteAdjointOption .eqv. .false.) then
 					call KSPCreate(MPIComm,KSPInstance_adjoint, ierr)
 					call preallocateMatrix(adjointMatrix,1)
 					call populateMatrix(adjointMatrix,4,dummyVec)
@@ -124,7 +124,7 @@ module solver
 !!$       ! Syntax for PETSc version 3.5 and later
 !!$       call KSPSetOperators(KSPInstance, matrix, preconditionerMatrix, ierr)
 !!$#endif
-					if (discreteAdjointOption==0) then
+					if (discreteAdjointOption .eqv. .false.) then
 #if (PETSC_VERSION_MAJOR < 3 || (PETSC_VERSION_MAJOR==3 && PETSC_VERSION_MINOR < 5))
 				 ! Syntax for PETSc versions up through 3.4
 						 call KSPSetOperators(KSPInstance_adjoint, adjointMatrix, adjointPreconditionerMatrix, SAME_PRECONDITIONER, ierr)
@@ -150,7 +150,7 @@ module solver
           call PetscViewerAndFormatCreate(PETSC_VIEWER_STDOUT_WORLD, PETSC_VIEWER_DEFAULT, vf, ierr) !!Added by AM 2016-07-06
           call KSPMonitorSet(KSPInstance, KSPMonitorDefault, vf, PetscViewerAndFormatDestroy, ierr) !!Added by AM 2016-07-06 
 #endif
-					if (discreteAdjointOption==0) then
+					if (discreteAdjointOption .eqv. .false.) then
 						call KSPGetPC(KSPInstance_adjoint, preconditionerContext_adjoint, ierr)
 						call PCSetType(preconditionerContext_adjoint, PCLU, ierr)
 						call KSPSetType(KSPInstance_adjoint, KSPGMRES, ierr)   ! Set the Krylov solver algorithm to GMRES
@@ -177,7 +177,7 @@ module solver
 !!$       ! Syntax for PETSc version 3.5 and later
 !!$       call KSPSetOperators(KSPInstance, matrix, matrix, ierr)
 !!$#endif
-			 if (discreteAdjointOption==0) then
+			 if (discreteAdjointOption .eqv. .false.) then
           ! Direct solver:
[…]

### 2019-02-12 fortran/version3/validateInput.F90

@@ -1225,7 +1225,7 @@ subroutine validateInput()
       end if
       stop
     end if
-    if (discreteAdjointOption==0 .and. ((((includeXDotTerm .eqv. .true.) .and. &
+    if ((discreteAdjointOption .eqv. .false.) .and. ((((includeXDotTerm .eqv. .true.) .and. &
       (useDKESExBDrift .eqv. .false.) .and. (includeElectricFieldTermInXiDot .eqv. .true.)) &
       .or. ((includeXDotTerm .eqv. .false.) .and. (useDKESExBDrift .eqv. .true.) .and. &
       (includeElectricFieldTermInXiDot .eqv. .false.))) .eqv. .false.)) then

### 2019-02-12 doc/manual/version3/inputParameters.tex
Added stellarator asymmetric option for geometryScheme 13.

@@
 \begin{gather}
-B(\theta_B,\zeta_B) = \sum_{m,n}\mathrm{\texttt{boozer\_bmnc}}(m,n) \cos(m\theta_B-n\mathrm{\texttt{Nperiods}} \zeta_B),
+B(\theta_B,\zeta_B) = \sum_{m,n}\mathrm{\texttt{boozer\_bmnc}}(m,n) \cos(m\theta_B-n\mathrm{\texttt{Nperiods}} \zeta_B) + \mathrm{\texttt{boozer\_bmnc}}(m,n) \sin(m\theta_B-n\mathrm{\texttt{Nperiods}} \zeta_B),
 \end{gather}
 The first index of \texttt{ Boozer\_bmnc} corresponds to $m \in $[0,40], and the second corresponds to $n \in $[-40,40]. Here \parlink{Nperiods} can be specified in the input file.
+}
+
+\myhrule
+
+\param{boozer\_bmns}
+{2D array of reals}
+{0}
+{Only when \parlink{geometryScheme} == 13. }
+{Fourier harmonics describing the magnetic field in Boozer coordinates,
+\begin{gather}
+B(\theta_B,\zeta_B) = \sum_{m,n}\mathrm{\texttt{boozer\_bmnc}}(m,n) \cos(m\theta_B-n\mathrm{\texttt{Nperiods}} \zeta_B) + \mathrm{\texttt{boozer\_bmnc}}(m,n) \sin(m\theta_B-n\mathrm{\texttt{Nperiods}} \zeta_B),
+\end{gather}
+The first index of \texttt{ Boozer\_bmns} corresponds to $m \in $[0,40], and the second corresponds to $n \in $[-40,40]. Here \parlink{Nperiods} can be specified in the input file.
 }

### 2019-02-12 fortran/version3/globalVariables.F90

@@ -73,6 +73,7 @@ module globalVariables
 	integer, parameter :: mmax_boozer = 40
 	integer, parameter :: nmax_boozer = 40
 	PetscScalar, dimension(0:mmax_boozer,-nmax_boozer:nmax_boozer) :: boozer_bmnc
+	PetscScalar, dimension(0:mmax_boozer,-nmax_boozer:nmax_boozer) :: boozer_bmns
 
   ! ********************************************************
   ! ********************************************************

### 2019-02-12 fortran/version3/readInput.F90

@@ -34,7 +34,7 @@ contains
          psiHat_wish, psiN_wish, rHat_wish, rN_wish, &
          VMECRadialOption, VMEC_Nyquist_option, rippleScale, EParallelHatSpec_bcdatFile, Nperiods, &
 !				 boozer_bmnc, mmax_boozer, nmax_boozer
-				 boozer_bmnc
+				 boozer_bmnc, boozer_bmns
 
     namelist / speciesParameters / mHats, Zs, nHats, THats, &
          dNHatdpsiHats, dTHatdpsiHats, &

### 2019-03-05 doc/manual/version3/equations.tex
Substantial revision to manual

@@
 $C_s$ to denote the collision operator, and $S_s$ to denote the source-sink term. 
+The magnetic moment is $\mu = v_{\perp}^2/\left(2 B\right)$ and the total energy is $W_s = v^2 / 2 + Z_s e \Phi / m_s$, 
+with the electrostatic potential $\Phi$. 
 A more detailed discussion of the equations implemented in the code can be found
@@
 
-\section{Default: Full $\vect{E}\times\vect{B}$ trajectories; no poloidal or toroidal magnetic drifts; flux function potential}
+\section{System of equations}
+Although the prime function of \sfincs~is to solve the drift-kinetic equation, when run in its most general way (\parlink{includePhi1} = \true and \parlink{readExternalPhi1} = \false) it simultaneously also solves the following equations:
+\begin{align}
+%\begin{equation}
+%\begin{dcases}
+	& \lambda + \sum_s Z_s \int d^3 v \, \left(f_{s 0} + f_{s 1} \right) = 0 \;\;\; (\mathrm{quasineutrality}), \label{eq:QuasineutralityEq}\\ %\nonumber \\
+	%
+	& %\left\langle \int d^3 v \, f_s \right\rangle = n_{0 s} \left(\psi\right), 
+	\left\langle \int d^3 v \, f_{s 1} \right\rangle = 0, 
+	\label{eq:DensityEq}\\ %\nonumber \\
+	%
+	& %\left\langle \frac{m_s}{3} \int d^3 v \, v^2 \, f_s \right\rangle = n_{0 s} \left(\psi\right) T_s \left(\psi\right),  
+	\left\langle \int d^3 v \, v^2 \, f_{s 1} \right\rangle = 0,
+	\label{eq:EnergyEq}\\
+	& \left\langle \Phi_1 \right\rangle = 0. \label{eq:Phi1Eq} 
+%\label{eq:SystemOfEquations} 
+%\end{dcases}
+%\end{equation}
+\end{align}
+If $\Phi_1$ is not included (\parlink{includePhi1} = \false) or if it is read from a file (\parlink{readExternalPhi1} = \false) then Eqs.~\eqref{eq:QuasineutralityEq} and \eqref{eq:Phi1Eq} are not solved.
+The unknowns in the system which are calculated are $\left\{f_{s1} \left(\theta, \zeta, x, \xi\right), \Phi_1 \left(\theta, \zeta\right), S_{s 1}, S_{s 2}, \lambda\right\}$ for the full system, and $\left\{f_{s1} \left(\theta, \zeta, x, \xi\right), S_{s 1}, S_{s 2}\right\}$ if $\Phi_1$ is not an unknown.
+The potential variation is defined through $\Phi = \Phi_0 + \Phi_1$, where $\Phi_0 = \Phi_0\left(r\right) = \left\langle \Phi \right\rangle$ and $\Phi_1 = \Phi_1\left(\theta, \zeta\right)$ with $\left|\Phi_1\right| \ll \left|\Phi_0\right|$ (the variation of the electrostatic potential along flux-surfaces is also assumed to be small compared to the radial variation).
+This implies that the radial electric field is $E_r = - d\Phi_0/dr$ (see \parlink{Er}).
+$S_{s 1}$ and $S_{s 2}$ represent the source-sink term (see section \ref{sec:source-sink}), and $\lambda$ (\parlink{lambda}) is a Lagrange multiplier which should come out as negligible in quasineutrality.
+We note that the number of degrees of freedom for the system is\newline 
+\Ntheta~$\times$~\Nzeta~$\times$~\Nx~$\times$~\Nxi~$\times$~\parlink{Nspecies}~+~\Ntheta~$\times$~\Nzeta~+~2~$\times$~\parlink{Nspecies}~+~1\newline 
+when $\Phi_1$ is included as an unknown and otherwise\newline
+\Ntheta~$\times$~\Nzeta~$\times$~\Nx~$\times$~\Nxi~$\times$~\parlink{Nspecies}~+~2~$\times$~\parlink{Nspecies}.
+
+\todo{It seems like Eqs.~\eqref{eq:DensityEq} and \eqref{eq:EnergyEq} are different for PAS}
+
+\section{$0^{\mathrm{th}}$ order distribution}
+The lowest-order distribution function is either a Maxwellian flux function $\displaystyle f_{s0}\left(r\right) = f_{sM}\left(r\right) = n_{s 0}\left(r\right) \left(\frac{m_s}{2 \pi T_s}\right)^{3/2} \exp \left(- x^2\right)$ if \parlink{includePhi1} = \false,
+or contains the flux-surface variation $\displaystyle f_{s0}\left(r,\theta,\zeta\right) = f_{sM}\left(r\right) \exp \left[- Z_s e \Phi_1(\theta,\zeta) / T_s \right]$ if \parlink{includePhi1} = \true. 
+Here $n_{s 0}$ represents the input density given by \parlink{nHats}. 
+Note that in a calculation with \parlink{includePhi1} = \true, although $n_{s 0}\left(r\right)$ is a flux function it is not neccessarily equal to the flux-surface-averaged lowest-order density (given by $\left\langle \int f_{s 0} \, d^3v \right\rangle$) since the condition $\displaystyle \left\langle \Phi_1 \right\rangle = 0$ does not imply that $\displaystyle \left\langle \exp \left(- Z_s e \Phi_1 / T_s \right) \right\rangle = 1$.
+
+\section{Collision operator}
+\label{sec:CollisionOperator}
+The total collision operator is a sum of linearized collision operators with each species: 
+%\begin{equation}
+$ \displaystyle
+C_s\left[f_s\right] \equiv \sum_b C_{s b}^{l}\left[f_s, f_b\right],
+$
+%\label{eq:CollisionOperatorTot}
+%\end{equation}
+where 
+%\begin{equation}
+$ \displaystyle
+C_{sb}^{l}\left[f_s, f_b\right] \equiv C_{s b}\left[f_{s 1}, f_{b M}\right] + C_{s b}\left[f_{s M}, f_{b 1}\right]
+$
+%\label{eq:LinearizedCollisionOperator}
+%\end{equation}
+and $C_{s b}\left[f_{s}, f_{b}\right]$ is the Fokker-Planck collision operator. 
+$C_{s b}\left[f_{s 1}, f_{b M}\right]$ %in Eq.~\eqref{eq:LinearizedCollisionOperator} 
+is referred to as the test particle part 
+and $C_{a b}\left[f_{a M}, f_{b 1}\right]$ is the field particle part.
+The linearized collision operator may be written as 
+%\begin{equation}
+$ \displaystyle
+C_{a b}^{l} = C_{a b}^{L} + C_{a b}^{E} + C_{a b}^{F},
+$
+%\label{eq:LinearizedCollisionOperatorParts}
+%\end{equation}
+where $C_{a b}^{L} + C_{a b}^{E}$ together represent the test particle part and $C_{a b}^{F}$ the field particle part. 
+Running \sfincs~with \parlink{collisionOperator}~=~1 only the Lorentz part $C_{a b}^{L}$ is kept and $C_{a b}^{E} = C_{a b}^{F} = 0$, 
+whereas when \parlink{collisionOperator}~=~0 all three terms in $C_{a b}^{l}$ are kept. 
+A specification of the three terms and a more detailed description can be found in Appendix~A of \cite{Mollen2015} (and in the technical documentation).
+
+Note that even if \parlink{includePhi1} = \true, the default collision operator is defined without the flux-surface variation, i.e.
+with respect to $f_{sM}\left(r\right)$ instead of $\displaystyle f_{s0}\left(r,\theta,\zeta\right) = f_{sM}\left(r\right) \exp \left[- Z_s e \Phi_1(\theta,\zeta) / T_s \right]$.
+To include the variation, and use 
+$ \displaystyle
+C_{sb}^{l}\left[f_s, f_b\right] \equiv C_{s b}\left[f_{s 1}, f_{b 0}\right] + C_{s b}\left[f_{s 0}, f_{b 1}\right],
+$
+set\newline \parlink{includePhi1InCollisionOperator} = \true (this is not default in \sfincs~because the memory and time requirements can increase significantly but the effect is usually small). 
+It is also possible to include the temperature equilibration term $C_{ab}[ f_{aM}, f_{bM}]$ in the collision operator (or $C_{ab}[ f_{a0}, f_{b0}]$ if \parlink{includePhi1InCollisionOperator} = \true) by using\newline \parlink{includeTemperatureEquilibrationTerm} = \true, but this term does not directly give any parallel or radial transport.
+
+\section{Source-sink term}
+\label{sec:source-sink}
+\todo{Can Matt check this section?}\newline
+The source-sink term $S_s$ is introduced together with the constraints given by Eqs.~\eqref{eq:DensityEq} and \eqref{eq:EnergyEq} to ensure that a steady-state solution exists. 
+This term is controlled by \parlink{constraintScheme}. 
+The default (recommended) setting \parlink{constraintScheme} = -1 implies that $S_s = \left(a_2 x^2 + a_0\right) e^{-x^2}$ represents a particle and heat source if the Fokker-Planck collision operator is used (\parlink{collisionOperator} = 0), where the coefficients $a_0$ and $a_2$ are calculated in the simulation (see \parlink{sources}).
+\newline\todo{Can Matt explain what is calculated when pitch-angle-scattering is used?}\newline
+Note that $\Phi_1$ is never included in the source-sink term irrespective of the type of calculation.
+
+\section{Description of various drift-kinetic equations}
+\label{sec:DKequations}
+
+Here we describe the most common forms of the drift-kinetic equation and their settings in \sfincs. 
+A more detailed description of the different forms can be found in \cite{sfincsPaper}, and in \cite{Mollen2018} for calculations including $\Phi_1$. 
+The total drift velocity is $\vect{v}_{ds} = \vect{v}_{ms} + \vect{v}_{E}$, i.e.
+the sum of the magnetic drift velocity $\vect{v}_{ms}$ and the $\vect{E} \times \vect{B}$-drift velocity $\vect{v}_{E}$.
+The radial magnetic drift is given by $\displaystyle \vect{v}_{ms} \cdot \nabla r = \left(m_s/Z_s e B^2\right) \left( v_{\perp}^2/2  + v_\|^2 \right) \vect{b}~\times~\nabla B~\cdot~\nabla r$. 
+The poloidal and toroidal magnetic drifts are set by \parlink{magneticDriftScheme}. 
+The $\vect{E} \times \vect{B}$-drift is $\vect{v}_E = \left(\vect{b} \times \nabla \Phi \right) / B$, and its radial part is only present if $\Phi_1$ is included $\displaystyle \vect{v}_{E} \cdot \nabla r = \left(\vect{b} \times \nabla \Phi_1 \cdot \nabla r\right) / B$
+
+\subsection{Default: Full $\vect{E}\times\vect{B}$ trajectories; no poloidal or toroidal magnetic drifts; flux function potential}
 
@@
 
-\section{DKES $\vect{E}\times\vect{B}$ trajectories; no poloidal or toroidal magnetic drifts; flux function potential}
+\subsection{DKES $\vect{E}\times\vect{B}$ trajectories; no poloidal or toroidal magnetic drifts; flux function potential}
 
@@
 
-\section{Full $\vect{E}\times\vect{B}$ trajectories; including poloidal and toroidal magnetic drifts; flux function potential}
+\subsection{Full $\vect{E}\times\vect{B}$ trajectories; including poloidal and toroidal magnetic drifts; flux function potential}
 
@@
 
-\section{Full $\vect{E}\times\vect{B}$ trajectories; no poloidal or toroidal magnetic drifts; leading $\Phi_1$ term}
+%%THE EQUATIONS IN THE FOLLOWING SUBSECTION WAS BASED ON J M García-Regaña et al 2013 Plasma Phys. Control. Fusion 55 074008
+%%HOWEVER, THESE EQUATIONS ARE NOT COMPLETELY CORRECT, AND CAN NO LONGER BE USED IN SFINCS
+%\subsection{Full $\vect{E}\times\vect{B}$ trajectories; no poloidal or toroidal magnetic drifts; leading $\Phi_1$ term}
+%
+%In this form of the drift-kinetic equation, we include the largest term involving $\Phi_1$,
+%associated with the adiabatic response.
+%However, other terms considered in Ref \cite{Regana2017} are not included.
+%
+%\begin{eqnarray}
+%&&\left(v_{||}\vect{b} + \frac{d\Phi_0}{dr} \frac{1}{B^2} \vect{B}\times\nabla r \right) \cdot \nabla f_{s1} \\
+%&&+ \left[ - \frac{(1-\xi^2)v}{2B} \nabla_{||} B
+%+\frac{(1-\xi^2)\xi}{2B^3} \frac{d\Phi_0}{dr}\vect{B}\times\nabla r\cdot\nabla B \right]
+% \frac{\partial f_{s1}}{\partial \xi} \nonumber \\
+%&&-(\vect{v}_{ms} \cdot\nabla r) \frac{Z_s e}{2 T_s x_s} \frac{d\Phi_0}{dr} \frac{\partial f_{s1}}{\partial x_s} \nonumber \\
+%&&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
+%&&+ (\vect{v}_{ms} \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM}
+% = C_s + S_s \nonumber
+%\end{eqnarray}
+%Note that this equation is equivalent to the following one, in which the independent variables
+%are $(\mu,v_{||})$ instead of $(\xi,x_s)$:
+%\begin{eqnarray}
+%&&\left(v_{||}\vect{b} + \frac{d\Phi_0}{dr} \frac{1}{B^2} \vect{B}\times\nabla r \right) \cdot (\nabla f_{s1})_{\mu, v_{||}} \\
+%&&+ \left[ - \mu \nabla_{||} B
+%-\frac{v_{||}}{B^2} \frac{d\Phi_0}{dr} \vect{b}\times\nabla B \times \nabla r \right]
+%\left( \frac{\partial f_{s1}}{\partial v_{||}} \right)_{\mu} \nonumber \\
+%&&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
+%&& + (\vect{v}_{ms} \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM}
+%=C_s + S_s \nonumber
+%\end{eqnarray}
+%These equivalent forms of the kinetic equation are selected using \\
+%\parlink{includeXDotTerm} = \true  \;\;\; (Default) \\
+%\parlink{includeElectricFieldTermInXiDot} = \true \;\;\; (Default) \\
+%\parlink{useDKESExBDrift} = \false \;\;\; (Default) \\
+%\parlink{magneticDriftScheme} = 0 \;\;\; (Default) \\
+%\parlink{includePhi1} = \true \;\;\; (Not default) %\\
+%%\parlink{includeRadialExBDrive} = \false \;\;\; (Default) \\
+%%\parlink{nonlinear} = \false \;\;\; (Default)
 
-In this form of the drift-kinetic equation, we include the largest term involving $\Phi_1$,
-associated with the adiabatic response.
-However, other terms considered in Ref \cite{Regana2017} are not included.
-\todo{This is not valid anymore}
+\subsection{Full $\vect{E}\times\vect{B}$ trajectories; no poloidal or toroidal magnetic drifts; Garc\'{i}a-Rega\~{n}a $\Phi_1$ terms}
 
-\begin{eqnarray}
-&&\left(v_{||}\vect{b} + \frac{d\Phi_0}{dr} \frac{1}{B^2} \vect{B}\times\nabla r \right) \cdot \nabla f_{s1} \\
-&&+ \left[ - \frac{(1-\xi^2)v}{2B} \nabla_{||} B
-+\frac{(1-\xi^2)\xi}{2B^3} \frac{d\Phi_0}{dr}\vect{B}\times\nabla r\cdot\nabla B \right]
- \frac{\partial f_{s1}}{\partial \xi} \nonumber \\
-&&-(\vect{v}_{ms} \cdot\nabla r) \frac{Z_s e}{2 T_s x_s} \frac{d\Phi_0}{dr} \frac{\partial f_{s1}}{\partial x_s} \nonumber \\
-&&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
-&&+ (\vect{v}_{ms} \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM}
- = C_s + S_s \nonumber
-\end{eqnarray}
-Note that this equation is equivalent to the following one, in which the independent variables
-are $(\mu,v_{||})$ instead of $(\xi,x_s)$:
-\begin{eqnarray}
-&&\left(v_{||}\vect{b} + \frac{d\Phi_0}{dr} \frac{1}{B^2} \vect{B}\times\nabla r \right) \cdot (\nabla f_{s1})_{\mu, v_{||}} \\
-&&+ \left[ - \mu \nabla_{||} B
--\frac{v_{||}}{B^2} \frac{d\Phi_0}{dr} \vect{b}\times\nabla B \times \nabla r \right]
-\left( \frac{\partial f_{s1}}{\partial v_{||}} \right)_{\mu} \nonumber \\
-&&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
-&& + (\vect{v}_{ms} \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM}
-=C_s + S_s \nonumber
-\end{eqnarray}
-These equivalent forms of the kinetic equation are selected using \\
-\parlink{includeXDotTerm} = \true  \;\;\; (Default) \\
-\parlink{includeElectricFieldTermInXiDot} = \true \;\;\; (Default) \\
-\parlink{useDKESExBDrift} = \false \;\;\; (Default) \\
-\parlink{magneticDriftScheme} = 0 \;\;\; (Default) \\
-\parlink{includePhi1} = \true \;\;\; (Not default) %\\
-%\parlink{includeRadialExBDrive} = \false \;\;\; (Default) \\
-%\parlink{nonlinear} = \false \;\;\; (Default)
+%This form of the drift-kinetic equation is nearly identical to equation (11) in Ref \cite{Regana2017}.
+%The one difference (which should be small) is that at the end of (11), Garc\'{i}a-Rega\~{n}a
+%has a term $\propto \vect{v}_m \cdot \nabla \Phi_1$, which is not (yet) in \sfincs.
 
-\section{Full $\vect{E}\times\vect{B}$ trajectories; no poloidal or toroidal magnetic drifts; Garc\'{i}a-Rega\~{n}a $\Phi_1$ terms}
+This form of the drift-kinetic equation is studied in Ref.~\cite{Mollen2018} (Eq.~(6)) and in Ref.~\cite{Regana2017} (Eq.~(16)).
 
-This form of the drift-kinetic equation is nearly identical to equation (11) in Ref \cite{Regana2017}.
-The one difference (which should be small) is that at the end of (11), Garc\'{i}a-Rega\~{n}a
-has a term $\propto \vect{v}_m \cdot \nabla \Phi_1$, which is not (yet) in \sfincs.
+\todo{In Eq. below check terms in left-hand-side multiplying $\frac{\partial f_{s1}}{\partial \xi}$ and $\frac{\partial f_{s1}}{\partial x_s}$. Define $v_s$?}\newline
 
@@
 \right] \frac{\partial f_{s1}}{\partial x_s} \nonumber \\
-&&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
-&& + (\vect{v}_E \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr}  + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM} \nonumber \\
-&&+ (\vect{v}_{ms} \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM}
+%&&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
+%&& + (\vect{v}_E \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr}  + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM} \nonumber \\
+&&+ \left(\vect{v}_{ms} + \vect{v}_E\right) \cdot \nabla r \left[ \frac{1}{n_{s 0}} \frac{dn_{s 0}}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2 - \frac{3}{2} + \frac{Z_s e}{T_s}\Phi_1\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{s0}
  = C_s + S_s \nonumber
 \end{eqnarray}
+Here $\displaystyle f_{s0} = f_{s0}\left(r,\theta,\zeta\right) = f_{sM}\left(r\right) \exp \left[- Z_s e \Phi_1(\theta,\zeta) / T_s \right]$, 
+whereas $n_{s 0} = n_{s 0} \left(r\right)$ represents the input density given by \parlink{nHats}. 
 Note that this equation is equivalent to the following one, in which the independent variables
@@
 - \mu \nabla_{||} B
--\frac{v_{||}}{B^2} \frac{d\Phi_0}{dr} \vect{b}\times\nabla B \times \nabla r \right]
+-\frac{v_{||}}{B^2} \frac{d\Phi_0}{dr} \vect{b}\times\nabla B \cdot \nabla r \right]
 \left( \frac{\partial f_{s1}}{\partial v_{||}} \right)_{\mu} \nonumber \\
-&&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
-&& + (\vect{v}_E \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr}  + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM} \nonumber \\
-&& + (\vect{v}_{ms} \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM}
+%&&+\frac{Z_s e}{T_s} f_{sM} v_{||} \nabla_{||} \Phi_1 \nonumber \\
+%&& + (\vect{v}_E \cdot \nabla r) \left[ \frac{1}{n_s} \frac{dn_s}{dr}  + \left(x_s^2-\frac{3}{2}\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{sM} \nonumber \\
+&& + \left(\vect{v}_{ms} + \vect{v}_E\right) \cdot \nabla r  \left[ \frac{1}{n_{s 0}} \frac{dn_{s 0}}{dr} + \frac{Z_s e}{T_s} \frac{d\Phi_0}{dr} + \left(x_s^2-\frac{3}{2} + \frac{Z_s e}{T_s}\Phi_1\right) \frac{1}{T_s} \frac{dT_s}{dr}\right] f_{s0}
 =C_s + S_s \nonumber
@@
 %\parlink{nonlinear} = \true \;\;\; (Not default)
+
+\section{Inductive electric field}
+\label{sec:Inductive-electric-field}
+Besides the terms described above, \sfincs~has the possibility to include an inductive parallel electric field in the drive part of the kinetic equation.
+There is typically no inductive electric field in a stellarator, but the term is retained since the transport driven by the inductive electric field is used for computing monoenergetic transport coefficients, and thus for comparison with other codes such as \dkes.
+This term has the form $\displaystyle \frac{Z_s e}{T_s} v_\| \frac{\left\langle \vect{E} \cdot \vect{B}\right\rangle B}{\left\langle B^2\right\rangle} f_{sM}$ and is set by \parlink{EParallelHat} (i.e.
+if non-zero than this term is added to the right-hand-side of all kinetic equations in Sec.~\ref{sec:DKequations}).
+Also see the input parameter \parlink{EParallelHatSpec}.

### 2019-03-05 doc/manual/version3/inputParameters.tex
Substantial revision to manual

--- 2018-12-14 doc/manual/version3/inputParameters.tex
@@
 \newcommand{\param}[5]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries \hypertarget{#1}{#1}}\\{\it Type}: #2\\{\it Default}: #3\\{\it When it matters}: #4\\{\it Meaning}: #5}}
+%%NOTE BY AM, the \param command does not work for defining parameters with an underscore in the name, for these parameters use \paramWithUnderscore.
+\newcommand{\paramWithUnderscore}[6]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries \hypertarget{#2}{#1}}\\{\it Type}: #3\\{\it Default}: #4\\{\it When it matters}: #5\\{\it Meaning}: #6}}
 \newcommand{\ssparam}[4]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries \hypertarget{#1}{#1}}\\{\it Type}: #2\\{\it When it matters}: #3\\{\it Meaning}: #4}}
+\newcommand{\ssparamWithUnderscore}[5]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries \hypertarget{#2}{#1}}\\{\it Type}: #3\\{\it When it matters}: #4\\{\it Meaning}: #5}}
 \newcommand{\PETScParam}[2]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries #1}\\{\it Meaning}: #2}}
@@
 
-{\ttfamily RHSMode} = 2: Solve for 3 right-hand sides to get the 3$\times$3 transport matrix. Presently implemented only for 1 species.\\
-
-{\ttfamily RHSMode} = 3: Solve for the 2$\times$2 monoenergetic transport coefficients. When this option is chosen, \Nx~is set to 1 and only 1 species is used. \\
-
-{\ttfamily RHSMode} = 4: Solve the adjoint equation to compute derivatives at constant $E_r$.
-Options related to the adjoint solve are detailed in section \ref{sec:adjointOptions}.
-{\ttfamily RHSMode} = 4 must be used with \parlink{includePhi1}=\false, \\ \parlink{EParallelHat}\texttt{=0}, \parlink{magneticDriftScheme}\texttt{>0}, \parlink{constraintScheme}\texttt{=-1} or 1, and \\ \parlink{collisionOperator}\texttt{=0}.
-\\
-
-{\ttfamily RHSMode} = 5: Solve the adjoint equation to compute derivatives at the ambipolar $E_r$.
-Options related to the adjoint solve are detailed in section \ref{sec:adjointOptions}.
-{\ttfamily RHSMode} = 5 must be used with \parlink{includePhi1}=\false, \parlink{EParallelHat}\texttt{=0}, \parlink{magneticDriftScheme}\texttt{>0}, \parlink{constraintScheme}\texttt{=-1} or 1, and \\ \parlink{collisionOperator}\texttt{=0}.\\
-
+{\ttfamily RHSMode} = 2: Solve for 3 right-hand sides to get the 3$\times$3 transport matrix. Presently implemented only for 1 species and \parlink{includePhi1} == \false\\
+
+{\ttfamily RHSMode} = 3: Solve for the 2$\times$2 monoenergetic transport coefficients. When this option is chosen, \Nx~is set to 1, only 1 species is used and \parlink{includePhi1} is set to \false
 }
@@
 
-\myhrule
-
-\param{ambipolarSolve}
-{Boolean}
-{\false}
-{When \parlink{RHSMode}==1, 4, or 5.}
-{When \true, a root finding method will be used to find the ambipolar \parlink{Er}.
-All output parameters will be included at the ambipolar \parlink{Er}.
-Options related to this parameter are \parlink{NEr\_ambipolarSolve}, \parlink{Er\_search\_tolerance\_dx}, \parlink{Er\_search\_tolerance\_f}, \parlink{ambipolarSolveOption}, \parlink{Er\_min}, and \parlink{Er\_max}.}
-
-\myhrule
-
-\param{NEr\_ambipolarSolve}
-{integer}
-{20}
-{When \parlink{ambipolarSolve} == \true.}
-{Maximum number of solves to allow while finding the ambipolar \parlink{Er}.}
-
-\myhrule
-
-\param{Er\_search\_tolerance\_dx}
-{real}
-{1.d-8}
-{When \parlink{ambipolarSolve} == \true and \parlink{ambipolarSolveOption}/=2.}
-{Tolerance used for ambipolar solve.
-If step size in \parlink{Er} falls below \parlink{Er\_search\_tolerance\_dx}, the ambipolar solve completes successfully.
-Note that this is not used for \parlink{ambipolarSolveOption}==2.}
-
-\myhrule
-
-\param{Er\_search\_tolerance\_f}
-{real}
-{1.d-10}
-{When \parlink{ambipolarSolve} == \true.}
-{Tolerance used for ambipolar solve. If the radial current,
-\begin{gather}
-J_r = \sum_s \texttt{Zs}(s) \texttt{particleFlux\_vm\_rN}(s),
-\end{gather}
-falls below \parlink{Er\_search\_tolerance\_f}, the ambipolar solve completes successfully.}
-
-\myhrule
-
-\param{ambipolarSolveOption}
-{integer}
-{1}
-{When \parlink{ambipolarSolve} == \true.}
-{Indicates which root solving algorithm to use for ambipolar solve.
-If \parlink{ambipolarSolveOption}==1 a Newton bisection method is used, \parlink{ambipolarSolveOption}==2 a Brent method is used, and if \parlink{ambipolarSolveOption}==3 a Newton method is used.
-If \parlink{ambipolarSolveOption}==1 or 3, an adjoint method will used to compute the $\partial J_r/\partial E_r$, so must be used with \parlink{includePhi1}==\false, \parlink{EParallelHat}==0, \parlink{magneticDriftScheme}>0, \parlink{constraintScheme}==-1 or 1, and \parlink{collisionOperator}==0.}
-
-\myhrule
-
-\param{Er\_min}
-{real}
-{-100}
-{When \parlink{ambipolarSolve} == \true and \parlink{ambipolarSolveOption} /= 3. }
-{Minimum value of \parlink{Er} used to bracket the ambipolar root. }
-
-\myhrule
-
-\param{Er\_max}
-{real}
-{100}
-{When \parlink{ambipolarSolve} == \true and \parlink{ambipolarSolveOption} /= 3. }
-{Maximum  value of \parlink{Er} used to bracket the ambipolar root. }
-
-\myhrule
-
 \section{The {\ttfamily geometryParameters} namelist}
@@
 B(\theta,\zeta)/\bar{B} 
-&=& (\mbox{\parlink{B0OverBBar}}) [1 + (\mbox{\parlink{epsilon\_t}}) \cos(\theta) \\
-&& + (\mbox{\parlink{epsilon\_h}}) \cos((\mbox{\parlink{helicity\_l}}) \theta - (\mbox{\parlink{helicity\_n}}) \zeta) \nonumber \\
-&&  + (\mbox{\parlink{epsilon\_antisymm}}) \nonumber \\
-&& \times \sin((\mbox{\parlink{helicity\_antisymm\_l}}) \theta - (\mbox{\parlink{helicity\_antisymm\_n}}) \zeta)] \nonumber
+&=& (\mbox{\parlink{B0OverBBar}}) [1 + (\mbox{\underscoreparlink{epsilon\_t}{epsilon_t}}) \cos(\theta) \\
+&& + (\mbox{\underscoreparlink{epsilon\_h}{epsilon_h}}) \cos((\mbox{\underscoreparlink{helicity\_l}{helicity_l}}) \theta - (\mbox{\underscoreparlink{helicity\_n}{helicity_n}}) \zeta) \nonumber \\
+&&  + (\mbox{\underscoreparlink{epsilon\_antisymm}{epsilon_antisymm}}) \nonumber \\
+&& \times \sin((\mbox{\underscoreparlink{helicity\_antisymm\_l}{helicity_antisymm_l}}) \theta - (\mbox{\underscoreparlink{helicity\_antisymm\_n}{helicity_antisymm_n}}) \zeta)] \nonumber
 \end{eqnarray}
@@
 The file is assumed to be stellarator-\emph{asymmetric}.
-
-{\ttfamily geometryScheme}==13: Read Boozer geometry from the {\ttfamily geometryParamters} namelist.
-Here the geometry parameters are specified by \parlink{boozer\_bmnc}, \parlink{GHat}, \parlink{IHat}, and \parlink{iota}.
-This has been implemented for use with STELLOPT for direct optimization of the Boozer spectrum.
 }
@@
 
-{\ttfamily inputRadialCoordinate}==0: Use the flux surface specified by \parlink{psiHat\_wish}
-for a single run, and use the range specified by \parlink{psiHat\_min} and \parlink{psiHat\_max} for radial scans.\\
-
-{\ttfamily inputRadialCoordinate}==1: Use the flux surface specified by \parlink{psiN\_wish}
-for a single run, and use the range specified by \parlink{psiN\_min} and \parlink{psiN\_max} for radial scans.\\
-
-{\ttfamily inputRadialCoordinate}==2: Use the flux surface specified by \parlink{rHat\_wish}
-for a single run, and use the range specified by \parlink{rHat\_min} and \parlink{rHat\_max} for radial scans.\\
-
-{\ttfamily inputRadialCoordinate}==3: Use the flux surface specified by \parlink{rN\_wish}
-for a single run, and use the range specified by \parlink{rN\_min} and \parlink{rN\_max} for radial scans. \\
+{\ttfamily inputRadialCoordinate}==0: Use the flux surface specified by \underscoreparlink{psiHat\_wish}{psiHat_wish}
+for a single run, and use the range specified by \underscoreparlink{psiHat\_min}{psiHat_min} and \underscoreparlink{psiHat\_max}{psiHat_max} for radial scans.\\
+
+{\ttfamily inputRadialCoordinate}==1: Use the flux surface specified by \underscoreparlink{psiN\_wish}{psiN_wish}
+for a single run, and use the range specified by \underscoreparlink{psiN\_min}{psiN_min} and \underscoreparlink{psiN\_max}{psiN_max} for radial scans.\\
+
+{\ttfamily inputRadialCoordinate}==2: Use the flux surface specified by \underscoreparlink{rHat\_wish}{rHat_wish}
+for a single run, and use the range specified by \underscoreparlink{rHat\_min}{rHat_min} and \underscoreparlink{rHat\_max}{rHat_max} for radial scans.\\
+
+{\ttfamily inputRadialCoordinate}==3: Use the flux surface specified by \underscoreparlink{rN\_wish}{rN_wish}
+for a single run, and use the range specified by \underscoreparlink{rN\_min}{rN_min} and \underscoreparlink{rN\_max}{rN_max} for radial scans. \\
 
@@
 
-\param{psiHat\_wish}
+\paramWithUnderscore{psiHat\_wish}{psiHat_wish}
 {real}
@@
 
-\param{psiN\_wish}
+\paramWithUnderscore{psiN\_wish}{psiN_wish}
 {real}
@@
 
-\param{rHat\_wish}
+\paramWithUnderscore{rHat\_wish}{rHat_wish}
 {real}
@@
 
-\param{rN\_wish}
+\paramWithUnderscore{rN\_wish}{rN_wish}
 {real}
@@
 {3.7481}
-{Only when \parlink{geometryScheme} == 1 or 13. Otherwise, {\ttfamily GHat} will be set according to the requested \parlink{geometryScheme}.}
+{Only when \parlink{geometryScheme} == 1. Otherwise, {\ttfamily GHat} will be set according to the requested \parlink{geometryScheme}.}
 {$G$ is $(c/2)\times$ the poloidal current outside the flux
@@
 {0.0}
-{Only when \parlink{geometryScheme} == 1 or 13. Otherwise, {\ttfamily IHat} will be set according to the requested \parlink{geometryScheme}.}
+{Only when \parlink{geometryScheme} == 1. Otherwise, {\ttfamily IHat} will be set according to the requested \parlink{geometryScheme}.}
 {$I$ is $(c/2)\times$ the toroidal current inside the flux
@@
 {0.4542}
-{Only when \parlink{geometryScheme} == 1 or 13. Otherwise, {\ttfamily iota} will be set according to the requested \parlink{geometryScheme}.}
+{Only when \parlink{geometryScheme} == 1. Otherwise, {\ttfamily iota} will be set according to the requested \parlink{geometryScheme}.}
 {Rotational transform (rationalized), equivalent to $1/q$ where $q$ is the safety factor.}
@@
 
-\param{epsilon\_t}
+\paramWithUnderscore{epsilon\_t}{epsilon_t}
 {real}
@@
 
-\param{epsilon\_h}
+\paramWithUnderscore{epsilon\_h}{epsilon_h}
 {real}
@@
 
-\param{epsilon\_antisymm}
+\paramWithUnderscore{epsilon\_antisymm}{epsilon_antisymm}
 {real}
@@
 
-\param{helicity\_l}
+\paramWithUnderscore{helicity\_l}{helicity_l}
 {integer}
 {2}
-{Only when \parlink{geometryScheme} == 1.}
+{Only when \parlink{geometryScheme} == 1, or when simultaneously \parlink{geometryScheme} == 5 and \parlink{rippleScale} != 1.0.}
 {Poloidal mode number of the helical variation in $B$, as defined by (\ref{eq:Bmodel}).}
@@
 
-\param{helicity\_n}
+\paramWithUnderscore{helicity\_n}{helicity_n}
 {integer}
 {10}
-{Only when \parlink{geometryScheme} == 1.}
+{Only when \parlink{geometryScheme} == 1, or when simultaneously \parlink{geometryScheme} == 5 and \parlink{rippleScale} != 1.0.}
 {Toroidal mode number of the helical variation in $B$, as defined by (\ref{eq:Bmodel}).}
@@
 
-\param{helicity\_antisymm\_l}
+\paramWithUnderscore{helicity\_antisymm\_l}{helicity_antisymm_l}
 {integer}
@@
 
-\param{helicity\_antisymm\_n}
+\paramWithUnderscore{helicity\_antisymm\_n}{helicity_antisymm_n}
 {integer}
@@
 {Toroidal mode number of the stellarator-antisymmetric variation in $B$, as defined by (\ref{eq:Bmodel}).
-Note that you can create an up-down asymmetric tokamak by setting \parlink{helicity\_antisymm\_n}=0, \parlink{epsilon\_h}=0,
-and \parlink{epsilon\_antisymm}$>$0.}
+Note that you can create an up-down asymmetric tokamak by setting \underscoreparlink{helicity\_antisymm\_n}{helicity_antisymm_n}=0, \underscoreparlink{epsilon\_h}{epsilon_h}=0,
+and \underscoreparlink{epsilon\_antisymm}{epsilon_antisymm}$>$0.}
 
@@
 
-\param{VMEC\_Nyquist\_option}
+\paramWithUnderscore{VMEC\_Nyquist\_option}{VMEC_Nyquist_option}
 {integer}
@@
 arrays but not in the {\ttfamily xm} and {\ttfamily xn} arrays.
-This this parameter only affects high-frequency components of the magnetic field, results of sfincs
+This parameter only affects high-frequency components of the magnetic field, results of sfincs
 should not be very sensitive to this parameter.
@@
 
-\param{min\_Bmn\_to\_load}
+\paramWithUnderscore{min\_Bmn\_to\_load}{min_Bmn_to_load}
 {real}
@@
 {1.0}
-{Only when \parlink{geometryScheme}==5. \parlink{helicity\_l} and \parlink{helicity\_n} must be specified.}
+{Only when \parlink{geometryScheme} == 5. \underscoreparlink{helicity\_l}{helicity_l} and \underscoreparlink{helicity\_n}{helicity_n} must be specified.}
 {Scales the values of components of VMEC geometry (modes of $B$, $B^{u}$, $B_{u}$, $B^{v}$, $B_{v}$, $B_{s}$, and $\sqrt{g}$) that do not satisfy a desired symmetry.
-If \parlink{helicity\_n}$==0$, modes for which $n \neq 0$ or $m \neq$ \parlink{helicity\_l} are multiplied by \parlink{rippleScale}.
-If \parlink{helicity\_n}$\neq0$, modes for which $m/n \neq$ \parlink{helicity\_l}/\parlink{helicity\_n} or $n==0$ are multiplied by \parlink{rippleScale}.
+If \underscoreparlink{helicity\_n}{helicity_n}$==0$, modes for which $n \neq 0$ or $m \neq$ \underscoreparlink{helicity\_l}{helicity_l} are multiplied by \parlink{rippleScale}.
+If \underscoreparlink{helicity\_n}{helicity_n}$\neq0$, modes for which $m/n \neq$ \underscoreparlink{helicity\_l}{helicity_l}/\underscoreparlink{helicity\_n}{helicity_n} or $n==0$ are multiplied by \parlink{rippleScale}.
 All other modes remain unchanged.}
@@
 
-\param{EParallelHatSpec\_bcdatFile}
+\paramWithUnderscore{EParallelHatSpec\_bcdatFile}{EParallelHatSpec_bcdatFile}
 {string}
@@
 {Only when \parlink{EParallelHatSpec} is non-zero
-  and \parlink{RHSmode}==1 \\and \parlink{EParallelHatSpec\_bcdatFile}
+  and \parlink{RHSmode}==1 \\and \underscoreparlink{EParallelHatSpec\_bcdatFile}{EParallelHatSpec_bcdatFile}
   is not == ``''}
@@
   (otherwise just proportional to $B$).}
-
-\myhrule
-
-\param{boozer\_bmnc}
-{2D array of reals}
-{0}
-{Only when \parlink{geometryScheme} == 13. }
-{Fourier harmonics describing the magnetic field in Boozer coordinates,
-\begin{gather}
-B(\theta_B,\zeta_B) = \sum_{m,n}\mathrm{\texttt{boozer\_bmnc}}(m,n) \cos(m\theta_B-n\mathrm{\texttt{Nperiods}} \zeta_B),
-\end{gather}
-The first index of \texttt{ Boozer\_bmnc} corresponds to $m \in $[0,\parlink{mmax\_boozer}], and the second corresponds to $n \in $[-\parlink{nmax\_boozer},\parlink{nmax\_boozer}].
-Here \parlink{Nperiods} can be specified in the input file.
-}
-
-\myhrule
-
-\param{mmax\_boozer}
-{integer}
-{40}
-{Only when \parlink{geometryScheme} == 13.}
-{Boozer geometry is described for $m \leq $ \texttt{mmax\_boozer}. }
-
-\myhrule
-
-\param{nmax\_boozer}
-{integer}
-{40}
-{Only when \parlink{geometryScheme} == 13.}
-{Boozer geometry is described for $|n| \leq $ \texttt{nmax\_boozer}. }
-
-\myhrule
-
-\param{Nperiods}
-{integer}
-{0}
-{Only when \parlink{geometryScheme} == 13.}
-{Boozer geometry will assume number of periods is \texttt{Nperiods}.}
 
@@
 {\false}
-{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and\\ \parlink{readExternalPhi1} == \false}
 {If this parameter is \true, an adiabatic species is added into the quasineutrality equation (but has no other effect).}
@@
 {-1.0}
-{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and \parlink{withAdiabatic} == \true}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and\\ \parlink{readExternalPhi1} == \false and \parlink{withAdiabatic} == \true}
 {Charge of adiabatic species, in units of the proton charge $e$.}
@@
 {5.44617e-4}
-{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and \parlink{withAdiabatic} == \true}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and\\ \parlink{readExternalPhi1} == \false and \parlink{withAdiabatic} == \true}
 {Mass of adiabatic species, in units of the reference mass $\bar{m}$.}
@@
 {1.0}
-{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and \parlink{withAdiabatic} == \true}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and\\ \parlink{readExternalPhi1} == \false and \parlink{withAdiabatic} == \true}
 {Density of adiabatic species, in units of the reference density $\bar{n}$.}
@@
 {1.0}
-{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and \parlink{withAdiabatic} == \true}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and\\ \parlink{readExternalPhi1} == \false and \parlink{withAdiabatic} == \true}
 {Temperature of adiabatic species, in units of the reference temperature $\bar{T}$.}
@@
 \param{withNBIspec} {Boolean} {\false} {Whenever \parlink{RHSMode} ==
-  1 and \parlink{includePhi1} == \true\\
-  ~and \parlink{quasineutralityOption} == 1} {When this is \true, an
+  1 and \parlink{includePhi1} == \true and\\
+  \parlink{readExternalPhi1} == \false and \parlink{quasineutralityOption} == 1} {When this is \true, an
   extra species (here called NBIspec) is added to the quasineutrality
@@
 {1.0}
-{Whenever \parlink{withNBIspec}== \true~ and \parlink{RHSMode} == 1 \\
-  and \parlink{includePhi1} == \true~ and \parlink{quasineutralityOption} == 1}
+{Whenever \parlink{withNBIspec}== \true and \parlink{RHSMode} == 1 and \parlink{includePhi1}\\
+   == \true and \parlink{readExternalPhi1} == \false and \parlink{quasineutralityOption} == 1}
 {Charge of the fast ion species, in units of the proton charge $e$, see \parlink{withNBIspec}.}
@@
 {0.0}
-{Whenever \parlink{withNBIspec}== \true~ and \parlink{RHSMode} == 1 \\
-and \parlink{includePhi1} == \true~ and \parlink{quasineutralityOption} == 1}
+{Whenever \parlink{withNBIspec}== \true and \parlink{RHSMode} == 1 and \parlink{includePhi1} \\
+== \true and \parlink{readExternalPhi1} == \false and \parlink{quasineutralityOption} == 1}
 {Density of the fast ion species, in units of the reference density
@@
 
-\param{nu\_n}
+%\param{nu\_n}
+\paramWithUnderscore{nu\_n}
+{nu_n}
 {real}
@@
 $\bar{m}$ = proton mass, $\bar{n}$ = $10^{20}$ m$^{-3}$, $\bar{T}$ = 1 keV, and $\ln\Lambda = 17$.
+If $\text{\ttfamily nu\_n} < 0$ in {\ttfamily input.namelist} then {\ttfamily nu\_n} will be calculated automatically using the density and temperature of the first species with the assumption that it is electrons, and assuming $\bar{R}$ = 1 meter, $\bar{m}$ = proton mass, $\bar{n}$ = $10^{20}$ m$^{-3}$, $\bar{T}$ = 1 keV.
 }
@@
 {Only when \parlink{RHSMode} == 3.}
-{Dimensionless collisionality used in place of \parlink{nHats}, \parlink{THats}, \parlink{mHats}, \parlink{Zs}, and \parlink{nu\_n} for computing monoenergetic transport coefficients.
+{Dimensionless collisionality used in place of \parlink{nHats}, \parlink{THats}, \parlink{mHats}, \parlink{Zs}, and \underscoreparlink{nu\_n}{nu_n} for computing monoenergetic transport coefficients.
 See section \ref{sec:monoenergetic} for more details.}
@@
 {\ttfamily collisionOperator} = 1: Pitch-angle scattering operator (with no momentum-conserving field term).
-}}
+}\newline
+\newline
+See section~\ref{sec:CollisionOperator} for more details about the collision operator.}
 
@@
 
-\param{include\_fDivVE\_term}
+\paramWithUnderscore{include\_fDivVE\_term}{include_fDivVE_term}
 {Boolean}
@@
 {If false, no terms involving $\Phi_1 = \Phi-\left<\Phi\right>$ are included in the kinetic equation,
-and the quasineutrality equation is not solved.  If true, then terms involving $\Phi_1$ are included
-in the kinetic equation if \parlink{includePhi1inKineticEquation} = \true, and the quasineutrality equation is solved at each point on the flux surface (make sure that your input densities fulfill quasineutrality, see \parlink{quasineutralityOption}).
-In this latter case, many more quantities are computed and saved in the output file, such as radial fluxes
-associated with the radial $\vect{E}\times\vect{B}$ drift.\newline
-%%Text added by AM 2016-07
-If true, the system becomes nonlinear in the unknowns $f_{s1}$ and $\Phi_1$. 
+and the quasineutrality equation is not solved.
+
+If true, and \parlink{readExternalPhi1} == \false, then terms involving $\Phi_1$ are included
+in the kinetic equation if \parlink{includePhi1InKineticEquation} = \true, and the quasineutrality equation is solved at each point on the flux surface 
+(make sure that your input densities fulfill quasineutrality, see \parlink{quasineutralityOption}) together with the condition $\left<\Phi_1\right> = 0$. 
+In this latter case, the system becomes nonlinear in the unknowns $f_{s1}$ and $\Phi_1$. 
 Newton's method will be used to solve the nonlinear system, meaning that the usual linear solve in \sfincs~must be iterated several times.
-}
-
-\myhrule
-
-\param{includePhi1inKineticEquation}
+
+If true, and \parlink{readExternalPhi1} == \true, then $\Phi_1$ is read from an input file (given by \parlink{externalPhi1Filename}), 
+the quasineutrality equation and the condition $\left<\Phi_1\right> = 0$ are ignored and the system is linear.\newline
+
+If true, many more quantities are computed and saved in the output file, such as radial fluxes
+associated with the radial $\vect{E}\times\vect{B}$ drift caused by $\Phi_1$ (see section~\ref{sec:RadialFluxesAndFlows}).
+More details on runs including $\Phi_1$ are found in section~\ref{sec:qn}.
+}
+
+\myhrule
+
+\param{readExternalPhi1}
+{Boolean}
+{\false}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true}
+{If true, \parlink{Phi1Hat} is read from the HDF5 file specified by \parlink{externalPhi1Filename}, 
+the quasineutrality equation and the condition $\left<\Phi_1\right> = 0$ are ignored and the solved system is linear.
+If true, and \parlink{includePhi1InKineticEquation} = \false, then $\Phi_1$ does not appear anywhere in the system of equations, 
+and $f_{s1}$ is unaffected by $\Phi_1$. 
+However, the fluxes are still affected by $\Phi_1$ (see section~\ref{sec:RadialFluxesAndFlows}).
+}
+
+\myhrule
+
+\param{externalPhi1Filename}
+{string}
+{``externalPhi1.h5''}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and\\ \parlink{readExternalPhi1} == \true}
+{Name of the HDF5 file from which \parlink{Phi1Hat} is read. The HDF5 file must contain the following fields specified in the same format as in the output file (\parlink{outputFileName}):
+\parlink{Ntheta}, \parlink{Nzeta}, \parlink{NIterations}, \parlink{NPeriods}, \parlink{theta}, \parlink{zeta}, \parlink{Phi1Hat}. 
+It is assumed that \parlink{theta} and \parlink{zeta} represent a 2D uniform grid on which \parlink{Phi1Hat} is given, but this grid can be different from the grid used in the calculation (i.e.
+\parlink{Ntheta}, \parlink{Nzeta} in the external HDF5 file do not have to be the same as specified in {\ttfamily input.namelist}).
+Linear interpolation is used to determine \parlink{Phi1Hat} in the calculation.}
+
+\myhrule
+
+\param{includePhi1InKineticEquation}
 {Boolean}
@@
 {If true, the terms containing $\Phi_1$ will be included in the kinetic equation. 
-If false, and \parlink{includePhi1} = \true, then $\Phi_1$ is only included in the quasineutrality equation. 
+If false, \parlink{includePhi1} == \true and \parlink{readExternalPhi1} == \false, then $\Phi_1$ is only included in the quasineutrality equation. 
 }
@@
 {\false}
-{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and\\ \parlink{includePhi1inKineticEquation} == \true}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and\\ \parlink{includePhi1InKineticEquation} == \true}
 {If true, the terms containing $\Phi_1$ will be included in the collision operator. 
-If false or \parlink{includePhi1inKineticEquation} == \false, and \parlink{includePhi1} = \true, then $\Phi_1$ is not included in the collision operator. 
-The default is false since it remains to show if this effect is important, and the memory and time requirements can increase significantly with this parameter.
+If false (or \parlink{includePhi1InKineticEquation} == \false) %, or \parlink{includePhi1} == \false, 
+then $\Phi_1$ is not included in the collision operator. 
+The default is false since this setting has yet only had a minor impact on the results, and the memory and time requirements can increase significantly when it is true.
+\newline \warning{Setting this to true is not yet compatible with \parlink{readExternalPhi1} == \true}
 }
@@
 {1}
-{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true}
+{Whenever \parlink{RHSMode} == 1 and \parlink{includePhi1} == \true and\\ \parlink{readExternalPhi1} == \false}
 {This variable controls which quasineutrality equation is used, and must be set to either 1 or 2. 
-Make sure that your input densities fulfill quasineutrality (also accounting for the adiabatic species if \parlink{withAdiabatic}=\true).\\
+If including $\Phi_1$ and not reading it from an external file (\parlink{readExternalPhi1} == \true), then make sure that your input densities fulfill quasineutrality (also accounting for the adiabatic species if \parlink{withAdiabatic}=\true).\\
 
@@
 {Whenever \parlink{RHSMode} == 1.}
-{When true, the term $C_{ab}[ f_{Ma}, f_{Mb}]$ is included in the kinetic equation,
+{When true, the term $C_{ab}[ f_{aM}, f_{bM}]$ is included in the kinetic equation,
 i.e. collisions between the leading-order Maxwellians of different species.  This term is nonzero
@@
 {Whenever \parlink{RHSMode} == 1.}
-{This variable controls the poloidal and magnetic drifts, and does not affect the radial magnetic drift.\\
+{This variable controls the poloidal and toroidal magnetic drifts, and does not affect the radial magnetic drift.\\
+\warning{Some terms involving both $\Phi_1$ and the magnetic drifts have not yet been implemented, 
+and therefore using another setting than 0 is incompatible with \parlink{includePhi1} == \true. 
+However, for testing purposes it is still allowed to use a non-zero setting together with \parlink{includePhi1} == \true, 
+but the results may not be physical.}\\
 
@@
 \parlink{EParallelHat} in the code. The electric field strength is
-proportional to $B$ unless explicitly specified using an input file \parlink{EParallelHatSpec\_bcdatFile}.}
+proportional to $B$ unless explicitly specified using an input file \underscoreparlink{EParallelHatSpec\_bcdatFile}{EParallelHatSpec_bcdatFile}.}
 
@@
 
-\param{xGrid\_k}
+\paramWithUnderscore{xGrid\_k}{xGrid_k}
 {integer}
@@
 
-\param{Nxi\_for\_x\_option}
+\paramWithUnderscore{Nxi\_for\_x\_option}{Nxi_for_x_option}
 {integer}
@@
 
-\param{preconditioner\_species}
+\paramWithUnderscore{preconditioner\_species}{preconditioner_species}
 {integer}
@@
 
-\param{preconditioner\_x}
+\paramWithUnderscore{preconditioner\_x}{preconditioner_x}
 {integer}
@@
 
-\param{preconditioner\_x\_min\_L}
+\paramWithUnderscore{preconditioner\_x\_min\_L}{preconditioner_x_min_L}
 {integer}
 {0}
-{Whenever \parlink{useIterativeLinearSolver} = \true~and \parlink{RHSMode} = 1 or 2 and \parlink{preconditioner\_x} $>$ 0.}
+{Whenever \parlink{useIterativeLinearSolver} = \true~and \parlink{RHSMode} = 1 or 2 and \underscoreparlink{preconditioner\_x}{preconditioner_x} $>$ 0.}
 {The $x$ structure of the matrix will only be simplified for Legendre index $L$ is $\ge$ this value.
@@
 
-\param{preconditioner\_theta}
+\paramWithUnderscore{preconditioner\_theta}{preconditioner_theta}
 {integer}
@@
 
-\param{preconditioner\_theta\_min\_L}
+\paramWithUnderscore{preconditioner\_theta\_min\_L}{preconditioner_theta_min_L}
 {integer}
 {0}
-{Whenever \parlink{useIterativeLinearSolver} = \true~and \parlink{preconditioner\_theta} $>$ 0.}
+{Whenever \parlink{useIterativeLinearSolver} = \true~and \underscoreparlink{preconditioner\_theta}{preconditioner_theta} $>$ 0.}
 {The $\theta$ structure of the matrix will only be simplified for Legendre index $L$ is $\ge$ this value.
@@
 
-\param{preconditioner\_zeta}
+\paramWithUnderscore{preconditioner\_zeta}{preconditioner_zeta}
 {integer}
@@
 
-\param{preconditioner\_zeta\_min\_L}
+\paramWithUnderscore{preconditioner\_zeta\_min\_L}{preconditioner_zeta_min_L}
 {integer}
 {0}
-{Whenever \parlink{useIterativeLinearSolver} = \true~and \parlink{preconditioner\_zeta} $>$ 0.}
+{Whenever \parlink{useIterativeLinearSolver} = \true~and \underscoreparlink{preconditioner\_zeta}{preconditioner_zeta} $>$ 0.}
 {The $\zeta$ structure of the matrix will only be simplified for Legendre index $L$ is $\ge$ this value.
@@
 
-\param{preconditioner\_xi}
+\paramWithUnderscore{preconditioner\_xi}{preconditioner_xi}
 {integer}
@@
 
-\param{preconditioner\_magnetic\_drifts\_max\_L}
+\paramWithUnderscore{preconditioner\_magnetic\_drifts\_max\_L}{preconditioner_magnetic_drifts_max_L}
 {integer}
@@
 {\true}
-{Only when \parlink{includePhi1} = \true}
+{Only when \parlink{includePhi1} = \true  and \parlink{readExternalPhi1} == \false}
 {If true, the nonlinear term will not be included in the preconditioner matrix, meaning the preconditioner matrix is the same
@@
 
-\param{export\_full\_f}
+\paramWithUnderscore{export\_full\_f}{export_full_f}
 {Boolean}
@@
 
-\param{export\_delta\_f}
+\paramWithUnderscore{export\_delta\_f}{export_delta_f}
 {Boolean}
@@
 
-\param{export\_f\_theta\_option}
+\paramWithUnderscore{export\_f\_theta\_option}{export_f_theta_option}
 {integer}
 {2}
-{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true}
+{Whenever \underscoreparlink{export\_full\_f}{export_full_f} or \underscoreparlink{export\_delta\_f}{export_delta_f} is \true}
 {Controls which grid in $\theta$ is used for exporting the distribution function.\\
@@
 
-{\ttfamily export\_f\_theta\_option} = 1: Interpolate to a different grid, specified by \parlink{export\_f\_theta}. Linear interpolation will be used.
+{\ttfamily export\_f\_theta\_option} = 1: Interpolate to a different grid, specified by \underscoreparlink{export\_f\_theta}{export_f_theta}. Linear interpolation will be used.
 No sorting of the requested values is performed.\\
@@
 values of the $\theta$ grid that are closest to the values requested
-in \parlink{export\_f\_theta}.  Values of $\theta$ will be in
+in \underscoreparlink{export\_f\_theta}{export_f_theta}.  Values of $\theta$ will be in
 increasing order. If multiple requested values are close to the same
@@
 
-For all of these options, you can see \parlink{export\_f\_theta} in {\ttfamily sfincsOutput.h5} for the actual grid used in the end.
-}
-
-\myhrule
-
-\param{export\_f\_zeta\_option}
+For all of these options, you can see \underscoreparlink{export\_f\_theta}{export_f_theta} in {\ttfamily sfincsOutput.h5} for the actual grid used in the end.
+}
+
+\myhrule
+
+\paramWithUnderscore{export\_f\_zeta\_option}{export_f_zeta_option}
 {integer}
 {2}
-{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true}
+{Whenever \underscoreparlink{export\_full\_f}{export_full_f} or \underscoreparlink{export\_delta\_f}{export_delta_f} is \true}
 {Controls which grid in $\zeta$ is used for exporting the distribution function.\\
@@
 
-{\ttfamily export\_f\_zeta\_option} = 1: Interpolate to a different grid, specified by \parlink{export\_f\_zeta}. Linear interpolation will be used.
+{\ttfamily export\_f\_zeta\_option} = 1: Interpolate to a different grid, specified by \underscoreparlink{export\_f\_zeta}{export_f_zeta}. Linear interpolation will be used.
 No sorting of the requested values is performed.\\
@@
 values of the $\zeta$ grid that are closest to the values requested
-in \parlink{export\_f\_zeta}.  Values of $\zeta$ will be in
+in \underscoreparlink{export\_f\_zeta}{export_f_zeta}.  Values of $\zeta$ will be in
 increasing order. If multiple requested values are close to the same
@@
 
-For all of these options, you can see \parlink{export\_f\_zeta} in {\ttfamily sfincsOutput.h5} for the actual grid used in the end.
-}
-
-\myhrule
-
-\param{export\_f\_theta}
+For all of these options, you can see \underscoreparlink{export\_f\_zeta}{export_f_zeta} in {\ttfamily sfincsOutput.h5} for the actual grid used in the end.
+}
+
+\myhrule
+
+\paramWithUnderscore{export\_f\_theta}{export_f_theta}
 {1D array of reals}
 {0.0}
-{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true, and \parlink{export\_f\_theta\_option} $>0$.}
-{Values of $\theta$ on which you want to save the distribution function.  modulo$(\ldots, 2\pi)$ will be applied.  See \parlink{export\_f\_theta\_option} for details}
-
-\myhrule
-
-\param{export\_f\_zeta}
+{Whenever \underscoreparlink{export\_full\_f}{export_full_f} or \underscoreparlink{export\_delta\_f}{export_delta_f} is \true, and \underscoreparlink{export\_f\_theta\_option}{export_f_theta_option} $>0$.}
+{Values of $\theta$ on which you want to save the distribution function.  modulo$(\ldots, 2\pi)$ will be applied.  See \underscoreparlink{export\_f\_theta\_option}{export_f_theta_option} for details}
+
+\myhrule
+
+\paramWithUnderscore{export\_f\_zeta}{export_f_zeta}
 {1D array of reals}
 {0.0}
-{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true, and \parlink{export\_f\_zeta\_option} $>0$.}
-{Values of $\zeta$ on which you want to save the distribution function.  modulo$(\ldots, 2\pi/\mbox{\ttfamily NPeriods})$ will be applied.  See \parlink{export\_f\_zeta\_option} for details}
-
-\myhrule
-
-\param{export\_f\_xi\_option}
+{Whenever \underscoreparlink{export\_full\_f}{export_full_f} or \underscoreparlink{export\_delta\_f}{export_delta_f} is \true, and \underscoreparlink{export\_f\_zeta\_option}{export_f_zeta_option} $>0$.}
+{Values of $\zeta$ on which you want to save the distribution function.
+modulo$(\ldots, 2\pi/\mbox{\ttfamily NPeriods})$ will be applied.
+See \underscoreparlink{export\_f\_zeta\_option}{export_f_zeta_option} for details}
+
+\myhrule
+
+\paramWithUnderscore{export\_f\_xi\_option}{export_f_xi_option}
 {integer}
 {1}
-{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true}
+{Whenever \underscoreparlink{export\_full\_f}{export_full_f} or \underscoreparlink{export\_delta\_f}{export_delta_f} is \true}
 {Controls which discretization in $\xi$ is used for exporting the distribution function.\\
@@
 
-{\ttfamily export\_f\_xi\_option} = 1: Report the distribution function on the values of $\xi$ specified by \parlink{export\_f\_xi}. No sorting of the requested values is performed.
-}
-
-\myhrule
-
-\param{export\_f\_xi}
+{\ttfamily export\_f\_xi\_option} = 1: Report the distribution function on the values of $\xi$ specified by \underscoreparlink{export\_f\_xi}{export_f_xi}.
+No sorting of the requested values is performed.
+}
+
+\myhrule
+
+\paramWithUnderscore{export\_f\_xi}{export_f_xi}
 {1D array of reals}
 {0.0}
-{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true, and \parlink{export\_f\_xi\_option} = 1.}
+{Whenever \underscoreparlink{export\_full\_f}{export_full_f} or \underscoreparlink{export\_delta\_f}{export_delta_f} is \true, and \underscoreparlink{export\_f\_xi\_option}{export_f_xi_option} = 1.}
 {Values of $\xi$ on which you want to save the distribution function.  Values must lie in the range $[-1,1]$.}
@@
 
-\param{export\_f\_x\_option}
+\paramWithUnderscore{export\_f\_x\_option}{export_f_x_option}
 {integer}
 {0}
-{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true}
+{Whenever \underscoreparlink{export\_full\_f}{export_full_f} or \underscoreparlink{export\_delta\_f}{export_delta_f} is \true}
 {Controls which grid in $x = v/\sqrt{2T/m}$ is used for exporting the distribution function.\\
@@
 
-{\ttfamily export\_f\_x\_option} = 1: Interpolate to a different grid, specified by \parlink{export\_f\_x}.  Polynomial spectral interpolation will be used.
+{\ttfamily export\_f\_x\_option} = 1: Interpolate to a different grid, specified by \underscoreparlink{export\_f\_x}{export_f_x}.  Polynomial spectral interpolation will be used.
   No sorting of the requested values is performed.\\
 
-{\ttfamily export\_f\_x\_option} = 2: Do not interpolate. Use the values of the internal $x$ grid that are closest to the values requested in \parlink{export\_f\_x}.
+{\ttfamily export\_f\_x\_option} = 2: Do not interpolate. Use the values of the internal $x$ grid that are closest to the values requested in \underscoreparlink{export\_f\_x}{export_f_x}.
   Values of $x$ will be in increasing order. If multiple requested values are close to the same grid point, the number of
@@
 
-For all of these options, you can see \parlink{export\_f\_x} in {\ttfamily sfincsOutput.h5} for the actual grid used in the end.
-}
-
-\myhrule
-
-\param{export\_f\_x}
+For all of these options, you can see \underscoreparlink{export\_f\_x}{export_f_x} in {\ttfamily sfincsOutput.h5} for the actual grid used in the end.
+}
+
+\myhrule
+
+\paramWithUnderscore{export\_f\_x}{export_f_x}
 {1D array of reals}
 {1.0}
-{Whenever \parlink{export\_full\_f} or \parlink{export\_delta\_f} is \true, and \parlink{export\_f\_x\_option} $>0$.}
+{Whenever \underscoreparlink{export\_full\_f}{export_full_f} or \underscoreparlink{export\_delta\_f}{export_delta_f} is \true, and \underscoreparlink{export\_f\_x\_option}{export_f_x_option} $>0$.}
 {Values of $x$ on which you want to save the distribution function.  Values must be $\ge 0$.}
-
-\section{The \texttt{adjointOptions} namelist}
-\label{sec:adjointOptions}
-
-\param{adjointBootstrapOption}
-{Boolean}
-{\false}
-{When \parlink{RHSMode} == 4 or 5}
-{If this parameter is \true, the derivative of the bootstrap current (\parlink{FSABjHatOverRootFSAB2}) is computed with respect to geometry parameters.
-If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
-Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
-
-\myhrule
-
-\param{adjointRadialCurrentOption}
-{Boolean}
-{\false}
-{When \parlink{RHSMode} == 4 or 5}
-{If this parameter is \true the derivative of the radial current ($J_r$) is computed with respect to geometry parameters,
-\begin{gather}
-J_r = \sum_s \texttt{Zs}(s) \texttt{particleFlux\_vm\_rN}(s),
-\end{gather}
-where \parlink{Zs} and \parlink{particleFlux\_vm\_rN} can be found elsewhere in this document.
-If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
-Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
-
-\myhrule
-
-\param{adjointTotalHeatFluxOption}
-{Boolean}
-{\false}
-{When \parlink{RHSMode} == 4 or 5}
-{If this parameter is \true, the derivative of the total heat flux ($Q$) is computed with respect to geometry parameters,
-\begin{gather}
-Q = \sum_s \texttt{Zs}(s) \texttt{heatFlux\_vm\_rN}(s),
-\end{gather}
-where \parlink{Zs} and \parlink{heatFlux\_vm\_rN} can be found elsewhere in this document.
-If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
-Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
-
-\myhrule
-
-\param{adjointHeatFluxOption}
-{1D array of booleans}
-{\false}
-{When \parlink{RHSMode} == 4 or 5}
-{This array is of the same length of {\ttfamily nHats} and other species parameters.
-If parameter is \true, the derivative of the heat flux (\parlink{heatFlux\_vm\_rN}) will be computed for this species with respect to geometry parameters.
-If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
-Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
-
-\myhrule
-
-\param{adjointParticleFluxOption}
-{1D array of booleans}
-{\false}
-{When \parlink{RHSMode} == 4 or 5}
-{This array is of the same length of {\ttfamily nHats} and other species parameters.
-If parameter is {\ttfamily .true.}, the derivative of the particle flux (\parlink{particleFlux\_vm\_rN}) will be computed for this species with respect to geometry parameters.
-If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
-Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
-
-\myhrule
-
-\param{adjointParallelFlowOption}
-{1D array of booleans}
-{\false}
-{When \parlink{RHSMode} == 4 or 5}
-{This array is of the same length of {\ttfamily nHats} and other species parameters.
-If parameter is {\ttfamily .true.}, the derivative of the parallel flow (\parlink{FSABVelocityUsingFSADensityOverRootFSAB2}) will be computed for this species with respect to geometry parameters.
-If \parlink{geometryScheme} == 5, then derivatives will be computed with respect to the quantities \texttt{bmnc}, \texttt{bsupthetamnc}, \texttt{bsupzetamnc}, \texttt{bsubthetamnc}, \texttt{bxubzetamnc}, and \texttt{gmnc} which are found in the wout file.
-Otherwise, derivatives will be computed with respect to the Boozer spectrum, \texttt{bmnc}.}
-
-\myhrule
-
-\param{nMaxAdjoint}
-{integer}
-{0}
-{When \parlink{RHSMode} == 4 or 5}
-{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \parlink{adjointParallelFlowOption} will computed for Fourier modes $|n| < $ \texttt{nMaxAdjoint}.
-}
-
-\myhrule
-
-\param{mMaxAdjoint}
-{integer}
-{0}
-{When \parlink{RHSMode} == 4 or 5}
-{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \parlink{adjointParallelFlowOption} will computed for Fourier modes $m < $ \texttt{mMaxAdjoint}.
-}
-
-\myhrule
-
-\param{nMinAdjoint}
-{integer}
-{0}
-{When \parlink{RHSMode} == 4 or 5}
-{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \parlink{adjointParallelFlowOption} will computed for Fourier modes $|n| > $ \texttt{nMaxAdjoint}.
-}
-
-\myhrule
-
-\param{mMinAdjoint}
-{integer}
-{0}
-{When \parlink{RHSMode} == 4 or 5}
-{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \parlink{adjointParallelFlowOption} will computed for Fourier modes $m > $ \texttt{mMaxAdjoint}.
-}
-
-\myhrule
-
-\param{debugAdjoint}
-{boolean}
-{\false}
-{When \parlink{RHSMode} == 4 or 5. }
-{If \texttt{.true.}, then derivatives computed with the adjoint method will be compared with forward difference derivatives with step size set by \parlink{deltaLambda}.}
-
-\myhrule
-
-\param{deltaLambda}
-{real}
-{1.d-4}
-{When \parlink{deltaLamda} == \texttt{.true.}}
-{The step size used for forward difference derivatives will be \texttt{deltaLambda} multiplied by the quantity that is being varied.
-For example, the step size for a Boozer harmonic, $B_{mn}$, would be $\Delta B_{mn} = $\texttt{deltaLambda}$B_{mn}$.}
-
-\myhrule
 
@@
 
-Notice that exactly 1 of the 4 variables \parlink{psiHat\_max}, \parlink{psiN\_max}, \parlink{rHat\_max}, and \parlink{rN\_max}
+Notice that exactly 1 of the 4 variables \underscoreparlink{psiHat\_max}{psiHat_max}, \underscoreparlink{psiN\_max}{psiN_max}, \underscoreparlink{rHat\_max}{rHat_max}, and \underscoreparlink{rN\_max}{rN_max}
 will be used, depending on
@@
 Similarly, exactly 1 of the 4 variables
-\parlink{psiHat\_min}, \parlink{psiN\_min}, \parlink{rHat\_min}, and \parlink{rN\_min}
+\underscoreparlink{psiHat\_min}{psiHat_min}, \underscoreparlink{psiN\_min}{psiN_min}, \underscoreparlink{rHat\_min}{rHat_min}, and \underscoreparlink{rN\_min}{rN_min}
 will be used.
@@
 
-\ssparam{psiHat\_max}
+\ssparamWithUnderscore{psiHat\_max}{psiHat_max}
 {real}
@@
 
-\ssparam{psiHat\_min}
+\ssparamWithUnderscore{psiHat\_min}{psiHat_min}
 {real}
@@
 
-\ssparam{psiN\_max}
+\ssparamWithUnderscore{psiN\_max}{psiN_max}
 {real}
@@
 
-\ssparam{psiN\_min}
+\ssparamWithUnderscore{psiN\_min}{psiN_min}
 {real}
@@
 
-\ssparam{rHat\_max}
+\ssparamWithUnderscore{rHat\_max}{rHat_max}
 {real}
@@
 
-\ssparam{rHat\_min}
+\ssparamWithUnderscore{rHat\_min}{rHat_min}
 {real}
@@
 
-\ssparam{rN\_max}
+\ssparamWithUnderscore{rN\_max}{rN_max}
 {real}
@@
 
-\ssparam{rN\_min}
+\ssparamWithUnderscore{rN\_min}{rN_min}
 {real}

### 2019-03-05 doc/manual/version3/outputParameters.tex
Substantial revision to manual

--- 2018-02-15 doc/manual/version3/outputParameters.tex
@@
 \newcommand{\outputparamWithUnderscore}[5]{{\setlength{\parindent}{0cm} {\ttfamily \bfseries \hypertarget{#2}{#1}}\\{\it Type}: #3\\{\it Available when}: #4\\{\it Meaning}: #5}}
-\newcommand{\underscoreparlink}[2]{{\ttfamily \hyperlink{#2}{#1}}}
+%\newcommand{\underscoreparlink}[2]{{\ttfamily \hyperlink{#2}{#1}}} %%Moved this to SFINCSUserManual.tex
 
-In this chapter we describe output parameters which are available in the \sfincs~output file \parlink{outputFileName} after a %successful 
+In this chapter we describe output parameters which are available in the \sfincs~output file (specified by \parlink{outputFileName}) after a %successful 
 \sfincs~run. 
@@
 \section{Convergence and test parameters}
+\label{sec:ConvergenceAndTestParameters}
 
@@
 {integer}
-{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true, \parlink{includePhi1} == \true and \parlink{finished} available.}
+{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true, \parlink{includePhi1} == \true, \parlink{readExternalPhi1} == \false and \parlink{finished} available.}
 {If this variable is \true~the nonlinear iteration has converged. Interpret this Boolean data by comparing to \parlink{integerToRepresentTrue}/\parlink{integerToRepresentFalse}.}
@@
 {real}
-{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true and \parlink{includePhi1} == \true.}
+{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true, \parlink{includePhi1} == \true and \parlink{readExternalPhi1} == \false}
 {Lagrange multiplier associated with the constraint that $\left\langle \Phi_1\right\rangle = 0$, which is included in the quasineutrality equation. %Should be within machine precision of 0.
 This parameter should come out as negligible in quasineutrality, e.g. its size should be significantly smaller than the maximum charge density of the plasma species $Z_s n_s$.
-\todo{Is this correct?}}
+%\todo{Is this correct?}
+}
 
@@
 {integer}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {Number of iterations in the nonlinear calculation. (In a linear calculation this will be 1.)}
@@
 \section{Radial fluxes and flows}
+\label{sec:RadialFluxesAndFlows}
 
@@
 Here we only present all possible combinations for the particle flux, but the pattern is the same for the momentum and energy flux.
+%
+\subsubsection{Classical fluxes}
+%
+The classical flux is often neglected as smaller than the neoclassical flux.
+However, in a stellarator that has been optimized to have a low ratio of parallel to perpendicular current (like W7-X), the classical flux can be of the same size as the neoclassical (particularly if the collisionality is large) \cite{Buller2018}.
+\sfincs~calculates the radial classical particle and heat fluxes, using the same normalizations as for the corresponding neoclassical fluxes. 
+Note that the classical fluxes (in contrast to the neoclassical) depend on the gyro-phase dependent part of the distribution function. 
+However, the first-order gyro-phase dependent part of the distribution function does not depend on the solution to the drift-kinetic equation in \sfincs, and the classical fluxes can thus be calculated to required order even without solving it.
+The classical fluxes do depend on $\Phi_1$ so if the quasi-neutrality equation is solved the fluxes will depend on the solution to the system of equations.
 
@@
 {particleFlux_vd_psiHat}
-{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true and \parlink{includePhi1} == \true.}
+{%$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ 
+\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true and \parlink{includePhi1} == \true}
 {$\displaystyle \text{\ttfamily particleFlux\_vd\_psiHat} = \frac{\bar{R}}{\bar{n} \bar{v}} \left\langle \int d^3v f_s \, \bm{v}_{ds} \cdot \nabla \hat{\psi}\right\rangle$ for all plasma species.
@@
 {particleFlux_vm_psiHat}
-{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {$\displaystyle \text{\ttfamily particleFlux\_vm\_psiHat} = \frac{\bar{R}}{\bar{n} \bar{v}} \left\langle \int d^3v f_s \, \bm{v}_{ms} \cdot \nabla \hat{\psi}\right\rangle$ for all plasma species.
@@
 {particleFlux_vE_psiHat}
-{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true and \parlink{includePhi1} == \true.}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true and \parlink{includePhi1} == \true}
 {$\displaystyle \text{\ttfamily particleFlux\_vE\_psiHat} = \frac{\bar{R}}{\bar{n} \bar{v}} \left\langle \int d^3v f_s \, \bm{v}_{E} \cdot \nabla \hat{\psi}\right\rangle$ for all plasma species.
@@
 {particleFlux_vm0_psiHat}
-{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {$\displaystyle \text{\ttfamily particleFlux\_vm0\_psiHat} = \frac{\bar{R}}{\bar{n} \bar{v}} \left\langle \int d^3v f_{s0} \, \bm{v}_{ms} \cdot \nabla \hat{\psi}\right\rangle$ for all plasma species.
 The final result is in the last index corresponding to the last nonlinear iteration.
-This variable should cancel \underscoreparlink{particleFlux\_vE0\_psiHat}{particleFlux_vE0_psiHat} in the last nonlinear iteration since $f_{s0}$ cannot drive radial transport (i.e.
-in a linear calculations this quantity should be 0 to high precision unless there is a radial current in the magnetic equilibrium).}
+This variable should cancel \underscoreparlink{particleFlux\_vE0\_psiHat}{particleFlux_vE0_psiHat} in the last nonlinear iteration since $f_{s0}$ cannot drive radial particle transport (in a calculation without $\Phi_1$ this quantity should be 0 to high precision unless there is a radial current in the magnetic equilibrium),\newline i.e.
+$\displaystyle \text{\ttfamily particleFlux\_vm0\_psiHat} = - \text{\ttfamily particleFlux\_vE0\_psiHat}$.}
 
@@
 {particleFlux_vE0_psiHat}
-{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true and \parlink{includePhi1} == \true.}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true and \parlink{includePhi1} == \true}
 {$\displaystyle \text{\ttfamily particleFlux\_vE0\_psiHat} = \frac{\bar{R}}{\bar{n} \bar{v}} \left\langle \int d^3v f_{s0} \, \bm{v}_{E} \cdot \nabla \hat{\psi}\right\rangle$ for all plasma species.
 The final result is in the last index corresponding to the last nonlinear iteration.
-This variable should cancel \underscoreparlink{particleFlux\_vm0\_psiHat}{particleFlux_vm0_psiHat} in the last nonlinear iteration since $f_{s0}$ cannot drive radial transport.}
+This variable should cancel \underscoreparlink{particleFlux\_vm0\_psiHat}{particleFlux_vm0_psiHat} in the last nonlinear iteration since $f_{s0}$ cannot drive radial particle transport, i.e.
+$\displaystyle \text{\ttfamily particleFlux\_vm0\_psiHat} = - \text{\ttfamily particleFlux\_vE0\_psiHat}$.}
 
@@
 {particleFlux_vd1_psiHat}
-{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true and \parlink{includePhi1} == \true.}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1, \parlink{solveSystem} == \true and \parlink{includePhi1} == \true}
 {$\displaystyle \text{\ttfamily particleFlux\_vd1\_psiHat} = \text{\ttfamily particleFlux\_vm\_psiHat}~+~\text{\ttfamily particleFlux\_vE0\_psiHat}$.
@@
 
+\outputparamWithUnderscore{classicalParticleFlux\_psiHat}
+{classicalParticleFlux_psiHat}
+{%$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ 
+\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
+{$\displaystyle \text{\ttfamily classicalParticleFlux\_psiHat} = \frac{\bar{R}}{\bar{n} \bar{v}} \frac{1}{Z_s e} \left\langle \frac{\vect{B} \times \nabla \hat{\psi}}{B^2} \cdot \vect{R}_s \right\rangle$ for all plasma species.
+$\displaystyle \vect{R}_s = m_s \sum_a \int d^3v \, \vect{v} C_s \left[f_s, f_a\right]$ is the friction force acting on the species.
+The final result is in the last index corresponding to the last nonlinear iteration.}
+
+\myhrule
+
 \outputparamWithUnderscore{particleFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}}
 {particleFlux_vXN_Y}
-{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{See above for various $\textcolor{magenta}{XN}$.}
-{For different options of $\textcolor{magenta}{XN}$ see output {\ttfamily particleFlux} parameters defined above. 
-For different options of $\textcolor{magenta}{Y}$ see flux-surface labels (section~\ref{sec:radialCoordinates}). 
-$\displaystyle \text{\ttfamily particleFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}} \left\langle \int d^3v f_{s\textcolor{magenta}{N}} \, \bm{v}_{\textcolor{magenta}{X}s} \cdot \nabla \textcolor{magenta}{Y}\right\rangle$ for all plasma species.
-The final result is in the last index corresponding to the last nonlinear iteration.}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{See above for various $\textcolor{magenta}{XN}$}
+{$\displaystyle \text{\ttfamily particleFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}} \left\langle \int d^3v f_{s\textcolor{magenta}{N}} \, \bm{v}_{\textcolor{magenta}{X}s} \cdot \nabla \textcolor{magenta}{Y}\right\rangle$ for all plasma species.
+The final result is in the last index corresponding to the last nonlinear iteration.\newline
+For different options of $\textcolor{magenta}{XN}$ see output {\ttfamily particleFlux} parameters defined above.\newline 
+For different options of $\textcolor{magenta}{Y}$ see flux-surface labels (section~\ref{sec:radialCoordinates}).}
+
+\myhrule
+
+\outputparamWithUnderscore{classicalParticleFlux\_\textcolor{magenta}{Y}}
+{classicalParticleFlux_Y}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
+{$\displaystyle \text{\ttfamily classicalParticleFlux\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}} \frac{1}{Z_s e} \left\langle \frac{\vect{B} \times \nabla \textcolor{magenta}{Y}}{B^2} \cdot \vect{R}_s \right\rangle$ for all plasma species.
+$\displaystyle \vect{R}_s = m_s \sum_a \int d^3v \, \vect{v} C_s \left[f_s, f_a\right]$ is the friction force acting on the species.
+The final result is in the last index corresponding to the last nonlinear iteration.\newline
+For different options of $\textcolor{magenta}{Y}$ see flux-surface labels (section~\ref{sec:radialCoordinates}).}
+
+\myhrule
+
+\outputparamWithUnderscore{classicalParticleFluxNoPhi1\_\textcolor{magenta}{Y}}
+{classicalParticleFluxNoPhi1_Y}
+{\parlink{Nspecies} array of reals}
+{\parlink{RHSMode} == 1}
+{$\displaystyle \text{\ttfamily classicalParticleFluxNoPhi1\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}} \frac{1}{Z_s e} \left\langle \frac{\vect{B} \times \nabla \textcolor{magenta}{Y}}{B^2} \cdot \vect{R}_{s,\,\mathrm{w/o}\,\Phi_1} \right\rangle$ for all plasma species.
+$\displaystyle \vect{R}_{s,\,\mathrm{w/o}\,\Phi_1} = m_s \sum_a \int d^3v \, \vect{v} C_s \left[f_{s,\,\mathrm{w/o}\,\Phi_1}, f_{a,\,\mathrm{w/o}\,\Phi_1}\right]$ is the friction force acting on the species, where the impact of $\Phi_1$ is neglected.\newline
+For different options of $\textcolor{magenta}{Y}$ see flux-surface labels (section~\ref{sec:radialCoordinates}).}
 
@@
 {momentumFlux_vXN_Y}
-{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{See above for various $\textcolor{magenta}{XN}$.}
-{For different options of $\textcolor{magenta}{XN}$ see  
-\underscoreparlink{particleFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}}{particleFlux_vXN_Y}.
-For different options of $\textcolor{magenta}{Y}$ see flux-surface labels (section~\ref{sec:radialCoordinates}). 
-$\displaystyle \text{\ttfamily momentumFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}^2 \bar{B} \bar{m}} \left\langle \int d^3v f_{s\textcolor{magenta}{N}} \, B m_s v_{\|} \bm{v}_{\textcolor{magenta}{X}s} \cdot \nabla \textcolor{magenta}{Y}\right\rangle$ for all plasma species.
-The final result is in the last index corresponding to the last nonlinear iteration.}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{See above for various $\textcolor{magenta}{XN}$}
+{$\displaystyle \text{\ttfamily momentumFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}^2 \bar{B} \bar{m}} \left\langle \int d^3v f_{s\textcolor{magenta}{N}} \, B m_s v_{\|} \bm{v}_{\textcolor{magenta}{X}s} \cdot \nabla \textcolor{magenta}{Y}\right\rangle$ for all plasma species.
+The final result is in the last index corresponding to the last nonlinear iteration.\newline
+For different options of $\textcolor{magenta}{XN}$ see  
+\underscoreparlink{particleFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}}{particleFlux_vXN_Y}.\newline
+For different options of $\textcolor{magenta}{Y}$ see flux-surface labels (section~\ref{sec:radialCoordinates}).}
 
@@
 {heatFlux_vXN_Y}
-{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{See above for various $\textcolor{magenta}{XN}$.}
-{For different options of $\textcolor{magenta}{XN}$ see  
-\underscoreparlink{particleFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}}{particleFlux_vXN_Y}.
-For different options of $\textcolor{magenta}{Y}$ see flux-surface labels (section~\ref{sec:radialCoordinates}). 
-$\displaystyle \text{\ttfamily heatFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}^3 \bar{m}} \left\langle \int d^3v f_{s\textcolor{magenta}{N}} \, \frac{m_s v^2}{2} \bm{v}_{\textcolor{magenta}{X}s} \cdot \nabla \textcolor{magenta}{Y}\right\rangle$ for all plasma species.
-The final result is in the last index corresponding to the last nonlinear iteration.}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{See above for various $\textcolor{magenta}{XN}$}
+{$\displaystyle \text{\ttfamily heatFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}^3 \bar{m}} \left\langle \int d^3v f_{s\textcolor{magenta}{N}} \, \frac{m_s v^2}{2} \bm{v}_{\textcolor{magenta}{X}s} \cdot \nabla \textcolor{magenta}{Y}\right\rangle$ for all plasma species.
+The final result is in the last index corresponding to the last nonlinear iteration.\newline
+For different options of $\textcolor{magenta}{XN}$ see  
+\underscoreparlink{particleFlux\_v\textcolor{magenta}{XN}\_\textcolor{magenta}{Y}}{particleFlux_vXN_Y}.\newline
+For different options of $\textcolor{magenta}{Y}$ see flux-surface labels (section~\ref{sec:radialCoordinates}).}
+
+\myhrule
+
+\outputparamWithUnderscore{classicalHeatFlux\_\textcolor{magenta}{Y}}
+{classicalHeatFlux_Y}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
+{\todo{Can Stefan update the text so it is correct?}\newline
+$\displaystyle \text{\ttfamily classicalHeatFlux\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}} \frac{1}{Z_s e} \left\langle \frac{\vect{B} \times \nabla \textcolor{magenta}{Y}}{B^2} \cdot \vect{R}_s \right\rangle$ for all plasma species.
+$\displaystyle \vect{R}_s = m_s \sum_a \int d^3v \, \vect{v} C_s \left[f_s, f_a\right]$ is the friction force acting on the species.
+The final result is in the last index corresponding to the last nonlinear iteration.\newline
+For different options of $\textcolor{magenta}{Y}$ see flux-surface labels (section~\ref{sec:radialCoordinates}).}
+
+\myhrule
+
+\outputparamWithUnderscore{classicalHeatFluxNoPhi1\_\textcolor{magenta}{Y}}
+{classicalHeatFluxNoPhi1_Y}
+{\parlink{Nspecies} array of reals}
+{\parlink{RHSMode} == 1}
+{\todo{Can Stefan update the text so it is correct?}\newline
+$\displaystyle \text{\ttfamily classicalHeatFluxNoPhi1\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}} \frac{1}{Z_s e} \left\langle \frac{\vect{B} \times \nabla \textcolor{magenta}{Y}}{B^2} \cdot \vect{R}_{s,\,\mathrm{w/o}\,\Phi_1} \right\rangle$ for all plasma species.
+$\displaystyle \vect{R}_{s,\,\mathrm{w/o}\,\Phi_1} = m_s \sum_a \int d^3v \, \vect{v} C_s \left[f_{s,\,\mathrm{w/o}\,\Phi_1}, f_{a,\,\mathrm{w/o}\,\Phi_1}\right]$ is the friction force acting on the species, where the impact of $\Phi_1$ is neglected.\newline
+For different options of $\textcolor{magenta}{Y}$ see flux-surface labels (section~\ref{sec:radialCoordinates}).}
 
@@
 \outputparam{flow}
-{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{Nzeta}~$\times$~\parlink{Ntheta}~$\times$~\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {$\displaystyle \text{\ttfamily flow} =  \frac{1}{\bar{n} \bar{v}} \int d^3v \, v_{\|} f_s$.}
@@
 \outputparam{velocityUsingFSADensity}
-{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{Nzeta}~$\times$~\parlink{Ntheta}~$\times$~\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {{\ttfamily velocityUsingFSADensity} =  \parlink{flow}~/~\parlink{nHats}.}
@@
 \outputparam{velocityUsingTotalDensity}
-{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{Nzeta}~$\times$~\parlink{Ntheta}~$\times$~\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {{\ttfamily velocityUsingTotalDensity} =  \parlink{flow}~/~\parlink{totalDensity}.}
@@
 \outputparam{MachUsingFSAThermalSpeed}
-{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{Nzeta}~$\times$~\parlink{Ntheta}~$\times$~\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {{\ttfamily MachUsingFSAThermalSpeed} =  (\parlink{mHats}~/~\parlink{THats})${}^{1/2}$~$\cdot$~\parlink{flow}~/~\parlink{nHats}.}
@@
 \outputparam{FSABFlow}
-{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {$\displaystyle \text{\ttfamily FSABFlow} =  \frac{1}{\bar{v} \bar{B} \bar{n}} \left\langle B \int d^3v \, v_{\|} f_s \right\rangle$.}
@@
 \outputparam{FSABVelocityUsingFSADensity}
-{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {{\ttfamily FSABVelocityUsingFSADensity} =  \parlink{FSABFlow}~/~\parlink{nHats}.}
@@
 \outputparam{FSABVelocityUsingFSADensityOverB0}
-{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {{\ttfamily FSABVelocityUsingFSADensityOverB0} =  ($\bar{B}$~/~$B_0$)~$\cdot$~\parlink{FSABFlow}~/~\parlink{nHats}.}
@@
 \outputparam{FSABVelocityUsingFSADensityOverRootFSAB2}
-{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {{\ttfamily FSABVelocityUsingFSADensityOverRootFSAB2} =  (1~/~\parlink{nHats})~$\cdot$~\parlink{FSABFlow}~/~(\parlink{FSABHat2})${}^{1/2}$.}
@@
 \outputparam{jHat}
-{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{Nzeta}~$\times$~\parlink{Ntheta}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {$\displaystyle \text{\ttfamily jHat} = \frac{\bm{j} \cdot \bm{b}}{e \bar{n} \bar{v}} = \frac{1}{e \bar{n} \bar{v}} \sum_s Z_s e \int d^3v \, v_{\|} f_s = \sum_s Z_s \cdot \text{\parlink{flow}}_s$.}
@@
 \outputparam{FSABjHat}
-{$\text{\ttfamily NIterations}$ array of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{NIterations} array of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {$\displaystyle \text{\ttfamily FSABjHat} =  \sum_s Z_s \cdot \text{\parlink{FSABFlow}}_s$.}
@@
 \outputparam{FSABjHatOverB0}
-{$\text{\ttfamily NIterations}$ array of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{NIterations} array of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {$\displaystyle \text{\ttfamily FSABjHatOverB0} =  \frac{\bar{B}}{B_0} \text{\parlink{FSABjHat}}$.}
@@
 \outputparam{FSABjHatOverRootFSAB2}
-{$\text{\ttfamily NIterations}$ array of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{NIterations} array of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {$\displaystyle \text{\ttfamily FSABjHatOverRootFSAB2} =  \frac{\bar{\text{\parlink{FSABjHat}}}}{\left(\text{\parlink{FSABHat2}}\right)^{1/2}}$.}
@@
 \section{Distributions}
+\label{sec:Distributions}
+
+\myhrule
+
+\outputparam{theta}
+{\parlink{Ntheta} array of reals}
+{Always}
+{Grid points in the poloidal angle, which runs from $0$ to $2 \pi$.}
+
+\myhrule
+
+\outputparam{zeta}
+{\parlink{Nzeta} array of reals}
+{Always}
+{Grid points in the toroidal angle, which runs from $0$ to $2 \pi/$ \parlink{NPeriods}.}
+
+\myhrule
+
+\outputparam{x}
+{\parlink{Nx} array of reals}
+{Always}
+{Grid points in normalized speed, $x_s = v / \sqrt{2 T_s / m_s}$, the same for each species $s$.}
 
@@
 \outputparam{densityPerturbation}
-{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{Nzeta}~$\times$~\parlink{Ntheta}~$\times$~\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {$\displaystyle \text{\ttfamily densityPerturbation} =  \frac{1}{\bar{n}} \int d^3v f_{s1}$.}
@@
 \outputparam{totalDensity}
-{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
-{{\ttfamily totalDensity} = \parlink{nHats}~+~\parlink{densityPerturbation}.}
+{\parlink{Nzeta}~$\times$~\parlink{Ntheta}~$\times$~\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
+{{\ttfamily totalDensity} = \parlink{nHats}$\cdot \exp \left(- \frac{Z e \Phi_1}{T}  \right)$~+~\parlink{densityPerturbation}.}
 
@@
 \outputparam{FSADensityPerturbation}
-{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {$\displaystyle \text{\ttfamily FSADensityPerturbation} =  \frac{1}{\bar{n}} \left\langle \int d^3v f_{s1} \right\rangle$. 
-Should be within machine precision of 0.}
+Should be %within machine precision of 
+nearly 0, within roundoff error.}
 
@@
 \outputparam{pressurePerturbation}
-{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{Nzeta}~$\times$~\parlink{Ntheta}~$\times$~\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {$\displaystyle \text{\ttfamily densityPerturbation} =  \frac{1}{\bar{n} \bar{T}} \frac{m_s}{3} \int d^3v \, v^2 f_{s1}$.}
@@
 \outputparam{totalPressure}
-{$\text{\ttfamily Nzeta} \times \text{\ttfamily Ntheta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
-{{\ttfamily totalPressure} = \parlink{nHats} $\cdot$ \parlink{THats}~+~\parlink{pressurePerturbation}.}
+{\parlink{Nzeta}~$\times$~\parlink{Ntheta}~$\times$~\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
+{{\ttfamily totalPressure} = \parlink{nHats}$\cdot \exp \left(- \frac{Z e \Phi_1}{T}  \right)$ $\cdot$ \parlink{THats}~+~\parlink{pressurePerturbation}.}
 
@@
 \outputparam{FSAPressurePerturbation}
-{$\text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true.}
+{\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
 {$\displaystyle \text{\ttfamily FSADensityPerturbation} =  \frac{1}{\bar{n} \bar{T}} \frac{m_s}{3} \left\langle \int d^3v \, v^2 f_{s1} \right\rangle$. 
-Should be within machine precision of 0.}
+Should be %within machine precision of 
+nearly 0, within roundoff error.}
 
@@
 
-\outputparam{delta\_f}
-{$M_x \times M_{\xi} \times M_{\zeta} \times M_{\theta} \times \text{\ttfamily Nspecies} \times \text{\ttfamily NIterations}$ matrix of reals}
-{\parlink{export\_delta\_f} == \true.}
-{?}
+\outputparamWithUnderscore{delta\_f}
+{delta_f}
+{$M_x \times M_{\xi} \times M_{\zeta} \times M_{\theta} \times$~\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{export\_delta\_f} == \true}
+{First order distribution function $f_{s1} = f_{s} - f_{s0}$ for each species, normalized by $\bar{n} / \bar{v}^3$. 
+The phase-space grid $\left(M_x, M_{\xi}, M_{\zeta}, M_{\theta}\right)$} is specified by the input parameters in the {\ttfamily \hyperref[sec:exportfParameters]{export\_f}} namelist.
+
+\myhrule
+
+\outputparamWithUnderscore{full\_f}
+{full_f}
+{$M_x \times M_{\xi} \times M_{\zeta} \times M_{\theta} \times$~\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{export\_full\_f} == \true}
+{Full distribution function $f_{s}$ for each species, normalized by $\bar{n} / \bar{v}^3$. 
+The phase-space grid $\left(M_x, M_{\xi}, M_{\zeta}, M_{\theta}\right)$} is specified by the input parameters in the {\ttfamily \hyperref[sec:exportfParameters]{export\_f}} namelist.
+
+\myhrule
+
+\outputparam{Phi1Hat}
+{\parlink{Nzeta}~$\times$~\parlink{Ntheta}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true and \parlink{includePhi1} == \true}
+{Normalized electrostatic potential minus its flux-surface-average, $\displaystyle \text{\ttfamily Phi1Hat} =  \Phi_1/\bar{\Phi}$ where $\Phi_1 = \Phi-\left<\Phi\right>$.}
+
+\myhrule
+
+\outputparam{sources}
+{$N_{\mathrm{sources}}$~$\times$~\parlink{Nspecies}~$\times$~\parlink{NIterations} matrix of reals}
+{\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true and \parlink{constraintScheme} != 0}
+{\todo{Need help from Matt to understand what is in the output. E.g. is the output the coefficients $a_0$, $a_2$, $a_4$ in \parlink{constraintScheme}} }
 
@@
 \section{Miscellaneous}
+\label{sec:Miscellaneous}
 
@@
 
+\outputparam{NPeriods}
+{positive integer}
+{Always}
+{Number of identical toroidal periods (e.g. 5 for W7-X, 10 for LHD, 4 for HSX), equivalent to the \vmec~variable nfp.}
+
+\myhrule
+
+\outputparam{Nspecies}
+{positive integer}
+{Always}
+{Number of particle species}
+
+\myhrule
+
 %The default values are usually best for the parameters in this namelist.

### 2019-03-05 doc/manual/version3/overview.tex
Working on manual
Substantial revision to manual

--- 2015-07-21 doc/manual/overview.tex
@@
 since a radially local approximation is made. 
+The solution is the first-order distribution function $f_{s1} \left(\theta, \zeta, x, \xi\right)$ (where the full distribution is $f_s = f_{s0} + f_{s1}$) on the flux surface for all species $s$, and if quasi-neutrality is solved also the electrostatic potential variation on the flux surface $\Phi_1 \left(\theta, \zeta\right) = \Phi-\left<\Phi\right>$.
 
@@
 \item
+You can choose to include or not include the poloidal and toroidal magnetic drifts in the kinetic equation.
+
+\item
 The code takes advantage of modern algorithms (GMRES) and parallelized libraries (\PETSc, \superludist, and \mumps).
@@
 Optional nonlinear terms in the kinetic equation (involving both the non-Maxwellian distribution function and poloidal/toroidal electric field) can be included using Newton's method.
+
+\item 
+The radial classical particle and heat fluxes are calculated.
 
@@
 
-The output from a single \sfincs~computation is saved in \HDF~format in the file {\ttfamily sfincsOutput.h5}.
-To browse this file you can enter {\ttfamily h5dump sfincsOutput.h5|less} from the command line.
+The output from a single \sfincs~computation is saved in \HDF~format in the file %{\ttfamily sfincsOutput.h5}.
+specified by \parlink{outputFileName}.
+To browse this file you can enter {\ttfamily h5dump outputFileName|less} %{\ttfamily h5dump sfincsOutput.h5|less} 
+from the command line.
 Every array saved in this file is annotated with strings that describe the array dimensions,
@@
 text that describes their meaning and normalization. 
+Some of the most frequently used output quantities are detailed in chapter~\ref{ch:output}.
 

### 2019-03-05 doc/manual/version3/runs.tex
Substantial revision to manual

@@
 The code ``knows'' about the reference values only through
-the 3 combinations \parlink{Delta}, \parlink{alpha}, and \parlink{nu\_n}
+the 3 combinations \parlink{Delta}, \parlink{alpha}, and \underscoreparlink{nu\_n}{nu_n}
 in the {\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}} namelist.
@@
 
-0: Specify the flux surface using \parlink{psiHat\_wish}.\\
+0: Specify the flux surface using \underscoreparlink{psiHat\_wish}{psiHat_wish}.\\
 
-1: Specify the flux surface using \parlink{psiN\_wish}.\\
+1: Specify the flux surface using \underscoreparlink{psiN\_wish}{psiN_wish}.\\
 
-2: Specify the flux surface using \parlink{rHat\_wish}.\\
+2: Specify the flux surface using \underscoreparlink{rHat\_wish}{rHat_wish}.\\
 
-3: Specify the flux surface using \parlink{rN\_wish}.\\
+3: Specify the flux surface using \underscoreparlink{rN\_wish}{rN_wish}.\\
 
@@
 %\section{Quasineutrality and variation of the electrostatic potential on the flux surface}
-\section{Nonlinear calculations, includePhi1 = {\ttfamily .true.}}
+\section{Calculations with $\Phi_1(\theta,\zeta)$, includePhi1 = {\ttfamily .true.}}
 \label{sec:qn}
@@
 to include variation on the flux surface of the electrostatic potential, $\Phi_1(\theta,\zeta)$.
-Such variation does occur to some degree in a real plasma, but it is 
-neglected in analytical theory and in many codes such as \dkes.  It can be proved
+Such variation does occur to some degree in a real plasma, but it has traditionally been  
+neglected in analytical theory and in many codes such as \dkes~(however, nowadays analytical modelling exists, see e.g. \cite{Buller2018} and references therein).  It can be proved
 that including $\Phi_1$ has no effect on the particle or heat fluxes, parallel flows, or bootstrap current
@@
 In \sfincs, you can choose whether or not to
-include $\Phi_1$ using the parameter {\ttfamily \hyperlink{includePhi1}{includePhi1}} in the 
+include $\Phi_1$ using the parameter \parlink{includePhi1} in the 
 {\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}}
 namelist. 
-Furthermore, the parameters \parlink{includePhi1inKineticEquation}, \parlink{includePhi1InCollisionOperator} and \parlink{quasineutralityOption} can be used to control how $\Phi_1$ is included in the system of equations.
-If and only if $\Phi_1$ is included, a quasineutrality equation is solved
+Furthermore, the parameters \parlink{includePhi1InKineticEquation}, \parlink{includePhi1InCollisionOperator} and \parlink{quasineutralityOption} can be used to control how $\Phi_1$ is included in the system of equations.
+\sfincs~also allows to read $\Phi_1$ as an input by using \parlink{readExternalPhi1}.
+This is useful for looking at the effect on the transport by a $\Phi_1$ generated by some mechanism not treated by \sfincs, or for using $\Phi_1$ calculated by another code.
+Moreover, with $\Phi_1$ as an input the runs are typically cheaper than if \sfincs~calculates $\Phi_1$ itself, so if the potential variation is not expected to change in a parameter scan this setting could be useful to reuse the same $\Phi_1$ and decrease the computational costs.
+If and only if $\Phi_1$ is included as an unknown (i.e. not if $\Phi_1$ is an input with \parlink{readExternalPhi1}), a quasineutrality equation is solved
 at each point on the flux surface.  Due to these extra unknowns ($\Phi_1$) and extra equations
@@
 Thus, there is typically not a substantial increase in memory requirements associated with \parlink{includePhi1} (however, setting the preconditioner option \parlink{reusePreconditioner} = \false~can lead to a significant increase in memory requirements).
-The time requirements will normally increase when \parlink{includePhi1} is \true~because the system of equations is then nonlinear. 
+The time requirements will normally increase when \parlink{includePhi1} is \true~because the system of equations is then nonlinear (except if \parlink{readExternalPhi1} is \true). 
 The nonlinear equations are solved with a Newton method using the Scalable Nonlinear Equations Solvers (SNES) in \PETSc. SNES internally employs KSP for the solution of its linear systems. 
@@
 The resolution parameters required for numerical convergence in nonlinear calculations are often similar to the parameters in the corresponding linear calculations (based on experience). 
-A sanity check to control that the nonlinear calculation is not incorrect (however it is not a guarantee for numerical convergence), is to check that the total flux-surface averaged particle flux for the lowest order distribution function $\displaystyle \left\langle \int d^3 v \, f_{s0} \left(\bm{v}_{\mathrm{d} s} + \bm{v}_{E}\right) \cdot \nabla \psi \right\rangle$ of each plasma species $s$ comes out as 0.
+A sanity check to control that the nonlinear calculation is not incorrect (however it is not a guarantee for numerical convergence), is to check that the total flux-surface averaged particle flux for the lowest order distribution function $\displaystyle \left\langle \int d^3 v \, f_{s0} \, \bm{v}_{\mathrm{d} s} \cdot \nabla \psi \right\rangle = \left\langle \int d^3 v \, f_{s0} \left(\bm{v}_{\mathrm{m} s} + \bm{v}_{E}\right) \cdot \nabla \psi \right\rangle$ of each plasma species $s$ comes out as 0.
 This can e.g.
-be done by comparing {\ttfamily particleFlux\_vm0\_psiHat} to {\ttfamily particleFlux\_vE0\_psiHat} in the \sfincs~output file \parlink{outputFileName} and see that they are equal but with opposite sign.
+be done by comparing \underscoreparlink{particleFlux\_vm0\_psiHat}{particleFlux_vm0_psiHat} to \underscoreparlink{particleFlux\_vE0\_psiHat}{particleFlux_vE0_psiHat} in the \sfincs~output file \parlink{outputFileName} and see that they are equal but with opposite sign.
 
@@
 These drifts are turned off by default.  To turn them on, all you need to do is
-set \parlink{magneticDriftScheme} = 1 in the {\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}} namelist.
-(Setting \parlink{magneticDriftScheme} = 2 uses a slightly different parallel magnetic drift, which gives
+set \parlink{magneticDriftScheme}~$>$~0 (with the default being \parlink{magneticDriftScheme}~=~1) in the {\ttfamily \hyperref[sec:physicsParameters]{physicsParameters}} namelist. 
+(Setting \parlink{magneticDriftScheme}~=~2 uses a slightly different parallel magnetic drift, which gives
 indistinguishable results to setting 1 for all cases examined so far, and which is in fact exactly identical to setting 1
@@
 If the poloidal/toroidal magnetic drifts are turned on, you must use VMEC
-geometry (\parlink{geometryScheme} = 5), since the magnetic drifts
+geometry (\parlink{geometryScheme}~=~5), since the magnetic drifts
 depend on various derivatives of the components of the magnetic field which
 are not available in the simplified geometry models. 
+(In fact, to include poloidal/toroidal magnetic drifts in a radially local code like \sfincs~is somewhat ambiguous since the approximation made when neglecting the radial derivatives of the unknowns is different in different coordinate system.)
 
@@
 If the electrostatic potential is not constant on flux surfaces and poloidal/toroidal magnetic drifts
-are included, certain terms exist in the kinetic equation which have not yet been implemented in the code.
-Thus, at present it is not strictly correct to simultaneously set  \parlink{magneticDriftScheme} $> 0$ 
-and  \parlink{includePhi1} = \true.
+are included, certain cross-terms exist in the kinetic equation which have not yet been implemented in \sfincs.
+Thus, at present it is not strictly correct to simultaneously set  \parlink{magneticDriftScheme}~$>$~0 
+and  \parlink{includePhi1} = \true, but it is allowed for testing purposes.
 
@@
 When \parlink{RHSMode} = 3, the values of \parlink{Zs}, \parlink{THats}, \parlink{nHats},
-\parlink{mHats}, \parlink{nu\_n}, and {\ttfamily dPhiHatdXXX} are all ignored.
+\parlink{mHats}, \underscoreparlink{nu\_n}{nu_n}, and {\ttfamily dPhiHatdXXX} are all ignored.
 Instead, the collisionality is set by \parlink{nuPrime}, and the radial electric field is set

### 2019-03-08 doc/classical_radial_fluxes_2019-01-17.tex
Updates classical heat flux to correct for an incorrect factor 2 in the normalization. Adds a notice that this correction has been applied to the output, for automatic correction. Updates documentation to reflect these changes.

@@
 \begin{document}
+\textcolor{red}{Update 2019-03-08: the normalized heat flux was a factor 2 too high due to using the wrong normalization. This has been corrected.}
+
 \section{Drift-kinetic classical transport}
@@
   \hat{\Gamma}_a &= \frac{\bar{R}}{\bar{n}\bar{v}} \frac{1}{Z_a e} \lang B^{-2} \left(\vec{B} \times \nabla \hat{\psi}\right) \cdot \vec{R}_a \rang, \\
-  \hat{Q}_a &= \frac{\bar{R}}{\bar{n}\bar{v} \bar{T}} \frac{1}{Z_a e} \lang B^{-2} \left(\vec{B} \times \nabla \hat{\psi}\right) \cdot \vec{G}_a \rang,
+  \hat{Q}_a &= \frac{\bar{R}}{\bar{n}\bar{m}\bar{v}^3} \frac{1}{Z_a e} \lang B^{-2} \left(\vec{B} \times \nabla \hat{\psi}\right) \cdot \vec{G}_a \rang,
 \end{align}
@@
 \begin{aligned}
-  \hspace*{-10ex} \hat{q}_a =  -\frac{\Delta^2 \nu_n}{2} \frac{Z_a\hat{m}_a^{1/2}}{\hat{T}_a^{1/2}} \sum_b Z_b^2 &\left[
+  \hspace*{-10ex} \hat{q}_a =  -\frac{\Delta^2 \nu_n}{\textcolor{red}{4}} \frac{Z_a\hat{m}_a^{1/2}}{\hat{T}_a^{1/2}} \sum_b Z_b^2 &\left[
     \lang \hat{n}_a \hat{n}_b\frac{|\bar{R}\nabla \hat{\psi}|^2}{\hat{B}^2 } \rang M_{ab}^{01}\left(\frac{\hat{T}_a}{Z_a}  \frac{\d \ln\eta_a}{\d \hat{\psi}} - \frac{\hat{T}_b}{Z_b}\frac{\d \ln\eta_b}{\d \hat{\psi}}\right) \right.\\
@@
 \begin{aligned}
-  \hspace*{-10ex} \hat{q}_a =  -\frac{\Delta^2 \nu_n}{2} \frac{Z_a\hat{m}_a^{1/2} \hat{\eta}_a}{\hat{T}_a^{1/2}} \sum_b Z_b^2 \hat{\eta}_b &\left[
+  \hspace*{-10ex} \hat{q}_a =  -\frac{\Delta^2 \nu_n}{\textcolor{red}{4}} \frac{Z_a\hat{m}_a^{1/2} \hat{\eta}_a}{\hat{T}_a^{1/2}} \sum_b Z_b^2 \hat{\eta}_b &\left[
     G_{ab}^{(1)}  M_{ab}^{01}\left(\frac{\hat{T}_a}{Z_a}  \frac{\d \ln\eta_a}{\d \hat{\psi}} - \frac{\hat{T}_b}{Z_b}\frac{\d \ln\eta_b}{\d \hat{\psi}}\right) \right.\\
@@
 \begin{equation}
-\hat{Q}_a = \hat{q}_a + \frac{5}{2} \hat{T}_a \hat{\Gamma}_a. 
+\hat{Q}_a = \hat{q}_a + \frac{5}{\textcolor{red}{4}} \hat{T}_a \hat{\Gamma}_a. 
 \end{equation}

### 2019-03-08 doc/manual/version3/outputParameters.tex
Updates classical heat flux to correct for an incorrect factor 2 in the normalization. Adds a notice that this correction has been applied to the output, for automatic correction. Updates documentation to reflect these changes.

@@
 {\parlink{RHSMode} == 1 and \parlink{solveSystem} == \true}
-{\todo{Can Stefan update the text so it is correct?}\newline
-$\displaystyle \text{\ttfamily classicalHeatFlux\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}} \frac{1}{Z_s e} \left\langle \frac{\vect{B} \times \nabla \textcolor{magenta}{Y}}{B^2} \cdot \vect{R}_s \right\rangle$ for all plasma species.
-$\displaystyle \vect{R}_s = m_s \sum_a \int d^3v \, \vect{v} C_s \left[f_s, f_a\right]$ is the friction force acting on the species.
+{$\displaystyle \text{\ttfamily classicalHeatFlux\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}^3 \bar{m}} \frac{1}{Z_s e} \left\langle \frac{\vect{B} \times \nabla \textcolor{magenta}{Y}}{B^2} \cdot \vect{G}_s \right\rangle$ for all plasma species.
+$\displaystyle \vect{G}_s = \frac{m_s^2}{2}\sum_a \int d^3v \, \vect{v} v^2 C_s \left[f_s, f_a\right]$ is the energy-weighted friction force acting on the species.
 The final result is in the last index corresponding to the last nonlinear iteration.\newline
@@
 {\parlink{RHSMode} == 1}
-{\todo{Can Stefan update the text so it is correct?}\newline
-$\displaystyle \text{\ttfamily classicalHeatFluxNoPhi1\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}} \frac{1}{Z_s e} \left\langle \frac{\vect{B} \times \nabla \textcolor{magenta}{Y}}{B^2} \cdot \vect{R}_{s,\,\mathrm{w/o}\,\Phi_1} \right\rangle$ for all plasma species.
-$\displaystyle \vect{R}_{s,\,\mathrm{w/o}\,\Phi_1} = m_s \sum_a \int d^3v \, \vect{v} C_s \left[f_{s,\,\mathrm{w/o}\,\Phi_1}, f_{a,\,\mathrm{w/o}\,\Phi_1}\right]$ is the friction force acting on the species, where the impact of $\Phi_1$ is neglected.\newline
+{$\displaystyle \text{\ttfamily classicalHeatFluxNoPhi1\_\textcolor{magenta}{Y}} = \frac{\bar{R}}{\bar{n} \bar{v}^3\bar{m}} \frac{1}{Z_s e} \left\langle \frac{\vect{B} \times \nabla \textcolor{magenta}{Y}}{B^2} \cdot \vect{G}_{s,\,\mathrm{w/o}\,\Phi_1} \right\rangle$ for all plasma species.
+$\displaystyle \vect{G}_{s,\,\mathrm{w/o}\,\Phi_1} = \frac{m_s^2}{2}\sum_a \int d^3v \, \vect{v} v^2 C_s \left[f_{s,\,\mathrm{w/o}\,\Phi_1}, f_{a,\,\mathrm{w/o}\,\Phi_1}\right]$ is the energy-weighted friction force acting on the species, where the impact of $\Phi_1$ is neglected.\newline
 For different options of $\textcolor{magenta}{Y}$ see flux-surface labels (section~\ref{sec:radialCoordinates}).}

### 2019-03-08 fortran/version3/writeHDF5Output.F90

@@ -323,10 +323,10 @@ contains
        call writeHDF5Field("classicalParticleFluxNoPhi1_rHat", classicalParticleFluxNoPhi1_rHat, dspaceIDForSpecies, dimForSpecies, "Classical particle flux projected onto \nabla rHat, calculcated, with Phi1 = 0.")
        call writeHDF5Field("classicalParticleFluxNoPhi1_rN", classicalParticleFluxNoPhi1_rN, dspaceIDForSpecies, dimForSpecies, "Classical particle flux projected onto \nabla rN, calculcated, with Phi1 = 0.")
        !!Added by SB 2019-01!!
-       call writeHDF5Field("classicalHeatFluxNoPhi1_psiHat", classicalHeatFluxNoPhi1_psiHat, dspaceIDForSpecies, dimForSpecies, "Classical heat flux projected onto \nabla psiHat, calculcated with Phi1 = 0.")
-       call writeHDF5Field("classicalHeatFluxNoPhi1_psiN", classicalHeatFluxNoPhi1_psiN, dspaceIDForSpecies, dimForSpecies, "Classical heat flux projected onto \nabla psiN, calculcated, with Phi1 = 0.")
-       call writeHDF5Field("classicalHeatFluxNoPhi1_rHat", classicalHeatFluxNoPhi1_rHat, dspaceIDForSpecies, dimForSpecies, "Classical heat flux projected onto \nabla rHat, calculcated, with Phi1 = 0.")
-       call writeHDF5Field("classicalHeatFluxNoPhi1_rN", classicalHeatFluxNoPhi1_rN, dspaceIDForSpecies, dimForSpecies, "Classical heat flux projected onto \nabla rN, calculcated, with Phi1 = 0.")
+       call writeHDF5Field("classicalHeatFluxNoPhi1_psiHat", classicalHeatFluxNoPhi1_psiHat, dspaceIDForSpecies, dimForSpecies, "Classical heat flux projected onto \nabla psiHat, calculcated with Phi1 = 0. Updated with correct normalization: 2019-03-08.")
+       call writeHDF5Field("classicalHeatFluxNoPhi1_psiN", classicalHeatFluxNoPhi1_psiN, dspaceIDForSpecies, dimForSpecies, "Classical heat flux projected onto \nabla psiN, calculcated, with Phi1 = 0. Updated with correct normalization: 2019-03-08.")
[…]

### 2019-04-30 doc/manual/version3/inputParameters.tex
Added adjoint and ambipolar options to manual.

@@
 
-{\ttfamily RHSMode} = 3: Solve for the 2$\times$2 monoenergetic transport coefficients. When this option is chosen, \Nx~is set to 1, only 1 species is used and \parlink{includePhi1} is set to \false
+{\ttfamily RHSMode} = 3: Solve for the 2$\times$2 monoenergetic transport coefficients.
+When this option is chosen, \Nx~is set to 1, only 1 species is used and \parlink{includePhi1} is set to \false \\
+
+{\ttfamily RHSMode} = 4: Solve the adjoint equation to compute derivatives at constant $E_r$.
+Options related to the adjoint solve are detailed in section \ref{sec:adjointOptions}.
+{\ttfamily RHSMode} = 4 must be used with \\ \parlink{includePhi1}=\false, \parlink{EParallelHat}\texttt{=0}, \parlink{magneticDriftScheme}\texttt{>0}, \\ \parlink{constraintScheme}\texttt{=-1} or 1, \parlink{collisionOperator}\texttt{=0}, and \parlink{geometryScheme}\texttt{/=5}.
+\\
+
+{\ttfamily RHSMode} = 5: Solve the adjoint equation to compute derivatives at the ambipolar $E_r$.
+Options related to the adjoint solve are detailed in section \ref{sec:adjointOptions}.
+{\ttfamily RHSMode} = 5 must be used with \\ \parlink{includePhi1}=\false, \parlink{EParallelHat}\texttt{=0}, \parlink{magneticDriftScheme}\texttt{>0}, \\ \parlink{constraintScheme}\texttt{=-1} or 1, \parlink{collisionOperator}\texttt{=0}, and \parlink{geometryScheme}\texttt{/=5}.
+\\
 }
@@
 
+\myhrule
+
+\param{ambipolarSolve}
+{Boolean}
+{\false}
+{When \parlink{RHSMode}==1, 4, or 5.}
+{When \true, a root finding method will be used to find the ambipolar \parlink{Er}.
+All output parameters will be included at the ambipolar \parlink{Er}.
+Options related to this parameter are \parlink{NEr\_ambipolarSolve}, \parlink{Er\_search\_tolerance\_dx}, \parlink{Er\_search\_tolerance\_f}, \\ \parlink{ambipolarSolveOption}, \parlink{Er\_min}, and \parlink{Er\_max}.}
+
+\myhrule
+
+\param{NEr\_ambipolarSolve}
+{integer}
+{20}
+{When \parlink{ambipolarSolve} == \true.}
+{Maximum number of solves to allow while finding the ambipolar \parlink{Er}.}
+
+\myhrule
+
+\param{Er\_search\_tolerance\_dx}
+{real}
+{1.d-8}
+{When \parlink{ambipolarSolve} == \true and \parlink{ambipolarSolveOption}/=2.}
+{Tolerance used for ambipolar solve.
+If step size in \parlink{Er} falls below \parlink{Er\_search\_tolerance\_dx}, the ambipolar solve completes successfully.
+Note that this is not used for \parlink{ambipolarSolveOption}==2.}
+
+\myhrule
+
+\param{Er\_search\_tolerance\_f}
+{real}
+{1.d-10}
+{When \parlink{ambipolarSolve} == \true.}
+{Tolerance used for ambipolar solve. If the radial current,
+\begin{gather}
+J_r = \sum_s \texttt{Zs}(s) \texttt{particleFlux\_vm\_rN}(s),
+\end{gather}
+falls below \parlink{Er\_search\_tolerance\_f}, the ambipolar solve completes successfully.}
+
+\myhrule
+
+\param{ambipolarSolveOption}
+{integer}
+{1}
+{When \parlink{ambipolarSolve} == \true}
+{Indicates which root solving algorithm to use for ambipolar solve. \\
+\parlink{ambipolarSolveOption}==1: a Newton bisection method is used \\ 
+\parlink{ambipolarSolveOption}==2: a Brent method is used \\
+\parlink{ambipolarSolveOption}==3: a Newton method is used.
+
+If \parlink{ambipolarSolveOption}==1 or 3, an adjoint method will used to compute the $\partial J_r/\partial E_r$, so must be used with \parlink{includePhi1}==\false, \parlink{EParallelHat}==0, \parlink{magneticDriftScheme}$>0$, \parlink{constraintScheme}==-1 or 1, and \parlink{collisionOperator}==0.}
+
+\myhrule
+
+\param{Er\_min}
+{real}
+{-100}
+{When \parlink{ambipolarSolve} == \true and \parlink{ambipolarSolveOption} /= 3. }
+{Minimum value of \parlink{Er} used to bracket the ambipolar root. }
+
+\myhrule
+
+\param{Er\_max}
+{real}
+{100}
+{When \parlink{ambipolarSolve} == \true and \parlink{ambipolarSolveOption} /= 3. }
+{Maximum  value of \parlink{Er} used to bracket the ambipolar root. }
+
+\myhrule
+
 \section{The {\ttfamily geometryParameters} namelist}
@@
 The file is assumed to be stellarator-\emph{asymmetric}.
+
+{\ttfamily geometryScheme}==13: Read Boozer geometry from the {\ttfamily geometryParamters} namelist.
+Here the geometry parameters are specified by \parlink{Nperiods}, \parlink{boozer\_bmnc}, \parlink{GHat}, \parlink{IHat}, and \parlink{iota}.
+This has been implemented for use with STELLOPT for direct optimization of the Boozer spectrum.
 }
@@
   (otherwise just proportional to $B$).}
+
+  \myhrule
+
+\param{boozer\_bmnc}
+{2D array of reals}
+{0}
+{Only when \parlink{geometryScheme} == 13. }
+{Fourier harmonics describing the magnetic field in Boozer coordinates,
+\begin{gather}
+B(\theta_B,\zeta_B) = \sum_{m,n}\mathrm{\texttt{boozer\_bmnc}}(m,n) \cos(m\theta_B-n\mathrm{\texttt{Nperiods}} \zeta_B) + \mathrm{\texttt{boozer\_bmnc}}(m,n) \sin(m\theta_B-n\mathrm{\texttt{Nperiods}} \zeta_B),
+\end{gather}
+The first index of \texttt{ Boozer\_bmnc} corresponds to $m \in $[0,40], and the second corresponds to $n \in $[-40,40]. Here \parlink{Nperiods} can be specified in the input file.
+}
+
+\myhrule
+
+\param{boozer\_bmns}
+{2D array of reals}
+{0}
+{Only when \parlink{geometryScheme} == 13. }
+{Fourier harmonics describing the magnetic field in Boozer coordinates,
+\begin{gather}
+B(\theta_B,\zeta_B) = \sum_{m,n}\mathrm{\texttt{boozer\_bmnc}}(m,n) \cos(m\theta_B-n\mathrm{\texttt{Nperiods}} \zeta_B) + \mathrm{\texttt{boozer\_bmnc}}(m,n) \sin(m\theta_B-n\mathrm{\texttt{Nperiods}} \zeta_B),
+\end{gather}
+The first index of \texttt{ Boozer\_bmns} corresponds to $m \in $[0,40], and the second corresponds to $n \in $[-40,40]. Here \parlink{Nperiods} can be specified in the input file.
+}
+
+\myhrule
+
+\param{Nperiods}
+{integer}
+{0}
+{Only when \parlink{geometryScheme} == 13.}
+{Boozer geometry will assume number of periods is \texttt{Nperiods}.}
+
+\myhrule
 
@@
 
+\section{The \texttt{adjointOptions} namelist}
+\label{sec:adjointOptions}
+
+This namelist allows for derivatives of the outputs to be computed with respect to geometric parameters. If \parlink{RHSMode}\texttt{<4}, this namelist will not be read.
+
+\param{adjointBootstrapOption}
+{Boolean}
+{\false}
+{When \parlink{RHSMode} == 4 or 5}
+{If this parameter is \true, the derivative of the bootstrap current (\parlink{FSABjHatOverRootFSAB2}) is computed with respect to geometry parameters: \parlink{GHat}, \parlink{IHat}, \parlink{iota}, and \parlink{bmnc}.}
+
+\myhrule
+
+\param{adjointRadialCurrentOption}
+{Boolean}
+{\false}
+{When \parlink{RHSMode} == 4 or 5}
+{If this parameter is \true the derivative of the radial current,
+\begin{gather}
+J_r = \sum_s \texttt{Zs}(s) \texttt{particleFlux\_vm\_rN}(s),
+\end{gather}
+is computed with respect to geometry parameters: \parlink{GHat}, \parlink{IHat}, \parlink{iota}, and \parlink{bmnc}.
+Here \parlink{Zs} and \parlink{particleFlux\_vm\_rN} can be found elsewhere in this document.}
+
+\myhrule
+
+\param{adjointTotalHeatFluxOption}
+{Boolean}
+{\false}
+{When \parlink{RHSMode} == 4 or 5}
+{If this parameter is \true, the derivative of the total heat flux,
+\begin{gather}
+Q = \sum_s \texttt{heatFlux\_vm\_rN}(s),
+\end{gather}
+is computed with respect to geometry parameters: \parlink{GHat}, \parlink{IHat}, \parlink{iota}, and \parlink{bmnc}. Here \parlink{heatFlux\_vm\_rN} can be found elsewhere in this document.}
+
+\myhrule
+
+\param{adjointHeatFluxOption}
+{1D array of booleans}
+{\false}
+{When \parlink{RHSMode} == 4 or 5}
+{This array is the same length as {\ttfamily nHats} and other species parameters.
+If parameter is \true, the derivative of the heat flux (\parlink{heatFlux\_vm\_rN}) will be computed for this species with respect to geometry parameters: \parlink{GHat}, \parlink{IHat}, \parlink{iota}, and \parlink{bmnc}.}
+
+\myhrule
+
+\param{adjointParticleFluxOption}
+{1D array of booleans}
+{\false}
+{When \parlink{RHSMode} == 4 or 5}
+{This array is the same length as {\ttfamily nHats} and other species parameters.
+If parameter is \true, the derivative of the particle flux (\parlink{particleFlux\_vm\_rN}) will be computed for this species with respect to geometry parameters: \parlink{GHat}, \parlink{IHat}, \parlink{iota}, and \parlink{bmnc}.}
+
+\myhrule
+
+\param{adjointParallelFlowOption}
+{1D array of booleans}
+{\false}
+{When \parlink{RHSMode} == 4 or 5}
+{This array is the same length as {\ttfamily nHats} and other species parameters.
+If parameter is \true, the derivative of the parallel flow (\parlink{FSABVelocityUsingFSADensityOverRootFSAB2}) will be computed for this species with respect to geometry parameters: \parlink{GHat}, \parlink{IHat}, \parlink{iota}, and \parlink{bmnc}.}
+
+\myhrule
+
+\param{discreteAdjointOption}
+{boolean}
+{\true}
+{When \parlink{RHSMode} == 4 or 5}
+{When \parlink{RHSMode} is \true, the discrete adjoint technique is used.
+If \false, the continuous technique is used.
+Both should give the same result within discretization error, but the discrete approach should be slightly faster.
+}
+
+\myhrule
+
+\param{nMaxAdjoint}
+{integer}
+{0}
+{When \parlink{RHSMode} == 4 or 5}
+{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \\ \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \\ \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \\ \parlink{adjointParallelFlowOption} will computed for Fourier modes $|n| \leq $ \texttt{nMaxAdjoint}.
+}
+
+\myhrule
+
+\param{mMaxAdjoint}
+{integer}
+{0}
+{When \parlink{RHSMode} == 4 or 5}
+{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \\ \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \\ \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \\ \parlink{adjointParallelFlowOption} will computed with respect to \parlink{bmnc} for Fourier modes $m \leq $ \texttt{mMaxAdjoint}.
+}
+
+\myhrule
+
+\param{nMinAdjoint}
+{integer}
+{0}
+{When \parlink{RHSMode} == 4 or 5}
+{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \\ \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \\ \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \\ \parlink{adjointParallelFlowOption} will computed with respect to \parlink{bmnc} for Fourier modes $|n| \geq $ \texttt{nMaxAdjoint}.
+}
+
+\myhrule
+
+\param{mMinAdjoint}
+{integer}
+{0}
+{When \parlink{RHSMode} == 4 or 5}
+{Derivatives of outputs specified by \parlink{adjointBootstrapOption}, \\ \parlink{adjointRadialCurrentOption}, \parlink{adjointTotalHeatFluxOption}, \\ \parlink{adjointHeatFluxOption}, \parlink{adjointParticleFluxOption}, and \\ \parlink{adjointParallelFlowOption} will computed with respect to \parlink{bmnc} for Fourier modes $m \geq $ \texttt{mMaxAdjoint}.
+}
+
+\myhrule
+
+\param{debugAdjoint}
+{boolean}
+{\false}
+{When \parlink{RHSMode} == 4 or 5. }
+{If \texttt{.true.}, then derivatives computed with the adjoint method will be compared with forward difference derivatives with step size set by \parlink{deltaLambda}.}
+
+\myhrule
+
+\param{deltaLambda}
+{real}
+{1.d-4}
+{When \parlink{deltaLamda} == \texttt{.true.}}
+{The step size used for forward difference derivatives will be \texttt{deltaLambda} multiplied by the quantity that is being varied.
+For example, the step size for a Boozer harmonic, $B_{mn}$, would be $\Delta B_{mn} = $\texttt{deltaLambda}$B_{mn}$.}
+
+\myhrule
+
 \section{Directives for \sfincsScan}
